---
title: "IRW Collections"
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(stringr)
library(glue)
library(htmltools)
library(irw)

reg <- irw_collections()
mem <- irw_collection_members()

##Exact-instrument collections (irw#1712): kind "instrument", curated table by table.
##They render as member lists below rather than as cards, since the point of them is
##which tables are in and how each was identified. Empty until they are uploaded.
exact <- reg$collection[reg$kind == "instrument" & reg$coverage == "curated-only"]

KIND_META <- tibble::tribble(
  ~kind,        ~heading,                     ~blurb,
  "design",     "Study &amp; instrument design",
    "How the study was run, and what structure the data carries. Derived from the columns each table actually has, so these cover every documented table.",
  "instrument", "Instrument family",
    "Which measure was administered. These group together tables that record the same instrument under different names &mdash; 63 Big Five tables are spread across more than forty spellings of the construct name.",
  "construct",  "Construct",
    "What is being measured, across instruments."
)
```

Collections are labelled groupings of IRW tables. Ask for a *type* of data ---
randomized trials, Big Five inventories, anything with a Q matrix --- and get
back the tables, ready to fetch.

A table belongs to as many collections as apply: `frac20` is in both
`math` and `q_matrix`.

```{r}
#| label: coverage-note
#| echo: false
#| results: asis
## The counts come from the note 10_collections.R appends to every
## tagged-subset-only definition: "(searched the N of M tables that have a
## construct name; ...)". Parsed here rather than hardcoded -- the last
## hardcoded pair (2,251 of 3,650) went stale within a month (irw#1719, 17.3).
## If no definition carries that shape, the prose below drops the numbers
## rather than showing old ones.
cov_m <- str_match(reg$definition,
                   "searched the ([0-9,]+) of ([0-9,]+) tables that have a construct name")
cov_m <- cov_m[!is.na(cov_m[, 1]), , drop = FALSE]
n_searched <- if (nrow(cov_m)) as.integer(gsub(",", "", cov_m[1, 2])) else NA_integer_
n_total    <- if (nrow(cov_m)) as.integer(gsub(",", "", cov_m[1, 3])) else NA_integer_
fmt <- function(n) format(n, big.mark = ",")
searched_txt <- if (is.na(n_searched)) "the tables" else
  glue("the **{fmt(n_searched)} of {fmt(n_total)}** tables")
n_part <- sum(reg$coverage == "tagged-subset-only")
cat(glue(
  '::: {{.callout-warning appearance="simple"}}\n',
  '## Please check a collection before relying on it\n\n',
  '**Membership is our judgment, and you should verify it for your own purposes.** ',
  'Most collections are assigned by a rule --- the presence of a column, or a pattern ',
  'matched against the construct name each table was annotated with. A pattern over ',
  'free text will sometimes catch a table you would exclude, and miss one you would ',
  'include. `continuous_response` and the exact-instrument collections are assigned by ',
  'hand, table by table, which makes them a judgment of a different kind rather than a ',
  'safer one: most exact-instrument members were identified from wording IRW supplied, ',
  'not from the study\'s own materials (see [Exact instruments](#exact-instruments)). ',
  'Every membership records the basis on which it was decided --- see ',
  '`irw_collection_members()` --- so you can inspect a collection and form your own view; ',
  'please do so before using one as the sample for an analysis.\n\n',
  '**Coverage is also uneven.** Collections built from the tag annotations searched only ',
  '{searched_txt} whose tags include a construct name, and that field is much sparser in ',
  'the newer warehouses. {n_part} of the ',
  '{nrow(reg)} collections below are marked `tagged-subset-only` for that reason: they are ',
  'useful starting points, but they are not exhaustive, and they lean toward older tables. ',
  'Treat them as "the ones we know of", not "all of them". The rest searched every ',
  'documented table.\n\n',
  'Corrections are welcome --- open an issue on ',
  '[ben-domingue/irw](https://github.com/ben-domingue/irw/issues).\n',
  ':::\n\n'
))
```

## Using a collection

::: panel-tabset

## R

```{r}
#| eval: false
#| echo: true
library(irw)

irw_collections()                    # what exists: kind, definition, coverage, size
tabs <- irw_collection("depression") # 70 table names, and a note on coverage

# Narrow to a modellable corpus. Note irw_filter()'s default density = c(0.5, 1),
# which is why 70 becomes 32 rather than the 41 that pass on sample size alone.
tabs <- irw_filter(collection = "depression", n_participants = c(500, Inf))

dat <- irw_fetch(tabs)               # named list, one long-format table each
irw_save_bibtex(tabs, output_file = "depression.bib")

# Cross-collection questions -- what no single filter argument reaches.
# Math assessments administered in booklets: linking and DIF work.
intersect(irw_collection("math"), irw_collection("multistage"))

# Note c(...) is a UNION, not an intersection:
irw_filter(collection = c("rct", "response_time"))

irw_collection_members(tables = "frac20")   # which collections is this table in?

# Exact-instrument collections: keep only tables whose own item text was checked.
m <- irw_collection_members(collection = "phq9")
m[startsWith(m$basis, "wording"), ]
```

## Python

```{python}
#| eval: false
#| echo: true
#| python.reticulate: false
import irw

irw.collections()
tabs = irw.collection("depression")

tabs = irw.filter(collection="depression", n_participants=[500, None])
dat  = irw.fetch(tabs)               # dict keyed by table name

sorted(set(irw.collection("math", quiet=True)) &
       set(irw.collection("multistage", quiet=True)))

irw.filter(collection=["rct", "response_time"])   # union, not intersection
irw.collection_members(tables="frac20")

m = irw.collection_members(collection="phq9")
m[m.basis.str.startswith("wording")]
```

:::

```{r}
#| label: cards
#| echo: false
#| results: asis
cat('
<style>
.coll-grid { display: grid; grid-template-columns: repeat(2, 1fr); gap: 1rem; margin-bottom: 2rem; }
@media (max-width: 768px) { .coll-grid { grid-template-columns: 1fr; } }
.coll-card { border: 1px solid #dee2e6; border-left: 4px solid #2780e3; border-radius: 6px;
  padding: 1rem 1.25rem; background: #fff; transition: box-shadow 0.15s;
  display: flex; flex-direction: column; }
.coll-card.partial { border-left-color: #c98a30; }
.coll-card:hover { box-shadow: 0 2px 12px rgba(0,0,0,0.08); }
.coll-name { font-family: var(--bs-font-monospace, monospace); font-size: 0.95rem;
  font-weight: 600; color: #212529; margin-bottom: 0.15rem; }
.coll-label { font-size: 0.8rem; color: #6c757d; margin-bottom: 0.5rem; }
.coll-desc { font-size: 0.85rem; color: #495057; flex: 1; line-height: 1.5; margin-bottom: 0.75rem; }
.coll-foot { margin-top: auto; padding-top: 0.6rem; display: flex; flex-wrap: wrap;
  gap: 0.5rem; align-items: baseline; }
.coll-n { font-size: 0.85rem; font-weight: 600; color: #212529; }
.coll-cov { font-size: 0.72rem; padding: 0.12rem 0.5rem; border-radius: 100px;
  background: #e8f1fb; color: #1a5c9e; white-space: nowrap; }
.coll-cov.partial { background: #f6ead6; color: #8f5410; }
.coll-cov.curated { background: #eceef0; color: #495057; }
.ex-controls { display: flex; flex-wrap: wrap; gap: 0.6rem; align-items: center; margin: 0.5rem 0 0.75rem; }
.ex-controls input { flex: 1 1 220px; min-width: 0; padding: 0.35rem 0.6rem; border: 1px solid #dee2e6; border-radius: 6px; }
.ex-chips { display: flex; gap: 0.35rem; flex-wrap: wrap; }
.ex-chip { font-size: 0.8rem; padding: 0.2rem 0.7rem; border-radius: 100px; border: 1px solid #dee2e6; background: #fff; color: #212529; }
.ex-chip[aria-pressed="true"] { background: #2f6b5c; border-color: #2f6b5c; color: #fff; }
.ex-list { margin-bottom: 2rem; border-top: 1px solid #dee2e6; }
.ex-coll { border-bottom: 1px solid #dee2e6; }
.ex-coll summary { cursor: pointer; display: grid; grid-template-columns: minmax(0, 1fr) auto; gap: 0.35rem 1rem;
  align-items: baseline; padding: 0.7rem 0; list-style: none; }
.ex-coll summary::-webkit-details-marker { display: none; }
.ex-title { display: flex; flex-wrap: wrap; gap: 0.6rem; align-items: baseline; min-width: 0; }
.ex-label { font-weight: 600; }
.ex-count { font-size: 0.85rem; color: #6c757d; white-space: nowrap; font-variant-numeric: tabular-nums; }
.ex-count b { color: #212529; }
.ex-bar { grid-column: 1 / -1; display: flex; height: 6px; border-radius: 3px; overflow: hidden; background: #eceef0; }
.ex-bar span { display: block; height: 100%; }
.ex-wording { background: #2f6b5c; } .ex-supplied { background: #7b93b8; } .ex-name { background: #c9a86a; }
.ex-body { padding: 0 0 1rem; }
.ex-what { font-size: 0.85rem; color: #495057; margin: 0 0 0.5rem; }
.ex-tbl { overflow-x: auto; }
.ex-tbl table { width: 100%; font-size: 0.85rem; }
.ex-tbl th { font-size: 0.72rem; text-transform: uppercase; letter-spacing: 0.04em; color: #6c757d; font-weight: 600; }
.ex-tbl td, .ex-tbl th { padding: 0.3rem 0.8rem 0.3rem 0; border-bottom: 1px solid #eceef0; vertical-align: top; }
.ex-pill { display: inline-flex; gap: 0.4rem; align-items: center; white-space: nowrap; }
.ex-sw { width: 9px; height: 9px; border-radius: 2px; display: inline-block; }
.ex-note { font-size: 0.78rem; background: #f8f9fa; border: 1px solid #dee2e6; border-radius: 4px; padding: 0 0.35rem; white-space: nowrap; }
</style>
')

BASES <- c("wording", "supplied", "name")
exact_instruments <- function() {
  ex <- reg[reg$collection %in% exact, ]
  m  <- mem[mem$collection %in% exact, ]
  m$b    <- str_trim(str_extract(m$basis, "^[^;]+"))
  m$note <- str_trim(str_match(m$basis, ";\\s*(.*)$")[, 2])
  ex <- ex[order(-ex$n_tables), ]
  cat(glue('\n### Exact instruments {{#exact-instruments}}\n\n',
    'Each of these lists the tables that administered one specific instrument: the full scale, ',
    'at least 80% of its items, or one whole subscale. Short forms, lengthened adaptations and ',
    'reworded versions are left out. Every membership was decided table by table, and its ',
    '`basis` says how the table was identified:\n\n',
    '- <span class="ex-sw ex-wording"></span> **wording**: the study\'s own item text matches the instrument. Identity is observed.\n',
    '- <span class="ex-sw ex-supplied"></span> **supplied**: IRW identified the instrument and supplied its canonical wording, so a ',
    'wording match only confirms that identification. Identity is asserted.\n',
    '- <span class="ex-sw ex-name"></span> **name**: there is no usable English item text; only the table\'s metadata names the instrument.\n\n',
    'A note after the basis marks a partial table (`partial: 18/20 items`) or a single subscale ',
    '(`subscale: anxiety (7/21)`). For the strictly verified set, keep members whose basis starts ',
    'with `wording`. The method is written up in the ',
    '[collections README](https://github.com/ben-domingue/irw/blob/main/src/collections/README).\n\n'))
  rows_html <- vapply(seq_len(nrow(ex)), function(i) {
    r  <- ex[i, ]
    mm <- m[m$collection == r$collection, ]
    mm <- mm[order(match(mm$b, BASES), tolower(mm$table)), ]
    n  <- nrow(mm)
    cnt <- vapply(BASES, function(b) sum(mm$b == b), 0L)
    bar <- paste0(vapply(BASES, function(b) if (cnt[[b]]) glue(
      '<span class="ex-{b}" style="width:{round(100 * cnt[[b]] / n, 2)}%" title="{b}: {cnt[[b]]}"></span>') else "", ""),
      collapse = "")
    what <- str_match(r$definition, "^Tables that administered (.*?): ")[, 2]
    trs <- paste0(glue_data(mm,
      '<tr data-b="{b}" data-t="{tolower(table)}"><td><a href="tables/{table}/"><code>{table}</code></a></td>',
      '<td><span class="ex-pill"><span class="ex-sw ex-{b}"></span>{b}</span></td>',
      '<td>{ifelse(is.na(note), "", paste0(\'<span class="ex-note">\', htmlEscape(note), "</span>"))}</td></tr>'),
      collapse = "")
    glue('<details class="ex-coll" data-slug="{r$collection}">',
      '<summary><span class="ex-title"><span class="ex-label">{htmlEscape(r$label)}</span>',
      '<code>{r$collection}</code></span>',
      '<span class="ex-count"><b>{n}</b> tables &middot; {cnt[["wording"]]} wording</span>',
      '<span class="ex-bar" aria-hidden="true">{bar}</span></summary>',
      '<div class="ex-body">{if (!is.na(what)) paste0("<p class=\\"ex-what\\">", htmlEscape(what), "</p>") else ""}',
      '<div class="ex-tbl"><table><thead><tr><th>Table</th><th>Basis</th><th>Note</th></tr></thead>',
      '<tbody>{trs}</tbody></table></div></div></details>')
  }, "")
  cat('<div class="ex-controls"><input id="ex-q" type="search" placeholder="Search table names" aria-label="Search table names">',
      '<div class="ex-chips" role="group" aria-label="Filter by basis">',
      paste0(vapply(c("all", BASES), function(b) glue(
        '<button type="button" class="ex-chip" data-b="{b}" aria-pressed="{tolower(b == "all")}">{b}</button>'), ""),
        collapse = ""),
      '</div></div>\n<div class="ex-list">', paste0(rows_html, collapse = ""), '</div>\n\n', sep = "")
  cat('<script>
(function () {
  var q = "", basis = "all";
  function apply() {
    document.querySelectorAll(".ex-coll").forEach(function (d) {
      var shown = 0;
      d.querySelectorAll("tbody tr").forEach(function (tr) {
        var ok = (basis === "all" || tr.dataset.b === basis) && tr.dataset.t.indexOf(q) !== -1;
        tr.hidden = !ok; if (ok) shown++;
      });
      d.hidden = shown === 0;
      if (q) d.open = true;
    });
  }
  document.getElementById("ex-q").addEventListener("input", function (e) { q = e.target.value.trim().toLowerCase(); apply(); });
  document.querySelectorAll(".ex-chip").forEach(function (b) {
    b.addEventListener("click", function () {
      basis = b.dataset.b;
      document.querySelectorAll(".ex-chip").forEach(function (x) { x.setAttribute("aria-pressed", x === b); });
      apply();
    });
  });
})();
</script>\n\n')
}

for (k in KIND_META$kind) {
  meta <- KIND_META[KIND_META$kind == k, ]
  rows <- reg[reg$kind == k & !reg$collection %in% exact, ]
  if (nrow(rows) == 0) next

  cat(glue('\n## {meta$heading}\n\n{meta$blurb}\n\n'))
  cat('<div class="coll-grid">\n')

  for (i in seq_len(nrow(rows))) {
    r <- rows[i, ]
    partial <- r$coverage == "tagged-subset-only"
    ##Strip the denominator clause the pipeline appends -- it is shown as a
    ##chip below instead, and reads as clutter inside the prose.
    desc <- str_squish(str_remove(r$definition, "\\s*\\(searched the.*?\\)\\s*$"))
    cov_txt <- dplyr::case_when(
      r$coverage == "tagged-subset-only" & !is.na(n_searched) ~ glue("of {fmt(n_searched)} searched"),
      r$coverage == "tagged-subset-only" ~ "partial search",
      r$coverage == "curated-only"       ~ "hand-curated",
      !is.na(n_total)                    ~ glue("of all {fmt(n_total)}"),
      TRUE                               ~ "of all documented"
    )
    cov_cls <- dplyr::case_when(
      r$coverage == "tagged-subset-only" ~ "partial",
      r$coverage == "curated-only"       ~ "curated",
      TRUE                               ~ ""
    )
    cat(as.character(HTML(glue(
      '<div class="coll-card{if (partial) " partial" else ""}">',
      '<div class="coll-name">{r$collection}</div>',
      '<div class="coll-label">{r$label}</div>',
      '<div class="coll-desc">{htmlEscape(desc)}</div>',
      '<div class="coll-foot">',
      '<span class="coll-n">{r$n_tables} tables</span>',
      '<span class="coll-cov {cov_cls}">{cov_txt}</span>',
      '</div></div>\n'
    ))))
  }
  cat('</div>\n\n')
  if (k == "instrument" && length(exact)) exact_instruments()
}
```

## How collections are built

Each collection is one row in a [version-controlled
registry](https://github.com/ben-domingue/irw/blob/main/src/collections/registry.csv)
carrying its definition and the rule that produces its membership --- the
presence of a column (`treat` for `rct`, `qmatrix*` for `q_matrix`), or a
pattern over the recorded construct name. `continuous_response` is assembled by
hand, because the available signals disagree too often for a rule to settle it.
The exact-instrument collections are also assembled by hand: a script proposes
candidates and checks their item text against the instrument's wording, and
each membership is then decided table by table (see
[issue #1712](https://github.com/ben-domingue/irw/issues/1712)).

Adding a collection is therefore a change to data, not to code, and every
membership records the basis on which it was decided. That last point is
deliberate: a rule-assigned membership and a hand-assigned one carry different
weight, and both are open to being wrong, so the record of *how* a table got
into a collection travels with the collection itself. See
[issue #1633](https://github.com/ben-domingue/irw/issues/1633).
