---
title: "IRW Collections"
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(stringr)
library(glue)
library(htmltools)
library(irw)

reg <- irw_collections()

KIND_META <- tibble::tribble(
  ~kind,        ~heading,                     ~blurb,
  "design",     "Study &amp; instrument design",
    "How the study was run, and what structure the data carries. Derived from the columns each table actually has, so these cover every documented table.",
  "instrument", "Instrument family",
    "Which measure was administered. These group together tables that record the same instrument under different names &mdash; 63 Big Five tables are spread across more than forty spellings of the construct name.",
  "construct",  "Construct",
    "What is being measured, across instruments."
)
```

Collections are labelled groupings of IRW tables. Ask for a *type* of data ---
randomized trials, Big Five inventories, anything with a Q matrix --- and get
back the tables, ready to fetch.

A table belongs to as many collections as apply: `frac20` is in both
`math` and `q_matrix`.

```{r}
#| label: coverage-note
#| echo: false
#| results: asis
n_tagged <- 2251L
n_total  <- 3650L
n_part <- sum(reg$coverage == "tagged-subset-only")
cat(glue(
  '::: {{.callout-warning appearance="simple"}}\n',
  '## Please check a collection before relying on it\n\n',
  '**Membership is our judgment, and you should verify it for your own purposes.** ',
  'Most collections are assigned by a rule --- the presence of a column, or a pattern ',
  'matched against the construct name each table was annotated with. A pattern over ',
  'free text will sometimes catch a table you would exclude, and miss one you would ',
  'include. `continuous_response` is assigned by hand, which makes it a judgment of a ',
  'different kind rather than a safer one. Nothing here has been checked table by table. ',
  'Every membership records the basis on which it was decided --- see ',
  '`irw_collection_members()` --- so you can inspect a collection and form your own view; ',
  'please do so before using one as the sample for an analysis.\n\n',
  '**Coverage is also uneven.** Collections built from the tag annotations searched only ',
  'the **{format(n_tagged, big.mark=",")} of {format(n_total, big.mark=",")}** tables that ',
  'carry tags, and tag coverage is much lower in the newer warehouses. {n_part} of the ',
  '{nrow(reg)} collections below are marked `tagged-subset-only` for that reason: they are ',
  'useful starting points, but they are not exhaustive, and they lean toward older tables. ',
  'Treat them as "the ones we know of", not "all of them". The rest searched every ',
  'documented table.\n\n',
  'Corrections are welcome --- open an issue on ',
  '[ben-domingue/irw](https://github.com/ben-domingue/irw/issues).\n',
  ':::\n\n'
))
```

## Using a collection

::: panel-tabset

## R

```{r}
#| eval: false
#| echo: true
library(irw)

irw_collections()                    # what exists: kind, definition, coverage, size
tabs <- irw_collection("depression") # 70 table names, and a note on coverage

# Narrow to a modellable corpus. Note irw_filter()'s default density = c(0.5, 1),
# which is why 70 becomes 32 rather than the 41 that pass on sample size alone.
tabs <- irw_filter(collection = "depression", n_participants = c(500, Inf))

dat <- irw_fetch(tabs)               # named list, one long-format table each
irw_save_bibtex(tabs, output_file = "depression.bib")

# Cross-collection questions -- what no single filter argument reaches.
# Math assessments administered in booklets: linking and DIF work.
intersect(irw_collection("math"), irw_collection("multistage"))

# Note c(...) is a UNION, not an intersection:
irw_filter(collection = c("rct", "response_time"))

irw_collection_members(tables = "frac20")   # which collections is this table in?
```

## Python

```{python}
#| eval: false
#| echo: true
#| python.reticulate: false
import irw

irw.collections()
tabs = irw.collection("depression")

tabs = irw.filter(collection="depression", n_participants=[500, None])
dat  = irw.fetch(tabs)               # dict keyed by table name

sorted(set(irw.collection("math", quiet=True)) &
       set(irw.collection("multistage", quiet=True)))

irw.filter(collection=["rct", "response_time"])   # union, not intersection
irw.collection_members(tables="frac20")
```

:::

```{r}
#| label: cards
#| echo: false
#| results: asis
cat('
<style>
.coll-grid { display: grid; grid-template-columns: repeat(2, 1fr); gap: 1rem; margin-bottom: 2rem; }
@media (max-width: 768px) { .coll-grid { grid-template-columns: 1fr; } }
.coll-card { border: 1px solid #dee2e6; border-left: 4px solid #2780e3; border-radius: 6px;
  padding: 1rem 1.25rem; background: #fff; transition: box-shadow 0.15s;
  display: flex; flex-direction: column; }
.coll-card.partial { border-left-color: #c98a30; }
.coll-card:hover { box-shadow: 0 2px 12px rgba(0,0,0,0.08); }
.coll-name { font-family: var(--bs-font-monospace, monospace); font-size: 0.95rem;
  font-weight: 600; color: #212529; margin-bottom: 0.15rem; }
.coll-label { font-size: 0.8rem; color: #6c757d; margin-bottom: 0.5rem; }
.coll-desc { font-size: 0.85rem; color: #495057; flex: 1; line-height: 1.5; margin-bottom: 0.75rem; }
.coll-foot { margin-top: auto; padding-top: 0.6rem; display: flex; flex-wrap: wrap;
  gap: 0.5rem; align-items: baseline; }
.coll-n { font-size: 0.85rem; font-weight: 600; color: #212529; }
.coll-cov { font-size: 0.72rem; padding: 0.12rem 0.5rem; border-radius: 100px;
  background: #e8f1fb; color: #1a5c9e; white-space: nowrap; }
.coll-cov.partial { background: #f6ead6; color: #8f5410; }
.coll-cov.curated { background: #eceef0; color: #495057; }
</style>
')

for (k in KIND_META$kind) {
  meta <- KIND_META[KIND_META$kind == k, ]
  rows <- reg[reg$kind == k, ]
  if (nrow(rows) == 0) next

  cat(glue('\n## {meta$heading}\n\n{meta$blurb}\n\n'))
  cat('<div class="coll-grid">\n')

  for (i in seq_len(nrow(rows))) {
    r <- rows[i, ]
    partial <- r$coverage == "tagged-subset-only"
    ##Strip the denominator clause the pipeline appends -- it is shown as a
    ##chip below instead, and reads as clutter inside the prose.
    desc <- str_squish(str_remove(r$definition, "\\s*\\(searched the.*?\\)\\s*$"))
    cov_txt <- dplyr::case_when(
      r$coverage == "tagged-subset-only" ~ glue("of {format(n_tagged, big.mark=',')} tagged"),
      r$coverage == "curated-only"       ~ "hand-curated",
      TRUE                               ~ glue("of all {format(n_total, big.mark=',')}")
    )
    cov_cls <- dplyr::case_when(
      r$coverage == "tagged-subset-only" ~ "partial",
      r$coverage == "curated-only"       ~ "curated",
      TRUE                               ~ ""
    )
    cat(as.character(HTML(glue(
      '<div class="coll-card{if (partial) " partial" else ""}">',
      '<div class="coll-name">{r$collection}</div>',
      '<div class="coll-label">{r$label}</div>',
      '<div class="coll-desc">{htmlEscape(desc)}</div>',
      '<div class="coll-foot">',
      '<span class="coll-n">{r$n_tables} tables</span>',
      '<span class="coll-cov {cov_cls}">{cov_txt}</span>',
      '</div></div>\n'
    ))))
  }
  cat('</div>\n\n')
}
```

## How collections are built

Each collection is one row in a [version-controlled
registry](https://github.com/ben-domingue/irw/blob/main/src/collections/registry.csv)
carrying its definition and the rule that produces its membership --- the
presence of a column (`treat` for `rct`, `qmatrix*` for `q_matrix`), or a
pattern over the recorded construct name. `continuous_response` is the one
collection assembled by hand, because the available signals disagree too often
for a rule to settle it.

Adding a collection is therefore a change to data, not to code, and every
membership records the basis on which it was decided. That last point is
deliberate: a rule-assigned membership and a hand-assigned one carry different
weight, and both are open to being wrong, so the record of *how* a table got
into a collection travels with the collection itself. See
[issue #1633](https://github.com/ben-domingue/irw/issues/1633).
