---
title: "The Competition Data Standard"
---

## The data standard

The IRW competition data standard is meant to describe data derived from pairwise comparisons or competitions. Such data are widely analyzed using Elo or Bradley-Luce-Terry approaches; more information about these approaches is available [here](https://www.annualreviews.org/content/journals/10.1146/annurev-statistics-040722-061813). The key distinction between these data and the conventional IRW data is that there are not `id` and `item` identifiers here but rather identifiers of the two `agents` that are involved in a given competition.

These tables form the competitions [data family](overview.qmd#families) (`source = "comp"`). Despite the name, the family covers more than contests between teams, players or models: it also holds choices between two fixed things, such as essays judged against each other or votes between two named options, where the same named entities are compared again and again. When the things compared are instead put together at random from a set of attributes for each comparison, the data come from a conjoint experiment and belong to the [conjoint family](conjoint_standard.qmd).

- `agent_a` name/id of first competitor
- `agent_b` name/id of second competitor
- `date` Time of the data point in UNIX format.
- `homefield` Indicator of agent that had advantage based on game features (e.g., where game was played, who played first, etc).
- `score_a` Score of `agent_a`.
- `score_b` Score of `agent_b`.
- `winner` Winner of the competition (typically based on a comparison of scores). Values will be `agent_a`, `agent_b` or `draw`.

Additional information that may be predictive of the outcome can be included as additional columns. 

## Accessing the competition data

The competitions data can be accessed from both the R and Python `irw` packages:

::: panel-tabset
## R
```{r}
#| eval: false
#| echo: true
# Install and load the package with `devtools::install_github("itemresponsewarehouse/Rpkg")`
library(irw)

# View available competition data
irw_list_tables(source="comp")

# Fetch competition data
df <- irw_fetch("collegefb_2021and2022", source="comp")
head(df)
```

## Python

```{python}
#| eval: false
#| echo: true
#| python.reticulate: false
# Install package with `pip install irw`
import irw

# View available competition data
irw.list_tables(source="comp")

# Fetch competition data
df = irw.fetch("collegefb_2021and2022", source="comp")
head(df)
```
:::


Documentation related to the tables (including table names as with `collegefb_2021and2022`) can be found [here](https://docs.google.com/spreadsheets/d/1WZZYyVC2cmw8CUJM69qP0F_ZlQjQfdkCZbdsG-8mUrs/edit?usp=sharing).

## What the family holds

{{< include _load-family-meta.qmd >}}

```{r}
#| label: comps-descriptives
comps <- irw_meta_table("comps_metadata")
comps$n_responses <- as.numeric(comps$n_responses)
comps$n_actors    <- as.numeric(comps$n_actors)
n_comps      <- nrow(comps)
n_contests   <- sum(comps$n_responses, na.rm = TRUE)
med_contests <- median(comps$n_responses, na.rm = TRUE)
med_actors   <- median(comps$n_actors, na.rm = TRUE)
n_kaggle     <- sum(grepl("^metakaggle_", comps$table))
share_kaggle <- sum(comps$n_responses[grepl("^metakaggle_", comps$table)], na.rm = TRUE) / n_contests
```

The family holds `r n_comps` tables and `r fmt_big(n_contests)` contests in all. Tables range widely in size: the median table has `r fmt_n(med_contests)` contests between `r fmt_n(med_actors)` competitors, while the largest, Kaggle simulation leaderboards drawn from Meta Kaggle (`r n_kaggle` tables), hold millions of games each and account for `r round(100 * share_kaggle)`% of all contests. Each table's own counts are in `irw_metadata(source = "comp")` (columns `n_responses` and `n_actors`).

```{r}
#| label: fig-comps-sizes
#| fig-cap: "Competition tables by number of contests and number of competitors (both on a log scale)."
#| fig-height: 3.6
#| fig-width: 9
size_bucket <- function(x, breaks, labels) {
  table(cut(x, breaks = breaks, labels = labels, right = FALSE))
}
par(mfrow = c(1, 2))
family_bar(size_bucket(comps$n_responses, c(0, 1e3, 1e4, 1e5, 1e6, Inf),
                       c("<1K", "1K-10K", "10K-100K", "100K-1M", "1M+")),
           xlab = "Contests per table")
family_bar(size_bucket(comps$n_actors, c(0, 10, 100, 1e3, 1e4, Inf),
                       c("<10", "10-100", "100-1K", "1K-10K", "10K+")),
           xlab = "Competitors per table")
```

## An example

Here is a first example wherein we fit an Elo model to data from the NFL.

::: panel-tabset
## R

```{r}
#| echo: true
library(elo)
library(BradleyTerry2)

# Fetch and prepare data
df <- irw::irw_fetch("fivethirtyeight_nfl_1920_2023", source = 'comp')
df <- df[df$season >= 2010 & df$season <= 2019, ]  # the 2010-2019 seasons, playoffs included
df <- df[order(df$date), ]

# ===== Elo ANALYSIS =====
er <- elo.run(score(score_a, score_b) ~ agent_a + agent_b, 
              data = df, k = 20)

# Convert Elo results to long format
elo_results <- as.data.frame(er)
elo_long <- lapply(seq_len(nrow(elo_results)), function(i) {
  data.frame(
    team = c(elo_results$team.A[i], elo_results$team.B[i]),
    elo = c(elo_results$elo.A[i], elo_results$elo.B[i]),
    date = df$date[i]
  )
})
elo_long <- do.call(rbind, elo_long)

# Convert dates to years since the start of the 2010 season
elo_long$year <- (elo_long$date - as.numeric(as.POSIXct("2010-09-01", tz = "UTC")))/(365.25*24*60*60)


elo_by_team <- split(elo_long, elo_long$team)

# ===== BRADLEY-TERRY ANALYSIS =====
# Prepare data for BT model
df_bt <- df[df$winner != 'draw', ]
df_bt$win <- as.numeric(df_bt$winner == 'agent_a')
df_bt$agent_a <- factor(df_bt$agent_a)
df_bt$agent_b <- factor(df_bt$agent_b)

# Fit BT models by season. br = TRUE uses bias-reduced estimation: a team that
# never won in a season (Cleveland in 2017) has no finite maximum-likelihood
# strength, and this keeps its estimate on the same scale as everyone else.
bt_by_season <- lapply(split(df_bt, df_bt$season), function(season_df) {
  mod <- BTm(win, agent_a, agent_b, data = season_df, id = "team", br = TRUE)
  co <- coef(mod)
  names(co) <- gsub("team", "", names(co))
  data.frame(
    season = unique(season_df$season),
    est = co,
    team = names(co),
    row.names = NULL
  )
})
bt_results <- do.call(rbind, bt_by_season)
bt_by_team <- split(bt_results, bt_results$team)

# ===== PLOTTING =====
par(mfrow = c(2, 1), mgp = c(2.5, 0.7, 0), mar = c(4, 4, 2, 1))

# Plot 1: Elo ratings over time
plot(NULL, 
     xlim = c(0, 10),
     ylim = c(1200,1800),
     xlab = "Years since start of 2010 season", 
     ylab = "Elo Rating",
     main = "NFL Team Elo Ratings (2010-2019)",
     xaxt = "n")
axis(1, at = 0:10)

for (team_data in elo_by_team) {
  lines(team_data$year, team_data$elo, col = rgb(0, 0, 0, 0.3))
  n <- nrow(team_data)
  text(team_data$year[n], team_data$elo[n], team_data$team[n], 
       pos = 4, cex = 0.6)
}

# Plot 2: Bradley-Terry coefficients by season
plot(NULL, 
     xlim = c(2010, 2020),
     ylim = range(bt_results$est,na.rm=TRUE),
     xlab = "Season", 
     ylab = "Bradley-Terry Coefficient",
     main = "NFL Team Strength by Season (Bradley-Terry Model)",
     xaxt = "n")
axis(1, at = 2010:2019)

for (team_data in bt_by_team) {
  lines(team_data$season, team_data$est, col = rgb(0, 0, 0, 0.3))
  n <- nrow(team_data)
  text(team_data$season[n], team_data$est[n], team_data$team[n], 
       pos = 4, cex = 0.6)
}
```
:::
