---
title: "Corrections"
---

```{r}
#| label: load-changes

# The change record is a tracked file in ben-domingue/irw, appended by hand
# when a correction is released (see red_up/README.md there). Read at render
# time, like the version manifest in components/data_version.R. A failed fetch
# prints a notice instead of an empty table: an empty table would read as
# "no corrections", which is a claim this page must never make by accident.
CHANGES_URL <- paste0(
  "https://raw.githubusercontent.com/ben-domingue/irw/main/",
  "metadata/table_changes.csv"
)

changes <- tryCatch({
  con <- url(CHANGES_URL)
  on.exit(try(close(con), silent = TRUE), add = TRUE)
  utils::read.csv(con, stringsAsFactors = FALSE, colClasses = "character",
                  na.strings = character(0))
}, error = function(e) NULL, warning = function(w) NULL)

if (!is.null(changes)) {
  changes <- changes[order(changes$date, changes$table, decreasing = c(TRUE, FALSE),
                           method = "radix"), ]
}
```

IRW tables are sometimes wrong: a processing script reads the wrong column, an
upload doubles a table, a response is miscoded. This page states how IRW
handles that, and lists corrections to published response tables.

## How IRW records change

1. **Released versions do not change.** Every IRW release is numbered, and the
   number always refers to the same data. Anything that pins a version
   (`irw_use_version()` in R, a citation) keeps getting exactly what it got.
2. **Corrections are new versions.** When a table is found to differ from its
   source, the corrected table is published in a later release, and the earlier
   version stays available at its release number. Pin a version for stability;
   take the latest for correctness (the same rule applies to
   [tags](tags_quality.qmd)).
3. **Every correction is recorded.** Each correction to a published response
   table adds a dated entry below: what changed, the IRW version in which it
   took effect, and the GitHub issue where it was reported.
4. **Renames and replacements point both ways.** A table that replaces or is
   renamed from another names the table it replaces, and the replaced table's
   entry names its successor.
5. **Duplicates are recorded, not adjudicated.** When two tables derive from
   the same source, both are kept unless one is a processing error. IRW does
   not rank independent datasets of the same instrument or population.
6. **Withdrawals remove material from the current release onward.** Material
   withdrawn for legal or rights reasons is removed from the current release
   and recorded internally; it is not listed here.
7. **IRW corrects fidelity to the source, not the source.** A table is corrected
   when it does not match what its source released. Doubts about whether the
   source data are good are for the source's authors and IRW's documentation,
   not grounds for altering or flagging a table.

Corrections to item text are listed separately, on [Known Issues](itemtext_issues.qmd).
If you think a table is wrong, open an issue on
[GitHub](https://github.com/ben-domingue/irw/issues).

## Change record

This record begins with corrections released on 2026-08-27, and entries before
2026-09-19 were reconstructed afterwards from the issue tracker and Redivis'
version history, so it may be incomplete for that period. Earlier corrections
are recorded only in the repositories' commit history and in the versioned data
itself. The IRW version is the first corpus version carrying the corrected
table; `irw_use_version()` with any earlier number returns the table as it was.

```{r}
#| label: changes-summary
#| results: asis

if (is.null(changes)) {
  cat("::: {.callout-warning}\nThe change record could not be loaded when this",
      "page was built. It is available directly at",
      sprintf("[`metadata/table_changes.csv`](%s).\n:::\n",
              "https://github.com/ben-domingue/irw/blob/main/metadata/table_changes.csv"))
} else {
  cat(sprintf("%d entries covering %d tables. Start typing to search.\n",
              nrow(changes), length(unique(changes$table))))
}
```

```{=html}
<input type="text" id="changeSearch" placeholder="Search corrections..." oninput="filterChanges()" style="width: 100%; padding: 10px; margin-bottom: 10px; font-size: 16px;">
<p id="change-count" style="color: #888; margin-bottom: 20px;"></p>

<style>
#changes-table { overflow-x: auto; }
#changes-table table { width: 100%; }
#changes-table th, #changes-table td { vertical-align: top; padding: 8px; }
#changes-table td:nth-child(3) { word-break: break-word; }
</style>

<script>
function filterChanges() {
  const filter = document.getElementById('changeSearch').value.toLowerCase().trim();
  const rows = document.querySelectorAll('#changes-table tbody tr');
  let shown = 0;
  rows.forEach(r => {
    const hit = filter === '' || (r.textContent || '').toLowerCase().includes(filter);
    r.style.display = hit ? '' : 'none';
    if (hit) shown++;
  });
  const count = document.getElementById('change-count');
  count.textContent = filter === '' ? '' :
    shown + (shown === 1 ? ' entry matches "' : ' entries match "') + filter + '"';
}
</script>
```

::: {#changes-table}

```{r}
#| label: changes-table
#| results: asis

if (!is.null(changes) && nrow(changes) > 0) {
  cell <- function(x) gsub("|", "\\|", gsub("\n", " ", x), fixed = TRUE)
  links <- function(col, lead) ifelse(nzchar(col), sprintf(" %s `%s`.", lead, col), "")
  issue <- ifelse(grepl("^#[0-9]+$", changes$issue),
                  sprintf("[%s](https://github.com/ben-domingue/irw/issues/%s)",
                          changes$issue, sub("#", "", changes$issue)),
                  changes$issue)
  cat("| Date | IRW version | Table | What changed | Issue |\n",
      "|---|---|---|---|---|\n", sep = "")
  cat(sprintf("| %s | %s | `%s` | %s%s%s | %s |",
              changes$date, changes$irw_version, changes$table,
              cell(changes$change),
              links(changes$replaces, "Replaces"),
              links(changes$replaced_by, "Replaced by"),
              issue),
      sep = "\n")
  cat("\n")
}
```

:::
