---
title: "Data Notes"
---

```{r}
#| label: load-notes

# Data notes are a tracked file in ben-domingue/irw (metadata/data_notes.csv,
# irw#2529), appended by hand. Read at render time, like the change record on
# corrections.qmd. A failed fetch prints a notice instead of an empty table: an
# empty table would read as "nothing to know about any table".
NOTES_URL <- paste0(
  "https://raw.githubusercontent.com/ben-domingue/irw/main/",
  "metadata/data_notes.csv"
)

notes <- tryCatch({
  con <- url(NOTES_URL)
  on.exit(try(close(con), silent = TRUE), add = TRUE)
  utils::read.csv(con, stringsAsFactors = FALSE, colClasses = "character",
                  na.strings = character(0), encoding = "UTF-8")
}, error = function(e) NULL, warning = function(w) NULL)

if (!is.null(notes)) {
  notes <- notes[order(tolower(notes$table), notes$date, method = "radix"), ]
}
```

Some things a user of a table should know cannot be expressed by the table
itself: an answer key in the source that looks wrong, covariates the source
filled with a mean, what a `wave` column counts, forms pooled under shared item
names. IRW reproduces its sources as released
([Corrections](corrections.qmd), rule 7), so these are recorded as notes rather
than fixed. Each note also appears on the table's own page.

A note is not a defect report. Tables with a known IRW error are flagged on
their page until the fix is released, and released fixes are listed on
[Corrections](corrections.qmd). Notes about item text are on
[Known Issues](itemtext_issues.qmd).

```{r}
#| label: notes-summary
#| results: asis

if (is.null(notes)) {
  cat("::: {.callout-warning}\nThe data notes could not be loaded when this",
      "page was built. They are available directly at",
      sprintf("[`metadata/data_notes.csv`](%s).\n:::\n",
              "https://github.com/ben-domingue/irw/blob/main/metadata/data_notes.csv"))
} else {
  cat(sprintf("%d notes covering %d tables. Start typing to search.\n",
              nrow(notes), length(unique(notes$table))))
}
```

```{=html}
<input type="text" id="noteSearch" placeholder="Search notes..." oninput="filterNotes()" style="width: 100%; padding: 10px; margin-bottom: 10px; font-size: 16px;">
<p id="note-count" style="color: #888; margin-bottom: 20px;"></p>

<style>
#notes-table { overflow-x: auto; }
#notes-table table { width: 100%; }
#notes-table th, #notes-table td { vertical-align: top; padding: 8px; }
#notes-table td:nth-child(2) { word-break: break-word; }
</style>

<script>
function filterNotes() {
  const filter = document.getElementById('noteSearch').value.toLowerCase().trim();
  const rows = document.querySelectorAll('#notes-table tbody tr');
  let shown = 0;
  rows.forEach(r => {
    const hit = filter === '' || (r.textContent || '').toLowerCase().includes(filter);
    r.style.display = hit ? '' : 'none';
    if (hit) shown++;
  });
  const count = document.getElementById('note-count');
  count.textContent = filter === '' ? '' :
    shown + (shown === 1 ? ' note matches "' : ' notes match "') + filter + '"';
}
</script>
```

::: {#notes-table}

```{r}
#| label: notes-table
#| results: asis

if (!is.null(notes) && nrow(notes) > 0) {
  cell <- function(x) gsub("|", "\\|", gsub("\n", " ", x), fixed = TRUE)
  issue <- ifelse(grepl("^#[0-9]+$", notes$issue),
                  sprintf("[%s](https://github.com/ben-domingue/irw/issues/%s)",
                          notes$issue, sub("#", "", notes$issue)),
                  notes$issue)
  cat("| Table | Note | Issue | Added |\n",
      "|---|---|---|---|\n", sep = "")
  cat(sprintf("| `%s` | %s | %s | %s |",
              notes$table, cell(notes$note), issue, notes$date),
      sep = "\n")
  cat("\n")
}
```

:::
