---
title: "The Item Response Warehouse"
---

The Item Response Warehouse (IRW) is a collection of open, harmonized item response datasets. These datasets span a wide range of measures; examples include measures of ability in education settings, measures of partisanship based on voting records, and personality measures based on survey questions. While they span a wide range of measures, the data are harmonized in a common format to allow for standardized analysis of larger quantities of data. The key element required of data for inclusion in the IRW is that the data need to report individual responses that are cross-classified by the focus of measurement (typically a person) and the measurement tool (typically an item). Responses can be scored in a variety of ways (dichotomous, polytomous, continuous) and may contain additional data (e.g., response time), but the cross-classification aspect is fundamental. 

```{r}
library(dplyr)
library(readr)
library(stringr)
```

```{r}
# set user and dataset
user <- redivis::user("stanford_templates")
dataset <- user$dataset("item_response_warehouse")

# get metadata table
metadata_table <- dataset$table("metadata")
metadata <- metadata_table$to_tibble()

# get dataset tables
datasets <- list(
  "4thgrade_math_sirt" = dataset$table("4thgrade_math_sirt")$to_tibble(),
  "florida_twins_behavior" = dataset$table("florida_twins_behavior")$to_tibble()
)

# pass data to ojs
ojs_define(metadata = metadata)
ojs_define(datasets = datasets)
```

## Individual datasets

```{ojs}
Plot = import("https://esm.sh/@observablehq/plot") 

// selector for dataset
ds = new Map(Object.entries(datasets))
viewof dataset = Inputs.select(ds, {label: "Dataset"})

// summarise items
item_summary = d3.rollup(transpose(dataset),
                         v => {return {mean: d3.mean(v, d => d.resp)}},
                         d => d.item)

// summarise subjects
// items = new Set(dataset.item)

subj_summary = d3.rollup(transpose(dataset),
                         v => {return {mean: d3.mean(v, d => d.resp)}},
                                       // n_items: v.length,
                                       // prop: d3.count(v, d => d.item) / items.size}},
                         d => d.id)

// plot items
viewof item_plot = Plot.plot({
  x: {label: "Item mean response"},
  y: {grid: true},
  marks: [
    Plot.rectY(item_summary.values(), Plot.binX({y: "count"}, {x: "mean"})),
    Plot.ruleY([0])
  ]
})

// plot subjects
viewof subj_plot = Plot.plot({
  x: {label: "Subject mean response"},
  y: {grid: true},
  marks: [
    Plot.rectY(subj_summary.values(), Plot.binX({y: "count"}, {x: "mean"})),
    Plot.ruleY([0])
  ]
})

// display item and subject plots side by side
html`<div style="display: flex;">
 <div style="flex-basis:50%">${viewof item_plot}</div>
 <div style="flex-basis:50%">${viewof subj_plot} </div>
</div>`
```

## Metadata

```{ojs}
// selectors for x and y variables
vars = new Map([["Number responses", "n_responses"],
                ["Number participants", "n_participants"],
                ["Number items", "n_items"],
                ["Responses per participant", "responses_per_participant"],
                ["Responses per item", "responses_per_item"],
                ["Sparsity", "sparsity"]])

viewof x_var = Inputs.select(vars, {value: "n_items", label: "X axis"})
viewof y_var = Inputs.select(vars, {value: "n_participants", label: "Y axis"})

// selector for color variable
default_color = "grey"
color_opts = new Map([["None", null],
                      ["Dichotomous vs. Polytomous", "partition"]])

viewof color_var = Inputs.select(color_opts, {label: "Color"})
```

```{ojs}
// histogram
viewof metadata_hist = Plot.plot({
  y: {grid: true},
  color: {legend: color_var != null},
  marks: [
    Plot.rectY(transpose(metadata), Plot.binX({y: "count"}, {x: x_var, fill: color_var || default_color})),
    Plot.ruleY([0])
  ]
})

// scatter plot
viewof metadata_scatter = Plot.plot({
  x: {type: "log"},
  y: {type: "log"},
  grid: true,
  color: {legend: color_var != null},
  marks: [
    Plot.dot(transpose(metadata), {
      x: x_var,
      y: y_var, 
      stroke: color_var || default_color
    }),
    Plot.tip(transpose(metadata), Plot.pointer({
      x: x_var,
      y: y_var,
      stroke: "partition",
      title: "dataset"
    })),
    Plot.linearRegressionY(transpose(metadata), {
      x: x_var, 
      y: y_var, 
      stroke: color_var || default_color
    })
  ]
})

// display histogram and scatter plots side by side
html`<div style="display: flex;">
 <div style="flex-basis:50%">${viewof metadata_hist}</div>
 <div style="flex-basis:50%">${viewof metadata_scatter} </div>
</div>`
```
