---
title: "Item Text Descriptives"
---

```{r}
#| label: load-data
#| include: false

library(redivis)

user <- redivis$user("bdomingu")
dataset <- user$dataset("irw_meta:bdxt")
table <- dataset$table("itemtext_metadata:drat")

data <- table$to_tibble()
data <- as.data.frame(data)
data$mean_word <- as.numeric(data$mean_word)
data$mean_character <- as.numeric(data$mean_character)
data$instrument <- gsub("\n", "", data$instrument)

n_tables      <- nrow(data)
n_instruments <- length(unique(data$instrument))
median_words  <- round(median(data$mean_word, na.rm = TRUE))
median_chars  <- round(median(data$mean_character, na.rm = TRUE))

ojs_define(plot_data = data)
```

The IRW item text resource contains full item wording for **`r n_tables` tables** spanning **`r n_instruments` distinct instruments**. Items have a median length of `r median_words` words (`r median_chars` characters). See the [Item Text Standard](itemtext.qmd) for a full description of the data structure.

## Tables with item text

```{ojs}
//| label: browse-table

plotData = transpose(plot_data)

viewof search = Inputs.search(plotData, {placeholder: "Search by table or instrument…"})

Inputs.table(search, {
  columns: ["table", "instrument", "mean_word", "mean_character"],
  header: {
    table: "Table",
    instrument: "Instrument",
    mean_word: "Mean words",
    mean_character: "Mean characters"
  },
  format: {
    mean_word: x => x.toFixed(1),
    mean_character: x => x.toFixed(1)
  },
  sort: "table"
})
```

## Item complexity

The scatterplot shows mean words vs. mean characters per item for each table; the histogram summarises the distribution of item length. Hover for table and instrument details.

```{ojs}
//| label: item-complexity

hist = Plot.plot({
  y: {grid: true},
  x: {label: "Mean words per item"},
  marks: [
    Plot.rectY(plotData, Plot.binX({y: "count"}, {x: "mean_word", fill: "steelblue", tip: true})),
    Plot.ruleY([0])
  ]
})

scatter = Plot.plot({
  grid: true,
  marks: [
    Plot.dot(plotData, {
      x: "mean_word",
      y: "mean_character",
      fill: "steelblue",
      fillOpacity: 0.6,
      r: 4
    }),
    Plot.tip(plotData, Plot.pointer({
      x: "mean_word",
      y: "mean_character",
      title: d => `${d.table}\n${d.instrument}\n${d.mean_word.toFixed(1)} words · ${d.mean_character.toFixed(1)} chars`
    }))
  ],
  x: {label: "Mean words per item"},
  y: {label: "Mean characters per item"}
})

html`<div style="display:flex;gap:1.5rem;">
  <div style="flex:1">${hist}</div>
  <div style="flex:1">${scatter}</div>
</div>`
```

## Accessing item text

::: panel-tabset
## R

```{r}
#| eval: false
#| echo: true
library(irw)

irw_list_itemtext_tables()                   # List tables that have item text
items <- irw_itemtext("gilbert_meta_49")     # Fetch item text for a specific table
```

## Python

```{python}
#| eval: false
#| echo: true
#| python.reticulate: false
import irw

irw.list_tables_with_itemtext()              # List tables that have item text
items = irw.itemtext("gilbert_meta_49")      # Fetch item text for a specific table
```
:::
