---
title: "The Nominal Data Standard"
---

## The data standard

The IRW nominal data standard is meant to describe data wherein responses are nominal and not ordinal. These tables form the nominal [data family](overview.qmd#families) (`source = "nom"`). There are two distinct types of such data. The first type involves responses in a necessarily finite (but nominal) set of mutually exclusive categories; a typical example here might be the specific option from a multiple choice item that is selected. The second type involves responses that are potentially infinite; examples here would be constructed responses like essays. 

- `text`: This is the nominal version of `resp`.

Where a nominal table has a counterpart in the main branch of the IRW, `text` holds the
same content that the standard table carries as `resp_raw` (see [the IRW data
standard](standard.qmd)). The two branches differ in what they treat as the response: the
standard table scores it and keeps the original alongside, while the nominal table takes
the original as the response itself. For the ENEM tables, for instance, the standard
table's `resp` is 0/1 for incorrect/correct with `resp_raw` giving the letter the
candidate marked, and the nominal table's `text` is that letter. Not every nominal table
is derived this way — constructed responses such as essays originate in the nominal
branch.

## Accessing the nominal data

Documentation related to the tables can be found [here](https://docs.google.com/spreadsheets/d/12tM4vADKcUm5LGOGRwQ5_HKkdYa3mZUaKbFUqgs2U_w/edit?gid=0#gid=0).

## What the family holds

{{< include _load-family-meta.qmd >}}

```{r}
#| label: nominal-descriptives
nom <- irw_meta_table("nominal_metadata")
for (v in c("n_responses", "n_participants", "n_items", "n_categories")) nom[[v]] <- as.numeric(nom[[v]])
n_nom       <- nrow(nom)
n_nom_resp  <- sum(nom$n_responses, na.rm = TRUE)
n_nom_items <- sum(nom$n_items, na.rm = TRUE)
is_enem     <- grepl("^enem", nom$table, ignore.case = TRUE)
n_enem      <- sum(is_enem)
share_enem  <- sum(nom$n_responses[is_enem], na.rm = TRUE) / n_nom_resp
n_open      <- sum(nom$n_categories > 20, na.rm = TRUE)
cat_counts  <- table(cut(nom$n_categories, breaks = c(0, 3, 4, 5, 6, 7, 10, 20, Inf),
                         labels = c("3", "4", "5", "6", "7", "8-10", "11-20", "20+"), right = TRUE))
```

The family holds `r n_nom` tables, `r fmt_big(n_nom_resp)` responses to `r fmt_n(n_nom_items)` items. One source dominates: the `r n_enem` ENEM tables (Brazil's national secondary exam, one table per year and test booklet, each holding the letter every candidate marked) account for `r round(100 * share_enem)`% of the responses. The remaining tables are smaller and more varied: option-level records of multiple-choice tests, stimulus-identification tasks, and `r n_open` tables of constructed responses (essays, free-text answers, drawn responses) where the number of distinct categories runs into the hundreds or thousands because nearly every response is unique. Each table's own counts are in `irw_metadata(source = "nom")`.

```{r}
#| label: fig-nominal-categories
#| fig-cap: "Nominal tables by number of distinct response categories. The 20+ group is the constructed-response tables."
#| fig-height: 3.4
#| fig-width: 7
family_bar(cat_counts, xlab = "Distinct response categories per table")
```

## An example

Let's look at an example of how we can fit the nominal response model to a table (`preference_inventory_nom`) from the nominal branch of the IRW and generated category response functions for all items.

::: panel-tabset
## R

```{r}
#| echo: true
df<-irw::irw_fetch("preference_inventory_nom",source='nom')
resp<-irw::irw_long2resp(df,resp_col='text')
resp$id<-NULL
library(mirt)
m<-mirt(resp,1,'nominal')
f<-function(mod,n) {
    extr <- extract.item(mod,n)
    Theta <- matrix(seq(-6,6, length.out=2000))
    pr <- probtrace(extr, Theta)
    list(Theta,pr)
}
out<-list()
for (n in 1:ncol(resp)) out[[n]]<-f(m,n)
##
par(mfrow=c(4,3),mgp=c(2,1,0),mar=c(3,3,1,1))
cols<-c("black","red","blue","green")
for (i in 1:length(out)) {
    plot(NULL,xlim=c(-5,5),ylim=0:1,xlab='theta',ylab='Pr')
    z<-out[[i]]
    for (j in 1:4) lines(z[[1]],z[[2]][,j],col=cols[j])
}
```
:::
