---
title: "Tag Quality"
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(stringr)
library(glue)
library(knitr)
library(redivis)

## Tables are addressed by NAME here, not by the `name:referenceId` form the
## other data pages use. Those reference ids rotate when the metadata pipeline
## rebuilds the dataset, and every one of them currently fails to resolve.
irw_meta <- redivis$user("datapages")$dataset("irw_meta:bdxt:latest")
metadata <- irw_meta$table("metadata")$to_tibble()
tag_table <- irw_meta$table("tags")$to_tibble()

## The version this page is describing, named on the page itself: coverage is
## whatever the released tags table holds at build time, and a reader comparing
## it against a download needs to know which release that was.
tag_version <- irw_meta$get()$properties$version$tag

## Every join between the tag sheet and the metadata is case-insensitive on
## purpose. Several hundred tag rows differ from their metadata row only by
## case, and a case-sensitive join silently drops all of them -- which is how
## tag coverage once read 53% instead of 62%.
key <- function(x) str_to_lower(str_trim(as.character(x)))

## Sentinels a rater typed instead of leaving a cell empty. They are not tags,
## and counting them as filled would overstate every column they appear in.
## Same list as `_load-data-explore.qmd`; keep the two in step.
NOT_A_TAG <- str_to_lower(c(
  "NA", "N/A", "no access to the osf page", "non-verbal task",
  "I can't find the description of this dataset", "missing description",
  "need help", "no link or info", "Missing (NA)"
))

filled <- function(x) {
  v <- str_trim(as.character(x))
  !is.na(v) & v != "" & !(str_to_lower(v) %in% NOT_A_TAG)
}

## The published tag table carries the eight columns in the order
## `metadata/03_tags.R` selects them, and that selection is POSITIONAL. So the
## labels below are matched to position and guarded: a schema change should
## break this page loudly rather than relabel a column silently.
tag_cols <- names(tag_table)[-1]
expect <- c("age", "child", "sample", "construct_type",
            "measurement", "item_format", "lang", "construct_name")
stopifnot(
  length(tag_cols) == 8,
  all(str_detect(str_to_lower(tag_cols), expect))
)

labels <- c("age range", "child age", "sample", "construct type",
            "measurement tool", "item format", "primary language(s)",
            "construct name")
classes <- c("Derived", "Derived", "Definitional", "Definitional",
             "Tagger", "Tagger", "Tagger", "Description")

live <- key(metadata$table)
n_tables <- length(live)
rows <- tag_table[key(tag_table$table) %in% live, , drop = FALSE]
n_rows <- nrow(rows)

n_filled <- vapply(tag_cols, function(c) sum(filled(rows[[c]])), integer(1))

## `child age` is only ever filled for tables whose sample includes children.
## Against all tables it reads ~17% and looks like the worst column on the site;
## against the tables eligible for a value it is the second-best covered. The
## denominator has to match the claim.
age_col <- tag_cols[1]
eligible_child <- sum(str_detect(as.character(rows[[age_col]]),
                                 fixed("Child (<18y)")) |
                      str_detect(as.character(rows[[age_col]]),
                                 fixed("Mixed")), na.rm = TRUE)

pct <- function(k, d = n_tables) sprintf("%.1f%%", 100 * k / d)
comma <- function(k) format(k, big.mark = ",")

coverage <- tibble(
  Column   = paste0("`", labels, "`"),
  `Produced by` = classes,
  Filled   = comma(n_filled),
  Coverage = ifelse(labels == "child age",
                    paste0(pct(n_filled, eligible_child), " *"),
                    pct(n_filled))
)
```

The IRW annotates every table with eight qualitative tags --- who was measured,
what was measured, and how. This page says what those tags are worth: how much
of each column is filled, how accurate it has been measured to be, who or what
wrote it, and what the evidence behind those numbers does not cover.

If you are about to filter on a tag, the section you need is
[Coverage, per column](#coverage-per-column) and then
[What the accuracy numbers do not cover](#what-the-accuracy-numbers-do-not-cover).

::: {.callout-important appearance="simple"}
## The row figure is not a coverage number

```{r}
#| label: headline
#| echo: false
#| results: asis
cat(glue(
  "**{comma(n_rows)} of the {comma(n_tables)} tables ({pct(n_rows)}) carry a tag row.** ",
  "A row is not an answer. Some columns in it are usually filled and some are ",
  "usually empty, so the row figure tells you almost nothing about the column ",
  "you intend to filter on. Read that column's own number below.\n"
))
```
:::

## Coverage, per column

```{r}
#| label: coverage-table
#| echo: false
kable(coverage, align = c("l", "l", "r", "r"))
```

```{r}
#| label: coverage-notes
#| echo: false
#| results: asis
cat(glue(
  "Computed at page build from **tags {tag_version}**, the released version you ",
  "can download today, against the {comma(n_tables)} tables in the IRW. ",
  "Values a rater typed to mean ",
  '"I could not find this" --- `need help`, `missing description` --- are ',
  "counted as empty rather than as tags.\n\n",
  "`*` **`child age` is shown against a different denominator**, and has to be: ",
  "it is only ever filled when the sample includes children. Its ",
  "{comma(n_filled[2])} values are counted against the ",
  "{comma(eligible_child)} tables whose `age range` is `Child (<18y)` or ",
  "`Mixed`. Against all {comma(n_tables)} tables it reads ",
  "{pct(n_filled[2])}, which would make it look like the worst-covered column ",
  "in the project. Blank is the correct value everywhere else.\n"
))
```

Two columns sit far below the rest, and they are not backlog --- see
[what is published and what is withheld](#what-is-published-and-what-is-withheld).
`construct name` is also not a coverage target in the way the others are: it is
free text naming an instrument, with most values used exactly once, so there is
no right answer being missed.

## Four kinds of column

A coverage figure only means something once you know what produced the value.
Each column belongs to exactly one of these classes, and each is held to its own
standard.

**Derived** --- `age range`, `child age`. Computed from the table's own
`cov_age` column rather than judged by anyone. `cov_age` is usable only when it
parses as numeric, has at least 30 non-missing values, lies within [0, 120] and
is not a banded code; a 2% floor stops three 17-year-olds in a 46,000-person
adult survey from making a table `Mixed`. This is the one place where a computed
value **outranks a human tag**.

**Tagger** --- `primary language(s)`, `item format`, `measurement tool`.
Answered from the study's source document against a fixed vocabulary, by a human
working the annotation sheet or by an automated tagger reading the same rules.

**Definitional** --- `sample`, `construct type`. The vocabulary is enforced, but
the boundaries between its values had to be written down before anyone, human or
machine, could apply them consistently. Writing `sample`'s rules and then
amending them moved its frame facet from 26.9% to 45.5% to 54.5% exact match
**with no change to the tagger at all**.

**Description** --- `construct name`. Free text: "Ages and Stages Questionnaire
(ASQ-3)", "International Math Olympiad problems". No enumeration, so no
precision to measure.

## Who wrote these tags

Every tag was written either by a human annotator working a spreadsheet or by an
automated tagger that reads the study's source document. **The published table
does not distinguish them**, so if provenance matters for your use, that
distinction lives in `tags/tags_auto.csv` in the
[processing repository](https://github.com/ben-domingue/irw), where every
machine-written row is stamped `Rater = claude-auto`. As of 2026-09-04 that
file holds 1,847 tagged tables, carrying:

| Column | Values written by the tagger |
|---|---:|
| `measurement tool` | 1,837 |
| `item format` | 1,725 |
| `primary language(s)` | 1,711 |
| `sample` (setting facet only) | 1,494 |

: A snapshot of the staging file, not of any published version. Staged values
reach a release only once that release is published, and a human row overrides
one wherever the two disagree.

Three rules govern the mix:

- **A human row wins.** The annotation sheet is authoritative for anything a
  person touched, and machine output is unioned underneath it.
- **A derived `age range` beats both.** It is computed from the responses
  themselves, which is better evidence than either rater had.
- **Nothing is permanent.** No published version of the tags is authoritative
  over a later, better one. A tag that can be *shown* wrong --- against the
  source, not merely against a different rater's judgment --- gets fixed. Every
  Redivis version is immutable and citable, so a correction produces a new
  version rather than editing the one you cited. Pin a version for stability;
  take the latest for correctness.

## What is published and what is withheld

The automated tagger fills every field it can support. A staging step then
decides what is allowed out, and blanks the rest before anything is written.
**Publishing is a decision made at assembly, not by the tagger**, and the bar is
90% per-atom precision against the human-annotated set.

| Machine-written column | | Why |
|---|---|---|
| `primary language(s)` | published | 100% per-atom precision against gold (n=33) |
| `item format` | published | 90.9% accurate when it commits |
| `measurement tool` | published | 86.8--93.3% accurate across runs |
| `sample` --- setting facet | published | 93.3% per-atom precision |
| `sample` --- frame facet | **withheld** | 59.1% precision; below the bar |
| `construct type` | **withheld** | 50.0--66.7% per-atom precision |
| `age range`, `child age` | **withheld** | the `cov_age` derivation owns these |
| `construct name` | **withheld** | free text; never scored |

This is why `construct type` sits near 55% while `measurement tool` sits near
99%. The tagger has an answer for `construct type` on most tables. It is not
good enough to publish, so it is held back rather than shipped with a caveat.

::: {.callout-warning appearance="simple"}
## One column carries two facets, and only one of them publishes

`sample` answers two questions at once: *where* the study happened --- the
**setting**: `Educational`, `Clinical`, `Workplace`, `Internet-based`,
`Program-based`, `Non-human` --- and *how broad* the sampling was --- the
**frame**: `Representative`, `Targeted/specific`, `General/non-specific`.

Only the setting facet publishes from the tagger. A machine-written
`Workplace, Targeted/specific` is staged as `Workplace` alone. So **a missing
frame value on a machine-tagged table means withheld, not unknown**. Human-tagged
rows carry both facets.
:::

## Measured accuracy

Scored blind against the human-annotated set: predictions were made without
access to the annotation sheet, by a tagger running the same vocabulary the
rules describe. Blanks count as abstentions rather than errors, because the
vocabulary asks the tagger to leave a field empty rather than guess --- so
*exact* means right when it commits.

| Column | Exact | Precision | Recall | n | How it fails |
|---|---:|---:|---:|---:|---|
| `primary language(s)` | 90.9% | 100.0% | 91.7% | 33 | never names a language that is not there; misses secondary ones |
| `item format` | 90.9% | --- | --- | 38 | abstains on 42% of tables rather than guess |
| `measurement tool` | 86.8% | --- | --- | 38 | commits on every table |
| `sample` --- setting | 87.5% | 93.3% | 87.5% | 16 | improved at every revision of the rules |
| `sample` --- frame | 54.5% | 59.1% | 56.5% | 22 | the weakest column; over-commits to `Representative` |
| `construct type` | 64.9% | 80.0% | 69.6% | 37 | under-tags: right when it names a facet, misses 30% of the facets gold holds |
| `age range` | 91.7% | --- | --- | 12 | scored against the `cov_age` derivation, not against the sheet |
| `child age` | --- | --- | --- | 6 | too few labelled rows to score |

: Measured 2026-09-01, on one 60-table sample. {tbl-colwidths="[20,9,10,9,6,46]"}

Precision and recall are per atom, because these columns are multi-select:
naming two facets of three should not score the same as naming something
unrelated. They are reported alongside exact match, never instead of it --- a
bar set on partial credit alone would pass a tagger that reliably names one
facet of three.

::: {.callout-caution appearance="simple"}
## Do not use whole-cell match on `sample`

Compared cell-for-cell, `sample` scores 14.7%, and that figure measures the
annotation sheet's blanks rather than the tagger. The tagger answers the frame
facet on 34 of 34 tables while the sheet answers it on 22 of 34, so whole-cell
matching penalises the tagger for filling a facet the key leaves empty --- 15 of
its 33 misses are supersets rather than wrong answers. Score this column one
facet at a time, on the tables where the sheet answered that facet.
:::

## What happens when the same pipeline runs twice

Accuracy is not the only question worth asking. On 2026-09-04, 60 already-tagged
tables were re-tagged blind, end to end, with cleared caches --- the same
pipeline, a second time. Separately, 369 tables that carried a language and
nothing else were re-tagged by agents who were not shown the existing value,
making that column an independent second opinion at six times the size.

| Column | Both runs answered | Identical | Second opinion at scale |
|---|---:|---:|---|
| `item format` | 49 | 100.0% | --- |
| `measurement tool` | 52 | 100.0% | --- |
| `primary language(s)` | 49 | 93.9% | 91.1% identical on 359 further tables |
| `sample` | 41 | 82.9% | --- |

: Run-to-run agreement, 2026-09-04.

**Stability is not correctness.** None of those re-tagged tables has a human
label to be right or wrong against. Two runs agreeing means the pipeline is
reproducible, not that it is right.

**A disagreement is sometimes a defect, and finding it is the point.** Of the 32
language disagreements in the 369-table comparison, ten were the same language
spelled two ways under ISO 639-2 --- `fra` against `fre`, `cze` against `ces`
--- since collapsed on export. The other 22 are substantive and have not been
adjudicated. One earlier disagreement was a plain error: a Brazilian validation
study recorded as Persian, found only because the same table was tagged twice.

**The two runs commit at different rates and nobody knows why.** Where one run
answered and the other left a blank, the second run was the one committing 11
times on `sample`, 8 on `item format` and 7 on `measurement tool`, against 0, 1
and 1 in the other direction. The asymmetry is real in this sample of 60; its
cause is not established.

## One column may be inferred, another may never be

The tagging rules are not just lists of permitted values, and two of them point
in deliberately opposite directions.

**`age range` may not be inferred.** With no usable `cov_age`, it is tagged only
from explicit age information in the source: a stated numeric inclusion
criterion, a reported minimum or maximum, a school grade that cannot include
18-year-olds. Never from a country, a survey programme's name, the word
"undergraduates" (which routinely includes 17-year-olds), or the construct.
A wrong value here silently mis-filters a sample, and a blank is cheaper to fix
than a wrong tag.

**`primary language(s)` may be inferred, and should be.** Where a source does
not state the language of administration, it is inferred from the country, the
population, a forward-backward translation, or the language of the instrument's
cited version. The corpus is majority non-English, and every missing value makes
it read as more English than it is --- so erring toward naming a non-English
language is the intended failure direction.

Two inferences stay off-limits even there, because neither is about the
respondents: a repository record's own language field describes the *record*
(Zenodo stamps "English" on deposits from Finnish universities), and the
language a paper is written in is not the language its questionnaire went out
in. What is recorded is what the respondents read.

## What the accuracy numbers do not cover

This section is the point of the page, not a disclaimer at the end of it.

1. **The evidence is one 60-table sample, scored three times.** Per-column n
   runs from 6 to 38. These are directional, not tight estimates. One model
   only; a stronger one would likely differ, and the gap is unmeasured.

2. **The labelled set is known to be wrong for `age range`.** Of 29 gold
   `Mixed` labels checkable against the table's own `cov_age`, 25 --- 86% ---
   are contradicted by it. That is why this column is now derived rather than
   tagged, and why its accuracy is scored against the derivation instead.

3. **No ceiling is known.** The human-annotated set is itself human work of
   unmeasured consistency. Nobody has measured how often two annotators
   disagree, so every figure above is scored against a standard whose own
   quality is unknown.

4. **Whole-cell `sample` match is a dead statistic** --- see the warning above.

5. **Coverage is not reach.** Between 8% and 37% of sampled tables produced
   nothing at all across runs: failed fetches, paywalls, deposit records with no
   usable description. Difficulty tracks the *shape of the source* rather than
   the warehouse a table sits in --- a national assessment programme published
   through a data portal is harder than a single-instrument psychometrics paper.

6. **Some subjectivity is irreducible.** The re-run found roughly five tables in
   twelve where a second rater could reasonably land somewhere else. Those are
   not errors, and rewriting them would be churn rather than repair.

## Where these numbers come from

Coverage on this page is computed at build time from the published tag table, so
it describes the version you can download today. The accuracy and reliability
figures are dated snapshots of specific scoring runs and are reproducible from
the scripts in
[`tags/scoring/`](https://github.com/ben-domingue/irw/tree/main/tags/scoring) in
the processing repository, alongside the predictions they scored. The rules
themselves live in that repository's tagging vocabulary, and the reasoning
behind each one --- the case as it was put, and what was decided --- in
`tags/decisions/`.

Corrections are welcome; open an issue on
[ben-domingue/irw](https://github.com/ben-domingue/irw/issues).
