---
title: "2PL Discrimination Parameters Across Cognitive Datasets"
description: |
  A Tier 2 IRW vignette demonstrating multi-dataset IRT analysis. We use `irw_filter()` 
  to select dichotomous cognitive/educational datasets, fit a 2PL model to each, and 
  examine what discrimination parameters look like in the wild.
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)

# Load precomputed results (produced by 2pl_across_datasets_compute.R)
.cache           <- readRDS("2pldata/2pl_across_datasets_results.rds")
all_results      <- .cache$all_results
cognitive_tables <- .cache$cognitive_tables
```

## Research question

Item discrimination — the *a* parameter in a 2PL model — captures how sharply an item
separates high- and low-ability respondents. Methodologists often assume *a* ≈ 1 as a
starting point (the Rasch model), but real data are messier. How variable is
discrimination across cognitive and educational instruments in practice? And how
consequential is the Rasch constraint of fixing *a* = 1?

The IRW makes this question tractable: we can filter for dichotomous cognitive datasets,
fit a 2PL to each, and examine the resulting distribution of *a* estimates across
instruments. What would otherwise require assembling dozens of datasets by hand is a few
lines of code.

---

## Step 1: Select datasets using metadata

We use `irw_filter()` to find dichotomous datasets (`n_categories == 2`) tagged as
**Cognitive/educational**, with at least 10 items and 500 respondents, and a response
density above 0.8 to avoid sparse matrices that destabilise 2PL estimation.

```{r}
#| label: filter-datasets
#| eval: false

cognitive_tables <- irw_filter(
  construct_type = "Cognitive/educational",
  n_categories   = 2,            # dichotomous only
  n_items        = c(10, 60),    # at least 10 items, at most 60
  n_participants = c(500, Inf),  # need enough respondents for stable estimation
  density        = c(0.8, 1.0)  # avoid sparse response matrices
)
```

This selected `r length(cognitive_tables)` datasets from the IRW
(as of `r format(.cache$date_run, "%B %Y")`).

::: {.callout-note}
To reproduce with current IRW holdings, re-run `2pl_across_datasets_compute.R` and
commit the updated `2pldata/2pl_across_datasets_results.rds`.
:::

---

## Step 2: Fit a 2PL to each dataset

We write a function that (1) fetches a table, (2) downsamples to at most 10,000
respondents to keep memory use manageable, (3) reshapes from long to wide format,
(4) fits a 2PL in `mirt`, and (5) extracts item parameter estimates. We then
`map()` over all selected tables.

The lognormal prior on *a* (`lnorm, 0.0, 1.0`) keeps discrimination estimates
positive and prevents runaway values for poorly behaved items, while remaining
weak enough to let the data speak.

```{r}
#| label: fit-function
#| eval: false

fit_2pl <- function(table_name) {
  mirt::mirt.options(cores = 1)  # prevent mirt from spawning its own threads

  df <- tryCatch(irw_fetch(table_name), error = function(e) NULL)
  if (is.null(df)) return(NULL)

  # Downsample to at most 10,000 respondents before reshaping
  unique_ids <- unique(df$id)
  if (length(unique_ids) > 10000) {
    df <- df[df$id %in% sample(unique_ids, 10000), ]
  }

  resp <- irw_long2resp(df)
  resp$id <- NULL

  # Drop zero-variance items
  resp <- resp[, sapply(resp, function(x) length(unique(na.omit(x))) > 1), drop = FALSE]
  if (ncol(resp) < 5) return(NULL)

  ni         <- ncol(resp)
  model_spec <- mirt.model(paste0(
    "F = 1-", ni, "\n",
    "PRIOR = (1-", ni, ", a1, lnorm, 0.0, 1.0)"
  ))

  fit <- tryCatch(
    mirt(resp, model_spec, itemtype = rep("2PL", ni),
         method = "EM", technical = list(NCYCLES = 500), verbose = FALSE),
    error = function(e) NULL
  )
  if (is.null(fit)) return(NULL)

  params <- coef(fit, IRTpars = TRUE, simplify = TRUE)$items
  tibble(
    table = table_name,
    item  = rownames(params),
    a     = params[, "a"],
    b     = params[, "b"]
  )
}

# Run in parallel; each result written to disk as it completes
# (see 2pl_across_datasets_compute.R for full details)
all_results <- map(cognitive_tables, fit_2pl) |>
  compact() |>
  bind_rows()
```

---

## Step 3: Examine discrimination distributions

### Distribution of *a* across all items

```{r}
#| label: fig-a-density
#| fig-cap: "Distribution of 2PL discrimination parameters (a) across cognitive datasets. Estimates > 4 are winsorised for display. Dashed line marks a = 1 (the Rasch assumption)."
#| fig-height: 4

all_results |>
  mutate(a_win = pmin(a, 4)) |>
  ggplot(aes(x = a_win)) +
  geom_density(fill = "#2166ac", colour = "#2166ac", alpha = 0.4, linewidth = 0.8) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  labs(
    x       = "Discrimination (a)",
    y       = "Density",
    caption = "Dashed line at a = 1 (Rasch assumption). Values winsorised at 4."
  ) +
  theme_minimal(base_size = 13)
```

### Dataset-level medians

Each point below is one IRW table. This shows whether the pattern is consistent
across datasets or driven by a few outliers.

```{r}
#| label: fig-dataset-medians
#| fig-cap: "Per-dataset median discrimination, ordered by value. Point size reflects number of items."
#| fig-height: 4

dataset_summary <- all_results |>
  group_by(table) |>
  summarise(
    median_a = median(a, na.rm = TRUE),
    n_items  = n(),
    .groups  = "drop"
  ) |>
  arrange(median_a) |>
  mutate(rank = row_number())

ggplot(dataset_summary, aes(x = rank, y = median_a)) +
  geom_point(aes(size = n_items), colour = "#2166ac", alpha = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  scale_size_continuous(range = c(2, 7), name = "# items") +
  labs(
    x = "Dataset (ranked by median a)",
    y = "Median discrimination (a)"
  ) +
  theme_minimal(base_size = 13)
```

### Summary statistics

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of 2PL discrimination estimates across cognitive datasets"

all_results |>
  summarise(
    Datasets      = n_distinct(table),
    Items         = n(),
    `Mean a`      = round(mean(a, na.rm = TRUE), 3),
    `Median a`    = round(median(a, na.rm = TRUE), 3),
    `SD a`        = round(sd(a, na.rm = TRUE), 3),
    `% above 1.0` = round(mean(a > 1, na.rm = TRUE) * 100, 1)
  )
```

---

## What to notice

**The Rasch assumption.** The dashed line at *a* = 1 marks where the Rasch model
parks every item. If the distribution is centred well above 1, or has a long right
tail, that suggests the equal-discrimination constraint is doing real work — and
possibly distorting person ability estimates for items at the extremes.

**Between-dataset variance.** The per-dataset median plot is often more informative
than the pooled density. If median *a* varies widely across datasets, that points to
instrument design or sample homogeneity as a driver, not just item-level noise.

**Limitations.** We downsample to 10,000 respondents per dataset, so estimates for
large-scale assessments are based on a random subset. For the purpose of characterising
the distribution of *a* this is acceptable — 10,000 respondents is more than enough
for stable 2PL estimation — but it means we are not using the full IRW data.
Some tables may also come from the same instrument administered in different
populations; `irw_info()` can help identify these.

---

## Extending this vignette

- **Model fit comparison.** For each dataset, compare Rasch vs. 2PL using `M2()`
  fit statistics or IMV (see the [IMV tutorial](imv.qmd)). Where does freeing
  discrimination actually improve fit?

- **Difficulty coverage.** The `all_results` object also contains *b* parameters.
  How does the range of item difficulties vary across instruments?

- **Construct type granularity.** Re-run with other `construct_type` values to
  compare discrimination patterns across domains.

---

## Reproducibility

Results on this page were computed on `r format(.cache$date_run, "%B %d, %Y")`
using `r length(cognitive_tables)` IRW tables. To reproduce:

```r
# 1. Re-run the compute script (from project root)
source("vignettes/2pl_across_datasets_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/2pl_across_datasets.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
for (tbl in cognitive_tables) {
  irw_save_bibtex(tbl, output_file = "2pldata/irw_references.bib", append = TRUE)
}
```
:::
