---
title: "AI Models as Respondents: An IMV Model Comparison"
description: |
  Does the extra structure of a 2PL, or of a second factor, improve
  prediction the same way when the respondents are AI models as when they are
  people? We fit Rasch, 2PL and two-factor models to item-level results from
  ten OpenEval benchmarks in IRW and to human IRW tables subsampled to the same
  size, and compare them with the cross-validated InterModel Vigorish (IMV).
date: today
bibliography: ai_benchmark_imv_data/references.bib
nocite: |
  @*
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(knitr)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "ai_benchmark_imv_data/ai_benchmark_imv_results.rds"
.cache     <- readRDS(cache_path)
results    <- .cache$results
folds      <- .cache$folds
prep       <- .cache$prep_counts
excluded   <- .cache$excluded
settings   <- .cache$settings
date_run   <- .cache$date_run
irw_ver    <- attr(.cache$irw_version, "irw_version")

fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
sgn <- function(x, d = 3) {
  x <- round(x, d)
  paste0(ifelse(x > 0, "+", ifelse(x < 0, "−", "")), fmt(abs(x), d))
}

kind_label <- c(ability = "ability", propensity = "propensity")
src_levels <- c("AI models", "Exam (choi)", "Vocabulary (previc_bohn)")
results <- results |>
  mutate(src = case_when(source == "ai" ~ "AI models",
                         startsWith(table, "choi") ~ "Exam (choi)",
                         TRUE ~ "Vocabulary (previc_bohn)"),
         src = factor(src, levels = rev(src_levels)),
         matrix = case_when(source == "ai" & draw == 0 ~ "AI: all items",
                            source == "ai" ~ "AI: 300-item draw",
                            TRUE ~ "Human draw"))

primary   <- prep$benchmark[prep$panel == "primary"]
secondary <- prep$benchmark[prep$panel == "secondary"]
bench_order <- prep |> arrange(panel, kind, benchmark) |> pull(benchmark)
n_ai <- function(b) prep$n_models[match(b, prep$benchmark)]

# "a to b", or one value when there is only one matrix.
span <- function(x) if (length(x) == 1) sgn(x) else paste0(sgn(min(x)), " to ", sgn(max(x)))

# Where a benchmark's AI matrices fall relative to its human draws.
rng <- function(b, s, col) {
  x <- results[[col]][results$benchmark == b & results$source == s]
  span(x)
}
position <- function(b, col) {
  ai <- results[[col]][results$benchmark == b & results$source == "ai"]
  hu <- results[[col]][results$benchmark == b & results$source == "human"]
  if (min(ai) > max(hu)) "above" else if (max(ai) < min(hu)) "below" else "overlap"
}
n_above <- function(benches, col) sum(vapply(benches, position, "", col = col) == "above")
which_not_above <- function(benches, col) {
  benches[vapply(benches, position, "", col = col) != "above"]
}
```

::: {.callout-note}
## Draft

This page is a draft and is not yet listed on the vignettes index. It reports
the full run specified in
[datapages/irw#186](https://github.com/datapages/irw/issues/186):
`r nrow(results)` matrices from `r nrow(prep)` benchmarks.
:::

## Motivation

Item-level results from AI benchmarks now form response matrices: models as
rows, benchmark items as columns, correct or not in the cells. Psychometric
tools are starting to be applied to them. OpenEval [@jiang2026aievaluationrequire]
reports classical test theory and item factor analysis, and names IRT as
unexplored.

A model comparison for these matrices has an obvious reference point: what the
same comparison gives on human data of the same shape. If going from a Rasch
model to a 2PL, or from one factor to two, buys the same predictive gain for AI
respondents as for people, the usual IRT toolkit carries over with its usual
trade-offs. If the gains differ, that difference says something about how AI
models "respond".

We use the InterModel Vigorish [IMV, @domingue2024imvlens], which scores models
by how well they predict held-out responses. With 50 to 170 respondents per
benchmark, a predictive, cross-validated criterion is a safer choice than
likelihood-ratio tests or limited-information fit statistics. Their asymptotics
assume far more respondents than this.

## Data

### AI benchmarks (OpenEval, in IRW)

Every benchmark is an IRW table, `jiang_2026_openeval_<name>`, built from
OpenEval's item-level release (CC BY-NC 4.0). The IRW processing keeps one run
and one metric per model and item, and one judge for judged benchmarks. Models
are kept if they answered at least `r 100 * settings$min_coverage`% of a
benchmark's items.

The **primary panel** has the seven benchmarks with 100 or more such models. The
**secondary panel** has three with 50 to 60. Results there are noisy, and are
shown separately.

```{r}
#| label: tbl-prep
#| tbl-cap: "OpenEval benchmarks in the comparison. \"Propensity\" benchmarks score a behaviour, not correctness, so 1 is not \"better\"."

prep |>
  arrange(match(benchmark, bench_order)) |>
  transmute(
    Benchmark = benchmark,
    Panel = panel,
    Kind = kind,
    `1 = ...` = coding,
    Models = n_models,
    `...aliases merged` = n_models_dealiased,
    Items = n_items,
    `Items varying` = n_items_varying,
    `Share 1` = fmt(p_correct, 2),
    Temperature = temperatures
  ) |>
  kable()
```

Three decisions shape these matrices.

- **IFEval is dichotomised.** Each prompt carries one to three instructions,
  and the IRW table records how many were followed. We use IFEval's
  prompt-level strict accuracy, which scores 1 only when every instruction is
  followed.
- **Four benchmarks measure a propensity, not an ability.** SORRY-Bench,
  AdvBench and HarmBench score whether a model complied with a harmful request.
  OR-Bench scores whether it refused a request that only looks harmful. A judge
  (a fine-tuned model, an LLM or a refusal-string match) assigns the score, and
  judges make mistakes.
- **Respondents are model names as OpenEval records them.** Dated snapshots
  and near-duplicates (for example `gpt-4o-mini` and `gpt-4o-mini-2024-07-18`)
  are kept as separate respondents. The table shows how many models remain
  after merging obvious aliases.

Other OpenEval tables in IRW were left out:

```{r}
#| label: tbl-excluded
#| tbl-cap: "OpenEval tables in IRW not used here."

excluded |> transmute(Table = paste0("jiang_2026_openeval_", table), Reason = reason) |> kable()
```

### Human comparison tables

Only a handful of binary IRW tables are naturally the size of these benchmarks,
and one of those is sparse. Instead, we take dense, right/wrong-scored human
tables and **subsample** them. Each draw picks as many respondents as the
benchmark has models, and then `r settings$j_common` items from those that vary
in the sample. Both sources have barely more than `r settings$j_common` items,
so a draw usually keeps every varying item instead
(`r min(results$J[results$source == "human"])`–`r max(results$J[results$source == "human"])`
items). HarmBench has only `r prep$n_items_varying[prep$benchmark == "harmbench"]`
varying items, so its human draws are cut to that number.

The AI matrices are cut the same way. Each primary benchmark gets
`r settings$n_draws` draws of `r settings$j_common` items, plus one fit on all
of its items. HarmBench gets only the full fit. Each secondary benchmark gets one
draw. Every benchmark gets `r settings$n_draws` draws from each human source.

```{r}
#| label: tbl-human
#| tbl-cap: "Human IRW tables subsampled for the comparison."

tibble(
  Table = c("choi_2026_cmsce_2019_1", "previc_bohn2023"),
  Content = c("Korean Clinical Medical Science Comprehensive Examination (one form)",
              "Parent-reported expressive vocabulary, children aged 3-8 (German)"),
  `Full size` = c("4,190 × 302", "1,190 × 379")
) |> kable()
```

We restricted the human sources to right/wrong-scored tests on purpose.
Personality inventories are larger and denser, but they contain reverse-keyed
items. A Rasch model or a positive slope prior fits those items badly, which
would inflate the 2PL's apparent gain for reasons that have nothing to do with
who the respondents are.

## Method

Each matrix gets the same treatment:

1. **Drop near-constant items.** An item needs at least
   `r settings$min_outcome` correct and `r settings$min_outcome` incorrect
   responses.
2. **Split responses into `r settings$nfold` folds.** Responses, not
   respondents, are split, so every model and every item appears in each
   training set. All three models share one fold assignment.
3. **In each fold, fit three models** with `mirt` to the training responses:
   - a Rasch model;
   - a 2PL with a lognormal(0, 1) prior on slopes, as in the site's
     [IMV tutorial](imv.qmd);
   - an exploratory two-factor 2PL, with the same prior on first-factor slopes
     and a normal(0, 1) prior on second-factor slopes.
4. **Score the held-out responses** using EAP ability estimates and each
   item's trace line.
5. **Compute the IMV** from Rasch to 2PL and from 2PL to two factors, then
   average over folds.

The IMV turns each model's held-out likelihood into the weight of a coin with
the same predictive accuracy, then reports the proportional change in that
weight. An IMV of 0.01 means the richer model is equivalent to a coin weighted
1% more heavily: 0.707 rather than 0.700, for example. Negative values mean the
richer model predicts held-out responses *worse*.

## Results

```{r}
#| label: plot-fn

comparison_levels <- c(imv_rasch_2pl = "Rasch → 2PL", imv_2pl_2f = "2PL → two factors")
matrix_shapes <- c("AI: all items" = 18, "AI: 300-item draw" = 16, "Human draw" = 15)

# One short line: plotly draws row strips rotated on the right.
strip_for <- function(b) paste0(b, " (", n_ai(b), ")")

to_long <- function(d) {
  d |>
    pivot_longer(c(imv_rasch_2pl, imv_2pl_2f), names_to = "comparison", values_to = "imv") |>
    mutate(comparison = factor(comparison_levels[comparison], levels = comparison_levels),
           bench = factor(strip_for(benchmark), levels = strip_for(bench_order)))
}

imv_figure <- function(benches, legend_y = -0.08) {
  r <- results |> filter(benchmark %in% benches)
  m <- to_long(r) |>
    mutate(tip = paste0(benchmark, ": ", src, " (", matrix, ")",
                        "<br>", N, " respondents × ", J, " items",
                        "<br>IMV ", sgn(imv, 4), " (mean of ", settings$nfold, " folds)"))
  f <- folds |>
    inner_join(select(r, key, benchmark, src), by = "key") |>
    to_long()

  set.seed(1)
  p <- ggplot(m, aes(x = imv, y = src, colour = src == "AI models")) +
    geom_vline(xintercept = 0, colour = irw_grey, linetype = "dashed") +
    geom_point(data = f, alpha = 0.25, size = 1,
               position = position_jitter(width = 0, height = 0.15)) +
    geom_point(aes(shape = matrix, text = tip), size = 3, stroke = 1) +
    scale_colour_manual(values = c(`TRUE` = irw_red, `FALSE` = irw_blue), guide = "none") +
    scale_shape_manual(values = matrix_shapes, name = NULL) +
    facet_grid(bench ~ comparison, scales = "free_x") +
    labs(x = "IMV (richer model vs simpler model)", y = NULL) +
    theme(legend.position = "bottom",
          panel.spacing.y = unit(0.6, "lines"),
          panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())

  plotly::ggplotly(p, tooltip = "text") |>
    plotly::layout(legend = list(orientation = "h", y = legend_y)) |>
    plotly::config(displayModeBar = FALSE)
}
```

### Overview

```{r}
#| label: fig-overview
#| fig-cap: "IMV for every primary-panel matrix. One dot per matrix: human draws matched to each benchmark's size, AI draws of 300 items, and AI fits on a benchmark's full item set (HarmBench's 223 items, which its human draws also use, counts as a full set). The black line is each row's median. The dashed line is no gain."
#| fig-height: 4.4

ov <- results |>
  filter(benchmark %in% primary) |>
  mutate(who = factor(case_when(source == "human" ~ "People",
                                draw == 0 ~ "AI: full item set",
                                TRUE ~ "AI: 300-item draws"),
                      levels = c("People", "AI: 300-item draws", "AI: full item set"))) |>
  pivot_longer(c(imv_rasch_2pl, imv_2pl_2f), names_to = "comparison", values_to = "imv") |>
  mutate(comparison = factor(comparison_levels[comparison], levels = comparison_levels),
         tip = paste0(benchmark, ": ", src, " (", matrix, ")<br>IMV ", sgn(imv, 4)))
ov_med <- ov |> group_by(comparison, who) |> summarise(imv = median(imv), .groups = "drop")

set.seed(2)
p_ov <- ggplot(ov, aes(x = imv, y = who, colour = who)) +
  geom_vline(xintercept = 0, colour = irw_grey, linetype = "dashed") +
  geom_point(aes(text = tip), size = 2.6, alpha = 0.75,
             position = position_jitter(width = 0, height = 0.18)) +
  geom_segment(data = ov_med, colour = "black", linewidth = 1.2,
               aes(x = imv, xend = imv, y = as.numeric(who) - 0.32, yend = as.numeric(who) + 0.32)) +
  scale_colour_manual(values = c("AI: full item set" = irw_red, "AI: 300-item draws" = irw_red,
                                 "People" = irw_blue), guide = "none") +
  facet_wrap(~ comparison, scales = "free_x") +
  labs(x = "IMV (richer model vs simpler model)", y = NULL) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        panel.spacing.x = unit(2, "lines"))

plotly::ggplotly(p_ov, tooltip = "text") |> plotly::config(displayModeBar = FALSE)
```

```{r}
#| label: ov-numbers
ov_n <- function(w) sum(ov$who == w & ov$comparison == comparison_levels[["imv_2pl_2f"]])
ov_m <- function(w, cmp) sgn(ov_med$imv[ov_med$who == w & ov_med$comparison == comparison_levels[[cmp]]])
```

The median gain from a second factor is `r ov_m("People", "imv_2pl_2f")` across
the `r ov_n("People")` human draws, `r ov_m("AI: 300-item draws", "imv_2pl_2f")`
across the `r ov_n("AI: 300-item draws")` AI draws of 300 items, and
`r ov_m("AI: full item set", "imv_2pl_2f")` across the
`r ov_n("AI: full item set")` full-item-set fits. So the gap does not come from
the AI matrices being larger: the size-matched draws show it too. For
Rasch → 2PL the rows largely overlap (medians
`r ov_m("People", "imv_rasch_2pl")`, `r ov_m("AI: 300-item draws", "imv_rasch_2pl")`
and `r ov_m("AI: full item set", "imv_rasch_2pl")`). GPQA's matrices are the AI
points far to the right.

### Primary panel by benchmark

Each large dot below is one matrix: a benchmark's full item set (diamond), a
300-item draw from it (circle), or a human table subsampled to the benchmark's
size (square). Its value is the IMV averaged over the `r settings$nfold`
cross-validation folds. The small faint dots are the individual folds.

```{r}
#| label: fig-primary
#| fig-cap: "Cross-validated IMV for the seven primary benchmarks, AI models (red) against human draws of the same size (blue). Rows are labelled with the benchmark and its number of models; the top three are ability benchmarks and the bottom four propensity benchmarks. The dashed line is no gain; left of it, the richer model predicts held-out responses worse."
#| fig-height: 13

imv_figure(primary)
```

```{r}
#| label: tbl-ranges
#| tbl-cap: "Range of fold-averaged IMV across matrices, by benchmark and source."

results |>
  group_by(benchmark, src) |>
  summarise(r2pl = span(imv_rasch_2pl), f2 = span(imv_2pl_2f),
            n = n(), .groups = "drop") |>
  mutate(src = factor(src, levels = src_levels)) |>
  arrange(match(benchmark, bench_order), src) |>
  transmute(Benchmark = benchmark, Respondents = as.character(src), Matrices = n,
            `Rasch → 2PL` = r2pl, `2PL → two factors` = f2) |>
  kable()
```

What the primary panel shows:

- **A second factor pays off for AI respondents and not for people.** For
  `r n_above(primary, "imv_2pl_2f")` of the `r length(primary)` primary
  benchmarks, every AI matrix gains more from a second factor than any human
  draw does. The exception is
  `r paste(which_not_above(primary, "imv_2pl_2f"), collapse = ", ")`, where one
  300-item draw gains only `r sgn(min(results$imv_2pl_2f[results$benchmark %in% which_not_above(primary, "imv_2pl_2f") & results$source == "ai"]), 4)`.
  On the medical-school exam a second factor always *hurts*
  (`r sgn(min(results$imv_2pl_2f[results$src == "Exam (choi)" & results$benchmark %in% primary]))` to
  `r sgn(max(results$imv_2pl_2f[results$src == "Exam (choi)" & results$benchmark %in% primary]))`).
  On the vocabulary checklist it is close to zero.
- **The size of the gain varies a lot between benchmarks.** It is largest for
  IFEval (`r rng("ifeval", "ai", "imv_2pl_2f")`) and HarmBench
  (`r rng("harmbench", "ai", "imv_2pl_2f")`), and smallest for AdvBench and
  TruthfulQA. Both kinds of benchmark appear at both ends, so the gain is not
  specific to propensity benchmarks.
- **Discrimination parameters mostly don't separate AI from people.** For
  Rasch → 2PL, the AI matrices fall within or below the human range for most
  benchmarks. They are above it only for
  `r paste(primary[vapply(primary, position, "", col = "imv_rasch_2pl") == "above"], collapse = " and ")`.
  GPQA stands out (`r rng("gpqa", "ai", "imv_rasch_2pl")`, against
  `r rng("gpqa", "human", "imv_rasch_2pl")` for its human draws). GPQA is
  four-option multiple choice. A quarter of its items are answered correctly by
  27% of models or fewer, close to the 25% chance rate. If models guess on
  those items, a 2PL can absorb some of that through flat slopes. A model with a
  lower asymptote would be the direct test.
- **Cutting a benchmark to 300 items changes little.** The 300-item draws
  usually land close to the fit on all of a benchmark's items. TruthfulQA is
  the exception for the second factor, which suggests the second dimension
  there is carried by a subset of items that some draws miss.

### Secondary panel

```{r}
#| label: fig-secondary
#| fig-cap: "Cross-validated IMV for the three secondary benchmarks (50 to 60 models). At this size the richer models overfit human draws, which pushes the human IMVs below zero."
#| fig-height: 6.5

imv_figure(secondary, legend_y = -0.2)
```

At 50 to 60 respondents every human draw goes negative for the second factor
(`r sgn(min(results$imv_2pl_2f[results$source == "human" & results$benchmark %in% secondary]))` to
`r sgn(max(results$imv_2pl_2f[results$source == "human" & results$benchmark %in% secondary]))`).
With so few people to estimate from, an extra dimension mostly fits noise.
Against that, two AI matrices still gain from it: BBQ
(`r rng("bbq-helm", "ai", "imv_2pl_2f")`, the largest gain on the page) and
CulturalBench (`r rng("culturalbench", "ai", "imv_2pl_2f")`). BBQ pairs
ambiguous questions, where the right answer is "unknown", with disambiguated
ones that test ordinary reading. Models are known to trade these off, which
makes it a plausible source of a second dimension. Hi-ToM gains nothing from
either step. Every Hi-ToM model was run at temperature 1, so its responses
carry more sampling noise.

## What this comparison can and cannot say

- **Two human sources.** IRW has few dense, right/wrong-scored tables with 300
  or more items. Apart from other forms of the same Korean exam, the only other
  candidate, `vermeiren_2022_vocab` (236 × 668), was dropped after its first
  fits. Its 668 item names turn out to be about 483 words with whitespace
  variants, and some 57,000 person-by-word cells appear more than once. The
  human reference is therefore a range set by two tests that also disagree with
  each other.
- **The human sources are not matched on content.** Benchmarks mix many kinds
  of task, such as instruction types in IFEval or harm categories in
  SORRY-Bench. The human tests were built to measure one thing. So a second
  factor that pays off only for AI could reflect benchmark content as much as
  respondent type. Same-item comparisons, where models take human IRW
  instruments, would separate the two
  ([datapages/irw#306](https://github.com/datapages/irw/issues/306)).
- **AI respondents are not independent.** Many models are fine-tunes or
  snapshots of the same base weights, so the effective number of respondents
  is smaller than the count. A second factor could pick up model families as
  well as item content.
- **Prior choice.** Both richer models use a slope prior. Without it, 2PL and
  two-factor fits at this sample size can overfit badly, and the IMV would then
  mostly measure that overfitting.
- **Decoding settings.** GPQA and Hi-ToM models were all run with non-zero
  temperature, and IFEval mixes the two. The other benchmarks are greedy.
- **Training-data contamination.** Some benchmark items are likely to have
  been seen in training, which would change how they behave for some models.

## Next steps

- Check what the second factor is: whether its loadings follow item content
  (BBQ's ambiguous vs disambiguated questions is the clearest test) and whether
  its scores follow model family.
- Fit a model with a lower asymptote to GPQA.
- Add a human test with deliberately multidimensional content, to see whether
  it lands among the AI points.
- Compare with and without the slope priors.

---

## Reproducibility

{{< include ../components/_source-links.qmd >}}

{{< include ../components/_data-version.qmd >}}

All data are read from IRW (version `r irw_ver` at compute time). AI benchmark
responses are the `jiang_2026_openeval_*` tables, from OpenEval's item-level
release (CC BY-NC 4.0).

Results computed `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
Rscript vignettes/ai_benchmark_imv_compute.R
quarto::quarto_render("vignettes/ai_benchmark_imv.qmd")
```

---

## References
