---
title: "AI Models as Respondents: An IMV Model Comparison"
description: |
  Does the extra structure of a 2PL, or of a second factor, improve
  prediction the same way when the respondents are AI models as when they are
  people? We fit Rasch, 2PL and two-factor models to item-level AI benchmark
  results from OpenEval and to human IRW tables subsampled to the same size,
  and compare them with the cross-validated InterModel Vigorish (IMV).
date: today
bibliography: ai_benchmark_imv_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(knitr)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "ai_benchmark_imv_data/ai_benchmark_imv_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache      <- readRDS(cache_path)
  results     <- .cache$results
  folds       <- .cache$folds
  prep        <- .cache$prep_counts
  settings    <- .cache$settings
  hf_revision <- .cache$hf_revision
  date_run    <- .cache$date_run
  pilot       <- .cache$pilot
} else {
  results  <- tibble(key = character(), source = character(), benchmark = character(),
                     table = character(), draw = integer(), N = integer(), J = integer(),
                     imv_rasch_2pl = numeric(), imv_2pl_2f = numeric(),
                     all_converged = logical())
  folds    <- tibble()
  prep     <- tibble(benchmark = character())
  settings <- list(j_common = 300, n_draws = 2, nfold = 5, min_coverage = 0.9, min_outcome = 5)
  hf_revision <- "23a1ded985b3c2cdaaddb27581a35bfabe0ad7e7"
  date_run <- Sys.Date()
  pilot    <- TRUE
}

fmt  <- function(x, d = 3) formatC(x, format = "f", digits = d)
pct  <- function(x) paste0(round(100 * x), "%")

group_label <- function(source, table, draw) {
  case_when(
    source == "ai" & draw == 0 ~ "AI: all items",
    source == "ai"             ~ "AI: item draws",
    TRUE                       ~ paste0("Human: ", table)
  )
}
results <- results |> mutate(group = group_label(source, table, draw))
```

::: {.callout-warning}
## Pilot

This is a draft built from a deliberately small run: two benchmarks, three
human source tables and `r settings$n_draws` draws of each. It exists to check
the pipeline and to see whether the question is worth a full run. The numbers
below should not be quoted as findings. See
[datapages/irw#186](https://github.com/datapages/irw/issues/186).
:::

## Motivation

Item-level results from AI benchmarks now form response matrices: models as
rows, benchmark items as columns, correct or not in the cells. Psychometric
tools are starting to be applied to them. OpenEval [@jiang2026aievaluationrequire]
reports classical test theory and item factor analysis, and names IRT as
unexplored.

A model comparison for these matrices has an obvious reference point: what the
same comparison gives on human data of the same shape. If going from a Rasch
model to a 2PL, or from one factor to two, buys the same predictive gain for AI
respondents as for people, the usual IRT toolkit carries over with its usual
trade-offs. If the gains differ, that difference says something about how AI
models "respond". Examples would be much larger gains from discrimination
parameters, or a second dimension that pays for itself only for AI.

We use the InterModel Vigorish [IMV, @domingue2024imvlens], which scores models
by how well they predict held-out responses. With only 100 to 150 respondents
per benchmark, a predictive, cross-validated criterion is a safer choice than
likelihood-ratio tests or limited-information fit statistics. Their asymptotics
assume far more respondents than this.

## Data

### AI benchmarks (OpenEval)

Responses come from OpenEval, read directly from Hugging Face
(`Open-Eval-Commons/OpenEval`), pinned to revision
`r substr(hf_revision, 1, 8)`. They are **not** part of IRW. The data is
licensed CC BY-NC 4.0, so only fitted summaries are kept here.

```{r}
#| label: tbl-prep
#| tbl-cap: "OpenEval benchmarks used in the pilot."

if (nrow(prep)) {
  prep |>
    transmute(
      Benchmark = benchmark,
      `Coding (1 = ...)` = coding,
      `Models (>= 90% coverage)` = n_models,
      `...after merging aliases` = n_models_dealiased,
      Items = n_items,
      `Items with >= 5 of each outcome` = n_items_varying,
      `Mean score` = fmt(p_correct, 2),
      `Replicate runs dropped` = n_replicate_rows
    ) |>
    kable()
}
```

Three coding decisions shape these matrices.

- **IFEval is dichotomised.** OpenEval stores the share of a prompt's
  instructions that were followed (0, 1/3, 1/2, 2/3 or 1).
  `r if (nrow(prep)) paste0(format(prep$n_fractional[prep$benchmark == "ifeval"], big.mark = ","), " of ", format(prep$n_rows[prep$benchmark == "ifeval"], big.mark = ","), " responses")` are fractional.
  We use IFEval's prompt-level strict accuracy, which scores 1 only when every
  instruction is followed.
- **SORRY-Bench measures a propensity, not an ability.** A fine-tuned judge
  labels each response as fulfilling (1) or refusing (0) a potentially unsafe
  request, so a "high score" means a less cautious model. The judge makes
  mistakes. About 3.5% of responses it labels as fulfilling begin with a
  refusal phrase such as "I can't", compared with 70% of those it labels as
  refusals.
- **Respondents are model names as OpenEval records them.** Dated snapshots
  and near-duplicates (for example `gpt-4o-mini` and `gpt-4o-mini-2024-07-18`)
  are kept as separate respondents. The table shows how many models remain
  after merging obvious aliases.

### Human comparison tables

Only a handful of binary IRW tables are naturally the size of these benchmarks
(roughly 120–150 respondents × 400+ items), and one of those is sparse.
Instead, we take dense, right/wrong-scored human tables and **subsample** them.
Each draw picks as many respondents as the benchmark has models, and then
`r settings$j_common` items from those that vary in the sample. The AI matrices
are cut the same way. Each benchmark gets `r settings$n_draws` draws of
`r settings$j_common` items, plus one fit on all of its items for reference.

```{r}
#| label: tbl-human

tibble(
  Table = c("choi_2026_cmsce_2019_1", "previc_bohn2023", "vermeiren_2022_vocab"),
  Content = c("Korean Clinical Medical Science Comprehensive Examination (one form)",
              "Parent-reported expressive vocabulary, children aged 3-8 (German)",
              "University students' English vocabulary"),
  `Full size` = c("4,190 × 302", "1,190 × 379", "236 × 668")
) |> kable()
```

We restricted the human sources to right/wrong-scored tests on purpose.
Personality inventories are larger and denser, but they contain reverse-keyed
items. A Rasch model or a positive slope prior fits those items badly, which
would inflate the 2PL's apparent gain for reasons that have nothing to do with
who the respondents are.

## Method

Each matrix gets the same treatment:

1. **Drop near-constant items.** An item needs at least
   `r settings$min_outcome` correct and `r settings$min_outcome` incorrect
   responses.
2. **Split responses into `r settings$nfold` folds.** Responses, not
   respondents, are split, so every model and every item appears in each
   training set. All three models share one fold assignment.
3. **In each fold, fit three models** with `mirt` to the training responses:
   - a Rasch model;
   - a 2PL with a lognormal(0, 1) prior on slopes, as in the site's
     [IMV tutorial](imv.qmd);
   - an exploratory two-factor 2PL, with the same prior on first-factor slopes
     and a normal(0, 1) prior on second-factor slopes.
4. **Score the held-out responses** using EAP ability estimates and each
   item's trace line.
5. **Compute the IMV** from Rasch to 2PL and from 2PL to two factors, then
   average over folds.

The IMV turns each model's held-out likelihood into the weight of a coin with
the same predictive accuracy, then reports the proportional change in that
weight. An IMV of 0.01 means the richer model is equivalent to a coin weighted
1% more heavily: 0.707 rather than 0.700, for example. Negative values mean the
richer model predicts held-out responses *worse*.

## Results

```{r}
#| label: fig-imv
#| fig-cap: "Cross-validated IMV for each matrix. Each point is one matrix (a full benchmark, an item draw from it, or a subsample of a human table), averaged over folds. Hollow points had at least one fit that did not converge."
#| fig-height: 6
#| fig-width: 10

if (nrow(results)) {
  plot_df <- results |>
    pivot_longer(c(imv_rasch_2pl, imv_2pl_2f), names_to = "comparison", values_to = "imv") |>
    mutate(
      comparison = recode(comparison,
                          imv_rasch_2pl = "Rasch → 2PL",
                          imv_2pl_2f    = "2PL → two factors"),
      comparison = factor(comparison, levels = c("Rasch → 2PL", "2PL → two factors")),
      group = factor(group, levels = rev(c("AI: all items", "AI: item draws",
                                           sort(unique(group[source == "human"])))))
    )

  ggplot(plot_df, aes(x = imv, y = group, colour = source)) +
    geom_vline(xintercept = 0, colour = irw_grey, linetype = "dashed") +
    geom_point(aes(shape = all_converged), size = 3, stroke = 1.1) +
    scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 1), guide = "none") +
    scale_colour_manual(values = c(ai = irw_red, human = irw_blue),
                        labels = c(ai = "AI models", human = "People")) +
    facet_grid(benchmark ~ comparison, scales = "free_x") +
    labs(x = "IMV (richer model vs simpler model)", y = NULL, colour = NULL) +
    theme(legend.position = "bottom", panel.spacing = unit(1.2, "lines"))
}
```

```{r}
#| label: summaries

med <- results |>
  filter(draw > 0) |>
  group_by(benchmark, source) |>
  summarise(rasch_2pl = median(imv_rasch_2pl), twopl_2f = median(imv_2pl_2f),
            n = n(), .groups = "drop")

getmed <- function(b, s, col) {
  x <- med[[col]][med$benchmark == b & med$source == s]
  if (length(x)) fmt(x) else "NA"
}
n_nonconv <- sum(!results$all_converged)
```

```{r}
#| label: tbl-medians
#| tbl-cap: "Median IMV across the matched-size draws (J items each)."

if (nrow(med)) {
  med |>
    transmute(Benchmark = benchmark,
              Respondents = if_else(source == "ai", "AI models", "People"),
              Draws = n,
              `Rasch → 2PL` = fmt(rasch_2pl),
              `2PL → two factors` = fmt(twopl_2f)) |>
    kable()
}
```

<!-- RESULTS-TEXT -->

## What this pilot can and cannot say

- **Small numbers of draws.** Each human source contributes only
  `r settings$n_draws` subsamples. The spread across draws, and across the
  three human tables, is a rough guide at best.
- **The human draws overlap.** `vermeiren_2022_vocab` has only 236
  respondents, so two draws of 120–150 of them share many people.
- **The human sources are not matched on content.** A medical licensing exam,
  a parent-report vocabulary checklist and a student vocabulary test are all
  cognitive, but none resembles instruction following or refusal behaviour.
  Matching on shape isolates respondent type from sample size, not from
  construct.
- **AI respondents are not independent.** Many models are fine-tunes or
  snapshots of the same base weights, so the effective number of respondents
  is smaller than the count. This is part of what makes AI respondents
  different, not a defect to correct.
- **Prior choice.** Both richer models use a slope prior. Without it, 2PL and
  two-factor fits at this sample size can overfit badly, and the IMV would then
  mostly measure that overfitting. A no-prior comparison is left for the full run.
- **Decoding settings are mixed.**
  `r if (nrow(prep)) prep$n_models_nongreedy[prep$benchmark == "ifeval"] else "Some"`
  IFEval models were run with non-zero temperature and the rest greedily. The
  pilot keeps them all.

## Next steps for a full run

- Add GPQA (`chain_of_thought_correctness`), OR-Bench (refusal) and, after
  checking its coding, TruthfulQA.
- Increase draws per human source, and add human tables whose content is closer
  to each benchmark where IRW has them.
- Run a greedy-only sensitivity analysis and a version with merged aliases.
- Compare with and without the slope priors.

---

## Reproducibility

{{< include ../components/_source-links.qmd >}}

{{< include ../components/_data-version.qmd >}}

AI benchmark responses: OpenEval, Hugging Face `Open-Eval-Commons/OpenEval` at
revision [`r substr(hf_revision, 1, 8)`](https://huggingface.co/datasets/Open-Eval-Commons/OpenEval/tree/`r hf_revision`),
read directly and not stored in IRW.

Results computed `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
Rscript vignettes/ai_benchmark_imv_compute.R
quarto::quarto_render("vignettes/ai_benchmark_imv.qmd")
```

---

## References
