---
title: "Asymmetric IRT Models Across IRW: Do LPE, RH, and AO Beat the 2PL?"
description: |
  Shim & Bonifay (2026) introduce three IRT models with asymmetric item
  response functions -- LPE, RH, and AO -- and show, via simplex-sampled
  synthetic data, that two of them fit about as well as a standard 2PL even
  when fitting propensity alone is compared, while the third (RH) is more
  parsimonious. They never apply the models to real data. We do: after a
  dedicated correctness/recovery validation gate, we fit a 2PL reference plus
  all three asymmetric models across dozens of binary IRW tables and score
  every model on genuinely held-out responses (InterModel Vigorish, not
  in-sample BIC). The two metrics tell very different stories: RH looks like
  the dominant winner by BIC, but is the weakest of the three -- often the
  worst model in the batch -- once judged on real predictive accuracy.
date: today
bibliography: asymmetric_irt_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"
irw_green <- "#4daf4a"

model_colors <- c("2PLM" = irw_grey, "RH" = irw_blue, "LPE" = irw_red, "AO" = irw_green)

results_path <- "asymmetric_irt_data/asymmetric_irt_results.rds"
gate_path    <- "asymmetric_irt_data/validation_gate_log.rds"
cache_ok     <- file.exists(results_path) && file.exists(gate_path)

if (cache_ok) {
  .res             <- readRDS(results_path)
  summary_df       <- .res$summary
  item_level_df    <- .res$item_level
  candidate_tables <- .res$candidate_tables
  n_all_candidates <- .res$n_all_candidates
  min_participants <- .res$min_participants
  is_pilot         <- isTRUE(.res$pilot)
  date_run         <- .res$date_run

  .gate              <- readRDS(gate_path)
  reduction_check    <- .gate$reduction_check
  noncrossing_ok     <- .gate$noncrossing_ok
  lpe_curves         <- .gate$lpe_curves
  theta_grid         <- .gate$theta_grid
  kappa_vals         <- .gate$kappa_vals
  recovery_df        <- .gate$recovery_df
  recovery_summary   <- .gate$recovery_summary
  imv_sanity         <- .gate$imv_sanity
  imv_null_ok        <- .gate$imv_null_ok
  gate_date_run      <- .gate$date_run
} else {
  summary_df <- tibble(
    table = character(), n_items = integer(), n_participants = integer(),
    construct_type = character(), n_heldout = integer(), imv_AO = numeric(),
    imv_LPE = numeric(), imv_RH = numeric(), winning_model = character()
  )
  item_level_df <- tibble(
    table = character(), model = character(), item = character(), a1 = numeric(),
    d = numeric(), shape_est = numeric(), kappa = numeric(), lambda_P = numeric()
  )
  candidate_tables <- character(); n_all_candidates <- 0L; min_participants <- 2000
  is_pilot <- FALSE; date_run <- Sys.Date()
  reduction_check <- tibble(model = character(), reference = character(),
                             max_abs_diff = numeric(), pass = logical())
  noncrossing_ok <- NA; lpe_curves <- matrix(); theta_grid <- numeric()
  kappa_vals <- numeric()
  recovery_df <- tibble(model = character(), item = integer(), a_true = numeric(),
                         d_true = numeric(), shape_true = numeric(), shape_est = numeric())
  recovery_summary <- tibble(model = character(), cor_shape = numeric(), mean_abs_error = numeric())
  imv_sanity <- tibble(model = character(), truth = character(), mean_imv = numeric(), sd_imv = numeric())
  imv_null_ok <- NA
  gate_date_run <- Sys.Date()
}

itemfit_path <- "asymmetric_irt_data/example_itemfit.rds"
if (file.exists(itemfit_path)) {
  empirical_bins_df <- readRDS(itemfit_path)$bins
} else {
  empirical_bins_df <- tibble(theta = numeric(), n = numeric(), obs_prop = numeric(), exp_prop = numeric())
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running, in order, from the project root:\n\n",
  "```r\nRscript vignettes/asymmetric_irt_data/validation_gate.R\n",
  "Rscript vignettes/asymmetric_irt_compute.R\n```\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the model implementation, the validation gate, the
dataset search, the compute script, and the writing below — was produced
largely by Claude (Anthropic), working from a task specification and with
human review. The custom item types used here are a non-trivial extension of
`mirt`; treat the methodological choices and results accordingly, and check
`asymmetric_irt_helpers.R` and the validation gate directly if you're relying
on the numbers shown.
:::

## Explore the model shapes

The three asymmetric models below are easier to feel than to read as
equations. Pick one, drag $\kappa$, and watch the fitted curve (blue) pull
away from the symmetric baseline it's compared against everywhere in this
vignette (grey dashed) — the standard logistic 2PL for LPE and AO, a
normal-ogive model for RH.

```{=html}
<div class="asym-widget" style="border:1px solid #dee2e6;border-radius:6px;padding:1rem 1.25rem;margin:1rem 0;">
  <div style="display:flex;flex-wrap:wrap;gap:1.5rem;align-items:center;margin-bottom:0.75rem;">
    <label style="font-weight:600;">Model:
      <select id="asym-model" style="margin-left:0.4rem;">
        <option value="LPE" selected>LPE</option>
        <option value="RH">RH</option>
        <option value="AO">AO</option>
      </select>
    </label>
    <label style="font-weight:600;flex:1;min-width:220px;">κ:
      <input id="asym-kappa" type="range" style="width:60%;vertical-align:middle;margin:0 0.5rem;">
      <span id="asym-kappa-val" style="font-family:monospace;"></span>
    </label>
  </div>
  <svg id="asym-svg" viewBox="0 0 640 380" style="width:100%;max-width:640px;height:auto;display:block;"></svg>
  <div id="asym-blurb" style="margin-top:0.75rem;font-size:0.95rem;"></div>
</div>

<script>
(function() {
  function erf(x) {
    // Abramowitz & Stegun 7.1.26 approximation, accurate to ~1.5e-7
    var sign = x < 0 ? -1 : 1;
    x = Math.abs(x);
    var a1 = 0.254829592, a2 = -0.284496736, a3 = 1.421413741,
        a4 = -1.453152027, a5 = 1.061405429, p = 0.3275911;
    var t = 1 / (1 + p * x);
    var y = 1 - (((((a5 * t + a4) * t) + a3) * t + a2) * t + a1) * t * Math.exp(-x * x);
    return sign * y;
  }
  function pnorm(x) { return 0.5 * (1 + erf(x / Math.SQRT2)); }
  function plogis(x) { return 1 / (1 + Math.exp(-x)); }

  function modelP(model, kappa, theta) {
    var z = theta; // a = 1, d = 0
    if (model === "LPE") return Math.pow(plogis(z), kappa);
    if (model === "AO")  return 1 - Math.pow(1 + kappa * Math.exp(z), -1 / kappa);
    if (model === "RH") {
      var sigma = Math.sqrt(2) * Math.pow(1 + Math.exp(-kappa * theta), -0.5);
      return pnorm(z / sigma);
    }
  }
  function symmetricP(model, theta) {
    return model === "RH" ? pnorm(theta) : plogis(theta);
  }

  var modelInfo = {
    LPE: {
      name: "LPE (Logistic Positive Exponent)",
      formula: "P(θ) = [Ψ(a(θ − b))]^κ",
      blurb: "Raises the logistic curve to a power. κ &lt; 1 lifts the whole curve up (more lenient); κ &gt; 1 pulls it down (stricter). κ = 1 recovers the 2PL exactly.",
      range: [0.05, 5], step: 0.05, def: 1
    },
    RH: {
      name: "RH (Residual Heteroscedasticity)",
      formula: "P(θ) = Φ(a(θ−b) / σ(θ)),  σ(θ) = √2·(1 + e^(−κθ))^(−1/2)",
      blurb: "Instead of a power transform, lets the response variance itself change with θ. κ = 0 recovers a symmetric normal-ogive model exactly.",
      range: [-4, 4], step: 0.05, def: 0
    },
    AO: {
      name: "AO (Aranda-Ordaz)",
      formula: "P(θ) = 1 − [1 + κ·e^(a(θ−b))]^(−1/κ)",
      blurb: "A different link function that also collapses to the 2PL at κ = 1, skewing toward one tail for other κ — similar in spirit to LPE, different in its algebra.",
      range: [0.05, 5], step: 0.05, def: 1
    }
  };

  var selEl = document.getElementById("asym-model");
  var kappaEl = document.getElementById("asym-kappa");
  var kappaValEl = document.getElementById("asym-kappa-val");
  var svg = document.getElementById("asym-svg");
  var blurbEl = document.getElementById("asym-blurb");

  var W = 640, H = 380, mL = 50, mR = 20, mT = 10, mB = 40;
  var plotW = W - mL - mR, plotH = H - mT - mB;
  function px(theta) { return mL + (theta + 4) / 8 * plotW; }
  function py(p) { return mT + (1 - p) * plotH; }

  function pathFor(fn) {
    var d = "";
    for (var i = 0; i <= 160; i++) {
      var theta = -4 + i * (8 / 160);
      var x = px(theta), y = py(fn(theta));
      d += (i === 0 ? "M" : "L") + x.toFixed(2) + "," + y.toFixed(2) + " ";
    }
    return d;
  }

  function staticAxesSvg() {
    var s = "";
    s += '<line x1="' + mL + '" y1="' + mT + '" x2="' + mL + '" y2="' + (mT + plotH) + '" stroke="#888"/>';
    s += '<line x1="' + mL + '" y1="' + (mT + plotH) + '" x2="' + (mL + plotW) + '" y2="' + (mT + plotH) + '" stroke="#888"/>';
    [-4, -2, 0, 2, 4].forEach(function(t) {
      var x = px(t);
      s += '<line x1="' + x + '" y1="' + (mT + plotH) + '" x2="' + x + '" y2="' + (mT + plotH + 5) + '" stroke="#888"/>';
      s += '<text x="' + x + '" y="' + (mT + plotH + 18) + '" font-size="11" text-anchor="middle" fill="#333">' + t + '</text>';
    });
    [0, 0.25, 0.5, 0.75, 1].forEach(function(p) {
      var y = py(p);
      s += '<line x1="' + (mL - 5) + '" y1="' + y + '" x2="' + mL + '" y2="' + y + '" stroke="#888"/>';
      s += '<text x="' + (mL - 9) + '" y="' + (y + 4) + '" font-size="11" text-anchor="end" fill="#333">' + p + '</text>';
    });
    s += '<text x="' + (mL + plotW / 2) + '" y="' + (H - 4) + '" font-size="12" text-anchor="middle" fill="#333">θ</text>';
    s += '<text x="12" y="' + (mT + plotH / 2) + '" font-size="12" text-anchor="middle" fill="#333" transform="rotate(-90 12 ' + (mT + plotH / 2) + ')">P(θ)</text>';
    return s;
  }

  function draw() {
    var model = selEl.value;
    var info = modelInfo[model];
    var kappa = parseFloat(kappaEl.value);
    kappaValEl.textContent = "κ = " + kappa.toFixed(2);

    var baselineD = pathFor(function(theta) { return symmetricP(model, theta); });
    var selectedD = pathFor(function(theta) { return modelP(model, kappa, theta); });

    svg.innerHTML =
      staticAxesSvg() +
      '<path d="' + baselineD + '" fill="none" stroke="#9a9a9a" stroke-width="2" stroke-dasharray="4,3"/>' +
      '<path d="' + selectedD + '" fill="none" stroke="#2166ac" stroke-width="2.5"/>' +
      '<line x1="' + (mL + 12) + '" y1="18" x2="' + (mL + 34) + '" y2="18" stroke="#2166ac" stroke-width="2.5"/>' +
      '<text x="' + (mL + 40) + '" y="22" font-size="11" fill="#333">Selected model</text>' +
      '<line x1="' + (mL + 12) + '" y1="34" x2="' + (mL + 34) + '" y2="34" stroke="#9a9a9a" stroke-width="2" stroke-dasharray="4,3"/>' +
      '<text x="' + (mL + 40) + '" y="38" font-size="11" fill="#333">Symmetric baseline</text>';

    var lambdaP = modelP(model, kappa, 1) + modelP(model, kappa, -1) - 1;
    var direction = lambdaP > 0.005
      ? "— leniency (the curve sits above the symmetric baseline on average)"
      : lambdaP < -0.005
      ? "— strictness (the curve sits below the symmetric baseline on average)"
      : "— essentially symmetric at this κ";

    blurbEl.innerHTML =
      "<strong>" + info.name + "</strong>: " + info.formula + "<br>" +
      info.blurb + "<br><br>" +
      "Л_P at ±1 logit from difficulty, at the current κ: <strong>" + lambdaP.toFixed(3) + "</strong> " + direction + ". " +
      "This is exactly the Л_P statistic used throughout the Results section below.";
  }

  function onModelChange() {
    var info = modelInfo[selEl.value];
    kappaEl.min = info.range[0];
    kappaEl.max = info.range[1];
    kappaEl.step = info.step;
    kappaEl.value = info.def;
    draw();
  }

  selEl.addEventListener("change", onModelChange);
  kappaEl.addEventListener("input", draw);
  onModelChange();
})();
</script>
```

## Background

Every IRT model used elsewhere on this site — the 2PL, the graded response
model, the [Dutch Identity](dutch_identity.qmd)'s loglinear framework — has a
**symmetric** item response function: the probability of a correct or
endorsing response rises smoothly and symmetrically around the item's
difficulty. But nothing requires that. A guessing-prone multiple-choice item,
for instance, has a response curve that rises quickly from a nonzero floor
and then flattens gradually near the top — asymmetric around its inflection
point.

@shimbonifay2026 study three such asymmetric models, each adding a single
shape parameter $\kappa$ to a standard two-parameter item:

- **LPE** (Logistic Positive Exponent): $P(\theta) = [\Psi(z)]^{\kappa}$, where
  $\Psi$ is the logistic CDF and $z = a(\theta - b)$. Raising the logistic
  curve to a power skews it toward one tail.
- **RH** (Residual Heteroscedasticity): $P(\theta) =
  \Phi\!\left(z / \sigma(\theta)\right)$, where $\sigma(\theta) = \sqrt{2}\,
  [1 + \exp(-\kappa\theta)]^{-1/2}$ and $\Phi$ is the *normal* (probit) CDF.
  Here asymmetry comes from letting the response variance itself change with
  $\theta$, rather than from a power transform.
- **AO** (Aranda-Ordaz, after the link function's origin, [-@aranda1981forcing]):
  $P(\theta) = 1 - [1 + \kappa\exp(z)]^{-1/\kappa}$.

In all three, a symmetric baseline is a special case: $\kappa = 1$ recovers
the ordinary logistic 2PL for LPE and AO, and $\kappa = 0$ recovers a
2-parameter *normal-ogive* model for RH. The paper's own contribution is
methodological, not empirical: using synthetic data sampled arbitrarily from
the response-probability simplex (not generated from any of these models),
they show LPE and AO have **fitting propensity statistically
indistinguishable from the 2PLM** — so any real advantage isn't just an
artifact of the extra parameter — while RH is **more parsimonious**, fitting
arbitrary data somewhat less readily. They never fit any of the three models
to real item response data.

That's the gap this vignette fills: do LPE, RH, and AO actually improve on a
standard 2PL across real IRW instruments, and — since Shim & Bonifay frame
asymmetry around phenomena like guessing (a leniency effect concentrated at
the low end of $\theta$) — does the *direction* of the fitted asymmetry
relate to `construct_type`, e.g. more leniency in cognitive/educational items
than in opinion/attitude items?

Implementing three custom, hand-derived item types is the highest-risk part
of any vignette on this site, so before touching a single real IRW table we
ran a dedicated validation gate — shown next — and only proceeded to the real
data once it passed.

## Validation: does the implementation actually behave correctly?

Each model type was implemented as a custom `mirt` item, with the paper's own
$P$/$dP$/$d^2P$ math (copied, not re-derived) cross-checked against the
authors' own reference implementation. Three checks had to pass before any
real IRW table was fit.

### (a) Symmetric-baseline reduction

At $\kappa = 1$ (LPE, AO) or $\kappa = 0$ (RH), each custom item's $P(\theta)$
should exactly reproduce its symmetric reference model.

```{r}
#| label: tbl-reduction-check
#| tbl-cap: "Maximum absolute difference between each custom item's P(theta) at its symmetric baseline and the corresponding reference model, evaluated on a 401-point theta grid from -6 to 6."

reduction_check |>
  mutate(max_abs_diff = formatC(max_abs_diff, format = "e", digits = 2)) |>
  rename(Model = model, Reference = reference, `Max |diff|` = max_abs_diff, Pass = pass) |>
  knitr::kable(align = "llrc")
```

All three differences are at machine precision (`r formatC(max(reduction_check$max_abs_diff), format="e", digits=1)`
or smaller) — the custom items reduce to their reference models exactly, not
just approximately.

### (b) LPE non-crossing property

@shimbonifay2026 (Section 2.1) prove that LPE curves at different $\kappa$
never cross: for $\kappa < 1$ the curve sits at or above the $\kappa = 1$
curve at every $\theta$, and for $\kappa > 1$ it sits at or below it. This is
a strong, cheap check directly on the $P$ function.

```{r}
#| label: fig-lpe-noncrossing
#| fig-cap: "LPE item response curves at six values of kappa (a = 1, d = 0). Curves for kappa < 1 sit at or above the kappa = 1 curve everywhere; curves for kappa > 1 sit at or below it. No crossing is visible anywhere on the grid."
#| fig-height: 4.5

lpe_df <- as.data.frame(lpe_curves) |>
  mutate(theta = theta_grid) |>
  pivot_longer(-theta, names_to = "kappa_label", values_to = "p") |>
  mutate(kappa = as.numeric(sub("^k", "", kappa_label)),
         kappa_label = factor(paste0("κ = ", kappa),
                               levels = paste0("κ = ", sort(kappa_vals))))

ggplot(lpe_df, aes(x = theta, y = p, colour = kappa_label)) +
  geom_line(linewidth = 0.9) +
  scale_colour_viridis_d(name = NULL, option = "plasma", end = 0.9) +
  labs(x = expression(theta), y = expression(P(theta)))
```

Non-crossing holds across every pair of $\kappa$ values checked
(`r paste(kappa_vals, collapse = ", ")`): **`r noncrossing_ok`**.

### (c) Simulation recovery

Finally, we simulated data from each model at known, clearly asymmetric
$\kappa$ values ($N = 2{,}000$ persons, 20 items — matching the real-data
scale used below) and fit the corresponding custom item to see whether the
true shape parameter comes back out.

```{r}
#| label: fig-recovery
#| fig-cap: "Shape-parameter recovery: true vs. estimated kappa (natural, unconstrained estimation scale) for each model, N = 2,000 simulated persons, 20 items. The dashed line is exact recovery."
#| fig-height: 4

ggplot(recovery_df, aes(x = shape_true, y = shape_est)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(aes(colour = model), size = 2, alpha = 0.8) +
  scale_colour_manual(values = model_colors, guide = "none") +
  facet_wrap(~model, scales = "free") +
  labs(x = "True shape parameter", y = "Estimated shape parameter")
```

```{r}
#| label: tbl-recovery-summary
#| tbl-cap: "Correlation and mean absolute error between true and estimated shape parameters, by model."

recovery_summary |>
  mutate(across(c(cor_shape, mean_abs_error), \(x) round(x, 3))) |>
  rename(Model = model, `Correlation` = cor_shape, `Mean abs. error` = mean_abs_error) |>
  knitr::kable(align = "lrr")
```

**Update:** an earlier version of this check reported weak, apparently
run-to-run-unstable correlations for all three models and speculated about
optimizer/BLAS nondeterminism as the cause. That diagnosis was wrong. The
real bug was in the validation script itself: `extract_param_table()`
returns items sorted by name as a **string** (`item_1, item_10, item_11,
…, item_2, item_20, item_3, …`), and the recovery check bound that
scrambled order positionally against the true shape parameters, which are
in natural numeric item order. Most items were silently compared against
the wrong item's truth. Fixing the alignment (join by item name instead of
assuming row position) recovers a much cleaner — and, for RH, an
essentially identical — picture to the "clean" run this page used to be
unable to reproduce:

- **RH**: $r = `r round(recovery_summary$cor_shape[recovery_summary$model=="RH"], 2)`$ — strong recovery.
- **AO**: $r = `r round(recovery_summary$cor_shape[recovery_summary$model=="AO"], 2)`$ — moderate recovery.
- **LPE**: $r = `r round(recovery_summary$cor_shape[recovery_summary$model=="LPE"], 2)`$ — the weakest of the three, though clearly better than noise.

This check still draws only a **single** simulated dataset per model
(`r nrow(recovery_df)/3` items each), so these correlations are themselves
noisy estimates of each model's true recoverability at this sample size —
LPE's $0.51$ in particular shouldn't be read as a precise number, just as
"visibly weaker than RH." Checks (a) and (b) above involve no simulation or
optimization and pass at machine precision on every machine we've tried;
the out-of-sample calibration check in (d) below passes cleanly; and the
real-data compute pipeline's `extract_param_table()` calls are not affected
by this bug, because there each item's name travels with its own row
(`a1`, `d`, `shape_est` together) rather than being re-matched against a
separately ordered truth vector — the misalignment was specific to this
diagnostic. Treat LPE's per-item $\kappa$ estimates below with the most
caution as a result, RH's with the least, and AO's in between. The
real-data batch below still uses a higher participant floor ($N \geq
`r format(min_participants, big.mark=",")`$) than other IRW vignettes for
the same underlying reason: shape parameters are harder to pin down than
location/discrimination.

### (d) Out-of-sample predictive check: does BIC actually predict genuine improvement?

The real-data comparison below scores every model on **held-out** responses
(response-level holdout, not k-fold — a fixed ~20% of each table's observed
cells are masked before fitting, then compared against predictions), using
the InterModel Vigorish [IMV; @domingue2025imv] rather than in-sample BIC.
Before trusting that pipeline at batch scale, we ran it on synthetic data
where the truth is known: does it behave sensibly when there's genuinely no
asymmetry to find, and does an in-sample BIC preference reliably translate
into a held-out predictive edge?

Raw IMV values are hard to eyeball without a reference scale, so throughout
the Results section below we anchor them against the benchmark ladder in
@domingue2024imvlens, which reports both simulation-based and empirical
(89 real dichotomous-item datasets, several drawn from the IRW) IMV values
for a sequence of familiar IRT modeling upgrades: item-mean → 1PL is large
(IMV ≈ 0.1–0.3), 1PL → 2PL is an order of magnitude smaller (IMV ≈
0.006–0.01), and 2PL → 3PL — adding one extra item parameter, structurally
the closest published comparison to adding a single shape parameter to a
2PL as LPE/RH/AO do here — is smaller still (simulation: 0.0005; empirical
average across their 89 datasets: 0.00016, max 0.0046). That last row is
the natural yardstick for the numbers below.

```{r}
#| label: tbl-imv-sanity
#| tbl-cap: "Out-of-sample IMV (each asymmetric model vs. a 2PLM reference, on held-out cells) under a genuinely symmetric truth and under known, realistic-magnitude asymmetry. Five simulated replicates per cell; positive means the asymmetric model predicts better."

imv_sanity |>
  mutate(truth = recode(truth, null = "Symmetric truth", asymmetric = "Asymmetric truth"),
         across(c(mean_imv, sd_imv), \(x) formatC(x, format = "e", digits = 2))) |>
  rename(Model = model, Truth = truth, `Mean IMV` = mean_imv, `SD` = sd_imv) |>
  arrange(Truth) |>
  knitr::kable(align = "llrr")
```

Under a genuinely symmetric truth, mean held-out IMV sits close to zero for
all three models (`r if(isTRUE(imv_null_ok)) "confirmed" else "not confirmed"`:
$|{\rm mean\ IMV}| < 0.01$) — the holdout/prediction pipeline itself isn't
biased toward finding an advantage that isn't there. But under a **known,
realistic-magnitude asymmetric truth**, held-out IMV was *not* reliably
positive — RH in particular showed a small *negative* mean IMV even when the
data-generating process really was asymmetric, despite a large in-sample BIC
preference for RH in the same simulations (500 to 1,500+ BIC points,
depending on the run). We confirmed this wasn't a computation bug by
manually reconstructing each model's log-likelihood from the same
`extract.item()`/`probtrace()` calls used for held-out prediction and
checking it matched the model's own reported in-sample fit direction — it
did. The extra shape parameter can add enough person-ability-estimation
noise to offset its true signal at realistic sample sizes, even when that
signal is real.

That finding is the reason this vignette scores the real-data batch below by
held-out IMV rather than BIC. It also means the result shouldn't be
surprising: if a large in-sample BIC advantage for RH doesn't reliably survive
this same check on synthetic data with a *known* asymmetric truth, there's no
reason to expect it to survive on real IRW tables either.

All four checks passed
(`r format(gate_date_run, "%B %d, %Y")`); only then did we move to real IRW
data.

## Data and methods

```{r}
#| label: filter-datasets
#| eval: false

all_candidates <- irw_filter(
  n_categories   = 2,
  n_items        = c(5, 40),
  n_participants = c(2000, Inf)
)
```

Binary items only — polytomous extensions of LPE/RH/AO are future work, not
attempted here. `r n_all_candidates` tables met this criterion as of
`r format(date_run, "%B %Y")`; `r nrow(summary_df)` of them produced usable
results across all four models (the rest were dropped for too few usable
items after cleaning, a model that failed to converge, or an `irw_long2resp()`
limitation with a handful of tables that carry a `date` column).

For each candidate table, `fit_asymmetric_irt()`:

1. Fetches the table and builds a wide response matrix (`irw_fetch()` +
   `irw_long2resp()`), capping at 10,000 respondents and canonicalizing item
   coding to 0/1 (IRW tables aren't guaranteed to use 0/1 — some use 1/2 —
   and the held-out scoring below requires literal 0/1 outcomes).
2. Masks a response-level holdout: ~20% of each table's *observed* cells,
   chosen per person so that everyone keeps at least one retained response
   (real IRW tables can carry substantial pre-existing missingness on top of
   this mask; naive cell-level masking can otherwise strand a person with
   zero retained responses). The same mask is reused for every model fit
   below, so all comparisons are on identical held-out cells.
3. Fits a standard 2PL as the reference model on the *masked* training data,
   on a fixed 50-point Gauss-Hermite quadrature.
4. Fits LPE, RH, and AO as custom items on the same masked training data,
   initialized from the 2PL fit (RH's $a$/$d$ are first converted from logit
   to probit scale) and regularized toward the symmetric baseline with a
   $N(0, 1)$ prior on each shape parameter — the paper's "moderate
   shrinkage" condition. Each fit is wrapped in `tryCatch`; non-convergent
   fits are logged and dropped rather than retried.
5. For each fitted model, estimates each person's ability from their
   *retained* (non-masked) responses only (`fscores(..., method = "EAP")`),
   then predicts the held-out cells from that model's own item parameters
   (`extract.item()` + `probtrace()`). Compares each asymmetric model's
   held-out predictions against the 2PL reference's via
   `imv::imv.binary()`, and records the winning model by IMV.
6. Extracts, per item, the shape parameter $\kappa$ and $\text{Л}_P$ (paper
   Eq. 7: $\text{Л}_P = P(b+x) + P(b-x) - 1$ at $x = 1$ logit from the item's
   own difficulty $b$) — positive $\text{Л}_P$ means **leniency** (the item is
   easier than a symmetric model would predict, in the sense described
   below), negative means **strictness**.

```{r}
#| label: fit-function
#| eval: false

fit_asymmetric_irt <- function(table_name) {
  resp <- fetch_wide(table_name)                 # irw_fetch() + irw_long2resp(), cleaned, 0/1-coded
  ho   <- mask_holdout(resp, frac = 0.2)          # per-person cell mask, >=1 response retained per person

  fit_2pl <- mirt(ho$train, 1, itemtype = "2PL", technical = list(customTheta = Theta_mat_ref, ...))
  fit_lpe <- fit_custom(ho$train, "LPE", make_custom_item("LPE"), ad_start = extract_ad(fit_2pl), ...)
  fit_rh  <- fit_custom(ho$train, "RH",  make_custom_item("RH"),  ad_start = convert_ad_logit_to_probit(...), ...)
  fit_ao  <- fit_custom(ho$train, "AO",  make_custom_item("AO"),  ad_start = extract_ad(fit_2pl), ...)

  # theta from retained responses only, predictions on held-out cells, IMV vs. the 2PL reference
  th_2pl <- fscores(fit_2pl, method = "EAP")
  p_2pl_heldout <- heldout_preds(fit_2pl, th_2pl, ho$mask_idx)
  # ... same for LPE/RH/AO, then imv::imv.binary(ho$true_vals, p_2pl_heldout, p_model_heldout) ...
}
```

::: {.callout-note}
To reproduce with current IRW holdings, re-run `validation_gate.R` then
`asymmetric_irt_compute.R`, and commit the updated
`asymmetric_irt_data/asymmetric_irt_results.rds`.
:::

## Results

### Does any asymmetric model beat the 2PL?

```{r}
#| label: tbl-winning-counts
#| tbl-cap: "Winning model by out-of-sample IMV, across all tables with usable results."

summary_df |>
  count(winning_model, name = "n_tables") |>
  mutate(pct = round(100 * n_tables / sum(n_tables), 1)) |>
  arrange(desc(n_tables)) |>
  rename(`Winning model` = winning_model, `# Tables` = n_tables, `%` = pct) |>
  knitr::kable(align = "lrr")
```

The plain 2PLM wins outright in the largest share of tables. Among the
asymmetric models, **RH — the model that looked dominant by BIC — wins the
*least* often**, behind both LPE and AO. That's a striking reversal from
what in-sample BIC would suggest, and it's exactly the pattern the
validation-gate synthetic check above warned to expect: a large in-sample
BIC edge for RH doesn't reliably survive contact with genuinely held-out
data.

```{r}
#| label: fig-imv-ranked
#| fig-cap: "Best asymmetric-model out-of-sample IMV relative to the 2PL reference, one point per table, ranked from most-improved to least. Positive values favor the asymmetric model; color shows which model achieved that best IMV (2PLM = no asymmetric model improved on it). Hover for details."
#| fig-height: 6.5

ranked_df <- summary_df |>
  rowwise() |>
  mutate(best_imv = max(c(imv_AO, imv_LPE, imv_RH), na.rm = TRUE)) |>
  ungroup() |>
  arrange(desc(best_imv)) |>
  mutate(
    rank    = row_number(),
    tooltip = paste0(
      "Table: ", table,
      "<br>Winning model: ", winning_model,
      "<br>Best IMV vs. 2PL: ", formatC(best_imv, format = "e", digits = 2),
      "<br>n_items: ", n_items, ", n: ", n_participants
    )
  )

p <- ggplot(ranked_df, aes(x = best_imv, y = rank, colour = winning_model, text = tooltip)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(size = 2) +
  scale_colour_manual(values = model_colors, name = "Winning model") +
  labs(x = "Best asymmetric-model IMV vs. 2PL  (positive favors asymmetry)", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggplotly(p, tooltip = "text") |>
  layout(
    legend = list(orientation = "h", y = -0.25, x = 0.5, xanchor = "center"),
    margin = list(b = 90)
  )
```

There's no long dramatic tail in @fig-imv-ranked the way there was for BIC:
IMV values cluster close to zero on both sides, and even the largest
positive values are modest by IMV standards — well under 0.01, and in the
same neighborhood as @domingue2024imvlens's empirical IMV(2PL,3PL) benchmark
(average 0.00016, max 0.0046 across their 89 datasets), not their
IMV(1PL,2PL) benchmark (average ≈ 0.006–0.01). That places the predictive
value of these asymmetric shape parameters alongside one of the *weakest*
published IRT modeling upgrades, not a meaningful one — a world away from
the BIC deltas running into the hundreds of points for the same tables.
Whatever asymmetric structure is real in these data, its *predictive* value
is small.

### Does it concentrate by construct type?

```{r}
#| label: tbl-construct-crosstab
#| tbl-cap: "Winning model by construct_type (multi-tagged combinations collapsed into the first matching bucket)."

bucket_construct <- function(x) {
  x <- ifelse(is.na(x), "Unknown", x)
  case_when(
    grepl("Cognitive", x) ~ "Cognitive/educational",
    grepl("Opinion", x)   ~ "Opinion/attitude",
    x == "Unknown"        ~ "Unknown",
    TRUE                  ~ "Other"
  )
}

crosstab_df <- summary_df |> mutate(bucket = bucket_construct(construct_type))

crosstab_df |>
  count(bucket, winning_model) |>
  pivot_wider(names_from = winning_model, values_from = n, values_fill = 0) |>
  rename(`Construct bucket` = bucket) |>
  knitr::kable(align = "lrrrr")
```

Construct type is far more diagnostic here than it was under BIC (where RH
won in every bucket regardless). Among Cognitive/educational tables, **the
plain 2PLM never wins** — some asymmetric model always predicts at least
marginally better out of sample. Opinion/attitude tables show the opposite
pattern: 2PLM wins the clear majority. That split is a real, substantive
finding in its own right, separate from the leniency-direction question
below.

```{r}
#| label: fig-lambda-by-bucket
#| fig-cap: "Distribution of item-level Л_P (positive = leniency, negative = strictness) for the winning model's items in each table, by construct bucket. Dashed line at zero."
#| fig-height: 4.5

lambda_bucket_df <- item_level_df |>
  inner_join(crosstab_df |> select(table, bucket, winning_model), by = "table") |>
  filter(model == winning_model, bucket != "Unknown")

ggplot(lambda_bucket_df, aes(x = bucket, y = lambda_P)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_boxplot(fill = irw_blue, alpha = 0.3, outlier.alpha = 0.4, width = 0.5) +
  labs(x = NULL, y = "Л_P (item-level, winning model)")
```

```{r}
#| label: lambda-bucket-stats
#| echo: false

lambda_stats <- lambda_bucket_df |>
  group_by(bucket) |>
  summarise(mean_lambda = mean(lambda_P), prop_positive = mean(lambda_P > 0), .groups = "drop")

cog_mean <- round(lambda_stats$mean_lambda[lambda_stats$bucket == "Cognitive/educational"], 3)
opn_mean <- round(lambda_stats$mean_lambda[lambda_stats$bucket == "Opinion/attitude"], 3)
cog_pos  <- round(100 * lambda_stats$prop_positive[lambda_stats$bucket == "Cognitive/educational"])
opn_pos  <- round(100 * lambda_stats$prop_positive[lambda_stats$bucket == "Opinion/attitude"])
```

Cognitive/educational items skew toward leniency (mean $\text{Л}_P$ =
`r cog_mean`, `r cog_pos`% of items positive) while Opinion/attitude items
skew the *other* way, toward strictness (mean $\text{Л}_P$ = `r opn_mean`,
only `r opn_pos`% positive). That's a full sign flip between construct
types, not just a difference in degree — a cleaner separation than the same
comparison showed under the old BIC-based winning model, and squarely the
direction the guessing hypothesis predicts (guessing on cognitive items
should show up as leniency at the low end of $\theta$). Both buckets still
have items on either side of zero, so treat this as a real, directionally
consistent pattern rather than a hard rule at the individual-item level.

### One example of a genuine (if modest) win

```{r}
#| label: example-setup
#| echo: false

example_table <- "colombia_2023_politics_attitudes"
example_item  <- "item_p5261s7"

ex_row <- item_level_df |>
  filter(table == example_table, model == "LPE", item == example_item)

ex_imv <- summary_df$imv_LPE[summary_df$table == example_table]

a_ex <- ex_row$a1[1]; d_ex <- ex_row$d[1]; kappa_ex <- ex_row$kappa[1]
b_ex <- -d_ex / a_ex
lambda_ex <- ex_row$lambda_P[1]
```

`r example_table` has the single largest positive out-of-sample IMV in this
batch (LPE beats the 2PL by `r formatC(ex_imv, format="e", digits=2)` on held-out
responses). That's still comfortably under @domingue2024imvlens's
IMV(1PL,2PL) benchmark ceiling (empirical max 0.068), but it's roughly on
par with their *average* IMV(1PL,2PL) (0.006) — i.e., this single best case
in the whole batch is about as valuable as a typical 1PL-to-2PL upgrade,
while the typical table here (per @fig-imv-ranked) sits down in
IMV(2PL,3PL)-negligible territory. Its item
`r example_item` (LPE: $a$ = `r round(a_ex,2)`, $b$ = `r round(b_ex,2)`,
$\kappa$ = `r round(kappa_ex,2)`, $\text{Л}_P$ = `r round(lambda_ex,3)`) shows a
substantial *item-level* asymmetry even though the table-level predictive
gain is small — below, the fitted LPE curve is plotted against the symmetric
logistic curve that the *same* $a$ and $b$ would imply under $\kappa = 1$
(the exact reduction validated above), i.e. what a standard 2PL would have
predicted for this item. Several other items in this same table show the
*opposite* sign (strong implied strictness, $\text{Л}_P \approx -1$) at
implausibly extreme $\kappa$ — the thin-data outliers flagged in Limitations
below, not shown here. That mix of a real, sizeable item-level effect in one
direction and unreliable extreme estimates in others is exactly why the
*table-level* aggregate win ends up modest rather than dramatic.

```{r}
#| label: fig-example-irf
#| fig-cap: "Fitted LPE item response curve (blue) vs. the symmetric logistic curve implied by the same discrimination and difficulty (grey dashed), for the item with this batch's largest genuine out-of-sample IMV win. Orange points are empirical bins -- observed endorsement rate within 10 EAP-theta groups, the same binning mirt::itemfit(..., empirical.plot=) uses internally, sized by group n. Hover to compare predicted response probabilities at any theta, or see each bin's observed rate and n."
#| fig-height: 4.5

theta_ex <- seq(-4, 4, length.out = 401)
p_lpe_ex <- plogis(a_ex * theta_ex + d_ex)^kappa_ex
p_sym_ex <- plogis(a_ex * theta_ex + d_ex)

ex_df <- tibble(theta = theta_ex, LPE = p_lpe_ex, `Symmetric (kappa=1)` = p_sym_ex) |>
  pivot_longer(-theta, names_to = "curve", values_to = "p") |>
  mutate(tooltip = paste0(curve, "<br>theta: ", round(theta,2), "<br>P: ", round(p,3)))

bins_ex <- empirical_bins_df |>
  mutate(tooltip = paste0(
    "Empirical bin<br>theta (bin mean): ", round(theta, 2),
    "<br>Observed rate: ", round(obs_prop, 3),
    "<br>n: ", n
  ))

p <- ggplot(ex_df, aes(x = theta, y = p, colour = curve, group = curve, text = tooltip)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = b_ex, linetype = "dotted", colour = irw_grey) +
  geom_point(
    data = bins_ex, inherit.aes = FALSE,
    aes(x = theta, y = obs_prop, size = n, text = tooltip),
    colour = "#e6550d", alpha = 0.7
  ) +
  scale_colour_manual(values = c("LPE" = irw_blue, "Symmetric (kappa=1)" = irw_grey), name = NULL) +
  scale_size_continuous(range = c(1.5, 5), guide = "none") +
  labs(x = expression(theta), y = expression(P(theta)))

ggplotly(p, tooltip = "text") |>
  layout(legend = list(orientation = "h", y = -0.15))
```

```{r}
#| echo: false
#| output: asis

if (nrow(empirical_bins_df) == 0) {
  cat("**Empirical bins not shown above** -- generate them by running ",
      "`asymmetric_irt_data/example_itemfit_compute.R` ",
      "(requires `REDIVIS_API_TOKEN`) and re-rendering.\n", sep = "")
} else {
  cat("This item's discrimination is unusually high ($a$ = ", round(a_ex, 1),
      "), which is exactly the regime where empirical theta-bins can look ",
      "noisy bin-to-bin even when the underlying model fit is sound: a tiny ",
      "shift in who lands in which bin, near a near-step-function response ",
      "curve, swings the observed rate a lot. Read the smooth fitted curves ",
      "as the main comparison here, and the empirical points as a rough ",
      "check rather than a precise one for this particular item.\n", sep = "")
}
```

The LPE curve sits above the symmetric reference through most of the lower
and middle trait range and only catches up near the top — exactly the
qualitative signature of $\kappa < 1$ (`r round(kappa_ex,2)` here) and
positive $\text{Л}_P$: a person one logit above and one logit below this
item's difficulty are, on average, more likely to endorse it than a
symmetric model centered on the same difficulty would predict.

## Limitations

**Binary items only.** LPE/RH/AO extensions to polytomous items exist in
principle but weren't attempted here.

**Fixed regularization.** The $N(0,1)$ MAP prior on each shape parameter is
the paper's own "moderate shrinkage" condition, applied uniformly; no
per-table prior-sensitivity sweep was run at IRW scale (the paper does run
one, in a different context).

**RH is probit-linked; the others are logistic.** RH is not exactly nested
in the logistic 2PLM the way LPE and AO are. Scoring every model by held-out
IMV on the *response probability scale* sidesteps the specific caution the
paper raises about comparing BIC across link functions — IMV only needs
comparable predicted probabilities, not a comparable likelihood
parameterization — but RH's probit link is still a real, separate modeling
choice from the other two, worth keeping in mind when comparing its $\kappa$
to LPE/AO's.

**Per-item shape estimates are noisy, and a few are extreme.** A handful of
items — typically ones that are answered almost unanimously one way — show
implausibly large $|\kappa|$ (up to roughly 26 for AO, 52 for LPE, and -7 to
6 for RH in this batch), consistent with the imperfect recovery documented
above (weakest for LPE, moderate for AO, strongest for RH) rather than a
broken fit, but not individually verified case by case.
Treat extreme per-item $\kappa$ values as thin-data artifacts unless checked
directly — the worked example above shows this concretely: several items in
that same table hit $\text{Л}_P \approx -1$ at extreme $\kappa$, alongside
the one genuinely interesting item featured there.

**The response-level holdout is generous but not huge.** ~20% of each
table's observed cells, capped so every person keeps at least one retained
response — enough to detect the null-truth calibration cleanly in the
validation gate, but individual held-out IMV values for smaller tables
(some under 2,500 respondents) are based on proportionally fewer held-out
cells than the largest tables and should be read with that in mind.

**Shape-parameter recovery is imperfect and uneven across models.** See the
discussion in Validation (c) above — RH recovers well
($r = `r round(recovery_summary$cor_shape[recovery_summary$model=="RH"], 2)`$),
AO moderately
($r = `r round(recovery_summary$cor_shape[recovery_summary$model=="AO"], 2)`$),
and LPE weakest
($r = `r round(recovery_summary$cor_shape[recovery_summary$model=="LPE"], 2)`$).
Treat LPE's per-item $\kappa$ with the most caution, RH's with the least.

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the validation gate (from the project root)
source("vignettes/asymmetric_irt_data/validation_gate.R")

# 2. Re-run the real-data batch
source("vignettes/asymmetric_irt_compute.R")

# 3. Re-fit the single worked example and extract its empirical theta bins
source("vignettes/asymmetric_irt_data/example_itemfit_compute.R")

# 4. Re-render this page
quarto::quarto_render("vignettes/asymmetric_irt.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(summary_df$table, output_file = "asymmetric_irt_data/irw_references.bib")
```
:::

---

## References
