---
title: "Asymmetric IRT Models Across IRW: Do LPE, RH, and AO Beat the 2PL?"
description: |
  Shim & Bonifay (2026) introduce three IRT models with asymmetric item
  response functions -- LPE, RH, and AO -- and show, via simplex-sampled
  synthetic data, that two of them fit about as well as a standard 2PL even
  when fitting propensity alone is compared, while the third (RH) is more
  parsimonious. They never apply the models to real data. We do: after a
  dedicated correctness/recovery validation gate, we fit a 2PL reference plus
  all three asymmetric models across dozens of binary IRW tables, to see
  whether the extra asymmetry parameter actually earns its keep, and whether
  its direction tracks known response phenomena like guessing.
date: today
bibliography: asymmetric_irt_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"
irw_green <- "#4daf4a"

model_colors <- c("2PLM" = irw_grey, "RH" = irw_blue, "LPE" = irw_red, "AO" = irw_green)

results_path <- "asymmetric_irt_data/asymmetric_irt_results.rds"
gate_path    <- "asymmetric_irt_data/validation_gate_log.rds"
cache_ok     <- file.exists(results_path) && file.exists(gate_path)

if (cache_ok) {
  .res             <- readRDS(results_path)
  summary_df       <- .res$summary
  item_level_df    <- .res$item_level
  candidate_tables <- .res$candidate_tables
  n_all_candidates <- .res$n_all_candidates
  min_participants <- .res$min_participants
  is_pilot         <- isTRUE(.res$pilot)
  date_run         <- .res$date_run

  .gate              <- readRDS(gate_path)
  reduction_check    <- .gate$reduction_check
  noncrossing_ok     <- .gate$noncrossing_ok
  lpe_curves         <- .gate$lpe_curves
  theta_grid         <- .gate$theta_grid
  kappa_vals         <- .gate$kappa_vals
  recovery_df        <- .gate$recovery_df
  recovery_summary   <- .gate$recovery_summary
  gate_date_run      <- .gate$date_run
} else {
  summary_df <- tibble(
    table = character(), n_items = integer(), n_participants = integer(),
    construct_type = character(), bic_2pl = numeric(), bic_delta_AO = numeric(),
    bic_delta_LPE = numeric(), bic_delta_RH = numeric(), winning_model = character()
  )
  item_level_df <- tibble(
    table = character(), model = character(), item = character(), a1 = numeric(),
    d = numeric(), shape_est = numeric(), kappa = numeric(), lambda_P = numeric()
  )
  candidate_tables <- character(); n_all_candidates <- 0L; min_participants <- 2000
  is_pilot <- FALSE; date_run <- Sys.Date()
  reduction_check <- tibble(model = character(), reference = character(),
                             max_abs_diff = numeric(), pass = logical())
  noncrossing_ok <- NA; lpe_curves <- matrix(); theta_grid <- numeric()
  kappa_vals <- numeric()
  recovery_df <- tibble(model = character(), item = integer(), a_true = numeric(),
                         d_true = numeric(), shape_true = numeric(), shape_est = numeric())
  recovery_summary <- tibble(model = character(), cor_shape = numeric(), mean_abs_error = numeric())
  gate_date_run <- Sys.Date()
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running, in order, from the project root:\n\n",
  "```r\nRscript vignettes/asymmetric_irt_data/validation_gate.R\n",
  "Rscript vignettes/asymmetric_irt_compute.R\n```\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the model implementation, the validation gate, the
dataset search, the compute script, and the writing below — was produced
largely by Claude (Anthropic), working from a task specification and with
human review. The custom item types used here are a non-trivial extension of
`mirt`; treat the methodological choices and results accordingly, and check
`asymmetric_irt_helpers.R` and the validation gate directly if you're relying
on the numbers shown.
:::

## Background

Every IRT model used elsewhere on this site — the 2PL, the graded response
model, the [Dutch Identity](dutch_identity.qmd)'s loglinear framework — has a
**symmetric** item response function: the probability of a correct or
endorsing response rises smoothly and symmetrically around the item's
difficulty. But nothing requires that. A guessing-prone multiple-choice item,
for instance, has a response curve that rises quickly from a nonzero floor
and then flattens gradually near the top — asymmetric around its inflection
point.

@shimbonifay2026 study three such asymmetric models, each adding a single
shape parameter $\kappa$ to a standard two-parameter item:

- **LPE** (logistic power expit): $P(\theta) = [\Psi(z)]^{\kappa}$, where
  $\Psi$ is the logistic CDF and $z = a(\theta - b)$. Raising the logistic
  curve to a power skews it toward one tail.
- **RH** (Ramsay/heteroscedastic, per the paper's naming): $P(\theta) =
  \Phi\!\left(z / \sigma(\theta)\right)$, where $\sigma(\theta) = \sqrt{2}\,
  [1 + \exp(-\kappa\theta)]^{-1/2}$ and $\Phi$ is the *normal* (probit) CDF.
  Here asymmetry comes from letting the response variance itself change with
  $\theta$, rather than from a power transform.
- **AO** (Aranda-Ordaz, after the link function's origin, [-@aranda1981forcing]):
  $P(\theta) = 1 - [1 + \kappa\exp(z)]^{-1/\kappa}$.

In all three, a symmetric baseline is a special case: $\kappa = 1$ recovers
the ordinary logistic 2PL for LPE and AO, and $\kappa = 0$ recovers a
2-parameter *normal-ogive* model for RH. The paper's own contribution is
methodological, not empirical: using synthetic data sampled arbitrarily from
the response-probability simplex (not generated from any of these models),
they show LPE and AO have **fitting propensity statistically
indistinguishable from the 2PLM** — so any real advantage isn't just an
artifact of the extra parameter — while RH is **more parsimonious**, fitting
arbitrary data somewhat less readily. They never fit any of the three models
to real item response data.

That's the gap this vignette fills: do LPE, RH, and AO actually improve on a
standard 2PL across real IRW instruments, and — since Shim & Bonifay frame
asymmetry around phenomena like guessing (a leniency effect concentrated at
the low end of $\theta$) — does the *direction* of the fitted asymmetry
relate to `construct_type`, e.g. more leniency in cognitive/educational items
than in opinion/attitude items?

Implementing three custom, hand-derived item types is the highest-risk part
of any vignette on this site, so before touching a single real IRW table we
ran a dedicated validation gate — shown next — and only proceeded to the real
data once it passed.

## Validation: does the implementation actually behave correctly?

Each model type was implemented as a custom `mirt` item, with the paper's own
$P$/$dP$/$d^2P$ math (copied, not re-derived) cross-checked against the
authors' own reference implementation. Three checks had to pass before any
real IRW table was fit.

### (a) Symmetric-baseline reduction

At $\kappa = 1$ (LPE, AO) or $\kappa = 0$ (RH), each custom item's $P(\theta)$
should exactly reproduce its symmetric reference model.

```{r}
#| label: tbl-reduction-check
#| tbl-cap: "Maximum absolute difference between each custom item's P(theta) at its symmetric baseline and the corresponding reference model, evaluated on a 401-point theta grid from -6 to 6."

reduction_check |>
  mutate(max_abs_diff = formatC(max_abs_diff, format = "e", digits = 2)) |>
  rename(Model = model, Reference = reference, `Max |diff|` = max_abs_diff, Pass = pass) |>
  knitr::kable(align = "llrc")
```

All three differences are at machine precision (`r formatC(max(reduction_check$max_abs_diff), format="e", digits=1)`
or smaller) — the custom items reduce to their reference models exactly, not
just approximately.

### (b) LPE non-crossing property

@shimbonifay2026 (Section 2.1) prove that LPE curves at different $\kappa$
never cross: for $\kappa < 1$ the curve sits at or above the $\kappa = 1$
curve at every $\theta$, and for $\kappa > 1$ it sits at or below it. This is
a strong, cheap check directly on the $P$ function.

```{r}
#| label: fig-lpe-noncrossing
#| fig-cap: "LPE item response curves at six values of kappa (a = 1, d = 0). Curves for kappa < 1 sit at or above the kappa = 1 curve everywhere; curves for kappa > 1 sit at or below it. No crossing is visible anywhere on the grid."
#| fig-height: 4.5

lpe_df <- as.data.frame(lpe_curves) |>
  mutate(theta = theta_grid) |>
  pivot_longer(-theta, names_to = "kappa_label", values_to = "p") |>
  mutate(kappa = as.numeric(sub("^k", "", kappa_label)),
         kappa_label = factor(paste0("κ = ", kappa),
                               levels = paste0("κ = ", sort(kappa_vals))))

ggplot(lpe_df, aes(x = theta, y = p, colour = kappa_label)) +
  geom_line(linewidth = 0.9) +
  scale_colour_viridis_d(name = NULL, option = "plasma", end = 0.9) +
  labs(x = expression(theta), y = expression(P(theta)))
```

Non-crossing holds across every pair of $\kappa$ values checked
(`r paste(kappa_vals, collapse = ", ")`): **`r noncrossing_ok`**.

### (c) Simulation recovery

Finally, we simulated data from each model at known, clearly asymmetric
$\kappa$ values ($N = 2{,}000$ persons, 20 items — matching the real-data
scale used below) and fit the corresponding custom item to see whether the
true shape parameter comes back out.

```{r}
#| label: fig-recovery
#| fig-cap: "Shape-parameter recovery: true vs. estimated kappa (natural, unconstrained estimation scale) for each model, N = 2,000 simulated persons, 20 items. The dashed line is exact recovery."
#| fig-height: 4

ggplot(recovery_df, aes(x = shape_true, y = shape_est)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(aes(colour = model), size = 2, alpha = 0.8) +
  scale_colour_manual(values = model_colors, guide = "none") +
  facet_wrap(~model, scales = "free") +
  labs(x = "True shape parameter", y = "Estimated shape parameter")
```

```{r}
#| label: tbl-recovery-summary
#| tbl-cap: "Correlation and mean absolute error between true and estimated shape parameters, by model."

recovery_summary |>
  mutate(across(c(cor_shape, mean_abs_error), \(x) round(x, 3))) |>
  rename(Model = model, `Correlation` = cor_shape, `Mean abs. error` = mean_abs_error) |>
  knitr::kable(align = "lrr")
```

RH recovers well; AO recovers decently; **LPE recovery is noticeably weaker**
at this sample size. That gap held up under scrutiny rather than being a bug:
the analytic LPE/AO gradients matched numerical finite-difference gradients
to several decimal places everywhere tested, and separate runs at larger $N$
showed recovery improving substantially (roughly $r \approx 0.4$ at $N =
800$ rising to $r \approx 0.7$ at $N = 5{,}000$) — consistent with a genuine
small-sample identifiability limitation of the LPE power-transform
parameterization, not an implementation error. Interestingly, this is the
*opposite* of what the paper's own indirect diagnostic (fitting-propensity
stability under arbitrary simplex-sampled data) suggests, which reads RH as
having comparatively less information. The two diagnostics ask different
questions — arbitrary-data fitting stability vs. direct generative-truth
recovery — and needn't agree; the tension is worth flagging rather than
resolving here. Practically, it means **per-item LPE $\kappa$ estimates below
should be read with more caution than RH's**, and it's why the real-data
batch below uses a higher participant floor ($N \geq `r format(min_participants, big.mark=",")`$)
than other IRW vignettes.

All three checks passed
(`r format(gate_date_run, "%B %d, %Y")`); only then did we move to real IRW
data.

## Data and methods

```{r}
#| label: filter-datasets
#| eval: false

all_candidates <- irw_filter(
  n_categories   = 2,
  n_items        = c(5, 40),
  n_participants = c(2000, Inf)
)
```

Binary items only — polytomous extensions of LPE/RH/AO are future work, not
attempted here. `r n_all_candidates` tables met this criterion as of
`r format(date_run, "%B %Y")`; `r nrow(summary_df)` of them produced usable
results across all four models (the rest were dropped for too few usable
items after cleaning, or a model that failed to converge).

For each candidate table, `fit_asymmetric_irt()`:

1. Fetches the table and builds a wide response matrix (`irw_fetch()` +
   `irw_long2resp()`), capping at 10,000 respondents.
2. Fits a standard 2PL as the reference model, on a fixed 50-point
   Gauss-Hermite quadrature (so later BIC comparisons aren't confounded by
   different integration grids).
3. Fits LPE, RH, and AO as custom items, initialized from the 2PL fit (RH's
   $a$/$d$ are first converted from logit to probit scale) and regularized
   toward the symmetric baseline with a $N(0, 1)$ prior on each shape
   parameter — the paper's "moderate shrinkage" condition. Each fit is
   wrapped in `tryCatch`; non-convergent fits are logged and dropped rather
   than retried.
4. Compares AIC/BIC/log-likelihood of each asymmetric model against the 2PL
   reference, and records the winning model by BIC.
5. Extracts, per item, the shape parameter $\kappa$ and $\Lambda_P$ (paper
   Eq. 7: $\Lambda_P = P(b+x) + P(b-x) - 1$ at $x = 1$ logit from the item's
   own difficulty $b$) — positive $\Lambda_P$ means **leniency** (the item is
   easier than a symmetric model would predict, in the sense described
   below), negative means **strictness**.

```{r}
#| label: fit-function
#| eval: false

fit_asymmetric_irt <- function(table_name) {
  resp    <- fetch_wide(table_name)              # irw_fetch() + irw_long2resp(), cleaned
  fit_2pl <- mirt(resp, 1, itemtype = "2PL", technical = list(customTheta = Theta_mat_ref, ...))
  fit_lpe <- fit_custom(resp, "LPE", make_custom_item("LPE"), ad_start = extract_ad(fit_2pl), ...)
  fit_rh  <- fit_custom(resp, "RH",  make_custom_item("RH"),  ad_start = convert_ad_logit_to_probit(...), ...)
  fit_ao  <- fit_custom(resp, "AO",  make_custom_item("AO"),  ad_start = extract_ad(fit_2pl), ...)
  # ... compare BIC, extract per-item kappa / lambda_P ...
}
```

::: {.callout-note}
To reproduce with current IRW holdings, re-run `validation_gate.R` then
`asymmetric_irt_compute.R`, and commit the updated
`asymmetric_irt_data/asymmetric_irt_results.rds`.
:::

## Results

### Does any asymmetric model beat the 2PL?

```{r}
#| label: tbl-winning-counts
#| tbl-cap: "Winning model by BIC, across all tables with usable results."

summary_df |>
  count(winning_model, name = "n_tables") |>
  mutate(pct = round(100 * n_tables / sum(n_tables), 1)) |>
  arrange(desc(n_tables)) |>
  rename(`Winning model` = winning_model, `# Tables` = n_tables, `%` = pct) |>
  knitr::kable(align = "lrr")
```

RH is the clear standout: it wins by BIC in a large share of tables, often by
a wide margin. LPE and AO win far less often — consistent with the paper's
own finding that RH behaves differently from the other two, though not in
the direction Shim & Bonifay's arbitrary-data diagnostic alone would suggest
(recall Section "Validation" above).

```{r}
#| label: fig-bic-delta-ranked
#| fig-cap: "Best asymmetric-model BIC delta relative to the 2PL, one point per table, ranked from most-improved to least. Negative values favor the asymmetric model; color shows which model achieved that best delta (2PLM = no asymmetric model improved on it). Hover for details."
#| fig-height: 6.5

ranked_df <- summary_df |>
  rowwise() |>
  mutate(best_delta = min(c(bic_delta_AO, bic_delta_LPE, bic_delta_RH), na.rm = TRUE)) |>
  ungroup() |>
  arrange(best_delta) |>
  mutate(
    rank    = row_number(),
    tooltip = paste0(
      "Table: ", table,
      "<br>Winning model: ", winning_model,
      "<br>Best BIC delta vs. 2PL: ", round(best_delta, 1),
      "<br>n_items: ", n_items, ", n: ", n_participants
    )
  )

p <- ggplot(ranked_df, aes(x = best_delta, y = rank, colour = winning_model, text = tooltip)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(size = 2) +
  scale_colour_manual(values = model_colors, name = "Winning model") +
  labs(x = "Best asymmetric BIC delta vs. 2PL  (negative favors asymmetry)", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggplotly(p, tooltip = "text") |>
  layout(legend = list(orientation = "h", y = -0.1))
```

The long tail on the left of @fig-bic-delta-ranked is almost entirely RH: for
a handful of tables the BIC improvement over the plain 2PL runs into the
hundreds of points, which is not a marginal preference.

### Does it concentrate by construct type?

```{r}
#| label: tbl-construct-crosstab
#| tbl-cap: "Winning model by construct_type (multi-tagged combinations collapsed into the first matching bucket)."

bucket_construct <- function(x) {
  x <- ifelse(is.na(x), "Unknown", x)
  case_when(
    grepl("Cognitive", x) ~ "Cognitive/educational",
    grepl("Opinion", x)   ~ "Opinion/attitude",
    x == "Unknown"        ~ "Unknown",
    TRUE                  ~ "Other"
  )
}

crosstab_df <- summary_df |> mutate(bucket = bucket_construct(construct_type))

crosstab_df |>
  count(bucket, winning_model) |>
  pivot_wider(names_from = winning_model, values_from = n, values_fill = 0) |>
  rename(`Construct bucket` = bucket) |>
  knitr::kable(align = "lrrrr")
```

RH wins across every construct bucket, so "does RH win" isn't itself very
diagnostic of construct type. The more targeted question is whether the
*direction* of asymmetry — leniency vs. strictness — differs by bucket, which
is closer to what Shim & Bonifay's guessing framing actually predicts.

```{r}
#| label: fig-lambda-by-bucket
#| fig-cap: "Distribution of item-level Lambda_P (positive = leniency, negative = strictness) for the winning model's items in each table, by construct bucket. Dashed line at zero."
#| fig-height: 4.5

lambda_bucket_df <- item_level_df |>
  inner_join(crosstab_df |> select(table, bucket, winning_model), by = "table") |>
  filter(model == winning_model, bucket != "Unknown")

ggplot(lambda_bucket_df, aes(x = bucket, y = lambda_P)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_boxplot(fill = irw_blue, alpha = 0.3, outlier.alpha = 0.4, width = 0.5) +
  labs(x = NULL, y = expression(Lambda[P]~"(item-level, winning model)"))
```

```{r}
#| label: lambda-bucket-stats
#| echo: false

lambda_stats <- lambda_bucket_df |>
  group_by(bucket) |>
  summarise(mean_lambda = mean(lambda_P), prop_positive = mean(lambda_P > 0), .groups = "drop")

cog_mean <- round(lambda_stats$mean_lambda[lambda_stats$bucket == "Cognitive/educational"], 3)
opn_mean <- round(lambda_stats$mean_lambda[lambda_stats$bucket == "Opinion/attitude"], 3)
cog_pos  <- round(100 * lambda_stats$prop_positive[lambda_stats$bucket == "Cognitive/educational"])
opn_pos  <- round(100 * lambda_stats$prop_positive[lambda_stats$bucket == "Opinion/attitude"])
```

Cognitive/educational items skew mildly toward leniency (mean
$\Lambda_P$ = `r cog_mean`, `r cog_pos`% of items positive) compared to
Opinion/attitude items (mean $\Lambda_P$ = `r opn_mean`, `r opn_pos`% positive).
That's the direction the guessing hypothesis predicts, and it's a real,
non-zero gap — but it is a **modest** one, well short of a clean separation,
and both buckets have plenty of items on either side of zero. Treat this as
weak, directionally consistent evidence rather than a strong confirmation.

### One convincing example

```{r}
#| label: example-setup
#| echo: false

example_table <- "colombia_2023_politics_attitudes"
example_item  <- "item_p5261s7"

ex_row <- item_level_df |>
  filter(table == example_table, model == "RH", item == example_item)

ex_delta <- summary_df$bic_delta_RH[summary_df$table == example_table]

a_ex <- ex_row$a1[1]; d_ex <- ex_row$d[1]; kappa_ex <- ex_row$kappa[1]
b_ex <- -d_ex / a_ex
lambda_ex <- ex_row$lambda_P[1]
```

`r example_table` is the table with the single largest RH BIC improvement in
this sample (`r round(ex_delta, 1)` BIC points below the 2PL). Its item
`r example_item` (RH: $a$ = `r round(a_ex,2)`, $b$ = `r round(b_ex,2)`,
$\kappa$ = `r round(kappa_ex,2)`, $\Lambda_P$ = `r round(lambda_ex,3)`) makes
the asymmetry concrete: below, the fitted RH curve is plotted against the
symmetric normal-ogive curve that the *same* $a$ and $b$ would imply under
$\kappa = 0$ (the exact reduction validated above) — i.e., what a standard
model would have predicted for this item.

```{r}
#| label: fig-example-irf
#| fig-cap: "Fitted RH item response curve (blue) vs. the symmetric normal-ogive curve implied by the same discrimination and difficulty (grey dashed), for one item where RH wins convincingly. Hover to compare predicted response probabilities at any theta."
#| fig-height: 4.5

theta_ex <- seq(-4, 4, length.out = 401)
sigma_ex <- sqrt(2) * (1 + exp(-kappa_ex * theta_ex))^(-1/2)
p_rh_ex  <- pnorm((a_ex * theta_ex + d_ex) / sigma_ex)
p_sym_ex <- pnorm(a_ex * theta_ex + d_ex)

ex_df <- tibble(theta = theta_ex, RH = p_rh_ex, `Symmetric (kappa=0)` = p_sym_ex) |>
  pivot_longer(-theta, names_to = "curve", values_to = "p") |>
  mutate(tooltip = paste0(curve, "<br>theta: ", round(theta,2), "<br>P: ", round(p,3)))

p <- ggplot(ex_df, aes(x = theta, y = p, colour = curve, group = curve, text = tooltip)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = b_ex, linetype = "dotted", colour = irw_grey) +
  scale_colour_manual(values = c("RH" = irw_blue, "Symmetric (kappa=0)" = irw_grey), name = NULL) +
  labs(x = expression(theta), y = expression(P(theta)))

ggplotly(p, tooltip = "text") |>
  layout(legend = list(orientation = "h", y = -0.15))
```

The RH curve rises more steeply through the middle of the trait range and
flattens earlier at the top than the symmetric reference would — exactly the
qualitative signature $\Lambda_P > 0$ describes: a person one logit above and
one logit below this item's difficulty are, on average, more likely to
endorse it than a symmetric model centered on the same difficulty would
predict.

## Limitations

**Binary items only.** LPE/RH/AO extensions to polytomous items exist in
principle but weren't attempted here.

**Fixed regularization.** The $N(0,1)$ MAP prior on each shape parameter is
the paper's own "moderate shrinkage" condition, applied uniformly; no
per-table prior-sensitivity sweep was run at IRW scale (the paper does run
one, in a different context).

**RH is probit-linked; the others are logistic.** RH is not exactly nested
in the logistic 2PLM the way LPE and AO are — its BIC comparisons deserve the
same caution the paper itself applies when comparing across link functions.

**Per-item shape estimates are noisy, and a few are extreme.** A handful of
items — typically ones that are answered almost unanimously one way — show
implausibly large $|\kappa|$ (LPE up to roughly 76, AO up to roughly 27 in
this batch), consistent with the recovery weakness documented above rather
than a broken fit, but not individually verified case by case. Treat
extreme per-item $\kappa$ values as thin-data artifacts unless checked
directly.

**LPE recovery is the weakest of the three at this sample size.** Per-item
LPE $\kappa$ values should be read with more caution than RH's; see the
validation section above.

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the validation gate (from the project root)
source("vignettes/asymmetric_irt_data/validation_gate.R")

# 2. Re-run the real-data batch
source("vignettes/asymmetric_irt_compute.R")

# 3. Re-render this page
quarto::quarto_render("vignettes/asymmetric_irt.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(summary_df$table, output_file = "asymmetric_irt_data/irw_references.bib")
```
:::

---

## References
