---
title: "Bits per Item: How Much Does a Test Tell You About a Person?"
description: |
  A test is a sequence of questions asked about someone we cannot see. Each
  answer narrows down where that person sits on the trait being measured.
  Information theory lets us count that narrowing in bits, the same unit that
  tells you how much a Wordle guess cuts down the list of possible words. We
  play the game on three real tests, then measure the bits delivered by more
  than two hundred dichotomous tests in the Item Response Warehouse.
date: today
bibliography: bits_per_item_data/references.bib
nocite: |
  @*
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)
library(patchwork)

theme_set(theme_minimal(base_size = 14) +
            theme(plot.title = element_text(face = "bold", size = 14),
                  plot.title.position = "plot",
                  panel.grid.minor = element_blank()))
irw_blue   <- "#2166ac"   # the population of theta groups
irw_red    <- "#d73027"   # surprise / the aberrant respondent
irw_grey   <- "grey50"
irw_accent <- "#1b9e77"   # information gained, everywhere on this page

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "bits_per_item_data/bits_per_item_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  summary_df$construct_type <- sub(",.*$", "", summary_df$construct_type)
  items_df         <- .cache$items
  curves_df        <- .cache$curves
  showcases        <- .cache$showcases
  candidate_tables <- .cache$candidate_tables
  date_run         <- .cache$date_run
  is_pilot         <- isTRUE(.cache$pilot)
  failures         <- unique(sub("^\\S+ \\S+ ", "", .cache$failures))
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_tables_run   <- length(.cache$tables_attempted %||% candidate_tables)
} else {
  summary_df <- tibble(table = character(), model = character(), I_X = numeric(),
                       I_S = numeric(), rho_marginal = numeric(), n_items = integer())
  items_df <- tibble(); curves_df <- tibble(); showcases <- list(); failures <- character()
  candidate_tables <- character()
  date_run         <- Sys.Date()
  is_pilot         <- FALSE
  r_all_candidates <- 0L
  r_n_tables_run   <- 0L
}

suff_path <- "bits_per_item_data/sufficiency_results.rds"
suff <- if (file.exists(suff_path)) readRDS(suff_path)$showcases else NULL
sim_path <- "bits_per_item_data/reliability_sim_results.rds"
rel_sim <- if (file.exists(sim_path)) readRDS(sim_path)$sim else NULL
knitr::opts_chunk$set(eval = cache_ok)
# Every ggplot printed on this page goes through wl().
registerS3method("knit_print", "ggplot", function(x, ...) print(wl(x)), envir = asNamespace("knitr"))

h2 <- function(p) { p <- pmin(pmax(p, 1e-12), 1 - 1e-12); -(p * log2(p)) - ((1 - p) * log2(1 - p)) }
show_titles <- c(fitz_2024_numeracy       = "Numeracy (11 items)",
                 machivallianism_test_vcl = "Vocabulary checklist (16 words)",
                 frac20                   = "Fraction subtraction (20 items)")
main <- "fitz_2024_numeracy"
two  <- filter(summary_df, model == "2PL")
fmt  <- function(x, d = 2) formatC(x, format = "f", digits = d)
wrap <- function(x, w = 95) if (is.null(x)) x else stringr::str_wrap(x, w)
# Wrap long titles, subtitles and axis labels so they fit the figure width.
wl <- function(p, w = 72) {
  for (nm in c("title", "subtitle", "x")) {
    lab <- p$labels[[nm]]
    if (is.character(lab)) p$labels[[nm]] <- wrap(lab, if (nm == "title") w else round(w * 1.15))
  }
  p
}
```

```{r}
#| label: cache-warning
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/bits_per_item_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, partial run.** This render uses a partial sample of ",
  r_n_tables_run, " tables out of ", r_all_candidates, " eligible tables. ",
  "All numbers here are real computed results, but treat the distributions ",
  "as a small, noisy preview. Re-run `bits_per_item_compute.R` without ",
  "`PILOT=1` to scale up.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`bits_per_item_compute.R`) directly if you're relying
on the numbers here.
:::

```{r}
#| label: ojs-data

ojs_payload <- lapply(names(showcases), function(nm) {
  s <- showcases[[nm]]
  list(
    key = nm, title = unname(show_titles[nm]), instrument = s$instrument,
    items = s$items[, c("label", "text", "label0", "label1", "a", "b", "bits")],
    pg = unname(round(s$group_probs, 5)),
    # group edges on the theta scale; the two open-ended tail groups are drawn to +/- 3.2
    bounds = pmin(pmax(s$group_bounds, -3.2), 3.2)
  )
})
names(ojs_payload) <- names(showcases)
ojs_define(showcases_r = ojs_payload, accent_r = irw_accent, blue_r = irw_blue)
```

## Guess the person

The curve below is the population: how people are spread along θ, the trait
the test measures, from low (left) to high (right). We have cut it into
**64 slices, each holding the same share of people**. The slices are narrow in
the crowded middle and wide in the thin tails. Somewhere in that population is
one person, and your job is to find out which slice they are in by asking
questions from a real test.

Before any question, all 64 slices are equally likely. Pinning down one slice
out of 64 takes **6 bits**, since 2^6^ = 64. After each answer the slices are
re-shaded, darker where the person is now more likely to be, and a line traces
the same thing as a density over θ. The counter shows how many bits that
answer bought.

Pick a question from the table; they are ranked by how many bits they are
*expected* to give. Then either answer it yourself, as the person might, or let
the simulated person answer. At the start we secretly draw one simulated
person from the population; they answer with the probabilities the model
gives someone in their slice. You can reveal where they are at any point.
The "best" and "worst" buttons ask the remaining question with the most or the
fewest expected bits, and the simulated person answers it.

```{ojs}
sc = showcases_r
keys = Object.keys(sc)
viewof tbl = Inputs.select(keys, {label: "Test", value: keys[0], format: k => sc[k].title})
```

```{ojs}
S = {
  const s = sc[tbl];
  return {items: transpose(s.items).map((d, j) => ({...d, j})), pg: s.pg, bounds: s.bounds, title: s.title};
}
h2 = p => (p <= 0 || p >= 1) ? 0 : -(p * Math.log2(p)) - (1 - p) * Math.log2(1 - p)
entropy = post => -post.reduce((s, p) => s + (p > 0 ? p * Math.log2(p) : 0), 0)
mutable hist = []
mutable hidden = Math.floor(Math.random() * 64)
reset_on_test = { tbl; mutable hist = []; mutable hidden = Math.floor(Math.random() * 64); return tbl; }
posterior_of = hs => {
  let p = Array(64).fill(1 / 64);
  for (const h of hs) {
    const like = S.pg.map(row => h.x === 1 ? row[h.j] : 1 - row[h.j]);
    const z = p.reduce((s, v, g) => s + v * like[g], 0);
    p = p.map((v, g) => v * like[g] / z);
  }
  return p;
}
post = { reset_on_test; return posterior_of(hist); }
exp_bits = j => h2(post.reduce((s, v, g) => s + v * S.pg[g][j], 0)) -
                post.reduce((s, v, g) => s + v * h2(S.pg[g][j]), 0)
asked = new Set(hist.map(h => h.j))
remaining = S.items.filter(d => !asked.has(d.j))
  .map(d => ({...d, expected: exp_bits(d.j)}))
  .sort((a, b) => b.expected - a.expected)
answer = (j, x) => {
  const p1 = post.reduce((s, v, g) => s + v * S.pg[g][j], 0);
  const p_obs = x === 1 ? p1 : 1 - p1;
  mutable hist = [...hist, {j, x, expected: exp_bits(j), before: entropy(post), surprisal: -Math.log2(p_obs)}];
}
hidden_answer = j => Math.random() < S.pg[hidden][j] ? 1 : 0
```

```{ojs}
viewof pick = Inputs.table(remaining, {
  columns: ["expected", "text"],
  header: {text: "Question", expected: "Exp. bits"},
  format: {expected: x => x.toFixed(3), text: (t, i, data) => t ?? data[i].label},
  width: {expected: 80},
  multiple: false,
  required: false,
  rows: 8,
  layout: "fixed"
})
```

```{ojs}
chosen = pick ?? remaining[0]
viewof act = Inputs.button(chosen ? [
    [`Answer: ${chosen.label1}`, () => answer(chosen.j, 1)],
    [`Answer: ${chosen.label0}`, () => answer(chosen.j, 0)],
    ["Let the simulated person answer", () => answer(chosen.j, hidden_answer(chosen.j))]
  ] : [["No questions left", () => null]])
viewof act2 = Inputs.button([
    ["Ask the best question", () => { if (remaining.length) { const j = remaining[0].j; answer(j, hidden_answer(j)); } }],
    ["Ask the worst question", () => { if (remaining.length) { const j = remaining[remaining.length - 1].j; answer(j, hidden_answer(j)); } }],
    ["Start over", () => { mutable hist = []; mutable hidden = Math.floor(Math.random() * 64); }]
  ])
viewof reveal = Inputs.toggle({label: "Show where the simulated person is"})
```

```{ojs}
bits_left = entropy(post)
gained = 6 - bits_left
theta_plot = {
  const b = S.bounds;
  // Bars never move: each slice is drawn at the population's height. Its shade
  // is the chance that this person is in it. The line is the updated density.
  const slices = post.map((v, g) => ({g, p: v, x1: b[g], x2: b[g + 1],
    pop: (1 / 64) / (b[g + 1] - b[g]), dens: v / (b[g + 1] - b[g])}));
  const line = slices.flatMap(d => [{x: d.x1, y: d.dens}, {x: d.x2, y: d.dens}]);
  const top = Math.max(0.42, ...slices.map(d => d.dens));
  return Plot.plot({
    width: Math.min(width, 680), height: 250, marginBottom: 40, marginTop: 24,
    x: {domain: [-3.2, 3.2], label: "θ: the trait the test measures (low → high)", labelAnchor: "center", labelArrow: "none"},
    y: {domain: [0, top * 1.08], axis: null},
    color: {type: "linear", domain: [0, Math.max(2 / 64, ...post)], range: ["#f4f7fb", "#08306b"]},
    marks: [
      Plot.rectY(slices, {x1: "x1", x2: "x2", y: "pop", fill: "p", inset: 0.3,
        title: d => `slice ${d.g + 1} of 64\nθ from ${d.x1.toFixed(2)} to ${d.x2.toFixed(2)}\nchance the person is here: ${(d.p * 100).toFixed(1)}%`}),
      hist.length ? Plot.line(line, {x: "x", y: "y", stroke: "#111", strokeWidth: 1.6}) : null,
      reveal ? Plot.ruleX([(slices[hidden].x1 + slices[hidden].x2) / 2], {stroke: "#d73027", strokeWidth: 2.5}) : null,
      reveal ? Plot.text([slices[hidden]], {x: d => (d.x1 + d.x2) / 2, y: top * 1.08, text: () => "the simulated person", dy: -4, lineAnchor: "bottom", fill: "#d73027"}) : null,
      Plot.ruleY([0])
    ]
  });
}
theta_key = html`<div style="font-size:0.85rem; color:#444; display:flex; flex-wrap:wrap; gap:0.4rem 1.2rem; max-width:680px; margin-bottom:2px;">
  <span><svg width="44" height="10"><defs><linearGradient id="kg"><stop offset="0" stop-color="#eef2f8"/><stop offset="1" stop-color="#08306b"/></linearGradient></defs><rect width="44" height="10" fill="url(#kg)" stroke="#ccc"/></svg>
    the population in 64 equal slices, shaded by the chance this person is in each (light → dark: less → more likely)</span>
  <span><svg width="26" height="10"><line x1="0" y1="5" x2="26" y2="5" stroke="#111" stroke-width="1.6"/></svg>
    where this person probably is, as a density (appears after the first answer)</span>
</div>`
counter = html`<div style="font-size:0.95rem; max-width:680px;">
  <div style="display:flex; justify-content:space-between;">
    <span><b style="color:${accent_r}">${gained.toFixed(2)} bits gained</b></span>
    <span>${bits_left.toFixed(2)} of 6 bits left</span></div>
  <div style="font-size:0.9rem; margin-top:2px;">About as certain as knowing the person is in one of
    <b>${Math.pow(2, bits_left).toFixed(Math.pow(2, bits_left) < 10 ? 1 : 0)}</b> equally likely slices (from 64).<sup>*</sup></div>
  <div style="height:14px; background:#e9ecef; border-radius:3px; overflow:hidden; margin:4px 0;">
    <div style="height:100%; width:${Math.max(0, gained) / 6 * 100}%; background:${accent_r};"></div></div>
  <div style="color:#666; font-size:0.85rem;">${hist.length} question${hist.length === 1 ? "" : "s"} asked.
  Before any answer every slice has the same shade: all are equally likely.
  Each answer darkens the slices that would most likely give it, and lightens the rest.</div>
  <div style="color:#666; font-size:0.8rem; margin-top:4px;"><sup>*</sup> An effective count, 2<sup>bits left</sup>.
  The chances are rarely spread evenly over a set of slices and zero elsewhere; they form a bump with tails.
  The count says how many equally likely slices would leave the same uncertainty (entropy) as that bump.</div></div>`
p1_now = chosen ? post.reduce((s, v, g) => s + v * S.pg[g][chosen.j], 0) : null
bars = chosen ? Plot.plot({
  width: Math.min(width, 380), height: 140, marginLeft: 90,
  x: {domain: [0, 1], label: "probability, given what we know", grid: true},
  y: {label: null},
  marks: [
    Plot.barX([{r: chosen.label1, p: p1_now}, {r: chosen.label0, p: 1 - p1_now}],
      {y: "r", x: "p", fill: blue_r}),
    Plot.text([{r: chosen.label1, p: p1_now}, {r: chosen.label0, p: 1 - p1_now}],
      {y: "r", x: "p", text: d => (d.p * 100).toFixed(0) + "%", dx: 16})
  ]
}) : html``
log_rows = hist.map((h, i) => ({
  "#": i + 1,
  answer: h.x === 1 ? S.items[h.j].label1 : S.items[h.j].label0,
  "expected bits": h.expected.toFixed(3),
  surprisal: h.surprisal.toFixed(2),
  question: S.items[h.j].text ?? S.items[h.j].label
}))
html`<div style="display:flex; flex-wrap:wrap; gap:1.5rem; align-items:flex-start;">
  <div>${theta_key}${theta_plot}${counter}</div>
  <div style="flex:1; min-width:260px;">
    ${chosen ? html`<p style="margin-bottom:0.3rem;"><b>Next question:</b> ${chosen.text ?? chosen.label}</p>
    <p style="color:#666; margin-top:0; font-size:0.9rem;">Expected to give
    <b style="color:${accent_r}">${chosen.expected.toFixed(3)} bits</b>, out of a possible 1:
    I(θ; X<sub>j</sub> | answers so far).</p>${bars}` :
    html`<p>Every question has been asked.</p>`}
  </div></div>`
```

```{ojs}
hist.length ? Inputs.table(log_rows, {rows: 6, layout: "fixed",
  width: {"#": 35, answer: 90, "expected bits": 100, surprisal: 80}}) : html``
```

A few things show up quickly. No real question is worth anywhere near a full
bit: the best ones promise about half a bit at most. A full bit would take a
perfectly sharp (Guttman) item whose cut falls exactly in the middle of what we
currently believe, splitting the remaining possibilities in half. Questions
lose value as you go, because the easy distinctions have already been made.
And now and then the shading spreads *out* after an answer. That happens when
the person says something the model did not expect.

::: {.callout-note}
## Which information? The quantities on this page

Several different "bits" appear below. They are not interchangeable.

- **I(θ; X)**: what the full answer pattern says about the person, on average
  over the population. This is "the test's information" in the warehouse
  figures.
- **I(θ; X~j~)**: what one item says *on its own*. Items repeat each other, so
  these do not add up to the test's total: I(θ; X) ≤ Σ~j~ I(θ; X~j~).
- **I(θ; X~j~ | answers so far)**: what one more item is expected to add, given
  the particular answers already seen. These are the "expected bits" in the game
  above, and the same item is worth less once earlier answers have done its job.
- **Bits gained by one answer**: H(θ | before) − H(θ | after). This is a realised
  change, not an average, so it can be negative. Averaged over possible answers
  it equals the expected bits.
- **Surprisal**: −log~2~ P(answer | answers so far), how unexpected one answer was.
  The [Surprise](#surprise) section conditions instead on all the person's other answers.
- **I(θ; S)** and **I(θ; X | S)**: what the sum score keeps, and what it drops.
:::

## What a bit is

```{r}
#| label: bit-strip
#| fig-height: 5.2
#| fig-width: 7.5

s <- showcases[[main]]
pg <- s$group_probs
pick3 <- s$items |> arrange(desc(bits)) |> slice(c(1, ceiling(n() / 2), n()))
bnd <- pmin(pmax(s$group_bounds, -3.2), 3.2)
cells <- tibble(g = 1:64, x1 = bnd[-65], x2 = bnd[-1])
ideal <- mutate(cells, row = "A perfect (Guttman) item at the median",
                p = rep(c(0, 1), each = 32), bits = 1)
real <- bind_rows(lapply(seq_len(nrow(pick3)), function(i) {
  it <- pick3[i, ]
  mutate(cells, row = paste0("\u201c", substr(it$text, 1, 52), if (nchar(it$text) > 52) "\u2026" else "", "\u201d"),
         p = pg[, it$item], bits = it$bits)
}))
strip <- bind_rows(ideal, real) |>
  mutate(row = factor(row, levels = unique(row)), y = 5 - as.integer(row),
         lab = paste0(fmt(bits), if_else(bits == 1, " bit", " bits")))
labs_df <- distinct(strip, row, y, lab)
ggplot(strip) +
  geom_rect(aes(xmin = x1, xmax = x2, ymin = y - 0.4, ymax = y + 0.4, fill = p),
            colour = "white", linewidth = 0.15) +
  geom_text(data = labs_df, aes(x = 3.35, y = y, label = lab),
            hjust = 0, colour = irw_accent, fontface = "bold", size = 4) +
  scale_fill_gradient(low = "#f4f6fa", high = irw_blue, limits = c(0, 1),
                      name = "P(yes / correct)\nin each slice") +
  scale_x_continuous(limits = c(-3.2, 4.3), breaks = -3:3, expand = c(0, 0)) +
  scale_y_continuous(breaks = labs_df$y, labels = stringr::str_wrap(as.character(labs_df$row), 34)) +
  labs(x = "θ, cut into 64 equally likely slices", y = NULL,
       title = sprintf("A perfect item halves the population; the best real numeracy item manages %s of a bit",
                       fmt(max(s$items$bits))),
       subtitle = "Each row is one question, shaded by how likely people in each slice are to say yes / get it right. Right: I(θ; Xⱼ), the item on its own.") +
  theme(panel.grid = element_blank(), legend.position = "bottom", legend.key.width = unit(1.2, "cm"))
```

A bit is the information in the answer to a question whose two answers are
equally likely and which splits the possibilities cleanly. The top row of the
figure does exactly that: it cuts 64 slices to 32. Real items fail in two ways.
Their split is blurred, because people in the same group sometimes answer
differently. And their split is often off-centre, because easy items mostly
separate the lowest groups from everyone else. Formally, the expected
information of item *j* is the entropy of its answer minus the noise in it:
<small>I(θ; X~j~) = h(P̄~j~) − E~θ~[h(P~j~(θ))], with h(p) = −p log~2~ p − (1−p) log~2~(1−p).</small>
Neither term can exceed one bit.

## Surprise

```{r}
#| label: surprise
#| fig-height: 4.2
#| fig-width: 7.5

reps <- s$replays
lab_of <- function(it) s$items$label[match(it, s$items$item)]
# Leave-one-out surprisal: -log2 P(x_j | every other answer), with the posterior
# over the 2^K theta groups built from all items except j. Order plays no part;
# items are shown easiest to hardest only for reading.
loo_surprisal <- function(x, PG) {
  ll <- log(PG) %*% diag(x, length(x)) + log(1 - PG) %*% diag(1 - x, length(x))  # groups x items
  tot <- rowSums(ll)
  vapply(seq_along(x), function(j) {
    lp <- tot - ll[, j]; post <- exp(lp - max(lp)); post <- post / sum(post)
    p1 <- sum(post * PG[, j])
    -log2(if (x[j] == 1) p1 else 1 - p1)
  }, numeric(1))
}
ord_b <- order(s$items$b)
surp <- bind_rows(lapply(reps, function(r) {
  x <- r$responses[s$items$item[ord_b]]
  tibble(respondent = r$respondent, kind = r$kind, step = seq_along(x), item = lab_of(names(x)),
         surprisal = loo_surprisal(x, s$group_probs[, names(x), drop = FALSE]), response = unname(x))
}))
ord_items <- surp |> filter(kind == "aberrant") |> arrange(step) |> pull(item)
b_of <- s$items$b[match(ord_items, s$items$label)]
ab <- filter(surp, kind == "aberrant") |> mutate(answer = if_else(response == 1, "correct", "incorrect"))
ab_resp <- reps[[which(sapply(reps, `[[`, "kind") == "aberrant")]]
n_easy_miss <- sum(ab$response[1:3] == 0)
n_hard_hit <- sum(ab$response[(nrow(ab) - 2):nrow(ab)] == 1)
ggplot(surp, aes(step, surprisal, group = respondent)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = irw_grey) +
  geom_step(data = filter(surp, kind != "aberrant"), colour = irw_grey, alpha = 0.5,
            direction = "mid") +
  geom_step(data = ab, colour = irw_red, linewidth = 1, direction = "mid") +
  geom_point(data = ab, aes(shape = answer), colour = irw_red, fill = "white", size = 2.6, stroke = 1.2) +
  scale_shape_manual(values = c(correct = 16, incorrect = 21), name = "aberrant respondent's answer") +
  scale_x_continuous(breaks = seq_along(ord_items),
                     labels = sprintf("%s\n(b = %s)", ord_items, formatC(b_of, format = "f", digits = 1))) +
  labs(x = "questions, from easiest (left) to hardest (right)",
       y = "surprisal (bits)",
       title = sprintf("The aberrant respondent misses %d of the 3 easiest questions and gets %d of the 3 hardest right",
                       n_easy_miss, n_hard_hit),
       subtitle = sprintf("Red: the respondent with the worst person-fit (Zh = %s, sum score %d of %d). Grey: five typical respondents. Surprisal = −log₂ P(answer | all their other answers). Dashed: 1 bit, a fair coin flip.",
                          fmt(ab_resp$zh, 1), ab_resp$sum_score, nrow(s$items))) +
  theme(legend.position = "top", axis.text.x = element_text(size = 7.5))
```

The surprisal of an answer is −log~2~ of the probability the model gave it.
An answer the model thought had a 50% chance costs 1 bit of surprise; one it
thought had a 1-in-8 chance costs 3 bits. Here each answer is judged against
*everything else the person answered*: hide one answer, work out from the
others where the person probably sits, and ask how likely the hidden answer
was from there. Nothing depends on the order of asking; the questions are
lined up from easiest to hardest only so the pattern is easy to read.

Most answers from typical respondents sit around or below a bit, because their
other answers place them well and each answer agrees with that placement. The
red respondent's answers disagree with each other. The rest of the pattern says
"able enough for the hard questions", which makes the misses on the easy ones
very unlikely, and vice versa, so the surprise lands on exactly the answers
that do not fit. This is the logic of person-fit statistics such as Zh, one
answer at a time.

Elsewhere on the page, as in the game at the top, surprisal is judged against
the answers *so far*. That version depends on order, and it adds up: the
surprisals of one ordering sum to the surprisal of the whole pattern. The
leave-one-out version does not add up that way, but it gives each answer one
number whatever the order, which is what a question about fit needs.

<!-- Expected versus actual: commented out 2026-10-06 (Ben); restore by removing this wrapper and the eval/include lines.

## Expected versus actual

```{r}
#| label: expected-actual
#| eval: false
#| include: false
#| fig-height: 4.6
#| fig-width: 7.5

ea <- bind_rows(lapply(reps[c(3, which(sapply(reps, `[[`, "kind") == "aberrant"))], function(r) {
  st <- r$fixed$steps[-1, ]
  lbl <- s$items$label[match(st$item, s$items$item)]
  tibble(who = if (r$kind == "aberrant") "Aberrant respondent" else "Typical respondent (median score)",
         step = st$step, item = lbl, expected = st$expected_bits, actual = st$actual_bits)
})) |>
  pivot_longer(c(expected, actual), names_to = "kind", values_to = "bits") |>
  mutate(kind = factor(kind, c("expected", "actual"), c("expected bits", "bits actually gained")))
tot <- ea |> group_by(who, kind) |> summarise(bits = sum(bits), .groups = "drop")
ggplot(ea, aes(factor(step), bits, fill = kind)) +
  geom_hline(yintercept = 0, colour = "grey30") +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~ factor(who, c("Typical respondent (median score)", "Aberrant respondent")), ncol = 1) +
  scale_fill_manual(values = c("expected bits" = "grey75", "bits actually gained" = irw_accent), name = NULL,
                    labels = c("expected bits: I(θ; Xⱼ | earlier answers)", "bits actually gained: H(θ | before) − H(θ | after)")) +
  labs(x = "question, in test order", y = "bits",
       title = "Information is a promise on average: a single answer can even lose bits",
       subtitle = sprintf("Totals over the test. Typical: expected %s, gained %s. Aberrant: expected %s, gained %s.",
                          fmt(tot$bits[tot$who != "Aberrant respondent" & tot$kind == "expected bits"]),
                          fmt(tot$bits[tot$who != "Aberrant respondent" & tot$kind == "bits actually gained"]),
                          fmt(tot$bits[tot$who == "Aberrant respondent" & tot$kind == "expected bits"]),
                          fmt(tot$bits[tot$who == "Aberrant respondent" & tot$kind == "bits actually gained"]))) +
  theme(legend.position = "top")
```

The expected bits of an item are an average over everyone who might be
answering. For a particular person, the drop in uncertainty can be larger
(a surprising answer that rules out a lot) or negative (an answer that
contradicts what earlier answers suggested, spreading the posterior back
out). Averaged over people, actual gains equal expected gains. That is the
sense in which I(θ; X) is a property of the test and the population, not of
any one respondent.

-->

## Watching uncertainty fall

```{r}
#| label: uncertainty
#| fig-height: 4.4
#| fig-width: 7.5

unc <- s$uncertainty |>
  mutate(order = factor(order, c("fixed", "random", "adaptive"),
                        c("test's own order", "random order", "adaptive: most informative next")))
trails <- bind_rows(lapply(reps, function(r) tibble(respondent = r$respondent,
                                                     k = r$fixed$steps$step, entropy = r$fixed$steps$entropy)))
end_bits <- 6 - filter(unc, k == max(k))$mean_entropy[1]
ad3 <- filter(unc, order == "adaptive: most informative next", k == 3)$mean_entropy
fx3 <- filter(unc, order == "test's own order", k == 3)$mean_entropy
ggplot() +
  geom_line(data = trails, aes(k, entropy, group = respondent), colour = irw_grey, alpha = 0.35) +
  geom_line(data = unc, aes(k, mean_entropy, colour = order, linetype = order), linewidth = 1.1) +
  scale_colour_manual(values = c(irw_blue, irw_grey, "black"), name = NULL) +
  scale_linetype_manual(values = c("solid", "dashed", "solid"), name = NULL) +
  scale_y_continuous(limits = c(NA, 6.05), breaks = 0:6) +
  scale_x_continuous(breaks = 0:nrow(s$items)) +
  labs(x = "questions asked", y = "bits left: H(slice | answers so far)",
       title = sprintf("After all %d numeracy questions, about %s of the 6 bits are resolved",
                       nrow(s$items), fmt(end_bits, 1)),
       subtitle = sprintf("Mean over 500 simulated respondents. Adaptive ordering front-loads the gains (%s vs %s bits left after 3 questions). Faint lines: six real respondents.",
                          fmt(ad3), fmt(fx3))) +
  theme(legend.position = "top")
```

Every order ends in the same place, because the same answers have been given.
What changes is how fast you get there. Asking the most informative question
next, given what is already known, buys in three questions what the fixed
order needs several more to reach. This is the logic of computerised adaptive
testing, and of Wordle solvers. The ending point is the real news, though:
a full pass through this test leaves most of the uncertainty about which of
64 groups a person belongs to unresolved.

## Patterns

```{r}
#| label: patterns
#| fig-height: 4.4
#| fig-width: 7.5

pt <- s$patterns |> mutate(rank = row_number())
top5 <- s$items |> arrange(desc(bits)) |> slice(1:5)
gth <- s$grid
P5 <- plogis(outer(gth$theta, top5$a) - matrix(top5$a * top5$b, nrow(gth), 5, byrow = TRUE))
H5 <- -sum(pt$implied * log2(pt$implied))
H5_noise <- sum(gth$w * rowSums(h2(P5)))
ggplot(pt, aes(rank, implied)) +
  geom_col(fill = irw_blue, width = 0.8) +
  geom_point(aes(y = observed), colour = "black", size = 1.8) +
  scale_x_continuous(breaks = pt$rank, labels = pt$pattern, expand = c(0.01, 0.01)) +
  labs(x = "answer pattern on the 5 most informative items (1 = correct)",
       y = "probability",
       title = sprintf("Five questions, 32 possible answer patterns, %s bits about the person",
                       fmt(H5 - H5_noise)),
       subtitle = sprintf("Bars: model-implied probability. Dots: observed share of %s respondents. Entropy of the bars %s bits, minus %s bits of noise. Items, in pattern order: %s.",
                          format(pt$n_obs[1], big.mark = ","), fmt(H5), fmt(H5_noise),
                          paste(top5$label, collapse = ", "))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, family = "mono", size = 8))
```

This is the Wordle bar chart translated. In Wordle, a guess is good when its
colour patterns are spread out, so that whichever one you see tells you a lot.
For a test, the spread of the 32 answer patterns is their entropy. But here,
unlike in Wordle, the same person could produce different patterns on
different days. That noise, H(X | θ), has to be subtracted before what is left
counts as information about the person. Most of the entropy in the chart is
noise.

## How the warehouse numbers were computed

Selection started from every dichotomous IRW table with 5–60 items and at
least 500 respondents. Each candidate was then fetched and checked. Tables
with waves or treatment arms were restricted to one administration: the
earliest wave, then the control arm. Tables with repeated person–item rows
(trial-level tasks) were excluded, as were tables with response density below
0.8. ENEM tables were left out for this version. That left
`r r_all_candidates` tables, of which `r n_distinct(two$table)` produced
results (`r length(failures)` failed: `r paste(unique(sub(" .*", "", failures)), collapse = ", ")`,
where every respondent gives the same answer to every item). Tables with more
than 10,000 respondents were randomly subsampled to 10,000
(`r sum(two$subsampled)` tables). Items answered by fewer than half the
respondents, and items with no variance, were dropped.

For each table we fit Rasch and 2PL models with `mirt`, using a lognormal(0, 1)
prior on 2PL slopes, and read the latent distribution from each fit: N(0, σ²)
with σ² estimated for Rasch, N(0, 1) for 2PL. All quantities are integrals over
that distribution on an 81-point grid spanning ±6 SD. In bits:

- **I(θ; S)**, the information in the sum score, via the Lord–Wingersky recursion.
- **I(θ; X)**, the information in the full response pattern. It is exact by
  enumerating all patterns when there are 16 or fewer items, and Monte Carlo
  above that (standard error below 0.005 bits). For Rasch the two are equal,
  because the sum score is sufficient.
- **H(X | θ)**, the noise, and H(X) = I(θ; X) + H(X | θ).
- Marginal reliability, the reliability of EAP sum scores, and their Gaussian
  bit equivalents −½ log~2~(1 − ρ).
- Per-item bits, length curves, and a sum-score model check.

Validation checks (entropy edge cases, single items under 1 bit, Rasch
sufficiency, I(θ; S) ≤ I(θ; X) on every table, Monte Carlo against enumeration,
binned against continuous, simulation recovery, grid stability) are in
`bits_per_item_checks.R`. All passed.

## Now, the whole warehouse

```{r}
#| label: corpus
#| fig-height: 4
#| fig-width: 7.5

med <- median(two$I_X)
p <- ggplot(two, aes(I_X)) +
  geom_histogram(binwidth = 0.1, boundary = 0, fill = irw_accent, colour = "white") +
  geom_vline(xintercept = med, linetype = "dashed") +
  annotate("text", x = med + 0.03, y = Inf, vjust = 1.5, hjust = 0,
           label = sprintf("median %s bits", fmt(med))) +
  scale_x_continuous(breaks = seq(0, 3, 0.5)) +
  labs(x = "information about the person, I(θ; X), in bits (2PL)", y = "tables",
       title = sprintf("Most tests tell you less than 2 bits about a person: %d of %d do",
                       sum(two$I_X < 2), nrow(two)),
       subtitle = sprintf("%d dichotomous IRW tables. Less than 2 bits means fewer than 4 equally likely groups resolved.",
                          nrow(two)))
p
```

Across `r nrow(two)` tables, the median test delivers `r fmt(med)` bits about
the person it measures: about as much as learning which half of the
population someone is in. The most informative,
`r arrange(two, desc(I_X))$table[1]` (`r arrange(two, desc(I_X))$n_items[1]` items),
reaches `r fmt(max(two$I_X))` bits. Tests are short and noisy relative to the
population spread, and bits grow only logarithmically with precision: each
extra bit requires halving the standard error.

```{r}
#| label: corpus-items
#| fig-height: 4.6
#| fig-width: 7.5

ceil <- tibble(n = 5:60, bits = log2(5:60 + 1))
share_ceiling <- two$I_X / log2(two$n_items + 1)
g_items <- ggplot(two, aes(n_items, I_X)) +
  geom_line(data = ceil, aes(n, bits), colour = "grey30", linetype = "dashed") +
  annotate("text", x = 40, y = log2(41) + 0.35, label = "ceiling: n perfect items, log₂(n + 1)",
           size = 3.4, colour = "grey30") +
  geom_point(aes(text = sprintf("%s\n%d items\nI = %.2f bits", table, n_items, I_X)),
             colour = irw_accent, alpha = 0.7, size = 1.8) +
  scale_x_continuous(breaks = c(5, 10, 20, 30, 40, 50, 60)) +
  scale_y_continuous(limits = c(0, 6.2), breaks = 0:6) +
  labs(x = "number of items", y = "I(θ; X), bits (2PL)",
       title = sprintf("Longer tests give more bits, but the median test reaches only %s of what that many perfect items could",
                       scales::percent(median(share_ceiling), accuracy = 1)))
g_items <- wl(g_items, 60)
g_items$labels$title <- gsub("\n", "<br>", g_items$labels$title)
ggplotly(g_items, tooltip = "text") |>
  layout(title = list(text = paste0("<b>", g_items$labels$title, "</b>"), x = 0, xanchor = "left",
                      font = list(size = 15)), margin = list(t = 70))
```

The dashed line is a hard ceiling. On a single trait, n items that never make
mistakes can produce only n + 1 different answer patterns, so they carry at
most log~2~(n + 1) bits, reached when the items cut the population into n + 1
equal parts. Real tables sit far below it, a median of
`r scales::percent(median(share_ceiling), accuracy = 1)` of the way up. The
missing bits are lost to noise, not to test length: answers are probabilistic,
so the same person could produce many patterns.

## Reliability in bits

Reliability and bits measure related things. If θ and a score were jointly
normal, with the score's correlation with θ equal to √ρ, the information
between them would be exactly −½ log~2~(1 − ρ). Item responses are binary, not
normal, so the question is how far real tests depart from that curve, and
why. We look first at synthetic tests, where we control the items, and then at
the warehouse.

Turning that formula around, b bits corresponds to a reliability of
1 − 4^−b^. To see what that buys, simulate people from the Gaussian picture
and compare each person's true rank with the rank the score gives them.

```{r}
#| label: rank-panels
#| fig-height: 6.6
#| fig-width: 7.5

set.seed(20261006)
bits_show <- c(0, 1, 2, 3, 4, 6)
rank_stats <- function(b, n = 1e6) {
  r <- sqrt(1 - 4^-b)
  th <- rnorm(n); sc <- r * th + sqrt(1 - r^2) * rnorm(n)
  c(rho = r^2, rank_err = 100 * mean(abs(pnorm(th) - pnorm(sc))),
    concord = 0.5 + asin(r) / pi)          # P(pair in the right order), exact
}
rs <- as_tibble(t(sapply(bits_show, rank_stats))) |> mutate(bits = bits_show)
rs <- rs |> mutate(lab = sprintf("%s %s · reliability %s\ntypical rank error %s points\n%.1f%% of pairs in the right order",
                                 bits, if_else(bits == 1, "bit", "bits"), signif(rho, 4),
                                 if_else(rank_err < 1, sprintf("%.1f", rank_err), sprintf("%.2g", rank_err)),
                                 100 * concord),
                   lab = factor(lab, lab))
dots <- rs |> select(bits, lab) |> rowwise() |>
  reframe(lab = lab, th = rnorm(600)) |>
  left_join(select(rs, lab, rho), by = "lab") |>
  mutate(sc = sqrt(rho) * th + sqrt(1 - rho) * rnorm(n()),
         x = 100 * pnorm(th), y = 100 * pnorm(sc))
ggplot(dots, aes(x, y)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(alpha = 0.5, size = 0.9) +
  facet_wrap(~ lab, ncol = 3) +
  scale_x_continuous(breaks = c(0, 50, 100)) + scale_y_continuous(breaks = c(0, 50, 100)) +
  labs(x = "person's true rank in the population (percentile)",
       y = "rank the test gives them (percentile)",
       title = "How well a test ranks people, by how much it tells you about them",
       subtitle = sprintf("Each panel: 600 simulated people, θ and score jointly normal. On the dashed line the ranking is perfect; each extra bit halves the scatter around it. Rank errors are in percentile points. The median warehouse test carries %s bits; the most informative, %s.",
                          fmt(median(two$I_X)), fmt(max(two$I_X)))) +
  theme(strip.text = element_text(hjust = 0, size = 9), panel.spacing = unit(0.8, "lines"))
```

Most of what a ranking needs arrives in the first bit or two: one bit already
puts five pairs in six in the right order. The panels show the Gaussian
picture, which is the dashed curve below. The warehouse tests sit a little
above that curve, so their bits are only approximately where they land in these
panels.

```{r}
#| label: reliability-sim
#| fig-height: 5.6
#| fig-width: 7.5

sim_sum <- rel_sim |> group_by(kind) |>
  summarise(dev_lo = min(I_X - gauss), dev_hi = max(I_X - gauss), .groups = "drop")
sim_dev <- rel_sim |> group_by(kind, n_items) |> summarise(dev = mean(I_X - gauss), .groups = "drop")
curve_s <- tibble(rho = seq(0, 0.97, by = 0.005), bits = -0.5 * log2(1 - rho))
ggplot(rel_sim, aes(rho, I_X, colour = kind)) +
  geom_line(data = curve_s, aes(rho, bits), inherit.aes = FALSE, colour = "grey30", linetype = "dashed") +
  geom_point(aes(size = n_items), alpha = 0.6) +
  scale_colour_manual(values = c(moderate = irw_blue, `sharp items` = irw_red, `one tail` = "#e6a100"),
                      labels = c(moderate = "moderate items, spread across θ",
                                 `sharp items` = "3 near-Guttman items",
                                 `one tail` = "items bunched at the high end"),
                      name = NULL) +
  scale_size_continuous(range = c(1, 3.5), breaks = c(5, 20, 60), name = "items") +
  labs(x = "marginal reliability", y = "I(θ; X), bits",
       title = "In simulation, ordinary items follow the Gaussian curve; sharp items and one-sided tests rise above it",
       subtitle = sprintf("Synthetic 2PL tests, θ ~ N(0, 1), 5–60 items, 10 random draws each. Gap above the curve: moderate %s–%s bits; sharp %s–%s; one tail %s–%s.",
                          fmt(max(0, sim_sum$dev_lo[sim_sum$kind == "moderate"])), fmt(sim_sum$dev_hi[sim_sum$kind == "moderate"]),
                          fmt(sim_sum$dev_lo[sim_sum$kind == "sharp items"]), fmt(sim_sum$dev_hi[sim_sum$kind == "sharp items"]),
                          fmt(sim_sum$dev_lo[sim_sum$kind == "one tail"]), fmt(sim_sum$dev_hi[sim_sum$kind == "one tail"]))) +
  guides(colour = guide_legend(nrow = 2, override.aes = list(size = 3))) +
  theme(legend.position = "top", legend.box = "vertical")
```

With ordinary items, moderately discriminating and spread across θ, the
Gaussian formula is close: reliability tells you the bits. Two departures push
tests above the curve. A few **near-Guttman items** split the population almost
deterministically. That is worth bits a reliability coefficient, built on
squared error, undercounts. The effect is largest in short tests
(`r fmt(sim_dev$dev[sim_dev$kind == "sharp items" & sim_dev$n_items == 5])` bits at 5 items)
and fades as ordinary items accumulate
(`r fmt(sim_dev$dev[sim_dev$kind == "sharp items" & sim_dev$n_items == 60])` at 60).
Items **bunched in one tail** make the score precise at one end and vague
elsewhere, and that gap grows with test length
(`r fmt(sim_dev$dev[sim_dev$kind == "one tail" & sim_dev$n_items == 5])` →
`r fmt(sim_dev$dev[sim_dev$kind == "one tail" & sim_dev$n_items == 60])` bits).

```{r}
#| label: reliability
#| fig-height: 5
#| fig-width: 7.5

curve_df <- tibble(rho = seq(0, 0.97, by = 0.005), bits = -0.5 * log2(1 - rho))
refs <- tibble(rho = c(0.8, 0.9, 0.95), bits = -0.5 * log2(1 - rho))
dev <- two$I_X - two$bits_rho_marginal
max_a_tbl <- tapply(items_df$a, items_df$table, max)
g <- ggplot(two, aes(rho_marginal, I_X)) +
  geom_line(data = curve_df, aes(rho, bits), colour = "grey30", linetype = "dashed") +
  geom_point(aes(text = sprintf("%s\n%d items\nρ = %.2f\nI = %.2f bits", table, n_items, rho_marginal, I_X)),
             colour = irw_accent, alpha = 0.7, size = 1.8) +
  geom_point(data = refs, aes(rho, bits), shape = 21, fill = "white", size = 3) +
  geom_text(data = refs, aes(rho, bits, label = sprintf("ρ = %.2f\n%.2f bits", rho, bits)),
            nudge_x = -0.07, nudge_y = 0.12, size = 3.2) +
  labs(x = "marginal reliability (2PL)", y = "I(θ; X), bits",
       title = sprintf("Reliability predicts bits closely, but the Gaussian formula undercounts by a median %s bits",
                       fmt(median(dev))),
       subtitle = "Dashed: −½ log₂(1 − ρ), the information if θ and its estimate were jointly normal.")
gauss_note <- "Dashed line: −½ log₂(1 − ρ), the information if θ and its estimate were jointly normal. Hover a point for the table."
g <- wl(g, 55)
g$labels$title <- gsub("\n", "<br>", g$labels$title)
g$labels$subtitle <- NULL
ggplotly(g, tooltip = "text") |>
  layout(title = list(text = paste0("<b>", g$labels$title, "</b>"), x = 0, xanchor = "left",
                      font = list(size = 15)), margin = list(t = 70))
```

`r gauss_note`

On the Gaussian curve, a reliability of .80 is worth `r fmt(-0.5 * log2(0.2))` bits, .90 is worth `r fmt(-0.5 * log2(0.1))`,
and .95 is worth `r fmt(-0.5 * log2(0.05))`. The tables follow the curve
(correlation `r fmt(cor(two$I_X, two$bits_rho_marginal))`), but almost all sit
above it: the exact information exceeds the Gaussian value in
`r sum(dev > 0)` of `r nrow(two)` tables, by up to `r fmt(max(dev))` bits.
Binary responses are not normal. The gap tracks the sharpest item in the
table (Spearman correlation `r fmt(cor(two$I_X - two$bits_rho_marginal, max_a_tbl[two$table], method = "spearman"))`
with the largest slope): a near-deterministic item splits the population
cleanly, which is worth bits that a reliability coefficient, built on
squared error, does not register.

## Diminishing returns

```{r}
#| label: length
#| fig-height: 4.6
#| fig-width: 7.5

kmax <- 40
med_curve <- curves_df |>
  group_by(k) |> filter(n() >= 20) |>
  summarise(random = median(random_mean), greedy = median(greedy), n = n(), .groups = "drop") |>
  filter(k <= kmax)
first <- curves_df |> filter(k == 1) |> pull(random_mean) |> median()
g5_all <- curves_df |> group_by(table) |> mutate(n = max(k), full = random_mean[k == n]) |>
  ungroup() |> filter(k == 5) |> mutate(greedy_share = greedy / full, random_share = random_mean / full)
g5 <- filter(g5_all, n >= 15)
at10 <- med_curve$random[med_curve$k == 10]
ggplot() +
  geom_line(data = filter(curves_df, k <= kmax), aes(k, random_mean, group = table),
            colour = irw_grey, alpha = 0.15) +
  geom_line(data = med_curve, aes(k, random, linetype = "median, random order"),
            colour = irw_accent, linewidth = 1.3) +
  geom_line(data = med_curve, aes(k, greedy, linetype = "median, greedy order"),
            colour = "black", linewidth = 0.9) +
  scale_linetype_manual(values = c("median, random order" = "solid", "median, greedy order" = "dashed"), name = NULL) +
  labs(x = "number of items in the sum score", y = "I(θ; Sₖ), bits",
       title = sprintf("The first item gives a median %s bits; the first ten together only %s",
                       fmt(first), fmt(at10)),
       subtitle = sprintf("Faint: each table's curve, mean of 5 random item orders. Medians use only lengths reached by at least 20 tables (k ≤ %d).",
                          max(med_curve$k))) +
  theme(legend.position = "top")
```

Each additional item adds less than the one before, because it mostly repeats
what earlier items already said about the person. The greedy order (best
item first, then the item that adds most to it, and so on) shows how much of a
test's information a short form could keep. Among the
`r nrow(g5)` tables with at least 15 items, the 5 greedy items carry a median
`r scales::percent(median(g5$greedy_share), accuracy = 1)` of the full sum
score's bits; 5 random items carry `r scales::percent(median(g5$random_share), accuracy = 1)`.

One curiosity: for a sum score, more items is not always better. Adding a
flat item to an unweighted sum adds noise without adding signal, so the
curve can turn down. In `r sum(g5_all$greedy_share > 1)` tables the best
five items alone carry more information than the full sum score, up to
`r fmt(max(g5_all$greedy_share), 1)` times as much
(`r g5_all$table[which.max(g5_all$greedy_share)]`).

## Where do the bits come from?

```{r}
#| label: item-bits
#| fig-height: 4.4
#| fig-width: 7.5

ib <- items_df |> filter(is.finite(b))
pa <- ggplot(ib, aes(a, bits)) +
  geom_bin_2d(bins = 50) +
  scale_fill_gradient(low = "#e5f5f0", high = irw_accent, trans = "log10", name = "items") +
  scale_x_log10() +
  labs(x = "discrimination a (log scale)", y = "I(θ; Xⱼ), the item alone (bits)")
pb <- ggplot(filter(ib, abs(b) < 6), aes(b, bits)) +
  geom_bin_2d(bins = 50) +
  scale_fill_gradient(low = "#e5f5f0", high = irw_accent, trans = "log10", name = "items") +
  labs(x = "difficulty b (θ units)", y = NULL) +
  guides(fill = "none")
(pa | pb) + plot_layout(guides = "collect") +
  plot_annotation(title = wrap(sprintf("Bits need both a sharp item and a central one; the median item gives %s bits",
                                  fmt(median(items_df$bits), 2))),
                  subtitle = sprintf("%s items from %d tables (2PL).", format(nrow(items_df), big.mark = ","),
                                     n_distinct(items_df$table)))
```

An item's bits depend on two things. Discrimination is how sharply the item
separates people just below its difficulty from people just above. Position
is where that cut falls. A highly discriminating item placed far in a tail
still says little about most people, because nearly everyone gives the same
answer to it. The ceiling is 1 bit, reached only by a perfectly sharp item at
the population median.

## What does sum-scoring cost?

```{r}
#| label: sum-gap
#| fig-height: 3.8
#| fig-width: 7.5

two <- mutate(two, gap = I_X - I_S, gap_prop = gap / I_X)
ggplot(two, aes(gap_prop)) +
  geom_histogram(binwidth = 0.02, boundary = 0, fill = irw_accent, colour = "white") +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "I(θ; X | S) as a share of I(θ; X): information lost by keeping only the sum score",
       y = "tables",
       title = sprintf("Using the sum score instead of the full pattern costs a median %s of the bits",
                       scales::percent(median(two$gap_prop), accuracy = 1)),
       subtitle = sprintf("Median I(θ; X | S) %s bits; largest %s bits (%s). 2PL.",
                          fmt(median(two$gap), 3), fmt(max(two$gap)),
                          arrange(two, desc(gap))$table[1]))
```

The gap between the full pattern and the sum score has a name: it is the
conditional mutual information I(θ; X | S), the information about θ in *which*
items a person got right beyond *how many*. Under the Rasch model it is zero:
the sum score carries every bit the pattern carries. Under
the 2PL, items with higher slopes deserve more weight, and an unweighted sum
throws some information away. The loss is small: a median of
`r scales::percent(median(two$gap_prop), accuracy = 1)` of the pattern's
information. It exceeds a quarter in `r sum(two$gap_prop > 0.25)` tables.
The loss tracks how unequal the slopes are (Spearman correlation
`r fmt(cor(two$gap_prop, tapply(log(items_df$a), items_df$table, sd)[two$table], method = "spearman"))`
with the spread of log slopes): tables that mix very sharp items with very
flat ones lose the most.

## Is the sum score enough?

Every score is computed from the answers, so θ → X → S is a chain: the score
can only know what the answers know. That is the **data processing
inequality**, I(θ; S) ≤ I(θ; X). The chain rule splits the full pattern's
information exactly into what the score keeps and what it drops:

<small>I(θ; X) = I(θ; S) + I(θ; X | S).</small>

A statistic is **sufficient** for θ exactly when nothing is dropped,
I(θ; X | S) = 0, *whatever the population* [@cover2006elements, §2.9].
Equivalently, once you know the score, the rest of the pattern says nothing more
about the person. Under the Rasch model the sum score is sufficient; under the
2PL the slope-weighted score Σ a~j~x~j~ is, and the plain sum is not.

```{r}
#| label: suff-bars
#| fig-height: 3.6
#| fig-width: 7.5

lad <- suff[[main]]$ladder
bars <- bind_rows(
  tibble(model = lad$model, part = "kept by the score: I(θ; S)", bits = lad$I_S),
  tibble(model = lad$model, part = "dropped by the score: I(θ; X | S)", bits = pmax(lad$I_X_given_S, 0))
) |>
  mutate(model = factor(model, c("Rasch", "2PL")),
         part = factor(part, c("dropped by the score: I(θ; X | S)", "kept by the score: I(θ; S)")))
full <- lad |> mutate(model = factor(model, c("Rasch", "2PL")))
ggplot(bars, aes(bits, model, fill = part)) +
  geom_col(width = 0.55) +
  geom_errorbarh(data = full, aes(xmin = 0, xmax = I_X, y = model), inherit.aes = FALSE,
                 height = 0.75, colour = "grey30") +
  geom_text(data = full, aes(x = I_X, y = model, label = sprintf("I(θ; X) = %s", fmt(I_X, 3))),
            inherit.aes = FALSE, hjust = -0.08, size = 3.6) +
  annotate("text", x = lad$I_X[lad$model == "2PL"], y = 2.42, hjust = 1, size = 3.4, colour = irw_red,
           label = sprintf("I(θ; X | S) = %s bits", fmt(lad$I_X_given_S[lad$model == "2PL"], 3))) +
  scale_fill_manual(values = c("kept by the score: I(θ; S)" = irw_accent,
                               "dropped by the score: I(θ; X | S)" = irw_red), name = NULL) +
  scale_x_continuous(limits = c(0, max(full$I_X) * 1.35), expand = c(0, 0)) +
  labs(x = "bits about θ (numeracy test)", y = NULL,
       title = "Under Rasch the sum score keeps every bit; under the 2PL a sliver is lost",
       subtitle = "θ → X → S. The bracket is everything the full answer pattern says about the person.") +
  theme(legend.position = "top", panel.grid.major.y = element_blank())
```

### For every population

Sufficiency is a statement about every prior, not just the population we
happened to sample. To see that, we hold each model's item parameters fixed and
recompute the information under 16 different populations: normal
distributions centred anywhere from −1.5 to 1.5 with SDs of 0.5, 1 and 2, plus
a two-humped mixture.

```{r}
#| label: suff-sweep
#| fig-height: 7.2
#| fig-width: 7.5

sw <- bind_rows(lapply(suff, `[[`, "sweep")) |>
  mutate(test = show_titles[table],
         test = if_else(table == "frac20", "Fraction subtraction (top 12 of 20 items)", test),
         sd_lab = if_else(prior_kind == "mixture", "bimodal mixture", paste0("SD ", prior_sd)),
         x = if_else(prior_kind == "mixture", 2.4, prior_mean),
         model = factor(model, c("Rasch", "2PL")))
rng <- sw |> filter(model == "2PL") |> group_by(test) |>
  summarise(lo = min(I_X_given_S), hi = max(I_X_given_S), .groups = "drop")
p_sweep <- ggplot(sw, aes(x, I_X_given_S, colour = sd_lab, shape = model, group = interaction(sd_lab, model))) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_line(data = filter(sw, prior_kind == "normal"), linewidth = 0.7) +
  geom_point(size = 2.2) +
  facet_wrap(~ test, ncol = 1, scales = "free_y") +
  scale_y_continuous(n.breaks = 3) +
  scale_x_continuous(breaks = c(-1.5, -0.5, 0, 0.5, 1.5, 2.4),
                     labels = c("−1.5", "−0.5", "0", "0.5", "1.5", "bimodal")) +
  scale_colour_manual(values = c("SD 0.5" = "#9ecae1", "SD 1" = irw_blue, "SD 2" = "#08306b",
                                 "bimodal mixture" = irw_red), name = "population") +
  scale_shape_manual(values = c(Rasch = 1, `2PL` = 16), name = "model") +
  labs(x = "population mean of θ", y = "I(θ; X | S), bits",
       title = "Rasch's sum score loses nothing in any population; the 2PL's loss depends on who is tested",
       subtitle = sprintf("Rasch: zero to machine precision in all 48 cases. 2PL: %s.",
                          paste(sprintf("%s %s–%s bits", sub(" \\(.*", "", rng$test), fmt(rng$lo, 3), fmt(rng$hi, 3)), collapse = "; "))) +
  theme(legend.position = "top", legend.box = "horizontal", legend.title = element_text(size = 10),
        legend.text = element_text(size = 9), panel.spacing.y = unit(1, "lines"))
th <- seq(-4, 4, length.out = 200)
shapes <- bind_rows(
  tibble(prior = "N(0, 0.5²)", theta = th, d = dnorm(th, 0, 0.5)),
  tibble(prior = "N(0, 1²)", theta = th, d = dnorm(th, 0, 1)),
  tibble(prior = "N(0, 2²)", theta = th, d = dnorm(th, 0, 2)),
  tibble(prior = "bimodal", theta = th, d = 0.5 * dnorm(th, -1.5, 0.5) + 0.5 * dnorm(th, 1.5, 0.5))
) |> mutate(prior = factor(prior, unique(prior)))
p_shapes <- ggplot(shapes, aes(theta, d, colour = prior)) +
  geom_line(linewidth = 0.8, show.legend = FALSE) +
  facet_wrap(~ prior, nrow = 1) +
  scale_colour_manual(values = c("#9ecae1", irw_blue, "#08306b", irw_red)) +
  labs(x = NULL, y = NULL) +
  theme_void(base_size = 10) + theme(strip.text = element_text(size = 9))
wl(p_sweep) / p_shapes + plot_layout(heights = c(6, 1))
```

The populations matter in two ways. First, the total I(θ; X) changes a lot with
the population: a narrow population has less to resolve, and a population far
from the items' difficulties gets mostly uninformative answers. Second, under
the 2PL the share of that information hiding in *which* items were answered
moves with it, because which items are informative depends on where people are.
Under Rasch none of this touches sufficiency: the gap is zero for every
population, as the definition requires.

### Compression

```{r}
#| label: suff-ladder
#| fig-height: 4.8
#| fig-width: 7.5

ladr <- bind_rows(lapply(suff, `[[`, "ladder")) |> filter(model == "Rasch") |>
  mutate(test = show_titles[table],
         test = if_else(table == "frac20", "Fraction subtraction (top 12 items)", test))
lad_long <- bind_rows(
  transmute(ladr, test, step = "H(X), patterns", bits = H_X),
  transmute(ladr, test, step = "H(S), score", bits = H_S),
  transmute(ladr, test, step = "I(θ; S) = I(θ; X)", bits = I_S)
) |> mutate(step = factor(step, rev(unique(step))))
ggplot(lad_long, aes(bits, step, fill = step)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = fmt(bits)), hjust = -0.15, size = 3.5) +
  facet_wrap(~ test, ncol = 1) +
  scale_fill_manual(values = c("H(X), patterns" = "grey70", "H(S), score" = "grey45",
                               "I(θ; S) = I(θ; X)" = irw_accent), guide = "none") +
  scale_x_continuous(limits = c(0, max(lad_long$bits) * 1.12), expand = c(0, 0)) +
  labs(x = "bits (Rasch)", y = NULL,
       title = sprintf("The score throws away %s–%s bits of pattern entropy and not one bit about the person",
                       fmt(min(ladr$H_X - ladr$H_S), 1), fmt(max(ladr$H_X - ladr$H_S), 1)),
       subtitle = "H(X): entropy of the answer patterns. H(S): entropy of the sum score. I(θ; S): what the score says about the person, which under Rasch equals I(θ; X).") +
  theme(panel.grid.major.y = element_blank())
```

Under Rasch the sum score is a **minimal sufficient statistic**: it collapses
up to 2^n^ patterns into n + 1 scores, the most compression possible without
losing anything about θ. For the numeracy test that is 2,048 patterns into 12
scores. In bits, the patterns carry `r fmt(ladr$H_X[ladr$table == main])` bits
of entropy, the score `r fmt(ladr$H_S[ladr$table == main])`, and both carry the
same `r fmt(ladr$I_S[ladr$table == main], 3)` bits about the person. Everything
else the patterns say is noise as far as θ is concerned.

### Same score, different patterns

```{r}
#| label: suff-same-score
#| fig-height: 5.2
#| fig-width: 7.5

ss <- suff[[main]]$same_score
bs <- suff[[main]]$by_s
k_show <- sort(unique(ss$s))[1]
n_show <- 15
items_lab <- s$items$label
top <- ss |> filter(s == k_show, rank <= n_show)
cells <- bind_rows(lapply(seq_len(nrow(top)), function(i)
  tibble(rank = top$rank[i], item = seq_along(items_lab),
         x = as.integer(strsplit(top$pattern[i], "")[[1]]))))
n_k <- choose(length(items_lab), k_show)
info_k <- bs |> filter(s == k_show) |> select(model, H_X_given_s, I_X_given_s)
p_cells <- ggplot(cells, aes(item, rank, fill = factor(x))) +
  geom_tile(colour = "white", linewidth = 0.6) +
  scale_fill_manual(values = c("0" = "#e8ecf2", "1" = irw_blue), labels = c("incorrect", "correct"), name = NULL) +
  scale_y_reverse(breaks = 1:n_show) +
  scale_x_continuous(breaks = seq_along(items_lab), labels = items_lab, expand = c(0, 0)) +
  labs(x = NULL, y = "pattern rank") +
  theme(panel.grid = element_blank(), axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
        legend.position = "top")
p_prob <- ggplot(top, aes(p_given_s, rank)) +
  geom_col(orientation = "y", fill = "grey55", width = 0.75) +
  scale_y_reverse() +
  scale_x_continuous(n.breaks = 2) +
  labs(x = "P(pattern | score)", y = NULL) +
  theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank())
(p_cells | p_prob) + plot_layout(widths = c(3, 1)) +
  plot_annotation(
    title = wrap(w = 72, sprintf("%s patterns share a score of %d; telling them apart is %s bits of entropy and %s bits about θ",
                    format(n_k, big.mark = ","), k_show, fmt(info_k$H_X_given_s[info_k$model == "Rasch"], 1),
                    fmt(abs(info_k$I_X_given_s[info_k$model == "Rasch"]), 0))),
    subtitle = wrap(sprintf("Numeracy test, Rasch. The %d most likely of the %s patterns with %d correct. Under the 2PL the same patterns carry %s bits about θ.",
                       n_show, format(n_k, big.mark = ","), k_show, fmt(info_k$I_X_given_s[info_k$model == "2PL"], 3)), 83))
```

Many patterns share each score: under Rasch, H(X | S = `r k_show`) is
`r fmt(info_k$H_X_given_s[info_k$model == "Rasch"], 2)` bits. Yet
I(θ; X | S = `r k_show`) is zero, because the probability of each pattern
*given its score* does not depend on θ at all. Two people with the same score
but very different θ produce these patterns in exactly the same proportions.
This θ-free conditional distribution is what conditional maximum likelihood
estimates item parameters from.

### Where misfit lives

```{r}
#| label: suff-misfit
#| fig-height: 5
#| fig-width: 7.5

wsh <- suff[[main]]$within_score_hist
rsp <- suff[[main]]$respondents
show_s <- sort(unique(rsp$sum_score))
hh <- wsh |> filter(s %in% show_s) |> mutate(mid = (lo + hi) / 2, s_lab = paste("score", s))
rsp <- rsp |> mutate(s_lab = paste("score", sum_score),
                     who = if_else(kind == "aberrant", "aberrant respondent", "typical respondents"))
ab_w <- filter(rsp, kind == "aberrant")
ggplot(hh, aes(mid, mass)) +
  geom_col(fill = "grey65", width = 0.24) +
  geom_vline(data = rsp, aes(xintercept = within_surprisal, colour = who), linewidth = 1) +
  facet_wrap(~ factor(s_lab, paste("score", show_s)), ncol = 1, strip.position = "left", scales = "free_y") +
  scale_colour_manual(values = c("aberrant respondent" = irw_red, "typical respondents" = "black"), name = NULL) +
  labs(x = "within-score surprisal, −log₂ P(pattern | score), bits (Rasch; no θ involved)", y = NULL,
       title = sprintf("Only %s of the %s patterns with a score of %d %s stranger than the aberrant respondent's",
                       ab_w$n_patterns_stranger, format(ab_w$n_patterns_at_score, big.mark = ","), ab_w$sum_score,
                       if (ab_w$n_patterns_stranger == 1) "is" else "are"),
       subtitle = sprintf("Numeracy test. Grey: how surprising patterns are among everyone with that score. Lines: the six replayed respondents. The aberrant pattern scores %s bits.",
                          fmt(ab_w$within_surprisal, 1))) +
  theme(legend.position = "top", strip.text.y.left = element_text(angle = 0), axis.text.y = element_blank(),
        panel.grid.major.y = element_blank())
```

The θ-free part of the data is where fit lives. A pattern can be very
surprising *given its score* while saying nothing about the person's level. This
is a different surprisal from the one in the [Surprise](#surprise) section,
which asked how unexpected each answer was given what we believed about θ.
Here θ never enters. The question is only whether this arrangement of correct
answers is a plausible one for someone with this many correct answers. The
aberrant respondent from the replays is, by this measure, among the least
plausible patterns at their score: right on hard items, wrong on easy ones.

### The Fisher view

```{r}
#| label: suff-fisher
#| fig-height: 4.4
#| fig-width: 7.5

fi <- suff[[main]]$fisher |> filter(abs(theta) <= 4)
f2 <- filter(fi, model == "2PL") |>
  pivot_longer(c(test_info, sum_score_info), names_to = "what", values_to = "info") |>
  mutate(what = factor(what, c("test_info", "sum_score_info"),
                       c("full answer pattern (test information)", "sum score only")))
fr <- filter(fi, model == "Rasch") |>
  pivot_longer(c(test_info, sum_score_info), names_to = "what", values_to = "info")
loss_peak <- f2 |> pivot_wider(names_from = what, values_from = info) |>
  mutate(loss = 1 - `sum score only` / `full answer pattern (test information)`)
main_p <- ggplot(f2, aes(theta, info, colour = what, linetype = what)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c(irw_blue, irw_red), name = NULL) +
  scale_linetype_manual(values = c("solid", "dashed"), name = NULL) +
  labs(x = "θ", y = "Fisher information",
       title = sprintf("Under the 2PL the sum score loses up to %s of the Fisher information at some θ",
                       scales::percent(max(loss_peak$loss), accuracy = 1)),
       subtitle = "Numeracy test. Inset: under Rasch the two curves are the same curve.") +
  theme(legend.position = "top")
inset <- ggplot(fr, aes(theta, info, colour = what, linetype = what)) +
  geom_line(linewidth = 0.8, show.legend = FALSE) +
  scale_colour_manual(values = c(irw_blue, irw_red)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = NULL, y = NULL, title = "Rasch") +
  theme_minimal(base_size = 9) + theme(plot.title = element_text(size = 9, face = "bold"),
                                       axis.text.y = element_blank(), panel.grid.minor = element_blank())
wl(main_p) + inset_element(inset, left = 0.68, bottom = 0.55, right = 0.99, top = 0.98)
```

Two classical facts say the same thing in other units. A sufficient statistic
preserves the Kullback–Leibler divergence between the response distributions of
any two values of θ, so any two people are exactly as distinguishable from their
score as from their full patterns. And it preserves Fisher information at every
θ, so the score's test information function equals the full data's. Under
the 2PL the plain sum score loses a little of both, and the slope-weighted
score keeps both [@cover2006elements].

## From groups to continuous θ

```{r}
#| label: bins
#| fig-height: 3.8
#| fig-width: 7.5

bc <- bind_rows(lapply(names(showcases), function(nm) {
  b <- showcases[[nm]]$bin_convergence
  cont <- b$bits[!is.finite(b$K)]
  tibble(test = show_titles[nm], K = b$K[is.finite(b$K)], bits = b$bits[is.finite(b$K)], cont = cont)
}))
ggplot(bc, aes(K, bits, colour = test)) +
  geom_hline(aes(yintercept = cont, colour = test), linetype = "dashed", alpha = 0.6) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_x_continuous(breaks = 1:8, labels = function(k) paste0(k, "\n(", 2^k, ")")) +
  scale_colour_manual(values = c(irw_blue, irw_accent, "black"), name = NULL) +
  labs(x = "K: number of bits needed to name a group (number of groups)", y = "I(group; X), bits",
       title = "By 64 groups, counting groups captures essentially all the information about θ",
       subtitle = "Dashed: information about continuous θ. Binned information never exceeds it.") +
  theme(legend.position = "top")
```

The game above asks "which of 64 groups?". The warehouse numbers ask about θ
itself, a continuous quantity. These are close: information about the group
can never exceed information about θ, since knowing θ tells you the group, and
it converges to it as the groups get finer. With 2 groups the numeracy test
gives `r fmt(showcases[[main]]$bin_convergence$bits[1])` bits. With 64 it gives
`r fmt(showcases[[main]]$bin_convergence$bits[6], 3)`, against
`r fmt(tail(showcases[[main]]$bin_convergence$bits, 1), 3)` for continuous θ.
Because the tests deliver only one or two bits, six bits of groups is far
more resolution than any of them can use.

## Limitations

- **Model dependence.** Every number is computed from a fitted model, not from
  the data directly. If a 2PL is wrong for a table (guessing, multiple
  dimensions, local dependence), the bits are wrong too. The sum-score check
  gives one warning sign: model-implied and observed score distributions
  diverge by more than 0.1 bit (KL) in `r sum(two$sum_kl_bits > 0.1, na.rm = TRUE)` tables.
- **Local dependence and sharp items.** Even with a slope prior,
  `r n_distinct(items_df$table[items_df$a > 5])` tables have at least one item
  with a slope above 5, often near-duplicate items. Such pairs make the model
  treat a repeated answer as two independent pieces of evidence, which
  inflates the bits.
- **Unidimensionality.** We fit one trait per table. A multidimensional test
  carries information about several traits, and the numbers here count only
  what projects onto the first.
- **The population matters.** Information is relative to the spread of
  people taking the test. The same items given to a more homogeneous group
  would deliver fewer bits, because there is less uncertainty to resolve to
  begin with.
- **Subsampling and missing data.** Large tables were subsampled to 10,000
  respondents; that affects parameter precision, not the definition of the
  quantity. Items answered by fewer than half the respondents were dropped,
  and I(θ; X) treats every respondent as answering every remaining item.
- **Convergence.** `r sum(!two$converged)` 2PL fit
  (`r paste(two$table[!two$converged], collapse = ", ")`) did not meet `mirt`'s
  convergence criterion; its results are included but should be read with care.
- **Enumeration in the sufficiency section.** Its quantities enumerate every
  answer pattern on a θ grid, so they are limited to 16 items. For the
  fraction-subtraction test they use the 12 items with the highest expected
  bits. The within-score surprisal needs no θ grid and uses all items.
- **Scope.** Only dichotomous items, and no ENEM tables in this version.
  Polytomous items can carry more than one bit each.

## Reproducibility

{{< include ../components/_source-links.qmd >}}

{{< include ../components/_data-version.qmd >}}

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 0. Optional: re-scan IRW for eligible tables (writes scout_tables.csv)
source("vignettes/bits_per_item_scout.R")

# 1. Validation checks, then the compute script (from the project root)
source("vignettes/bits_per_item_checks.R")
source("vignettes/bits_per_item_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/bits_per_item.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(unique(summary_df$table), output_file = "bits_per_item_data/references.bib")
```
:::

Visual inspiration: 3Blue1Brown, ["Solving Wordle using information
theory"](https://www.youtube.com/watch?v=v68zYyaEmEA). No material from the
video is reused here.

---

## References

::: {#refs}
:::
