---
title: "Bits per Item: How Much Does a Test Tell You About a Person?"
description: |
  A test is a sequence of questions asked about someone we cannot see. Each
  answer narrows down where that person sits on the trait being measured.
  Information theory lets us count that narrowing in bits, the same unit that
  tells you how much a Wordle guess cuts down the list of possible words. We
  play the game on three real tests, then measure the bits delivered by more
  than two hundred dichotomous tests in the Item Response Warehouse.
date: today
bibliography: bits_per_item_data/references.bib
nocite: |
  @*
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)
library(patchwork)

theme_set(theme_minimal(base_size = 14) +
            theme(plot.title = element_text(face = "bold", size = 14),
                  plot.title.position = "plot",
                  panel.grid.minor = element_blank()))
irw_blue   <- "#2166ac"   # the population of theta groups
irw_red    <- "#d73027"   # surprise / the aberrant respondent
irw_grey   <- "grey50"
irw_accent <- "#1b9e77"   # information gained, everywhere on this page

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "bits_per_item_data/bits_per_item_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  summary_df$construct_type <- sub(",.*$", "", summary_df$construct_type)
  items_df         <- .cache$items
  curves_df        <- .cache$curves
  showcases        <- .cache$showcases
  candidate_tables <- .cache$candidate_tables
  date_run         <- .cache$date_run
  is_pilot         <- isTRUE(.cache$pilot)
  failures         <- unique(sub("^\\S+ \\S+ ", "", .cache$failures))
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_tables_run   <- length(.cache$tables_attempted %||% candidate_tables)
} else {
  summary_df <- tibble(table = character(), model = character(), I_X = numeric(),
                       I_S = numeric(), rho_marginal = numeric(), n_items = integer())
  items_df <- tibble(); curves_df <- tibble(); showcases <- list(); failures <- character()
  candidate_tables <- character()
  date_run         <- Sys.Date()
  is_pilot         <- FALSE
  r_all_candidates <- 0L
  r_n_tables_run   <- 0L
}

sim_path <- "bits_per_item_data/reliability_sim_results.rds"
rel_sim <- if (file.exists(sim_path)) readRDS(sim_path)$sim else NULL
knitr::opts_chunk$set(eval = cache_ok)
# Every ggplot printed on this page goes through wl().
registerS3method("knit_print", "ggplot", function(x, ...) print(wl(x)), envir = asNamespace("knitr"))

h2 <- function(p) { p <- pmin(pmax(p, 1e-12), 1 - 1e-12); -(p * log2(p)) - ((1 - p) * log2(1 - p)) }
show_titles <- c(fitz_2024_numeracy       = "Numeracy (11 items)",
                 machivallianism_test_vcl = "Vocabulary checklist (16 words)",
                 frac20                   = "Fraction subtraction (20 items)")
main <- "fitz_2024_numeracy"
two  <- filter(summary_df, model == "2PL")
fmt  <- function(x, d = 2) formatC(x, format = "f", digits = d)
wrap <- function(x, w = 95) if (is.null(x)) x else stringr::str_wrap(x, w)
# Wrap long titles, subtitles and axis labels so they fit the figure width.
wl <- function(p, w = 72) {
  for (nm in c("title", "subtitle", "x")) {
    lab <- p$labels[[nm]]
    if (is.character(lab)) p$labels[[nm]] <- wrap(lab, if (nm == "title") w else round(w * 1.15))
  }
  p
}
```

```{r}
#| label: cache-warning
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/bits_per_item_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, partial run.** This render uses a partial sample of ",
  r_n_tables_run, " tables out of ", r_all_candidates, " eligible tables. ",
  "All numbers here are real computed results, but treat the distributions ",
  "as a small, noisy preview. Re-run `bits_per_item_compute.R` without ",
  "`PILOT=1` to scale up.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`bits_per_item_compute.R`) directly if you're relying
on the numbers here.
:::

```{r}
#| label: ojs-data

ojs_payload <- lapply(names(showcases), function(nm) {
  s <- showcases[[nm]]
  list(
    key = nm, title = unname(show_titles[nm]), instrument = s$instrument,
    items = s$items[, c("label", "text", "label0", "label1", "a", "b", "bits")],
    pg = unname(round(s$group_probs, 5)),
    # group edges on the theta scale; the two open-ended tail groups are drawn to +/- 3.2
    bounds = pmin(pmax(s$group_bounds, -3.2), 3.2)
  )
})
names(ojs_payload) <- names(showcases)
ojs_define(showcases_r = ojs_payload, accent_r = irw_accent, blue_r = irw_blue)
```

## Cutting up the population

Start with an imaginary test whose questions are perfect. People are spread
along θ, the trait being measured, in the familiar bell shape. Each question
is a **cut**: everyone above it answers right, everyone below answers wrong.
Asking it tells you which side of the cut a person is on, and nothing else.

**Click on the chart to place a cut** (or use the slider), and watch the bits.
With perfect cuts the arithmetic is simple: the cuts chop the population into
groups, and the bits are the entropy of the groups' shares,
−Σ p log~2~ p. One cut at the median makes two halves: 1 bit (each half holds
p = ½ of people, so −½ log~2~ ½ − ½ log~2~ ½ = ½ + ½ = 1). Three cuts at
the quartiles make four quarters: 2 bits (4 × −¼ log~2~ ¼ = 4 × ½ = 2). In general k cuts are worth at most
log~2~(k + 1) bits, reached only when they make equal groups.

Then turn up the **blur**. A blurred cut is a real question: near the cut,
people sometimes land on the wrong side, so the answer is partly luck. The
luck is noise, and it comes off the top.

```{ojs}
cx_pnorm = x => {
  const t = 1 / (1 + 0.3275911 * Math.abs(x) / Math.SQRT2);
  const y = 1 - (((((1.061405429 * t - 1.453152027) * t) + 1.421413741) * t - 0.284496736) * t + 0.254829592) * t * Math.exp(-x * x / 2);
  return x >= 0 ? (1 + y) / 2 : (1 - y) / 2;
}
cx_qnorm = p => {   // Acklam's approximation
  const a = [-3.969683028665376e+01, 2.209460984245205e+02, -2.759285104469687e+02, 1.383577518672690e+02, -3.066479806614716e+01, 2.506628277459239e+00];
  const b = [-5.447609879822406e+01, 1.615858368580409e+02, -1.556989798598866e+02, 6.680131188771972e+01, -1.328068155288572e+01];
  const c = [-7.784894002430293e-03, -3.223964580411365e-01, -2.400758277161838e+00, -2.549732539343734e+00, 4.374664141464968e+00, 2.938163982698783e+00];
  const d = [7.784695709041462e-03, 3.224671290700398e-01, 2.445134137142996e+00, 3.754408661907416e+00];
  const lo = 0.02425;
  if (p < lo) { const q = Math.sqrt(-2 * Math.log(p));
    return (((((c[0]*q+c[1])*q+c[2])*q+c[3])*q+c[4])*q+c[5]) / ((((d[0]*q+d[1])*q+d[2])*q+d[3])*q+1); }
  if (p > 1 - lo) { const q = Math.sqrt(-2 * Math.log(1 - p));
    return -(((((c[0]*q+c[1])*q+c[2])*q+c[3])*q+c[4])*q+c[5]) / ((((d[0]*q+d[1])*q+d[2])*q+d[3])*q+1); }
  const q = p - 0.5, r = q * q;
  return (((((a[0]*r+a[1])*r+a[2])*r+a[3])*r+a[4])*r+a[5])*q / (((((b[0]*r+b[1])*r+b[2])*r+b[3])*r+b[4])*r+1);
}
cx_M = 400
cx_nodes = Array.from({length: cx_M}, (_, i) => cx_qnorm((i + 0.5) / cx_M))
cx_h2 = p => (p <= 0 || p >= 1) ? 0 : -(p * Math.log2(p)) - (1 - p) * Math.log2(1 - p)
cx_ent = ps => -ps.reduce((s, p) => s + (p > 0 ? p * Math.log2(p) : 0), 0)
cx_shares = cuts => {
  const c = [...cuts].sort((x, y) => x - y);
  const edges = [0, ...c.map(cx_pnorm), 1];
  return edges.slice(1).map((e, i) => e - edges[i]);
}
// Exact I(theta; X) for k cuts with logistic blur s (s = 0: perfect cuts).
cx_info = (cuts, s) => {
  const k = cuts.length;
  if (k === 0) return {I: 0, H: 0, N: 0, live: 1};
  if (s === 0) {
    const sh = cx_shares(cuts), H = cx_ent(sh);
    return {I: H, H, N: 0, live: sh.filter(p => p > 1e-9).length};
  }
  const n = 1 << k, px = new Float64Array(n), d = new Float64Array(n);
  let noise = 0;
  for (const t of cx_nodes) {
    d.fill(0); d[0] = 1;
    for (let j = 0; j < k; j++) {
      const p = 1 / (1 + Math.exp(-(t - cuts[j]) / s)), step = 1 << j;
      noise += cx_h2(p);
      for (let i = 0; i < step; i++) { d[i + step] = d[i] * p; d[i] *= 1 - p; }
    }
    for (let i = 0; i < n; i++) px[i] += d[i] / cx_M;
  }
  noise /= cx_M;
  const H = cx_ent(Array.from(px));
  return {I: H - noise, H, N: noise, live: Array.from(px).filter(p => p > 0.001).length};
}
cx_max = 12
mutable cx_cuts = []
cx_add = t => { if (cx_cuts.length < cx_max) mutable cx_cuts = [...cx_cuts, Math.round(t * 100) / 100]; }
cx_numeracy = { const it = showcases_r["fitz_2024_numeracy"].items;
  const a = [...it.a].sort((x, y) => x - y), med = a[Math.floor(a.length / 2)];
  return {cuts: it.b.map(b => Math.round(b * 100) / 100).slice(0, cx_max), blur: Math.round(20 / med) / 20}; }
cx_setblur = v => { viewof cx_blur.value = v; viewof cx_blur.dispatchEvent(new Event("input", {bubbles: true})); }
```

```{ojs}
viewof cx_blur = Inputs.range([0, 1.5], {step: 0.05, value: 0, label: "Blur (0 = perfect cuts)"})
viewof cx_place = Inputs.range([-3, 3], {step: 0.05, value: 0, label: "New cut at θ ="})
viewof cx_b1 = Inputs.button([
  ["Add cut", () => cx_add(cx_place)],
  ["Undo", () => { mutable cx_cuts = cx_cuts.slice(0, -1); }],
  ["Clear", () => { mutable cx_cuts = []; }]
])
viewof cx_b2 = Inputs.button([
  ["Median", () => { mutable cx_cuts = [0]; }],
  ["Quartiles", () => { mutable cx_cuts = [0, -0.67, 0.67]; }],
  ["Octiles", () => { mutable cx_cuts = [0, -0.67, 0.67, -1.15, -0.32, 0.32, 1.15]; }],
  ["3 near the median", () => { mutable cx_cuts = [0, -0.1, 0.1]; }],
  ["3 in the low tail", () => { mutable cx_cuts = [-1.5, -1.8, -2.1]; }],
  ["Numeracy test", () => { mutable cx_cuts = cx_numeracy.cuts; cx_setblur(cx_numeracy.blur); }]
], {label: "Try"})
```

```{ojs}
cx_res = cx_info(cx_cuts, cx_blur)
cx_rows = cx_cuts.map((c, i) => {
  const before = i === 0 ? 0 : cx_info(cx_cuts.slice(0, i), cx_blur).I;
  const after = cx_info(cx_cuts.slice(0, i + 1), cx_blur).I;
  const prev = cx_cuts.slice(0, i), pc = cx_pnorm(c);
  const lo = Math.max(0, ...prev.map(cx_pnorm).filter(q => q <= pc)), hi = Math.min(1, ...prev.map(cx_pnorm).filter(q => q > pc));
  return {"#": i + 1, "θ": c.toFixed(2),
    "group it splits": `${((hi - lo) * 100).toFixed(0)}% → ${((pc - lo) * 100).toFixed(0)}% + ${((hi - pc) * 100).toFixed(0)}%`,
    "bits added": (after - before).toFixed(3), "on its own": cx_info([c], cx_blur).I.toFixed(3)};
})
cx_plot = {
  const W = Math.min(width, 680), mL = 20, mR = 20, dom = [-3.2, 3.2];
  const xs = d3.range(dom[0], dom[1] + 1e-9, 0.02);
  const dn = x => Math.exp(-x * x / 2) / Math.sqrt(2 * Math.PI);
  const sorted = [...cx_cuts].sort((x, y) => x - y), edges = [dom[0], ...sorted, dom[1]];
  const groups = edges.slice(1).map((e, i) => ({x1: edges[i], x2: e, g: i,
    share: cx_pnorm(e === dom[1] ? 9 : e) - cx_pnorm(edges[i] === dom[0] ? -9 : edges[i])}));
  const area = groups.flatMap(gr => xs.filter(x => x >= gr.x1 && x <= gr.x2).concat([gr.x2])
    .map(x => ({x, y: dn(x), g: gr.g})));
  const s = cx_blur;
  const plt = Plot.plot({
    width: W, height: 230, marginLeft: mL, marginRight: mR, marginTop: 26, marginBottom: 40,
    x: {domain: dom, label: "θ: the trait being measured (low → high)", labelAnchor: "center", labelArrow: "none"},
    y: {domain: [0, 0.46], axis: null},
    marks: [
      Plot.areaY(area, {x: "x", y: "y", z: "g", fill: d => d.g % 2 ? "#92c5de" : "#d1e5f0", curve: "linear"}),
      s > 0 ? Plot.rect(cx_cuts, {x1: c => c - 2.2 * s, x2: c => c + 2.2 * s, y1: 0, y2: 0.42, fill: "#d73027", fillOpacity: 0.08}) : null,
      Plot.ruleX(cx_cuts, {x: c => c, y1: 0, y2: 0.42, stroke: "#d73027", strokeWidth: 1.8}),
      Plot.text(cx_cuts.map((c, i) => ({c, i})), {x: "c", y: 0.42, text: d => String(d.i + 1), dy: -8, fill: "#d73027", fontWeight: "bold"}),
      Plot.text(groups.filter(gr => gr.share > 0.04), {x: gr => (Math.max(gr.x1, -2.6) + Math.min(gr.x2, 2.6)) / 2, y: 0.03,
        text: gr => (gr.share * 100).toFixed(0) + "%", fill: "#333", fontSize: 11}),
      Plot.ruleY([0])
    ]
  });
  plt.style.cursor = "crosshair";
  plt.addEventListener("click", e => {
    const [px] = d3.pointer(e, plt);
    const t = dom[0] + (px - mL) / (W - mL - mR) * (dom[1] - dom[0]);
    if (t > -3.1 && t < 3.1) cx_add(t);
  });
  return plt;
}
cx_ceiling = Math.log2(cx_cuts.length + 1)
cx_readout = html`<div style="font-size:0.95rem; max-width:680px;">
  <div style="display:flex; justify-content:space-between; flex-wrap:wrap; gap:0.3rem 1rem;">
    <span><b style="color:${accent_r}; font-size:1.15rem;">${cx_res.I.toFixed(2)} bits</b> from ${cx_cuts.length} cut${cx_cuts.length === 1 ? "" : "s"}</span>
    <span>ceiling for ${cx_cuts.length} cut${cx_cuts.length === 1 ? "" : "s"}: log₂(${cx_cuts.length + 1}) = ${cx_ceiling.toFixed(2)} bits</span></div>
  <div style="height:14px; background:#e9ecef; border-radius:3px; overflow:hidden; margin:4px 0; position:relative;">
    <div style="height:100%; width:${Math.min(100, cx_res.I / Math.log2(cx_max + 1) * 100)}%; background:${accent_r};"></div>
    <div style="position:absolute; top:0; bottom:0; left:${cx_ceiling / Math.log2(cx_max + 1) * 100}%; border-left:2px dashed #333;"></div></div>
  <div style="color:#555; font-size:0.88rem;">Spread H(X) ${cx_res.H.toFixed(2)} bits − noise H(X | θ) ${cx_res.N.toFixed(2)} bits.
  As certain as knowing which of <b>${Math.pow(2, cx_res.I).toFixed(1)}</b> equal groups the person is in.
  Answer patterns that actually occur: <b>${cx_res.live}</b> of ${Math.pow(2, cx_cuts.length).toLocaleString()} possible${cx_blur > 0 ? " (with a chance above 0.1%)" : ""}.</div>
  <div style="color:#888; font-size:0.8rem; margin-top:2px;">Bar: bits so far, on a scale to log₂(${cx_max + 1}), the most ${cx_max} cuts can give. Dashed: the ceiling for this many cuts.${cx_blur > 0 ? " Pale red bands: where a blurred cut is uncertain (from a 10% to a 90% chance of answering right)." : ""}</div></div>`
html`<div>${cx_plot}${cx_readout}</div>`
```

```{ojs}
cx_rows.length ? Inputs.table(cx_rows, {rows: 6, layout: "fixed",
  width: {"#": 35, "θ": 60, "bits added": 90, "on its own": 90}}) :
  html`<p style="color:#666;"><i>No cuts yet. Click the chart.</i></p>`
```

Some things to try:

- **Median, then add another cut at 0.** The second cut adds nothing: it asks
  a question that has already been answered. Redundancy is the first way bits
  are lost.
- **Quartiles.** Each cut's "bits added" depends on what was there before. The
  first cut is worth a full bit; the next two split halves, and are worth half a
  bit each, because each only matters for the half of the population it falls in.
- **3 near the median.** Cuts at −0.1, 0 and +0.1. The outer two each split
  off a sliver of about 4% of people, so together the three are worth 1.40
  bits, well short of the 2 that three spread-out cuts can give. A cut close to
  an existing one asks nearly the same question.
- **3 in the low tail.** Each cut splits off a sliver. Cuts where few people
  are say little about most people.
- **Quartiles, then blur.** The bits fall below 2 even though the cuts are
  perfectly placed, and patterns that perfect cuts cannot produce (a right answer
  above a wrong one) start to occur. That is the noise.
- **Numeracy test.** Cuts at the difficulty of each of its items, blurred by
  its typical slope. They pile up between about −1 and −0.5, and blurred, the
  whole test is worth little more than one well-placed perfect question. The
  real test, with each item's own slope, carries `r fmt(two$I_X[two$table == main])` bits.


## Answer patterns

Everything on this page comes from one object: the **answer pattern**, the
string of right and wrong answers a person gives. List every pattern a test
could produce and two quantities decide how much the test tells you about a
person:

- **Spread**, H(X): how varied the patterns are across the population. If
  everyone gave the same pattern, seeing it would tell you nothing.
- **Noise**, H(X | θ): how varied they would be for one person, answering again
  and again. Some of the spread is just luck, not differences between people.

The information about the person is what is left: **I(θ; X) = H(X) − H(X | θ)**,
in bits. The cuts above were the special case with no noise. Now we build
this up on a real numeracy test: one question first, then three, five, and the
whole test.

### One question

With one question there are only two patterns, right and wrong. To picture the
population we cut θ, the trait being measured, into **64 slices, each holding
the same share of people**: narrow in the crowded middle, wide in the thin
tails. Pinning down one slice out of 64 would take 6 bits, since 2^6^ = 64.

```{r}
#| label: patterns-setup

s <- showcases[[main]]
top5 <- s$items |> arrange(desc(bits)) |> slice(1:5)
gth <- s$grid                                    # the 2PL's N(0, 1) quadrature grid
# P(correct | theta) for a set of items: length(theta) x items
p_items <- function(it, theta) plogis(outer(theta, it$a) - matrix(it$a * it$b, length(theta), nrow(it), byrow = TRUE))
# P(pattern | theta) for 0/1 pattern rows: patterns x theta
pat_given_theta <- function(X, P) exp(X %*% t(log(P)) + (1 - X) %*% t(log(1 - P)))
```

```{r}
#| label: bit-strip
#| fig-height: 5.2
#| fig-width: 7.5

pg <- s$group_probs
pick3 <- s$items |> arrange(desc(bits)) |> slice(c(1, ceiling(n() / 2), n()))
bnd <- pmin(pmax(s$group_bounds, -3.2), 3.2)
cells <- tibble(g = 1:64, x1 = bnd[-65], x2 = bnd[-1])
ideal <- mutate(cells, row = "A perfect (Guttman) item at the median",
                p = rep(c(0, 1), each = 32), bits = 1)
real <- bind_rows(lapply(seq_len(nrow(pick3)), function(i) {
  it <- pick3[i, ]
  mutate(cells, row = paste0("\u201c", substr(it$text, 1, 52), if (nchar(it$text) > 52) "\u2026" else "", "\u201d"),
         p = pg[, it$item], bits = it$bits)
}))
strip <- bind_rows(ideal, real) |>
  mutate(row = factor(row, levels = unique(row)), y = 5 - as.integer(row),
         lab = paste0(fmt(bits), if_else(bits == 1, " bit", " bits")))
labs_df <- distinct(strip, row, y, lab)
ggplot(strip) +
  geom_rect(aes(xmin = x1, xmax = x2, ymin = y - 0.4, ymax = y + 0.4, fill = p),
            colour = "white", linewidth = 0.15) +
  geom_text(data = labs_df, aes(x = 3.35, y = y, label = lab),
            hjust = 0, colour = irw_accent, fontface = "bold", size = 4) +
  scale_fill_gradient(low = "#f4f6fa", high = irw_blue, limits = c(0, 1),
                      name = "P(yes / correct)\nin each slice") +
  scale_x_continuous(limits = c(-3.2, 4.3), breaks = -3:3, expand = c(0, 0)) +
  scale_y_continuous(breaks = labs_df$y, labels = stringr::str_wrap(as.character(labs_df$row), 34)) +
  labs(x = "θ, cut into 64 equally likely slices", y = NULL,
       title = sprintf("A perfect item halves the population; the best real numeracy item manages %s of a bit",
                       fmt(max(s$items$bits))),
       subtitle = "Each row is one question, shaded by how likely people in each slice are to say yes / get it right. Right: I(θ; Xⱼ), the item on its own.") +
  theme(panel.grid = element_blank(), legend.position = "bottom", legend.key.width = unit(1.2, "cm"))
```

A bit is the information in the answer to a question whose two answers are
equally likely and which splits the possibilities cleanly. The top row of the
figure does exactly that: it cuts 64 slices to 32. Real items fail in two ways.
Their split is blurred, because people in the same group sometimes answer
differently. And their split is often off-centre, because easy items mostly
separate the lowest groups from everyone else. In pattern terms, the
spread is the entropy of the answer across the population, the noise is the
entropy of the answer for someone at a fixed θ, and the information is the gap:
<small>I(θ; X~j~) = h(P̄~j~) − E~θ~[h(P~j~(θ))], with h(p) = −p log~2~ p − (1−p) log~2~(1−p).</small>
Neither term can exceed one bit.

### Three questions

```{r}
#| label: patterns-three
#| fig-height: 6.4
#| fig-width: 7.5

top3 <- top5[1:3, ]
th3 <- seq(-3, 3, by = 0.02)
X3 <- as.matrix(expand.grid(rep(list(0:1), 3)))
X3 <- X3[order(rowSums(X3), -X3 %*% c(4, 2, 1)), ]
key3 <- apply(X3, 1, paste, collapse = "")
pxt3 <- pat_given_theta(X3, p_items(top3, th3))           # 8 x theta
area <- tibble(pattern = rep(key3, times = length(th3)), theta = rep(th3, each = 8),
               p = as.vector(pxt3), score = rep(rowSums(X3), times = length(th3))) |>
  group_by(theta) |> mutate(ymax = cumsum(p), ymin = ymax - p) |> ungroup()   # 000 at the bottom
labs3 <- area |> filter(abs(theta) <= 2.6) |> group_by(pattern) |>
  slice_max(p, n = 1, with_ties = FALSE) |> ungroup() |> mutate(mid = (ymin + ymax) / 2)
w3 <- gth$w
marg3 <- as.vector(pat_given_theta(X3, p_items(top3, gth$theta)) %*% w3)
H3 <- -sum(marg3 * log2(marg3))
pxt3g <- pat_given_theta(X3, p_items(top3, gth$theta))
N3 <- -sum((pxt3g * log2(pxt3g)) %*% w3)
score_fill <- c(`0` = "#d1e5f0", `1` = "#92c5de", `2` = "#4393c3", `3` = irw_blue)
p_top <- ggplot(area, aes(theta, fill = factor(score), group = pattern)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), colour = "white", linewidth = 0.4) +
  geom_text(data = labs3, aes(theta, mid, label = pattern), inherit.aes = FALSE,
            family = "mono", size = 3.6, colour = if_else(labs3$score >= 2, "white", "grey15")) +
  scale_fill_manual(values = score_fill, name = "number correct") +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(x = "θ", y = "chance of each pattern",
       title = sprintf("Three questions: the mix of patterns shifts with θ, and that shift is %s bits", fmt(H3 - N3)),
       subtitle = wrap(sprintf("Top: read straight up from any θ to see the patterns a person there would give, answering again and again. Bottom: the same mix averaged over the population. Items, in pattern order: %s.",
                          paste(top3$label, collapse = ", ")), 78)) +
  theme(legend.position = "top")
bars3 <- tibble(pattern = factor(key3, key3), p = marg3, score = factor(rowSums(X3)))
p_bot <- ggplot(bars3, aes(pattern, p, fill = score)) +
  geom_col(width = 0.75, show.legend = FALSE) +
  geom_text(aes(label = fmt(p, 2)), vjust = -0.4, size = 3.3) +
  scale_fill_manual(values = score_fill) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "pattern (1 = correct)", y = "population share") +
  theme(axis.text.x = element_text(family = "mono"), panel.grid.major.x = element_blank())
p_top / p_bot + plot_layout(heights = c(2.2, 1))
```

Read the top panel straight up from any point on θ. Far to the left almost
everyone gives `r key3[1]`; far to the right almost everyone gives
`r key3[8]`. In between, a person at a fixed θ could give several patterns,
and which one comes out is partly luck. That luck is the noise, H(X | θ) =
`r fmt(N3)` bits on average. The bottom panel is what you see when you do not
know θ: the mix averaged over everyone, whose entropy, the spread, is
H(X) = `r fmt(H3)` bits. The information is the gap, `r fmt(H3 - N3)` bits,
and it comes entirely from the mix *changing* along θ. If every θ produced the
same mix, the top panel would be flat bands, the spread and the noise would be
equal, and seeing a pattern would tell you nothing about the person.

### Five questions

```{r}
#| label: patterns
#| fig-height: 4.6
#| fig-width: 7.5

pt <- s$patterns |> mutate(rank = row_number())
P5 <- p_items(top5, gth$theta)
X5 <- do.call(rbind, lapply(strsplit(pt$pattern, ""), as.integer))
pxt5 <- pat_given_theta(X5, P5)                       # patterns x theta nodes
stopifnot(max(abs(as.vector(pxt5 %*% gth$w) - pt$implied)) < 1e-8)
spread_j <- -pt$implied * log2(pt$implied)
noise_j <- -as.vector((pxt5 * log2(pxt5)) %*% gth$w)
H5 <- sum(spread_j); H5_noise <- sum(noise_j)
post5 <- sweep(pxt5, 2, gth$w, "*"); post5 <- post5 / rowSums(post5)
q_theta <- function(pr, q) gth$theta[which(cumsum(pr) >= q)[1]]
pt <- pt |> mutate(
  spread = spread_j, noise = noise_j, net = spread - noise,
  who = sapply(seq_len(n()), function(i) sprintf("θ mean %s, middle 80%% %s to %s",
          fmt(sum(post5[i, ] * gth$theta), 1), fmt(q_theta(post5[i, ], .1), 1), fmt(q_theta(post5[i, ], .9), 1))),
  answers = sapply(seq_len(n()), function(i) paste(sprintf("%s: %s", top5$label,
          if_else(X5[i, ] == 1, "correct", "incorrect")), collapse = "<br>")),
  tip = sprintf("<b>pattern %s</b><br>%s<br><br>model probability %s<br>observed %s (%s of %s)<br>who gives it: %s<br><br>spread  −p log₂ p = %s bits<br>noise = %s bits<br>net = %s bits",
                pattern, answers, fmt(implied, 3), fmt(observed, 3),
                format(round(observed * n_obs), big.mark = ","), format(n_obs, big.mark = ","),
                who, fmt(spread, 3), fmt(noise, 3), fmt(net, 3)))
g <- ggplot(pt, aes(rank, implied, text = tip)) +
  geom_col(fill = irw_blue, width = 0.8) +
  geom_point(aes(y = observed), colour = "black", size = 1.8) +
  scale_x_continuous(breaks = pt$rank, labels = pt$pattern, expand = c(0.01, 0.01)) +
  labs(x = "answer pattern on the 5 most informative items (1 = correct)", y = "probability",
       title = sprintf("Five questions, 32 possible answer patterns, %s bits about the person", fmt(H5 - H5_noise))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, family = "mono", size = 8))
g <- wl(g, 55)
ggplotly(g, tooltip = "text") |>
  layout(title = list(text = paste0("<b>", gsub("\n", "<br>", g$labels$title), "</b>"), x = 0, xanchor = "left",
                      font = list(size = 15)), margin = list(t = 70),
         hoverlabel = list(align = "left"))
```

Bars: the model's probability of each pattern. Dots: the share of the
`r format(pt$n_obs[1], big.mark = ",")` respondents who actually gave it.
Items, in pattern order: `r paste(top5$label, collapse = ", ")`. **Hover over
a bar** to see its pieces: which answers the pattern contains, who tends to
give it, and how much it contributes to the spread, to the noise, and to the
net bits. Summed over the 32 bars, the spread is `r fmt(H5)` bits and the
noise `r fmt(H5_noise)`, leaving `r fmt(H5 - H5_noise)` bits about the person.
Most of the spread is noise.

```{r}
#| label: worked-example
#| results: asis

i_ex <- 1                                                # the most likely pattern
x_ex <- X5[i_ex, ]
th_ex <- c(-1, 0, 1)
P_ex <- p_items(top5, th_ex)                             # 3 x 5
p_ans <- t(ifelse(matrix(x_ex, 3, 5, byrow = TRUE) == 1, P_ex, 1 - P_ex))   # 5 x 3
prod_ex <- apply(p_ans, 2, prod)
tab <- tibble(item = top5$label, answer = if_else(x_ex == 1, "correct", "incorrect"),
              `θ = −1` = fmt(p_ans[, 1], 3), `θ = 0` = fmt(p_ans[, 2], 3), `θ = +1` = fmt(p_ans[, 3], 3))
tab <- bind_rows(tab, tibble(item = "**product**", answer = "",
                             `θ = −1` = paste0("**", fmt(prod_ex[1], 3), "**"),
                             `θ = 0` = paste0("**", fmt(prod_ex[2], 3), "**"),
                             `θ = +1` = paste0("**", fmt(prod_ex[3], 3), "**")))
cat("::: {.callout-tip collapse=\"true\"}\n## How one bar is computed: pattern ", pt$pattern[i_ex], "\n\n", sep = "")
cat("**1. The chance of the pattern for one person.** The model gives each item a probability of a correct answer ",
    "that depends on θ. For a person at a given θ, the chance of the whole pattern is the product of the chances ",
    "of its five answers, because the model treats answers as independent once θ is fixed. Here it is for three people:\n\n", sep = "")
print(knitr::kable(tab, align = "llrrr"))
cat("\n\n**2. Average over the population.** Do the same at every θ and average the products, weighting each θ by ",
    "how many people sit there (a normal curve, on an ", nrow(gth), "-point grid). That average, ", fmt(pt$implied[i_ex], 3),
    ", is the height of the bar. ", fmt(100 * pt$observed[i_ex], 0), "% of respondents actually gave this pattern.\n\n", sep = "")
cat("**3. Spread.** Each bar adds −p log₂ p to the spread: here −", fmt(pt$implied[i_ex], 3), " × log₂ ",
    fmt(pt$implied[i_ex], 3), " = ", fmt(pt$spread[i_ex], 3), " bits. Summed over all 32 bars that is H(X) = ",
    fmt(H5), " bits.\n\n", sep = "")
cat("**4. Noise.** The same calculation *within* each θ, using the products from step 1 instead of the average, ",
    "then averaged over the population: ", fmt(pt$noise[i_ex], 3), " bits from this pattern, H(X | θ) = ",
    fmt(H5_noise), " bits in all. This is how unpredictable the pattern is even for someone whose θ we know.\n\n", sep = "")
cat("**5. Information.** Spread minus noise: I(θ; X) = ", fmt(H5), " − ", fmt(H5_noise), " = ", fmt(H5 - H5_noise),
    " bits. A single bar's net can be negative; only the total has to be positive.\n:::\n", sep = "")
```

### The whole test

The numeracy test has `r nrow(s$items)` items, so `r format(2^nrow(s$items), big.mark = ",")`
possible patterns: too many to draw, but not too many to count. The same
calculation gives a spread of
`r fmt(two$H_X[two$table == main])` bits, noise of
`r fmt(two$H_X_given_theta[two$table == main])` bits, and so
`r fmt(two$I_X[two$table == main])` bits about the person. Adding six items to
the five above bought about
`r fmt(two$I_X[two$table == main] - (H5 - H5_noise))` bits more. The sum score
gathers the `r format(2^nrow(s$items), big.mark = ",")` patterns into
`r nrow(s$items) + 1` piles, one per score, and keeps
`r fmt(two$I_S[two$table == main])` of those bits; [Is the sum score enough?](#is-the-sum-score-enough)
asks what the piling loses. For long tests the patterns cannot be listed, and
the same quantities are estimated by simulation (see
[How the warehouse numbers were computed](#how-the-warehouse-numbers-were-computed)).

## Guess the person

The patterns above are averages over everyone. Now play the test forward for
one person, an answer at a time. The curve below is the population along θ,
cut into the same 64 equally likely slices. Somewhere in it is one person, and
your job is to find out which slice they are in by asking questions from a
real test.

Before any question, all 64 slices are equally likely, so there are 6 bits to
find. After each answer the slices are
re-shaded, darker where the person is now more likely to be, and a line traces
the same thing as a density over θ. The counter shows how many bits that
answer bought.

Pick a question from the table; they are ranked by how many bits they are
*expected* to give. Then either answer it yourself, as the person might, or let
the simulated person answer. At the start we secretly draw one simulated
person from the population; they answer with the probabilities the model
gives someone in their slice. You can reveal where they are at any point.
The "best" and "worst" buttons ask the remaining question with the most or the
fewest expected bits, and the simulated person answers it.

```{ojs}
sc = showcases_r
keys = Object.keys(sc)
viewof tbl = Inputs.select(keys, {label: "Test", value: keys[0], format: k => sc[k].title})
```

```{ojs}
S = {
  const s = sc[tbl];
  return {items: transpose(s.items).map((d, j) => ({...d, j})), pg: s.pg, bounds: s.bounds, title: s.title};
}
h2 = p => (p <= 0 || p >= 1) ? 0 : -(p * Math.log2(p)) - (1 - p) * Math.log2(1 - p)
entropy = post => -post.reduce((s, p) => s + (p > 0 ? p * Math.log2(p) : 0), 0)
mutable hist = []
mutable hidden = Math.floor(Math.random() * 64)
reset_on_test = { tbl; mutable hist = []; mutable hidden = Math.floor(Math.random() * 64); return tbl; }
posterior_of = hs => {
  let p = Array(64).fill(1 / 64);
  for (const h of hs) {
    const like = S.pg.map(row => h.x === 1 ? row[h.j] : 1 - row[h.j]);
    const z = p.reduce((s, v, g) => s + v * like[g], 0);
    p = p.map((v, g) => v * like[g] / z);
  }
  return p;
}
post = { reset_on_test; return posterior_of(hist); }
exp_bits = j => h2(post.reduce((s, v, g) => s + v * S.pg[g][j], 0)) -
                post.reduce((s, v, g) => s + v * h2(S.pg[g][j]), 0)
asked = new Set(hist.map(h => h.j))
remaining = S.items.filter(d => !asked.has(d.j))
  .map(d => ({...d, expected: exp_bits(d.j)}))
  .sort((a, b) => b.expected - a.expected)
answer = (j, x) => {
  const p1 = post.reduce((s, v, g) => s + v * S.pg[g][j], 0);
  const p_obs = x === 1 ? p1 : 1 - p1;
  mutable hist = [...hist, {j, x, expected: exp_bits(j), before: entropy(post), surprisal: -Math.log2(p_obs)}];
}
hidden_answer = j => Math.random() < S.pg[hidden][j] ? 1 : 0
```

```{ojs}
viewof pick = Inputs.table(remaining, {
  columns: ["expected", "text"],
  header: {text: "Question", expected: "Exp. bits"},
  format: {expected: x => x.toFixed(3), text: (t, i, data) => t ?? data[i].label},
  width: {expected: 80},
  multiple: false,
  required: false,
  rows: 8,
  layout: "fixed"
})
```

```{ojs}
chosen = pick ?? remaining[0]
viewof act = Inputs.button(chosen ? [
    [`Answer: ${chosen.label1}`, () => answer(chosen.j, 1)],
    [`Answer: ${chosen.label0}`, () => answer(chosen.j, 0)],
    ["Let the simulated person answer", () => answer(chosen.j, hidden_answer(chosen.j))]
  ] : [["No questions left", () => null]])
viewof act2 = Inputs.button([
    ["Ask the best question", () => { if (remaining.length) { const j = remaining[0].j; answer(j, hidden_answer(j)); } }],
    ["Ask the worst question", () => { if (remaining.length) { const j = remaining[remaining.length - 1].j; answer(j, hidden_answer(j)); } }],
    ["Start over", () => { mutable hist = []; mutable hidden = Math.floor(Math.random() * 64); }]
  ])
viewof reveal = Inputs.toggle({label: "Show where the simulated person is"})
```

```{ojs}
bits_left = entropy(post)
gained = 6 - bits_left
theta_plot = {
  const b = S.bounds;
  // Bars never move: each slice is drawn at the population's height. Its shade
  // is the chance that this person is in it. The line is the updated density.
  const slices = post.map((v, g) => ({g, p: v, x1: b[g], x2: b[g + 1],
    pop: (1 / 64) / (b[g + 1] - b[g]), dens: v / (b[g + 1] - b[g])}));
  const line = slices.flatMap(d => [{x: d.x1, y: d.dens}, {x: d.x2, y: d.dens}]);
  const top = Math.max(0.42, ...slices.map(d => d.dens));
  return Plot.plot({
    width: Math.min(width, 680), height: 250, marginBottom: 40, marginTop: 24,
    x: {domain: [-3.2, 3.2], label: "θ: the trait the test measures (low → high)", labelAnchor: "center", labelArrow: "none"},
    y: {domain: [0, top * 1.08], axis: null},
    color: {type: "linear", domain: [0, Math.max(2 / 64, ...post)], range: ["#f4f7fb", "#08306b"]},
    marks: [
      Plot.rectY(slices, {x1: "x1", x2: "x2", y: "pop", fill: "p", inset: 0.3,
        title: d => `slice ${d.g + 1} of 64\nθ from ${d.x1.toFixed(2)} to ${d.x2.toFixed(2)}\nchance the person is here: ${(d.p * 100).toFixed(1)}%`}),
      hist.length ? Plot.line(line, {x: "x", y: "y", stroke: "#111", strokeWidth: 1.6}) : null,
      reveal ? Plot.ruleX([(slices[hidden].x1 + slices[hidden].x2) / 2], {stroke: "#d73027", strokeWidth: 2.5}) : null,
      reveal ? Plot.text([slices[hidden]], {x: d => (d.x1 + d.x2) / 2, y: top * 1.08, text: () => "the simulated person", dy: -4, lineAnchor: "bottom", fill: "#d73027"}) : null,
      Plot.ruleY([0])
    ]
  });
}
theta_key = html`<div style="font-size:0.85rem; color:#444; display:flex; flex-wrap:wrap; gap:0.4rem 1.2rem; max-width:680px; margin-bottom:2px;">
  <span><svg width="44" height="10"><defs><linearGradient id="kg"><stop offset="0" stop-color="#eef2f8"/><stop offset="1" stop-color="#08306b"/></linearGradient></defs><rect width="44" height="10" fill="url(#kg)" stroke="#ccc"/></svg>
    the population in 64 equal slices, shaded by the chance this person is in each (light → dark: less → more likely)</span>
  <span><svg width="26" height="10"><line x1="0" y1="5" x2="26" y2="5" stroke="#111" stroke-width="1.6"/></svg>
    where this person probably is, as a density (appears after the first answer)</span>
</div>`
counter = html`<div style="font-size:0.95rem; max-width:680px;">
  <div style="display:flex; justify-content:space-between;">
    <span><b style="color:${accent_r}">${gained.toFixed(2)} bits gained</b></span>
    <span>${bits_left.toFixed(2)} of 6 bits left</span></div>
  <div style="font-size:0.9rem; margin-top:2px;">About as certain as knowing the person is in one of
    <b>${Math.pow(2, bits_left).toFixed(Math.pow(2, bits_left) < 10 ? 1 : 0)}</b> equally likely slices (from 64).<sup>*</sup></div>
  <div style="height:14px; background:#e9ecef; border-radius:3px; overflow:hidden; margin:4px 0;">
    <div style="height:100%; width:${Math.max(0, gained) / 6 * 100}%; background:${accent_r};"></div></div>
  <div style="color:#666; font-size:0.85rem;">${hist.length} question${hist.length === 1 ? "" : "s"} asked.
  Before any answer every slice has the same shade: all are equally likely.
  Each answer darkens the slices that would most likely give it, and lightens the rest.</div>
  <div style="color:#666; font-size:0.8rem; margin-top:4px;"><sup>*</sup> An effective count, 2<sup>bits left</sup>.
  The chances are rarely spread evenly over a set of slices and zero elsewhere; they form a bump with tails.
  The count says how many equally likely slices would leave the same uncertainty (entropy) as that bump.</div></div>`
p1_now = chosen ? post.reduce((s, v, g) => s + v * S.pg[g][chosen.j], 0) : null
bars = chosen ? Plot.plot({
  width: Math.min(width, 380), height: 140, marginLeft: 90,
  x: {domain: [0, 1], label: "probability, given what we know", grid: true},
  y: {label: null},
  marks: [
    Plot.barX([{r: chosen.label1, p: p1_now}, {r: chosen.label0, p: 1 - p1_now}],
      {y: "r", x: "p", fill: blue_r}),
    Plot.text([{r: chosen.label1, p: p1_now}, {r: chosen.label0, p: 1 - p1_now}],
      {y: "r", x: "p", text: d => (d.p * 100).toFixed(0) + "%", dx: 16})
  ]
}) : html``
log_rows = hist.map((h, i) => ({
  "#": i + 1,
  answer: h.x === 1 ? S.items[h.j].label1 : S.items[h.j].label0,
  "expected bits": h.expected.toFixed(3),
  surprisal: h.surprisal.toFixed(2),
  question: S.items[h.j].text ?? S.items[h.j].label
}))
html`<div style="display:flex; flex-wrap:wrap; gap:1.5rem; align-items:flex-start;">
  <div>${theta_key}${theta_plot}${counter}</div>
  <div style="flex:1; min-width:260px;">
    ${chosen ? html`<p style="margin-bottom:0.3rem;"><b>Next question:</b> ${chosen.text ?? chosen.label}</p>
    <p style="color:#666; margin-top:0; font-size:0.9rem;">Expected to give
    <b style="color:${accent_r}">${chosen.expected.toFixed(3)} bits</b>, out of a possible 1:
    I(θ; X<sub>j</sub> | answers so far).</p>${bars}` :
    html`<p>Every question has been asked.</p>`}
  </div></div>`
```

```{ojs}
hist.length ? Inputs.table(log_rows, {rows: 6, layout: "fixed",
  width: {"#": 35, answer: 90, "expected bits": 100, surprisal: 80}}) : html``
```

A few things show up quickly. No real question is worth anywhere near a full
bit: the best ones promise about half a bit at most. A full bit would take a
perfectly sharp (Guttman) item whose cut falls exactly in the middle of what we
currently believe, splitting the remaining possibilities in half. Questions
lose value as you go, because the easy distinctions have already been made.
And now and then the shading spreads *out* after an answer. That happens when
the person says something the model did not expect.

::: {.callout-note}
## Which information? The quantities on this page

Several different "bits" appear below. They are not interchangeable.

- **I(θ; X)**: what the full answer pattern says about the person, on average
  over the population. This is "the test's information" in the warehouse
  figures.
- **I(θ; X~j~)**: what one item says *on its own*. Items repeat each other, so
  these do not add up to the test's total: I(θ; X) ≤ Σ~j~ I(θ; X~j~).
- **I(θ; X~j~ | answers so far)**: what one more item is expected to add, given
  the particular answers already seen. These are the "expected bits" in the game
  above, and the same item is worth less once earlier answers have done its job.
- **Bits gained by one answer**: H(θ | before) − H(θ | after). This is a realised
  change, not an average, so it can be negative. Averaged over possible answers
  it equals the expected bits.
- **Surprisal**: −log~2~ P(answer | answers so far), how unexpected one answer was.
  The [Surprise](#surprise) section conditions instead on all the person's other answers.
- **I(θ; S)** and **I(θ; X | S)**: what the sum score keeps, and what it drops.
:::

## Surprise

```{r}
#| label: surprise
#| fig-height: 4.2
#| fig-width: 7.5

reps <- s$replays
lab_of <- function(it) s$items$label[match(it, s$items$item)]
# Leave-one-out surprisal: -log2 P(x_j | every other answer), with the posterior
# over the 2^K theta groups built from all items except j. Order plays no part;
# items are shown easiest to hardest only for reading.
loo_surprisal <- function(x, PG) {
  ll <- log(PG) %*% diag(x, length(x)) + log(1 - PG) %*% diag(1 - x, length(x))  # groups x items
  tot <- rowSums(ll)
  vapply(seq_along(x), function(j) {
    lp <- tot - ll[, j]; post <- exp(lp - max(lp)); post <- post / sum(post)
    p1 <- sum(post * PG[, j])
    -log2(if (x[j] == 1) p1 else 1 - p1)
  }, numeric(1))
}
ord_b <- order(s$items$b)
surp <- bind_rows(lapply(reps, function(r) {
  x <- r$responses[s$items$item[ord_b]]
  tibble(respondent = r$respondent, kind = r$kind, step = seq_along(x), item = lab_of(names(x)),
         surprisal = loo_surprisal(x, s$group_probs[, names(x), drop = FALSE]), response = unname(x))
}))
ord_items <- surp |> filter(kind == "aberrant") |> arrange(step) |> pull(item)
b_of <- s$items$b[match(ord_items, s$items$label)]
ab <- filter(surp, kind == "aberrant") |> mutate(answer = if_else(response == 1, "correct", "incorrect"))
ab_resp <- reps[[which(sapply(reps, `[[`, "kind") == "aberrant")]]
n_easy_miss <- sum(ab$response[1:3] == 0)
n_hard_hit <- sum(ab$response[(nrow(ab) - 2):nrow(ab)] == 1)
ggplot(surp, aes(step, surprisal, group = respondent)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = irw_grey) +
  geom_step(data = filter(surp, kind != "aberrant"), colour = irw_grey, alpha = 0.5,
            direction = "mid") +
  geom_step(data = ab, colour = irw_red, linewidth = 1, direction = "mid") +
  geom_point(data = ab, aes(shape = answer), colour = irw_red, fill = "white", size = 2.6, stroke = 1.2) +
  scale_shape_manual(values = c(correct = 16, incorrect = 21), name = "aberrant respondent's answer") +
  scale_x_continuous(breaks = seq_along(ord_items),
                     labels = sprintf("%s\n(b = %s)", ord_items, formatC(b_of, format = "f", digits = 1))) +
  labs(x = "questions, from easiest (left) to hardest (right)",
       y = "surprisal (bits)",
       title = sprintf("The aberrant respondent misses %d of the 3 easiest questions and gets %d of the 3 hardest right",
                       n_easy_miss, n_hard_hit),
       subtitle = sprintf("Red: the respondent with the worst person-fit (Zh = %s, sum score %d of %d). Grey: five typical respondents. Surprisal = −log₂ P(answer | all their other answers). Dashed: 1 bit, a fair coin flip.",
                          fmt(ab_resp$zh, 1), ab_resp$sum_score, nrow(s$items))) +
  theme(legend.position = "top", axis.text.x = element_text(size = 7.5))
```

The surprisal of an answer is −log~2~ of the probability the model gave it.
An answer the model thought had a 50% chance costs 1 bit of surprise; one it
thought had a 1-in-8 chance costs 3 bits. Here each answer is judged against
*everything else the person answered*: hide one answer, work out from the
others where the person probably sits, and ask how likely the hidden answer
was from there. Nothing depends on the order of asking; the questions are
lined up from easiest to hardest only so the pattern is easy to read.

Most answers from typical respondents sit around or below a bit, because their
other answers place them well and each answer agrees with that placement. The
red respondent's answers disagree with each other. The rest of the pattern says
"able enough for the hard questions", which makes the misses on the easy ones
very unlikely, and vice versa, so the surprise lands on exactly the answers
that do not fit. This is the logic of person-fit statistics such as Zh, one
answer at a time.

Elsewhere on the page, as in the game at the top, surprisal is judged against
the answers *so far*. That version depends on order, and it adds up: the
surprisals of one ordering sum to the surprisal of the whole pattern. The
leave-one-out version does not add up that way, but it gives each answer one
number whatever the order, which is what a question about fit needs.

<!-- Expected versus actual: commented out 2026-10-06 (Ben); restore by removing this wrapper and the eval/include lines.

## Expected versus actual

```{r}
#| label: expected-actual
#| eval: false
#| include: false
#| fig-height: 4.6
#| fig-width: 7.5

ea <- bind_rows(lapply(reps[c(3, which(sapply(reps, `[[`, "kind") == "aberrant"))], function(r) {
  st <- r$fixed$steps[-1, ]
  lbl <- s$items$label[match(st$item, s$items$item)]
  tibble(who = if (r$kind == "aberrant") "Aberrant respondent" else "Typical respondent (median score)",
         step = st$step, item = lbl, expected = st$expected_bits, actual = st$actual_bits)
})) |>
  pivot_longer(c(expected, actual), names_to = "kind", values_to = "bits") |>
  mutate(kind = factor(kind, c("expected", "actual"), c("expected bits", "bits actually gained")))
tot <- ea |> group_by(who, kind) |> summarise(bits = sum(bits), .groups = "drop")
ggplot(ea, aes(factor(step), bits, fill = kind)) +
  geom_hline(yintercept = 0, colour = "grey30") +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~ factor(who, c("Typical respondent (median score)", "Aberrant respondent")), ncol = 1) +
  scale_fill_manual(values = c("expected bits" = "grey75", "bits actually gained" = irw_accent), name = NULL,
                    labels = c("expected bits: I(θ; Xⱼ | earlier answers)", "bits actually gained: H(θ | before) − H(θ | after)")) +
  labs(x = "question, in test order", y = "bits",
       title = "Information is a promise on average: a single answer can even lose bits",
       subtitle = sprintf("Totals over the test. Typical: expected %s, gained %s. Aberrant: expected %s, gained %s.",
                          fmt(tot$bits[tot$who != "Aberrant respondent" & tot$kind == "expected bits"]),
                          fmt(tot$bits[tot$who != "Aberrant respondent" & tot$kind == "bits actually gained"]),
                          fmt(tot$bits[tot$who == "Aberrant respondent" & tot$kind == "expected bits"]),
                          fmt(tot$bits[tot$who == "Aberrant respondent" & tot$kind == "bits actually gained"]))) +
  theme(legend.position = "top")
```

The expected bits of an item are an average over everyone who might be
answering. For a particular person, the drop in uncertainty can be larger
(a surprising answer that rules out a lot) or negative (an answer that
contradicts what earlier answers suggested, spreading the posterior back
out). Averaged over people, actual gains equal expected gains. That is the
sense in which I(θ; X) is a property of the test and the population, not of
any one respondent.

-->

## Watching uncertainty fall

```{r}
#| label: uncertainty
#| fig-height: 4.4
#| fig-width: 7.5

unc <- s$uncertainty |>
  mutate(order = factor(order, c("fixed", "random", "adaptive"),
                        c("test's own order", "random order", "adaptive: most informative next")))
trails <- bind_rows(lapply(reps, function(r) tibble(respondent = r$respondent,
                                                     k = r$fixed$steps$step, entropy = r$fixed$steps$entropy)))
end_bits <- 6 - filter(unc, k == max(k))$mean_entropy[1]
ad3 <- filter(unc, order == "adaptive: most informative next", k == 3)$mean_entropy
fx3 <- filter(unc, order == "test's own order", k == 3)$mean_entropy
ggplot() +
  geom_line(data = trails, aes(k, entropy, group = respondent), colour = irw_grey, alpha = 0.35) +
  geom_line(data = unc, aes(k, mean_entropy, colour = order, linetype = order), linewidth = 1.1) +
  scale_colour_manual(values = c(irw_blue, irw_grey, "black"), name = NULL) +
  scale_linetype_manual(values = c("solid", "dashed", "solid"), name = NULL) +
  scale_y_continuous(limits = c(NA, 6.05), breaks = 0:6) +
  scale_x_continuous(breaks = 0:nrow(s$items)) +
  labs(x = "questions asked", y = "bits left: H(slice | answers so far)",
       title = sprintf("After all %d numeracy questions, about %s of the 6 bits are resolved",
                       nrow(s$items), fmt(end_bits, 1)),
       subtitle = sprintf("Mean over 500 simulated respondents. Adaptive ordering front-loads the gains (%s vs %s bits left after 3 questions). Faint lines: six real respondents.",
                          fmt(ad3), fmt(fx3))) +
  theme(legend.position = "top")
```

Every order ends in the same place, because the same answers have been given.
What changes is how fast you get there. Asking the most informative question
next, given what is already known, buys in three questions what the fixed
order needs several more to reach. This is the logic of computerised adaptive
testing, and of Wordle solvers. The ending point is the real news, though:
a full pass through this test leaves most of the uncertainty about which of
64 groups a person belongs to unresolved.

## How the warehouse numbers were computed

Selection started from every dichotomous IRW table with 5–60 items and at
least 500 respondents. Each candidate was then fetched and checked. Tables
with waves or treatment arms were restricted to one administration: the
earliest wave, then the control arm. Tables with repeated person–item rows
(trial-level tasks) were excluded, as were tables with response density below
0.8. ENEM tables were left out for this version. That left
`r r_all_candidates` tables, of which `r n_distinct(two$table)` produced
results (`r length(failures)` failed: `r paste(unique(sub(" .*", "", failures)), collapse = ", ")`,
where every respondent gives the same answer to every item). Tables with more
than 10,000 respondents were randomly subsampled to 10,000
(`r sum(two$subsampled)` tables). Items answered by fewer than half the
respondents, and items with no variance, were dropped.

For each table we fit Rasch and 2PL models with `mirt`, using a lognormal(0, 1)
prior on 2PL slopes, and read the latent distribution from each fit: N(0, σ²)
with σ² estimated for Rasch, N(0, 1) for 2PL. All quantities are integrals over
that distribution on an 81-point grid spanning ±6 SD. In bits:

- **I(θ; S)**, the information in the sum score, via the Lord–Wingersky recursion.
- **I(θ; X)**, the information in the full response pattern. It is exact by
  enumerating all patterns when there are 16 or fewer items, and Monte Carlo
  above that (standard error below 0.005 bits). For Rasch the two are equal,
  because the sum score is sufficient.
- **H(X | θ)**, the noise, and H(X) = I(θ; X) + H(X | θ).
- Marginal reliability, the reliability of EAP sum scores, and their Gaussian
  bit equivalents −½ log~2~(1 − ρ).
- Per-item bits, length curves, and a sum-score model check.

Validation checks (entropy edge cases, single items under 1 bit, Rasch
sufficiency, I(θ; S) ≤ I(θ; X) on every table, Monte Carlo against enumeration,
binned against continuous, simulation recovery, grid stability) are in
`bits_per_item_checks.R`. All passed.

## Now, the whole warehouse

```{r}
#| label: corpus
#| fig-height: 4
#| fig-width: 7.5

med <- median(two$I_X)
p <- ggplot(two, aes(I_X)) +
  geom_histogram(binwidth = 0.1, boundary = 0, fill = irw_accent, colour = "white") +
  geom_vline(xintercept = med, linetype = "dashed") +
  annotate("text", x = med + 0.03, y = Inf, vjust = 1.5, hjust = 0,
           label = sprintf("median %s bits", fmt(med))) +
  scale_x_continuous(breaks = seq(0, 3, 0.5)) +
  labs(x = "information about the person, I(θ; X), in bits (2PL)", y = "tables",
       title = sprintf("Most tests tell you less than 2 bits about a person: %d of %d do",
                       sum(two$I_X < 2), nrow(two)),
       subtitle = sprintf("%d dichotomous IRW tables. Less than 2 bits means fewer than 4 equally likely groups resolved.",
                          nrow(two)))
p
```

Across `r nrow(two)` tables, the median test delivers `r fmt(med)` bits about
the person it measures: about as much as learning which half of the
population someone is in. The most informative,
`r arrange(two, desc(I_X))$table[1]` (`r arrange(two, desc(I_X))$n_items[1]` items),
reaches `r fmt(max(two$I_X))` bits. Tests are short and noisy relative to the
population spread, and bits grow only logarithmically with precision: each
extra bit requires halving the standard error.

```{r}
#| label: corpus-items
#| fig-height: 4.6
#| fig-width: 7.5

ceil <- tibble(n = 5:60, bits = log2(5:60 + 1))
share_ceiling <- two$I_X / log2(two$n_items + 1)
g_items <- ggplot(two, aes(n_items, I_X)) +
  geom_line(data = ceil, aes(n, bits), colour = "grey30", linetype = "dashed") +
  annotate("text", x = 40, y = log2(41) + 0.35, label = "ceiling: n perfect items, log₂(n + 1)",
           size = 3.4, colour = "grey30") +
  geom_point(aes(text = sprintf("%s\n%d items\nI = %.2f bits", table, n_items, I_X)),
             colour = irw_accent, alpha = 0.7, size = 1.8) +
  scale_x_continuous(breaks = c(5, 10, 20, 30, 40, 50, 60)) +
  scale_y_continuous(limits = c(0, 6.2), breaks = 0:6) +
  labs(x = "number of items", y = "I(θ; X), bits (2PL)",
       title = sprintf("Longer tests give more bits, but the median test reaches only %s of what that many perfect items could",
                       scales::percent(median(share_ceiling), accuracy = 1)))
g_items <- wl(g_items, 60)
g_items$labels$title <- gsub("\n", "<br>", g_items$labels$title)
ggplotly(g_items, tooltip = "text") |>
  layout(title = list(text = paste0("<b>", g_items$labels$title, "</b>"), x = 0, xanchor = "left",
                      font = list(size = 15)), margin = list(t = 70))
```

The dashed line is a hard ceiling. On a single trait, n items that never make
mistakes can produce only n + 1 different answer patterns, so they carry at
most log~2~(n + 1) bits, reached when the items cut the population into n + 1
equal parts. Real tables sit far below it, a median of
`r scales::percent(median(share_ceiling), accuracy = 1)` of the way up. The
missing bits are lost to noise, not to test length: answers are probabilistic,
so the same person could produce many patterns.

## Reliability in bits

Reliability and bits measure related things. If θ and a score were jointly
normal, with the score's correlation with θ equal to √ρ, the information
between them would be exactly −½ log~2~(1 − ρ). Item responses are binary, not
normal, so the question is how far real tests depart from that curve, and
why. We look first at synthetic tests, where we control the items, and then at
the warehouse.

Turning that formula around, b bits corresponds to a reliability of
1 − 4^−b^. To see what that buys, simulate people from the Gaussian picture
and compare each person's true rank with the rank the score gives them.

```{r}
#| label: rank-panels
#| fig-height: 6.6
#| fig-width: 7.5

set.seed(20261006)
bits_show <- c(0, 1, 2, 3, 4, 6)
rank_stats <- function(b, n = 1e6) {
  r <- sqrt(1 - 4^-b)
  th <- rnorm(n); sc <- r * th + sqrt(1 - r^2) * rnorm(n)
  c(rho = r^2, rank_err = 100 * mean(abs(pnorm(th) - pnorm(sc))),
    concord = 0.5 + asin(r) / pi)          # P(pair in the right order), exact
}
rs <- as_tibble(t(sapply(bits_show, rank_stats))) |> mutate(bits = bits_show)
rs <- rs |> mutate(lab = sprintf("%s %s · reliability %s\ntypical rank error %s points\n%.1f%% of pairs in the right order",
                                 bits, if_else(bits == 1, "bit", "bits"), signif(rho, 4),
                                 if_else(rank_err < 1, sprintf("%.1f", rank_err), sprintf("%.2g", rank_err)),
                                 100 * concord),
                   lab = factor(lab, lab))
dots <- rs |> select(bits, lab) |> rowwise() |>
  reframe(lab = lab, th = rnorm(600)) |>
  left_join(select(rs, lab, rho), by = "lab") |>
  mutate(sc = sqrt(rho) * th + sqrt(1 - rho) * rnorm(n()),
         x = 100 * pnorm(th), y = 100 * pnorm(sc))
ggplot(dots, aes(x, y)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(alpha = 0.5, size = 0.9) +
  facet_wrap(~ lab, ncol = 3) +
  scale_x_continuous(breaks = c(0, 50, 100)) + scale_y_continuous(breaks = c(0, 50, 100)) +
  labs(x = "person's true rank in the population (percentile)",
       y = "rank the test gives them (percentile)",
       title = "How well a test ranks people, by how much it tells you about them",
       subtitle = sprintf("Each panel: 600 simulated people, θ and score jointly normal. On the dashed line the ranking is perfect; each extra bit halves the scatter around it. Rank errors are in percentile points. The median warehouse test carries %s bits; the most informative, %s.",
                          fmt(median(two$I_X)), fmt(max(two$I_X)))) +
  theme(strip.text = element_text(hjust = 0, size = 9), panel.spacing = unit(0.8, "lines"))
```

Most of what a ranking needs arrives in the first bit or two: one bit already
puts five pairs in six in the right order. The panels show the Gaussian
picture, which is the dashed curve below. The warehouse tests sit a little
above that curve, so their bits are only approximately where they land in these
panels.

```{r}
#| label: reliability-sim
#| fig-height: 5.6
#| fig-width: 7.5

sim_sum <- rel_sim |> group_by(kind) |>
  summarise(dev_lo = min(I_X - gauss), dev_hi = max(I_X - gauss), .groups = "drop")
sim_dev <- rel_sim |> group_by(kind, n_items) |> summarise(dev = mean(I_X - gauss), .groups = "drop")
curve_s <- tibble(rho = seq(0, 0.97, by = 0.005), bits = -0.5 * log2(1 - rho))
ggplot(rel_sim, aes(rho, I_X, colour = kind)) +
  geom_line(data = curve_s, aes(rho, bits), inherit.aes = FALSE, colour = "grey30", linetype = "dashed") +
  geom_point(aes(size = n_items), alpha = 0.6) +
  scale_colour_manual(values = c(moderate = irw_blue, `sharp items` = irw_red, `one tail` = "#e6a100"),
                      labels = c(moderate = "moderate items, spread across θ",
                                 `sharp items` = "3 near-Guttman items",
                                 `one tail` = "items bunched at the high end"),
                      name = NULL) +
  scale_size_continuous(range = c(1, 3.5), breaks = c(5, 20, 60), name = "items") +
  labs(x = "marginal reliability", y = "I(θ; X), bits",
       title = "In simulation, ordinary items follow the Gaussian curve; sharp items and one-sided tests rise above it",
       subtitle = sprintf("Synthetic 2PL tests, θ ~ N(0, 1), 5–60 items, 10 random draws each. Gap above the curve: moderate %s–%s bits; sharp %s–%s; one tail %s–%s.",
                          fmt(max(0, sim_sum$dev_lo[sim_sum$kind == "moderate"])), fmt(sim_sum$dev_hi[sim_sum$kind == "moderate"]),
                          fmt(sim_sum$dev_lo[sim_sum$kind == "sharp items"]), fmt(sim_sum$dev_hi[sim_sum$kind == "sharp items"]),
                          fmt(sim_sum$dev_lo[sim_sum$kind == "one tail"]), fmt(sim_sum$dev_hi[sim_sum$kind == "one tail"]))) +
  guides(colour = guide_legend(nrow = 2, override.aes = list(size = 3))) +
  theme(legend.position = "top", legend.box = "vertical")
```

With ordinary items, moderately discriminating and spread across θ, the
Gaussian formula is close: reliability tells you the bits. Two departures push
tests above the curve. A few **near-Guttman items** split the population almost
deterministically. That is worth bits a reliability coefficient, built on
squared error, undercounts. The effect is largest in short tests
(`r fmt(sim_dev$dev[sim_dev$kind == "sharp items" & sim_dev$n_items == 5])` bits at 5 items)
and fades as ordinary items accumulate
(`r fmt(sim_dev$dev[sim_dev$kind == "sharp items" & sim_dev$n_items == 60])` at 60).
Items **bunched in one tail** make the score precise at one end and vague
elsewhere, and that gap grows with test length
(`r fmt(sim_dev$dev[sim_dev$kind == "one tail" & sim_dev$n_items == 5])` →
`r fmt(sim_dev$dev[sim_dev$kind == "one tail" & sim_dev$n_items == 60])` bits).

```{r}
#| label: reliability
#| fig-height: 5
#| fig-width: 7.5

curve_df <- tibble(rho = seq(0, 0.97, by = 0.005), bits = -0.5 * log2(1 - rho))
refs <- tibble(rho = c(0.8, 0.9, 0.95), bits = -0.5 * log2(1 - rho))
dev <- two$I_X - two$bits_rho_marginal
max_a_tbl <- tapply(items_df$a, items_df$table, max)
g <- ggplot(two, aes(rho_marginal, I_X)) +
  geom_line(data = curve_df, aes(rho, bits), colour = "grey30", linetype = "dashed") +
  geom_point(aes(text = sprintf("%s\n%d items\nρ = %.2f\nI = %.2f bits", table, n_items, rho_marginal, I_X)),
             colour = irw_accent, alpha = 0.7, size = 1.8) +
  geom_point(data = refs, aes(rho, bits), shape = 21, fill = "white", size = 3) +
  geom_text(data = refs, aes(rho, bits, label = sprintf("ρ = %.2f\n%.2f bits", rho, bits)),
            nudge_x = -0.07, nudge_y = 0.12, size = 3.2) +
  labs(x = "marginal reliability (2PL)", y = "I(θ; X), bits",
       title = sprintf("Reliability predicts bits closely, but the Gaussian formula undercounts by a median %s bits",
                       fmt(median(dev))),
       subtitle = "Dashed: −½ log₂(1 − ρ), the information if θ and its estimate were jointly normal.")
gauss_note <- "Dashed line: −½ log₂(1 − ρ), the information if θ and its estimate were jointly normal. Hover a point for the table."
g <- wl(g, 55)
g$labels$title <- gsub("\n", "<br>", g$labels$title)
g$labels$subtitle <- NULL
ggplotly(g, tooltip = "text") |>
  layout(title = list(text = paste0("<b>", g$labels$title, "</b>"), x = 0, xanchor = "left",
                      font = list(size = 15)), margin = list(t = 70))
```

`r gauss_note`

On the Gaussian curve, a reliability of .80 is worth `r fmt(-0.5 * log2(0.2))` bits, .90 is worth `r fmt(-0.5 * log2(0.1))`,
and .95 is worth `r fmt(-0.5 * log2(0.05))`. The tables follow the curve
(correlation `r fmt(cor(two$I_X, two$bits_rho_marginal))`), but almost all sit
above it: the exact information exceeds the Gaussian value in
`r sum(dev > 0)` of `r nrow(two)` tables, by up to `r fmt(max(dev))` bits.
Binary responses are not normal. The gap tracks the sharpest item in the
table (Spearman correlation `r fmt(cor(two$I_X - two$bits_rho_marginal, max_a_tbl[two$table], method = "spearman"))`
with the largest slope): a near-deterministic item splits the population
cleanly, which is worth bits that a reliability coefficient, built on
squared error, does not register.

## Diminishing returns

```{r}
#| label: length
#| fig-height: 4.6
#| fig-width: 7.5

kmax <- 40
med_curve <- curves_df |>
  group_by(k) |> filter(n() >= 20) |>
  summarise(random = median(random_mean), greedy = median(greedy), n = n(), .groups = "drop") |>
  filter(k <= kmax)
first <- curves_df |> filter(k == 1) |> pull(random_mean) |> median()
g5_all <- curves_df |> group_by(table) |> mutate(n = max(k), full = random_mean[k == n]) |>
  ungroup() |> filter(k == 5) |> mutate(greedy_share = greedy / full, random_share = random_mean / full)
g5 <- filter(g5_all, n >= 15)
at10 <- med_curve$random[med_curve$k == 10]
ggplot() +
  geom_line(data = filter(curves_df, k <= kmax), aes(k, random_mean, group = table),
            colour = irw_grey, alpha = 0.15) +
  geom_line(data = med_curve, aes(k, random, linetype = "median, random order"),
            colour = irw_accent, linewidth = 1.3) +
  geom_line(data = med_curve, aes(k, greedy, linetype = "median, greedy order"),
            colour = "black", linewidth = 0.9) +
  scale_linetype_manual(values = c("median, random order" = "solid", "median, greedy order" = "dashed"), name = NULL) +
  labs(x = "number of items in the sum score", y = "I(θ; Sₖ), bits",
       title = sprintf("The first item gives a median %s bits; the first ten together only %s",
                       fmt(first), fmt(at10)),
       subtitle = sprintf("Faint: each table's curve, mean of 5 random item orders. Medians use only lengths reached by at least 20 tables (k ≤ %d).",
                          max(med_curve$k))) +
  theme(legend.position = "top")
```

Each additional item adds less than the one before, because it mostly repeats
what earlier items already said about the person. The greedy order (best
item first, then the item that adds most to it, and so on) shows how much of a
test's information a short form could keep. Among the
`r nrow(g5)` tables with at least 15 items, the 5 greedy items carry a median
`r scales::percent(median(g5$greedy_share), accuracy = 1)` of the full sum
score's bits; 5 random items carry `r scales::percent(median(g5$random_share), accuracy = 1)`.

One curiosity: for a sum score, more items is not always better. Adding a
flat item to an unweighted sum adds noise without adding signal, so the
curve can turn down. In `r sum(g5_all$greedy_share > 1)` tables the best
five items alone carry more information than the full sum score, up to
`r fmt(max(g5_all$greedy_share), 1)` times as much
(`r g5_all$table[which.max(g5_all$greedy_share)]`).

## Where do the bits come from?

```{r}
#| label: item-bits
#| fig-height: 4.4
#| fig-width: 7.5

ib <- items_df |> filter(is.finite(b))
pa <- ggplot(ib, aes(a, bits)) +
  geom_bin_2d(bins = 50) +
  scale_fill_gradient(low = "#e5f5f0", high = irw_accent, trans = "log10", name = "items") +
  scale_x_log10() +
  labs(x = "discrimination a (log scale)", y = "I(θ; Xⱼ), the item alone (bits)")
pb <- ggplot(filter(ib, abs(b) < 6), aes(b, bits)) +
  geom_bin_2d(bins = 50) +
  scale_fill_gradient(low = "#e5f5f0", high = irw_accent, trans = "log10", name = "items") +
  labs(x = "difficulty b (θ units)", y = NULL) +
  guides(fill = "none")
(pa | pb) + plot_layout(guides = "collect") +
  plot_annotation(title = wrap(sprintf("Bits need both a sharp item and a central one; the median item gives %s bits",
                                  fmt(median(items_df$bits), 2))),
                  subtitle = sprintf("%s items from %d tables (2PL).", format(nrow(items_df), big.mark = ","),
                                     n_distinct(items_df$table)))
```

An item's bits depend on two things. Discrimination is how sharply the item
separates people just below its difficulty from people just above. Position
is where that cut falls. A highly discriminating item placed far in a tail
still says little about most people, because nearly everyone gives the same
answer to it. The ceiling is 1 bit, reached only by a perfectly sharp item at
the population median.

## What does sum-scoring cost?

```{r}
#| label: sum-gap
#| fig-height: 3.8
#| fig-width: 7.5

two <- mutate(two, gap = I_X - I_S, gap_prop = gap / I_X)
ggplot(two, aes(gap_prop)) +
  geom_histogram(binwidth = 0.02, boundary = 0, fill = irw_accent, colour = "white") +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "share of the response pattern's information lost by using the sum score",
       y = "tables",
       title = sprintf("Using the sum score instead of the full pattern costs a median %s of the bits",
                       scales::percent(median(two$gap_prop), accuracy = 1)),
       subtitle = sprintf("Median loss %s bits; largest %s bits (%s). 2PL.",
                          fmt(median(two$gap), 3), fmt(max(two$gap)),
                          arrange(two, desc(gap))$table[1]))
```

Under the Rasch model the sum score carries every bit the pattern carries. Under
the 2PL, items with higher slopes deserve more weight, and an unweighted sum
throws some information away. The loss is small: a median of
`r scales::percent(median(two$gap_prop), accuracy = 1)` of the pattern's
information. It exceeds a quarter in `r sum(two$gap_prop > 0.25)` tables.
The loss tracks how unequal the slopes are (Spearman correlation
`r fmt(cor(two$gap_prop, tapply(log(items_df$a), items_df$table, sd)[two$table], method = "spearman"))`
with the spread of log slopes): tables that mix very sharp items with very
flat ones lose the most.

## From groups to continuous θ

```{r}
#| label: bins
#| fig-height: 3.8
#| fig-width: 7.5

bc <- bind_rows(lapply(names(showcases), function(nm) {
  b <- showcases[[nm]]$bin_convergence
  cont <- b$bits[!is.finite(b$K)]
  tibble(test = show_titles[nm], K = b$K[is.finite(b$K)], bits = b$bits[is.finite(b$K)], cont = cont)
}))
ggplot(bc, aes(K, bits, colour = test)) +
  geom_hline(aes(yintercept = cont, colour = test), linetype = "dashed", alpha = 0.6) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_x_continuous(breaks = 1:8, labels = function(k) paste0(k, "\n(", 2^k, ")")) +
  scale_colour_manual(values = c(irw_blue, irw_accent, "black"), name = NULL) +
  labs(x = "K: number of bits needed to name a group (number of groups)", y = "I(group; X), bits",
       title = "By 64 groups, counting groups captures essentially all the information about θ",
       subtitle = "Dashed: information about continuous θ. Binned information never exceeds it.") +
  theme(legend.position = "top")
```

The game above asks "which of 64 groups?". The warehouse numbers ask about θ
itself, a continuous quantity. These are close: information about the group
can never exceed information about θ, since knowing θ tells you the group, and
it converges to it as the groups get finer. With 2 groups the numeracy test
gives `r fmt(showcases[[main]]$bin_convergence$bits[1])` bits. With 64 it gives
`r fmt(showcases[[main]]$bin_convergence$bits[6], 3)`, against
`r fmt(tail(showcases[[main]]$bin_convergence$bits, 1), 3)` for continuous θ.
Because the tests deliver only one or two bits, six bits of groups is far
more resolution than any of them can use.

## Limitations

- **Model dependence.** Every number is computed from a fitted model, not from
  the data directly. If a 2PL is wrong for a table (guessing, multiple
  dimensions, local dependence), the bits are wrong too. The sum-score check
  gives one warning sign: model-implied and observed score distributions
  diverge by more than 0.1 bit (KL) in `r sum(two$sum_kl_bits > 0.1, na.rm = TRUE)` tables.
- **Local dependence and sharp items.** Even with a slope prior,
  `r n_distinct(items_df$table[items_df$a > 5])` tables have at least one item
  with a slope above 5, often near-duplicate items. Such pairs make the model
  treat a repeated answer as two independent pieces of evidence, which
  inflates the bits.
- **Unidimensionality.** We fit one trait per table. A multidimensional test
  carries information about several traits, and the numbers here count only
  what projects onto the first.
- **The population matters.** Information is relative to the spread of
  people taking the test. The same items given to a more homogeneous group
  would deliver fewer bits, because there is less uncertainty to resolve to
  begin with.
- **Subsampling and missing data.** Large tables were subsampled to 10,000
  respondents; that affects parameter precision, not the definition of the
  quantity. Items answered by fewer than half the respondents were dropped,
  and I(θ; X) treats every respondent as answering every remaining item.
- **Convergence.** `r sum(!two$converged)` 2PL fit
  (`r paste(two$table[!two$converged], collapse = ", ")`) did not meet `mirt`'s
  convergence criterion; its results are included but should be read with care.
- **Scope.** Only dichotomous items, and no ENEM tables in this version.
  Polytomous items can carry more than one bit each.

## Reproducibility

{{< include ../components/_source-links.qmd >}}

{{< include ../components/_data-version.qmd >}}

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 0. Optional: re-scan IRW for eligible tables (writes scout_tables.csv)
source("vignettes/bits_per_item_scout.R")

# 1. Validation checks, then the compute script (from the project root)
source("vignettes/bits_per_item_checks.R")
source("vignettes/bits_per_item_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/bits_per_item.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(unique(summary_df$table), output_file = "bits_per_item_data/references.bib")
```
:::

Visual inspiration: 3Blue1Brown, ["Solving Wordle using information
theory"](https://www.youtube.com/watch?v=v68zYyaEmEA). No material from the
video is reused here.

---

## References

::: {#refs}
:::
