---
title: "Comparing Predictive Performance of CFA and IRT: Limited vs. Full Information"
description: |
  Person-level cross-validation across 10 IRW datasets comparing DWLS-estimated
  CFA to MML-estimated 2PL IRT, using the InterModel Vigorish (IMV) as a common
  predictive metric. The primary comparison is the direct head-to-head IMV of
  IRT vs. CFA, motivated by Cho (2026).
date: today
bibliography: cfairtdata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_classic(base_size = 13))

cache_path <- "cfairtdata/cfa_irt_imv_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache          <- readRDS(cache_path)
  table_names     <- .cache$table_names
  dataset_results <- .cache$dataset_results
  dataset_summary <- .cache$dataset_summary
  nfold           <- .cache$nfold
  date_run        <- .cache$date_run
  n_datasets      <- length(table_names)
} else {
  table_names <- character(); dataset_results <- list()
  dataset_summary <- data.frame(); nfold <- 5; n_datasets <- 0
  date_run <- Sys.Date()
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures will not appear. ",
  "Generate results by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/cfa_irt_imv_compute.R\n```\n",
  ":::\n", sep = ""
)
```

This vignette compares response-level predictive accuracy from a confirmatory factor analysis (CFA) model estimated with limited-information methods (DWLS) to that of an item response theory (IRT) model estimated with full-information methods (marginal maximum likelihood; MML). We use the InterModel Vigorish [@domingue2024imv] as the comparison metric, specifically the **direct head-to-head IMV of IRT vs. CFA** — measuring how much better IRT predicts held-out responses relative to CFA as a baseline.

The question is motivated by @cho2026, who showed via simulation that full-information estimation generally recovers item parameters more accurately than limited-information estimation. Here we ask: does that parameter-recovery advantage translate into better **out-of-sample predictive accuracy** on real data?

Both models assume a **single latent dimension**: the CFA specifies one factor, and the IRT specifies a one-dimensional 2PL. Model structure is therefore held constant across the comparison — only the estimation method varies. CFA and IRT are theoretically equivalent under this shared assumption, with interchangeable parameters [@takane1987], but they differ in how they extract information from the data during estimation. DWLS operates on pairwise item relationships (polychoric correlations); MML uses the full multivariate response pattern. We run the comparison across `r n_datasets` IRW datasets to assess consistency.

::: {.callout-note}
Results are pre-computed from `cfa_irt_imv_compute.R`
(`r n_datasets` datasets, `r nfold`-fold person-level CV, computed `r format(date_run, "%B %Y")`).
:::

---

## Models and estimation

### CFA (DWLS)

We specify a **single-factor** CFA model with all items loading on one latent variable. Items are treated as ordered (binary), and parameters are estimated with DWLS under the delta parameterization [@rosseel2012].

```{r}
#| eval: false
#| echo: true

library(irw); library(lavaan); library(mirt); library(imv)

# Dataset list — change to re-run on different IRW datasets
table_names <- c(
  "gilbert_meta_2", "gilbert_meta_1", "gilbert_meta_8",
  "gilbert_meta_12", "gilbert_meta_15", "gilbert_meta_37",
  "fims_tam", "psychtools_ability",
  "much_tte_2025_matrixreasoning", "naep_multilcirt"
)
```

### IRT (MML)

We fit a **unidimensional** 2PL model with a lognormal prior on discrimination parameters, consistent with the existing IMV vignette on this site. Both models therefore share the same latent-variable structure; the comparison isolates estimation method, not model complexity.

---

## IMV comparison

**CFA prediction** follows Eq. 6 from @zhang2026: for a test person with factor score $\hat\theta_i$, the predicted probability for item $j$ is

$$\Pr(X_{ij} = 1 \mid \hat\theta_i) = \Phi\!\left(\frac{\lambda_j \hat\theta_i - \tau_j}{\sqrt{\psi_j}}\right)$$

where $\lambda_j$ is the factor loading, $\tau_j$ is the threshold, $\psi_j$ is the residual variance (delta parameterization), and $\Phi$ is the standard normal CDF.

**IRT prediction** uses the 2PL ICC evaluated at the EAP-estimated $\hat\theta_i$ for each test person.

The **primary metric** is the direct head-to-head IMV of IRT vs. CFA:

$$\text{IMV}(\text{CFA}, \text{IRT}) = \text{imv.binary}(y, \hat p_{\text{CFA}}, \hat p_{\text{IRT}})$$

A positive value means IRT predicts better than CFA; negative means CFA predicts better. We also report each model's IMV relative to a prevalence baseline for context.

---

## Cross-validation design

For each dataset, a single person-level 5-fold CV loop assigns identical fold splits to both models. Neither model has any individual-level information about test persons beyond the population-level item parameters estimated on training persons.

```{r}
#| eval: false
#| echo: true

predict_cfa <- function(train, test, model, vary) {
  fit         <- cfa(model, data = train, ordered = vary,
                     parameterization = "delta", estimator = "DWLS")
  ov_pred     <- lavPredict(fit, newdata = test, type = "ov")
  y_threshold <- inspect(fit, what = "est")$tau
  y_residual  <- inspect(fit, what = "est")$theta
  yprob <- matrix(NA, nrow = nrow(test), ncol = length(vary),
                  dimnames = list(NULL, vary))
  for (j in seq_along(vary)) {
    num        <- ov_pred[, j] - y_threshold[j]
    yprob[, j] <- pnorm(num / sqrt(y_residual[j, j]))
  }
  yprob
}

predict_irt <- function(train_resp, test_resp, mirt_model) {
  fit <- mirt(train_resp, mirt_model,
              itemtype  = rep("2PL", ncol(train_resp)),
              method    = "EM",
              technical = list(NCYCLES = 10000),
              verbose   = FALSE)
  theta_test <- fscores(fit, response.pattern = test_resp, method = "EAP")[, 1]
  probs      <- probtrace(fit, theta_test)
  p1_cols    <- seq(2, ncol(probs), by = 2)
  yprob      <- probs[, p1_cols, drop = FALSE]
  colnames(yprob) <- colnames(train_resp)
  yprob
}
```

```{r}
#| eval: false
#| echo: true

# Inside the CV loop — three IMV quantities per fold per item:
for (j in seq_along(items)) {
  y_j   <- test_df[, j]
  prev_j <- rep(prev_train[j], length(y_j))
  item_cfa_v_irt[j]  <- imv.binary(y_j, yprob_cfa[, j], yprob_irt[, j])  # primary
  item_cfa_v_prev[j] <- imv.binary(y_j, prev_j, yprob_cfa[, j])           # context
  item_irt_v_prev[j] <- imv.binary(y_j, prev_j, yprob_irt[, j])           # context
}
```

---

## Results

### IMV(CFA, IRT) across datasets

```{r}
#| label: tbl-dataset-summary
#| tbl-cap: "Scale-level IMV across datasets. IMV(CFA, IRT) is the primary comparison: how much better IRT predicts than CFA. Context columns show each model's gain over a prevalence baseline."

ds <- dataset_summary |>
  mutate(
    ci_lo = scale_cfa_v_irt_mean - 1.96 * scale_cfa_v_irt_sd / sqrt(K),
    ci_hi = scale_cfa_v_irt_mean + 1.96 * scale_cfa_v_irt_sd / sqrt(K),
    imv_direct = sprintf("%.4f [%.4f, %.4f]",
                         scale_cfa_v_irt_mean, ci_lo, ci_hi),
    cfa_prev   = sprintf("%.4f", scale_cfa_v_prev_mean),
    irt_prev   = sprintf("%.4f", scale_irt_v_prev_mean)
  ) |>
  arrange(desc(scale_cfa_v_irt_mean)) |>
  select(table_name, N, J, imv_direct, cfa_prev, irt_prev)

knitr::kable(ds,
             col.names = c("Dataset", "N", "J",
                           "IMV(CFA, IRT) [95% CI]",
                           "IMV(prev, CFA)",
                           "IMV(prev, IRT)"),
             align = "lrrlrr")
```

### Distribution of IMV(CFA, IRT)

```{r}
#| label: fig-imv-direct
#| fig-cap: "Scale-level IMV(CFA, IRT) with 95% CI across datasets, ordered by point estimate. Positive values indicate IRT predicts better than CFA."
#| fig-height: 5

plot_df <- dataset_summary |>
  mutate(
    ci_lo      = scale_cfa_v_irt_mean - 1.96 * scale_cfa_v_irt_sd / sqrt(K),
    ci_hi      = scale_cfa_v_irt_mean + 1.96 * scale_cfa_v_irt_sd / sqrt(K),
    table_name = reorder(table_name, scale_cfa_v_irt_mean),
    pos        = scale_cfa_v_irt_mean >= 0
  )

p_main <- ggplot(plot_df,
                 aes(x = scale_cfa_v_irt_mean, y = table_name, color = pos)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = c("TRUE" = "#d6604d", "FALSE" = "#4393c3"),
                     guide = "none") +
  labs(x = "IMV(CFA, IRT)  [positive = IRT better]", y = NULL)

p_main
```

### Context: each model vs. prevalence baseline

```{r}
#| label: fig-vs-prev
#| fig-cap: "Scale-level IMV of each model against a prevalence baseline, across datasets. Both models substantially outperform simple prevalence prediction."
#| fig-height: 5

prev_df <- dataset_summary |>
  select(table_name, N, J,
         CFA = scale_cfa_v_prev_mean,
         IRT = scale_irt_v_prev_mean) |>
  pivot_longer(c(CFA, IRT), names_to = "model", values_to = "imv") |>
  mutate(table_name = reorder(table_name, imv, mean))

p_prev <- ggplot(prev_df,
                 aes(x = imv, y = table_name, color = model, group = model)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("CFA" = "#4393c3", "IRT" = "#d6604d"),
                     name = NULL) +
  labs(x = "Scale-level IMV vs. prevalence baseline", y = NULL) +
  theme(legend.position = "bottom")

p_prev
```

```{r}
#| label: summary-stats
#| echo: false

med_imv  <- median(dataset_summary$scale_cfa_v_irt_mean)
mean_imv <- mean(dataset_summary$scale_cfa_v_irt_mean)
pct_pos  <- mean(dataset_summary$scale_cfa_v_irt_mean > 0) * 100
```

Across `r n_datasets` datasets, the median IMV(CFA, IRT) is `r round(med_imv, 4)` (mean `r round(mean_imv, 4)`). IRT outperforms CFA in `r round(pct_pos, 0)`% of datasets.

---

## Discussion

**IMV(CFA, IRT) is the direct test.** Rather than comparing each model's gain over prevalence and taking the difference, `imv.binary(y, p_CFA, p_IRT)` directly measures how much additional predictive value IRT provides over CFA — expressed in the same interpretable betting-return units.

**Both models substantially outperform prevalence.** The vs-prevalence IMVs confirm that knowing a person's estimated latent ability (from either model) is genuinely informative compared to predicting item means for everyone.

**The CFA–IRT gap is generally small.** The small IMV(CFA, IRT) values across most datasets are consistent with their theoretical equivalence [@takane1987]: once item parameters and factor scores are estimated, the prediction formula is structurally the same. The estimation advantage of MML documented by @cho2026 appears to matter more for parameter recovery than for response-level prediction at these sample sizes.

**Estimation method matters when the unidimensional assumption is strained.** Both models assume a single latent dimension, so any difference in predictive accuracy reflects estimation robustness, not model structure. When DWLS performs dramatically worse than MML — as with `gilbert_meta_37` — it suggests that DWLS's reliance on pairwise polychoric correlations is more brittle under the conditions that stress the unidimensional assumption (e.g., extreme item prevalences, near-zero correlations). Neither model is detecting multidimensionality; one is simply more numerically stable when a unidimensional fit is imperfect.

---

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
Rscript vignettes/cfa_irt_imv_compute.R
quarto::quarto_render("vignettes/cfa_irt_imv.qmd")
```

---

## References
