---
title: "Comparing Predictive Performance of CFA and IRT: Limited vs. Full Information"
description: |
  Person-level cross-validation across 10 IRW datasets comparing DWLS-estimated
  CFA to MML-estimated 2PL IRT, using the InterModel Vigorish (IMV) as a common
  predictive metric. Motivated by Cho (2026), who showed full-information
  estimation recovers parameters more accurately — here we ask whether that
  advantage extends to out-of-sample prediction.
date: today
bibliography: cfairtdata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_classic(base_size = 13))

cache_path <- "cfairtdata/cfa_irt_imv_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache          <- readRDS(cache_path)
  table_names     <- .cache$table_names
  dataset_results <- .cache$dataset_results
  dataset_summary <- .cache$dataset_summary
  nfold           <- .cache$nfold
  date_run        <- .cache$date_run
  n_datasets      <- length(table_names)
} else {
  table_names <- character(); dataset_results <- list()
  dataset_summary <- data.frame(); nfold <- 5; n_datasets <- 0
  date_run <- Sys.Date()
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures will not appear. ",
  "Generate results by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/cfa_irt_imv_compute.R\n```\n",
  ":::\n", sep = ""
)
```

This vignette compares response-level predictive accuracy from a confirmatory factor analysis (CFA) model estimated with limited-information methods (DWLS) to that of an item response theory (IRT) model estimated with full-information methods (marginal maximum likelihood; MML), using the InterModel Vigorish [@domingue2024imv] as a common metric. We run the comparison across `r n_datasets` IRW datasets to assess how consistent any differences are.

The contrast is motivated by @cho2026, who showed via simulation that full-information estimation generally recovers item parameters more accurately than limited-information estimation — particularly with smaller samples, more items, and skewed trait distributions. Here we ask: does that parameter-recovery advantage translate into better **out-of-sample predictive accuracy** on real data?

CFA and IRT are theoretically equivalent models with interchangeable parameters [@takane1987], but they differ in how they use information during estimation. DWLS operates on pairwise item relationships (polychoric correlations); MML uses the full multivariate response pattern.

::: {.callout-note}
Results are pre-computed from `cfa_irt_imv_compute.R`
(`r n_datasets` datasets, `r nfold`-fold person-level CV, computed `r format(date_run, "%B %Y")`).
:::

---

## Models and estimation

### CFA (DWLS)

We specify a single-factor CFA model with all items loading on one latent variable. Items are treated as ordered (binary), and parameters are estimated with DWLS under the delta parameterization [@rosseel2012].

```{r}
#| eval: false
#| echo: true

library(irw); library(lavaan); library(mirt); library(imv)

# Dataset list — change to re-run on different IRW datasets
table_names <- c(
  "gilbert_meta_2", "gilbert_meta_1", "gilbert_meta_8",
  "gilbert_meta_12", "gilbert_meta_15", "gilbert_meta_37",
  "fims_tam", "psychtools_ability",
  "much_tte_2025_matrixreasoning", "naep_multilcirt"
)
```

### IRT (MML)

We fit a 2PL model with a lognormal prior on discrimination parameters, consistent with the existing IMV vignette on this site.

---

## Prediction formulas

**CFA prediction** follows Eq. 6 from @zhang2026: for a test person with factor score $\hat\theta_i$, the predicted probability for item $j$ is

$$\Pr(X_{ij} = 1 \mid \hat\theta_i) = \Phi\!\left(\frac{\lambda_j \hat\theta_i - \tau_j}{\sqrt{\psi_j}}\right)$$

where $\lambda_j$ is the factor loading, $\tau_j$ is the threshold, $\psi_j$ is the residual variance (delta parameterization), and $\Phi$ is the standard normal CDF.

**IRT prediction** uses the 2PL ICC evaluated at the EAP-estimated $\hat\theta_i$ for each test person.

---

## Cross-validation design

For each dataset, a single person-level 5-fold CV loop assigns identical fold splits to both models. Neither model has any individual-level information about test persons beyond the population-level item parameters estimated on training persons.

```{r}
#| eval: false
#| echo: true

predict_cfa <- function(train, test, model, vary) {
  fit         <- cfa(model, data = train, ordered = vary,
                     parameterization = "delta", estimator = "DWLS")
  ov_pred     <- lavPredict(fit, newdata = test, type = "ov")
  y_threshold <- inspect(fit, what = "est")$tau
  y_residual  <- inspect(fit, what = "est")$theta
  yprob <- matrix(NA, nrow = nrow(test), ncol = length(vary),
                  dimnames = list(NULL, vary))
  for (j in seq_along(vary)) {
    num        <- ov_pred[, j] - y_threshold[j]
    yprob[, j] <- pnorm(num / sqrt(y_residual[j, j]))
  }
  yprob
}

predict_irt <- function(train_resp, test_resp, mirt_model) {
  fit <- mirt(train_resp, mirt_model,
              itemtype  = rep("2PL", ncol(train_resp)),
              method    = "EM",
              technical = list(NCYCLES = 10000),
              verbose   = FALSE)
  theta_test <- fscores(fit, response.pattern = test_resp, method = "EAP")[, 1]
  probs      <- probtrace(fit, theta_test)
  p1_cols    <- seq(2, ncol(probs), by = 2)
  yprob      <- probs[, p1_cols, drop = FALSE]
  colnames(yprob) <- colnames(train_resp)
  yprob
}
```

The baseline for each IMV calculation is item prevalence in the training fold.

---

## Results

### Scale-level IMV across datasets

```{r}
#| label: tbl-dataset-summary
#| tbl-cap: "Scale-level IMV for CFA and IRT across datasets. 95% CI computed as mean ± 1.96 × sd/√K over valid folds."

ds <- dataset_summary |>
  mutate(
    cfa_ci = sprintf("%.4f [%.4f, %.4f]",
                     scale_cfa_mean,
                     scale_cfa_mean - 1.96 * scale_cfa_sd / sqrt(K),
                     scale_cfa_mean + 1.96 * scale_cfa_sd / sqrt(K)),
    irt_ci = sprintf("%.4f [%.4f, %.4f]",
                     scale_irt_mean,
                     scale_irt_mean - 1.96 * scale_irt_sd / sqrt(K),
                     scale_irt_mean + 1.96 * scale_irt_sd / sqrt(K)),
    adv    = sprintf("%+.4f", advantage)
  ) |>
  select(table_name, N, J, cfa_ci, irt_ci, adv)

knitr::kable(ds,
             col.names = c("Dataset", "N", "J",
                           "CFA IMV [95% CI]", "IRT IMV [95% CI]",
                           "IRT − CFA"),
             align = "lrrllr")
```

### Dataset-level comparison

```{r}
#| label: fig-dataset-imv
#| fig-cap: "Scale-level IMV for each dataset: CFA (blue) vs. IRT (red), with 95% fold-based confidence intervals. Datasets are ordered by CFA IMV."
#| fig-height: 5.5

plot_df <- dataset_summary |>
  mutate(
    cfa_lo = scale_cfa_mean - 1.96 * scale_cfa_sd / sqrt(K),
    cfa_hi = scale_cfa_mean + 1.96 * scale_cfa_sd / sqrt(K),
    irt_lo = scale_irt_mean - 1.96 * scale_irt_sd / sqrt(K),
    irt_hi = scale_irt_mean + 1.96 * scale_irt_sd / sqrt(K),
    table_name = reorder(table_name, scale_cfa_mean)
  ) |>
  pivot_longer(
    cols      = c(scale_cfa_mean, scale_irt_mean),
    names_to  = "model",
    values_to = "imv"
  ) |>
  mutate(
    lo    = ifelse(model == "scale_cfa_mean", cfa_lo, irt_lo),
    hi    = ifelse(model == "scale_cfa_mean", cfa_hi, irt_hi),
    model = recode(model,
                   scale_cfa_mean = "CFA (DWLS)",
                   scale_irt_mean = "IRT (MML)")
  )

p_forest <- ggplot(plot_df,
                   aes(x = imv, y = table_name, color = model, group = model)) +
  geom_errorbarh(aes(xmin = lo, xmax = hi),
                 height = 0.25, position = position_dodge(width = 0.5)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("CFA (DWLS)" = "#4393c3",
                                 "IRT (MML)"  = "#d6604d"),
                     name = NULL) +
  labs(x = "Scale-level IMV (vs. prevalence baseline)", y = NULL) +
  theme(legend.position = "bottom")

p_forest
```

### CFA vs. IRT per dataset

```{r}
#| label: fig-cfa-vs-irt
#| fig-cap: "Dataset-level scale IMV: CFA on x-axis, IRT on y-axis. Points above the diagonal favor IRT. Datasets labeled by name."
#| fig-height: 4.5

p_scatter <- ggplot(dataset_summary,
                    aes(x = scale_cfa_mean, y = scale_irt_mean)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 3, color = "#4393c3", alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = table_name),
                            size = 3, color = "grey30", max.overlaps = 20) +
  labs(x = "Scale IMV — CFA (DWLS)",
       y = "Scale IMV — IRT (MML)",
       subtitle = "Points above dashed line favor IRT") +
  coord_fixed()

p_scatter
```

```{r}
#| label: fig-advantage-dist
#| fig-cap: "Distribution of IRT − CFA IMV advantage across datasets. Positive = IRT predicts better."
#| fig-height: 3.5

p_adv <- ggplot(dataset_summary, aes(x = advantage)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_dotplot(fill = "#4393c3", color = "white", binwidth = 0.0005) +
  labs(x = "IMV advantage (IRT − CFA)", y = NULL) +
  scale_y_continuous(breaks = NULL)

p_adv
```

```{r}
#| label: advantage-stats
#| echo: false

med_adv  <- median(dataset_summary$advantage)
mean_adv <- mean(dataset_summary$advantage)
pct_pos  <- mean(dataset_summary$advantage > 0) * 100
```

Across `r n_datasets` datasets, the median IRT advantage is `r round(med_adv, 4)` IMV units (mean `r round(mean_adv, 4)`). IRT outperforms CFA in `r round(pct_pos, 0)`% of datasets.

---

## Discussion

**Both models substantially outperform the prevalence baseline.** Scale-level IMV values in the range seen here reflect meaningful gains from knowing a person's estimated latent ability, regardless of estimation method.

**The IRT advantage is consistently small.** The narrow gap between IRT and CFA is consistent with their theoretical equivalence [@takane1987]: once item parameters and factor scores are estimated, the prediction formula is the same. The estimation advantage of MML documented by @cho2026 appears to matter more for parameter recovery than for response-level prediction, at least for the sample sizes and factor structures represented here.

**Interpretive caveats.** The CV scheme is person-level — a strict test. Both models assume a single latent dimension. Results may differ for multidimensional instruments or for datasets with smaller samples where MML's efficiency advantage is larger.

---

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
Rscript vignettes/cfa_irt_imv_compute.R
quarto::quarto_render("vignettes/cfa_irt_imv.qmd")
```

---

## References
