---
title: "Comparing Predictive Performance of CFA and IRT: Limited vs. Full Information"
description: |
  Person-level cross-validation across 10 IRW datasets comparing DWLS-estimated
  CFA to MML-estimated 2PL IRT, using the InterModel Vigorish (IMV) as a common
  predictive metric. The primary comparison is the direct head-to-head IMV of
  IRT vs. CFA, motivated by Cho (2026).
date: today
bibliography: cfairtdata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)

theme_set(theme_classic(base_size = 13))

cache_path <- "cfairtdata/cfa_irt_imv_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache             <- readRDS(cache_path)
  table_names        <- .cache$table_names
  dataset_results    <- .cache$dataset_results
  dataset_results_mr <- .cache$dataset_results_mr %||% list()
  dataset_summary    <- .cache$dataset_summary
  nfold              <- .cache$nfold
  nfold_mr           <- .cache$nfold_mr %||% 5L
  date_run           <- .cache$date_run
  n_datasets         <- length(table_names)
  has_mr             <- "scale_mr_cfa_v_irt_mean" %in% names(dataset_summary) &&
                        any(!is.na(dataset_summary$scale_mr_cfa_v_irt_mean))
} else {
  table_names <- character(); dataset_results <- list(); dataset_results_mr <- list()
  dataset_summary <- data.frame(); nfold <- 5; nfold_mr <- 5; n_datasets <- 0
  date_run <- Sys.Date(); has_mr <- FALSE
}

`%||%` <- function(x, y) if (is.null(x)) y else x

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures will not appear. ",
  "Generate results by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/cfa_irt_imv_compute.R\n```\n",
  ":::\n", sep = ""
)
```

This vignette compares response-level predictive accuracy from a confirmatory factor analysis (CFA) model estimated with limited-information methods (DWLS) to that of an item response theory (IRT) model estimated with full-information methods (marginal maximum likelihood; MML). We use the InterModel Vigorish [@domingue2024imv] as the comparison metric, specifically the **direct head-to-head IMV of IRT vs. CFA**.

The question is motivated by @cho2026, who showed via simulation that full-information estimation generally recovers item parameters more accurately than limited-information estimation. Here we ask: does that parameter-recovery advantage translate into better **out-of-sample predictive accuracy** on real data?

Both models assume a **single latent dimension**. Model structure is therefore held constant — only the estimation method varies. We run the comparison under **two out-of-sample paradigms** [@stenhaug2022], which correspond to different definitions of what "new data" means for an item response matrix:

- **Missing person (MP):** entire persons are held out. The model predicts a new person's full response vector, integrating over their unknown ability. This is the natural paradigm for ability estimation in traditional testing and corresponds to person-level cross-validation.

- **Missing response (MR):** individual item responses are held out for persons who are otherwise observed. The model predicts a person's response to a specific item using their responses to the remaining items to pin down their ability. This is the natural paradigm for adaptive testing and item-level scoring.

We run both paradigms across `r n_datasets` IRW datasets and compare the conclusions.

::: {.callout-note}
Results are pre-computed from `cfa_irt_imv_compute.R`
(`r n_datasets` datasets, `r nfold`-fold person-level CV, computed `r format(date_run, "%B %Y")`).
:::

---

## Models and estimation

### CFA (DWLS)

We specify a **single-factor** CFA model with all items loading on one latent variable. Items are treated as ordered (binary), and parameters are estimated with DWLS under the delta parameterization [@rosseel2012].

```{r}
#| eval: false
#| echo: true

library(irw); library(lavaan); library(mirt); library(imv)

# Dataset list — change to re-run on different IRW datasets
table_names <- c(
  "gilbert_meta_2", "gilbert_meta_1", "gilbert_meta_8",
  "gilbert_meta_12", "gilbert_meta_15", "gilbert_meta_37",
  "fims_tam", "psychtools_ability",
  "much_tte_2025_matrixreasoning", "naep_multilcirt"
)
```

### IRT (MML)

We fit a **unidimensional** 2PL model with a lognormal prior on discrimination parameters, consistent with the existing IMV vignette on this site. Both models therefore share the same latent-variable structure; the comparison isolates estimation method, not model complexity.

---

## IMV comparison

**CFA prediction** follows Eq. 6 from @zhang2026: for a test person with factor score $\hat\theta_i$, the predicted probability for item $j$ is

$$\Pr(X_{ij} = 1 \mid \hat\theta_i) = \Phi\!\left(\frac{\lambda_j \hat\theta_i - \tau_j}{\sqrt{\psi_j}}\right)$$

where $\lambda_j$ is the factor loading, $\tau_j$ is the threshold, $\psi_j$ is the residual variance (delta parameterization), and $\Phi$ is the standard normal CDF.

**IRT prediction** uses the 2PL ICC evaluated at the EAP-estimated $\hat\theta_i$ for each test person.

The **primary metric** is the direct head-to-head IMV of IRT vs. CFA:

$$\text{IMV}(\text{CFA}, \text{IRT}) = \text{imv.binary}(y, \hat p_{\text{CFA}}, \hat p_{\text{IRT}})$$

A positive value means IRT predicts better than CFA; negative means CFA predicts better. We also report each model's IMV relative to a prevalence baseline for context.

---

## Cross-validation design

We run two CV loops per dataset with identical IMV evaluation at each step.

**Missing person (MP):** `r nfold`-fold person-level CV. Entire persons are withheld; item parameters are estimated from training persons; test-person ability is unknown to the model, so prediction integrates over the ability distribution $\hat{g}(\theta)$.

**Missing response (MR):** `r nfold_mr`-fold item-response-level CV. Fold membership is assigned at the cell level, stratified by person (so each person contributes to every fold). Held-out cells are set to NA in the training matrix; both models are refit on that matrix; each person's held-in item responses are used to estimate their ability, which is then applied to predict the held-out items.

```{r}
#| eval: false
#| echo: true

predict_cfa <- function(train, test, model, vary) {
  fit         <- cfa(model, data = train, ordered = vary,
                     parameterization = "delta", estimator = "DWLS")
  ov_pred     <- lavPredict(fit, newdata = test, type = "ov")
  y_threshold <- inspect(fit, what = "est")$tau
  y_residual  <- inspect(fit, what = "est")$theta
  yprob <- matrix(NA, nrow = nrow(test), ncol = length(vary),
                  dimnames = list(NULL, vary))
  for (j in seq_along(vary)) {
    num        <- ov_pred[, j] - y_threshold[j]
    yprob[, j] <- pnorm(num / sqrt(y_residual[j, j]))
  }
  yprob
}

predict_irt <- function(train_resp, test_resp, mirt_model) {
  fit <- mirt(train_resp, mirt_model,
              itemtype  = rep("2PL", ncol(train_resp)),
              method    = "EM",
              technical = list(NCYCLES = 10000),
              verbose   = FALSE)
  theta_test <- fscores(fit, response.pattern = test_resp, method = "EAP")[, 1]
  probs      <- probtrace(fit, theta_test)
  p1_cols    <- seq(2, ncol(probs), by = 2)
  yprob      <- probs[, p1_cols, drop = FALSE]
  colnames(yprob) <- colnames(train_resp)
  yprob
}
```

```{r}
#| eval: false
#| echo: true

# Inside the CV loop — three IMV quantities per fold per item:
for (j in seq_along(items)) {
  y_j   <- test_df[, j]
  prev_j <- rep(prev_train[j], length(y_j))
  item_cfa_v_irt[j]  <- imv.binary(y_j, yprob_cfa[, j], yprob_irt[, j])  # primary
  item_cfa_v_prev[j] <- imv.binary(y_j, prev_j, yprob_cfa[, j])           # context
  item_irt_v_prev[j] <- imv.binary(y_j, prev_j, yprob_irt[, j])           # context
}
```

---

## Results

### IMV(CFA, IRT) across datasets

```{r}
#| label: tbl-dataset-summary
#| tbl-cap: "Scale-level IMV(CFA, IRT) under both prediction paradigms. MP = missing person (person-level CV); MR = missing response (item-response-level CV). Positive values indicate IRT predicts better than CFA."

ds <- dataset_summary |>
  mutate(
    ci_lo_mp = scale_cfa_v_irt_mean - 1.96 * scale_cfa_v_irt_sd / sqrt(K),
    ci_hi_mp = scale_cfa_v_irt_mean + 1.96 * scale_cfa_v_irt_sd / sqrt(K),
    imv_mp   = sprintf("%.4f [%.4f, %.4f]",
                       scale_cfa_v_irt_mean, ci_lo_mp, ci_hi_mp),
    imv_mr   = if ("scale_mr_cfa_v_irt_mean" %in% names(dataset_summary) & has_mr)
                 sprintf("%.4f [%.4f, %.4f]",
                         scale_mr_cfa_v_irt_mean,
                         scale_mr_cfa_v_irt_mean - 1.96 * scale_mr_cfa_v_irt_sd / sqrt(K_mr),
                         scale_mr_cfa_v_irt_mean + 1.96 * scale_mr_cfa_v_irt_sd / sqrt(K_mr))
               else NA_character_
  ) |>
  arrange(desc(scale_cfa_v_irt_mean)) |>
  select(table_name, N, J, imv_mp, imv_mr)

knitr::kable(ds,
             col.names = c("Dataset", "N", "J",
                           "MP: IMV(CFA, IRT) [95% CI]",
                           "MR: IMV(CFA, IRT) [95% CI]"),
             align = "lrrll")
```

### Distribution of IMV(CFA, IRT)

```{r}
#| label: fig-imv-direct
#| fig-cap: "Scale-level IMV(CFA, IRT) with 95% CI across datasets, ordered by point estimate. Positive values indicate IRT predicts better than CFA."
#| fig-height: 5

plot_df <- dataset_summary |>
  mutate(
    ci_lo      = scale_cfa_v_irt_mean - 1.96 * scale_cfa_v_irt_sd / sqrt(K),
    ci_hi      = scale_cfa_v_irt_mean + 1.96 * scale_cfa_v_irt_sd / sqrt(K),
    table_name = reorder(table_name, scale_cfa_v_irt_mean),
    pos        = scale_cfa_v_irt_mean >= 0
  )

p_main <- ggplot(plot_df,
                 aes(x = scale_cfa_v_irt_mean, y = table_name, color = pos)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = c("TRUE" = "#d6604d", "FALSE" = "#4393c3"),
                     guide = "none") +
  labs(x = "IMV(CFA, IRT)  [positive = IRT better]", y = NULL)

p_main
```

### Context: each model vs. prevalence baseline

```{r}
#| label: fig-vs-prev
#| fig-cap: "Scale-level IMV of each model against a prevalence baseline, across datasets. Both models substantially outperform simple prevalence prediction."
#| fig-height: 5

prev_df <- dataset_summary |>
  select(table_name, N, J,
         CFA = scale_cfa_v_prev_mean,
         IRT = scale_irt_v_prev_mean) |>
  pivot_longer(c(CFA, IRT), names_to = "model", values_to = "imv") |>
  mutate(table_name = reorder(table_name, imv, mean))

p_prev <- ggplot(prev_df,
                 aes(x = imv, y = table_name, color = model, group = model)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("CFA" = "#4393c3", "IRT" = "#d6604d"),
                     name = NULL) +
  labs(x = "Scale-level IMV vs. prevalence baseline", y = NULL) +
  theme(legend.position = "bottom")

p_prev
```

```{r}
#| label: summary-stats
#| echo: false

med_imv  <- median(dataset_summary$scale_cfa_v_irt_mean)
mean_imv <- mean(dataset_summary$scale_cfa_v_irt_mean)
pct_pos  <- mean(dataset_summary$scale_cfa_v_irt_mean > 0) * 100
```

Across `r n_datasets` datasets, the median IMV(CFA, IRT) is `r round(med_imv, 4)` (mean `r round(mean_imv, 4)`). IRT outperforms CFA in `r round(pct_pos, 0)`% of datasets.

---

## Missing-response (MR) results

```{r}
#| label: fig-mp-vs-mr
#| fig-cap: "Direct comparison of IMV(CFA, IRT) under missing-person (MP) and missing-response (MR) paradigms. Each point is one dataset. Points above the diagonal indicate stronger IRT advantage under MR than MP."
#| fig-height: 4.5

if (has_mr) {
  both_df <- dataset_summary |>
    filter(!is.na(scale_mr_cfa_v_irt_mean)) |>
    mutate(pos_mp = scale_cfa_v_irt_mean >= 0,
           pos_mr = scale_mr_cfa_v_irt_mean >= 0)

  lim <- range(c(both_df$scale_cfa_v_irt_mean, both_df$scale_mr_cfa_v_irt_mean),
               na.rm = TRUE)

  ggplot(both_df, aes(x = scale_cfa_v_irt_mean, y = scale_mr_cfa_v_irt_mean,
                      label = table_name)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
    geom_vline(xintercept = 0, colour = "grey80") +
    geom_hline(yintercept = 0, colour = "grey80") +
    geom_point(size = 3, colour = "#d6604d") +
    ggrepel::geom_text_repel(size = 3, colour = "grey30", max.overlaps = 10) +
    coord_equal(xlim = lim, ylim = lim) +
    labs(x = "IMV(CFA, IRT)  [missing person]",
         y = "IMV(CFA, IRT)  [missing response]")
}
```

```{r}
#| label: mr-summary-stats
#| echo: false

if (has_mr) {
  mr_valid <- dataset_summary[!is.na(dataset_summary$scale_mr_cfa_v_irt_mean), ]
  med_mr  <- median(mr_valid$scale_mr_cfa_v_irt_mean)
  mean_mr <- mean(mr_valid$scale_mr_cfa_v_irt_mean)
  pct_pos_mr <- mean(mr_valid$scale_mr_cfa_v_irt_mean > 0) * 100
}
```

`r if (has_mr) sprintf("Under the MR paradigm, the median IMV(CFA, IRT) is %.4f (mean %.4f). IRT outperforms CFA in %.0f%% of datasets.", med_mr, mean_mr, pct_pos_mr) else "*MR results pending.*"`

---

## Dimensionality check

The models above assume a single latent dimension. The ratio of the first to second eigenvalue of the inter-item correlation matrix provides a quick signal of how dominant the first factor is: a high ratio is consistent with unidimensionality; a low ratio suggests additional structure that a single-factor model may be missing for both CFA and IRT.

```{r}
#| label: tbl-eigenvalues
#| tbl-cap: "Eigenvalue ratio (λ₁/λ₂) and variance explained by the first factor (% of total) for each dataset. High ratios indicate a dominant first factor; values below 3–4 suggest potential multidimensionality."

if (has_mr && "ev_ratio" %in% names(dataset_summary)) {
  ev_df <- dataset_summary |>
    filter(!is.na(ev_ratio)) |>
    mutate(ev_pct1_pct = round(ev_pct1 * 100, 1)) |>
    arrange(ev_ratio) |>
    select(table_name, N, J, ev_ratio, ev_pct1_pct)

  knitr::kable(ev_df,
               col.names = c("Dataset", "N", "J", "λ₁/λ₂", "% var (F1)"),
               digits = c(0, 0, 0, 2, 1),
               align = "lrrrr")
} else {
  cat("*Eigenvalue results not yet available.*\n")
}
```

Datasets with low λ₁/λ₂ are candidates for a multidimensional model comparison. A natural next step — using the same IMV and MR framework — is to fit a two-factor IRT (or bifactor CFA) for those datasets and test whether the additional dimension improves out-of-sample prediction.

---

## Discussion

**IMV(CFA, IRT) is the direct test.** Rather than comparing each model's gain over prevalence and taking the difference, `imv.binary(y, p_CFA, p_IRT)` directly measures how much additional predictive value IRT provides over CFA — expressed in the same interpretable betting-return units.

**MP vs. MR can tell different stories.** The two paradigms weight model features differently [@stenhaug2022]. MP prediction requires integrating over the unknown ability of a new person, which rewards accurate marginal likelihood estimation (MML's strength). MR prediction uses the person's held-in responses to pin down their ability, which reduces the gap between the two estimation methods — both can estimate ability reliably from partial response vectors. When MP and MR conclusions agree, the finding is robust; when they diverge, the discrepancy reveals which aspect of estimation is driving the difference.

**Both models substantially outperform prevalence.** The vs-prevalence IMVs confirm that knowing a person's estimated latent ability (from either model) is genuinely informative compared to predicting item means for everyone.

**The CFA–IRT gap is generally small.** The small IMV(CFA, IRT) values across most datasets are consistent with their theoretical equivalence [@takane1987]: once item parameters and factor scores are estimated, the prediction formula is structurally the same. The estimation advantage of MML documented by @cho2026 appears to matter more for parameter recovery than for response-level prediction at these sample sizes.

**Estimation method matters when the unidimensional assumption is strained.** Both models assume a single latent dimension, so any difference in predictive accuracy reflects estimation robustness, not model structure. When DWLS performs dramatically worse than MML — as with `gilbert_meta_37` — it suggests that DWLS's reliance on pairwise polychoric correlations is more brittle under the conditions that stress the unidimensional assumption (e.g., extreme item prevalences, near-zero correlations). Neither model is detecting multidimensionality; one is simply more numerically stable when a unidimensional fit is imperfect.

**Dimensionality as a next step.** The eigenvalue ratios provide a quick filter for which datasets are robustly unidimensional and which warrant a multidimensional model comparison. For datasets with a low λ₁/λ₂, fitting a two-factor IRT and comparing it to the one-factor model using both MP and MR IMV would directly test whether additional dimensions improve predictive accuracy — and whether both paradigms agree on the answer.

---

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
Rscript vignettes/cfa_irt_imv_compute.R
quarto::quarto_render("vignettes/cfa_irt_imv.qmd")
```

---

## References
