---
title: "Conjoint Experiments as Paired Comparisons"
description: |
  A conjoint experiment asks people to choose between pairs of randomly built
  profiles: two immigrants, two candidates, two vaccines. The usual summary is
  the average effect of each attribute. IRW holds 849 conjoint experiments, and
  read as item response data each choice is a paired comparison. That lets us
  ask two measurement questions across all of them: how well do the attributes
  predict a choice nobody has seen yet, and how much do individual respondents'
  own preferences add to that? Start by taking one of the experiments yourself.
date: today
bibliography: conjoint_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "conjoint_data/conjoint_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  irt_df           <- .cache$irt_summary
  hhy_amce         <- .cache$hhy_amce
  widget_tasks     <- .cache$widget_tasks
  support          <- .cache$support
  candidate_tables <- .cache$candidate_tables
  date_run         <- .cache$date_run
  is_pilot         <- isTRUE(.cache$pilot)
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_conj         <- .cache$n_conj_tables
  r_irt_pool       <- .cache$irt_pool_n
  r_n_tables_run   <- nrow(summary_df)
} else {
  summary_df <- tibble(table = character(), n_resp = integer(), n_tasks_used = integer(),
                       n_attr = integer(), n_terms = integer(), restrictions = character(),
                       country = character(), imv_attr = numeric(), w_base = numeric(),
                       w_attr = numeric(), imv_indiv = numeric(), w_shared = numeric(),
                       w_indiv = numeric(), lambda = numeric(), n_resp_q2 = integer(),
                       mean_tasks = numeric(), reference = character())
  irt_df <- tibble(table = character(), persons = integer(), items = integer(),
                   per_person = numeric(), lambda = numeric(), imv_indiv = numeric(),
                   w_shared = numeric(), w_indiv = numeric())
  hhy_amce <- tibble(attribute = character(), level = character(), baseline = logical(),
                     published_amce = numeric(), published_se = numeric(),
                     irw_amce = numeric(), irw_se = numeric(), dim = numeric())
  widget_tasks <- tibble(task_key = integer(), profile = integer(), choice = integer())
  support <- tibble(attribute = character(), level = character(), other = character(),
                    other_levels = list())
  candidate_tables <- character()
  date_run         <- Sys.Date()
  is_pilot         <- FALSE
  r_all_candidates <- 0L
  r_n_conj         <- 0L
  r_irt_pool       <- 0L
  r_n_tables_run   <- 0L
}

# Display names for the Hainmueller, Hopkins & Yamamoto attributes and levels,
# in the order the paper lists them (the first level of each is the baseline).
hhy_labels <- list(
  attr_education = c("No formal education" = "No formal education",
    "Equivalent to completing fourth grade in the US" = "4th grade",
    "Equivalent to completing eighth grade in the US" = "8th grade",
    "Equivalent to completing high school in the US" = "High school",
    "Equivalent to completing two years of college in the US" = "Two years of college",
    "Equivalent to completing a college degree in the US" = "College degree",
    "Equivalent to completing a graduate degree in the US" = "Graduate degree"),
  attr_gender = c("female" = "Female", "male" = "Male"),
  attr_origin = c("Germany" = "Germany", "France" = "France", "Mexico" = "Mexico",
    "Philippines" = "Philippines", "Poland" = "Poland", "India" = "India", "China" = "China",
    "Sudan" = "Sudan", "Somalia" = "Somalia", "Iraq" = "Iraq"),
  attr_application_reason = c("Reunite with family members already in the U.S." = "Reunite with family",
    "Seek better job in U.S." = "Seek a better job",
    "Escape political/religious persecution" = "Escape persecution"),
  attr_profession = c("Janitor" = "Janitor", "Waiter" = "Waiter",
    "Child care provider" = "Child care provider", "Gardener" = "Gardener",
    "Financial analyst" = "Financial analyst", "Construction worker" = "Construction worker",
    "Teacher" = "Teacher", "Computer programmer" = "Computer programmer", "Nurse" = "Nurse",
    "Research scientist" = "Research scientist", "Doctor" = "Doctor"),
  attr_job_experience = c("No job training or prior experience" = "None",
    "One or two years of job training and experience" = "1-2 years",
    "Three to five years of job training and experience" = "3-5 years",
    "More than five years of job training and experience" = "More than 5 years"),
  attr_job_plans = c("Has a contract with a U.S. employer" = "Contract with U.S. employer",
    "Does not have a contract with a U.S. employer but has done job interviews" = "Interviews, no contract",
    "Will look for work after arriving in the U.S." = "Will look for work",
    "Has no plans to look for work at this time" = "No plans to look for work"),
  attr_prior_trips = c("Never been to the U.S." = "Never",
    "Entered U.S. once before on a tourist visa" = "Once, tourist visa",
    "Has visited the U.S. many times before on tourist visas" = "Many times, tourist visas",
    "Spent six months with family members in the U.S" = "Six months with family",
    "Entered the U.S. once before without legal authorization" = "Once, without authorization"),
  attr_language = c("During the admission interview, this applicant spoke fluent English" = "Fluent English",
    "During admission interview, this applicant spoke broken English" = "Broken English",
    "During admission interview, this applicant tried to speak English but was unable" = "Tried English, unable",
    "During admission interview, this applicant spoke [language] and used an interpreter" = "Used an interpreter"))
hhy_attr_names <- c(attr_education = "Education", attr_gender = "Gender",
  attr_origin = "Country of origin", attr_application_reason = "Reason for applying",
  attr_profession = "Profession", attr_job_experience = "Job experience",
  attr_job_plans = "Job plans", attr_prior_trips = "Prior trips to the U.S.",
  attr_language = "Language skills")

level_key <- bind_rows(lapply(names(hhy_labels), function(a)
  tibble(attribute = a, attr_label = hhy_attr_names[[a]], level = names(hhy_labels[[a]]),
         short = unname(hhy_labels[[a]]), order_in_attr = seq_along(hhy_labels[[a]]))))
level_key$row <- seq_len(nrow(level_key))

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/conjoint_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, partial run.** This render uses a partial sample of ",
  r_n_tables_run, " tables out of ", r_all_candidates, " eligible tables, ",
  "to validate the pipeline and page layout before the full batch finishes. ",
  "All numbers here are real computed results -- not placeholders -- but ",
  "treat the *distributions* as a small, noisy preview rather than a ",
  "representative picture. Re-run `conjoint_compute.R` (with `CONJ_PILOT` ",
  "unset) to scale up.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`conjoint_compute.R`) directly if you're relying
on the numbers here.
:::

## Motivation

A conjoint experiment shows a respondent two profiles side by side and asks
which one they prefer. Each profile is a bundle of attributes (an immigrant's
education, profession, country of origin and so on), and every attribute is
drawn at random, independently for each profile. The respondent sees a handful
of such pairs. Because the attributes are randomized, comparing how often
profiles with one level are chosen against profiles with another level
estimates that level's average effect on the choice, the *average marginal
component effect* or AMCE [@Hainmueller_2014]. The design is popular in
political science because it lets one experiment estimate the effects of many
attributes at once.

The easiest way to see how this works is to take part in one. The pairs below
are real tasks from Hainmueller, Hopkins & Yamamoto's (2014) study of
immigrant admission, in which 1,396 Americans each compared five pairs of
applicants. Each time, choose the applicant you would admit. After each
choice you will see what the person who actually saw that pair chose, and
the chart below the cards turns your choices into AMCEs and sets them against
the published ones.

```{r}
#| label: widget-data
#| echo: false

w_tasks <- widget_tasks
for (a in names(hhy_labels)) {
  w_tasks[[a]] <- sub("[language]", "a language other than English", w_tasks[[a]], fixed = TRUE)
}
w_amce <- hhy_amce |>
  inner_join(level_key, by = c("attribute", "level")) |>
  arrange(row) |>
  transmute(attribute, attr_label, level, short, baseline,
            published = published_amce, published_se, row)
w_support <- support |>
  mutate(other = ifelse(is.na(other), "", other)) |>
  unnest_longer(other_levels, keep_empty = TRUE) |>
  mutate(other_levels = ifelse(is.na(other_levels), "", other_levels))
w_attrs <- tibble(attribute = names(hhy_attr_names), label = unname(hhy_attr_names))
ojs_define(w_tasks_r = w_tasks, w_amce_r = w_amce, w_support_r = w_support, w_attrs_r = w_attrs)
```

```{ojs}
//| echo: false

w_tasks   = transpose(w_tasks_r)
w_amce    = transpose(w_amce_r)
w_support = transpose(w_support_r)
w_attrs   = transpose(w_attrs_r)

// one entry per task: [profile 1, profile 2], in a random order for this reader
w_pairs = {
  const by = d3.group(w_tasks, d => d.task_key);
  const pairs = Array.from(by.values()).map(v => v.sort((a, b) => a.profile - b.profile));
  return d3.shuffle(pairs.slice());
}

// for each non-baseline level: which values of the restricted partner attribute
// it is compared over (empty = all), and the attribute's baseline level
w_rules = {
  const m = new Map();
  for (const s of w_support) {
    const k = s.attribute + "|" + s.level;
    if (!m.has(k)) m.set(k, {other: s.other, allowed: new Set()});
    if (s.other_levels !== "") m.get(k).allowed.add(s.other_levels);
  }
  return m;
}
w_base = new Map(w_amce.filter(d => d.baseline).map(d => [d.attribute, d.level]))
```

```{ojs}
//| echo: false

viewof w_answers = {
  const el = html`<div class="cj-widget"></div>`;
  el.value = [];
  let i = 0, last = null;
  const cell = (txt) => html`<td style="padding:4px 10px;border-top:1px solid #ddd;">${txt}</td>`;
  function render() {
    const pair = w_pairs[i % w_pairs.length];
    const pick = (p) => {
      last = {mine: p, theirs: pair[0].choice === 1 ? 1 : 2};
      el.value = [...el.value, {pair, chosen: p}];
      i += 1;
      el.dispatchEvent(new Event("input", {bubbles: true}));
      render();
    };
    const btn = (p) => {
      const b = html`<button class="btn btn-outline-primary btn-sm">Admit applicant ${p}</button>`;
      b.onclick = () => pick(p);
      return b;
    };
    const fb = last === null ? "" :
      `On the last pair you chose applicant ${last.mine}; the respondent who saw it chose applicant ${last.theirs}` +
      (last.mine === last.theirs ? " too." : ".");
    const agree = el.value.filter(a => (a.pair[0].choice === 1 ? 1 : 2) === a.chosen).length;
    el.replaceChildren(html`<div>
      <p style="margin-bottom:6px;"><strong>Pair ${i + 1}.</strong> Which applicant would you admit to the United States?</p>
      <table style="border-collapse:collapse;max-width:760px;width:100%;font-size:0.92em;">
        <thead><tr><th></th><th style="padding:4px 10px;">Applicant 1</th><th style="padding:4px 10px;">Applicant 2</th></tr></thead>
        <tbody>${w_attrs.map(a => html`<tr>${cell(html`<strong>${a.label}</strong>`)}${cell(pair[0][a.attribute])}${cell(pair[1][a.attribute])}</tr>`)}</tbody>
        <tfoot><tr><td></td><td style="padding:8px 10px;">${btn(1)}</td><td style="padding:8px 10px;">${btn(2)}</td></tr></tfoot>
      </table>
      <p style="margin-top:6px;color:#555;font-size:0.9em;">${fb}${el.value.length ? ` You have made ${el.value.length} choice${el.value.length > 1 ? "s" : ""} and agreed with the original respondent on ${agree}.` : ""}</p>
    </div>`);
  }
  render();
  return el;
}
```

```{ojs}
//| echo: false

// The reader's AMCEs: for each level, the share of profiles with that level the
// reader chose, minus the share for the baseline level. For the two restricted
// attributes the comparison is made only among profiles where both levels could
// appear, which is what makes the simple difference an AMCE in this design.
w_mine = {
  const profs = w_answers.flatMap(a => a.pair.map(p => ({...p, chosen: p.profile === a.chosen ? 1 : 0})));
  return w_amce.filter(d => !d.baseline).map(d => {
    const rule = w_rules.get(d.attribute + "|" + d.level);
    const ok = p => !rule || rule.other === "" || rule.allowed.has(p[rule.other]);
    const g1 = profs.filter(p => ok(p) && p[d.attribute] === d.level);
    const g0 = profs.filter(p => ok(p) && p[d.attribute] === w_base.get(d.attribute));
    if (!g1.length || !g0.length) return {...d, mine: null};
    const r = g => (d3.sum(g, p => p.chosen) + 0.5) / (g.length + 1);
    const est = d3.mean(g1, p => p.chosen) - d3.mean(g0, p => p.chosen);
    const se = Math.sqrt(r(g1) * (1 - r(g1)) / g1.length + r(g0) * (1 - r(g0)) / g0.length);
    return {...d, mine: est, lo: est - 2 * se, hi: est + 2 * se};
  });
}
```

```{ojs}
//| echo: false

{
  const n = w_answers.length;
  const rows = w_amce.map(d => d.attr_label + ": " + d.short);
  const pub = w_amce.filter(d => !d.baseline);
  const base = w_amce.filter(d => d.baseline);
  const mine = w_mine.filter(d => d.mine !== null);
  const xs = mine.map(d => d.mine), ys = mine.map(d => d.published);
  const mx = d3.mean(xs), my = d3.mean(ys);
  const cor = xs.length > 5 ? d3.sum(xs.map((x, i) => (x - mx) * (ys[i] - my))) /
    Math.sqrt(d3.sum(xs.map(x => (x - mx) ** 2)) * d3.sum(ys.map(y => (y - my) ** 2))) : null;
  const note = n < 5
    ? `Make at least five choices to see your estimates (${n} so far).`
    : `Your estimates from ${n} choices (red, ±2 SE) against the published AMCEs from 1,396 respondents (blue, ±2 SE)` +
      (cor !== null && isFinite(cor) ? `. Across the ${mine.length} levels you have estimates for, the two correlate at r = ${cor.toFixed(2)}.` : ".");
  const plot = Plot.plot({
    width: 760,
    height: 22 * rows.length + 40,
    marginLeft: 260,
    x: {label: "Change in the chance an applicant is chosen", domain: [-0.8, 0.8], grid: true},
    clip: true,
    y: {domain: rows, label: null},
    marks: [
      Plot.ruleX([0], {stroke: "#999"}),
      Plot.dot(base, {x: 0, y: d => d.attr_label + ": " + d.short, symbol: "diamond", fill: "#999", r: 3, title: () => "Baseline level"}),
      n >= 5 ? Plot.ruleY(mine, {y: d => d.attr_label + ": " + d.short, x1: "lo", x2: "hi", stroke: "#d73027", strokeOpacity: 0.5}) : null,
      n >= 5 ? Plot.dot(mine, {x: "mine", y: d => d.attr_label + ": " + d.short, fill: "#d73027", r: 3.5,
        title: d => `You: ${d.mine.toFixed(2)}\nPublished: ${d.published.toFixed(2)}`}) : null,
      Plot.ruleY(pub, {y: d => d.attr_label + ": " + d.short, x1: d => d.published - 2 * d.published_se,
        x2: d => d.published + 2 * d.published_se, stroke: "#2166ac", strokeWidth: 2}),
      Plot.dot(pub, {x: "published", y: d => d.attr_label + ": " + d.short, fill: "#2166ac", r: 3.5,
        title: d => `Published: ${d.published.toFixed(3)}`})
    ]
  });
  return html`<div><p style="font-size:0.92em;color:#333;">${note}</p>${plot}</div>`;
}
```

Diamonds mark each attribute's baseline level, against which the others are
measured. Two things stand out after ten or twenty choices. Your own estimates
are noisy: the original respondents each made five choices, which is very
little information about one person, while the published estimates are precise
because they pool 1,396 people. And the published effects are modest next to
that noise. None moves the chance of admission by more than
`r round(max(abs(hhy_amce$published_amce)), 2)`, so learning how one
respondent's priorities differ from the average would take many more pairs than
a conjoint usually asks for. That point returns below.

### The published analysis reproduces from the IRW table

The IRW table `hainmueller_2014_immigrant` has one row per respondent, task
and profile, with the respondent's choice and the profile's attributes in
`attr_*` columns. It can be read without a Redivis account, from the same
link its [table page](https://itemresponsewarehouse.org/tables/hainmueller_2014_immigrant/)
offers:

```{r}
#| label: read-hhy
#| eval: false
#| code-fold: false

url <- paste0("https://redivis.com/api/v1/tables/datapages.irw_conjoint:v4_2.",
              "hainmueller_2014_immigrant/rows?format=csv")
hhy <- read.csv(url)
# with a Redivis account: irw::irw_fetch("hainmueller_2014_immigrant", source = "conj")
```

The AMCEs come from a linear probability model of the choice on every
attribute, with standard errors clustered by respondent. Two combinations
could never appear (the four high-skill professions without at least two years
of college, and fleeing persecution from the six countries other than China,
Sudan, Somalia and Iraq), so those two pairs of attributes get an interaction
and each level's effect is averaged over the levels of its partner where both
it and the baseline could occur. This is what the `cjoint` package does.

```{r}
#| label: hhy-model
#| eval: false

fit <- lm(choice ~ attr_education + attr_gender + attr_origin + attr_application_reason +
            attr_profession + attr_job_experience + attr_job_plans + attr_prior_trips +
            attr_language + attr_education:attr_profession +
            attr_origin:attr_application_reason, data = hhy)
V <- sandwich::vcovCL(fit, cluster = ~ id, type = "HC1")
# then, for a restricted attribute, AMCE = main effect + the mean of its
# interaction terms over the partner levels where it and the baseline can occur
```

```{r}
#| label: hhy-check
#| echo: false

hhy_diff <- max(abs(hhy_amce$irw_amce - hhy_amce$published_amce), na.rm = TRUE)
hhy_se_diff <- max(abs(hhy_amce$irw_se - hhy_amce$published_se), na.rm = TRUE)
hhy_dim_diff <- max(abs(hhy_amce$dim - hhy_amce$irw_amce), na.rm = TRUE)
```

All `r sum(!hhy_amce$baseline)` AMCEs and their standard errors match the
published estimates (computed by `cjoint` from the authors' replication data)
to `r if (max(hhy_diff, hhy_se_diff) < 1e-10) "at least ten decimal places" else signif(max(hhy_diff, hhy_se_diff), 1)`, that is, exactly. The
simple difference in choice rates the widget uses, applied to all 1,396
respondents, is within `r round(hhy_dim_diff, 3)` of every AMCE.

The rest of this page steps back from the average effects. Read as item
response data, each task is a paired comparison: two objects, one chosen. That
is the same shape as the [comparisons](intransitivity.qmd) elsewhere in IRW,
except that the objects are built at random rather than fixed (the [conjoint
standard](../conjoint_standard.qmd) describes the format). So we can ask two
predictive questions, scored with the InterModel Vigorish
[IMV, @domingue2025imv; @domingue2024imvlens], which the [IMV vignette](imv.qmd)
introduces:

1. **How much do the attributes explain?** How much better than a coin flip
   does a model of the attributes predict choices it was not fitted to?
2. **How much do people differ?** Given that model, how much do
   respondent-specific preferences add? And how does that compare with what a
   person's ability adds over item difficulties in ordinary item response
   data?

## Data and methods

### Dataset selection

The IRW conjoint source holds `r r_n_conj` tables in version 4.1, which this page uses. We use
those with a forced choice between exactly two profiles and at least 500
usable tasks, where a usable task is one in which exactly one of the two
profiles was chosen (this drops opt-out and unanswered tasks):

```{r}
#| label: filter-datasets
#| eval: false

meta <- read.csv(paste0("https://redivis.com/api/v1/tables/datapages.irw_meta:v41_0.",
                        "conj_metadata/rows?format=csv"))
candidates <- subset(meta, grepl("(^|;)choice(;|$)", outcomes) & n_profiles == 2 &
                           n_respondents * n_tasks >= 500)$table
```

`r r_all_candidates` tables met this screen as of
`r format(date_run, "%B %Y")`, `r if (r_n_tables_run == r_all_candidates) "all of them with" else paste(r_n_tables_run, "with")`
at least 500 usable tasks. Tables with more than 5,000 respondents were subsampled to 5,000 at
random.

### Per-table computation

For each table, the two profiles in a task are coded as attribute dummies and
the model works with their difference, so a task becomes one binary outcome:
was the first profile chosen? Attributes that were not shown on a profile get
a level of their own.

1. **Attributes.** An L2-penalised logistic regression of the choice on the
   difference in attribute dummies (penalty 1, intercept unpenalised) is fitted
   on four fifths of the respondents and predicts the other fifth, rotating
   over five folds. The baseline is a coin weighted to the training share of
   first-profile choices. The IMV compares the two.
2. **Respondent preferences.** Among respondents with at least three tasks,
   one task in five (by position, so each respondent keeps most of their tasks
   for fitting) is held out at a time. The shared model is the attribute model
   above, fitted to the training tasks. The individual model keeps it as an
   offset and gives each respondent their own deviation on every attribute term
   and on the intercept, shrunk towards zero by an L2 penalty $\lambda$. This
   is the posterior mode of a random-coefficient logit.
   $\lambda \in \{1, 3, 10, 30, 100, 300, 1000\}$ is chosen on a random 20% of
   respondents and the IMV of the individual over the shared model is reported
   on the other 80%. Tables need at least 200 such respondents.

   The two steps split the data differently on purpose. Attribute effects are
   meant to generalise to new people, so step 1 holds out respondents. Step 2
   asks how much a respondent's own earlier choices predict their later ones,
   the same question a person parameter in IRT answers when it is estimated on
   some of a person's items and scored on the rest, so it holds out tasks
   within respondents. Holding out whole respondents there would leave each new
   respondent with no training tasks, shrink every deviation to zero and make
   the individual model identical to the shared one: the IMV would be zero by
   construction. The held-out choice itself never enters the fit: the
   deviation is estimated from the respondent's training tasks only, the shared
   offset from training tasks only, and $\lambda$ on separate respondents.

```{r}
#| label: fit-function
#| eval: false

# Respondent i's deviation delta_i, given the shared model's linear predictor
# `off` on their training tasks Z (attribute differences plus a 1 for their own
# intercept). The penalised optimum lies in the span of Z's rows, so we solve for
# alpha with delta_i = t(Z) %*% alpha: a system the size of the number of tasks.
K <- tcrossprod(Z); a <- numeric(nrow(Z))
for (it in 1:50) {
  p <- plogis(off + drop(K %*% a)); W <- p * (1 - p)
  step <- solve(W * K + diag(lambda, nrow(Z)), y - p - lambda * a)
  a <- a + step
  if (max(abs(step)) < 1e-8) break
}
p_test <- plogis(off_test + drop(Z_test %*% t(Z) %*% a))
```

For the comparison with ordinary item response data we sampled
80 tables at random from the `r r_irt_pool` dichotomous core IRW
tables that are cross-sectional, at least 90% complete, have 300 to 20,000
respondents and 5 to 100 items, and applied the same design with items in place
of tasks: one fifth of each person's responses held out at a time, a shared
model of item difficulties (the logit of each item's training mean), and an
individual model adding a person ability with the same kind of penalty,
$\lambda \in \{0.3, 1, 3, 10, 30, 100\}$, chosen the same way. `r nrow(irt_df)`
of the 80 gave results; the other (`vocabulary_iq`) has six response values
although its metadata lists two.

::: {.callout-note}
To reproduce with current IRW holdings, re-run `conjoint_compute.R`
and commit the updated `conjoint_data/conjoint_results.rds`. Every table is read
without a login.
:::

## Results

### How much do the attributes explain?

```{r}
#| label: q1-numbers
#| echo: false

q1 <- summary_df |> filter(!is.na(imv_attr))
q1_med <- median(q1$imv_attr)
q1_w   <- median(q1$w_attr)
q1_rho <- cor(q1$imv_attr, q1$n_attr, method = "spearman")
```

Across `r nrow(q1)` experiments the median IMV of the attribute model over a
coin flip is `r round(q1_med, 2)`. In accuracy terms, the median attribute model
predicts a held-out choice like a coin that comes up right
`r round(100 * q1_w)`% of the time. Designs with more attributes are not more
predictable (Spearman $\rho$ = `r round(q1_rho, 2)` between the IMV and the
number of attributes).

```{r}
#| label: fig-q1
#| fig-cap: "Held-out IMV of the attribute model over a coin flip, one point per experiment, by the number of attributes in the design. Hover for the table."

p1 <- ggplot(q1, aes(n_attr, imv_attr,
                     text = paste0(table, "<br>IMV = ", round(imv_attr, 3),
                                   "<br>attributes: ", n_attr, "<br>usable tasks: ", n_tasks_used))) +
  geom_jitter(width = 0.15, height = 0, colour = irw_blue, alpha = 0.7, size = 2) +
  geom_hline(yintercept = q1_med, linetype = "dashed", colour = irw_grey) +
  labs(x = "Number of attributes", y = "IMV, attributes over a coin flip")
ggplotly(p1, tooltip = "text")
```

### How much do people differ?

```{r}
#| label: q2-numbers
#| echo: false

q2 <- summary_df |> filter(!is.na(imv_indiv))
q2_med  <- median(q2$imv_indiv)
irt_med <- median(irt_df$imv_indiv)
q2_share_small <- mean(q2$imv_indiv < 0.01)
q2_top <- q2 |> arrange(desc(imv_indiv)) |> slice_head(n = 5)
both <- bind_rows(
  q2 |> transmute(table, imv = imv_indiv, family = "Conjoint: respondent preferences",
                  n_per = mean_tasks, detail = paste0("tasks per respondent: ", round(mean_tasks, 1))),
  irt_df |> transmute(table, imv = imv_indiv, family = "Dichotomous IRT: person ability",
                      n_per = per_person, detail = paste0("items per person: ", round(per_person, 1))))
```

Respondent-specific preferences add much less. Over the shared attribute model
the median IMV is `r round(q2_med, 3)` across `r nrow(q2)` experiments, and in
`r round(100 * q2_share_small)`% of them it is below 0.01. In the
`r nrow(irt_df)` dichotomous IRT tables, knowing the person adds a median of
`r round(irt_med, 3)` over item difficulties alone.

```{r}
#| label: fig-q2
#| fig-cap: "Held-out IMV of the individual model over the shared model: respondent-specific preferences in conjoint experiments, person ability in dichotomous IRW tables. One point per table; the diamond marks the median."

meds <- both |> group_by(family) |> summarise(m = median(imv))
p2 <- ggplot(both, aes(imv, family, text = paste0(table, "<br>IMV = ", round(imv, 3), "<br>", detail))) +
  geom_jitter(aes(colour = family), height = 0.2, width = 0, alpha = 0.6, size = 2) +
  geom_point(data = meds, aes(m, family), inherit.aes = FALSE, shape = 23, size = 4, fill = "black") +
  geom_vline(xintercept = 0, colour = irw_grey) +
  scale_colour_manual(values = c(irw_red, irw_blue), guide = "none") +
  labs(x = "IMV, individual over shared model", y = NULL)
ggplotly(p2, tooltip = "text")
```

Most of the experiments where preferences add the most are ones where
respondents take sides: choosing between candidates who differ by party
(`titelman_2024_party_labels`, `mummolo_2021_partisan_loyalty`,
`carey_2020_democracy_public_oct`, `driscoll_2023_court_curbing`) and choosing
which side of a dispute to sympathize with (`inouye_2026_sympathies_2023` and
`_2025`). There an attribute's average effect mixes respondents pulling in
opposite directions, and knowing which way a given respondent leans helps
predict their next choice. The exceptions near the top are purchases:
`zou_2020_ev_purchase` (new and used electric cars) and
`gan_2026_alternative_meat` (plant-based and cultivated-meat burgers), where
buyers weigh the same features differently.

```{r}
#| label: tbl-q2-top
#| echo: false

q2_top |>
  transmute(Table = table, `IMV, preferences` = round(imv_indiv, 3),
            `IMV, attributes` = round(imv_attr, 3), Respondents = n_resp,
            `Tasks per respondent` = round(mean_tasks, 1)) |>
  knitr::kable()
```

Put together: in a typical conjoint experiment the attributes do a fair amount
of work, and the average effect captures most of what can be learned about how
a given respondent will choose. For person-level prediction, a conjoint's
handful of tasks per respondent buys less than a test's items do.

## Limitations

::: {.callout-important}
**Draft, for review.** This section was written by Claude and has not yet been
checked by the IRW team.
:::

- **Few tasks per respondent.** A typical conjoint gives each respondent about
  five tasks, against a dozen or more terms for their preferences; a test gives
  a person a dozen or more items for one ability. The preference IMV measures
  how much can be learned about a respondent from the tasks they did, not how
  much respondents truly differ, and it rises with the number of tasks
  (Spearman $\rho$ = `r round(cor(q2$imv_indiv, q2$mean_tasks, method = "spearman"), 2)`).
  The comparison with IRT is between what each kind of design lets you learn
  about a person, not between quantities of the same size.
- **One model of preferences.** The individual model is additive in the
  attributes with one isotropic penalty. Preferences that are structured (for
  example, a respondent's party driving several attributes at once) would be
  captured better by respondent covariates or a low-rank model, and would
  likely add more.
- **Noise in choosing the penalty.** $\lambda$ is chosen on one fifth of
  respondents and scored on the rest, so in a few tables the chosen value fits
  the scoring respondents worse than the shared model does:
  `r sum(q2$imv_indiv < 0)` conjoint and `r sum(irt_df$imv_indiv < 0)` IRT tables
  have a slightly negative IMV. These are left in, since dropping them would
  bias both medians upward.
- **Repeated pairs help.** Some designs show a respondent the same pair twice
  (the Inouye experiments repeat their first pair, flipped, as the eleventh). A
  respondent-specific model can use that, so those tables partly measure
  consistency rather than preference.
- **Held-out tasks by position.** Conjoint tasks are held out by their position
  in the sequence and IRT items at random, so order effects (fatigue, learning)
  enter the conjoint comparison slightly differently.
- **Forced choices only.** Ratings, opt-out tasks and designs with one or three
  profiles are left out; so are tables with fewer than 500 usable tasks.
- **A random IRT sample.** The `r nrow(irt_df)` IRT tables are a random draw from
  dichotomous core tables, not matched to the conjoint experiments by sample,
  domain or length.
- **One version of the source.** Everything here is `irw_conjoint` v4.2 and
  `irw_conjoint_2` v1.0 (849 tables in all). Later versions will add
  experiments; rerunning means changing the version strings at the top of
  `conjoint_compute.R`.
- **One table left out.** `joo_2026_descriptive_rep` meets the criteria but is
  excluded: in its deposit every respondent's chosen candidate has the same
  position in all six tasks, so its outcomes do not line up with its profiles.
- **The widget.** Its pairs are a random 400 of the 6,980 tasks, its estimator
  is a simple difference in choice rates, and the published respondents are a
  2012 sample of U.S. adults.

## Reproducibility

{{< include ../components/_source-links.qmd >}}

{{< include ../components/_data-version.qmd >}}

Results were computed on `r format(date_run, "%B %d, %Y")` from
`irw_conjoint` v4.2, `irw_conjoint_2` v1.0 and `irw_meta` v41.0. To regenerate:

```r
# 1. Re-run the compute script (from the project root)
source("vignettes/conjoint_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/conjoint.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(summary_df$table, source = "conj", output_file = "conjoint_data/references.bib")
```
:::

---

## References
