---
title: "Bounded Continuous Responses: Comparing Beta IRT, Samejima's CRM, and Mueller's CRSM"
description: |
  Slider and visual-analogue-scale items produce a response that is
  continuous but lives in a fixed, interpretable range, like [0,1] or
  [0,100] -- not the open-ended continuous case standard CFA/linear-factor
  models assume. We compare four published ways of modeling that: a naive
  linear baseline that ignores the bounds, Noel and Dauvier's Beta IRT
  model, Samejima's Continuous Response Model, and Mueller's Continuous
  Rating Scale Model, on one simulated dataset and one real IRW table.
date: today
bibliography:
  - continuous_bounded_data/references.bib
  - continuous_bounded_data/irw_references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`continuous_bounded_compute.R`) directly if
you're relying on the numbers here.
:::

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 13))
irw_blue   <- "#2166ac"
irw_red    <- "#d73027"
irw_green  <- "#4daf4a"
irw_purple <- "#984ea3"
irw_grey   <- "#808080"

model_colors <- c(
  "naive_linear"              = irw_grey,
  "beta_irt_item_dispersion"  = irw_blue,
  "beta_irt_fixed_dispersion" = irw_purple,
  "samejima_crm"              = irw_red,
  "muller_corsm"              = irw_green
)
model_labels <- c(
  "naive_linear"              = "Naive linear",
  "beta_irt_item_dispersion"  = "Beta IRT (item disp.)",
  "beta_irt_fixed_dispersion" = "Beta IRT (fixed disp.)",
  "samejima_crm"              = "Samejima CRM",
  "muller_corsm"               = "Mueller CRSM"
)

cache_path <- "continuous_bounded_data/continuous_bounded_results.rds"
cache_ok <- file.exists(cache_path)
if (cache_ok) res <- readRDS(cache_path)

grid_cache_path <- "continuous_bounded_data/continuous_bounded_grid_results.rds"
grid_cache_ok <- file.exists(grid_cache_path)
if (grid_cache_ok) grid_res <- readRDS(grid_cache_path)
```

## Motivation

Visual analogue scales, slider ratings, and continuous confidence or
percentage judgments are common: a respondent marks a point on a
fixed, labeled segment (0 to 100, say), and the raw response is a
continuous number that is *bounded* at both ends. This is a different
measurement problem from the unbounded continuous case (response times,
open-ended magnitude estimates), which reduces to standard linear-factor
or CFA modeling and is covered elsewhere on this site -- the bounds
themselves are the interesting part here, because they create floor/ceiling
effects and non-constant response variance near the edges that a plain
linear model doesn't see.

Three published models handle this directly. Noel and Dauvier's Beta IRT
model [@noel2007beta] treats the response as Beta-distributed given
ability, with the two natural Beta parameters reparameterized into an
item difficulty and a dispersion. Samejima's Continuous Response Model
[@samejima1973homogeneous] models a monotonic transform of the response as
normal given ability -- the historically original approach, predating Beta
IRT by over 30 years. Mueller's Continuous Rating Scale Model
[@mueller1987rasch] takes a Rasch-family, specific-objectivity approach,
estimated by pairwise conditional likelihood; Verhelst's (2019) exponential
family framework [@verhelst2019exponential] is the modern theoretical
generalization of Mueller's model, though the model actually fitted here is
Mueller's original. We compare these three against a naive linear (Gaussian,
identity-link) baseline that ignores the bounds entirely -- the "what if you
just treat it like ordinary continuous data" control, playing the same role
the Rasch baseline plays in the site's guessing-models vignette.

## Data and methods

### The four core specs

1. **Naive linear baseline** -- single-factor CFA via `lavaan`, identity
   link, Gaussian residual, fit directly on the raw (rescaled-to-[0,1])
   response. Ignores boundedness entirely.
2. **Beta IRT** [@noel2007beta] -- fit as a discretized custom `mirt`
   itemtype wrapping `sirt::brm.irf()`, following `sirt`'s own documented
   mirt-estimation recipe (`?sirt::brm.sim`). Two variants: **item-varying
   dispersion** (each item gets its own dispersion parameter, the model's
   default parameterization) and **fixed dispersion** (one dispersion
   parameter shared across all items, via an `mirt` parameter constraint).
3. **Samejima's Continuous Response Model** [@samejima1973homogeneous] --
   `EstCRM::EstCRMitem()` / `EstCRMperson()`.
4. **Mueller's Continuous Rating Scale Model** [@mueller1987rasch] -- a
   patched copy of `pcIRT::CRSM()` / a from-scratch reimplementation of its
   person-scoring equation (see [Limitations](#limitations) for why
   patching was necessary).

Ferrando's (2001) nonlinear congeneric model [@ferrando2001nonlinear] is an
optional stretch spec with no package implementation (it would need a
custom truncated-normal likelihood via `optim`) and has not been attempted
here -- see [Limitations](#limitations).

### Comparing the item response functions across models

Before fitting anything to data, it helps to see what each model's own
equation implies about the relationship between a respondent's ability
$\theta$ and their expected response. All three bounded-response models
share a common building block -- a linear "ability minus difficulty" term,
$a(\theta-b)$ -- but wire it into the response very differently. Drag the
sliders below to see how difficulty ($b$) and discrimination ($a$) reshape
each model's mean response curve.

```{ojs}
//| echo: false
viewof irf_b = Inputs.range([-2, 2], {value: 0, step: 0.1, label: "Item difficulty (b)"})
```

```{ojs}
//| echo: false
viewof irf_a = Inputs.range([0.3, 2.5], {value: 1, step: 0.1, label: "Discrimination (a) -- no effect on Mueller CRSM"})
```

```{ojs}
//| echo: false

function normalPDF(z, mu, sigma) {
  return Math.exp(-0.5 * ((z - mu) / sigma) ** 2) / (sigma * Math.sqrt(2 * Math.PI));
}

function simpson(f, lo, hi, n) {
  const h = (hi - lo) / n;
  let s = f(lo) + f(hi);
  for (let i = 1; i < n; i++) {
    s += (i % 2 === 0 ? 2 : 4) * f(lo + i * h);
  }
  return (s * h) / 3;
}

irf_sigma = 1
irf_lambda = 2

function betaIrtMean(theta, a, b) {
  return 1 / (1 + Math.exp(-a * (theta - b)));
}

function samejimaMean(theta, a, b, sigma) {
  const loc = a * (theta - b);
  return simpson((z) => (1 / (1 + Math.exp(-z))) * normalPDF(z, loc, sigma), loc - 6 * sigma, loc + 6 * sigma, 60);
}

function muellerMean(theta, b, lambda) {
  const g = (x) => Math.exp(x * (theta - b) + x * (1 - x) * lambda);
  const Z = simpson(g, 0, 1, 100);
  return simpson((x) => x * g(x), 0, 1, 100) / Z;
}

irf_thetas = d3.range(-3, 3.01, 0.1)

irf_curve_data = irf_thetas.flatMap((theta) => [
  { theta, model: "naive_linear", y: 0.5 + irf_a * (theta - irf_b) },
  { theta, model: "beta_irt", y: betaIrtMean(theta, irf_a, irf_b) },
  { theta, model: "samejima_crm", y: samejimaMean(theta, irf_a, irf_b, irf_sigma) },
  { theta, model: "muller_corsm", y: muellerMean(theta, irf_b, irf_lambda) }
])
```

```{ojs}
//| echo: false

irf_colors = ({
  naive_linear: "#808080",
  beta_irt: "#2166ac",
  samejima_crm: "#d73027",
  muller_corsm: "#4daf4a"
})

irf_labels = ({
  naive_linear: "Naive linear",
  beta_irt: "Beta IRT",
  samejima_crm: "Samejima CRM",
  muller_corsm: "Mueller CRSM"
})

Plot.plot({
  width: 700,
  height: 420,
  x: {label: "Ability (θ)", domain: [-3, 3]},
  y: {label: "Expected response (rescaled to [0,1])", domain: [-0.15, 1.15]},
  color: {
    domain: Object.keys(irf_colors),
    range: Object.values(irf_colors),
    legend: true,
    tickFormat: (d) => irf_labels[d]
  },
  marks: [
    Plot.ruleY([0, 1], {stroke: "#999", strokeDasharray: "2,2"}),
    Plot.line(irf_curve_data.filter((d) => d.model === "naive_linear"), {
      x: "theta", y: "y", stroke: "model", strokeDasharray: "4,3"
    }),
    Plot.line(irf_curve_data.filter((d) => d.model !== "naive_linear"), {
      x: "theta", y: "y", stroke: "model"
    })
  ]
})
```

The naive linear line (dashed) is the only curve not actually confined to
$[0,1]$: away from $\theta=b$ it predicts responses outside the possible
range entirely, which is the exact failure mode the other three models are
built to avoid by leveling off smoothly at the boundary instead. Mueller's
CRSM has no discrimination parameter at all -- being a Rasch-family,
specific-objectivity model [@mueller1987rasch], moving the $a$ slider only
changes the other three curves. Dispersion parameters are held fixed here
at representative values ($\sigma=1$ for Samejima CRM, $\lambda=2$ for
Mueller CRSM) rather than exposed as sliders: Beta IRT's mean curve doesn't
depend on its dispersion parameter at all (only its spread does), while
Samejima's and Mueller's dispersion parameters do shift the mean curve
itself once backtransformed through a nonlinear link, not just its spread.

### Estimation and model comparison

Every model's held-out fit is reported as the **log-probability of the
response falling in a small interval**, not a raw point density. This
choice exists because the 4 models don't report likelihood on the same
scale to begin with: reading `EstCRM::EstCRMitem()`'s source shows it
transforms the raw response to $Z = \ln\!\big(X/(K-X)\big)$ and computes
its entire EM log-likelihood on that transformed scale, not the original
response scale the Beta IRT and naive linear fits use. Comparing those two
log-likelihoods directly would be comparing different units -- like
comparing an error reported in inches to one reported in centimeters
without converting first. The usual fix is a model-specific Jacobian
correction; instead, every model's held-out score here is
$\log P(\text{bin}_{lo} < X < \text{bin}_{hi})$, using the same bin edges
(on the raw, rescaled-to-$[0,1]$ scale) for all 4 models. A probability that
a response lands in a fixed interval means the same thing regardless of
which internal scale a model uses to compute it, so this sidesteps the
rescaling problem entirely rather than requiring a per-model correction.

Held-out log-probability, built this way, is the metric used to compare all
4 specs throughout this vignette. It's a standard, interpretable scale for
predictive fit, though not a validated best-practice specific to this
bounded-continuous setting -- flagged as an ad hoc choice worth revisiting
(see [Limitations](#limitations)).

### Held-out design: person-level, not cell-level

The project's usual convention for held-out evaluation on IRW data is a
response-level (missing-cell) holdout: mask a random subset of person x item
cells, fit on the rest, score the masked cells. That isn't viable here for
2 of the 4 specs -- reading `EstCRM::EstCRMitem()` and `pcIRT::CRSM()`'s
fitting code directly shows **neither has any missing-data handling at
all** (both do matrix/pairwise algebra that silently breaks with `NA`
cells, confirmed by testing).

We substituted a **person-level train/test split** instead: fit item
parameters on a training subset of persons (a complete matrix, no missing
cells), then estimate ability for held-out test persons (from their own
full response vector, which none of the 4 fitting calls ever saw) and score
each model's implied probability for those same responses. This is a
standard "do item parameters generalize to new examinees" cross-validation
design, but it's different, and weaker, evidence than true cell-level
held-out-response prediction. Flagged here as a deviation driven by package
constraints, not a silent substitution.

### Simulation

**A single illustrative condition** ($N=300$, $J=10$, Beta IRT DGP, fixed
dispersion) is shown first below, generated directly from the Beta IRT
model's own equations (`sim_beta_irt()` in `continuous_bounded_helpers.R`
-- a bug-fixed reimplementation of `sirt::brm.sim()`'s logic; see
[Limitations](#limitations) for why the reimplementation was necessary).

**The full recovery/misspecification grid** cross-fits all 4 core specs
against 3 DGPs, generated from each model's own equations: Beta IRT
(`sim_beta_irt()`), Samejima CRM (`sim_samejima_crm()`: $Z_{ij} =
\alpha_j(\theta_i - b_j) + e_{ij}$, $e_{ij}\sim N(0,\sigma_j^2)$, then
$X_{ij}=1/(1+e^{-Z_{ij}})$ -- the exact inverse of EstCRM's own $Z=\ln(X/(1-X))$
transform), and Mueller CRSM (`sim_muller_corsm()`: grid-based inverse-CDF
sampling from the same $\exp[x(\theta-b)+x(1-x)\lambda]$ density used for
scoring, since no closed-form sampler exists for this family). The
naive-linear DGP is skipped: an unbounded-Gaussian DGP clipped to $[0,1]$ to
make it "bounded" would introduce an artificial boundary spike
unrepresentative of what a linear model actually generates.

Grid: $N\in\{300,1000\}$, $J\in\{8,20\}$, and, for the Beta IRT DGP
specifically, fixed vs. item-varying dispersion -- 8 + 4 + 4 = 16
conditions, plus 1 boundary-inflation condition (10% of responses forced to
exactly 0, 10% forced to exactly 1, layered on the Beta IRT DGP at
$N=300$/$J=8$) for 17 total. Run via `furrr::future_pmap()` +
`plan(multisession)`, cached separately from the real-data results below
(own skip-if-exists gate in the compute script).

### Real-data selection

#### Finding candidate tables: the scouting process

IRW has no field that directly flags "this table's responses are bounded
continuous" -- that has to be inferred. The scouting process that produced
the candidate list lives entirely in this repo, **not** in the `irw`
R package (`irw_metadata()` / `irw_fetch()` / `irw_tags()` are the package
functions it's built on top of; the scout scripts and their output are
project-local artifacts, not something `library(irw)` gives you). Anyone
extending this vignette to more tables, or adapting the approach for a
different response format, should start from these files directly rather
than re-deriving the logic:

- `vignettes/continuous_scout_fetch.R` -- the expensive stage.
- `vignettes/continuous_scout_classify.R` -- the cheap, rerun-freely stage.
- Cached raw output: `vignettes/continuous_scout_data/continuous_scout_raw.rds`.
- Final table: **`continuous_scout_candidates.csv`, at the repo root**
  (not inside a `vignettes/` subfolder, and not published anywhere --
  regenerate it locally with the two scripts above if you don't have it).

The filtering logic is more than a single `range(resp)` check, precisely
because raw min/max alone can't distinguish a genuine fixed-format scale
from a derived/open-ended score. It runs in three steps:

1. **Cheap metadata filter, no fetching.** Keep tables where
   `irw_metadata()$n_categories >= 12` -- the threshold this site's other
   summaries already use as the boundary of "cleanly ordinal" (see
   `dimensionality.qmd`, `local_dependence_compute.R`). This is exactly
   where many-category and continuous candidates both live, so it's a
   cheap way to shortlist before touching Redivis at all.
2. **Fetch and summarize.** For every table that survives step 1, fetch
   `resp` and compute, per table, `n_unique`/`min_resp`/`max_resp`/
   `range_resp`/`pct_noninteger`, and, per item within the table,
   `item_min`/`item_max`/`item_n`/`pct_noninteger`.
3. **Classify from two independent signals**, combined so that agreement
   is stronger evidence than either alone and disagreement gets downgraded
   to `unclear` rather than resolved automatically:
   - *resp-based*: are the table's observed min/max at a "round,"
     interpretable endpoint (0, 1, 5, 10, 100, ...), **and** do at least
     70% of items individually come within 15% of the table's overall
     range at both ends? That item-consistency check matters --  an
     earlier version compared raw per-item extremes directly and
     misclassified clean tables purely from small-sample noise (e.g. an
     11-item 0-100 VAS table where 9 items hit exactly [0,100] but 2
     items topped out at 92/98 got flagged "inconsistent"). Checking
     whether each item's range comes *within a tolerance* of the table's
     range, instead, tolerates that kind of sampling noise while still
     catching an item on a genuinely different scale (e.g. a 1-9 item
     mixed into a 0-100 table).
   - *metadata-based*: IRW's own curated `item_format` tag from
     `irw_tags()` -- `"Slider/continuous"` is direct evidence for bounded,
     `"Constructed Response"` for unbounded.

Of the 55 tables with $\geq$12 categories, this produced 24 `bounded`, 18
`unbounded`, 13 `unclear` (full breakdown and per-table notes in
`vignettes/continuous_scout_data/continuous_scout_handoff.md`).

#### From 24 candidates to the 7 tables analyzed

Not every scouted candidate is usable as-is. Two structural checks rule
tables out regardless of anything else: an **ipsative/compositional**
constraint (row sums across items are a constant, e.g. a fixed
point-allocation task), which violates the conditional-independence
assumption every one of these 4 models makes; and **non-unique
`(id, item)` pairs**, which usually means the table holds multiple
measurement occasions per item, not a single-occasion battery, and needs
reducing to one occasion before analysis -- IRW's own data standard
(`standard.qmd`) documents `wave` (repeated administrations) and `rater`
(multiple raters per unit) as the columns that explain this in practice,
and several tables that looked unusable at first check were reclaimed after
filtering to a single wave/rater.

Applying those checks, plus a by-hand look at fractional-value evidence (a
table classified `bounded` from its min/max endpoints can still turn out to
be 100% integer-valued on inspection -- weaker evidence of genuine
continuity than a fine-grained scale, since classification from metadata
alone can't tell "true VAS" from "a very fine 0-100 integer rating"),
narrowed the 24 `bounded` candidates to the 7 tables used for every
real-data result below:

| Table | N | J | Note |
|---|---|---|---|
| `lsbq_maleki_2025_non_persian_proficiency` | 308 | 3 | 0-10 scale, ~40% non-integer responses -- primary evidence of genuine continuity |
| `ai_fear_dong_2026_ai` (Doctor facet, subsampled) | 2,000 | 8 | 0-100, integer-valued; large-*N* structural check |
| `much_tte_2025_currentmotivation` | 1,244 | 3 | wave-filtered to one occasion |
| `test_taking_much_2025_cm` | 1,244 | 9 | -- |
| `nas_rogoza_2024_study5_nas` | 277 | 4 | wave-filtered; Narcissistic Antagonism subscale |
| `nas_rogoza_2024_study5_ngs` | 274 | 4 | wave-filtered; Narcissistic Grandiosity subscale |
| `nas_rogoza_2024_study5_nvs` | 278 | 4 | wave-filtered; Narcissistic Vulnerability subscale |

Tables left out along the way fell into a few categories: ipsative/
point-allocation formats, item ranges that vary within a table, formats
mixed within a single table, too few items to fit an IRT model, and
repeated-measures structure too irregular to reduce to a single
wave/rater. Two tables that pass every structural check
(`westhoff2023_{stopd,pbat}`, `gilbert_meta_95`) hit an unresolved Beta IRT
scoring issue instead, described in [Limitations](#limitations). Full
per-table vetting code and output, including the exact reason for every
excluded table, live in `vignettes/continuous_bounded_data/vet_candidates.R`
/ `vet_results.csv` and `vet_candidates2.R` / `vet_results2.csv`.

Two of the 7 needed an additional judgment call beyond the structural
checks above. `ai_fear_dong_2026_ai` is a 48-item, 6-target x 8-trait
factorial design (Care/Doctor/Journalist/Judge/Manager/Religious x
competent/warm/etc.), not one construct, so we used one target's 8 trait
items (`Doctor_*`) as a single "overall impression of Doctor" construct --
trait ratings like these reliably load on one general "halo" factor even
when they nominally span separate facets, but it's a real
construct-validity call, not something the scouting script could resolve
automatically -- and it was subsampled from N=10,000 to N=2,000 for compute
tractability. `lsbq_maleki_2025_non_persian_proficiency` needed no such
call; it's the scout's own top pick and the only one of the 7 with
substantial fractional-value evidence.

## Results

### Simulated data: does the correctly-specified model win, and does it recover parameters?

```{r}
#| label: fig-sim-scores
if (cache_ok) {
  res$sim$scores |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = sprintf("%.2f", mean_ll)), vjust = 1.4, color = "white", size = 3.5) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = "Simulated Beta-IRT data: held-out fit by model") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

On data simulated directly from the Beta IRT model, the two Beta IRT
variants have the best (least negative) held-out fit, essentially tied with
Samejima CRM and Mueller CRSM, and all three clearly beat the naive linear
baseline. This is the expected pattern for a correctly-specified-model
recovery check: the models built for bounded continuous responses recover
the truth better than the one that ignores the bounds.

Item-difficulty recovery for the item-varying-dispersion Beta IRT fit
(correlation between estimated and true $\delta$) was
`r if (cache_ok) sprintf("%.3f", res$sim$recovery_cor_delta) else "NA"`.

### The full recovery grid: 16 conditions x 3 DGPs

Parameter recovery for the correctly-specified model was excellent
everywhere -- every one of the 16 non-inflation conditions recovered its
true item parameters at $r>0.96$, most above $r>0.99$:

```{r}
#| label: tbl-grid-recovery
if (grid_cache_ok) {
  grid_res$scores |>
    filter(!boundary_inflation) |>
    distinct(dgp, N, J, dispersion_mode, recovery_cor) |>
    arrange(dgp, N, J, dispersion_mode) |>
    mutate(recovery_cor = sprintf("%.3f", recovery_cor)) |>
    knitr::kable(col.names = c("DGP", "N", "J", "Beta IRT dispersion", "Recovery r"))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

More useful than "does recovery work" (it does, everywhere) is the
cross-fitting question: when data comes from one DGP, which of the 4
*fitted* specs wins the held-out comparison?

```{r}
#| label: fig-grid-crossfit
if (grid_cache_ok) {
  grid_res$scores |>
    filter(!boundary_inflation) |>
    mutate(model = factor(model, levels = names(model_colors)),
           dgp_label = recode(dgp, beta_irt = "DGP: Beta IRT", samejima_crm = "DGP: Samejima CRM",
                               muller_corsm = "DGP: Mueller CRSM")) |>
    group_by(dgp_label, model) |>
    summarise(mean_ll = mean(mean_ll, na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    facet_wrap(~dgp_label, ncol = 3) +
    labs(x = NULL, y = "Mean held-out log-probability\n(averaged across N/J conditions)",
         title = "Cross-fitting: held-out fit by model, averaged across all N/J conditions per DGP") +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

The pattern is consistent across all 3 DGPs and matches every real-data
result above: **the 3 bounded-response models cluster tightly together and
substantially outperform the naive linear baseline, regardless of which of
the 3 actually generated the data.** The correctly-specified model is
usually numerically best or tied for best (Samejima CRM wins clearly under
its own DGP; Mueller CRSM wins clearly under its own DGP), but the margin
between the 3 bounded-response models is small compared to the gap to naive
linear -- these 3 models are more interchangeable with each other than any
of them is with a model that ignores the bounds. One fit failed outright:
Samejima CRM's EM did not converge (a `missing value where TRUE/FALSE
needed` error partway through) on one Mueller-CRSM-generated dataset
($N=300$, $J=8$) -- a real misspecification-robustness finding, not
papered over; see [Limitations](#limitations).

### Boundary inflation: does floor/ceiling piling break anything?

Beyond fitting, we asked how each spec's held-out fit *degrades* when 20%
of responses are forced to the exact boundary (10% to 0, 10% to 1) on top
of the Beta IRT DGP -- diagnostic information, not a 5th model to fit:

```{r}
#| label: fig-grid-inflation
if (grid_cache_ok) {
  grid_res$scores |>
    filter(dgp == "beta_irt", N == 300, J == 8, dispersion_mode == "fixed") |>
    mutate(model = factor(model, levels = names(model_colors)),
           condition = ifelse(boundary_inflation, "20% boundary-inflated", "Not inflated")) |>
    ggplot(aes(x = model, y = mean_ll, fill = model, alpha = condition)) +
    geom_col(position = "dodge", width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_alpha_manual(values = c("Not inflated" = 1, "20% boundary-inflated" = 0.55), name = NULL) +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability",
         title = "Beta IRT DGP (N=300, J=8): held-out fit with vs. without boundary inflation") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

Every spec's held-out fit gets worse under inflation -- none of the 4 core
specs explicitly models a zero-one-inflated mixture, so this is expected --
but **not by the same amount**. The naive linear baseline degrades by
about 23% (mean log-probability drops from -1.57 to -1.93); the 3
bounded-response models degrade by roughly 15-16% each (e.g. Mueller CRSM:
-1.25 to -1.43). The bounded-response models' smooth density still puts
more mass near the boundary than a Gaussian does, so they absorb some of
the extra piling without a dedicated inflation term, while the naive linear
model has nothing in its structure to absorb it at all. None of the 4
handle it *well* -- a true zero-one-inflated extension (Molenaar et al.,
2022) would be needed for that -- but the bounded-response models are
comparatively more robust to it than the baseline that ignores boundedness
altogether.

### Real data: does floor/ceiling piling look the way the models expect?

Before comparing model fits, it helps to see what these tables' response
distributions actually look like. This is the primary table's ~40%
non-integer evidence from [Real-data selection](#real-data-selection) made
visual, and it's also the direct reason the naive linear baseline
underperforms below: a Gaussian model expects mass concentrated
symmetrically around the mean, not piled up against a hard boundary.

```{r}
#| label: fig-real-dist
if (cache_ok) {
  as.data.frame(res$real$Y_raw) |>
    tidyr::pivot_longer(everything(), names_to = "item", values_to = "resp") |>
    ggplot(aes(x = resp)) +
    geom_histogram(binwidth = 0.5, boundary = 0, fill = irw_blue, color = "white", linewidth = 0.1) +
    facet_wrap(~item, ncol = 3) +
    labs(x = sprintf("Raw response (%g-%g scale)", res$real$low, res$real$high),
         y = "Count",
         title = sprintf("%s: raw response distribution by item", res$real$table))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

All 3 items show a similar shape: a broad hump across the mid-range plus a
sharp spike at the upper boundary. Between 23% and 44% of responses per
item sit at *exactly* 10, the scale ceiling (vs. under 9% at exactly 0) --
respondents who are, or report being, fully proficient pile up exactly at
the boundary, which a symmetric Gaussian around a single mean cannot
represent. The same ceiling pile-up shows up, to varying degrees, in the
other 6 tables analyzed below -- it's the general floor/ceiling behavior
the 3 bounded-response models are built to handle and the naive linear
model is not.

### Held-out fit across all 7 real tables

```{r}
#| label: prep-real-scores
if (cache_ok) {
  all_real_scores <- dplyr::bind_rows(
    res$real$scores |> dplyr::mutate(N = res$real$N, J = res$real$J),
    res$real2$scores |> dplyr::mutate(
      N = res$real2$N, J = res$real2$J,
      label = paste0(label, " (", res$real2$facet, ")")
    ),
    res$batch$scores
  )
}
```

```{r}
#| label: fig-real-all-scores
if (cache_ok) {
  all_real_scores |>
    mutate(model = factor(model, levels = names(model_colors)),
           label = sprintf("%s\n(N=%d, J=%d)", label, N, J)) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    facet_wrap(~label, ncol = 4) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = "Held-out fit by model, across all 7 real IRW tables analyzed") +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

The pattern is consistent across all 7 tables and matches the simulation
results above: the naive linear baseline is clearly worst everywhere, and
the 3 bounded-response models are tightly clustered together. No single
bounded-response spec wins consistently -- Samejima CRM comes out slightly
ahead on the primary `lsbq_maleki_2025_non_persian_proficiency` table,
while Mueller CRSM edges ahead on `ai_fear_dong_2026_ai` (Doctor facet) and
on `test_taking_much_2025_cm` and 2 of the 3 `nas_rogoza` subscales -- but
the gaps between the 3 bounded-response models are small throughout, and
each table has only one train/test split (see
[Limitations](#limitations)). The main takeaway is qualitative, not "which
of the 3 wins": all 3 bounded-response models consistently and
substantially beat the naive linear baseline across every real table
analyzed.

## Limitations {#limitations}

Known gaps this vignette does not yet address:

- **Ferrando's (2001) nonlinear congeneric model** [@ferrando2001nonlinear]
  -- the optional stretch spec, comparable to the treatment in Ferrando
  (2002) [@ferrando2002theoretical] -- has not been attempted at all. Specs
  1-4 are now complete and validated (the grid ran cleanly), so its own
  sequencing rule no longer blocks it -- it just hasn't been picked up yet.
- **Partial missingness in real data.** All 7 real tables used complete
  cases only; IRW tables with partially missing item responses aren't
  handled yet.
- **Only one train/test split per condition** (real or simulated), not a
  repeated/cross-validated design -- so none of the held-out numbers above
  have an uncertainty band, and a single unlucky split could shift a close
  call between the 3 bounded-response models.
- **One grid fit failed outright**: Samejima CRM's EM did not converge
  (`missing value where TRUE/FALSE needed`) on one Mueller-CRSM-generated
  condition ($N=300$, $J=8$, no boundary inflation). Not re-run with a
  different seed or diagnosed further -- reported as a real
  misspecification-robustness finding (EstCRM's EM can fail outright, not
  just fit poorly, on data far from its assumptions) rather than treated as
  a bug to route around.
- **The grid's boundary-inflation condition covers only 1 of the 3 DGPs**
  (Beta IRT, $N=300$/$J=8$) -- not repeated for Samejima CRM or Mueller
  CoRSM, or at other N/J combinations.

**Package-availability and correctness issues found while building this**
(worth knowing before extending this vignette):

- `pcIRT::CRSM()` (Mueller's model) was **dropped from the package
  entirely** in the last CRAN release (0.2.4, 2019); it only exists in
  pcIRT $\leq$ 0.2.3. `pcIRT` itself was later archived by CRAN entirely
  (2025-11-11). This vignette's compute script installs 0.2.3 from the
  CRAN archive, not the current CRAN release -- a real risk for
  reproducing this vignette in the future if the archive changes.
- That same version's `CRSM()` has a convergence-check bug
  (`is.na(para) || ...` where `para` becomes a 2-vector after the first
  iteration) that is a hard error under R $\geq$ 4.3 and had to be patched
  locally before it would run at all.
- `sirt::brm.sim()` hardcodes a 9-item assumption in its column-naming
  code and errors for any other item count; `sim_beta_irt()` in the
  helpers file reimplements its generative logic directly rather than
  patching `sirt`.
- `pcIRT::person_par.CRSM()` only scores the same persons used in the
  original fit (no missing-data handling, no new-data argument) and could
  not be reused for the person-level holdout design above; a generalized
  version of its estimating equation was written from scratch instead.
- The from-scratch Mueller CRSM person-scoring routine
  (`theta_new_muller_corsm()`) could diverge to a `theta` extreme enough to
  overflow `exp()` inside `integrate()`, which throws a hard "non-finite
  function value" error rather than the graceful NA `stop.on.error=FALSE`
  is supposed to allow -- this crashed the *entire* held-out scoring pass
  for a table (not just Mueller CRSM) before being caught and fixed by
  clamping `theta` and wrapping each `integrate()` call individually.
- Similarly, a Beta IRT scoring-stage failure (see below) was originally
  written so it could crash naive-linear/Samejima/Mueller's results for the
  same table too -- every model's fit-and-score block is now wrapped in its
  own `tryCatch` so one model's failure can't take down the other 3.
- **Unresolved**: `westhoff2023_{stopd,pbat}` and `gilbert_meta_95` pass
  every structural check (single wave, unique `(id, item)`, no ipsative
  constraint) but still fail at the Beta IRT scoring step -- `mirt`
  compacts a training split's observed categories to consecutive integers
  per item, and if a held-out test person's response lands in a category
  with zero training-split observations, `fscores(response.pattern=...)`
  rejects it. Lowering the shared bin count from 12 to 6 fixed this for
  every other table here but not these 3 (`pbat` also fails
  Samejima CRM fitting with an unrelated "missing value where TRUE/FALSE
  needed" error, not yet diagnosed). Left out of the real-data set rather
  than chasing further.

**Explicitly out of scope** (not implemented, not planned for this
vignette): DIF for continuous responses [@finch2023impact]; dual-range
slider / interval-response models and the Interval Consensus Model
[@kloft2024measuring; @kloft2023dirichlet; @kloft2025interval] -- a
different response format (an interval, not a point) and, for the
Consensus Model, a different modeling paradigm (cultural consensus
theory) entirely; Zopluoglu's mixture CRM [@zopluoglu2020finite] and
Noel's beta unfolding model [@noel2014beta] -- each adds a full second axis
of complexity (latent mixture classes; unfolding vs. dominance response
process), and unfolding in particular is a more natural extension of the
site's existing unfolding-vs-dominance vignette than a duplicate here;
Ferrando and Navarro-Gonzalez's (2021) multidimensional Dual Thurstonian
model [@ferrando2021multidimensional].

**The comparison metric (held-out log-likelihood) is an ad hoc choice**,
not a validated best-practice for this setting -- flagged explicitly above,
worth revisiting.

## Reproducibility

```r
# 1. Run the compute script (from the project root)
source("vignettes/continuous_bounded_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/continuous_bounded.qmd")
```

::: {.callout-tip}
To cite the real-data table used here, run:

```r
irw_save_bibtex("lsbq_maleki_2025_non_persian_proficiency",
                 output_file = "continuous_bounded_data/irw_references.bib")
```
:::

---

## References
