---
title: "Bounded Continuous Responses: Comparing Beta IRT, Samejima's CRM, and Mueller's CRSM"
description: |
  Slider and visual-analogue-scale items produce a response that is
  continuous but lives in a fixed, interpretable range, like [0,1] or
  [0,100] -- not the open-ended continuous case standard CFA/linear-factor
  models assume. We compare four published ways of modeling that: a naive
  linear baseline that ignores the bounds, Noel and Dauvier's Beta IRT
  model, Samejima's Continuous Response Model, and Mueller's Continuous
  Rating Scale Model, on one simulated dataset and one real IRW table.
date: today
bibliography:
  - continuous_bounded_data/references.bib
  - continuous_bounded_data/irw_references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`continuous_bounded_compute.R`) directly if
you're relying on the numbers here.
:::

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue   <- "#2166ac"
irw_red    <- "#d73027"
irw_green  <- "#4daf4a"
irw_purple <- "#984ea3"
irw_grey   <- "#808080"

model_colors <- c(
  "naive_linear"              = irw_grey,
  "beta_irt_item_dispersion"  = irw_blue,
  "beta_irt_fixed_dispersion" = irw_purple,
  "samejima_crm"              = irw_red,
  "muller_corsm"              = irw_green
)
model_labels <- c(
  "naive_linear"              = "Naive linear",
  "beta_irt_item_dispersion"  = "Beta IRT (item disp.)",
  "beta_irt_fixed_dispersion" = "Beta IRT (fixed disp.)",
  "samejima_crm"              = "Samejima CRM",
  "muller_corsm"               = "Mueller CRSM"
)

cache_path <- "continuous_bounded_data/continuous_bounded_results.rds"
cache_ok <- file.exists(cache_path)
if (cache_ok) res <- readRDS(cache_path)

grid_cache_path <- "continuous_bounded_data/continuous_bounded_grid_results.rds"
grid_cache_ok <- file.exists(grid_cache_path)
if (grid_cache_ok) grid_res <- readRDS(grid_cache_path)

info_cache_path <- "continuous_bounded_data/continuous_bounded_info_results.rds"
info_cache_ok <- file.exists(info_cache_path)
if (info_cache_ok) info_res <- readRDS(info_cache_path)
```

## Motivation

Visual analogue scales, slider ratings, and continuous confidence or
percentage judgments are common: a respondent marks a point on a
fixed, labeled segment (0 to 100, say), and the raw response is a
continuous number that is *bounded* at both ends. This is a different
measurement problem from the unbounded continuous case (response times,
open-ended magnitude estimates), which reduces to standard linear-factor
or CFA modeling and is covered elsewhere on this site -- the bounds
themselves are the interesting part here, because they create floor/ceiling
effects and non-constant response variance near the edges that a plain
linear model doesn't see.

Three published models handle this directly. Noel and Dauvier's Beta IRT
model [@noel2007beta] treats the response as Beta-distributed given
ability, with the two natural Beta parameters reparameterized into an
item difficulty and a dispersion. Samejima's Continuous Response Model
[@samejima1973homogeneous] models a monotonic transform of the response as
normal given ability -- the historically original approach, predating Beta
IRT by over 30 years. Mueller's Continuous Rating Scale Model
[@mueller1987rasch] takes a Rasch-family, specific-objectivity approach,
estimated by pairwise conditional likelihood; Verhelst's (2019) exponential
family framework [@verhelst2019exponential] is the modern theoretical
generalization of Mueller's model, though the model actually fitted here is
Mueller's original. We compare these three against a naive linear (Gaussian,
identity-link) baseline that ignores the bounds entirely -- the "what if you
just treat it like ordinary continuous data" control, playing the same role
the Rasch baseline plays in the site's guessing-models vignette.

## Data and methods

### The four core specs

1. **Naive linear baseline** -- single-factor CFA via `lavaan`, identity
   link, Gaussian residual, fit directly on the raw (rescaled-to-[0,1])
   response. Ignores boundedness entirely.
2. **Beta IRT** [@noel2007beta] -- fit as a discretized custom `mirt`
   itemtype wrapping `sirt::brm.irf()`, following `sirt`'s own documented
   mirt-estimation recipe (`?sirt::brm.sim`). Two variants: **item-varying
   dispersion** (each item gets its own dispersion parameter, the model's
   default parameterization) and **fixed dispersion** (one dispersion
   parameter shared across all items, via an `mirt` parameter constraint).
3. **Samejima's Continuous Response Model** [@samejima1973homogeneous] --
   `EstCRM::EstCRMitem()` / `EstCRMperson()`.
4. **Mueller's Continuous Rating Scale Model** [@mueller1987rasch] -- a
   patched copy of `pcIRT::CRSM()` / a from-scratch reimplementation of its
   person-scoring equation (see [Limitations](#limitations) for why
   patching was necessary).

### Model equations

Each spec below is written exactly as it is fit in the compute script --
same parameter names, same functional form -- so the equations here can be
checked directly against `continuous_bounded_helpers.R`. Ability
$\theta_i$ itself is never bounded -- every widget below spans
$\theta\in[-3,3]$ -- it is only the *response* $X_{ij}$ that each model
constrains (or, for the naive baseline, fails to constrain) to $[0,1]$.
Each equation is followed by a small widget plotting that model's own
$E[X_{ij}\mid\theta_i]$ curve, so you can see what the equation implies
before any fitting happens; dispersion parameters that don't have their
own slider are held fixed at a representative value, noted in each case.

```{ojs}
//| echo: false
//| output: false

function normalPDF(z, mu, sigma) {
  return Math.exp(-0.5 * ((z - mu) / sigma) ** 2) / (sigma * Math.sqrt(2 * Math.PI));
}

function simpson(f, lo, hi, n) {
  const h = (hi - lo) / n;
  let s = f(lo) + f(hi);
  for (let i = 1; i < n; i++) {
    s += (i % 2 === 0 ? 2 : 4) * f(lo + i * h);
  }
  return (s * h) / 3;
}

function logit(x) {
  return Math.log(x / (1 - x));
}

function normalQuantile(p) {
  const a = [-3.969683028665376e+01, 2.209460984245205e+02, -2.759285104469687e+02, 1.383577518672690e+02, -3.066479806614716e+01, 2.506628277459239e+00];
  const b = [-5.447609879822406e+01, 1.615858368580409e+02, -1.556989798598866e+02, 6.680131188771972e+01, -1.328068155288572e+01];
  const c = [-7.784894002430293e-03, -3.223964580411365e-01, -2.400758277161838e+00, -2.549732539343734e+00, 4.374664141464968e+00, 2.938163982698783e+00];
  const d = [7.784695709041462e-03, 3.224671290700398e-01, 2.445134137142996e+00, 3.754408661907416e+00];
  const plow = 0.02425;
  const phigh = 1 - plow;
  if (p < plow) {
    const q = Math.sqrt(-2 * Math.log(p));
    return (((((c[0]*q+c[1])*q+c[2])*q+c[3])*q+c[4])*q+c[5]) /
           ((((d[0]*q+d[1])*q+d[2])*q+d[3])*q+1);
  } else if (p <= phigh) {
    const q = p - 0.5;
    const r = q * q;
    return (((((a[0]*r+a[1])*r+a[2])*r+a[3])*r+a[4])*r+a[5])*q /
           (((((b[0]*r+b[1])*r+b[2])*r+b[3])*r+b[4])*r+1);
  } else {
    const q = Math.sqrt(-2 * Math.log(1 - p));
    return -(((((c[0]*q+c[1])*q+c[2])*q+c[3])*q+c[4])*q+c[5]) /
             ((((d[0]*q+d[1])*q+d[2])*q+d[3])*q+1);
  }
}

function lgamma(x) {
  const g = 7;
  const c = [
    0.99999999999980993, 676.5203681218851, -1259.1392167224028,
    771.32342877765313, -176.61502916214059, 12.507343278686905,
    -0.13857109526572012, 9.9843695780195716e-6, 1.5056327351493116e-7
  ];
  if (x < 0.5) {
    return Math.log(Math.PI / Math.sin(Math.PI * x)) - lgamma(1 - x);
  }
  x -= 1;
  let a = c[0];
  const t = x + g + 0.5;
  for (let i = 1; i < g + 2; i++) {
    a += c[i] / (x + i);
  }
  return 0.5 * Math.log(2 * Math.PI) + (x + 0.5) * Math.log(t) - t + Math.log(a);
}

function naiveMean(theta, b, a) {
  return 0.5 + a * (theta - b);
}

function naiveDensity(x, theta, b, a, sigma) {
  return normalPDF(x, naiveMean(theta, b, a), sigma);
}

function betaShape1(theta, delta, tau) {
  return Math.exp((theta - delta + tau) / 2);
}

function betaShape2(theta, delta, tau) {
  return Math.exp((-theta + delta + tau) / 2);
}

function betaIrtMean(theta, delta) {
  return 1 / (1 + Math.exp(-(theta - delta)));
}

function betaPDF(x, a, b) {
  if (x <= 0 || x >= 1) return 0;
  const logB = lgamma(a) + lgamma(b) - lgamma(a + b);
  return Math.exp((a - 1) * Math.log(x) + (b - 1) * Math.log(1 - x) - logB);
}

function betaIrtDensity(x, theta, delta, tau) {
  return betaPDF(x, betaShape1(theta, delta, tau), betaShape2(theta, delta, tau));
}

function samejimaMean(theta, alpha, b, sigma) {
  const loc = alpha * (theta - b);
  return simpson((z) => (1 / (1 + Math.exp(-z))) * normalPDF(z, loc, sigma), loc - 6 * sigma, loc + 6 * sigma, 60);
}

function samejimaDensity(x, theta, alpha, b, sigma) {
  if (x <= 0 || x >= 1) return 0;
  return normalPDF(logit(x), alpha * (theta - b), sigma) / (x * (1 - x));
}

function muellerUnnorm(x, theta, b, lambda) {
  return Math.exp(x * (theta - b) + x * (1 - x) * lambda);
}

function muellerNormalizer(theta, b, lambda) {
  return simpson((x) => muellerUnnorm(x, theta, b, lambda), 0, 1, 100);
}

function muellerMean(theta, b, lambda) {
  return simpson((x) => x * muellerUnnorm(x, theta, b, lambda), 0, 1, 100) / muellerNormalizer(theta, b, lambda);
}

function muellerDensity(x, theta, b, lambda) {
  if (x <= 0 || x >= 1) return 0;
  return muellerUnnorm(x, theta, b, lambda) / muellerNormalizer(theta, b, lambda);
}

model_colors = ({
  naive_linear: "#808080",
  beta_irt: "#2166ac",
  samejima_crm: "#d73027",
  muller_corsm: "#4daf4a"
})

model_labels = ({
  naive_linear: "Naive linear",
  beta_irt: "Beta IRT",
  samejima_crm: "Samejima CRM",
  muller_corsm: "Mueller CRSM"
})

theta_grid = d3.range(-3, 3.01, 0.1)

fixed_nl_sigma = 0.15
fixed_sam_sigma = 1
fixed_mu_lambda = 2
```

**Naive linear baseline.** Ordinary single-factor CFA on the raw
(rescaled-to-$[0,1]$) response: $X_{ij} = \nu_j + \lambda_j\theta_i +
\varepsilon_{ij}$, $\varepsilon_{ij}\sim N(0,\sigma_j^2)$, fit via
`lavaan::cfa()`. Nothing here constrains $X_{ij}$ to $[0,1]$, as the widget
below makes obvious once $\theta$ moves far enough from $b$.

```{ojs}
//| echo: false

nl_b_input = Inputs.range([-2, 2], {value: 0, step: 0.1, label: "Location (b)"})
nl_a_input = Inputs.range([0.3, 2.5], {value: 1, step: 0.1, label: "Discrimination (a)"})
nl_b = Generators.input(nl_b_input)
nl_a = Generators.input(nl_a_input)

html`<div style="display:flex; flex-wrap:wrap; gap:2rem; align-items:center;">${nl_b_input}${nl_a_input}</div>`
```

```{ojs}
//| echo: false

Plot.plot({
  width: 800,
  height: 280,
  x: {label: "Ability (θ)", domain: [-3, 3]},
  y: {label: "Expected response", domain: [-0.6, 1.6]},
  marks: [
    Plot.ruleY([0, 1], {stroke: "#999", strokeDasharray: "2,2"}),
    Plot.line(theta_grid.map((theta) => ({theta, y: naiveMean(theta, nl_b, nl_a)})), {
      x: "theta", y: "y", stroke: model_colors.naive_linear, strokeDasharray: "4,3"
    })
  ]
})
```

**Beta IRT** [@noel2007beta], as `sirt::brm.irf()` actually parameterizes
it: $X_{ij}\mid\theta_i \sim \text{Beta}(\text{shape1}_{ij},
\text{shape2}_{ij})$, with
$$\text{shape1}_{ij} = \exp\!\left(\frac{\theta_i-\delta_j+\tau_j}{2}\right),
\qquad
\text{shape2}_{ij} = \exp\!\left(\frac{-\theta_i+\delta_j+\tau_j}{2}\right),$$
where $\delta_j$ is item difficulty and $\tau_j$ is dispersion (fixed
across items in the "fixed dispersion" variant, free in the "item-varying"
variant). Working out this Beta distribution's mean shows it reduces
exactly to $E[X_{ij}\mid\theta_i] = \text{logistic}(\theta_i-\delta_j)$,
independent of $\tau_j$ -- **there is no discrimination parameter**, so
unlike the other 3 widgets on this page, this one has only a single
slider. $\tau_j$ is left out of it entirely, not just discrimination:
since $\tau_j$ enters $\text{shape1}_{ij}$ and $\text{shape2}_{ij}$ with
the *same* sign, it changes their sum (the Beta distribution's
concentration) without changing their ratio (the mean), so it would have
no visible effect on this mean-curve widget either. It does change the
*density*'s spread, though -- see the $\tau_j$ slider in [Comparing
response densities across models](#comparing-response-densities-across-models)
below.

```{ojs}
//| echo: false
viewof beta_delta = Inputs.range([-2, 2], {value: 0, step: 0.1, label: "Item difficulty (δⱼ)"})
```

```{ojs}
//| echo: false

Plot.plot({
  width: 800,
  height: 280,
  x: {label: "Ability (θ)", domain: [-3, 3]},
  y: {label: "Expected response", domain: [-0.15, 1.15]},
  marks: [
    Plot.ruleY([0, 1], {stroke: "#999", strokeDasharray: "2,2"}),
    Plot.line(theta_grid.map((theta) => ({theta, y: betaIrtMean(theta, beta_delta)})), {
      x: "theta", y: "y", stroke: model_colors.beta_irt
    })
  ]
})
```

**Samejima's Continuous Response Model** [@samejima1973homogeneous], via
`EstCRM::EstCRMitem()`: the raw response is transformed to
$Z_{ij}=\ln\!\big(X_{ij}/(1-X_{ij})\big)$, and $Z_{ij} = \alpha_j(\theta_i -
b_j) + e_{ij}$, $e_{ij}\sim N(0,\sigma_j^2)$, i.e. an ordinary linear factor
model *on the transformed scale*. $\alpha_j$ is discrimination and $b_j$ is
difficulty. The observed-scale mean response is the backtransform of this
normal model's mean through the logistic link (dispersion held fixed at
$\sigma=1$ below).

```{ojs}
//| echo: false

sam_b_input = Inputs.range([-2, 2], {value: 0, step: 0.1, label: "Difficulty (b)"})
sam_alpha_input = Inputs.range([0.3, 2.5], {value: 1, step: 0.1, label: "Discrimination (α)"})
sam_b = Generators.input(sam_b_input)
sam_alpha = Generators.input(sam_alpha_input)

html`<div style="display:flex; flex-wrap:wrap; gap:2rem; align-items:center;">${sam_b_input}${sam_alpha_input}</div>`
```

```{ojs}
//| echo: false

Plot.plot({
  width: 800,
  height: 280,
  x: {label: "Ability (θ)", domain: [-3, 3]},
  y: {label: "Expected response", domain: [-0.15, 1.15]},
  marks: [
    Plot.ruleY([0, 1], {stroke: "#999", strokeDasharray: "2,2"}),
    Plot.line(theta_grid.map((theta) => ({theta, y: samejimaMean(theta, sam_alpha, sam_b, fixed_sam_sigma)})), {
      x: "theta", y: "y", stroke: model_colors.samejima_crm
    })
  ]
})
```

**Mueller's Continuous Rating Scale Model** [@mueller1987rasch], via
`pcIRT::CRSM()`: a Rasch-family (specific-objectivity) density,
$$f(x_{ij}\mid\theta_i) \;\propto\; \exp\!\big[x_{ij}(\theta_i-b_j) +
x_{ij}(1-x_{ij})\lambda_j\big], \quad x_{ij}\in[0,1],$$
with a single item-location parameter $b_j$ and a dispersion parameter
$\lambda_j$ -- no discrimination parameter, by construction of the
Rasch-family specific-objectivity property. Unlike Beta IRT, $\lambda_j$
*does* move the mean here (the widget below lets it, rather than holding
it fixed), since the backtransform from the exponential-tilt density to
$E[X_{ij}\mid\theta_i]$ is nonlinear in $\lambda_j$.

```{ojs}
//| echo: false

mu_b_input = Inputs.range([-2, 2], {value: 0, step: 0.1, label: "Location (b)"})
mu_lambda_input = Inputs.range([-3, 5], {value: 2, step: 0.1, label: "Dispersion (λ)"})
mu_b = Generators.input(mu_b_input)
mu_lambda = Generators.input(mu_lambda_input)

html`<div style="display:flex; flex-wrap:wrap; gap:2rem; align-items:center;">${mu_b_input}${mu_lambda_input}</div>`
```

```{ojs}
//| echo: false

Plot.plot({
  width: 800,
  height: 280,
  x: {label: "Ability (θ)", domain: [-3, 3]},
  y: {label: "Expected response", domain: [-0.15, 1.15]},
  marks: [
    Plot.ruleY([0, 1], {stroke: "#999", strokeDasharray: "2,2"}),
    Plot.line(theta_grid.map((theta) => ({theta, y: muellerMean(theta, mu_b, mu_lambda)})), {
      x: "theta", y: "y", stroke: model_colors.muller_corsm
    })
  ]
})
```

### Comparing response densities across models

The 4 widgets above each plot a single line -- a model's mean response --
but every one of them is really a full probability distribution over
$X_{ij}\in[0,1]$ (or, for naive linear, over all of $\mathbb{R}$). Because
all 4 responses live on the same $[0,1]$ scale, their densities can be
overlaid directly. Rather than fix $\theta$ at a specific, unbounded
value, the slider below is expressed as an **ability percentile**, mapped
to $\theta$ through the standard normal quantile function -- the same
$\theta_i\sim N(0,1)$ assumed everywhere else in this vignette (e.g. the
simulation grid's `theta_true <- rnorm(N)`). 50% is $\theta=0$, i.e. the
person of average ability the earlier fixed value silently assumed;
dragging toward the tails reaches every $\theta$ a percentile-bounded
slider can, without the awkwardness of a slider whose bounds are
themselves an arbitrary choice on an unbounded scale. Dispersion is held
at the same representative values used above ($\sigma=0.15$ for naive
linear on the raw scale, $\sigma=1$ for Samejima CRM, $\lambda=2$ for
Mueller CRSM); drag difficulty ($b$) and discrimination ($a$, Beta IRT's
$\delta_j$ and Mueller CRSM have no discrimination term) to see how each
model's *whole distribution*, not just its mean, responds. Beta IRT gets
one more slider here than in its own widget above: $\tau_j$, which has no
effect on Beta IRT's mean curve (see [Model equations](#model-equations))
but a real one on its density's concentration, visible only in a density
view like this one.

```{ojs}
//| echo: false

cmp_percentile_input = Inputs.range([1, 99], {value: 50, step: 1, label: "Ability percentile"})
cmp_b_input = Inputs.range([-2, 2], {value: 0, step: 0.1, label: "Item difficulty / δⱼ (b)"})
cmp_a_input = Inputs.range([0.3, 2.5], {value: 1, step: 0.1, label: "Discrimination (a)"})
cmp_tau_input = Inputs.range([-2, 4], {value: 1, step: 0.1, label: "Beta IRT dispersion (τⱼ)"})
cmp_percentile = Generators.input(cmp_percentile_input)
cmp_b = Generators.input(cmp_b_input)
cmp_a = Generators.input(cmp_a_input)
cmp_tau = Generators.input(cmp_tau_input)

html`<div style="display:flex; flex-wrap:wrap; gap:2rem; align-items:center;">${cmp_percentile_input}${cmp_b_input}${cmp_a_input}${cmp_tau_input}</div>`
```

```{ojs}
//| echo: false

cmp_theta = normalQuantile(cmp_percentile / 100)

md`${cmp_percentile}th percentile ability &rarr; θ = ${cmp_theta.toFixed(2)}. Discrimination (a) has no effect on Beta IRT or Mueller CRSM; τⱼ has no effect on Beta IRT's mean, only its density (see panel below).`
```

```{ojs}
//| echo: false

cmp_x_grid = d3.range(-0.4, 1.401, 0.004)

cmp_density_data = cmp_x_grid.flatMap((x) => [
  { x, model: "naive_linear", density: naiveDensity(x, cmp_theta, cmp_b, cmp_a, fixed_nl_sigma) },
  { x, model: "beta_irt", density: betaIrtDensity(x, cmp_theta, cmp_b, cmp_tau) },
  { x, model: "samejima_crm", density: samejimaDensity(x, cmp_theta, cmp_a, cmp_b, fixed_sam_sigma) },
  { x, model: "muller_corsm", density: muellerDensity(x, cmp_theta, cmp_b, fixed_mu_lambda) }
])

Plot.plot({
  width: 900,
  height: 420,
  x: {label: "Response (x)", domain: [-0.4, 1.4]},
  y: {label: "Density"},
  color: {
    domain: Object.keys(model_colors),
    range: Object.values(model_colors),
    legend: true,
    tickFormat: (d) => model_labels[d]
  },
  marks: [
    Plot.ruleX([0, 1], {stroke: "#999", strokeDasharray: "2,2"}),
    Plot.line(cmp_density_data.filter((d) => d.model === "naive_linear"), {
      x: "x", y: "density", stroke: "model", strokeDasharray: "4,3"
    }),
    Plot.line(cmp_density_data.filter((d) => d.model !== "naive_linear"), {
      x: "x", y: "density", stroke: "model"
    })
  ]
})
```

The dashed vertical lines at $x=0$ and $x=1$ mark the response range; naive
linear (dashed gray) is the only density that isn't actually confined
inside them -- exactly the failure mode its widget above shows for the
mean curve, now visible in the distribution itself. The other 3 densities
all integrate to 1 strictly within $[0,1]$: Beta IRT's shifts left/right
with $\delta_j$ without changing shape, but *narrows or widens in place* as
$\tau_j$ moves -- large $\tau_j$ concentrates it tightly around the mean,
small (or negative) $\tau_j$ spreads it out and can even pull mass toward
both boundaries at once, all while the mean itself never moves, exactly as
the equation in [Model equations](#model-equations) implies. Samejima
CRM's and Mueller CRSM's shapes visibly skew as $b$ moves them toward
either boundary, a consequence of the nonlinear backtransform in both
models that a mean-curve-only view can't show. Moving $a$ redistributes
naive linear's and Samejima CRM's mass without translating it (steeper
slope on the same $\theta$), and does nothing to Beta IRT's or Mueller
CRSM's density, consistent with neither model having a discrimination
parameter.

### Comparing information curves across models {#information-curves}

The densities above show *where* probability mass sits; item information
shows *how precisely* each model can pin down $\theta$ from a response, as
a function of $\theta$ itself. This is worth looking at on its own, because
information curves can diverge sharply across these 4 models even for
items with matched difficulty/discrimination, in a way the mean and
density widgets above don't make obvious -- a point raised directly by
Dylan Molenaar (author of the `BoundIRT` package discussed in
[Limitations](#limitations) [@molenaar2026boundirt]) in feedback on an
earlier draft of this vignette.

Fisher information is computed the same way for all 4 models here, without
relying on a model-specific closed form: $I(\theta) = E\!\left[\left(\dfrac{\partial}{\partial\theta}\log
f(X\mid\theta)\right)^{2}\right]$, with the score obtained by central finite
difference on each model's own log-density (the same `*Density()` functions
used for the plots above) and the expectation evaluated by numerical
integration over $X$. This is a numerical approximation, not an
analytically derived information function -- adequate for a qualitative
comparison of *shape* across models, but not a substitute for each model's
published closed-form information function if a precise value is needed.

```{ojs}
//| echo: false
//| output: false

function scoreSquaredTimesDensity(x, theta, densityOfTheta, h) {
  const fC = densityOfTheta(x, theta);
  if (fC <= 1e-10) return 0;
  const fP = densityOfTheta(x, theta + h);
  const fM = densityOfTheta(x, theta - h);
  const logFP = Math.log(Math.max(fP, 1e-300));
  const logFM = Math.log(Math.max(fM, 1e-300));
  const score = (logFP - logFM) / (2 * h);
  return score * score * fC;
}

function fisherInformationAt(densityOfTheta, theta, lo, hi, n, h = 0.001) {
  return simpson((x) => scoreSquaredTimesDensity(x, theta, densityOfTheta, h), lo, hi, n);
}

// For the 3 bounded-response models, integrating over x directly on a fixed
// grid (as fisherInformationAt above does) silently fails at extreme
// parameter values: a steep model-implied slope can concentrate almost all
// response density into an x-window far narrower than the grid resolves,
// so the integral is undercounted and the curve falsely collapses toward 0
// -- caught by checking Samejima CRM against its own closed form (Fisher
// info there is analytically alpha^2/sigma^2, exactly constant in theta,
// since Z is a Gaussian location family and Fisher information is
// invariant under the 1-to-1 Z->X transform); the naive grid showed a false
// collapse near the theta range's edges at high slider values where the
// true curve is flat. Substituting z = logit(x) as the integration
// variable fixes this generically: a spike near x=0 or x=1 becomes a
// normal-width bump on the z scale, which a uniform grid over z resolves
// correctly regardless of how extreme the original x-concentration was.
function fisherInformationBoundedAt(densityOfTheta, theta, h = 0.001, zlo = -30, zhi = 30, n = 400) {
  const integrand = (z) => {
    const x = 1 / (1 + Math.exp(-z));
    const jac = x * (1 - x); // dx/dz
    const fC = densityOfTheta(x, theta) * jac;
    if (fC <= 1e-300) return 0;
    const fP = densityOfTheta(x, theta + h) * jac;
    const fM = densityOfTheta(x, theta - h) * jac;
    const score = (Math.log(Math.max(fP, 1e-300)) - Math.log(Math.max(fM, 1e-300))) / (2 * h);
    return score * score * fC;
  };
  return simpson(integrand, zlo, zhi, n);
}
```

```{ojs}
//| echo: false

info_b_input = Inputs.range([-2, 2], {value: 0, step: 0.1, label: "Item difficulty / δⱼ (b)"})
info_a_input = Inputs.range([0.3, 2.5], {value: 1, step: 0.1, label: "Discrimination (a)"})
info_tau_input = Inputs.range([-2, 4], {value: 1, step: 0.1, label: "Beta IRT dispersion (τⱼ)"})
info_b = Generators.input(info_b_input)
info_a = Generators.input(info_a_input)
info_tau = Generators.input(info_tau_input)

html`<div style="display:flex; flex-wrap:wrap; gap:2rem; align-items:center;">${info_b_input}${info_a_input}${info_tau_input}</div>`
```

```{ojs}
//| echo: false

info_naive_of    = (x, th) => naiveDensity(x, th, info_b, info_a, fixed_nl_sigma)
info_beta_of     = (x, th) => betaIrtDensity(x, th, info_b, info_tau)
info_samejima_of = (x, th) => samejimaDensity(x, th, info_a, info_b, fixed_sam_sigma)
info_mueller_of  = (x, th) => muellerDensity(x, th, info_b, fixed_mu_lambda)

info_curve_data = theta_grid.flatMap((theta) => {
  // naive linear is unbounded, so its integration window has to track the
  // (possibly far-off-[0,1]) mean as theta/b/a move, unlike the 3 bounded
  // models, whose domain is fixed at [0,1] regardless of parameters
  const nlMean = naiveMean(theta, info_b, info_a);
  const nlLo = nlMean - 8 * fixed_nl_sigma;
  const nlHi = nlMean + 8 * fixed_nl_sigma;
  return [
    { theta, model: "naive_linear", info: fisherInformationAt(info_naive_of, theta, nlLo, nlHi, 100) },
    { theta, model: "beta_irt",     info: fisherInformationBoundedAt(info_beta_of, theta) },
    { theta, model: "samejima_crm", info: fisherInformationBoundedAt(info_samejima_of, theta) },
    { theta, model: "muller_corsm", info: fisherInformationBoundedAt(info_mueller_of, theta) }
  ];
})

Plot.plot({
  width: 900,
  height: 380,
  x: {label: "Ability (θ)", domain: [-3, 3]},
  y: {label: "Fisher information I(θ)"},
  color: {
    domain: Object.keys(model_colors),
    range: Object.values(model_colors),
    legend: true,
    tickFormat: (d) => model_labels[d]
  },
  marks: [
    Plot.line(info_curve_data.filter((d) => d.model === "naive_linear"), {
      x: "theta", y: "info", stroke: "model", strokeDasharray: "4,3"
    }),
    Plot.line(info_curve_data.filter((d) => d.model !== "naive_linear"), {
      x: "theta", y: "info", stroke: "model"
    })
  ]
})
```

With difficulty/discrimination matched across all 4 models (drag $b$ and
$a$ above and watch every curve move together), the shapes still don't
match -- and not along the split you might expect from "3 bounded models
vs. 1 unbounded baseline." **Two of the 4 curves are exactly flat in
$\theta$, for two different reasons.** Naive linear's information (dashed
grey) is constant because it's a plain Gaussian location family with
linear mean, where information reduces to $a^2/\sigma^2$ independent of
$\theta$ -- unsurprising, since the model is unbounded and pays no
attention to $[0,1]$ at all. Samejima CRM's information (red) is *also*
exactly constant, at $\alpha^2/\sigma^2$ -- a much less obvious fact, since
the model's raw-scale density is visibly non-Gaussian and does change
shape with $\theta$ (see the density widget above). The resolution: CRM
models a Gaussian location family on the *transformed* scale
($Z=\alpha(\theta-b)+e$), and Fisher information is invariant under any
one-to-one transform of the observed data -- so observing the bounded
$X=\text{logistic}(Z)$ carries exactly the same information about $\theta$
as observing $Z$ directly would, at every $\theta$, regardless of how
skewed $X$'s own density looks. **Only Beta IRT and Mueller CRSM actually
have $\theta$-varying information**, and for different reasons than each
other: Beta IRT's (blue/purple) is single-peaked and comes closest to the
"more informative in the middle of the range" intuition information curves
usually carry, with $\tau_j$ (which has no effect on any model's mean)
reshaping its height and width without moving it; Mueller CRSM's (green)
is much flatter overall but still genuinely slopes, asymmetrically once
$\lambda\neq0$, rather than sitting dead level like naive linear's or
Samejima's. So two analysts fitting ostensibly comparable "bounded IRT"
models to the same item can walk away with opposite conclusions about
whether that item's precision even depends on ability at all --
exactly the kind of divergence a mean-curve or single-density snapshot
doesn't surface, and the reason it's included here as its own comparison
rather than folded into an existing widget.

See [Empirical information curves](#empirical-information-curves-the-primary-real-table)
in [Real data](#real-data) below for the same comparison using item
parameters actually fitted to a real IRW table, rather than the
hand-set values above -- including a genuine estimation failure the
sliders here can't produce.

### Estimation and model comparison

Every model's held-out fit is reported as the **log-probability of the
response falling in a small interval**, not a raw point density. This
choice exists because the 4 models don't report likelihood on the same
scale to begin with: reading `EstCRM::EstCRMitem()`'s source shows it
transforms the raw response to $Z = \ln\!\big(X/(K-X)\big)$ and computes
its entire EM log-likelihood on that transformed scale, not the original
response scale the Beta IRT and naive linear fits use. Comparing those two
log-likelihoods directly would be comparing different units -- like
comparing an error reported in inches to one reported in centimeters
without converting first. The usual fix is a model-specific Jacobian
correction; instead, every model's held-out score here is
$\log P(\text{bin}_{lo} < X < \text{bin}_{hi})$, using the same bin edges
(on the raw, rescaled-to-$[0,1]$ scale) for all 4 models. A probability that
a response lands in a fixed interval means the same thing regardless of
which internal scale a model uses to compute it, so this sidesteps the
rescaling problem entirely rather than requiring a per-model correction.

Held-out log-probability, built this way, is the metric used to compare all
4 specs throughout this vignette. It's a standard, interpretable scale for
predictive fit, though not a validated best-practice specific to this
bounded-continuous setting -- flagged as an ad hoc choice worth revisiting
(see [Limitations](#limitations)).

This same design also makes scores **comparable across the 11 real tables**,
not just across the 4 models within a table -- a separate problem the
rescaling was never explicitly built for but solves anyway.
`rescale01()` maps every table's raw response to $[0,1]$ using that
table's own bounds before anything else happens (`continuous_bounded_helpers.R`),
and the bin edges used for scoring are equal-width on that shared $[0,1]$
scale with the same bin count ($K=6$) for every table, regardless of its
raw range. A table scored on a raw 0-10 scale and one scored on a raw
0-100 scale both get bin width exactly $1/6$ in the space the score is
actually computed in -- so a wide bar or point for one table isn't wider
raw units in disguise, and height differences across tables in the figures
below reflect genuine differences in predictive fit, not unit mismatches.

### Held-out design: genuine missing-response prediction

The project's usual convention for held-out evaluation on IRW data is a
response-level (missing-cell) holdout: mask a random subset of person x item
cells, fit on the rest, score the masked cells. Fitting item parameters this
way isn't viable for 2 of the 4 specs -- reading `EstCRM::EstCRMitem()` and
`pcIRT::CRSM()`'s fitting code directly shows **neither has any
missing-data handling at all** (both do matrix/pairwise algebra that
silently breaks with `NA` cells, confirmed by testing) -- so item
parameters for all 4 specs are still fit on a complete training-person
matrix, no missing cells.

*Scoring*, once item parameters are fixed, is a different matter: all 4
specs' person-location estimators turn out to be NA-tolerant.
`mirt::fscores(response.pattern = ...)` handles missing cells natively (a
core mirt feature, not something added here); `lavaan::lavPredict()`
computes factor scores from an incomplete `newdata` row the same way;
`EstCRM::EstCRMperson()` skips `NA` columns per person when forming its
theta estimate (confirmed by reading its source); and the from-scratch
Mueller CRSM person-scoring routine (see [Limitations](#limitations)) was
written to do the same. Each was smoke-tested directly (masked vs.
full-data theta estimates, correlation $>0.98$ in every case) before
being used for real.

That makes true cell-level prediction possible after all: for every
held-out **test person**, a random subset of their items (25%, clamped to
at least 1 kept and at least 1 held out) is hidden before estimating that
person's ability from the remaining items alone, and only the hidden items
are scored. The model never sees the scored value, directly or indirectly,
at any stage -- this is genuine missing-response prediction, not the
weaker "estimate ability from a person's full vector, then score those same
items" self-consistency check an earlier version of this design used.

## Simulation {#simulation}

Before turning to real IRW tables, this section checks the 4 specs against
data simulated directly from known, correctly-specified generating models
-- design first, then results. The short version, expanded below: every
correctly-specified model recovers its true item parameters cleanly, the 3
bounded-response models beat the naive linear baseline on held-out
log-probability regardless of which of them generated the data, and
boundary inflation degrades every spec somewhat but the 3 bounded-response
models noticeably less than naive linear. Real-data selection and results
follow in the next section, once this baseline is established.

### Design

**A single illustrative condition** ($N=300$, $J=10$, Beta IRT DGP, fixed
dispersion) is shown first below, generated directly from the Beta IRT
model's own equations (`sim_beta_irt()` in `continuous_bounded_helpers.R`
-- a bug-fixed reimplementation of `sirt::brm.sim()`'s logic; see
[Limitations](#limitations) for why the reimplementation was necessary).

**The full recovery/misspecification grid** cross-fits all 4 core specs
against 3 DGPs, generated from each model's own equations: Beta IRT
(`sim_beta_irt()`), Samejima CRM (`sim_samejima_crm()`: $Z_{ij} =
\alpha_j(\theta_i - b_j) + e_{ij}$, $e_{ij}\sim N(0,\sigma_j^2)$, then
$X_{ij}=1/(1+e^{-Z_{ij}})$ -- the exact inverse of EstCRM's own $Z=\ln(X/(1-X))$
transform), and Mueller CRSM (`sim_muller_corsm()`: grid-based inverse-CDF
sampling from the same $\exp[x(\theta-b)+x(1-x)\lambda]$ density used for
scoring, since no closed-form sampler exists for this family). The
naive-linear DGP is skipped: an unbounded-Gaussian DGP clipped to $[0,1]$ to
make it "bounded" would introduce an artificial boundary spike
unrepresentative of what a linear model actually generates.

Grid: $N\in\{300,1000\}$, $J\in\{8,20\}$, and, for the Beta IRT DGP
specifically, fixed vs. item-varying dispersion -- 8 + 4 + 4 = 16
conditions, plus 1 boundary-inflation condition (10% of responses forced to
exactly 0, 10% forced to exactly 1, layered on the Beta IRT DGP at
$N=300$/$J=8$) for 17 total. Run via `furrr::future_pmap()` +
`plan(multisession)`, cached separately from the real-data results below
(own skip-if-exists gate in the compute script).

### Does the correctly-specified model win, and does it recover parameters?

```{r}
#| label: fig-sim-scores
if (cache_ok) {
  res$sim$scores |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = sprintf("%.2f", mean_ll)), vjust = 1.4, color = "white", size = 3.5) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = "Simulated Beta-IRT data: held-out fit by model") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

On data simulated directly from the Beta IRT model, the two Beta IRT
variants have the best (least negative) held-out fit, essentially tied with
Samejima CRM and Mueller CRSM, and all three clearly beat the naive linear
baseline. This is the expected pattern for a correctly-specified-model
recovery check: the models built for bounded continuous responses recover
the truth better than the one that ignores the bounds.

Item-difficulty recovery for the item-varying-dispersion Beta IRT fit
(correlation between estimated and true $\delta$) was
`r if (cache_ok) sprintf("%.3f", res$sim$recovery_cor_delta) else "NA"`.

### The full recovery grid: 16 conditions x 3 DGPs

**"Recovery" here means one specific, narrow comparison**: for each
condition, only the fitted model that matches that condition's DGP is
checked -- the correlation between its estimated item-difficulty
parameters and the true difficulties used to simulate the data (e.g. for
a Beta-IRT-generated condition, only the Beta IRT fit's estimated
$\delta_j$ vs. true $b_j$ is computed). The other 3 fitted models aren't
included in this table at all: their difficulty-type parameters aren't on
a directly comparable scale to the true simulation parameters without a
model-specific transform, so a recovery correlation for them isn't
computed here. (Whether all 4 models can nonetheless fit the same data
well is a separate question, answered by the cross-fitting comparison
just below, not by this table.) With that scope in mind, parameter
recovery for the correctly-specified model was excellent everywhere --
every one of the 16 non-inflation conditions recovered its true item
parameters at $r>0.96$, most above $r>0.99$:

```{r}
#| label: tbl-grid-recovery
if (grid_cache_ok) {
  grid_res$scores |>
    filter(!boundary_inflation, !response_heaping) |>
    distinct(dgp, N, J, dispersion_mode, recovery_cor) |>
    arrange(dgp, N, J, dispersion_mode) |>
    mutate(recovery_cor = sprintf("%.3f", recovery_cor)) |>
    knitr::kable(col.names = c("DGP", "N", "J", "Beta IRT dispersion", "Recovery r (correctly-specified model only)"))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

More useful than "does recovery work" (it does, everywhere) is the
cross-fitting question: when data comes from one DGP, which of the 4
*fitted* specs wins the held-out comparison?

```{r}
#| label: fig-grid-crossfit
if (grid_cache_ok) {
  # marks the correctly-specified model(s) in each DGP facet with a vertical
  # line; the Beta IRT DGP panel averages over both dispersion_mode
  # conditions, so both Beta IRT variants are correctly specified somewhere
  # in that average and both get a line
  correct_model_df <- data.frame(
    dgp_label = c("DGP: Beta IRT", "DGP: Beta IRT", "DGP: Samejima CRM", "DGP: Mueller CRSM"),
    model = factor(c("beta_irt_item_dispersion", "beta_irt_fixed_dispersion", "samejima_crm", "muller_corsm"),
                   levels = names(model_colors))
  )

  grid_res$scores |>
    filter(!boundary_inflation, !response_heaping) |>
    mutate(model = factor(model, levels = names(model_colors)),
           dgp_label = recode(dgp, beta_irt = "DGP: Beta IRT", samejima_crm = "DGP: Samejima CRM",
                               muller_corsm = "DGP: Mueller CRSM")) |>
    group_by(dgp_label, model) |>
    summarise(mean_ll = mean(mean_ll, na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_vline(data = correct_model_df, aes(xintercept = as.numeric(model)),
               color = "black", linewidth = 0.9) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    facet_wrap(~dgp_label, ncol = 3) +
    labs(x = NULL, y = "Mean held-out log-probability\n(averaged across N/J conditions)",
         title = "Cross-fitting: held-out fit by model, averaged across all N/J conditions per DGP",
         subtitle = "Black line marks the model(s) correctly specified for that panel's DGP") +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

The pattern is consistent across all 3 DGPs, and, as shown in [Real
data](#real-data) below, the same pattern holds up on real IRW tables too:
**the 3 bounded-response models cluster tightly together and substantially
outperform the naive linear baseline, regardless of which of the 3
actually generated the data.** "Correctly specified wins" is a much
weaker pattern here than under a self-consistency check, though: under the
Beta IRT DGP, both Beta IRT variants edge out Samejima CRM by only
$\approx$0.003 in mean held-out log-probability -- a three-way tie in
practice. Under the Samejima CRM DGP, Samejima does come out slightly
ahead, but by less than 0.01. And under the Mueller CRSM DGP, Mueller CRSM
is **not** the best-fitting model -- Samejima CRM scores marginally better
than Mueller's own generating model, a genuine (if small) misspecification
surprise, not papered over here. The margin between the 3 bounded-response
models is small compared to the gap to naive linear in every case -- these
3 models are more interchangeable with each other than any of them is with
a model that ignores the bounds, but which one "wins" under true
missing-response prediction is close to a coin flip. One fit failed
outright:
Samejima CRM's EM did not converge (a `missing value where TRUE/FALSE
needed` error partway through) on one Mueller-CRSM-generated dataset
($N=300$, $J=8$) -- a real misspecification-robustness finding, not
papered over; see [Limitations](#limitations).

### Boundary inflation: does floor/ceiling piling break anything?

Beyond fitting, we asked how each of the 4 specs' held-out fit *degrades*
when 20% of responses are forced to the exact boundary (10% to 0, 10% to
1) on top of the Beta IRT DGP. This is a stress test of the 4 existing
specs, not a new model: none of them includes a dedicated zero-one-inflated
component, so the question is how much their held-out fit suffers without
one, not which model handles inflation best.

```{r}
#| label: fig-grid-inflation
if (grid_cache_ok) {
  grid_res$scores |>
    filter(dgp == "beta_irt", N == 300, J == 8, dispersion_mode == "fixed", !response_heaping) |>
    mutate(model = factor(model, levels = names(model_colors)),
           condition = ifelse(boundary_inflation, "20% boundary-inflated", "Not inflated")) |>
    ggplot(aes(x = model, y = mean_ll, fill = model, alpha = condition)) +
    geom_col(position = "dodge", width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_alpha_manual(values = c("Not inflated" = 1, "20% boundary-inflated" = 0.55), name = NULL) +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability",
         title = "Beta IRT DGP (N=300, J=8): held-out fit with vs. without boundary inflation") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

Every spec's held-out fit gets worse under inflation -- none of the 4 core
specs explicitly models a zero-one-inflated mixture, so this is expected --
but **not by the same amount**. The naive linear baseline degrades by
about 20% (mean log-probability drops from -1.74 to -2.09); the 3
bounded-response models degrade by roughly 8-10% each (e.g. Mueller CRSM:
-1.46 to -1.61). The bounded-response models' smooth density still puts
more mass near the boundary than a Gaussian does, so they absorb some of
the extra piling without a dedicated inflation term, while the naive linear
model has nothing in its structure to absorb it at all. None of the 4
handle it *well* -- a true zero-one-inflated extension [@molenaar2022zero]
would be needed for that -- but the bounded-response models are
comparatively more robust to it than the baseline that ignores boundedness
altogether.

### Response heaping: does round-number clumping break anything?

Boundary inflation isn't the only way real responses depart from a smooth
continuous distribution. As [Real data](#real-data) below shows in detail,
the primary real table's responses heap not just at the scale ceiling but
at the scale *midpoint and quarter-points* too -- a multi-modal pattern no
unimodal density, bounded or not, is built to represent. This condition
tests that directly: with probability 0.4 per cell, `sim_beta_irt()`'s
continuous draw is thrown away and replaced with a value drawn uniformly
at random from 5 fixed points spanning the range ($0, 0.25, 0.5, 0.75, 1$),
independent of the underlying $\theta$ (`apply_response_heaping()` in
`continuous_bounded_helpers.R`) -- same base condition as boundary
inflation above ($N=300$, $J=8$, Beta IRT DGP, fixed dispersion), so the
two are directly comparable.

```{r}
#| label: fig-grid-heaping
if (grid_cache_ok) {
  grid_res$scores |>
    filter(dgp == "beta_irt", N == 300, J == 8, dispersion_mode == "fixed", !boundary_inflation) |>
    mutate(model = factor(model, levels = names(model_colors)),
           condition = ifelse(response_heaping, "40% round-number heaping", "Not heaped")) |>
    ggplot(aes(x = model, y = mean_ll, fill = model, alpha = condition)) +
    geom_col(position = "dodge", width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_alpha_manual(values = c("Not heaped" = 1, "40% round-number heaping" = 0.55), name = NULL) +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability",
         title = "Beta IRT DGP (N=300, J=8): held-out fit with vs. without round-number heaping") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

```{r}
#| label: calc-heaping-degradation
if (grid_cache_ok) {
  heap_calc <- grid_res$scores |>
    filter(dgp == "beta_irt", N == 300, J == 8, dispersion_mode == "fixed", !boundary_inflation) |>
    select(model, response_heaping, mean_ll) |>
    tidyr::pivot_wider(names_from = response_heaping, values_from = mean_ll,
                        names_prefix = "heap_") |>
    mutate(pct_change = 100 * (heap_TRUE - heap_FALSE) / abs(heap_FALSE))
}
```

**Every spec degrades under heaping, naive linear is the worst of the 5,
but the gap is modest -- not the dramatic split boundary inflation
produces.** Naive linear degrades by
`r if (grid_cache_ok) sprintf("%.0f%%", -heap_calc$pct_change[heap_calc$model == "naive_linear"]) else "NA"`,
against
`r if (grid_cache_ok) sprintf("%.0f-%.0f%%", min(-heap_calc$pct_change[heap_calc$model != "naive_linear"]), max(-heap_calc$pct_change[heap_calc$model != "naive_linear"])) else "NA"`
for the other 4 -- naive linear is worst, but by a few percentage points,
not the roughly 2x gap boundary inflation produced (20% vs. 8-10%). On
`rmse_pred`, there isn't even a consistent naive-linear penalty: all 5
specs degrade by 41-55%, and naive linear (46%) is squarely in the middle
of that range, not at either end.

Flagging this now, ahead of the real data: [the real-data comparison
below](#real-data) will show naive linear catastrophically worse on one
table (`lsbq`) and fully competitive on another
(`ai_fear_dong_2026_ai`) that piles up at its ceiling almost as severely --
a real difference this simulated condition, designed specifically to test
that mechanism, does **not** come close to reproducing. A few possible
reasons, none confirmed here: (1) this condition replaces heaped cells
with responses
drawn *independent of $\theta$* -- pure noise -- while real heaping
likely still correlates with the respondent's true trait level (someone
who rounds to "10" is probably genuinely higher-ability, not a random
draw); a noise-injection DGP may punish every model's predictive ability
fairly evenly, precisely because none of the 4 specs are built to exploit
theta-correlated structure in heaped cells either way. (2) A "snap each
value to its nearest anchor" version of this same function was tried
first (rounding, which *does* preserve the theta correlation) and
degraded every spec by under 3% -- rounding to 1 of 5 evenly-spaced
points moves a value by at most 0.125, a far gentler perturbation than
boundary inflation's full jump to 0 or 1, so it barely registered. Neither
version bridges the gap to what `lsbq` actually shows. (3) `lsbq`'s
heaping is concentrated much more heavily at the ceiling specifically
(44% at exactly 10, vs. 9-15% at the interior quarter-points) than this
condition's evenly-weighted 5-point design, so the real table may be
functioning more like an especially severe version of boundary inflation
with some extra interior noise on top, rather than a cleanly separate
"multi-modality" phenomenon. Untangling which of these actually explains
the real-data gap would need its own dedicated simulation design, not
attempted here.

### Predicted responses in simulation: point estimates, not just probabilities

Held-out log-probability answers "how much probability mass did the model
put on the interval the true response landed in" -- a distributional
question. It doesn't directly answer the more concrete question the site's
binary-response vignettes usually lead with (e.g. the [asymmetric IRT
vignette](asymmetric_irt.html)'s predicted-vs-actual convention): if you
had to guess the actual missing
value, how close would you get? So alongside every log-probability number
above, each model's fitted parameters and estimated theta also produce a
genuine **point prediction** for the same held-out cells -- $E[X_{ij}\mid
\theta_i]$, the closed-form or numerically-integrated mean derived in
[Model equations](#model-equations) for each spec -- compared directly to
the actual masked value via RMSE and correlation. The same
`rmse_pred`/`cor_pred` metrics were computed for every condition in [the
full recovery grid](#the-full-recovery-grid-16-conditions-x-3-dgps) above:

```{r}
#| label: fig-grid-rmse-crossfit
if (grid_cache_ok) {
  grid_res$scores |>
    filter(!boundary_inflation) |>
    mutate(model = factor(model, levels = names(model_colors)),
           dgp_label = recode(dgp, beta_irt = "DGP: Beta IRT", samejima_crm = "DGP: Samejima CRM",
                               muller_corsm = "DGP: Mueller CRSM")) |>
    group_by(dgp_label, model) |>
    summarise(rmse_pred = mean(rmse_pred, na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = model, y = rmse_pred, fill = model)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    facet_wrap(~dgp_label, ncol = 3) +
    labs(x = NULL, y = "Mean RMSE of predicted vs. actual response\n(averaged across N/J conditions)",
         title = "Cross-fitting: predicted-response RMSE by model, averaged across all N/J conditions per DGP") +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

Where the log-probability cross-fit chart showed real (if small) separation
between the 3 bounded-response models and a large, consistent gap to naive
linear, RMSE shows almost none of that: all 5 specs land within about 0.01
of each other under every DGP (e.g. under the Beta IRT DGP, everything is
within $[0.217, 0.222]$; under Samejima CRM's own DGP, within $[0.176,
0.182]$). Naive linear is never the worst of the 5 on this metric, and under
Mueller CRSM's own DGP it's actually the numerically *best* (0.270 vs.
Mueller CRSM's own 0.278). Point-prediction accuracy simply isn't where
these 4 specs differ; distributional fit is -- the same disconnect shows up
again on real data, [below](#predicted-responses-in-real-data-point-estimates-not-just-probabilities).

Boundary inflation degrades RMSE for every spec by a roughly similar amount
too (Beta IRT DGP, $N=300$/$J=8$: naive linear 0.229 $\to$ 0.359, the 3
bounded-response models 0.229-0.235 $\to$ 0.351-0.378) -- unlike the
log-probability version of this comparison, there's no clear asymmetry
where naive linear degrades disproportionately more. That asymmetry was a
distributional-fit story (a Gaussian's tails handle boundary piling worse
than the other 3 models' densities do); it doesn't show up in a metric that
only scores the point estimate.

### Summary

- **Parameter recovery works.** Every correctly-specified model recovered
  its true item-difficulty parameters cleanly across all 16 non-inflation
  grid conditions ($r>0.96$, mostly $r>0.99$) -- the fitting code itself is
  sound.
- **On distributional fit (held-out log-probability), the 3
  bounded-response models beat naive linear consistently, regardless of
  which of the 3 generated the data**, and are close to interchangeable
  with each other: "correctly specified wins" holds only weakly, and under
  the Mueller CRSM DGP it doesn't hold at all (Samejima CRM edges out
  Mueller's own generating model).
- **On point predictions (RMSE), the 3 bounded-response models bring no
  advantage over naive linear** -- all 4 specs land within about 0.01 of
  each other under every DGP, and naive linear is never the worst. Point
  estimates and distributional fit are answering different questions, and
  only the second one is where these models actually differ.
- **Boundary inflation hurts every spec's log-probability, but hurts naive
  linear roughly twice as much** (~20% degradation vs. ~8-10% for the 3
  bounded-response models) -- none of the 4 specs has a dedicated
  zero-one-inflated component, so none handles it *well*, but the 3
  bounded-response models absorb it better. RMSE shows no such asymmetry.
- **Round-number heaping (multi-modal clumping at the floor, midpoint, and
  quarter-points, simulated as noise independent of $\theta$) degrades
  every spec, with naive linear worst but only modestly so** (~17% vs.
  ~12-15% for the 3 bounded-response models) -- a real but much smaller
  gap than boundary inflation's roughly 2x split, and one that doesn't
  come close to reproducing the size of the real-data gap it was designed
  to test (see [Real data](#real-data) below). What specifically closes
  that gap in real data -- theta-correlated heaping, ceiling-dominant
  heaping severity, or something else -- isn't resolved here.

Taken together: under known, controlled conditions, the 3 bounded-response
models do what they're built to do -- they fit the shape of a bounded
response better than a model that ignores the bounds, especially once
responses pile up at the boundary or heap at multiple round points at
once. Whether that translates to real IRW tables, where the true
generating process isn't known and neither dial is under our control, is
the question [Real data](#real-data) below takes up next.

## Real data {#real-data}

Simulation above established that every correctly-specified model recovers
its parameters and that the 3 bounded-response models beat naive linear on
held-out fit under known, controlled conditions. This section checks
whether the same patterns hold on real IRW tables, where the true
generating process is unknown.

### Real-data selection

Candidate tables were identified by scouting IRW for tables with
`irw_metadata()$n_categories >= 12` whose response range and `irw_tags()`
item-format both point to a bounded, fixed-scale response, then narrowed to
tables that pass structural fitness checks for these 4 models specifically
(unique `(id, item)` pairs -- reducing to a single wave/rater where IRW's
`wave`/`rater` columns require it, no ipsative/compositional row-sum
constraint, and genuine fractional-value evidence rather than a merely
fine-grained integer scale). 11 tables survive and are used for every
real-data result below -- 7 with all 4 core specs fitting cleanly, and 4
more (added in a later pass) where one spec fails to fit for a specific,
documented reason (see [Limitations](#limitations)) but the rest of the
comparison remains valid:

| Table | N | J | Note |
|---|---|---|---|
| `lsbq_maleki_2025_non_persian_proficiency` | 308 | 3 | 0-10 scale, ~40% non-integer responses -- primary evidence of genuine continuity |
| `ai_fear_dong_2026_ai` (Doctor facet, subsampled) | 2,000 | 8 | 0-100, integer-valued; large-*N* structural check |
| `much_tte_2025_currentmotivation` | 1,244 | 3 | wave-filtered to one occasion |
| `test_taking_much_2025_cm` | 1,244 | 9 | -- |
| `nas_rogoza_2024_study5_nas` | 277 | 4 | wave-filtered; Narcissistic Antagonism subscale |
| `nas_rogoza_2024_study5_ngs` | 274 | 4 | wave-filtered; Narcissistic Grandiosity subscale |
| `nas_rogoza_2024_study5_nvs` | 278 | 4 | wave-filtered; Narcissistic Vulnerability subscale |
| `mendes_2019_snycq` | 157 | 12 | wave-filtered; 0-100, ~27% non-integer; Samejima CRM fails to fit |
| `westhoff2023_stopd` | 109 | 5 | wave-filtered; Beta IRT fails to fit (both variants) |
| `westhoff2023_pbat` | 109 | 18 | wave-filtered; Samejima CRM fails to fit |
| `gilbert_meta_95` | 115 | 12 | wave-filtered; Beta IRT fails to fit (both variants) |

`ai_fear_dong_2026_ai` is a 48-item, 6-target x 8-trait factorial design
(Care/Doctor/Journalist/Judge/Manager/Religious x competent/warm/etc.); only
one target's 8 items (`Doctor_*`) are used here as a single "overall
impression of Doctor" construct, not the full table.

Full scouting and per-table vetting code, cached candidate lists, and the
exact reason every non-included candidate was set aside all live in
`vignettes/continuous_scout_*` and `vignettes/continuous_bounded_data/
vet_candidates*.R` / `vet_results*.csv` -- start there rather than
re-deriving the logic if extending this vignette to more tables.

### Real data: what do these response distributions actually look like?

```{r}
#| label: fig-real-dist
if (cache_ok) {
  as.data.frame(res$real$Y_raw) |>
    tidyr::pivot_longer(everything(), names_to = "item", values_to = "resp") |>
    ggplot(aes(x = resp)) +
    geom_histogram(binwidth = 0.5, boundary = 0, fill = irw_blue, color = "white", linewidth = 0.1) +
    facet_wrap(~item, ncol = 3) +
    labs(x = sprintf("Raw response (%g-%g scale)", res$real$low, res$real$high),
         y = "Count",
         title = sprintf("%s:\nraw response distribution by item", res$real$table)) +
    theme(plot.title = element_text(size = 12))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

The primary table's histogram above isn't just "a hump plus a boundary
spike" -- look at which *exact values* are most common, not just the
bins. For all 3 items, the top 5 response values by frequency are, in
order, **10 (the ceiling), 5 (the midpoint), 7.5, 2.5, and 0** -- the 5
quarter-points of the 0-10 scale, in descending order of how salient each
one is as a "round number." On item `17.1.S`, that's 44.5% of responses
at exactly 10 and another 9.1% at exactly 5; combined, over half of every
item's responses land on one of these 5 points. This is **response
heaping**: respondents (or whatever process generated these ratings)
gravitating to salient round fractions of the scale, not spreading
continuously across it -- a well-documented phenomenon in self-report
scales, here compounded by the fact that one of the 5 heap points happens
to sit at the scale's hard boundary. The result is a genuinely
**multi-modal** response distribution, and multi-modality is a problem
none of this vignette's 4 specs is built to solve -- Beta IRT, Samejima
CRM, and Mueller CRSM's conditional response densities are just as
unimodal as naive linear's Gaussian is, only bounded and more flexibly
shaped.

That distinction -- boundary piling vs. multi-modal heaping across the
whole scale -- turns out to matter a lot, and a second table makes the
contrast concrete. `ai_fear_dong_2026_ai` (Doctor facet, 0-100 scale, 8
items, $N=2{,}000$) also piles up near its ceiling, comparably severely
on some items, but its next-most-common values are 98, 99, 97, 96,
95, 90 -- a smooth decay of nearby-to-the-ceiling integers, not a
separate spike at the scale's midpoint:

```{r}
#| label: fig-real-dist2
if (cache_ok) {
  as.data.frame(res$real2$Y_raw) |>
    tidyr::pivot_longer(everything(), names_to = "item", values_to = "resp") |>
    ggplot(aes(x = resp)) +
    geom_histogram(binwidth = 5, boundary = 0, fill = irw_green, color = "white", linewidth = 0.1) +
    facet_wrap(~item, ncol = 4) +
    labs(x = sprintf("Raw response (%g-%g scale)", res$real2$low, res$real2$high),
         y = "Count",
         title = sprintf("%s (%s facet):\nraw response distribution by item", res$real2$table, res$real2$facet)) +
    theme(plot.title = element_text(size = 12),
          strip.text = element_text(size = 7))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

Counting the same top-of-scale histogram bin for both tables, `lsbq`'s 3
items put 22-47% of responses in the top 5% of the range (mean 31%);
`ai_fear_dong_2026_ai`'s 8 items range 14-46% (mean 27%) -- close enough
that peak boundary-bin height alone doesn't distinguish them. Despite
that similarity, naive linear's held-out log-probability is night and day
between the two: -1.50 on `ai_fear_dong_2026_ai`, in the same range as
the 3 bounded-response models (-1.18 to -1.31), against `lsbq`'s
catastrophic -6.08 (see [Held-out fit across all 11 real
tables](#held-out-fit-across-all-11-real-tables) below). Multi-modal
heaping, present on `lsbq` and absent on `ai_fear_dong_2026_ai`, is the
most obvious candidate explanation, but it's a hypothesis, not a
demonstrated mechanism: [the Simulation section's response-heaping
condition](#response-heaping-does-round-number-clumping-break-anything),
built specifically to test it, only reproduces a small fraction of this
gap. Something about real heaping -- its correlation with $\theta$, its
concentration at the ceiling specifically, or some combination -- matters
in a way the simulated version doesn't fully capture; see that section
for the actual numbers and what was tried.

### Empirical information curves: the primary real table

The [interactive widget in Model equations](#comparing-information-curves-across-models)
uses illustrative, hand-set item parameters. Whether the same divergence
shows up with *actually fitted* item parameters is a separate, more
useful question -- so all 4 core specs (5 counting both Beta IRT
dispersion variants) were refit on the full primary real table,
`lsbq_maleki_2025_non_persian_proficiency` (3 items, $N=308$; no train/test
split here, since this is about what the fitted item parameters imply, not
held-out prediction), and each fitted item's Fisher information curve
computed the same numerical way as the widget above (central-difference
score, adaptive quadrature over $X$) -- see
`continuous_bounded_info_compute.R`.

```{r}
#| label: fig-real-info-curves
#| fig-width: 9
#| fig-height: 5
if (info_cache_ok) {
  ic <- info_res$info_curves |>
    mutate(model = factor(model, levels = names(model_colors)))
  # naive linear's item 17.1.W is a Heywood case (see note below) whose
  # information is ~7 orders of magnitude larger than every other
  # curve -- clipping the y-axis (not the underlying data) so the other
  # 14 curves stay legible; the clipped point is called out in the prose
  # just below instead of a plot caption, which would clip the same way
  # the title/legend did at this facet width.
  ic |>
    ggplot(aes(x = theta, y = info, color = model)) +
    geom_line(linewidth = 0.8) +
    scale_color_manual(values = model_colors, labels = model_labels, name = NULL) +
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
    coord_cartesian(ylim = c(0, 8)) +
    facet_wrap(~item, ncol = 3) +
    labs(x = expression(paste("Ability (", theta, ")")), y = "Fisher information I(θ)",
         title = "Empirical item information, all 4 fitted specs",
         subtitle = info_res$table) +
    theme(legend.position = "bottom", plot.title = element_text(size = 13))
} else {
  cat("Info cache not found -- run continuous_bounded_info_compute.R first.")
}
```

Naive linear's curve for item `17.1.W` is off this scale entirely
(~$1.2\times10^7$, a Heywood case -- see below) and isn't visible in the
plot above; the y-axis is clipped at 8 to keep the other 14 curves legible.
Three things stand out, confirming the widget's story on real fitted
items and adding one it can't show:

- **Naive linear and Samejima CRM are exactly flat here too, on every one
  of the 3 real items** (their `min`/`median`/`max` in the underlying data
  agree to 4+ significant figures per item -- see
  `continuous_bounded_info_compute.R`'s output). This is the same
  transform-invariance fact from the widget above, not a coincidence of
  round slider values: CRM's information genuinely doesn't depend on which
  real item or table it's fit to, only on that item's own $\alpha_j/\sigma_j$
  ratio.
- **Beta IRT is the only spec that visibly peaks and tapers across all 3
  items, and *where* it peaks moves a lot from item to item.** Item-varying
  dispersion peaks around $\theta\approx-0.75$ on `17.1.R`, $\theta\approx
  -0.25$ on `17.1.W`, but $\theta\approx-2.5$ on `17.1.S` -- the same model
  spec, refit per item, locating its most-informative region in
  substantially different places. Mueller CRSM is technically non-constant
  too on every item, but far more mildly (its range is under a factor of
  1.6 on each item, against Beta IRT's factor of roughly 2.3-7 across the
  2 dispersion variants and 3 items), consistent with
  the widget's own comparison above.
- **Naive linear's item `17.1.W` fit is degenerate.** Its CFA residual
  variance estimate for that item is slightly *negative*
  ($\widehat{\sigma}^2\approx -0.006$, a lavaan Heywood case -- the fit
  itself warns `some estimated ov variances are negative`), which the
  information formula ($a^2/\sigma^2$ for this model, floored at a tiny
  positive value once $\sigma^2$ is clamped away from 0) turns into an
  information curve on the order of $10^7$ -- not a real measurement
  property, just a division-by-near-zero artifact of an unstable variance
  estimate on this particular item. This is a genuine failure mode the
  hand-set widget above can't show (its sliders never let $\sigma\to0$),
  and a concrete reason to distrust naive linear's information (not just
  its distributional shape, already shown unreliable elsewhere in this
  section) on any item where its own fit is this close to a boundary case.

This vignette's other 10 real tables were not checked for the same
pattern -- this is a one-table illustration, not a systematic audit of
every fitted item across all 11 tables.

### Held-out fit across all 11 real tables

```{r}
#| label: prep-real-scores
if (cache_ok) {
  all_real_scores <- dplyr::bind_rows(
    res$real$scores |> dplyr::mutate(N = res$real$N, J = res$real$J),
    res$real2$scores |> dplyr::mutate(
      N = res$real2$N, J = res$real2$J,
      label = paste0(label, " (", res$real2$facet, ")")
    ),
    res$batch$scores
  )
}

# One row per model, one point per table on that model's row -- lets you
# hover a point to see which table/condition it is, rather than reading
# facet titles across a grid of small bar charts.
make_strip_plot <- function(df, value_col, x_lab, title_txt) {
  df <- df |>
    dplyr::mutate(
      model = factor(model, levels = names(model_colors)),
      model_label = factor(model_labels[as.character(model)],
                            levels = rev(model_labels[names(model_colors)])),
      tbl_label = sprintf("%s (N=%d, J=%d)", label, N, J),
      value = .data[[value_col]]
    )
  # title is set via plotly::layout() below, not ggplot's own labs(title=...):
  # ggplotly() converting a ggplot title into a plotly annotation was
  # clipping the text at the plot edge. The modebar (camera/zoom icon row)
  # is disabled below too -- at narrower viewport widths it overlaps the
  # right end of a long title, obscuring the last word or two; these are
  # simple exploratory charts that don't need pan/zoom/export controls.
  p <- ggplot(df, aes(x = value, y = model_label, color = model,
                       text = sprintf("%s<br>%s: %.3f", tbl_label, x_lab, value))) +
    geom_point(size = 2, alpha = 0.85,
               position = position_jitter(height = 0.1, width = 0, seed = 1)) +
    scale_color_manual(values = model_colors, guide = "none") +
    labs(x = x_lab, y = NULL)
  plotly::ggplotly(p, tooltip = "text", height = 300) |>
    plotly::layout(
      title = list(text = title_txt, x = 0.02, xanchor = "left", font = list(size = 14)),
      margin = list(t = 45, l = 150, r = 20, b = 45),
      hoverlabel = list(align = "left")
    ) |>
    plotly::config(displayModeBar = FALSE)
}
```

```{r}
#| label: fig-real-all-scores
if (cache_ok) {
  make_strip_plot(all_real_scores, "mean_ll",
                   "Mean held-out log-probability (per response)",
                   "Held-out fit by model, across all 11 real IRW tables analyzed")
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

Each row above is one model; each point is one of the 11 real tables --
hover a point for the table name and exact value. The pattern is
consistent across all 11 tables and matches the simulation results above:

- **The naive linear baseline is clearly worst everywhere it's compared**,
  often by a wide margin -- e.g. on the primary
  `lsbq_maleki_2025_non_persian_proficiency` table, naive linear's mean
  held-out log-probability is -6.08, against -1.13 to -2.51 for the 3
  bounded-response models. Peak boundary-bin height alone doesn't explain
  the gap: `ai_fear_dong_2026_ai` has comparably severe *peak* ceiling
  piling on some items yet naive linear scores -1.50 there, close to the
  3 bounded-response models' own range. Multi-modal heaping -- `lsbq` also
  has a secondary spike at the scale midpoint, `ai_fear_dong_2026_ai`
  doesn't -- is the leading candidate explanation (see [the distribution
  comparison](#real-data-what-do-these-response-distributions-actually-look-like)
  above), though a simulation built to test it only partially confirms it.
- **Among the 7 tables where all 3 bounded specs fit successfully,
  Samejima CRM is the worst of the 3 on every single one** -- sometimes by
  a wide margin (`lsbq`: -2.51 vs. -1.13 to -1.51 for the other two),
  sometimes narrowly (`ai_fear_dong_2026_ai`: -1.31 vs. -1.18).
- **Beta IRT (either dispersion variant, interchangeably) wins outright on
  6 of the 11 tables**: `lsbq`, `ai_fear_dong_2026_ai`,
  `much_tte_2025_currentmotivation`, `test_taking_much_2025_cm`,
  `mendes_2019_snycq`, and `westhoff2023_pbat` -- the last 2 are tables
  where Samejima CRM wasn't fit at all, see below.
- **Mueller CRSM wins the 3 narcissism subscales** from
  `nas_rogoza_2024_study5` (`nas`/`ngs`/`nvs`).
- **Samejima CRM's only 2 wins** (`gilbert_meta_95`, `westhoff2023_stopd`)
  are both tables where Beta IRT itself failed to fit -- among tables
  where all 3 bounded specs were actually in the running, Samejima never
  wins. One plausible reason: its held-out score depends on an
  empirically-estimated residual SD on the transformed $Z$ scale (see
  `.crm_resid_sd()` in `continuous_bounded_helpers.R`, chosen because
  EstCRM's own EM-shrinkage precision term isn't straightforward to
  reverse-engineer), which may generalize to genuinely new items less well
  than the other 2 specs' fitted dispersion parameters do -- a real cost
  of the workaround, not asserted with certainty.

Each table has only one train/test split (see [Limitations](#limitations)).
The main takeaway holds regardless of which of the 3 wins: all 3
bounded-response models consistently and substantially beat the naive
linear baseline across every real table analyzed, on genuinely held-out
item responses.

### Predicted responses in real data: point estimates, not just probabilities

As in [Simulation](#predicted-responses-in-simulation-point-estimates-not-just-probabilities)
above, held-out log-probability only answers "how much probability mass did
the model put on the interval the true response landed in" -- a
distributional question, not the more concrete "if you had to guess the
actual missing value, how close would you get?" So alongside every
log-probability number above, each model's fitted parameters and estimated
theta also produce a genuine **point prediction** for the same held-out
cells -- $E[X_{ij}\mid\theta_i]$, the closed-form or numerically-integrated
mean derived in [Model equations](#model-equations) for each spec --
compared directly to the actual masked value via RMSE and correlation.

```{r}
#| label: fig-real-rmse
if (cache_ok) {
  make_strip_plot(all_real_scores, "rmse_pred",
                   "RMSE of predicted vs. actual response (held-out)",
                   "Predicted-response accuracy by model, across all 11 real IRW tables")
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

Same layout as above -- one row per model, one point per table, hoverable.
This is where the story genuinely changes:

- **Naive linear's point predictions are competitive with, and often
  better than, all 3 bounded-response models** -- the opposite of the
  log-probability picture. It has the *lowest* RMSE of all specs it's
  compared against on 7 of the 11 tables: `much_tte_2025_currentmotivation`
  (0.0999), `test_taking_much_2025_cm` (0.1048),
  `nas_rogoza_2024_study5_ngs` (0.1799), `westhoff2023_stopd` (0.1684),
  `westhoff2023_pbat` (0.1970, vs. Beta IRT's 0.3302 -- a large gap in
  naive's favor on that table specifically), `mendes_2019_snycq` (0.2666),
  and, most strikingly, the primary
  `lsbq_maleki_2025_non_persian_proficiency` table (0.2238, ahead of
  Samejima CRM's 0.2331, Beta IRT's 0.2459/0.2498, and Mueller's 0.2680).
- **That's despite naive linear having by far the worst log-probability
  everywhere**, including a catastrophic -6.08 on `lsbq` against -1.13 to
  -2.51 for the 3 bounded-response models on that same table.
- **It's essentially tied for best on `ai_fear_dong_2026_ai`** (0.1961 vs.
  Beta IRT's 0.1958), and only clearly beaten on 3 of the 11 (`nas` and
  `nvs`, by Mueller CRSM and Beta IRT respectively; `gilbert_meta_95`, by
  Samejima CRM).

This isn't a contradiction -- it's what the two metrics are built to catch.
RMSE only checks whether the *central tendency* is right; log-probability
also checks the *shape* of the distribution around it (tail behavior,
non-constant variance near the boundaries, how much mass piles up at 0
and 1). A linear model's conditional mean can be a perfectly reasonable
guess at where a response will land even while its Gaussian, constant-variance
error structure is badly wrong about how *confident* to be near a boundary --
exactly the boundary and midpoint heaping shown in [the histograms
above](#real-data-what-do-these-response-distributions-actually-look-like).
So: if the applied question is "predict this person's missing response as a
single number," the extra machinery of a bounded-response model buys little
to nothing here. If the applied question involves *uncertainty* --
uses fitted probabilities, is calibration-sensitive, or the response
sits near a boundary and getting the shape right actually matters -- the
naive baseline's advantage disappears entirely and the 3 bounded models
are the clear right choice.

```{r}
#| label: fig-real-pred-scatter
if (cache_ok) {
  res$real$predictions |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = actual, y = predicted, color = model)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(alpha = 0.6, size = 1.6) +
    scale_color_manual(values = model_colors, labels = model_labels, guide = "none") +
    facet_wrap(~model, ncol = 3, labeller = as_labeller(model_labels)) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(x = "Actual (held-out) response", y = "Predicted response",
         title = sprintf("Predicted vs. actual held-out response:\n%s", res$real$table)) +
    theme(plot.title = element_text(size = 12),
          panel.spacing = unit(1.4, "lines"))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

The scatter for the primary table makes the RMSE numbers concrete: all 5
specs track the dashed 45-degree line reasonably well across the middle of
the range, and none of them is dramatically better calibrated to the *point
values themselves* than the others -- the real differentiator among models
in this vignette is the log-probability metric, not this one. That's just
one table, though -- the strip plot below shows the same predicted-vs-actual
correlation (`cor_pred`) for every model on every one of the 11 real
tables, the same layout as the RMSE and log-probability strip plots above:

```{r}
#| label: fig-real-cor
if (cache_ok) {
  make_strip_plot(all_real_scores, "cor_pred",
                   "Correlation of predicted vs. actual response (held-out)",
                   "Predicted-response correlation by model, across all 11 real IRW tables")
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

Same story as the RMSE strip plot: **all 5 specs land in a broadly
overlapping band on this metric**, with no consistent separation between
the 3 bounded-response models and naive linear the way there is on
log-probability. Naive linear isn't even reliably the worst here --
consistent with the point-prediction metrics simply not being where
these models differentiate themselves.

The same `rmse_pred`/`cor_pred` metrics were computed for every condition
in [the full recovery grid](#the-full-recovery-grid-16-conditions-x-3-dgps)
in the [Simulation](#simulation) section above, including under boundary
inflation -- the same disconnect between RMSE/correlation and log-probability
held up there too: naive linear was competitive or better on point
predictions despite its consistently worse log-probability.

### Summary

- **The 3 bounded-response models consistently and substantially beat
  naive linear on held-out log-probability, on every one of the 11 real
  tables** -- the simulation's central finding holds up on real data, not
  just under known, controlled conditions.
- **Among the 3 bounded-response models, no single one dominates.** Beta
  IRT wins outright on 6 of 11 tables, Mueller CRSM wins the 3 narcissism
  subscales, and Samejima CRM -- worst of the 3 whenever all 3 are
  actually in the running -- only "wins" on the 2 tables where Beta IRT
  itself failed to fit. As in simulation, these 3 are more interchangeable
  with each other than any of them is with naive linear.
- **How much worse naive linear does varies enormously by table, and
  tracks response heaping more than boundary piling alone.** `lsbq`'s
  catastrophic -6.08 and `ai_fear_dong_2026_ai`'s mild -1.50 have
  comparably severe *peak* boundary piling; the difference is `lsbq`'s
  extra multi-modal heaping at the scale midpoint. [Simulation's
  response-heaping
  condition](#response-heaping-does-round-number-clumping-break-anything),
  built to test exactly this, only partially reproduces it -- a real,
  named, still-open question this vignette raises but doesn't close.
- **On point-prediction metrics (RMSE and correlation), naive linear is
  competitive with, and sometimes better than, all 3 bounded-response
  models** -- the reverse of the log-probability picture, and consistent
  with the same disconnect found in simulation. If the applied question
  needs calibrated uncertainty, the 3 bounded-response models are the
  clear choice; if it's just a point guess, the extra machinery buys
  little here.
- **2 of the 4 specs fail to fit outright on some real tables**, for
  specific, diagnosed reasons (Samejima CRM's EM on 2 tables, Beta IRT's
  category-compaction issue on 2 different tables) -- documented in
  [Limitations](#limitations) below, not silently routed around.

## Limitations {#limitations}

Known gaps this vignette does not yet address:

- **Ferrando's (2001) nonlinear congeneric model** [@ferrando2001nonlinear]
  -- the optional stretch spec, comparable to the treatment in Ferrando
  (2002) [@ferrando2002theoretical] -- has not been attempted at all. Specs
  1-4 are now complete and validated (the grid ran cleanly), so its own
  sequencing rule no longer blocks it -- it just hasn't been picked up yet.
- **Partial missingness in real data.** All 11 real tables used complete
  cases only *before* the evaluation design's own synthetic masking is
  applied; IRW tables whose respondents have naturally-occurring partial
  missingness (as opposed to the missingness this vignette introduces
  on purpose, for evaluation) aren't handled yet.
- **The 25% per-person item-masking fraction (`MASK_FRAC`) was not
  varied or sensitivity-tested** -- a different fraction could shift how
  much information theta estimation has to work with, and thus the
  absolute (though probably not the relative, across-model) held-out
  numbers above.
- **Only one train/test split per condition** (real or simulated), not a
  repeated/cross-validated design -- so none of the held-out numbers above
  have an uncertainty band, and a single unlucky split could shift a close
  call between the 3 bounded-response models.
- **Samejima CRM's EM is prone to failing outright, not just fitting
  poorly.** One grid condition (Mueller-CRSM-generated, $N=300$, $J=8$, no
  boundary inflation) hit `missing value where TRUE/FALSE needed` during
  EM. This isn't a one-off: the identical error recurred on 2 of the 4 real
  tables added in the follow-up vetting pass, `westhoff2023_pbat` and
  `mendes_2019_snycq` -- both otherwise clean, wave-filtered, structurally
  valid tables (see [Real-data selection](#real-data-selection)). None of
  these were re-run with a different seed or diagnosed further; reported as
  a real, recurring misspecification/convergence fragility in `EstCRM`'s EM
  rather than something to route around.
- **Beta IRT fitting fails outright on some real tables, for a specific,
  confirmed reason.** `westhoff2023_stopd` and `gilbert_meta_95` (both
  dispersion variants) fail with `response.pattern input contains more
  response categories than defined by the model` -- `mirt` compacts each
  training split's observed categories to consecutive integers per item,
  and a held-out person whose response lands in a category the training
  split never populated gets rejected outright rather than scored as
  missing. This is the same failure mode described just below (originally
  found and fixed for a different table during development); it simply
  wasn't fixable for these 2 by lowering the bin count the way it was
  elsewhere. Confirmed by actually running `fit_score_all()` on both tables
  (2026-08-01), not asserted from an earlier, untested claim.
- **The grid's boundary-inflation condition covers only 1 of the 3 DGPs**
  (Beta IRT, $N=300$/$J=8$) -- not repeated for Samejima CRM or Mueller
  CoRSM, or at other N/J combinations.
- **Potential extension: the `BoundIRT` R package** (Molenaar, CRAN, July
  2026; documented in a preprint aiming for journal submission
  [@molenaar2026boundirt]) fits Beta IRT, Samejima IRT, Simplex IRT, and
  Normal IRT under one Bayesian (Stan/MCMC) framework, which would let the
  held-out comparison above be replaced with a principled fully-marginalized
  log-likelihood (bridge sampling) or LOOIC/WAIC -- a direct fix for the ad
  hoc bin-probability metric flagged just below -- and it natively models
  zero/one inflation (Molenaar et al.'s (2022) extension [@molenaar2022zero]), which none of the
  4 specs here handle despite the boundary piling documented above. It does
  not implement Mueller's CRSM, so that spec would still need the current
  `pcIRT`-based code, and MCMC fitting is considerably slower than the
  current MLE/EM approach across 11 real tables and a 17-condition
  simulation grid. Not attempted here; worth a small pilot (refit Beta IRT
  + Samejima on the primary `lsbq` table and compare) before a full
  rewrite.

**Package-availability and correctness issues found while building this**
(worth knowing before extending this vignette):

- `pcIRT::CRSM()` (Mueller's model) was **dropped from the package
  entirely** in the last CRAN release (0.2.4, 2019); it only exists in
  pcIRT $\leq$ 0.2.3. `pcIRT` itself was later archived by CRAN entirely
  (2025-11-11). This vignette's compute script installs 0.2.3 from the
  CRAN archive, not the current CRAN release -- a real risk for
  reproducing this vignette in the future if the archive changes.
- That same version's `CRSM()` has a convergence-check bug
  (`is.na(para) || ...` where `para` becomes a 2-vector after the first
  iteration) that is a hard error under R $\geq$ 4.3 and had to be patched
  locally before it would run at all.
- `sirt::brm.sim()` hardcodes a 9-item assumption in its column-naming
  code and errors for any other item count; `sim_beta_irt()` in the
  helpers file reimplements its generative logic directly rather than
  patching `sirt`.
- `pcIRT::person_par.CRSM()` only scores the same persons used in the
  original fit and has no new-data argument, so it could not be reused for
  scoring held-out test persons; a generalized version of its estimating
  equation was written from scratch instead (`theta_new_muller_corsm()`
  below) -- this turned out to be a genuine improvement over the original,
  since it also skips `NA` items per person, which is exactly what made
  cell-level missing-response prediction possible for this spec.
- The from-scratch Mueller CRSM person-scoring routine
  (`theta_new_muller_corsm()`) could diverge to a `theta` extreme enough to
  overflow `exp()` inside `integrate()`, which throws a hard "non-finite
  function value" error rather than the graceful NA `stop.on.error=FALSE`
  is supposed to allow -- this crashed the *entire* held-out scoring pass
  for a table (not just Mueller CRSM) before being caught and fixed by
  clamping `theta` and wrapping each `integrate()` call individually.
- Similarly, a Beta IRT scoring-stage failure (`mirt` compacts a training
  split's observed categories to consecutive integers per item, and rejects
  a held-out response that lands in a category with zero training-split
  observations) was originally written so it could crash
  naive-linear/Samejima/Mueller's results for the same table too -- every
  model's fit-and-score block is now wrapped in its own `tryCatch` so one
  model's failure can't take down the other 3.
- **The current 11 real tables are not necessarily the full set of viable
  candidates.** See the vetting scripts/CSVs referenced in
  [Real-data selection](#real-data-selection) for the working record of
  what's been checked so far.

**Explicitly out of scope** (not implemented, not planned for this
vignette): DIF for continuous responses [@finch2023impact]; dual-range
slider / interval-response models and the Interval Consensus Model
[@kloft2024measuring; @kloft2023dirichlet; @kloft2025interval] -- a
different response format (an interval, not a point) and, for the
Consensus Model, a different modeling paradigm (cultural consensus
theory) entirely; Zopluoglu's mixture CRM [@zopluoglu2020finite] and
Noel's beta unfolding model [@noel2014beta] -- each adds a full second axis
of complexity (latent mixture classes; unfolding vs. dominance response
process), and unfolding in particular is a more natural extension of the
site's existing unfolding-vs-dominance vignette than a duplicate here;
Ferrando and Navarro-Gonzalez's (2021) multidimensional Dual Thurstonian
model [@ferrando2021multidimensional].

**The comparison metric (held-out log-likelihood) is an ad hoc choice**,
not a validated best-practice for this setting -- flagged explicitly above,
worth revisiting.

## Reproducibility

```r
# 1. Run the compute script (from the project root)
source("vignettes/continuous_bounded_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/continuous_bounded.qmd")
```

::: {.callout-tip}
To cite the real-data table used here, run:

```r
irw_save_bibtex("lsbq_maleki_2025_non_persian_proficiency",
                 output_file = "continuous_bounded_data/irw_references.bib")
```
:::

---

## References
