---
title: "Bounded Continuous Responses: Comparing Beta IRT, Samejima's CRM, and Mueller's CRSM"
description: |
  Slider and visual-analogue-scale items produce a response that is
  continuous but lives in a fixed, interpretable range, like [0,1] or
  [0,100] -- not the open-ended continuous case standard CFA/linear-factor
  models assume. We compare four published ways of modeling that: a naive
  linear baseline that ignores the bounds, Noel and Dauvier's Beta IRT
  model, Samejima's Continuous Response Model, and Mueller's Continuous
  Rating Scale Model, on one simulated dataset and one real IRW table.
date: today
bibliography:
  - continuous_bounded_data/references.bib
  - continuous_bounded_data/irw_references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-important}
**This is a draft pilot, not the final vignette.** It exists to get feedback
on the modeling approach before investing in the full study. What's here:
real fits of all 4 core model specs, on one simulated dataset and one real
IRW table, with a working held-out comparison. What's *not* here yet: the
full simulation/recovery grid (multiple DGPs x sample sizes x item counts),
a boundary-inflation (zero/one-piling) condition, and Ferrando's (2001)
optional stretch model. See [Limitations](#limitations) for the complete
list and the reasoning behind several scope decisions made along the way.
:::

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 13))
irw_blue   <- "#2166ac"
irw_red    <- "#d73027"
irw_green  <- "#4daf4a"
irw_purple <- "#984ea3"
irw_grey   <- "#808080"

model_colors <- c(
  "naive_linear"              = irw_grey,
  "beta_irt_item_dispersion"  = irw_blue,
  "beta_irt_fixed_dispersion" = irw_purple,
  "samejima_crm"              = irw_red,
  "muller_corsm"              = irw_green
)
model_labels <- c(
  "naive_linear"              = "Naive linear",
  "beta_irt_item_dispersion"  = "Beta IRT (item disp.)",
  "beta_irt_fixed_dispersion" = "Beta IRT (fixed disp.)",
  "samejima_crm"              = "Samejima CRM",
  "muller_corsm"               = "Mueller CRSM"
)

cache_path <- "continuous_bounded_data/continuous_bounded_results.rds"
cache_ok <- file.exists(cache_path)
if (cache_ok) res <- readRDS(cache_path)

grid_cache_path <- "continuous_bounded_data/continuous_bounded_grid_results.rds"
grid_cache_ok <- file.exists(grid_cache_path)
if (grid_cache_ok) grid_res <- readRDS(grid_cache_path)
```

## Motivation

Visual analogue scales, slider ratings, and continuous confidence or
percentage judgments are common in IRW: a respondent marks a point on a
fixed, labeled segment (0 to 100, say), and the raw response is a
continuous number that is *bounded* at both ends. This is a different
measurement problem from the unbounded continuous case (response times,
open-ended magnitude estimates), which reduces to standard linear-factor
or CFA modeling and is covered elsewhere on this site -- the bounds
themselves are the interesting part here, because they create floor/ceiling
effects and non-constant response variance near the edges that a plain
linear model doesn't see.

Three published models handle this directly. Noel and Dauvier's Beta IRT
model [@noel2007beta] treats the response as Beta-distributed given
ability, with the two natural Beta parameters reparameterized into an
item difficulty and a dispersion. Samejima's Continuous Response Model
[@samejima1973homogeneous] models a monotonic transform of the response as
normal given ability -- the historically original approach, predating Beta
IRT by over 30 years. Mueller's Continuous Rating Scale Model
[@mueller1987rasch] takes a Rasch-family, specific-objectivity approach,
estimated by pairwise conditional likelihood; Verhelst's (2019) exponential
family framework [@verhelst2019exponential] is the modern theoretical
generalization of Mueller's model, though the model actually fitted here is
Mueller's original. We compare these three against a naive linear (Gaussian,
identity-link) baseline that ignores the bounds entirely -- the "what if you
just treat it like ordinary continuous data" control, playing the same role
the Rasch baseline plays in the site's guessing-models vignette.

## Data and methods

### The four core specs

1. **Naive linear baseline** -- single-factor CFA via `lavaan`, identity
   link, Gaussian residual, fit directly on the raw (rescaled-to-[0,1])
   response. Ignores boundedness entirely.
2. **Beta IRT** [@noel2007beta] -- fit as a discretized custom `mirt`
   itemtype wrapping `sirt::brm.irf()`, following `sirt`'s own documented
   mirt-estimation recipe (`?sirt::brm.sim`). Two variants: **item-varying
   dispersion** (each item gets its own dispersion parameter, the model's
   default parameterization) and **fixed dispersion** (one dispersion
   parameter shared across all items, via an `mirt` parameter constraint).
3. **Samejima's Continuous Response Model** [@samejima1973homogeneous] --
   `EstCRM::EstCRMitem()` / `EstCRMperson()`.
4. **Mueller's Continuous Rating Scale Model** [@mueller1987rasch] -- a
   patched copy of `pcIRT::CRSM()` / a from-scratch reimplementation of its
   person-scoring equation (see [Limitations](#limitations) for why
   patching was necessary).

Ferrando's (2001) nonlinear congeneric model [@ferrando2001nonlinear] is an
optional stretch spec with no package implementation (it would need a
custom truncated-normal likelihood via `optim`); per the original scope for
this vignette it's only attempted after specs 1-4 are complete and
validated, and **has not been attempted in this draft** -- see
[Limitations](#limitations).

### Why "interval probability," not raw log-likelihood

Before any model comparison, we checked whether all 4 models report
likelihood on the same scale. They don't: reading `EstCRM::EstCRMitem()`'s
source directly shows it transforms the raw response to
$Z = \ln\!\big(X/(K-X)\big)$ and fits/reports its EM log-likelihood entirely
on that $Z$-scale, with no Jacobian term anywhere in the function. A raw
log-likelihood from EstCRM is therefore not comparable to, say, the Beta
IRT model's density on the raw response, without a correction.

Rather than hand-deriving a per-model Jacobian correction, every model's
held-out score in this vignette is a **binned interval log-probability**:
$\log P(\text{bin}_{lo} < X < \text{bin}_{hi})$, using the same bin edges
(on the raw, rescaled-to-$[0,1]$ scale) for all 4 models. Interval
probability is invariant to a monotonic reparameterization of the
intervening likelihood -- the Jacobian exactly cancels under integration --
so this sidesteps the Jacobian bookkeeping entirely. This is a deliberate
substitution for "compare raw densities with a Jacobian correction," flagged
here because the original scope for this vignette didn't spell out this
exact resolution.

### Comparison metric: held-out log-likelihood, not IMV

We use held-out log-likelihood rather than IMV (InterModel Vigorish), the
metric used in most other vignettes on this site. `imv4sem` and
`imv.mirt` are not on CRAN, and no citable generalization of IMV to
continuous or Beta-type outcomes was found. **This is flagged as an ad hoc
decision to revisit, not a settled one** -- unlike IMV's binary/count case,
there's no obviously "natural" coin-flip baseline for a continuous
response, so held-out log-likelihood is a reasonable default rather than a
demonstrably right answer.

### Held-out design: person-level, not cell-level

The project's usual convention for held-out evaluation on IRW data is a
response-level (missing-cell) holdout: mask a random subset of person x item
cells, fit on the rest, score the masked cells. That isn't viable here for
2 of the 4 specs -- reading `EstCRM::EstCRMitem()` and `pcIRT::CRSM()`'s
fitting code directly shows **neither has any missing-data handling at
all** (both do matrix/pairwise algebra that silently breaks with `NA`
cells, confirmed by testing).

We substituted a **person-level train/test split** instead: fit item
parameters on a training subset of persons (a complete matrix, no missing
cells), then estimate ability for held-out test persons (from their own
full response vector, which none of the 4 fitting calls ever saw) and score
each model's implied probability for those same responses. This is a
standard "do item parameters generalize to new examinees" cross-validation
design, but it's different, and weaker, evidence than true cell-level
held-out-response prediction. Flagged here as a deviation driven by package
constraints, not a silent substitution.

### Real-data selection

#### Finding candidate tables: the scouting process

IRW has no field that directly flags "this table's responses are bounded
continuous" -- that has to be inferred. The scouting process that produced
the candidate list lives entirely in this repo, **not** in the `irw`
R package (`irw_metadata()` / `irw_fetch()` / `irw_tags()` are the package
functions it's built on top of; the scout scripts and their output are
project-local artifacts, not something `library(irw)` gives you). Anyone
extending this vignette to more tables, or adapting the approach for a
different response format, should start from these files directly rather
than re-deriving the logic:

- `vignettes/continuous_scout_fetch.R` -- the expensive stage.
- `vignettes/continuous_scout_classify.R` -- the cheap, rerun-freely stage.
- Cached raw output: `vignettes/continuous_scout_data/continuous_scout_raw.rds`.
- Final table: **`continuous_scout_candidates.csv`, at the repo root**
  (not inside a `vignettes/` subfolder, and not published anywhere --
  regenerate it locally with the two scripts above if you don't have it).

The filtering logic is more than a single `range(resp)` check, precisely
because raw min/max alone can't distinguish a genuine fixed-format scale
from a derived/open-ended score. It runs in three steps:

1. **Cheap metadata filter, no fetching.** Keep tables where
   `irw_metadata()$n_categories >= 12` -- the threshold this site's other
   summaries already use as the boundary of "cleanly ordinal" (see
   `dimensionality.qmd`, `local_dependence_compute.R`). This is exactly
   where many-category and continuous candidates both live, so it's a
   cheap way to shortlist before touching Redivis at all.
2. **Fetch and summarize.** For every table that survives step 1, fetch
   `resp` and compute, per table, `n_unique`/`min_resp`/`max_resp`/
   `range_resp`/`pct_noninteger`, and, per item within the table,
   `item_min`/`item_max`/`item_n`/`pct_noninteger`.
3. **Classify from two independent signals**, combined so that agreement
   is stronger evidence than either alone and disagreement gets downgraded
   to `unclear` rather than resolved automatically:
   - *resp-based*: are the table's observed min/max at a "round,"
     interpretable endpoint (0, 1, 5, 10, 100, ...), **and** do at least
     70% of items individually come within 15% of the table's overall
     range at both ends? That item-consistency check matters --  an
     earlier version compared raw per-item extremes directly and
     misclassified clean tables purely from small-sample noise (e.g. an
     11-item 0-100 VAS table where 9 items hit exactly [0,100] but 2
     items topped out at 92/98 got flagged "inconsistent"). Checking
     whether each item's range comes *within a tolerance* of the table's
     range, instead, tolerates that kind of sampling noise while still
     catching an item on a genuinely different scale (e.g. a 1-9 item
     mixed into a 0-100 table).
   - *metadata-based*: IRW's own curated `item_format` tag from
     `irw_tags()` -- `"Slider/continuous"` is direct evidence for bounded,
     `"Constructed Response"` for unbounded.

Of the 55 tables with $\geq$12 categories, this produced 24 `bounded`, 18
`unbounded`, 13 `unclear` (full breakdown and per-table notes in
`vignettes/continuous_scout_data/continuous_scout_handoff.md`).

#### Choosing among the 24 bounded candidates

The scout's own top pick -- `lsbq_maleki_2025_non_persian_proficiency` --
is what we used for this pilot: 3 items, 312 respondents, a 0-10 scale with
**~40% genuinely non-integer response values**. That fractional-value
evidence mattered more than sample size or item count in the final choice:
every 0-100 "Slider/continuous"-tagged table we checked
(`emoji_scheffler_2024`, the `ai_fear_dong_2026_*` family, etc.) came back
**100% integer-valued** on inspection, which is weaker evidence of genuine
continuity than a fine-grained integer scale -- classification from
min/max and tags alone can't tell "true VAS" from "a very fine 0-100
integer rating," so this had to be checked by hand.

Two other scout candidates were checked and rejected before landing on this
one, both for reasons specific to what these 4 models assume:

- `eammi_grahe_2018_marriage_identity_allocation` -- row sums across items
  are a constant 100. This is an ipsative/compositional point-allocation
  task, which violates the conditional-independence assumption every one of
  these 4 models makes (a higher score on one item mechanically forces
  lower scores on the others).
- `much_tte_2025_currentmotivation` / `test_taking_much_2025_cm` -- person
  x item cells are not unique (repeated measurement occasions per item, not
  a single-occasion multi-item battery).

### Simulation

**A single illustrative condition** ($N=300$, $J=10$, Beta IRT DGP, fixed
dispersion) is shown first below, generated directly from the Beta IRT
model's own equations (`sim_beta_irt()` in `continuous_bounded_helpers.R`
-- a bug-fixed reimplementation of `sirt::brm.sim()`'s logic; see
[Limitations](#limitations) for why the reimplementation was necessary).

**The full recovery/misspecification grid** cross-fits all 4 core specs
against 3 DGPs, generated from each model's own equations: Beta IRT
(`sim_beta_irt()`), Samejima CRM (`sim_samejima_crm()`: $Z_{ij} =
\alpha_j(\theta_i - b_j) + e_{ij}$, $e_{ij}\sim N(0,\sigma_j^2)$, then
$X_{ij}=1/(1+e^{-Z_{ij}})$ -- the exact inverse of EstCRM's own $Z=\ln(X/(1-X))$
transform), and Mueller CRSM (`sim_muller_corsm()`: grid-based inverse-CDF
sampling from the same $\exp[x(\theta-b)+x(1-x)\lambda]$ density used for
scoring, since no closed-form sampler exists for this family). The
naive-linear DGP is skipped, per the original scope's own allowance: an
unbounded-Gaussian DGP clipped to $[0,1]$ to make it "bounded" would
introduce an artificial boundary spike unrepresentative of what a linear
model actually generates -- the exact awkwardness the scope flagged as a
reason to skip it.

Grid: $N\in\{300,1000\}$, $J\in\{8,20\}$, and, for the Beta IRT DGP
specifically, fixed vs. item-varying dispersion -- 8 + 4 + 4 = 16
conditions, plus 1 boundary-inflation condition (10% of responses forced to
exactly 0, 10% forced to exactly 1, layered on the Beta IRT DGP at
$N=300$/$J=8$) for 17 total. Run via `furrr::future_pmap()` +
`plan(multisession)`, cached separately from the real-data pilot above
(own skip-if-exists gate in the compute script).

## Results

### Simulated data: does the correctly-specified model win, and does it recover parameters?

```{r}
#| label: fig-sim-scores
if (cache_ok) {
  res$sim$scores |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = sprintf("%.2f", mean_ll)), vjust = 1.4, color = "white", size = 3.5) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = "Simulated Beta-IRT data: held-out fit by model") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

On data simulated directly from the Beta IRT model, the two Beta IRT
variants have the best (least negative) held-out fit, essentially tied with
Samejima CRM and Mueller CRSM, and all three clearly beat the naive linear
baseline. This is the expected pattern for a correctly-specified-model
recovery check: the models built for bounded continuous responses recover
the truth better than the one that ignores the bounds.

Item-difficulty recovery for the item-varying-dispersion Beta IRT fit
(correlation between estimated and true $\delta$) was
`r if (cache_ok) sprintf("%.3f", res$sim$recovery_cor_delta) else "NA"`.

### The full recovery grid: 16 conditions x 3 DGPs

Parameter recovery for the correctly-specified model was excellent
everywhere -- every one of the 16 non-inflation conditions recovered its
true item parameters at $r>0.96$, most above $r>0.99$:

```{r}
#| label: tbl-grid-recovery
if (grid_cache_ok) {
  grid_res$scores |>
    filter(!boundary_inflation) |>
    distinct(dgp, N, J, dispersion_mode, recovery_cor) |>
    arrange(dgp, N, J, dispersion_mode) |>
    mutate(recovery_cor = sprintf("%.3f", recovery_cor)) |>
    knitr::kable(col.names = c("DGP", "N", "J", "Beta IRT dispersion", "Recovery r"))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

More useful than "does recovery work" (it does, everywhere) is the
cross-fitting question: when data comes from one DGP, which of the 4
*fitted* specs wins the held-out comparison?

```{r}
#| label: fig-grid-crossfit
if (grid_cache_ok) {
  grid_res$scores |>
    filter(!boundary_inflation) |>
    mutate(model = factor(model, levels = names(model_colors)),
           dgp_label = recode(dgp, beta_irt = "DGP: Beta IRT", samejima_crm = "DGP: Samejima CRM",
                               muller_corsm = "DGP: Mueller CRSM")) |>
    group_by(dgp_label, model) |>
    summarise(mean_ll = mean(mean_ll, na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    facet_wrap(~dgp_label, ncol = 3) +
    labs(x = NULL, y = "Mean held-out log-probability\n(averaged across N/J conditions)",
         title = "Cross-fitting: held-out fit by model, averaged across all N/J conditions per DGP") +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

The pattern is consistent across all 3 DGPs and matches every real-data
result above: **the 3 bounded-response models cluster tightly together and
substantially outperform the naive linear baseline, regardless of which of
the 3 actually generated the data.** The correctly-specified model is
usually numerically best or tied for best (Samejima CRM wins clearly under
its own DGP; Mueller CRSM wins clearly under its own DGP), but the margin
between the 3 bounded-response models is small compared to the gap to naive
linear -- these 3 models are more interchangeable with each other than any
of them is with a model that ignores the bounds. One fit failed outright:
Samejima CRM's EM did not converge (a `missing value where TRUE/FALSE
needed` error partway through) on one Mueller-CRSM-generated dataset
($N=300$, $J=8$) -- a real misspecification-robustness finding, not
papered over; see [Limitations](#limitations).

### Boundary inflation: does floor/ceiling piling break anything?

Beyond fitting, we asked how each spec's held-out fit *degrades* when 20%
of responses are forced to the exact boundary (10% to 0, 10% to 1) on top
of the Beta IRT DGP -- diagnostic information, not a 5th model to fit:

```{r}
#| label: fig-grid-inflation
if (grid_cache_ok) {
  grid_res$scores |>
    filter(dgp == "beta_irt", N == 300, J == 8, dispersion_mode == "fixed") |>
    mutate(model = factor(model, levels = names(model_colors)),
           condition = ifelse(boundary_inflation, "20% boundary-inflated", "Not inflated")) |>
    ggplot(aes(x = model, y = mean_ll, fill = model, alpha = condition)) +
    geom_col(position = "dodge", width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_alpha_manual(values = c("Not inflated" = 1, "20% boundary-inflated" = 0.55), name = NULL) +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability",
         title = "Beta IRT DGP (N=300, J=8): held-out fit with vs. without boundary inflation") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

Every spec's held-out fit gets worse under inflation -- none of the 4 core
specs explicitly models a zero-one-inflated mixture, so this is expected --
but **not by the same amount**. The naive linear baseline degrades by
about 23% (mean log-probability drops from -1.57 to -1.93); the 3
bounded-response models degrade by roughly 15-16% each (e.g. Mueller CRSM:
-1.25 to -1.43). The bounded-response models' smooth density still puts
more mass near the boundary than a Gaussian does, so they absorb some of
the extra piling without a dedicated inflation term, while the naive linear
model has nothing in its structure to absorb it at all. None of the 4
handle it *well* -- a true zero-one-inflated extension (Molenaar et al.,
2022) would be needed for that -- but the bounded-response models are
comparatively more robust to it than the baseline that ignores boundedness
altogether.

### Real data: `r if (cache_ok) res$real$table else ""`

What does this table's response distribution actually look like? This is
the ~40% non-integer evidence from [Real-data selection](#real-data-selection)
made visual, and it's also the direct reason the naive linear baseline
underperforms below: a Gaussian model expects mass concentrated symmetrically
around the mean, not piled up against a hard boundary.

```{r}
#| label: fig-real-dist
if (cache_ok) {
  as.data.frame(res$real$Y_raw) |>
    tidyr::pivot_longer(everything(), names_to = "item", values_to = "resp") |>
    ggplot(aes(x = resp)) +
    geom_histogram(binwidth = 0.5, boundary = 0, fill = irw_blue, color = "white", linewidth = 0.1) +
    facet_wrap(~item, ncol = 3) +
    labs(x = sprintf("Raw response (%g-%g scale)", res$real$low, res$real$high),
         y = "Count",
         title = sprintf("%s: raw response distribution by item", res$real$table))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

All 3 items show a similar shape: a broad hump across the mid-range plus a
sharp spike at the upper boundary. Between 23% and 44% of responses per
item sit at *exactly* 10, the scale ceiling (vs. under 9% at exactly 0) --
respondents who are, or report being, fully proficient pile up exactly at
the boundary, which a symmetric Gaussian around a single mean cannot
represent. This ceiling pile-up is precisely the floor/ceiling behavior
the 3 bounded-response models are built to handle and the naive linear
model is not.

```{r}
#| label: fig-real-scores
if (cache_ok) {
  res$real$scores |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = sprintf("%.2f", mean_ll)), vjust = 1.4, color = "white", size = 3.5) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = sprintf("%s: held-out fit by model", res$real$table)) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

On the real data
(N=`r if (cache_ok) res$real$N else "?"`,
J=`r if (cache_ok) res$real$J else "?"`), the naive linear baseline is
substantially worse than all 3 bounded-response models -- unsurprising,
given this is exactly a 0-10 scale with real floor/ceiling behavior.
Samejima CRM comes out slightly ahead of the other 3 in this single pilot
run; with only 3 items and one train/test split, this ordering shouldn't be
over-read (see [Limitations](#limitations)).

### Vetting the other 20 bounded candidates (two rounds -- the first round had a real bug)

The scout's `bounded` classification covers 24 tables total. Before adding
more real-data examples, we screened the other 20 (everything besides the
one already used, plus 1 rejected during the initial selection) for
structural problems like the one that ruled out
`eammi_grahe_2018_marriage_identity_allocation` above.

**Round 1** (`vet_candidates.R`) flagged 7 tables as unusable because
`(id, item)` pairs repeated many times per table -- more rows than
`n_persons x n_items` would predict. That conclusion was wrong for most of
them. IRW's own data standard (`standard.qmd`) documents `wave` (repeated/
longitudinal administrations) and `rater` (multiple raters per unit) as
standard optional columns for exactly this scenario, and every one of the
7 flagged tables turned out to have one:

| Table | Column that explains the "duplication" |
|---|---|
| `tears` | `rater`, `stimulus`, `phase` |
| `nas_rogoza_2024_study5_{nas,ngs,nvs}` | `wave` |
| `emoji_scheffler_2024` | `rater` |
| `opentsstvr_linnig_2025_vas` | `wave` |
| `westhoff2023_{stopd,pbat}` | `wave` |
| `gilbert_meta_95` | `wave` |
| `much_tte_2025_currentmotivation` | `wave` |

Round 1's check only ever grouped by `(id, item)` and never looked for
these columns -- so it was discarding information IRW had correctly
stored, not finding a flaw in the data. Separately, `test_taking_much_2025_cm`
had been excluded on an untested assumption (grouped with `much_tte` by
name similarity, never actually checked) -- it turns out to have zero
duplicate rows and needs no filtering at all.

**Round 2** (`vet_candidates2.R`) filters each of these 8 tables to a
single wave/rater occasion (or, for `test_taking_much_2025_cm`, uses it
as-is) and re-checks. **5 of the 8 came out clean** -- unique `(id, item)`
pairs, a real N, and no ipsative row-sum constraint -- and are used below.
The other 3 (`tears`, `emoji_scheffler_2024`, `opentsstvr_linnig_2025_vas`,
plus `double_marking_steele_2022` and `thomeczek2025_les`, which weren't
wave/rater cases to begin with) remain excluded even after this
correction: `tears` is a fully crossed rater x stimulus x phase design that
doesn't reduce to a single filter; `emoji_scheffler_2024`'s `rater` column
(153 distinct values against 107 `id`s) doesn't divide cleanly into one
occasion per person; `opentsstvr_linnig_2025_vas` still has duplicate
`(id, item)` pairs even within one wave; and `double_marking_steele_2022` /
`thomeczek2025_les` have too few observations per rater once split (8 and
6 dissertations/sessions respectively).

The 3 tables from the **original** round-1 screen that were correctly
excluded regardless of the wave/rater oversight: `klippel_irw`
(item-varying ranges), `ehealth_rioux_2025_{triplep,beam,abiliti}` (mixed
0-100/1-7 formats within one table), and
`mclaughlin_samuel_2025_auditory_session_1` (1 item -- unusable for IRT
regardless of anything else). Full vetting code and output:
`vignettes/continuous_bounded_data/vet_candidates.R` /`vet_results.csv`
(round 1) and `vet_candidates2.R` / `vet_results2.csv` (round 2).

### Real data #2: a large-*N*, integer-valued check (`ai_fear_dong_2026_ai`)

Of the tables that were structurally clean from the start,
`ai_fear_dong_2026_*` (4 tables, N=10,000, exactly `n_persons x n_items`
rows, 0 duplicates) stood out but needed one more decision: each is a
48-item, 6-target x 8-trait factorial design (Care/Doctor/Journalist/
Judge/Manager/Religious x competent/warm/etc.), not one construct. We used
one target's 8 trait items -- `Doctor_*` (competent, determined, fair,
imaginative, intelligent, sincere, tolerant, warm) -- treated as a single
"overall impression of Doctor" construct. This is a defensible
simplification (trait ratings like these reliably load on one general
"halo"/evaluation factor even when they nominally span warmth and
competence facets) but is a real construct-validity judgment call, not
something the scouting script could resolve automatically.

Unlike `lsbq_maleki_2025_non_persian_proficiency`, this table is **100%
integer-valued** -- weaker evidence of genuine continuity (see
[Real-data selection](#real-data-selection)) -- so it's included as a
large-*N* structural check, not a second demonstration of fractional-value
continuity. We subsampled to N=2,000 (from 10,000) for compute
tractability in this pilot.

```{r}
#| label: fig-real2-dist
if (cache_ok) {
  as.data.frame(res$real2$Y_raw) |>
    tidyr::pivot_longer(everything(), names_to = "item", values_to = "resp") |>
    mutate(item = sub("^Doctor_", "", item)) |>
    ggplot(aes(x = resp)) +
    geom_histogram(binwidth = 5, boundary = 0, fill = irw_blue, color = "white", linewidth = 0.1) +
    facet_wrap(~item, ncol = 4) +
    labs(x = sprintf("Raw response (%g-%g scale)", res$real2$low, res$real2$high),
         y = "Count",
         title = sprintf("%s (%s facet): raw response distribution by item", res$real2$table, res$real2$facet))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

Ceiling piling shows up here too (1.7%-27% of responses land at exactly
100, item-dependent), though less uniformly across items than in the
primary table.

```{r}
#| label: fig-real2-scores
if (cache_ok) {
  res$real2$scores |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = sprintf("%.2f", mean_ll)), vjust = 1.4, color = "white", size = 3.5) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = sprintf("%s (%s facet): held-out fit by model", res$real2$table, res$real2$facet)) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

Same qualitative pattern as the primary table: naive linear is clearly
worst, the 3 bounded-response models are close together, with Mueller CRSM
slightly ahead this time (vs. Samejima CRM in the primary table) -- with
N=2,000 this ordering is more stable than the 3-item pilot above, but it's
still one train/test split on one facet of one table, not a claim that
Mueller CRSM generally wins.

### 5 more real tables (round-2 vetting survivors)

`much_tte_2025_currentmotivation` (3 items, N=1,244), `test_taking_much_2025_cm`
(9 items, N=1,244), and the 3 `nas_rogoza_2024_study5_*` subscales (NAS/NGS/NVS,
4 items each, N~270-280, from the same daily-diary narcissism study --
Narcissistic Antagonism, Grandiosity, and Vulnerability) all passed round-2
vetting and fit cleanly:

```{r}
#| label: fig-batch-scores
if (cache_ok) {
  res$batch$scores |>
    mutate(model = factor(model, levels = names(model_colors)),
           label = sprintf("%s\n(N=%d, J=%d)", label, N, J)) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    facet_wrap(~label, ncol = 3) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = "Held-out fit by model, across 5 additional real tables") +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

The naive linear baseline is worst everywhere, consistent with the 2 tables
above. Among the 3 bounded-response models, the ranking isn't stable across
tables -- Mueller CRSM edges out the others on `test_taking_much_2025_cm`
and 2 of the 3 `nas_rogoza` subscales, but the gaps are small throughout and
this is still one train/test split per table. The main takeaway is
qualitative, not "which of the 3 wins": all 3 bounded-response models
consistently and substantially beat the naive linear baseline across every
real table tried so far.

`westhoff2023_{stopd,pbat}` and `gilbert_meta_95` passed the same
structural (wave) fix but hit an unresolved numerical issue in the Beta IRT
scoring step (and, for `pbat`, Samejima CRM fitting too) even after
coarsening the shared bin grid from 12 to 6 bins -- left out of this pass
rather than force partial results into the comparison (see
[Limitations](#limitations)).

## Limitations {#limitations}

**This is still a pilot, not the final study,** though the simulation side
is now complete per the original scope. Concretely not yet done:

- **Ferrando's (2001) nonlinear congeneric model** [@ferrando2001nonlinear]
  -- the optional stretch spec, comparable to the treatment in Ferrando
  (2002) [@ferrando2002theoretical] -- has not been attempted at all. Specs
  1-4 are now complete and validated (the grid ran cleanly), so its own
  sequencing rule no longer blocks it -- it just hasn't been picked up yet.
- **Partial missingness in real data.** All 7 real tables used complete
  cases only; IRW tables with partially missing item responses aren't
  handled yet.
- **Only one train/test split per condition** (real or simulated), not a
  repeated/cross-validated design -- so none of the held-out numbers above
  have an uncertainty band, and a single unlucky split could shift a close
  call between the 3 bounded-response models.
- **One grid fit failed outright**: Samejima CRM's EM did not converge
  (`missing value where TRUE/FALSE needed`) on one Mueller-CRSM-generated
  condition ($N=300$, $J=8$, no boundary inflation). Not re-run with a
  different seed or diagnosed further -- reported as a real
  misspecification-robustness finding (EstCRM's EM can fail outright, not
  just fit poorly, on data far from its assumptions) rather than treated as
  a bug to route around.
- **The grid's boundary-inflation condition covers only 1 of the 3 DGPs**
  (Beta IRT, $N=300$/$J=8$) -- not repeated for Samejima CRM or Mueller
  CoRSM, or at other N/J combinations.

**Package-availability and correctness issues found while building this**
(worth knowing before extending this pilot):

- `pcIRT::CRSM()` (Mueller's model) was **dropped from the package
  entirely** in the last CRAN release (0.2.4, 2019); it only exists in
  pcIRT $\leq$ 0.2.3. `pcIRT` itself was later archived by CRAN entirely
  (2025-11-11). This vignette's compute script installs 0.2.3 from the
  CRAN archive, not the current CRAN release -- a real risk for
  reproducing this vignette in the future if the archive changes.
- That same version's `CRSM()` has a convergence-check bug
  (`is.na(para) || ...` where `para` becomes a 2-vector after the first
  iteration) that is a hard error under R $\geq$ 4.3 and had to be patched
  locally before it would run at all.
- `sirt::brm.sim()` hardcodes a 9-item assumption in its column-naming
  code and errors for any other item count; `sim_beta_irt()` in the
  helpers file reimplements its generative logic directly rather than
  patching `sirt`.
- `pcIRT::person_par.CRSM()` only scores the same persons used in the
  original fit (no missing-data handling, no new-data argument) and could
  not be reused for the person-level holdout design above; a generalized
  version of its estimating equation was written from scratch instead.
- The from-scratch Mueller CRSM person-scoring routine
  (`theta_new_muller_corsm()`) could diverge to a `theta` extreme enough to
  overflow `exp()` inside `integrate()`, which throws a hard "non-finite
  function value" error rather than the graceful NA `stop.on.error=FALSE`
  is supposed to allow -- this crashed the *entire* held-out scoring pass
  for a table (not just Mueller CRSM) before being caught and fixed by
  clamping `theta` and wrapping each `integrate()` call individually.
- Similarly, a Beta IRT scoring-stage failure (see below) was originally
  written so it could crash naive-linear/Samejima/Mueller's results for the
  same table too -- every model's fit-and-score block is now wrapped in its
  own `tryCatch` so one model's failure can't take down the other 3.
- **Unresolved**: `westhoff2023_{stopd,pbat}` and `gilbert_meta_95` pass
  every structural check (single wave, unique `(id, item)`, no ipsative
  constraint) but still fail at the Beta IRT scoring step -- `mirt`
  compacts a training split's observed categories to consecutive integers
  per item, and if a held-out test person's response lands in a category
  with zero training-split observations, `fscores(response.pattern=...)`
  rejects it. Lowering the shared bin count from 12 to 6 fixed this for
  every other table in this pilot but not these 3 (`pbat` also fails
  Samejima CRM fitting with an unrelated "missing value where TRUE/FALSE
  needed" error, not yet diagnosed). Left out of the real-data set rather
  than chasing further.

**Explicitly out of scope** (not implemented, not planned for this
vignette): DIF for continuous responses [@finch2023impact]; dual-range
slider / interval-response models and the Interval Consensus Model
[@kloft2024measuring; @kloft2023dirichlet; @kloft2025interval] -- a
different response format (an interval, not a point) and, for the
Consensus Model, a different modeling paradigm (cultural consensus
theory) entirely; Zopluoglu's mixture CRM [@zopluoglu2020finite] and
Noel's beta unfolding model [@noel2014beta] -- each adds a full second axis
of complexity (latent mixture classes; unfolding vs. dominance response
process), and unfolding in particular is a more natural extension of the
site's existing unfolding-vs-dominance vignette than a duplicate here;
Ferrando and Navarro-Gonzalez's (2021) multidimensional Dual Thurstonian
model [@ferrando2021multidimensional].

**The comparison metric (held-out log-likelihood) is an ad hoc choice**,
not a validated best-practice for this setting -- flagged explicitly above,
worth revisiting before this becomes the final vignette.

## Reproducibility

```{r}
#| echo: false
run_note <- if (cache_ok) res$scope_note else "not yet run"
```

`r run_note`.

```r
# 1. Run the compute script (from the project root)
source("vignettes/continuous_bounded_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/continuous_bounded.qmd")
```

::: {.callout-tip}
To cite the real-data table used here, run:

```r
irw_save_bibtex("lsbq_maleki_2025_non_persian_proficiency",
                 output_file = "continuous_bounded_data/irw_references.bib")
```
:::

---

## References
