---
title: "Bounded Continuous Responses: Comparing Beta IRT, Samejima's CRM, and Mueller's CRSM"
description: |
  Slider and visual-analogue-scale items produce a response that is
  continuous but lives in a fixed, interpretable range, like [0,1] or
  [0,100] -- not the open-ended continuous case standard CFA/linear-factor
  models assume. We compare four published ways of modeling that: a naive
  linear baseline that ignores the bounds, Noel and Dauvier's Beta IRT
  model, Samejima's Continuous Response Model, and Mueller's Continuous
  Rating Scale Model, on one simulated dataset and one real IRW table.
date: today
bibliography: continuous_bounded_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-important}
**This is a draft pilot, not the final vignette.** It exists to get feedback
on the modeling approach before investing in the full study. What's here:
real fits of all 4 core model specs, on one simulated dataset and one real
IRW table, with a working held-out comparison. What's *not* here yet: the
full simulation/recovery grid (multiple DGPs x sample sizes x item counts),
a boundary-inflation (zero/one-piling) condition, and Ferrando's (2001)
optional stretch model. See [Limitations](#limitations) for the complete
list and the reasoning behind several scope decisions made along the way.
:::

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 13))
irw_blue   <- "#2166ac"
irw_red    <- "#d73027"
irw_green  <- "#4daf4a"
irw_purple <- "#984ea3"
irw_grey   <- "#808080"

model_colors <- c(
  "naive_linear"              = irw_grey,
  "beta_irt_item_dispersion"  = irw_blue,
  "beta_irt_fixed_dispersion" = irw_purple,
  "samejima_crm"              = irw_red,
  "muller_corsm"              = irw_green
)
model_labels <- c(
  "naive_linear"              = "Naive linear",
  "beta_irt_item_dispersion"  = "Beta IRT (item disp.)",
  "beta_irt_fixed_dispersion" = "Beta IRT (fixed disp.)",
  "samejima_crm"              = "Samejima CRM",
  "muller_corsm"               = "Mueller CRSM"
)

cache_path <- "continuous_bounded_data/continuous_bounded_results.rds"
cache_ok <- file.exists(cache_path)
if (cache_ok) res <- readRDS(cache_path)
```

## Motivation

Visual analogue scales, slider ratings, and continuous confidence or
percentage judgments are common in IRW: a respondent marks a point on a
fixed, labeled segment (0 to 100, say), and the raw response is a
continuous number that is *bounded* at both ends. This is a different
measurement problem from the unbounded continuous case (response times,
open-ended magnitude estimates), which reduces to standard linear-factor
or CFA modeling and is covered elsewhere on this site -- the bounds
themselves are the interesting part here, because they create floor/ceiling
effects and non-constant response variance near the edges that a plain
linear model doesn't see.

Three published models handle this directly. Noel and Dauvier's Beta IRT
model [@noel2007beta] treats the response as Beta-distributed given
ability, with the two natural Beta parameters reparameterized into an
item difficulty and a dispersion. Samejima's Continuous Response Model
[@samejima1973homogeneous] models a monotonic transform of the response as
normal given ability -- the historically original approach, predating Beta
IRT by over 30 years. Mueller's Continuous Rating Scale Model
[@mueller1987rasch] takes a Rasch-family, specific-objectivity approach,
estimated by pairwise conditional likelihood; Verhelst's (2019) exponential
family framework [@verhelst2019exponential] is the modern theoretical
generalization of Mueller's model, though the model actually fitted here is
Mueller's original. We compare these three against a naive linear (Gaussian,
identity-link) baseline that ignores the bounds entirely -- the "what if you
just treat it like ordinary continuous data" control, playing the same role
the Rasch baseline plays in the site's guessing-models vignette.

## Data and methods

### The four core specs

1. **Naive linear baseline** -- single-factor CFA via `lavaan`, identity
   link, Gaussian residual, fit directly on the raw (rescaled-to-[0,1])
   response. Ignores boundedness entirely.
2. **Beta IRT** [@noel2007beta] -- fit as a discretized custom `mirt`
   itemtype wrapping `sirt::brm.irf()`, following `sirt`'s own documented
   mirt-estimation recipe (`?sirt::brm.sim`). Two variants: **item-varying
   dispersion** (each item gets its own dispersion parameter, the model's
   default parameterization) and **fixed dispersion** (one dispersion
   parameter shared across all items, via an `mirt` parameter constraint).
3. **Samejima's Continuous Response Model** [@samejima1973homogeneous] --
   `EstCRM::EstCRMitem()` / `EstCRMperson()`.
4. **Mueller's Continuous Rating Scale Model** [@mueller1987rasch] -- a
   patched copy of `pcIRT::CRSM()` / a from-scratch reimplementation of its
   person-scoring equation (see [Limitations](#limitations) for why
   patching was necessary).

Ferrando's (2001) nonlinear congeneric model [@ferrando2001nonlinear] is an
optional stretch spec with no package implementation (it would need a
custom truncated-normal likelihood via `optim`); per the original scope for
this vignette it's only attempted after specs 1-4 are complete and
validated, and **has not been attempted in this draft** -- see
[Limitations](#limitations).

### Why "interval probability," not raw log-likelihood

Before any model comparison, we checked whether all 4 models report
likelihood on the same scale. They don't: reading `EstCRM::EstCRMitem()`'s
source directly shows it transforms the raw response to
$Z = \ln\!\big(X/(K-X)\big)$ and fits/reports its EM log-likelihood entirely
on that $Z$-scale, with no Jacobian term anywhere in the function. A raw
log-likelihood from EstCRM is therefore not comparable to, say, the Beta
IRT model's density on the raw response, without a correction.

Rather than hand-deriving a per-model Jacobian correction, every model's
held-out score in this vignette is a **binned interval log-probability**:
$\log P(\text{bin}_{lo} < X < \text{bin}_{hi})$, using the same bin edges
(on the raw, rescaled-to-$[0,1]$ scale) for all 4 models. Interval
probability is invariant to a monotonic reparameterization of the
intervening likelihood -- the Jacobian exactly cancels under integration --
so this sidesteps the Jacobian bookkeeping entirely. This is a deliberate
substitution for "compare raw densities with a Jacobian correction," flagged
here because the original scope for this vignette didn't spell out this
exact resolution.

### Comparison metric: held-out log-likelihood, not IMV

We use held-out log-likelihood rather than IMV (InterModel Vigorish), the
metric used in most other vignettes on this site. `imv4sem` and
`imv.mirt` are not on CRAN, and no citable generalization of IMV to
continuous or Beta-type outcomes was found. **This is flagged as an ad hoc
decision to revisit, not a settled one** -- unlike IMV's binary/count case,
there's no obviously "natural" coin-flip baseline for a continuous
response, so held-out log-likelihood is a reasonable default rather than a
demonstrably right answer.

### Held-out design: person-level, not cell-level

The project's usual convention for held-out evaluation on IRW data is a
response-level (missing-cell) holdout: mask a random subset of person x item
cells, fit on the rest, score the masked cells. That isn't viable here for
2 of the 4 specs -- reading `EstCRM::EstCRMitem()` and `pcIRT::CRSM()`'s
fitting code directly shows **neither has any missing-data handling at
all** (both do matrix/pairwise algebra that silently breaks with `NA`
cells, confirmed by testing).

We substituted a **person-level train/test split** instead: fit item
parameters on a training subset of persons (a complete matrix, no missing
cells), then estimate ability for held-out test persons (from their own
full response vector, which none of the 4 fitting calls ever saw) and score
each model's implied probability for those same responses. This is a
standard "do item parameters generalize to new examinees" cross-validation
design, but it's different, and weaker, evidence than true cell-level
held-out-response prediction. Flagged here as a deviation driven by package
constraints, not a silent substitution.

### Real-data selection

We used the scouting pass over IRW documented in
`continuous_scout_candidates.csv` / `continuous_scout_data/continuous_scout_handoff.md`
(55 tables with $\geq$12 response categories, classified bounded /
unbounded / unclear from observed min/max patterns and IRW's `item_format`
tag). Of the 24 tables classified `bounded`, the scout's own top pick --
`lsbq_maleki_2025_non_persian_proficiency` -- is what we used: 3 items, 312
respondents, a 0-10 scale with **~40% genuinely non-integer response
values**. That fractional-value evidence mattered more than sample size or
item count in the final choice: every 0-100 "Slider/continuous"-tagged
table we checked (`emoji_scheffler_2024`, the `ai_fear_dong_2026_*` family,
etc.) came back **100% integer-valued** on inspection, which is weaker
evidence of genuine continuity than a fine-grained integer scale.

Two other scout candidates were checked and rejected before landing on this
one, both for reasons specific to what these 4 models assume:

- `eammi_grahe_2018_marriage_identity_allocation` -- row sums across items
  are a constant 100. This is an ipsative/compositional point-allocation
  task, which violates the conditional-independence assumption every one of
  these 4 models makes (a higher score on one item mechanically forces
  lower scores on the others).
- `much_tte_2025_currentmotivation` / `test_taking_much_2025_cm` -- person
  x item cells are not unique (repeated measurement occasions per item, not
  a single-occasion multi-item battery).

### Simulation

A single condition, not yet the full grid (see
[Limitations](#limitations)): $N=300$ persons, $J=10$ items, generated
directly from the Beta IRT model's own equations (`sim_beta_irt()` in
`continuous_bounded_helpers.R` -- a bug-fixed reimplementation of
`sirt::brm.sim()`'s logic; see [Limitations](#limitations) for why the
reimplementation was necessary), fixed dispersion $\tau=0.6$ across items,
item difficulties spaced evenly on $[-1.5, 1.5]$.

## Results

### Simulated data: does the correctly-specified model win, and does it recover parameters?

```{r}
#| label: fig-sim-scores
if (cache_ok) {
  res$sim$scores |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = sprintf("%.2f", mean_ll)), vjust = 1.4, color = "white", size = 3.5) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = "Simulated Beta-IRT data: held-out fit by model") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

On data simulated directly from the Beta IRT model, the two Beta IRT
variants have the best (least negative) held-out fit, essentially tied with
Samejima CRM and Mueller CRSM, and all three clearly beat the naive linear
baseline. This is the expected pattern for a correctly-specified-model
recovery check: the models built for bounded continuous responses recover
the truth better than the one that ignores the bounds.

Item-difficulty recovery for the item-varying-dispersion Beta IRT fit
(correlation between estimated and true $\delta$) was
`r if (cache_ok) sprintf("%.3f", res$sim$recovery_cor_delta) else "NA"`.

### Real data: `r if (cache_ok) res$real$table else ""`

```{r}
#| label: fig-real-scores
if (cache_ok) {
  res$real$scores |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = sprintf("%.2f", mean_ll)), vjust = 1.4, color = "white", size = 3.5) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = sprintf("%s: held-out fit by model", res$real$table)) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

On the real data
(N=`r if (cache_ok) res$real$N else "?"`,
J=`r if (cache_ok) res$real$J else "?"`), the naive linear baseline is
substantially worse than all 3 bounded-response models -- unsurprising,
given this is exactly a 0-10 scale with real floor/ceiling behavior.
Samejima CRM comes out slightly ahead of the other 3 in this single pilot
run; with only 3 items and one train/test split, this ordering shouldn't be
over-read (see [Limitations](#limitations)).

## Limitations {#limitations}

**This is a single-shot pilot, not the full study.** Concretely not yet
done:

- **The full recovery/misspecification grid.** The original scope called
  for cross-fitting all 4 specs against 3-4 DGPs (Beta IRT, Samejima CRM,
  Mueller CoRSM), varying N (300/1000), J (8/20), and Beta IRT dispersion
  structure (fixed/item-varying), plus a boundary-inflation (zero/one
  piling) condition to see how each spec degrades qualitatively. None of
  that is here yet -- this draft has one simulated dataset under one DGP.
- **Ferrando's (2001) nonlinear congeneric model** [@ferrando2001nonlinear]
  -- the optional stretch spec, comparable to the treatment in Ferrando
  (2002) [@ferrando2002theoretical] -- has not been attempted at all,
  correctly per its own sequencing rule (only after specs 1-4 are complete
  and validated).
- **Partial missingness in real data.** The real-data pilot used complete
  cases only; IRW tables with partially missing item responses aren't
  handled yet.
- **Only one train/test split**, not a repeated/cross-validated design.

**Package-availability and correctness issues found while building this**
(worth knowing before extending this pilot):

- `pcIRT::CRSM()` (Mueller's model) was **dropped from the package
  entirely** in the last CRAN release (0.2.4, 2019); it only exists in
  pcIRT $\leq$ 0.2.3. `pcIRT` itself was later archived by CRAN entirely
  (2025-11-11). This vignette's compute script installs 0.2.3 from the
  CRAN archive, not the current CRAN release -- a real risk for
  reproducing this vignette in the future if the archive changes.
- That same version's `CRSM()` has a convergence-check bug
  (`is.na(para) || ...` where `para` becomes a 2-vector after the first
  iteration) that is a hard error under R $\geq$ 4.3 and had to be patched
  locally before it would run at all.
- `sirt::brm.sim()` hardcodes a 9-item assumption in its column-naming
  code and errors for any other item count; `sim_beta_irt()` in the
  helpers file reimplements its generative logic directly rather than
  patching `sirt`.
- `pcIRT::person_par.CRSM()` only scores the same persons used in the
  original fit (no missing-data handling, no new-data argument) and could
  not be reused for the person-level holdout design above; a generalized
  version of its estimating equation was written from scratch instead.

**Explicitly out of scope** (not implemented, not planned for this
vignette): DIF for continuous responses [@finch2023impact]; dual-range
slider / interval-response models and the Interval Consensus Model
[@kloft2024measuring; @kloft2023dirichlet; @kloft2025interval] -- a
different response format (an interval, not a point) and, for the
Consensus Model, a different modeling paradigm (cultural consensus
theory) entirely; Zopluoglu's mixture CRM [@zopluoglu2020finite] and
Noel's beta unfolding model [@noel2014beta] -- each adds a full second axis
of complexity (latent mixture classes; unfolding vs. dominance response
process), and unfolding in particular is a more natural extension of the
site's existing unfolding-vs-dominance vignette than a duplicate here;
Ferrando and Navarro-Gonzalez's (2021) multidimensional Dual Thurstonian
model [@ferrando2021multidimensional].

**The comparison metric (held-out log-likelihood) is an ad hoc choice**,
not a validated best-practice for this setting -- flagged explicitly above,
worth revisiting before this becomes the final vignette.

## Reproducibility

```{r}
#| echo: false
run_note <- if (cache_ok) res$scope_note else "not yet run"
```

`r run_note`.

```r
# 1. Run the compute script (from the project root)
source("vignettes/continuous_bounded_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/continuous_bounded.qmd")
```

::: {.callout-tip}
To cite the real-data table used here, run:

```r
irw_save_bibtex("lsbq_maleki_2025_non_persian_proficiency",
                 output_file = "continuous_bounded_data/irw_references.bib")
```
:::

---

## References
