---
title: "Bounded Continuous Responses: Comparing Beta IRT, Samejima's CRM, and Mueller's CRSM"
description: |
  Slider and visual-analogue-scale items produce a response that is
  continuous but lives in a fixed, interpretable range, like [0,1] or
  [0,100] -- not the open-ended continuous case standard CFA/linear-factor
  models assume. We compare four published ways of modeling that: a naive
  linear baseline that ignores the bounds, Noel and Dauvier's Beta IRT
  model, Samejima's Continuous Response Model, and Mueller's Continuous
  Rating Scale Model, on one simulated dataset and one real IRW table.
date: today
bibliography:
  - continuous_bounded_data/references.bib
  - continuous_bounded_data/irw_references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`continuous_bounded_compute.R`) directly if
you're relying on the numbers here.
:::

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 13))
irw_blue   <- "#2166ac"
irw_red    <- "#d73027"
irw_green  <- "#4daf4a"
irw_purple <- "#984ea3"
irw_grey   <- "#808080"

model_colors <- c(
  "naive_linear"              = irw_grey,
  "beta_irt_item_dispersion"  = irw_blue,
  "beta_irt_fixed_dispersion" = irw_purple,
  "samejima_crm"              = irw_red,
  "muller_corsm"              = irw_green
)
model_labels <- c(
  "naive_linear"              = "Naive linear",
  "beta_irt_item_dispersion"  = "Beta IRT (item disp.)",
  "beta_irt_fixed_dispersion" = "Beta IRT (fixed disp.)",
  "samejima_crm"              = "Samejima CRM",
  "muller_corsm"               = "Mueller CRSM"
)

cache_path <- "continuous_bounded_data/continuous_bounded_results.rds"
cache_ok <- file.exists(cache_path)
if (cache_ok) res <- readRDS(cache_path)

grid_cache_path <- "continuous_bounded_data/continuous_bounded_grid_results.rds"
grid_cache_ok <- file.exists(grid_cache_path)
if (grid_cache_ok) grid_res <- readRDS(grid_cache_path)
```

## Motivation

Visual analogue scales, slider ratings, and continuous confidence or
percentage judgments are common: a respondent marks a point on a
fixed, labeled segment (0 to 100, say), and the raw response is a
continuous number that is *bounded* at both ends. This is a different
measurement problem from the unbounded continuous case (response times,
open-ended magnitude estimates), which reduces to standard linear-factor
or CFA modeling and is covered elsewhere on this site -- the bounds
themselves are the interesting part here, because they create floor/ceiling
effects and non-constant response variance near the edges that a plain
linear model doesn't see.

Three published models handle this directly. Noel and Dauvier's Beta IRT
model [@noel2007beta] treats the response as Beta-distributed given
ability, with the two natural Beta parameters reparameterized into an
item difficulty and a dispersion. Samejima's Continuous Response Model
[@samejima1973homogeneous] models a monotonic transform of the response as
normal given ability -- the historically original approach, predating Beta
IRT by over 30 years. Mueller's Continuous Rating Scale Model
[@mueller1987rasch] takes a Rasch-family, specific-objectivity approach,
estimated by pairwise conditional likelihood; Verhelst's (2019) exponential
family framework [@verhelst2019exponential] is the modern theoretical
generalization of Mueller's model, though the model actually fitted here is
Mueller's original. We compare these three against a naive linear (Gaussian,
identity-link) baseline that ignores the bounds entirely -- the "what if you
just treat it like ordinary continuous data" control, playing the same role
the Rasch baseline plays in the site's guessing-models vignette.

## Data and methods

### The four core specs

1. **Naive linear baseline** -- single-factor CFA via `lavaan`, identity
   link, Gaussian residual, fit directly on the raw (rescaled-to-[0,1])
   response. Ignores boundedness entirely.
2. **Beta IRT** [@noel2007beta] -- fit as a discretized custom `mirt`
   itemtype wrapping `sirt::brm.irf()`, following `sirt`'s own documented
   mirt-estimation recipe (`?sirt::brm.sim`). Two variants: **item-varying
   dispersion** (each item gets its own dispersion parameter, the model's
   default parameterization) and **fixed dispersion** (one dispersion
   parameter shared across all items, via an `mirt` parameter constraint).
3. **Samejima's Continuous Response Model** [@samejima1973homogeneous] --
   `EstCRM::EstCRMitem()` / `EstCRMperson()`.
4. **Mueller's Continuous Rating Scale Model** [@mueller1987rasch] -- a
   patched copy of `pcIRT::CRSM()` / a from-scratch reimplementation of its
   person-scoring equation (see [Limitations](#limitations) for why
   patching was necessary).

Ferrando's (2001) nonlinear congeneric model [@ferrando2001nonlinear] is an
optional stretch spec with no package implementation (it would need a
custom truncated-normal likelihood via `optim`) and has not been attempted
here -- see [Limitations](#limitations).

### Model equations

Each spec below is written exactly as it is fit in the compute script --
same parameter names, same functional form -- so the equations here can be
checked directly against `continuous_bounded_helpers.R`.

**Naive linear baseline.** Ordinary single-factor CFA on the raw
(rescaled-to-$[0,1]$) response: $X_{ij} = \nu_j + \lambda_j\theta_i +
\varepsilon_{ij}$, $\varepsilon_{ij}\sim N(0,\sigma_j^2)$, fit via
`lavaan::cfa()`. Nothing here constrains $X_{ij}$ to $[0,1]$.

**Beta IRT** [@noel2007beta], as `sirt::brm.irf()` actually parameterizes
it: $X_{ij}\mid\theta_i \sim \text{Beta}(\text{shape1}_{ij},
\text{shape2}_{ij})$, with
$$\text{shape1}_{ij} = \exp\!\left(\frac{\theta_i-\delta_j+\tau_j}{2}\right),
\qquad
\text{shape2}_{ij} = \exp\!\left(\frac{-\theta_i+\delta_j+\tau_j}{2}\right),$$
where $\delta_j$ is item difficulty and $\tau_j$ is dispersion (fixed
across items in the "fixed dispersion" variant, free in the "item-varying"
variant). Working out this Beta distribution's mean shows it reduces
exactly to $E[X_{ij}\mid\theta_i] = \text{logistic}(\theta_i-\delta_j)$,
independent of $\tau_j$ -- **there is no discrimination parameter**;
dispersion changes the curve's spread around that mean, not its
steepness or location. This is why the widget below has no effect from
the discrimination slider on the Beta IRT curve.

**Samejima's Continuous Response Model** [@samejima1973homogeneous], via
`EstCRM::EstCRMitem()`: the raw response is transformed to
$Z_{ij}=\ln\!\big(X_{ij}/(1-X_{ij})\big)$, and $Z_{ij} = \alpha_j(\theta_i -
b_j) + e_{ij}$, $e_{ij}\sim N(0,\sigma_j^2)$, i.e. an ordinary linear factor
model *on the transformed scale*. $\alpha_j$ is discrimination and $b_j$ is
difficulty. The observed-scale mean response is the backtransform of this
normal model's mean through the logistic link.

**Mueller's Continuous Rating Scale Model** [@mueller1987rasch], via
`pcIRT::CRSM()`: a Rasch-family (specific-objectivity) density,
$$f(x_{ij}\mid\theta_i) \;\propto\; \exp\!\big[x_{ij}(\theta_i-b_j) +
x_{ij}(1-x_{ij})\lambda_j\big], \quad x_{ij}\in[0,1],$$
with a single item-location parameter $b_j$ and a dispersion parameter
$\lambda_j$ -- no discrimination parameter, by construction of the
Rasch-family specific-objectivity property, which is why moving the
discrimination slider below has no effect on this curve either.

### Comparing the item response functions across models

Before fitting anything to data, it helps to see what each model's own
equation implies about the relationship between a respondent's ability
$\theta$ and their expected response. Naive linear and Samejima CRM both
have a genuine discrimination parameter ($a$ in the naive case, $\alpha$
for Samejima); Beta IRT and Mueller CRSM do not, as shown above. Drag the
sliders below to see how difficulty ($b$) and discrimination ($a$) reshape
each model's mean response curve.

```{ojs}
//| echo: false
viewof irf_b = Inputs.range([-2, 2], {value: 0, step: 0.1, label: "Item difficulty (b)"})
```

```{ojs}
//| echo: false
viewof irf_a = Inputs.range([0.3, 2.5], {value: 1, step: 0.1, label: "Discrimination (a) -- no effect on Beta IRT or Mueller CRSM"})
```

```{ojs}
//| echo: false

function normalPDF(z, mu, sigma) {
  return Math.exp(-0.5 * ((z - mu) / sigma) ** 2) / (sigma * Math.sqrt(2 * Math.PI));
}

function simpson(f, lo, hi, n) {
  const h = (hi - lo) / n;
  let s = f(lo) + f(hi);
  for (let i = 1; i < n; i++) {
    s += (i % 2 === 0 ? 2 : 4) * f(lo + i * h);
  }
  return (s * h) / 3;
}

irf_sigma = 1
irf_lambda = 2

function betaIrtMean(theta, b) {
  return 1 / (1 + Math.exp(-(theta - b)));
}

function samejimaMean(theta, a, b, sigma) {
  const loc = a * (theta - b);
  return simpson((z) => (1 / (1 + Math.exp(-z))) * normalPDF(z, loc, sigma), loc - 6 * sigma, loc + 6 * sigma, 60);
}

function muellerMean(theta, b, lambda) {
  const g = (x) => Math.exp(x * (theta - b) + x * (1 - x) * lambda);
  const Z = simpson(g, 0, 1, 100);
  return simpson((x) => x * g(x), 0, 1, 100) / Z;
}

irf_thetas = d3.range(-3, 3.01, 0.1)

irf_curve_data = irf_thetas.flatMap((theta) => [
  { theta, model: "naive_linear", y: 0.5 + irf_a * (theta - irf_b) },
  { theta, model: "beta_irt", y: betaIrtMean(theta, irf_b) },
  { theta, model: "samejima_crm", y: samejimaMean(theta, irf_a, irf_b, irf_sigma) },
  { theta, model: "muller_corsm", y: muellerMean(theta, irf_b, irf_lambda) }
])
```

```{ojs}
//| echo: false

irf_colors = ({
  naive_linear: "#808080",
  beta_irt: "#2166ac",
  samejima_crm: "#d73027",
  muller_corsm: "#4daf4a"
})

irf_labels = ({
  naive_linear: "Naive linear",
  beta_irt: "Beta IRT",
  samejima_crm: "Samejima CRM",
  muller_corsm: "Mueller CRSM"
})

Plot.plot({
  width: 700,
  height: 420,
  x: {label: "Ability (θ)", domain: [-3, 3]},
  y: {label: "Expected response (rescaled to [0,1])", domain: [-0.15, 1.15]},
  color: {
    domain: Object.keys(irf_colors),
    range: Object.values(irf_colors),
    legend: true,
    tickFormat: (d) => irf_labels[d]
  },
  marks: [
    Plot.ruleY([0, 1], {stroke: "#999", strokeDasharray: "2,2"}),
    Plot.line(irf_curve_data.filter((d) => d.model === "naive_linear"), {
      x: "theta", y: "y", stroke: "model", strokeDasharray: "4,3"
    }),
    Plot.line(irf_curve_data.filter((d) => d.model !== "naive_linear"), {
      x: "theta", y: "y", stroke: "model"
    })
  ]
})
```

The naive linear line (dashed) is the only curve not actually confined to
$[0,1]$: away from $\theta=b$ it predicts responses outside the possible
range entirely, which is the exact failure mode the other three models are
built to avoid by leveling off smoothly at the boundary instead. Neither
Beta IRT (see [Model equations](#model-equations)) nor Mueller's CRSM --
being a Rasch-family, specific-objectivity model [@mueller1987rasch] -- has
a discrimination parameter, so moving the $a$ slider only changes the
naive linear and Samejima CRM curves. Dispersion parameters are held fixed here
at representative values ($\sigma=1$ for Samejima CRM, $\lambda=2$ for
Mueller CRSM) rather than exposed as sliders: Beta IRT's mean curve doesn't
depend on its dispersion parameter at all (only its spread does), while
Samejima's and Mueller's dispersion parameters do shift the mean curve
itself once backtransformed through a nonlinear link, not just its spread.

### Estimation and model comparison

Every model's held-out fit is reported as the **log-probability of the
response falling in a small interval**, not a raw point density. This
choice exists because the 4 models don't report likelihood on the same
scale to begin with: reading `EstCRM::EstCRMitem()`'s source shows it
transforms the raw response to $Z = \ln\!\big(X/(K-X)\big)$ and computes
its entire EM log-likelihood on that transformed scale, not the original
response scale the Beta IRT and naive linear fits use. Comparing those two
log-likelihoods directly would be comparing different units -- like
comparing an error reported in inches to one reported in centimeters
without converting first. The usual fix is a model-specific Jacobian
correction; instead, every model's held-out score here is
$\log P(\text{bin}_{lo} < X < \text{bin}_{hi})$, using the same bin edges
(on the raw, rescaled-to-$[0,1]$ scale) for all 4 models. A probability that
a response lands in a fixed interval means the same thing regardless of
which internal scale a model uses to compute it, so this sidesteps the
rescaling problem entirely rather than requiring a per-model correction.

Held-out log-probability, built this way, is the metric used to compare all
4 specs throughout this vignette. It's a standard, interpretable scale for
predictive fit, though not a validated best-practice specific to this
bounded-continuous setting -- flagged as an ad hoc choice worth revisiting
(see [Limitations](#limitations)).

### Held-out design: genuine missing-response prediction

The project's usual convention for held-out evaluation on IRW data is a
response-level (missing-cell) holdout: mask a random subset of person x item
cells, fit on the rest, score the masked cells. Fitting item parameters this
way isn't viable for 2 of the 4 specs -- reading `EstCRM::EstCRMitem()` and
`pcIRT::CRSM()`'s fitting code directly shows **neither has any
missing-data handling at all** (both do matrix/pairwise algebra that
silently breaks with `NA` cells, confirmed by testing) -- so item
parameters for all 4 specs are still fit on a complete training-person
matrix, no missing cells.

*Scoring*, once item parameters are fixed, is a different matter: all 4
specs' person-location estimators turn out to be NA-tolerant.
`mirt::fscores(response.pattern = ...)` handles missing cells natively (a
core mirt feature, not something added here); `lavaan::lavPredict()`
computes factor scores from an incomplete `newdata` row the same way;
`EstCRM::EstCRMperson()` skips `NA` columns per person when forming its
theta estimate (confirmed by reading its source); and the from-scratch
Mueller CRSM person-scoring routine (see [Limitations](#limitations)) was
written to do the same. Each was smoke-tested directly (masked vs.
full-data theta estimates, correlation $>0.98$ in every case) before
being used for real.

That makes true cell-level prediction possible after all: for every
held-out **test person**, a random subset of their items (25%, clamped to
at least 1 kept and at least 1 held out) is hidden before estimating that
person's ability from the remaining items alone, and only the hidden items
are scored. The model never sees the scored value, directly or indirectly,
at any stage -- this is genuine missing-response prediction, not the
weaker "estimate ability from a person's full vector, then score those same
items" self-consistency check an earlier version of this design used.

### Simulation

**A single illustrative condition** ($N=300$, $J=10$, Beta IRT DGP, fixed
dispersion) is shown first below, generated directly from the Beta IRT
model's own equations (`sim_beta_irt()` in `continuous_bounded_helpers.R`
-- a bug-fixed reimplementation of `sirt::brm.sim()`'s logic; see
[Limitations](#limitations) for why the reimplementation was necessary).

**The full recovery/misspecification grid** cross-fits all 4 core specs
against 3 DGPs, generated from each model's own equations: Beta IRT
(`sim_beta_irt()`), Samejima CRM (`sim_samejima_crm()`: $Z_{ij} =
\alpha_j(\theta_i - b_j) + e_{ij}$, $e_{ij}\sim N(0,\sigma_j^2)$, then
$X_{ij}=1/(1+e^{-Z_{ij}})$ -- the exact inverse of EstCRM's own $Z=\ln(X/(1-X))$
transform), and Mueller CRSM (`sim_muller_corsm()`: grid-based inverse-CDF
sampling from the same $\exp[x(\theta-b)+x(1-x)\lambda]$ density used for
scoring, since no closed-form sampler exists for this family). The
naive-linear DGP is skipped: an unbounded-Gaussian DGP clipped to $[0,1]$ to
make it "bounded" would introduce an artificial boundary spike
unrepresentative of what a linear model actually generates.

Grid: $N\in\{300,1000\}$, $J\in\{8,20\}$, and, for the Beta IRT DGP
specifically, fixed vs. item-varying dispersion -- 8 + 4 + 4 = 16
conditions, plus 1 boundary-inflation condition (10% of responses forced to
exactly 0, 10% forced to exactly 1, layered on the Beta IRT DGP at
$N=300$/$J=8$) for 17 total. Run via `furrr::future_pmap()` +
`plan(multisession)`, cached separately from the real-data results below
(own skip-if-exists gate in the compute script).

### Real-data selection

Candidate tables were identified by scouting IRW for tables with
`irw_metadata()$n_categories >= 12` whose response range and `irw_tags()`
item-format both point to a bounded, fixed-scale response, then narrowed to
tables that pass structural fitness checks for these 4 models specifically
(unique `(id, item)` pairs -- reducing to a single wave/rater where IRW's
`wave`/`rater` columns require it, no ipsative/compositional row-sum
constraint, and genuine fractional-value evidence rather than a merely
fine-grained integer scale). The 7 tables that survive, used for every
real-data result below:

| Table | N | J | Note |
|---|---|---|---|
| `lsbq_maleki_2025_non_persian_proficiency` | 308 | 3 | 0-10 scale, ~40% non-integer responses -- primary evidence of genuine continuity |
| `ai_fear_dong_2026_ai` (Doctor facet, subsampled) | 2,000 | 8 | 0-100, integer-valued; large-*N* structural check |
| `much_tte_2025_currentmotivation` | 1,244 | 3 | wave-filtered to one occasion |
| `test_taking_much_2025_cm` | 1,244 | 9 | -- |
| `nas_rogoza_2024_study5_nas` | 277 | 4 | wave-filtered; Narcissistic Antagonism subscale |
| `nas_rogoza_2024_study5_ngs` | 274 | 4 | wave-filtered; Narcissistic Grandiosity subscale |
| `nas_rogoza_2024_study5_nvs` | 278 | 4 | wave-filtered; Narcissistic Vulnerability subscale |

`ai_fear_dong_2026_ai` is a 48-item, 6-target x 8-trait factorial design
(Care/Doctor/Journalist/Judge/Manager/Religious x competent/warm/etc.); only
one target's 8 items (`Doctor_*`) are used here as a single "overall
impression of Doctor" construct, not the full table.

Full scouting and per-table vetting code, cached candidate lists, and the
exact reason every non-included candidate was set aside all live in
`vignettes/continuous_scout_*` and `vignettes/continuous_bounded_data/
vet_candidates*.R` / `vet_results*.csv` -- start there rather than
re-deriving the logic if extending this vignette to more tables.

## Results

### Simulated data: does the correctly-specified model win, and does it recover parameters?

```{r}
#| label: fig-sim-scores
if (cache_ok) {
  res$sim$scores |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = sprintf("%.2f", mean_ll)), vjust = 1.4, color = "white", size = 3.5) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = "Simulated Beta-IRT data: held-out fit by model") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

On data simulated directly from the Beta IRT model, the two Beta IRT
variants have the best (least negative) held-out fit, essentially tied with
Samejima CRM and Mueller CRSM, and all three clearly beat the naive linear
baseline. This is the expected pattern for a correctly-specified-model
recovery check: the models built for bounded continuous responses recover
the truth better than the one that ignores the bounds.

Item-difficulty recovery for the item-varying-dispersion Beta IRT fit
(correlation between estimated and true $\delta$) was
`r if (cache_ok) sprintf("%.3f", res$sim$recovery_cor_delta) else "NA"`.

### The full recovery grid: 16 conditions x 3 DGPs

Parameter recovery for the correctly-specified model was excellent
everywhere -- every one of the 16 non-inflation conditions recovered its
true item parameters at $r>0.96$, most above $r>0.99$:

```{r}
#| label: tbl-grid-recovery
if (grid_cache_ok) {
  grid_res$scores |>
    filter(!boundary_inflation) |>
    distinct(dgp, N, J, dispersion_mode, recovery_cor) |>
    arrange(dgp, N, J, dispersion_mode) |>
    mutate(recovery_cor = sprintf("%.3f", recovery_cor)) |>
    knitr::kable(col.names = c("DGP", "N", "J", "Beta IRT dispersion", "Recovery r"))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

More useful than "does recovery work" (it does, everywhere) is the
cross-fitting question: when data comes from one DGP, which of the 4
*fitted* specs wins the held-out comparison?

```{r}
#| label: fig-grid-crossfit
if (grid_cache_ok) {
  grid_res$scores |>
    filter(!boundary_inflation) |>
    mutate(model = factor(model, levels = names(model_colors)),
           dgp_label = recode(dgp, beta_irt = "DGP: Beta IRT", samejima_crm = "DGP: Samejima CRM",
                               muller_corsm = "DGP: Mueller CRSM")) |>
    group_by(dgp_label, model) |>
    summarise(mean_ll = mean(mean_ll, na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    facet_wrap(~dgp_label, ncol = 3) +
    labs(x = NULL, y = "Mean held-out log-probability\n(averaged across N/J conditions)",
         title = "Cross-fitting: held-out fit by model, averaged across all N/J conditions per DGP") +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

The pattern is consistent across all 3 DGPs and matches every real-data
result above: **the 3 bounded-response models cluster tightly together and
substantially outperform the naive linear baseline, regardless of which of
the 3 actually generated the data.** "Correctly specified wins" is a much
weaker pattern here than under a self-consistency check, though: under the
Beta IRT DGP, both Beta IRT variants edge out Samejima CRM by only
$\approx$0.003 in mean held-out log-probability -- a three-way tie in
practice. Under the Samejima CRM DGP, Samejima does come out slightly
ahead, but by less than 0.01. And under the Mueller CRSM DGP, Mueller CRSM
is **not** the best-fitting model -- Samejima CRM scores marginally better
than Mueller's own generating model, a genuine (if small) misspecification
surprise, not papered over here. The margin between the 3 bounded-response
models is small compared to the gap to naive linear in every case -- these
3 models are more interchangeable with each other than any of them is with
a model that ignores the bounds, but which one "wins" under true
missing-response prediction is close to a coin flip. One fit failed
outright:
Samejima CRM's EM did not converge (a `missing value where TRUE/FALSE
needed` error partway through) on one Mueller-CRSM-generated dataset
($N=300$, $J=8$) -- a real misspecification-robustness finding, not
papered over; see [Limitations](#limitations).

### Boundary inflation: does floor/ceiling piling break anything?

Beyond fitting, we asked how each spec's held-out fit *degrades* when 20%
of responses are forced to the exact boundary (10% to 0, 10% to 1) on top
of the Beta IRT DGP -- diagnostic information, not a 5th model to fit:

```{r}
#| label: fig-grid-inflation
if (grid_cache_ok) {
  grid_res$scores |>
    filter(dgp == "beta_irt", N == 300, J == 8, dispersion_mode == "fixed") |>
    mutate(model = factor(model, levels = names(model_colors)),
           condition = ifelse(boundary_inflation, "20% boundary-inflated", "Not inflated")) |>
    ggplot(aes(x = model, y = mean_ll, fill = model, alpha = condition)) +
    geom_col(position = "dodge", width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_alpha_manual(values = c("Not inflated" = 1, "20% boundary-inflated" = 0.55), name = NULL) +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Mean held-out log-probability",
         title = "Beta IRT DGP (N=300, J=8): held-out fit with vs. without boundary inflation") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  cat("Grid cache not found -- run continuous_bounded_compute.R first.")
}
```

Every spec's held-out fit gets worse under inflation -- none of the 4 core
specs explicitly models a zero-one-inflated mixture, so this is expected --
but **not by the same amount**. The naive linear baseline degrades by
about 20% (mean log-probability drops from -1.74 to -2.09); the 3
bounded-response models degrade by roughly 8-10% each (e.g. Mueller CRSM:
-1.46 to -1.61). The bounded-response models' smooth density still puts
more mass near the boundary than a Gaussian does, so they absorb some of
the extra piling without a dedicated inflation term, while the naive linear
model has nothing in its structure to absorb it at all. None of the 4
handle it *well* -- a true zero-one-inflated extension (Molenaar et al.,
2022) would be needed for that -- but the bounded-response models are
comparatively more robust to it than the baseline that ignores boundedness
altogether.

### Real data: does floor/ceiling piling look the way the models expect?

Before comparing model fits, it helps to see what these tables' response
distributions actually look like. This is the primary table's ~40%
non-integer evidence from [Real-data selection](#real-data-selection) made
visual, and it's also the direct reason the naive linear baseline
underperforms below: a Gaussian model expects mass concentrated
symmetrically around the mean, not piled up against a hard boundary.

```{r}
#| label: fig-real-dist
if (cache_ok) {
  as.data.frame(res$real$Y_raw) |>
    tidyr::pivot_longer(everything(), names_to = "item", values_to = "resp") |>
    ggplot(aes(x = resp)) +
    geom_histogram(binwidth = 0.5, boundary = 0, fill = irw_blue, color = "white", linewidth = 0.1) +
    facet_wrap(~item, ncol = 3) +
    labs(x = sprintf("Raw response (%g-%g scale)", res$real$low, res$real$high),
         y = "Count",
         title = sprintf("%s: raw response distribution by item", res$real$table))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

All 3 items show a similar shape: a broad hump across the mid-range plus a
sharp spike at the upper boundary. Between 23% and 44% of responses per
item sit at *exactly* 10, the scale ceiling (vs. under 9% at exactly 0) --
respondents who are, or report being, fully proficient pile up exactly at
the boundary, which a symmetric Gaussian around a single mean cannot
represent. The same ceiling pile-up shows up, to varying degrees, in the
other 6 tables analyzed below -- it's the general floor/ceiling behavior
the 3 bounded-response models are built to handle and the naive linear
model is not.

### Held-out fit across all 7 real tables

```{r}
#| label: prep-real-scores
if (cache_ok) {
  all_real_scores <- dplyr::bind_rows(
    res$real$scores |> dplyr::mutate(N = res$real$N, J = res$real$J),
    res$real2$scores |> dplyr::mutate(
      N = res$real2$N, J = res$real2$J,
      label = paste0(label, " (", res$real2$facet, ")")
    ),
    res$batch$scores
  )
}
```

```{r}
#| label: fig-real-all-scores
if (cache_ok) {
  all_real_scores |>
    mutate(model = factor(model, levels = names(model_colors)),
           label = sprintf("%s\n(N=%d, J=%d)", label, N, J)) |>
    ggplot(aes(x = model, y = mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    facet_wrap(~label, ncol = 4) +
    labs(x = NULL, y = "Mean held-out log-probability (per response)",
         title = "Held-out fit by model, across all 7 real IRW tables analyzed") +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

The pattern is consistent across all 7 tables and matches the simulation
results above: the naive linear baseline is clearly worst everywhere, often
by a wide margin (e.g. on the primary
`lsbq_maleki_2025_non_persian_proficiency` table, naive linear's mean
held-out log-probability is -6.08, against -1.13 to -2.51 for the 3
bounded-response models). Unlike the simulation grid, a real ordering does
emerge among the 3 bounded-response models here: **Samejima CRM is the
worst of the 3 on every single one of the 7 real tables** -- sometimes by a
wide margin (`lsbq`: -2.51 vs. -1.13 to -1.51 for the other two), sometimes
narrowly (`ai_fear_dong_2026_ai`: -1.31 vs. -1.18). Beta IRT (either
dispersion variant, interchangeably) wins outright on 4 of 7 tables
(`lsbq`, `ai_fear_dong_2026_ai`, `much_tte_2025_currentmotivation`,
`test_taking_much_2025_cm`); Mueller CRSM wins the other 3, all narcissism
subscales from `nas_rogoza_2024_study5` (`nas`/`ngs`/`nvs`), with Beta IRT
a close second in each. One plausible reason Samejima CRM trails here but
not in simulation: its held-out score depends on an empirically-estimated
residual SD on the transformed $Z$ scale (see `.crm_resid_sd()` in
`continuous_bounded_helpers.R`, chosen because EstCRM's own EM-shrinkage
precision term isn't straightforward to reverse-engineer), which may
generalize to genuinely new items less well than the other 2 specs' fitted
dispersion parameters do -- a real cost of the workaround, not asserted
with certainty. Each table has only one train/test split (see
[Limitations](#limitations)). The main takeaway holds regardless of which
of the 3 wins: all 3 bounded-response models consistently and
substantially beat the naive linear baseline across every real table
analyzed, on genuinely held-out item responses.

### Predicted responses: point estimates, not just probabilities

Held-out log-probability answers "how much probability mass did the model
put on the interval the true response landed in" -- a distributional
question. It doesn't directly answer the more concrete question the site's
binary-response vignettes usually lead with (e.g. `asymmetric_irt.qmd`'s
predicted-vs-actual convention): if you had to guess the actual missing
value, how close would you get? So alongside every log-probability number
above, each model's fitted parameters and estimated theta also produce a
genuine **point prediction** for the same held-out cells -- $E[X_{ij}\mid
\theta_i]$, the closed-form or numerically-integrated mean derived in
[Model equations](#model-equations) for each spec -- compared directly to
the actual masked value via RMSE and correlation.

```{r}
#| label: fig-real-rmse
if (cache_ok) {
  all_real_scores |>
    mutate(model = factor(model, levels = names(model_colors)),
           label = sprintf("%s\n(N=%d, J=%d)", label, N, J)) |>
    ggplot(aes(x = model, y = rmse_pred, fill = model)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = model_colors, labels = model_labels, guide = "none") +
    scale_x_discrete(labels = model_labels) +
    facet_wrap(~label, ncol = 4) +
    labs(x = NULL, y = "RMSE of predicted vs. actual response (held-out)",
         title = "Predicted-response accuracy by model, across all 7 real IRW tables") +
    theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

This is where the story genuinely changes. **Naive linear's point
predictions are competitive with, and often better than, all 3
bounded-response models** -- the opposite of the log-probability picture.
Naive linear has the *lowest* RMSE of all 5 specs on 4 of the 7 tables --
`much_tte_2025_currentmotivation` (0.0999), `test_taking_much_2025_cm`
(0.1048), `nas_rogoza_2024_study5_ngs` (0.1799), and, most strikingly, the
primary `lsbq_maleki_2025_non_persian_proficiency` table (0.2238, ahead of
Samejima CRM's 0.2331, Beta IRT's 0.2459/0.2498, and Mueller's 0.2680) --
despite having by far the worst log-probability everywhere, including a
catastrophic -6.08 on `lsbq` against -1.13 to -2.51 for the 3
bounded-response models on that same table. It's essentially tied for best
on `ai_fear_dong_2026_ai` (0.1961 vs. Beta IRT's 0.1958), and only clearly
middling on 2 of the 7 (`nas`, `nvs`).

This isn't a contradiction -- it's what the two metrics are built to catch.
RMSE only checks whether the *central tendency* is right; log-probability
also checks the *shape* of the distribution around it (tail behavior,
non-constant variance near the boundaries, how much mass piles up at 0
and 1). A linear model's conditional mean can be a perfectly reasonable
guess at where a response will land even while its Gaussian, constant-variance
error structure is badly wrong about how *confident* to be near a boundary --
exactly the floor/ceiling piling shown in [the histogram
above](#real-data-does-floorceiling-piling-look-the-way-the-models-expect).
So: if the applied question is "predict this person's missing response as a
single number," the extra machinery of a bounded-response model buys little
to nothing here. If the applied question involves *uncertainty* --
uses fitted probabilities, is calibration-sensitive, or the response
sits near a boundary and getting the shape right actually matters -- the
naive baseline's advantage disappears entirely and the 3 bounded models
are the clear right choice.

```{r}
#| label: fig-real-pred-scatter
if (cache_ok) {
  res$real$predictions |>
    mutate(model = factor(model, levels = names(model_colors))) |>
    ggplot(aes(x = actual, y = predicted, color = model)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(alpha = 0.6, size = 1.6) +
    scale_color_manual(values = model_colors, labels = model_labels, guide = "none") +
    facet_wrap(~model, ncol = 3, labeller = as_labeller(model_labels)) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(x = "Actual (held-out) response", y = "Predicted response",
         title = sprintf("Predicted vs. actual held-out response: %s", res$real$table))
} else {
  cat("Cache not found -- run continuous_bounded_compute.R first.")
}
```

The scatter for the primary table makes the RMSE numbers concrete: all 5
specs track the dashed 45-degree line reasonably well across the middle of
the range, and none of them is dramatically better calibrated to the *point
values themselves* than the others -- the real differentiator among models
in this vignette is the log-probability metric, not this one.

## Limitations {#limitations}

Known gaps this vignette does not yet address:

- **Ferrando's (2001) nonlinear congeneric model** [@ferrando2001nonlinear]
  -- the optional stretch spec, comparable to the treatment in Ferrando
  (2002) [@ferrando2002theoretical] -- has not been attempted at all. Specs
  1-4 are now complete and validated (the grid ran cleanly), so its own
  sequencing rule no longer blocks it -- it just hasn't been picked up yet.
- **Partial missingness in real data.** All 7 real tables used complete
  cases only *before* the evaluation design's own synthetic masking is
  applied; IRW tables whose respondents have naturally-occurring partial
  missingness (as opposed to the missingness this vignette introduces
  on purpose, for evaluation) aren't handled yet.
- **The 25% per-person item-masking fraction (`MASK_FRAC`) was not
  varied or sensitivity-tested** -- a different fraction could shift how
  much information theta estimation has to work with, and thus the
  absolute (though probably not the relative, across-model) held-out
  numbers above.
- **Only one train/test split per condition** (real or simulated), not a
  repeated/cross-validated design -- so none of the held-out numbers above
  have an uncertainty band, and a single unlucky split could shift a close
  call between the 3 bounded-response models.
- **One grid fit failed outright**: Samejima CRM's EM did not converge
  (`missing value where TRUE/FALSE needed`) on one Mueller-CRSM-generated
  condition ($N=300$, $J=8$, no boundary inflation). Not re-run with a
  different seed or diagnosed further -- reported as a real
  misspecification-robustness finding (EstCRM's EM can fail outright, not
  just fit poorly, on data far from its assumptions) rather than treated as
  a bug to route around.
- **The grid's boundary-inflation condition covers only 1 of the 3 DGPs**
  (Beta IRT, $N=300$/$J=8$) -- not repeated for Samejima CRM or Mueller
  CoRSM, or at other N/J combinations.

**Package-availability and correctness issues found while building this**
(worth knowing before extending this vignette):

- `pcIRT::CRSM()` (Mueller's model) was **dropped from the package
  entirely** in the last CRAN release (0.2.4, 2019); it only exists in
  pcIRT $\leq$ 0.2.3. `pcIRT` itself was later archived by CRAN entirely
  (2025-11-11). This vignette's compute script installs 0.2.3 from the
  CRAN archive, not the current CRAN release -- a real risk for
  reproducing this vignette in the future if the archive changes.
- That same version's `CRSM()` has a convergence-check bug
  (`is.na(para) || ...` where `para` becomes a 2-vector after the first
  iteration) that is a hard error under R $\geq$ 4.3 and had to be patched
  locally before it would run at all.
- `sirt::brm.sim()` hardcodes a 9-item assumption in its column-naming
  code and errors for any other item count; `sim_beta_irt()` in the
  helpers file reimplements its generative logic directly rather than
  patching `sirt`.
- `pcIRT::person_par.CRSM()` only scores the same persons used in the
  original fit and has no new-data argument, so it could not be reused for
  scoring held-out test persons; a generalized version of its estimating
  equation was written from scratch instead (`theta_new_muller_corsm()`
  below) -- this turned out to be a genuine improvement over the original,
  since it also skips `NA` items per person, which is exactly what made
  cell-level missing-response prediction possible for this spec.
- The from-scratch Mueller CRSM person-scoring routine
  (`theta_new_muller_corsm()`) could diverge to a `theta` extreme enough to
  overflow `exp()` inside `integrate()`, which throws a hard "non-finite
  function value" error rather than the graceful NA `stop.on.error=FALSE`
  is supposed to allow -- this crashed the *entire* held-out scoring pass
  for a table (not just Mueller CRSM) before being caught and fixed by
  clamping `theta` and wrapping each `integrate()` call individually.
- Similarly, a Beta IRT scoring-stage failure (`mirt` compacts a training
  split's observed categories to consecutive integers per item, and rejects
  a held-out response that lands in a category with zero training-split
  observations) was originally written so it could crash
  naive-linear/Samejima/Mueller's results for the same table too -- every
  model's fit-and-score block is now wrapped in its own `tryCatch` so one
  model's failure can't take down the other 3.
- **The current 7 real tables are not the full set of viable candidates.**
  A handful of additional tables are plausible additions but need a fresh
  vetting pass before inclusion; see the vetting scripts/CSVs referenced in
  [Real-data selection](#real-data-selection) for the working record of
  what's been checked so far.

**Explicitly out of scope** (not implemented, not planned for this
vignette): DIF for continuous responses [@finch2023impact]; dual-range
slider / interval-response models and the Interval Consensus Model
[@kloft2024measuring; @kloft2023dirichlet; @kloft2025interval] -- a
different response format (an interval, not a point) and, for the
Consensus Model, a different modeling paradigm (cultural consensus
theory) entirely; Zopluoglu's mixture CRM [@zopluoglu2020finite] and
Noel's beta unfolding model [@noel2014beta] -- each adds a full second axis
of complexity (latent mixture classes; unfolding vs. dominance response
process), and unfolding in particular is a more natural extension of the
site's existing unfolding-vs-dominance vignette than a duplicate here;
Ferrando and Navarro-Gonzalez's (2021) multidimensional Dual Thurstonian
model [@ferrando2021multidimensional].

**The comparison metric (held-out log-likelihood) is an ad hoc choice**,
not a validated best-practice for this setting -- flagged explicitly above,
worth revisiting.

## Reproducibility

```r
# 1. Run the compute script (from the project root)
source("vignettes/continuous_bounded_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/continuous_bounded.qmd")
```

::: {.callout-tip}
To cite the real-data table used here, run:

```r
irw_save_bibtex("lsbq_maleki_2025_non_persian_proficiency",
                 output_file = "continuous_bounded_data/irw_references.bib")
```
:::

---

## References
