---
title: "Dimensionality Across IRW: How Often Does Unidimensionality Hold?"
description: |
  Every unidimensional IRT model used elsewhere on this site -- 1PL, 2PL,
  the IMV comparisons -- assumes item responses are explained by a single
  latent trait. We compute item correlation matrices for every eligible
  IRW table and run two classic exploratory diagnostics, the eigenvalue
  ratio and parallel analysis, to see how often that assumption actually
  holds and whether it varies by construct type.
date: today
bibliography: dimensionality_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "dimensionality_data/dimensionality_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  # construct_type from IRW's tags is sometimes a comma-joined multi-label
  # string (e.g. "Cognitive/educational, Behavioral"); group by the first
  # (primary) label so the by-construct breakdown below isn't fragmented
  # into dozens of near-singleton combinations.
  summary_df$construct_type <- sub(",.*$", "", summary_df$construct_type)
  eigenvalues_list <- .cache$eigenvalues %||% list()
  candidate_tables <- .cache$candidate_tables
  date_run         <- .cache$date_run
  is_pilot         <- isTRUE(.cache$pilot)
  ratio_cutoff     <- .cache$ratio_cutoff %||% 4
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_tables_run   <- length(candidate_tables)
} else {
  summary_df <- tibble(
    table = character(), n_items = integer(), n_participants = integer(),
    n_categories = integer(), cor_method = character(), ratio_12 = numeric(),
    prop_var_1 = numeric(), nfact_suggested = integer(),
    unidimensional_pa = logical(), construct_type = character(),
    item_format = character()
  )
  eigenvalues_list <- list()
  candidate_tables <- character()
  date_run         <- Sys.Date()
  is_pilot         <- FALSE
  ratio_cutoff     <- 4
  r_all_candidates <- 0L
  r_n_tables_run   <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/dimensionality_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, partial run.** This render uses a partial sample of ",
  r_n_tables_run, " tables out of ", r_all_candidates, " eligible tables ",
  "(either a small hand-picked/random pilot, or a snapshot taken while the ",
  "full batch was still running against Redivis), to validate the pipeline ",
  "and page layout before the full batch finishes. All numbers here are ",
  "real computed results -- not placeholders -- but treat the ",
  "*distributions* as a small, noisy preview rather than a representative ",
  "picture of dimensionality across the IRW. Re-run ",
  "`dimensionality_compute.R` (with `PILOT <- FALSE` and the full batch ",
  "complete) to scale up.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`dimensionality_compute.R`) directly if you're
relying on the numbers here.
:::

## Motivation

Every unidimensional IRT model used elsewhere on this site — the [2PL fits
across cognitive datasets](2pl_across_datasets.qmd), the [IMV
comparisons](imv.qmd), the [Dutch Identity](dutch_identity.qmd) — rests on
the same starting assumption: a single latent trait explains the pattern of
item correlations. That assumption is a convenience as much as a claim
about the world. Real instruments often bundle related but distinct
sub-facets (a "self-regulation" battery with attention, inhibition, and
working-memory items), or are built explicitly as multi-factor scales (a
personality inventory with five named traits). Fitting a 1PL/2PL to
multidimensional data doesn't produce an error — it produces a single
composite ability estimate that quietly averages over whatever structure
was actually there. This vignette asks a simple empirical question: across
real IRW instruments, how often does the data actually look
one-dimensional, and does that vary by construct type?

We use two classic *exploratory* diagnostics, not a confirmatory test: the
ratio of the first to second eigenvalue of the item correlation matrix
(larger = more consistent with one dominant factor), and parallel analysis
[@horn1965rationale] — comparing the observed eigenvalues to those expected
from random data. Both flag *candidates* for multidimensionality; neither
tells you what the extra dimensions are or whether they're substantively
meaningful. The [CFA vignette](cfa.qmd) shows the confirmatory next step for
any table flagged here.

## Data and methods

### Dataset selection

```{r}
#| label: filter-datasets
#| eval: false

all_candidates <- irw_filter(
  n_items        = c(5, 40),
  n_participants = c(200, Inf)
)
```

We cap items at 40 (and require at least 5) for compute reasons: tetrachoric
and polychoric correlation matrices, and parallel analysis on top of them,
both scale poorly as item count grows, and this needs to run across every
eligible IRW table. `r r_all_candidates` tables met this criterion as of
`r format(date_run, "%B %Y")`.
`r if (is_pilot) "This page currently runs on a small pilot subset (see the warning above); a full pass analyzes all of them." else paste0("This pass analyzes all ", r_n_tables_run, " of them.")`

### Per-table computation

For each candidate table, `fit_dimensionality()`:

1. Fetches the table and builds a wide response matrix (`irw_fetch()` +
   `irw_long2resp()`), downsampling to at most 10,000 respondents and
   dropping zero-variance items, mirroring the [2PL across
   datasets](2pl_across_datasets.qmd) and [local
   dependence](local_dependence.qmd) pipelines.
2. Builds an item correlation matrix appropriate to the response type:
   **tetrachoric** (`psych::tetrachoric()`) for binary items, **polychoric**
   (`psych::polychoric()`) for 3-10 response categories, and a **Pearson**
   fallback (logged explicitly) for anything outside that range — mostly
   continuous slider items, which show up as tables with a large number of
   distinct response values.
3. Eigen-decomposes the correlation matrix and records the ratio of the
   1st to 2nd eigenvalue and the proportion of total variance the 1st
   eigenvalue explains.
4. Runs parallel analysis (`psych::fa.parallel()`, factor extraction, no
   plot) against a simulated random-data baseline, recording the suggested
   number of factors. Wrapped in `tryCatch` — this can fail on small or
   ill-conditioned matrices, in which case the table is logged and dropped
   rather than crashing the batch.
5. Records, per table: item/respondent counts, correlation method used,
   eigenvalue ratio, proportion of variance on factor 1, parallel-analysis
   suggested factor count, plus `construct_type` and `item_format` from
   IRW's tags for the breakdown below.

```{r}
#| label: fit-function
#| eval: false

fit_dimensionality <- function(table_name) {
  resp   <- fetch_wide(table_name)  # irw_fetch() + irw_long2resp(), cleaned
  cormat <- if (n_categories == 2) tetrachoric(resp)$rho
            else if (n_categories <= 10) polychoric(resp, max.cat = 10)$rho
            else cor(resp, use = "pairwise.complete.obs")
  ev <- eigen(cormat, symmetric = TRUE, only.values = TRUE)$values
  pa <- fa.parallel(cormat, n.obs = nrow(resp), fa = "fa", plot = FALSE)
  # ... ratio_12 <- ev[1] / ev[2]; suggested factors <- pa$nfact ...
}
```

Of the `r r_n_tables_run` candidate tables in this pass, `r nrow(summary_df)`
produced usable results (the rest were dropped for too few usable items or
a correlation/eigendecomposition that failed outright).

::: {.callout-note}
To reproduce with current IRW holdings, re-run `dimensionality_compute.R`
and commit the updated
`dimensionality_data/dimensionality_results.rds`.
:::

## Results

### How concentrated is the variance in one factor?

```{r}
#| label: fig-ratio-hist
#| fig-cap: "Ratio of the 1st to 2nd eigenvalue of the item correlation matrix, one table per bar. The dashed line marks a ratio of 4, near the top of the ~3-to-1-or-greater range sometimes read as a rough sign of approximate unidimensionality [@embretson2000item] -- treat it as a rule of thumb, not a hard rule."
#| fig-height: 4.5

ggplot(summary_df, aes(x = pmin(ratio_12, 15))) +
  geom_histogram(bins = 40, fill = irw_blue, colour = "white") +
  geom_vline(xintercept = ratio_cutoff, linetype = "dashed", colour = irw_red) +
  labs(x = "1st / 2nd eigenvalue ratio (capped at 15 for display)", y = "Number of tables")
```

```{r}
#| label: ratio-summary-inline
#| echo: false

n_above_cutoff <- sum(summary_df$ratio_12 > ratio_cutoff, na.rm = TRUE)
pct_above_cutoff <- round(100 * n_above_cutoff / nrow(summary_df))
n_pa_unidim <- sum(summary_df$unidimensional_pa, na.rm = TRUE)
pct_pa_unidim <- round(100 * n_pa_unidim / nrow(summary_df))
```

`r pct_above_cutoff`% of tables (`r n_above_cutoff` of `r nrow(summary_df)`)
clear the ratio > `r ratio_cutoff` heuristic in @fig-ratio-hist. Parallel
analysis is much stricter: only `r pct_pa_unidim`% of tables
(`r n_pa_unidim`) get a suggested factor count of exactly 1. That gap
between the two diagnostics is itself the most interesting finding here —
see [below](#two-diagnostics-one-disagreement).

### Does it vary by construct type?

```{r}
#| label: fig-ratio-by-construct
#| fig-cap: "Eigenvalue ratio by construct_type (primary label, for tables tagged with more than one). Each point is one table; diamonds mark the median within each type. Hover over a point to see which table it is."
#| fig-height: 5

construct_order <- summary_df %>%
  filter(!is.na(construct_type)) %>%
  group_by(construct_type) %>%
  summarise(med = median(ratio_12), .groups = "drop") %>%
  arrange(med) %>%
  pull(construct_type)

plot_df <- summary_df %>%
  filter(!is.na(construct_type)) %>%
  mutate(
    construct_type = factor(construct_type, levels = construct_order),
    tooltip = paste0(
      "Table: ", table,
      "<br>Ratio: ", round(ratio_12, 2),
      "<br>PA factors: ", nfact_suggested,
      "<br>n_items: ", n_items,
      "<br>Construct: ", construct_type
    )
  )

medians_df <- plot_df %>%
  group_by(construct_type) %>%
  summarise(med = median(ratio_12), .groups = "drop") %>%
  mutate(tooltip = paste0("Median (", construct_type, "): ", round(med, 2)))

p <- ggplot(plot_df, aes(x = pmin(ratio_12, 15), y = construct_type, text = tooltip)) +
  geom_jitter(height = 0.15, colour = irw_blue, alpha = 0.5, size = 1.7) +
  geom_point(data = medians_df, aes(x = pmin(med, 15), y = construct_type, text = tooltip),
             shape = 18, size = 4, colour = irw_red) +
  labs(x = "1st / 2nd eigenvalue ratio (capped at 15 for display)", y = NULL)

ggplotly(p, tooltip = "text")
```

```{r}
#| label: tbl-by-construct
#| tbl-cap: "Eigenvalue ratio and parallel-analysis factor counts by construct_type (primary label)."

summary_df %>%
  filter(!is.na(construct_type)) %>%
  group_by(construct_type) %>%
  summarise(
    n_tables            = n(),
    median_ratio         = median(ratio_12),
    prop_ratio_above_cut = mean(ratio_12 > ratio_cutoff),
    median_nfact         = median(nfact_suggested, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(median_ratio)) %>%
  knitr::kable(digits = 2, col.names = c("Construct type", "N tables", "Median ratio",
                                          "Prop. ratio > cutoff", "Median PA factors"))
```

Read @fig-ratio-by-construct and @tbl-by-construct together: constructs
built as a single narrow skill or trait tend to sit toward the high-ratio
end, while constructs assembled from several named sub-facets (a
multi-scale personality or psychopathology battery folded into one IRW
table) tend to sit lower. Neither diagnostic is a clean binary split —
there's substantial within-type spread — but the ranking across types is
informative on its own.

### Every table, ranked

```{r}
#| label: fig-ratio-ranked
#| fig-cap: "Every table's eigenvalue ratio, ranked from lowest to highest. Hover over a point to see which table it is."
#| fig-height: 6

ranked_df <- summary_df %>%
  arrange(ratio_12) %>%
  mutate(
    rank    = row_number(),
    tooltip = paste0(
      "Table: ", table,
      "<br>Ratio: ", round(ratio_12, 2),
      "<br>PA factors: ", nfact_suggested,
      "<br>n_items: ", n_items,
      "<br>Construct: ", construct_type
    )
  )

p <- ggplot(ranked_df, aes(x = pmin(ratio_12, 15), y = rank, text = tooltip)) +
  geom_point(colour = irw_blue, size = 1.6, alpha = 0.7) +
  geom_vline(xintercept = ratio_cutoff, linetype = "dashed", colour = irw_red) +
  labs(x = "1st / 2nd eigenvalue ratio (capped at 15 for display)", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggplotly(p, tooltip = "text") |>
  layout(margin = list(t = 60))
```

### Sanity check: does the diagnostic recover a known answer?

```{r}
#| label: fig-scree-sanity
#| fig-cap: "Scree plots for a known multi-factor personality battery (left, psychtools_bfi -- 25 items, 5 named traits) and a single-construct math test (right, 4thgrade_math_sirt), for comparison. The dashed line at eigenvalue = 1 is a visual reference (the classic Kaiser criterion), not the decision rule used elsewhere in this vignette."
#| fig-height: 5

scree_tables <- c("psychtools_bfi" = "psychtools_bfi (Big Five personality, 5 known factors)",
                   "4thgrade_math_sirt" = "4thgrade_math_sirt (single-construct math test)")

scree_df <- bind_rows(lapply(names(scree_tables), function(tbl) {
  ev <- eigenvalues_list[[tbl]]
  if (is.null(ev)) return(NULL)
  tibble(table = scree_tables[[tbl]], component = seq_along(ev), eigenvalue = ev)
}))

if (nrow(scree_df) > 0) {
  ggplot(scree_df, aes(x = component, y = eigenvalue)) +
    geom_hline(yintercept = 1, linetype = "dashed", colour = irw_grey) +
    geom_line(colour = irw_blue) +
    geom_point(colour = irw_blue) +
    facet_wrap(~table, scales = "free_x", labeller = label_wrap_gen(width = 22)) +
    labs(x = "Component", y = "Eigenvalue") +
    theme(strip.text = element_text(size = 10, lineheight = 1.1))
} else {
  cat("Sanity-check tables not present in this run.")
}
```

```{r}
#| label: sanity-inline
#| echo: false

bfi_row  <- summary_df %>% filter(table == "psychtools_bfi")
math_row <- summary_df %>% filter(table == "4thgrade_math_sirt")
```

`r if (nrow(bfi_row) > 0) paste0("The Big Five inventory -- 5 named traits by construction -- has a 1st/2nd eigenvalue ratio of ", round(bfi_row$ratio_12, 2), " and parallel analysis suggests ", bfi_row$nfact_suggested, " factors: both diagnostics correctly flag it as clearly multidimensional.") else "psychtools_bfi is not present in this run."` `r if (nrow(math_row) > 0) paste0("The math test has a ratio of ", round(math_row$ratio_12, 2), ", well above the personality battery's -- the diagnostic moves in the expected direction even though (see below) parallel analysis alone still suggests more than one factor for it.") else ""`

### Two diagnostics, one disagreement {#two-diagnostics-one-disagreement}

The eigenvalue ratio and parallel analysis agree on *direction* — the math
test above scores more unidimensional than the personality battery on both
— but they disagree sharply on the *absolute* call. Parallel analysis
suggests exactly one factor for only `r pct_pa_unidim`% of tables in this
sample, including tables like the math test above that most researchers
would treat as reasonably unidimensional in practice. This isn't a bug in
the pipeline: it's a documented property of parallel analysis run on
tetrachoric/polychoric correlation matrices. Garrido, Abad, and Ponsoda
[-@garrido2013new] show in a large simulation study that PA on polychoric
correlations systematically **over-extracts** factors relative to PA on
Pearson correlations, especially with few response categories, skewed
items, or smaller samples — exactly the conditions common across IRW
tables. The polychoric/tetrachoric correlation estimates themselves carry
sampling noise that inflates later eigenvalues, and parallel analysis's
random-data baseline doesn't fully correct for it.

Practically: treat the eigenvalue ratio in @fig-ratio-hist as the more
robust of the two signals for *ranking* tables against each other, and
treat parallel analysis's suggested factor count as a noisy, generally
inflated upper bound rather than a literal count of a table's dimensions.
A table is worth a closer confirmatory look (with the [CFA
vignette](cfa.qmd)) when *both* diagnostics point the same direction —
low ratio and a PA count well above 1 — rather than off either one alone.

## Limitations

**Exploratory, not confirmatory.** Neither diagnostic tells you what the
extra dimensions are, whether they're substantively meaningful, or whether
a bifactor/testlet structure (rather than several independent traits)
better describes the data. The [CFA vignette](cfa.qmd) is the natural next
step for any table flagged here.

**Parallel analysis over-extracts on ordinal data**, as discussed above
[@garrido2013new] — don't read the suggested factor count as a literal
answer to "how many dimensions does this table have."

**A single ratio collapses a lot of structure.** Two tables with the same
1st/2nd eigenvalue ratio can have very different shapes underneath — one
might have one dominant factor and many small residual ones, another two
comparably-sized factors and nothing else. The full eigenvalue spectrum
(as in the scree plots above) carries more information than the ratio
alone; the ratio is used here as a single sortable number because it needs
to summarize `r nrow(summary_df)` tables in one figure.

**Local dependence and low dimensionality aren't the same thing.** A table
can show local item dependence (testlets, shared stems) without failing a
global dimensionality check, and vice versa. See the [local dependence
vignette](local_dependence.qmd) for the complementary residual-correlation
diagnostic.

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the compute script (from the project root)
source("vignettes/dimensionality_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/dimensionality.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(summary_df$table, output_file = "dimensionality_data/references.bib")
```
:::

---

## References
