---
title: "The Dutch Identity: Manifest Probabilities and IRT Models"
description: |
  A tour of Holland's (1990) Dutch Identity — the result that log-manifest
  probabilities of dichotomous response patterns are approximately log-quadratic
  in the item responses. We ground each theoretical claim in IRW data: enumerating
  manifest probabilities directly, fitting loglinear models, diagnosing the Rasch
  model via total-score frequencies, assessing dimensionality from second-order
  interactions, and testing Holland's conservation law for model complexity.
date: today
bibliography: dutchdata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "dutchdata/dutch_identity_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache   <- readRDS(cache_path)
  s1_3     <- .cache$s1_3
  s4       <- .cache$s4
  s5       <- .cache$s5
  s6       <- .cache$s6
  date_run <- .cache$date_run
} else {
  # Placeholder objects so inline R expressions in prose don't crash
  na_dataset <- list(table_name = "—", J = NA_integer_, N = NA_integer_,
                     ll_diff = NA_real_, r2_quad = NA_real_,
                     gamma_k_data = NULL, cor_eigval = numeric(),
                     soe_eigval = NULL)
  s1_3 <- c(na_dataset, list(patterns = NULL, gamma_matrix = NULL,
                              eigenvalues = NULL, bic_table = NULL))
  s4   <- list(good_rasch = na_dataset, poor_rasch = na_dataset,
               rasch_meta = tibble(table_name = character()))
  s5   <- list(unidim = na_dataset, multidim = na_dataset)
  s6   <- tibble(table_name = character(), J = integer(),
                 imv_1v2 = numeric(), imv_2v3 = numeric())
  date_run <- Sys.Date()
}

# Disable all subsequent code chunks when cache is missing
knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures will not appear. ",
  "Generate results by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/dutch_identity_compute.R\n```\n",
  ":::\n", sep = ""
)
```

## Background

@holland1990dutch introduced the **Dutch Identity** — a result about the
mathematical structure that any unidimensional IRT model imposes on the joint
distribution of dichotomous item responses. The key insight is that the
*manifest* (unconditional) probability of a response pattern $\mathbf{x} =
(x_1, \ldots, x_J) \in \{0,1\}^J$ can be written as an exponential family
whose sufficient statistics are functions of the item responses. For large tests
with smooth item response functions (IRFs), this reduces to a strikingly simple
approximation:

$$\log p(\mathbf{x}) \approx \alpha + \sum_{j=1}^{J} \beta_j x_j + \frac{1}{2}\!\left(\sum_{j=1}^{J} \gamma_j x_j\right)^{\!2}$$

This is log-quadratic in the $x_j$, requires only $2J + 1$ parameters regardless
of $J$, and makes precise predictions about the structure of second-order
interactions between items. The sections below work through the implications one
step at a time, using IRW datasets to make each idea concrete.

::: {.callout-note}
All figures use pre-computed results from `dutch_identity_compute.R`. The
dataset used in Sections 1–3 is **`r s1_3$table_name`**
($J = `r s1_3$J`$ items, $N = `r format(s1_3$N, big.mark=",")`$ respondents,
computed `r format(date_run, "%B %Y")`).
:::

---

## Section 1: The manifest probabilities directly

For a test of $J$ binary items, a respondent's answers define a response pattern
$\mathbf{x} \in \{0,1\}^J$. There are $2^J$ possible patterns, each occurring
with manifest probability $p(\mathbf{x})$. Before modeling, it is worth asking:
what do these probabilities actually look like? Most patterns are extremely rare;
a handful dominate. @holland1990dutch motivates the loglinear approach precisely
because the $p(\mathbf{x})$ are too numerous and sparse to model directly for
large $J$.

For $J \leq 10$ items and $N \geq 2{,}000$ respondents the full distribution is
enumerable. Below we tabulate every $2^J$ pattern and display the empirical
$p(\mathbf{x})$.

```{r}
#| label: fig-manifest-dist
#| fig-cap: "Distribution of empirical manifest probabilities across all 2^J response patterns (log scale). Most patterns are never observed (grey); a few common patterns carry most of the probability mass."
#| fig-height: 4

patterns <- s1_3$patterns |>
  arrange(desc(count)) |>
  mutate(
    rank        = row_number(),
    p_obs       = count / s1_3$N,
    observed    = count > 0,
    pattern_str = apply(pick(all_of(s1_3$item_cols)), 1, paste, collapse = "")
  )

n_observed   <- sum(patterns$observed)
n_unobserved <- sum(!patterns$observed)

ggplot(patterns, aes(x = rank, y = p_obs, colour = observed)) +
  geom_point(size = 1.2, alpha = 0.7) +
  scale_colour_manual(
    values = c("TRUE" = irw_blue, "FALSE" = irw_grey),
    labels = c("TRUE" = "Observed", "FALSE" = "Never seen"),
    name   = NULL
  ) +
  scale_y_log10(labels = scales::label_scientific()) +
  labs(
    x       = "Pattern rank (most to least common)",
    y       = "Empirical p(x)  [log scale]",
    caption = paste0(
      n_observed, " of ", 2^s1_3$J, " patterns observed; ",
      n_unobserved, " never appeared."
    )
  )
```

```{r}
#| label: fig-score-dist
#| fig-cap: "Total probability mass at each sum score k = Σ xⱼ. Patterns within each score group contribute additively."
#| fig-height: 4

patterns |>
  filter(observed) |>
  ggplot(aes(x = factor(total), y = p_obs, fill = p_obs)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  scale_fill_gradient(low = "#d1e5f0", high = irw_blue) +
  labs(
    x       = "Total score k",
    y       = "p(x)",
    caption = "Each bar is the total probability mass at that score level."
  )
```

**What to notice.** Probability mass is extremely concentrated: the most common
pattern can account for a large share of the total. The long tail of near-zero
patterns means a $2^J$-parameter saturated model is hopeless — most cells have
zero or one observation. This motivates the loglinear parameterisation in
Section 2.

---

## Section 2: Log-manifest probabilities and the independence baseline

Under item *independence* — no latent variable, no associations — $\log
p(\mathbf{x})$ is a linear function of the item responses:

$$\log p(\mathbf{x}) = \alpha + \sum_{j=1}^{J} \beta_j x_j$$

This **independence loglinear model** has $J + 1$ parameters. It is equivalent
to a Poisson regression of pattern counts on the $J$ binary item predictors.
When it fits, all structure in the data is captured by item marginals alone.

Any IRT model implies that $\log p(\mathbf{x})$ *deviates* from this baseline —
the latent variable induces positive associations among items that the
main-effects model misses. Plotting observed versus fitted $\log p(\mathbf{x})$
makes those deviations visible.

```{r}
#| label: fig-independence-fit
#| fig-cap: "Observed vs. fitted log p(x) under the independence loglinear model. Points on the diagonal are fully explained by item marginals; points below the line are more probable than independence predicts (positive local dependence)."
#| fig-height: 4.5

plot_dat <- patterns |>
  filter(observed) |>
  mutate(residual = log_p_obs - log_p_ind)

r2_ind <- cor(plot_dat$log_p_obs, plot_dat$log_p_ind, use = "complete.obs")^2

ggplot(plot_dat, aes(x = log_p_ind, y = log_p_obs)) +
  geom_abline(slope = 1, intercept = 0, colour = irw_grey, linetype = "dashed") +
  geom_point(aes(size = p_obs), colour = irw_blue, alpha = 0.6) +
  scale_size_continuous(range = c(1, 5), name = "p(x)", guide = "none") +
  annotate(
    "text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4,
    label = paste0("R² = ", round(r2_ind, 3)),
    size = 4, colour = "grey30"
  ) +
  labs(
    x = "Fitted log p(x)  [independence model]",
    y = "Observed log p(x)"
  )
```

**What to notice.** Systematic curvature away from the diagonal — especially
for extreme total-score patterns — signals that main effects alone are not
enough. The points that deviate most are those where all or nearly all items are
answered the same way: item correlation inflates these relative to the
independence prediction. This is exactly the second-order structure that
@holland1990dutch formalises.

---

## Section 3: Fitting the second-order exponential model

The full **second-order exponential (SOE) model** augments the independence
model with all $\binom{J}{2}$ pairwise interaction terms:

$$\log p(\mathbf{x}) = \alpha + \sum_j \beta_j x_j + \sum_{j < k} \gamma_{jk}\, x_j x_k$$

This is still a Poisson GLM, now with $1 + J + \binom{J}{2}$ parameters. The
$J \times J$ matrix $\boldsymbol{\Gamma}$ (with $\Gamma_{jk} = \gamma_{jk}$ and
$\Gamma_{jj} = 0$) encodes all pairwise item associations after removing main
effects.

The Dutch Identity [@holland1990dutch, Theorem 1] predicts that
$\boldsymbol{\Gamma}$ is *approximately rank-1*: $\Gamma_{jk} \approx \gamma_j
\gamma_k$. Under this constraint, the interaction structure collapses to:

$$\log p(\mathbf{x}) \approx \alpha + \sum_j \beta_j' x_j + \frac{1}{2}\!\left(\sum_j \gamma_j x_j\right)^{\!2}$$

the **restricted log-quadratic** model ($1 + 2J$ parameters). We fit all three
models and compare.

### Model comparison

```{r}
#| label: tbl-bic
#| tbl-cap: "Log-likelihood and BIC for three nested loglinear models. The restricted log-quadratic uses 2J + 1 parameters; the full SOE uses 1 + J + J(J-1)/2."

s1_3$bic_table |>
  mutate(
    across(c(Log_lik, BIC), \(x) round(x, 1)),
    BIC_delta = round(BIC - min(BIC), 1)
  ) |>
  rename(`ΔBIC vs. best` = BIC_delta, `Log-lik` = Log_lik) |>
  knitr::kable(align = "lrrr")
```

### Structure of the Γ matrix

```{r}
#| label: fig-gamma-heatmap
#| fig-cap: "Heatmap of the J×J interaction matrix Γ from the full SOE model. Under the Dutch Identity, off-diagonal entries should follow a rank-1 pattern γⱼγₖ — all rows should look like rescaled versions of each other."
#| fig-height: 4

gmat      <- s1_3$gamma_matrix
gmat_long <- as.data.frame(as.table(gmat)) |>
  rename(item1 = Var1, item2 = Var2, gamma = Freq) |>
  filter(item1 != item2)

ggplot(gmat_long, aes(x = item2, y = item1, fill = gamma)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(
    low = irw_red, mid = "white", high = irw_blue,
    midpoint = 0, name = "γⱼₖ"
  ) +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid  = element_blank()
  )
```

### Eigenvalue decay of Γ

```{r}
#| label: fig-gamma-scree
#| fig-cap: "Eigenvalue scree plot of the SOE interaction matrix Γ. The Dutch Identity predicts a single dominant eigenvalue (Γ is approximately rank-1)."
#| fig-height: 3.5

eig_df <- tibble(
  component  = seq_along(s1_3$eigenvalues),
  eigenvalue = s1_3$eigenvalues,
  pct_var    = abs(s1_3$eigenvalues) / sum(abs(s1_3$eigenvalues))
)

ggplot(eig_df, aes(x = component, y = eigenvalue)) +
  geom_col(fill = irw_blue, alpha = 0.7, width = 0.6) +
  geom_hline(yintercept = 0, colour = irw_grey) +
  scale_x_continuous(breaks = seq_len(s1_3$J)) +
  labs(
    x       = "Component",
    y       = "Eigenvalue of Γ",
    caption = paste0(
      "Leading eigenvalue explains ",
      round(eig_df$pct_var[1] * 100, 1),
      "% of |eigenvalue| sum."
    )
  )
```

**What to notice.** If the Dutch Identity holds, the restricted log-quadratic
model ($1 + 2J$ parameters) should close most of the gap between the independence
model and the full SOE ($1 + J + \binom{J}{2}$ parameters), despite using far
fewer parameters. The eigenvalue plot shows how nearly rank-1 $\boldsymbol{\Gamma}$
is: one dominant positive eigenvalue with the rest near zero is the hallmark of
a unidimensional instrument.

---

## Section 4: The γₖ diagnostic for the Rasch model

@holland1990dutch Corollary 2 gives a striking prediction about the Rasch model.
If ability $\theta \sim N(\mu, \sigma^2)$, the probability of obtaining total
score $k$ is:

$$\log \pi_k \approx C + \mu(k - \bar{y}_+) + \frac{\sigma^2}{2}(k - \bar{y}_+)^2$$

where $\bar{y}_+ = \sum_j \bar{p}_j$ is the expected total score at the mean
ability. In other words, $\log \pi_k$ is a *quadratic* function of $k$, centred
on $\bar{y}_+$. Systematic deviations indicate either non-normality of the
ability distribution or Rasch misfit.

We compare two IRW datasets: one where the Rasch model fits well
(**`r s4$good_rasch$table_name`**, $J = `r s4$good_rasch$J`$,
$N = `r format(s4$good_rasch$N, big.mark=",")`$,
$\Delta\ell = `r round(s4$good_rasch$ll_diff, 1)`$ 2PL vs. Rasch),
and one where it does not
(**`r s4$poor_rasch$table_name`**, $J = `r s4$poor_rasch$J`$,
$N = `r format(s4$poor_rasch$N, big.mark=",")`$,
$\Delta\ell = `r round(s4$poor_rasch$ll_diff, 1)`$).

```{r}
#| label: plot-gk-fn
#| echo: false

plot_gk <- function(gk_data) {
  ggplot(gk_data, aes(x = k, y = gamma_k)) +
    geom_point(colour = irw_blue, size = 2.5) +
    geom_line(aes(y = gamma_k_fitted), colour = irw_red,
              linewidth = 0.9) +
    labs(
      x       = "Total score k",
      y       = expression(log~pi[k]),
      caption = "Red curve: fitted quadratic."
    )
}
```

```{r}
#| label: fig-gamma-k-good
#| fig-cap: "Total-score log-frequencies for the dataset where Rasch fits well. Points follow a smooth quadratic, consistent with Holland's Corollary 2."
#| fig-height: 4

plot_gk(s4$good_rasch$gamma_k_data)
```

The quadratic $R^2$ is `r round(s4$good_rasch$r2_quad, 3)` for this dataset.

```{r}
#| label: fig-gamma-k-poor
#| fig-cap: "Total-score log-frequencies for the dataset where Rasch fits poorly. Deviations from the quadratic suggest non-constant discrimination or a non-normal ability distribution."
#| fig-height: 4

plot_gk(s4$poor_rasch$gamma_k_data)
```

The quadratic $R^2$ is `r round(s4$poor_rasch$r2_quad, 3)` for this dataset.

**What to notice.** When Rasch holds, $\log \pi_k$ traces a clean inverted
parabola. Deviations can take two forms: (1) a sharp peak at the mode with thin
tails, suggesting a more leptokurtic ability distribution; or (2) asymmetry or a
kink, consistent with varying item discriminations. The $R^2$ from the quadratic
regression is a quick, model-free diagnostic: values above 0.99 are consistent
with Rasch; lower values flag something worth investigating.

---

## Section 5: Dimensionality via second-order interaction matrices

Holland ([-@holland1990dutch], Section 3.3) shows that for a $D$-dimensional IRT
model, the matrix $\boldsymbol{\Gamma}$ of pairwise SOE interactions has rank at
most $D$. This gives a direct route to dimensionality assessment: fit the SOE
loglinear model, extract $\boldsymbol{\Gamma}$, and read off the number of
non-negligible eigenvalues. We compare this to the standard approach of examining
eigenvalues of the inter-item correlation matrix.

We use **`r s5$unidim$table_name`** ($J = `r s5$unidim$J`$) as the
unidimensional candidate and **`r s5$multidim$table_name`**
($J = `r s5$multidim$J`$) as the more complex one.

```{r}
#| label: fig-scree-comparison
#| fig-cap: "Scree plots of inter-item correlation matrix eigenvalues. A sharply dominant first eigenvalue indicates unidimensionality; multiple eigenvalues above 1 indicate multidimensionality."
#| fig-height: 5.5

make_scree_df <- function(dim_result, label) {
  tibble(
    label      = label,
    component  = seq_along(dim_result$cor_eigval),
    eigenvalue = dim_result$cor_eigval,
    pct_var    = dim_result$cor_eigval / sum(dim_result$cor_eigval)
  )
}

scree_df <- bind_rows(
  make_scree_df(
    s5$unidim,
    paste0("Unidimensional: ", s5$unidim$table_name, " (J=", s5$unidim$J, ")")
  ),
  make_scree_df(
    s5$multidim,
    paste0("Complex: ", s5$multidim$table_name, " (J=", s5$multidim$J, ")")
  )
) |> filter(component <= 10)

ggplot(scree_df, aes(x = component, y = eigenvalue)) +
  geom_col(fill = irw_blue, alpha = 0.7, width = 0.6) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = irw_grey) +
  scale_x_continuous(breaks = 1:10) +
  facet_wrap(~label, ncol = 1, scales = "free_y") +
  labs(
    x       = "Component",
    y       = "Eigenvalue",
    caption = "Dashed line at 1 (Kaiser criterion)."
  )
```

```{r}
#| label: fig-gamma-scree-dim
#| fig-cap: "Eigenvalue scree plot of the SOE interaction matrix Γ for the unidimensional dataset. A single dominant eigenvalue is consistent with D = 1."
#| fig-height: 3.5

if (!is.null(s5$unidim$soe_eigval)) {
  tibble(
    component  = seq_along(s5$unidim$soe_eigval),
    eigenvalue = s5$unidim$soe_eigval
  ) |>
    ggplot(aes(x = component, y = eigenvalue)) +
    geom_col(fill = irw_blue, alpha = 0.7, width = 0.6) +
    geom_hline(yintercept = 0, colour = irw_grey) +
    scale_x_continuous(breaks = seq_len(s5$unidim$J)) +
    labs(x = "Component", y = "Eigenvalue of Γ  (SOE model)")
} else {
  message("SOE eigenvalues not available for this dataset (J too large for pattern enumeration).")
}
```

**What to notice.** A clearly unidimensional instrument shows one eigenvalue
well above the rest, with a sharp elbow. A multidimensional instrument shows
multiple eigenvalues above 1, with a more gradual decline. The SOE
$\boldsymbol{\Gamma}$ scree plot makes a stronger theoretical claim: the number
of non-trivial eigenvalues *equals* the number of latent dimensions, not just the
number that exceed an arbitrary threshold.

---

## Section 6: How many parameters does a long test support?

@holland1990dutch establishes a **conservation law**: a $D$-dimensional test of
$J$ items supports at most $(D+1)J$ parameters as $J \to \infty$. For $D = 1$
this means at most $2J$ parameters — exactly what the 2PL provides. The 3PL has
$3J$ parameters, exceeding the limit. Holland's prediction is therefore that
**the marginal gain from 2PL → 3PL should shrink as $J$ grows**, and may
become negative (overfitting) for long tests.

We test this using the InterModel Vigorish [IMV; see also @holland1990dutch] — a
log-score measure of predictive improvement — across IRW datasets spanning a
range of test lengths.

```{r}
#| label: fig-conservation
#| fig-cap: "IMV gain from 1PL → 2PL (blue) and 2PL → 3PL (red) as a function of test length J. Holland's conservation law predicts the 2PL → 3PL gain shrinks toward zero as J grows."
#| fig-height: 4.5

cons_df <- s6 |>
  filter(!is.na(imv_1v2)) |>
  pivot_longer(
    cols      = c(imv_1v2, imv_2v3),
    names_to  = "comparison",
    values_to = "imv"
  ) |>
  filter(!is.na(imv)) |>
  mutate(
    comparison = recode(
      comparison,
      imv_1v2 = "1PL → 2PL  (free discrimination)",
      imv_2v3 = "2PL → 3PL  (add guessing)"
    )
  )

ggplot(cons_df, aes(x = J, y = imv, colour = comparison)) +
  geom_jitter(width = 0.3, height = 0, alpha = 0.6, size = 2) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.15, linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = irw_grey) +
  scale_colour_manual(
    values = c(
      "1PL → 2PL  (free discrimination)" = irw_blue,
      "2PL → 3PL  (add guessing)"        = irw_red
    ),
    name = NULL
  ) +
  labs(
    x       = "Number of items (J)",
    y       = "IMV gain",
    caption = paste0(
      "Based on ", n_distinct(s6$table_name[!is.na(s6$imv_1v2)]),
      " IRW datasets. 3PL fitted for J ≤ 20 only."
    )
  ) +
  theme(legend.position = "bottom")
```

**What to notice.** The 1PL → 2PL gain (blue) is typically positive and stable
across $J$: freeing discrimination improves fit regardless of test length,
consistent with real variation in item discriminations. The 2PL → 3PL gain (red)
should, if Holland is right, trend toward zero as $J$ grows. If it remains
positive at large $J$, that suggests genuine multidimensionality or pervasive
guessing. The pattern across the IRW is the empirical verdict.

---

::: {.callout-warning}
## Limitations

**Pattern sparsity.** Sections 1–3 require $J \leq 10$ for pattern enumeration
to be feasible. The SOE loglinear approach does not generalise to large $J$.

**Normal ability assumption.** The quadratic $\log \pi_k$ result (Section 4)
holds asymptotically under normality; skewed or multimodal ability distributions
can mimic the pattern for the wrong reasons.

**3PL instability.** The 3PL is difficult to estimate for short tests and small
samples. IMV values for 2PL → 3PL should be interpreted cautiously for $J < 10$.

**Dataset selection.** `irw_filter()` conditions are heuristic. The
"unidimensional" and "complex" candidates in Section 5 were selected on
eigenvalue structure alone. Users with substantive knowledge of specific IRW
tables should override the automatic selection.
:::

---

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the compute script (from the project root)
source("vignettes/dutch_identity_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/dutch_identity.qmd")
```

::: {.callout-tip}
To cite the IRW datasets used, run:

```r
for (tbl in unique(c(s1_3$table_name, s4$rasch_meta$table_name,
                      s5$unidim$table_name, s5$multidim$table_name,
                      s6$table_name))) {
  irw_save_bibtex(tbl, output_file = "vignettes/dutchdata/irw_references.bib",
                  append = TRUE)
}
```
:::

---

## References
