---
title: "What Is a Floor, Actually?"
description: |
  Experience-sampling responses pile up at the bottom of the scale, and a recent
  preprint argues that ignoring that pile biases the person-level correlations
  that dynamic structural equation models are fitted to estimate. Its remedy is a
  censored model. But censoring is only one theory of what a floor is — an
  appropriate link models the boundedness directly, and a graded model says the
  floor was never special to begin with. This vignette measures how much floor
  mass IRW's intensive longitudinal datasets actually carry, shows that the
  composite scores the field models hide almost all of it, and puts the competing
  models against each other on a study that randomised its participants to a
  Likert or a slider version of the same protocol.
date: today
bibliography: esm_floor_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)
library(knitr)

theme_set(theme_minimal(base_size = 13))
irw_blue  <- "#2166ac"
irw_red   <- "#d73027"
irw_amber <- "#b8791f"
irw_grey  <- "grey50"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "esm_floor_data/esm_floor_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  floor_summary    <- .cache$floor_summary
  item_detail      <- .cache$item_detail
  stage_b          <- .cache$stage_b %||% tibble()
  candidate_tables <- .cache$candidate_tables
  esm_slider       <- .cache$esm_slider
  esm_ordinal      <- .cache$esm_ordinal
  date_run         <- .cache$date_run
  is_pilot         <- isTRUE(.cache$pilot)
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_tables_run   <- nrow(floor_summary)
} else {
  floor_summary <- tibble(
    table = character(), n_items = integer(), n_persons = integer(),
    n_responses = integer(), occasion_col = character(),
    floor_min = numeric(), floor_max = numeric(), floor_median = numeric(),
    composite_floor = numeric(), format = character()
  )
  item_detail <- tibble(
    table = character(), item = character(), n = integer(),
    lo = numeric(), hi = numeric(), n_cat = integer(),
    floor_pct = numeric(), ceil_pct = numeric()
  )
  stage_b <- tibble(
    item = character(), arm = character(), family = character(),
    n_obs = integer(), n_persons = integer(), floor_pct = numeric(),
    phi_ar1 = numeric(), phi_ar1_lo = numeric(), phi_ar1_hi = numeric(),
    cor_mean_disp = numeric(), loo_elpd = numeric(),
    rhat_ar = numeric(), ess_ar = numeric(), n_divergent = numeric(),
    converged = logical()
  )
  candidate_tables <- character(); esm_slider <- character(); esm_ordinal <- character()
  date_run <- Sys.Date(); is_pilot <- FALSE
  r_all_candidates <- 0L; r_n_tables_run <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis
if (!cache_ok) {
  cat("::: {.callout-warning}\n## Results cache not found\n\n",
      "This page renders from `", cache_path, "`, which is not present. ",
      "Run `Rscript vignettes/esm_floor_compute.R` from the repository root ",
      "(with `REDIVIS_API_TOKEN` set) to generate it. Every results chunk below ",
      "is skipped until then.\n:::\n", sep = "")
}
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: !expr cache_ok
#| output: asis
if (is_pilot) {
  cat("::: {.callout-note}\n## Partial run\n\n",
      "This page was built from a **pilot** run: ", r_n_tables_run, " of ",
      r_all_candidates, " candidate tables profiled, and the model comparison ",
      "restricted to a few headline items. Numbers will shift when the full batch ",
      "completes.\n:::\n", sep = "")
}
```

::: {.callout-note appearance="simple"}
This vignette was drafted by Claude with human review. The scoping decisions —
which datasets qualify as experience sampling, and which models are being
compared — came out of a written brief circulated to IRW collaborators, and the
model set below was revised in response to their comments.
:::

## Motivation

Ask people how they feel several times a day for a fortnight and a lot of their
answers land on the bottom of the scale. Not anxious. Not stressed. Not at all.
That pile is a nuisance for the models the field fits to such data: dynamic
structural equation models (DSEM) treat each person's series as a continuous
process with a mean, an autoregression, and an innovation variance, and then
estimate how those person-level quantities correlate.

A recent preprint — *Beneath the Floor: Censored DSEM Models for Analyzing ESM
Data with Floor Effects* [@beneaththefloor2026] — argues that ignoring the pile
biases exactly those correlations, most sharply the one between a person's mean
level and their innovation variance, and recommends modelling the censoring
explicitly.

That recommendation carries an assumption worth naming. **A censored model says
there is a latent continuum extending below the boundary that we simply cannot
observe** — the person is somewhere below "not at all anxious", and the scale
truncated them. That is a specific and contestable claim about what a floor
*is*, and it is not the only one available:

- **Censoring.** A continuum runs past the boundary; the floor is truncation.
  Someone at the floor could have been lower.
- **Zero-inflation.** Sitting *at* the floor is a different kind of event from
  sitting just above it. "Not at all" is a state, not a smaller amount of
  something. Under an inflated beta model the boundary gets its own probability
  and the interior gets a bounded continuous density.
- **Graded response.** The answer was categorical the whole time. The floor is
  the bottom category, and nothing about it needs special handling.

These are not three ways of patching the same nuisance. They are three theories
of the data, they imply different generative stories, and — usefully — they make
different predictions about the very correlation the preprint reports.

IRW is a reasonable place to adjudicate because it holds intensive longitudinal
data in both ordinal and near-continuous response formats, in one schema. The
two vignettes closest to this one are
[Bounded Continuous Responses](./continuous_bounded.qmd), which compares
bounded-response model families cross-sectionally, and
[Dimensionality Across IRW](./dimensionality.qmd), which is the template for an
assumption audit run across many tables at once. This page is the time-domain
sibling of the first.

That earlier vignette also left this exact question open. It stress-tested four
bounded-response specifications against 20% boundary inflation and concluded that
all of them degrade, because *none* of them includes a zero-one-inflated
component — "a true zero-one-inflated extension would be needed for that"
[@molenaar2022zero]. It named the missing model and stopped. This page fits it.

## Data and methods

### Finding the datasets at all

IRW has no reliable handle for intensive longitudinal data, which is the first
finding and the reason the pool below had to be built by hand.

- `metadata$longitudinal` flags 703 tables, but it is a text search for `wave` or
  `date` in the column list, so it matches `cov_birthdate` and `cov_startdate`.
- The `intensive_longitudinal` collection requires `date` **and** `wave`
  together. Of its 81 members, 72 are one COVID panel study at about 2.6 waves
  per person, and exactly one is experience sampling. Its own definition says as
  much.
- There is no temporal tag. The tag vocabulary covers age, sample, construct,
  tool, format and language — nothing about design over time.

Across 4,238 tables, 551 carry `wave` and 224 carry `date`, but only 81 carry
both, so requiring both excludes most genuine ESM studies while admitting a large
panel study.

What works is **occasions per person** — responses per participant divided by
item count — followed by reading each surviving table's processing script. That
screen produced the pool used here.

```{r}
#| label: tbl-pool
#| echo: true
if (nrow(floor_summary) > 0) {
  floor_summary |>
    mutate(`occ/person` = round(n_responses / n_persons / n_items, 1)) |>
    select(Table = table, Format = format, Persons = n_persons, Items = n_items,
           `occ/person`, `Time index` = occasion_col) |>
    arrange(desc(`occ/person`)) |>
    kable(caption = "The curated ESM pool, ordered by occasions per person.")
}
```

Confirming each table by script mattered: four candidates that pass any
column-based screen fail on inspection. `narcissism_schneider_2025_study2_*`
(18 tables) falls back to a survey disposition code as `id`, so `id` is not a
person and the recorded participant count is 3.
`moralvignettes_rakhmankulova_2025` has `id` as a video and `rater` as the
person. `political_psychology` drops `wave` in processing. And about 21
cognitive trial-level tasks have many measurements per person, but those are
trials, not occasions.

Two more were caught only after they had been profiled, which is worth
recording because it shows what the screen alone cannot do.
`emoji_scheffler_2024` (`id` is the emoji, `rater` the person) and `tears`
(`id` is a video, `rater` the person) are stimulus-rating studies: many
responses per `id`, no time dimension at all. They scored well on
occasions-per-person precisely *because* many raters judged each stimulus. Both
were dropped.

### The floor is an item-level property

The field reports floor rates for a composite — the mean of the items
administered at each beep. Because items within one instrument sit at very
different points on the scale, that average can drive the composite's floor mass
to almost nothing while individual items remain heavily piled.

It does, and by a lot. The gap is not uniform, though, and the pattern is
informative: **compositing hides the floor precisely when the items disagree
about where the floor is.** Where every item in an instrument is heavily floored,
the composite stays floored and nothing is concealed. Where one item floors and
its neighbours do not, the average washes it out entirely.

```{r}
#| label: fig-item-vs-composite
#| echo: true
if (nrow(floor_summary) > 0) {
  p <- floor_summary |>
    filter(!is.na(composite_floor)) |>
    mutate(text = sprintf("%s\nitems: %.1f%%–%.1f%% at floor\ncomposite: %.3f%%",
                          table, floor_min, floor_max, composite_floor)) |>
    ggplot(aes(x = floor_max, y = composite_floor, colour = format, text = text)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = irw_grey) +
    geom_point(size = 2.6, alpha = .85) +
    scale_colour_manual(values = c("Ordinal" = irw_amber,
                                   "Slider / continuous" = irw_blue), name = NULL) +
    labs(x = "Highest per-item floor mass in the table (%)",
         y = "Floor mass of the composite (%)",
         title = "Averaging items hides the floor",
         subtitle = "Points below the dashed line: the composite understates its own worst item") +
    theme(legend.position = "top")
  ggplotly(p, tooltip = "text")
}
```

```{r}
#| label: hiding-readout
#| echo: false
#| output: asis
if (nrow(floor_summary) > 0) {
  fs <- floor_summary |> filter(!is.na(composite_floor))
  n_half <- sum(fs$composite_floor < fs$floor_max / 2)
  n_hid  <- sum(fs$composite_floor < 1 & fs$floor_max > 20)
  worst  <- fs |> mutate(ratio = floor_max / pmax(composite_floor, .001)) |>
    arrange(desc(ratio)) |> slice(1)
  cat("In **", n_half, " of ", nrow(fs), "** tables the composite's floor mass is ",
      "less than half that of the table's most-floored item. In **", n_hid,
      "** of them the composite sits below 1% while an item is above 20%. ",
      "The extreme case is `", worst$table, "`, where an item at ",
      sprintf("%.1f%%", worst$floor_max), " floor becomes a composite at ",
      sprintf("%.3f%%", worst$composite_floor), " — a factor of roughly ",
      format(round(worst$ratio), big.mark = ","), ".\n\n", sep = "")
}
```

```{r}
#| label: tbl-hiding
#| echo: true
if (nrow(floor_summary) > 0) {
  floor_summary |>
    filter(!is.na(composite_floor)) |>
    mutate(hiding = floor_max - composite_floor) |>
    arrange(desc(hiding)) |>
    transmute(Table = table, Format = format,
              `Worst item (%)` = round(floor_max, 1),
              `Median item (%)` = round(floor_median, 1),
              `Composite (%)`   = round(composite_floor, 3),
              `Hidden (pp)`     = round(hiding, 1)) |>
    head(12) |>
    kable(caption = "Where compositing conceals the most floor mass.")
}
```

```{r}
#| label: fig-heterogeneity
#| echo: true
if (nrow(floor_summary) > 0) {
  p <- floor_summary |>
    filter(!is.na(composite_floor)) |>
    mutate(spread = floor_max - floor_min,
           ratio  = floor_max / pmax(composite_floor, .001),
           text = sprintf("%s\nitem floor spread: %.1f pp\nworst item %.1f%% -> composite %.3f%%",
                          table, spread, floor_max, composite_floor)) |>
    ggplot(aes(x = spread, y = ratio, colour = format, text = text)) +
    geom_point(size = 2.6, alpha = .85) +
    scale_y_log10() +
    scale_colour_manual(values = c("Ordinal" = irw_amber,
                                   "Slider / continuous" = irw_blue), name = NULL) +
    labs(x = "Spread of per-item floor mass within the table (percentage points)",
         y = "Worst item floor / composite floor (log scale)",
         title = "Compositing hides the floor when items disagree about where it is",
         subtitle = "Instruments whose items all floor together conceal nothing") +
    theme(legend.position = "top")
  ggplotly(p, tooltip = "text")
}
```

`sondell_2018_dementia_motivation` is absent from the composite comparisons for
a different and more interesting reason: it has a perfectly good `wave`, but
across all 5,312 person-occasions **only ever one item is answered at a time**.
Its two items were never co-administered, so there is no occasion at which a
composite could be formed. It stays in the per-item figures.

### The model comparison, and why it runs on this table

`vollbracht_et_al_2026_ambulatory_assessment` [@vollbracht2026] randomised 406
participants to complete the same ambulatory protocol with either a Likert or a
slider response format, then measured them roughly 37 times each.

One correction to how this was originally scoped: **the format is a
between-person randomisation, not a within-person crossing.** Every one of the
406 participants has exactly one `cov_group` value — 200 in the Likert arm, 206
in the slider arm. That is weaker than a crossed design for isolating format, but
it is still a randomised manipulation of response format holding the protocol,
the items and the sampling schedule fixed, which is rare.

The table also supplies a control that nobody designed. **The `stress` item was
never sliderised** — it is a 1–5 item in both arms. So it functions as a
within-study placebo: whatever changes for the items whose format actually
changed, but not for `stress`, is attributable to format rather than to the
people in each arm.

```{r}
#| label: tbl-format-floor
#| echo: true
if (nrow(item_detail) > 0) {
  v <- item_detail |> filter(table == "vollbracht_et_al_2026_ambulatory_assessment")
  if (nrow(v) > 0) {
    v |> transmute(Item = item, `Scale low` = lo, `Scale high` = hi,
                   Categories = n_cat, `At floor (%)` = round(floor_pct, 1),
                   `At ceiling (%)` = round(ceil_pct, 1)) |>
      arrange(desc(`At floor (%)`)) |>
      kable(caption = "Per-item floor mass in vollbracht, pooled across arms.")
  }
}
```

Every model uses the same mean structure — a person-level random intercept and a
within-person AR(1) — and adds a person-level dispersion random effect correlated
with that intercept, because the mean-level-to-innovation-variance correlation is
the parameter the preprint reports as biased.

```{r}
#| label: model-spec
#| eval: false
#| echo: true
# Shared across families; only the response distribution changes.
brms::bf(resp ~ 1 + ar(time = occ, gr = id) + (1 | p | id),
         sigma ~ 1 + (1 | p | id))                      # gaussian, and censored
brms::bf(resp | cens(cens_ind) ~ ..., sigma ~ ...)      # floor = left-censored
brms::bf(resp01 ~ ..., phi ~ ...)                       # zero-one-inflated beta
brms::bf(resp_ord ~ ..., disc ~ ...)                    # cumulative (graded)
```

Which families are even *admissible* depends on the format, and that is part of
the finding rather than an inconvenience. A zero-one-inflated beta cannot be
fitted to a five-category item; a graded model is impractical on a 101-point
slider. So "apply the censoring fix to both formats" — the field's current
habit — is not a choice between equally available options. For much of this data
the alternatives are ruled out before anyone compares fit.

## Results

### How much floor mass is actually out there

```{r}
#| label: fig-floor-spread
#| echo: true
if (nrow(item_detail) > 0) {
  ord <- floor_summary |> arrange(floor_max) |> pull(table)
  p <- item_detail |>
    left_join(select(floor_summary, table, format), by = "table") |>
    mutate(table = factor(table, levels = ord),
           text = sprintf("%s\nitem %s: %.1f%% at floor\nscale %g–%g (%d categories)",
                          table, item, floor_pct, lo, hi, n_cat)) |>
    ggplot(aes(x = floor_pct, y = table, colour = format, text = text)) +
    geom_point(alpha = .7, size = 1.9) +
    scale_colour_manual(values = c("Ordinal" = irw_amber,
                                   "Slider / continuous" = irw_blue), name = NULL) +
    labs(x = "Responses at the item's observed minimum (%)", y = NULL,
         title = "Per-item floor mass within each ESM table") +
    theme(legend.position = "top", axis.text.y = element_text(size = 8))
  ggplotly(p, tooltip = "text")
}
```

Two things to read off this, one of them not what we expected.

Floor mass varies enormously *within* a single instrument, so a table-level
summary is close to meaningless — which is the case for treating the floor as an
item property throughout.

The second is that **the two response formats carry about the same amount of
it.** Going continuous does not empty the bottom of the scale in practice; a
slider simply relocates where people pile up. That undercuts a tempting
shortcut — "use a VAS and the problem goes away" — and it is why the model
comparison below has to be run in both formats rather than only in the ordinal
one.

### Does the model you choose change the answer?

```{r}
#| label: tbl-stage-b
#| echo: true
if (nrow(stage_b) > 0 && "phi_ar1" %in% names(stage_b)) {
  stage_b |>
    filter(!is.na(phi_ar1)) |>
    transmute(Item = item, Arm = arm, Model = family,
              `Floor (%)` = round(floor_pct, 1),
              `AR(1)` = sprintf("%.3f [%.3f, %.3f]", phi_ar1, phi_ar1_lo, phi_ar1_hi),
              `cor(mean, disp)` = round(cor_mean_disp, 3),
              `elpd (see note)` = round(loo_elpd, 1),
              `Rhat` = round(rhat_ar, 3), `ESS` = round(ess_ar),
              `Div.` = n_divergent,
              OK = ifelse(converged, "yes", "**no**")) |>
    arrange(Item, Arm, Model) |>
    kable(caption = "Same data, competing theories of the floor. AR(1) is comparable across families. The correlation is not strictly so, and elpd is comparable ONLY within a response representation — see Limitations. Rows marked no in the OK column failed the convergence screen and are not estimates.")
}
```

```{r}
#| label: fig-ar1
#| echo: true
if (nrow(stage_b) > 0 && "phi_ar1" %in% names(stage_b)) {
  d <- stage_b |> filter(!is.na(phi_ar1), converged)
  if (nrow(d) > 0) {
    ggplot(d, aes(x = family, y = phi_ar1, colour = arm)) +
      geom_pointrange(aes(ymin = phi_ar1_lo, ymax = phi_ar1_hi),
                      position = position_dodge(width = .4)) +
      facet_wrap(~ item) +
      scale_colour_manual(values = c("Likert" = irw_amber, "Slider" = irw_blue),
                          name = NULL) +
      labs(x = NULL, y = "AR(1) coefficient (95% CrI)",
           title = "Carryover estimated under each theory of the floor",
           subtitle = "stress kept the same 1–5 format in both arms; att1 did not") +
      theme(legend.position = "top",
            axis.text.x = element_text(angle = 20, hjust = 1))
  }
}
```

```{r}
#| label: stage-b-readout
#| echo: false
#| output: asis
if (nrow(stage_b) > 0 && "phi_ar1" %in% names(stage_b)) {
  n_bad <- sum(!stage_b$converged, na.rm = TRUE)
  if (n_bad > 0) {
    cat("::: {.callout-warning}\n## ", n_bad, " fit(s) did not converge\n\n",
        "They are listed in the table above with `OK = no` and are excluded from ",
        "the figure. A short draft chain is the likely cause; they need rerunning ",
        "at higher iterations before anything is concluded from them.\n:::\n\n", sep = "")
  }
  d <- stage_b |> filter(!is.na(phi_ar1), converged)
  if (nrow(d) > 0) {
    sp <- d |> group_by(item, arm) |>
      summarise(spread = diff(range(phi_ar1)), .groups = "drop") |>
      arrange(desc(spread))
    cat("Across the fits that converged, the AR(1) estimate moves by up to **",
        sprintf("%.3f", max(sp$spread)), "** between model families on the same ",
        "data (", sp$item[1], ", ", sp$arm[1], " arm). ",
        "That is the practical question for anyone fitting these models: the ",
        "carryover parameter is not invariant to a modelling choice the field ",
        "usually makes by convention.\n\n", sep = "")
  }
}
```

```{r}
#| label: stage-b-missing
#| echo: false
#| output: asis
if (nrow(stage_b) == 0 || !"phi_ar1" %in% names(stage_b)) {
  cat("::: {.callout-note}\n## Model comparison not yet in the cache\n\n",
      "Stage A (the floor profiles above) is complete, but the Stage B model ",
      "fits are absent from this cache — either the run is still in progress or ",
      "`STAGE_B` was set to `FALSE`. The descriptive result stands on its own; ",
      "the comparison below fills in once the fits land.\n:::\n", sep = "")
}
```

## Limitations

**`elpd` cannot be compared across all four families.** The Gaussian and
censored models are fitted to a standardised continuous response and their
log-densities are directly comparable to each other. The cumulative model is
fitted to an ordinal probability mass function and the inflated beta to a
density on the unit interval — different response representations, so their
elpd values are not on a common scale and a "best fit" ranking across the whole
table would be meaningless. The column is shown because the within-pair
comparison (naive vs censored) is informative, not because the column has a
maximum worth reading.

**The dispersion parameter is not the same object across families.** The Gaussian
and censored models have a residual standard deviation; the inflated beta has a
precision, and the cumulative model a discrimination. Their person-level
correlations with the intercept play analogous roles but are not numerically the
same quantity, so the `cor(mean, dispersion)` column should be read for sign and
rough magnitude, not compared digit for digit. The AR(1) coefficient *is*
comparable across all four, which is why it carries the headline. This is a real
obstacle to the preprint's framing, not just to ours: the parameter it reports as
biased is defined inside a Gaussian latent-variable metric, and the alternatives a
reader might reach for do not all have one.

**The format contrast is between persons.** Randomisation makes it a fair
comparison, but it is not the within-person crossing that would let us hold the
participant fixed. The `stress` item mitigates this — same format in both arms,
so arm differences on it are attributable to sample rather than format — but one
control item is not a full design.

**These are bipolar scales.** Both the Likert (−2…5) and slider (−50…100) items
run from a negative to a positive pole, so the "floor" is an extreme negative
response, not the absence of a symptom. The preprint's motivating case —
depression and anxiety ESM, where the floor means "no symptoms today" and is very
common — is the unipolar situation. The ordinal tables in the pool above include
that case; vollbracht does not. Results here should not be read as settling what
happens on a unipolar symptom scale.

**Response direction is not harmonised in IRW.** "Floor" throughout means the
observed minimum of that item. For a reverse-keyed item that is the substantive
ceiling. This is reported as found rather than silently corrected, because
recovering the intended keying reliably is its own problem.

**The target parameter is hard to estimate at ESM sample sizes.** The first
draft run of this page produced Rhat 1.8, a bulk effective sample size of 3, and
416 divergent transitions on a fit that nonetheless reported a tight-looking
interval. The location-scale structure the preprint's parameter requires — each
person's mean correlated with their own residual variability — is only weakly
identified with tens of people and tens of occasions, and flat priors leave the
sampler nowhere to stand. The fits here use weakly informative priors, a
standardised response, and `adapt_delta = 0.99`. Convergence is reported per
fit rather than assumed, and non-converging fits are excluded from the figure.
This is a practical caveat on the preprint's recommendation as much as on this
page: a model can be the right one and still be beyond what the data will
support.

## Reproducibility

```{r}
#| label: repro
#| echo: false
#| output: asis
if (cache_ok) {
  cat("Computed on **", format(date_run), "**, profiling ", r_n_tables_run,
      " of ", r_all_candidates, " candidate tables.\n\n", sep = "")
}
```

Run `Rscript vignettes/esm_floor_compute.R` from the repository root, with
`REDIVIS_API_TOKEN` set to a read-scoped token. The script caches every table
profile and every model fit separately under `esm_floor_data/fits/`, so an
interrupted run resumes rather than restarting.

{{< include ../components/_data-version.qmd >}}

{{< include ../components/_source-links.qmd >}}

## References

::: {#refs}
:::
