---
title: "Statistical Fit vs. Predictive Value: M2/Item-Fit and IMV"
description: |
  Traditional IRT fit statistics (M2, item-level S-X2) test whether a model
  is statistically distinguishable from the data-generating process. IMV
  asks a different question: how much does the model actually improve
  prediction? We compute both across IRW tables and ask whether
  statistically-flagged tables actually show worse IMV, or whether the two
  diagnostics diverge.
date: today
bibliography: fit_vs_imv_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "fit_vs_imv_data/fit_vs_imv_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  candidate_tables <- .cache$candidate_tables
  date_run         <- .cache$date_run
  is_pilot         <- isTRUE(.cache$pilot)
  sig_alpha        <- .cache$sig_alpha %||% 0.05
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_tables_run   <- length(candidate_tables)
} else {
  summary_df <- tibble(
    table = character(), n_items = integer(), n_participants = integer(),
    construct_type = character(), imv_mean = numeric(), imv_sd = numeric(),
    imv_ci_lower = numeric(), imv_ci_upper = numeric(), M2 = numeric(),
    M2_df = numeric(), M2_p = numeric(), M2_RMSEA = numeric(),
    n_items_flagged = integer(), prop_items_flagged = numeric(),
    M2_flagged = logical(), itemfit_flagged = logical()
  )
  candidate_tables <- character()
  date_run         <- Sys.Date()
  is_pilot         <- FALSE
  sig_alpha        <- 0.05
  r_all_candidates <- 0L
  r_n_tables_run   <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/fit_vs_imv_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, pilot run.** This render uses a small pilot sample of ",
  r_n_tables_run, " tables (out of ", r_all_candidates, " eligible dichotomous ",
  "tables) to validate the pipeline and page layout before a full batch run. ",
  "All numbers here are real computed results — not placeholders — but treat ",
  "the *distributions* as a small, noisy preview rather than a representative ",
  "picture. Re-run `fit_vs_imv_compute.R` with `PILOT <- FALSE` to scale up.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`fit_vs_imv_compute.R`) directly if you're
relying on the numbers here.
:::

## Motivation

Traditional IRT fit assessment asks whether a model's predictions are
*statistically* distinguishable from the data — limited-information tests
like M2 [@sinharay2025assessment], item-level statistics like S-X2, and
residual diagnostics all test some version of "is there detectable misfit?"
But as Sinharay's [-@sinharay2025assessment] review of the field discusses,
statistical misfit and *practically consequential* misfit are not the same
thing: a large-sample table can show significant misfit that barely changes
person scores or predictions, while a smaller table's real misfit can go
statistically undetected.

The **InterModel Vigorish** [IMV; @domingue2025imv] offers a different lens:
instead of testing whether two models differ statistically, it measures how
much one model's held-out predictive accuracy improves on another's — see
the [IMV tutorial](imv.qmd) for the calculation itself on a single dataset.
This vignette asks two questions across many IRW tables at once: (1) how
large is IMV, typically, when a 2PL adds a discrimination parameter to a
Rasch/1PL baseline? and (2) do tables that traditional statistics flag as
poor-fitting actually show lower IMV gains from that richer model — or do
the two diagnostics diverge, flagging different tables as "notable" for
different reasons?

---

## Data and methods

### Dataset selection

```{r}
#| label: filter-datasets
#| eval: false

all_candidates <- irw_filter(
  n_categories   = 2,
  n_items        = c(5, 40),
  n_participants = c(500, Inf)
)
```

This pilot is restricted to **dichotomous tables only**, matching `imv.qmd`'s
own Rasch-vs-2PL comparison exactly. Polytomous tables would need a
different baseline/richer-model pair (e.g. a constrained vs. a graded
response model) that hasn't been validated here yet — left for a future
extension rather than attempted with an unverified comparison.
`r r_all_candidates` dichotomous tables met the item/participant criteria as
of `r format(date_run, "%B %Y")`.
`r if (is_pilot) "This page currently runs on a small pilot subset (see the warning above); a full pass would analyze all of them." else paste0("This pass analyzes all ", r_n_tables_run, " of them.")`

### Per-table computation

For each candidate table, `fit_and_compare()`:

1. Fetches the table and builds a wide response matrix, downsampling to at
   most 10,000 respondents and dropping zero-variance items (mirrors the
   [2PL across datasets](2pl_across_datasets.qmd) pipeline).
2. Fits a Rasch/1PL baseline and a 2PL model with a weak lognormal prior on
   discrimination — the identical specification used in `imv.qmd` and the
   2PL vignette.
3. Computes **IMV(Rasch, 2PL)** via the `imv` package's `imv()` function,
   which performs 5-fold cross-validation internally.
4. Computes two statistical fit diagnostics on the 2PL model, per
   Sinharay's survey: **M2** (`mirt::M2()`, a limited-information overall
   fit test) and **item-level S-X2** (`mirt::itemfit()`, default
   `fit_stats = "S_X2"`). These two are chosen for being well-supported and
   computationally tractable at this scale — not an attempt at Sinharay's
   full survey of fit procedures.
5. Classifies the table as statistically flagged or not by each method,
   using the conventional (uncorrected) `r sig_alpha` significance
   threshold — an arbitrary convention, not a validated cutoff, the same
   caveat that applies to the [local dependence](local_dependence.qmd)
   vignette's Q3 threshold.
6. Records: item/respondent counts, `construct_type`, IMV (mean, SD, and
   95% CI across folds), M2 (statistic, df, p-value, RMSEA), and the
   proportion of items individually flagged by S-X2.

```{r}
#| label: fit-function
#| eval: false

fit_and_compare <- function(table_name) {
  resp <- fetch_wide(table_name)  # irw_fetch() + irw_long2resp(), cleaned
  m0   <- mirt(resp, 1, "Rasch")
  m1   <- mirt(resp, model_2pl_with_lognormal_prior, itemtype = "2PL")
  imv_res  <- imv(m0, m1)                 # 5-fold CV internally
  m2_res   <- M2(m1, calcNull = FALSE)    # limited-information fit
  ifit_res <- itemfit(m1)                 # per-item S-X2
  # ... store imv_res$mean, m2_res$p, m2_res$RMSEA, prop(ifit_res$p.S_X2 < .05) ...
}
```

Of the `r r_n_tables_run` candidate tables in this pass, `r nrow(summary_df)`
produced usable results (the rest were dropped for too few usable items or a
model/diagnostic that failed outright).

::: {.callout-note}
To reproduce with current IRW holdings, re-run `fit_vs_imv_compute.R` and
commit the updated `fit_vs_imv_data/fit_vs_imv_results.rds`.
:::

## Results

### How large is IMV, across real datasets?

```{r}
#| label: fig-imv-dist
#| fig-cap: "Distribution of IMV(Rasch, 2PL) across sampled IRW tables. Each point is one table; error bars show the 95% CI across the 5 cross-validation folds."
#| fig-height: 4.5

imv_plot_df <- summary_df |>
  arrange(imv_mean) |>
  mutate(
    rank    = row_number(),
    tooltip = paste0(
      "Table: ", table,
      "<br>IMV: ", round(imv_mean, 4),
      "<br>95% CI: [", round(imv_ci_lower, 4), ", ", round(imv_ci_upper, 4), "]",
      "<br>n_items: ", n_items,
      "<br>N: ", n_participants
    )
  )

p <- ggplot(imv_plot_df, aes(x = imv_mean, y = rank, text = tooltip)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_errorbarh(aes(xmin = imv_ci_lower, xmax = imv_ci_upper), height = 0, colour = irw_grey) +
  geom_point(colour = irw_blue, size = 2.2) +
  labs(x = "IMV(Rasch, 2PL)", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggplotly(p, tooltip = "text")
```

```{r}
#| label: imv-dist-inline
#| echo: false

med_imv <- median(summary_df$imv_mean, na.rm = TRUE)
```

The median IMV gain from adding a 2PL discrimination parameter over Rasch is
`r round(med_imv, 4)` across this sample. For context, `gilbert_meta_2` — the
same table used in the [IMV tutorial](imv.qmd) — has an IMV of
`r if ("gilbert_meta_2" %in% summary_df$table) round(summary_df$imv_mean[summary_df$table == "gilbert_meta_2"], 4) else "not in this pilot sample"`
here, which should match that vignette's own number up to Monte Carlo /
cross-validation-fold noise.

### Does statistical misfit track predictive gain?

```{r}
#| label: fig-imv-vs-m2
#| fig-cap: "IMV(Rasch, 2PL) against the 2PL model's M2-based RMSEA. If statistical and practical significance agreed perfectly, worse-fitting (higher-RMSEA) tables would show systematically higher IMV. Colour marks whether the table's item-level S-X2 flags any items."
#| fig-height: 4.5

div_plot_df <- summary_df |>
  filter(!is.na(M2_RMSEA), !is.na(imv_mean)) |>
  mutate(
    itemfit_label = if_else(itemfit_flagged, "≥1 item flagged (S-X2)", "No items flagged"),
    tooltip = paste0(
      "Table: ", table,
      "<br>IMV: ", round(imv_mean, 4),
      "<br>M2 RMSEA: ", round(M2_RMSEA, 4),
      "<br>M2 p: ", signif(M2_p, 3),
      "<br>Items flagged: ", n_items_flagged, " / ", n_items
    )
  )

p2 <- ggplot(div_plot_df, aes(x = M2_RMSEA, y = imv_mean, colour = itemfit_label, text = tooltip)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(size = 2.8, alpha = 0.85) +
  scale_colour_manual(values = c("≥1 item flagged (S-X2)" = irw_red, "No items flagged" = irw_blue), name = NULL) +
  labs(x = "M2 RMSEA (2PL model)", y = "IMV(Rasch, 2PL)") +
  theme(legend.position = "bottom")

ggplotly(p2, tooltip = "text") |>
  layout(legend = list(orientation = "h", y = -0.2), margin = list(b = 80))
```

`r sum(summary_df$M2_flagged, na.rm = TRUE)` of `r sum(!is.na(summary_df$M2_flagged))`
tables are flagged by M2 at α = `r sig_alpha`, and
`r sum(summary_df$itemfit_flagged, na.rm = TRUE)` of `r sum(!is.na(summary_df$itemfit_flagged))`
have at least one item individually flagged by S-X2. With only
`r nrow(summary_df)` tables in this pilot, this figure is meant to confirm
the pipeline produces sensible, sane numbers — not to support a real
divergence claim. A full batch run is needed before drawing conclusions
about whether statistically-flagged tables systematically differ in IMV.

### All tables

```{r}
#| label: tbl-all
#| tbl-cap: "All tables in this pilot run: item/respondent counts, IMV, and statistical fit diagnostics."

summary_df |>
  transmute(
    Table = table,
    Items = n_items,
    N = n_participants,
    `Construct type` = construct_type,
    IMV = round(imv_mean, 4),
    `M2 RMSEA` = round(M2_RMSEA, 4),
    `M2 p` = signif(M2_p, 3),
    `Items flagged (S-X2)` = paste0(n_items_flagged, " / ", n_items)
  ) |>
  arrange(desc(IMV)) |>
  knitr::kable()
```

---

## Limitations

**This is a pilot, not a full run.** `r nrow(summary_df)` tables is nowhere
near enough to characterize the real relationship between statistical fit
and IMV, or to identify genuine divergent cases — the point of this pass is
to confirm the pipeline (fetch → fit → IMV → M2/itemfit → store) works
end-to-end and produces sane numbers, matching known results
(`gilbert_meta_2` against `imv.qmd`) where possible.

**Dichotomous tables only.** Polytomous tables are excluded until a
validated baseline/richer-model pair exists for them.

**Two fit statistics, not Sinharay's full survey.** M2 and S-X2 are two
well-supported, computationally tractable diagnostics; residual-based and
other limited-information diagnostics Sinharay discusses are not
implemented here.

**Uncorrected significance threshold.** α = `r sig_alpha` per table (M2) and
per item (S-X2) with no multiple-comparison correction — a stated
convention, not a validated cutoff.

**IMV has its own assumptions.** Cross-validated predictive gain depends on
sample size, item count, and the specific models compared; it is one useful
lens on model value, not a final verdict on which diagnostic is "right."

---

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the compute script (from the project root)
source("vignettes/fit_vs_imv_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/fit_vs_imv.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(summary_df$table, output_file = "fit_vs_imv_data/references.bib")
```
:::

---

## References
