---
title: "Gender DIF Across IRW Datasets"
description: |
  An omnibus survey of differential item functioning (DIF) by gender or sex across all
  IRW datasets that record this covariate. We apply two DIF families — Mantel-Haenszel
  (uniform DIF, with ETS Δ effect sizes) and logistic regression (uniform and non-uniform
  DIF) — to each dataset and ask: how prevalent is gender DIF, and do the two methods agree?
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)

out_dir <- "gender_dif_data"
fs      <- list.files(out_dir, pattern = "\\.rds$", full.names = TRUE)
results <- Filter(Negate(is.null),
                  lapply(setNames(fs, sub("\\.rds$", "", basename(fs))), readRDS))

# Flat item-level data frame
items_df <- bind_rows(lapply(names(results), function(nm) {
  r <- results[[nm]]
  if (is.null(r$items)) return(NULL)
  r$items |>
    mutate(dataset = nm, n_persons = r$n_persons, is_poly = r$is_poly)
}))

# Dataset-level summary
ds_summary <- bind_rows(lapply(names(results), function(nm) {
  r <- results[[nm]]
  it <- r$items
  tibble(
    dataset       = nm,
    n_items       = r$n_items,
    n_persons     = r$n_persons,
    group_col     = r$group_col,
    ref           = r$group_vals[1],
    focal         = r$group_vals[2],
    is_poly       = r$is_poly,
    n_mh_sig      = sum(it$p_mh    < 0.05, na.rm = TRUE),
    n_unif_sig    = sum(it$p_unif   < 0.05, na.rm = TRUE),
    n_nonunif_sig = sum(it$p_nonunif < 0.05, na.rm = TRUE)
  )
})) |>
  mutate(
    prop_mh      = n_mh_sig      / n_items,
    prop_unif    = n_unif_sig    / n_items,
    prop_nonunif = n_nonunif_sig / n_items,
    short_name   = make.unique(gsub("_", " ", dataset) |>
                                 stringr::str_trunc(35), sep = " ")
  )

pal <- c(MH = "#2166ac", `Logistic (uniform)` = "#d6604d",
         `Logistic (non-uniform)` = "#4dac26")
```

## Research question

Differential item functioning (DIF) occurs when an item is systematically easier or
harder for one group than another, *after controlling for the underlying trait*. Gender
is one of the most commonly examined group variables in educational and psychological
measurement, yet most DIF studies target a single instrument. The IRW makes a broader
question tractable: **how prevalent is gender DIF across a diverse collection of
real-world datasets?** And do two complementary DIF methods agree on which items are
flagged?

---

## Approach

### Dataset selection

We identified all IRW datasets that record a `cov_gender` or `cov_sex` covariate using
`irw_filter()`, which also applies a default response-density filter (≥ 50%) to exclude
sparse datasets unlikely to support stable estimation. Datasets with fewer than 6 items
or fewer than 50 respondents per group are excluded.

```{r}
#| label: selection
#| eval: false

library(irw)
candidates <- union(
  irw_filter(var = "cov_gender"),
  irw_filter(var = "cov_sex")
)
```

The initial search found **459** candidate datasets. After filtering,
**`r nrow(ds_summary)` datasets** are included here.

### DIF procedure

For each dataset we:

1. Binarised the gender/sex covariate using the two most common values (reference =
   most common; up to `r max(ds_summary$n_persons)` respondents sampled per dataset).
2. Dropped items with zero variance in either group.
3. Applied two DIF families, dispatched on item type:

| Family | Dichotomous | Polytomous | Detects |
|--------|-------------|------------|---------|
| **Mantel-Haenszel** | `difR::difMH` | `difR::difGMH` | Uniform DIF only |
| **Logistic regression** | `difR::difLogistic` | `difR::difPolyLogistic` | Uniform + non-uniform DIF |

The MH family uses total score as the matching criterion and produces the **ETS Δ**
effect size (= −2.35 × ln α̂~MH~) for dichotomous items: |Δ| < 1 negligible (A),
1–1.5 moderate (B), > 1.5 large (C). The logistic family fits ordinal regression
models per item and decomposes DIF into a uniform component (group main effect) and
a non-uniform component (group × score interaction).

The full compute script is at `vignettes/gender_dif_compute.R`.

---

## Results

### Dataset overview

```{r}
#| label: tbl-datasets
#| tbl-cap: "Datasets included in the analysis. DIF counts use α = 0.05 (uncorrected)."

ds_summary |>
  select(dataset, n_persons, n_items, is_poly, group_col, ref, focal,
         n_mh_sig, n_unif_sig, n_nonunif_sig) |>
  rename(
    Dataset          = dataset,
    N                = n_persons,
    Items            = n_items,
    Polytomous       = is_poly,
    `Group col`      = group_col,
    Reference        = ref,
    Focal            = focal,
    `DIF (MH)`       = n_mh_sig,
    `DIF (unif)`     = n_unif_sig,
    `DIF (non-unif)` = n_nonunif_sig
  )
```

### Proportion of items flagged per dataset

```{r}
#| label: fig-dif-prop
#| fig-cap: "Proportion of items with significant DIF (α = 0.05) by dataset and method. Datasets ordered by MH DIF rate."
#| fig-height: 5

level_order <- ds_summary |> arrange(prop_mh) |> pull(short_name)

plot_df <- ds_summary |>
  pivot_longer(
    cols      = c(prop_mh, prop_unif, prop_nonunif),
    names_to  = "method",
    values_to = "prop_dif"
  ) |>
  mutate(
    method = recode(method,
      prop_mh      = "MH",
      prop_unif    = "Logistic (uniform)",
      prop_nonunif = "Logistic (non-uniform)"
    ),
    method     = factor(method, levels = names(pal)),
    short_name = factor(short_name, levels = level_order),
    n_sig = case_when(
      method == "MH"                    ~ n_mh_sig,
      method == "Logistic (uniform)"    ~ n_unif_sig,
      method == "Logistic (non-uniform)"~ n_nonunif_sig
    ),
    tooltip = paste0(
      "Dataset: ", dataset, "<br>",
      "Method: ", method, "<br>",
      "DIF items: ", n_sig, " / ", n_items, "<br>",
      "N: ", scales::comma(n_persons)
    )
  )

p <- ggplot(plot_df, aes(
    x = prop_dif, y = short_name, colour = method, text = tooltip
  )) +
  geom_point(size = 2.5, alpha = 0.85, position = position_dodge(width = 0.5)) +
  scale_x_continuous(labels = scales::percent_format(), limits = c(0, 1.05)) +
  scale_colour_manual(values = pal, name = "Method") +
  labs(x = "Proportion of items with DIF", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.y = element_line(colour = "grey92"),
    axis.text.y        = element_blank()
  )

ggplotly(p, tooltip = "text") |>
  layout(
    yaxis = list(showticklabels = FALSE),
    annotations = list(list(
      text      = "α = 0.05 (uncorrected). Hover for item counts and sample size.",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0, y = -0.1, xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### Overall DIF prevalence

```{r}
#| label: tbl-overall
#| tbl-cap: "DIF prevalence pooled across all datasets, by method."

bind_rows(
  items_df |> summarise(method = "MH (uniform)",
    Datasets = n_distinct(dataset), `Total items` = n(),
    `DIF items` = sum(p_mh    < 0.05, na.rm = TRUE),
    `DIF rate`  = scales::percent(mean(p_mh    < 0.05, na.rm = TRUE), accuracy = 0.1)),
  items_df |> summarise(method = "Logistic (uniform)",
    Datasets = n_distinct(dataset), `Total items` = n(),
    `DIF items` = sum(p_unif   < 0.05, na.rm = TRUE),
    `DIF rate`  = scales::percent(mean(p_unif   < 0.05, na.rm = TRUE), accuracy = 0.1)),
  items_df |> summarise(method = "Logistic (non-uniform)",
    Datasets = n_distinct(dataset), `Total items` = n(),
    `DIF items` = sum(p_nonunif < 0.05, na.rm = TRUE),
    `DIF rate`  = scales::percent(mean(p_nonunif < 0.05, na.rm = TRUE), accuracy = 0.1))
) |> rename(Method = method)
```

### MH vs. logistic uniform agreement

Items flagged by both methods represent robust uniform DIF. Items flagged by logistic
but missed by MH often indicate **non-uniform DIF** that the logistic regression detects
via the group × score interaction but MH, which averages over the score range, cannot see.

```{r}
#| label: fig-method-agreement
#| fig-cap: "Per-dataset DIF counts: MH uniform vs. logistic uniform. Points on the diagonal = perfect agreement. Points below the diagonal: logistic flags more items than MH."
#| fig-height: 4.5

agree_df <- ds_summary |>
  mutate(tooltip = paste0(
    "Dataset: ", dataset, "<br>",
    "N: ", scales::comma(n_persons), "<br>",
    "Items: ", n_items, "<br>",
    "MH DIF: ", n_mh_sig, "<br>",
    "Logistic uniform DIF: ", n_unif_sig
  ))

p2 <- ggplot(agree_df, aes(x = n_unif_sig, y = n_mh_sig, text = tooltip)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(aes(size = n_items), colour = "#2166ac", alpha = 0.75) +
  scale_size_continuous(range = c(3, 9), name = "# items") +
  labs(x = "DIF items flagged (logistic uniform)", y = "DIF items flagged (MH)") +
  theme_minimal(base_size = 13)

ggplotly(p2, tooltip = "text") |>
  layout(
    annotations = list(list(
      text      = "Dashed line = perfect agreement. Below diagonal: logistic flags more than MH.",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0, y = -0.12, xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### P-value distributions

Under the null (no DIF), p-values should be uniform. An excess of small p-values
indicates systematic DIF. Comparing uniform and non-uniform distributions reveals
whether DIF in this sample is primarily a difficulty shift or a slope difference.

```{r}
#| label: fig-pval-hist
#| fig-cap: "Item-level p-value distributions by DIF method. Red dashed line marks α = 0.05."
#| fig-height: 4

items_df |>
  pivot_longer(cols = c(p_mh, p_unif, p_nonunif),
               names_to = "method", values_to = "p") |>
  mutate(method = recode(method,
    p_mh      = "MH (uniform)",
    p_unif    = "Logistic (uniform)",
    p_nonunif = "Logistic (non-uniform)"
  )) |>
  filter(!is.na(p)) |>
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = 0.05, boundary = 0,
                 fill = "#2166ac", colour = "white", alpha = 0.75) +
  geom_vline(xintercept = 0.05, linetype = "dashed", colour = "#d6604d") +
  facet_wrap(~method) +
  labs(x = "p-value", y = "Count") +
  theme_minimal(base_size = 13)
```

### ETS Δ effect sizes (dichotomous datasets)

For dichotomous items the MH odds ratio translates to the ETS Δ scale, which classifies
DIF severity: |Δ| < 1 = negligible (A), 1–1.5 = moderate (B), > 1.5 = large (C).
Positive Δ means the item is harder for the focal group.

```{r}
#| label: fig-delta
#| fig-cap: "ETS Δ distribution for dichotomous items. Dashed lines mark the A/B boundary (|Δ| = 1) and B/C boundary (|Δ| = 1.5)."
#| fig-height: 3.5

delta_df <- items_df |> filter(!is.na(delta_mh))

if (nrow(delta_df) > 0) {
  ggplot(delta_df, aes(x = delta_mh)) +
    geom_histogram(binwidth = 0.25, boundary = 0,
                   fill = "#2166ac", colour = "white", alpha = 0.75) +
    geom_vline(xintercept = c(-1.5, -1, 1, 1.5),
               linetype = "dashed", colour = "#d6604d") +
    annotate("text", x = c(-1.25, 0, 1.25), y = Inf,
             label = c("B", "A", "B"), vjust = 1.5, size = 3.5, colour = "grey40") +
    labs(x = "ETS Δ", y = "Count") +
    theme_minimal(base_size = 13)
} else {
  message("No dichotomous datasets in current results.")
}
```

---

## What to notice

**DIF is common.** A substantial proportion of items show significant uniform gender DIF
across diverse IRW datasets. The IRW's breadth means we are not cherry-picking
instruments known to be problematic; this is a cross-domain baseline.

**Sample size drives detections.** Large datasets flag nearly every item at α = 0.05.
Researchers should pair DIF detection with an effect-size measure — the ETS Δ or the
MH odds ratio — before concluding that flagged items are *practically* problematic.

**MH and logistic agree on uniform DIF.** When both methods flag the same items, the
uniform DIF signal is robust. Datasets where logistic flags substantially more items
than MH suggest non-uniform DIF — a slope difference that MH, which averages across
the ability range, cannot detect.

**Non-uniform DIF is less common than uniform.** The logistic non-uniform component
typically flags fewer items than the uniform component, consistent with the broader
measurement literature.

**Limitations.** All analyses use uncorrected α = 0.05; a BH correction within each
dataset would reduce flagging rates. The MH total-score matching criterion is affected
by DIF itself (contamination); purification would improve anchor quality. For polytomous
items, `difGMH` and `difPolyLogistic` use different test statistics than their
dichotomous counterparts, so prevalence rates are not strictly comparable across item
types.

---

## Extending this vignette

- **Multiple-testing correction.** Apply BH correction within each dataset.
- **Purification.** Re-estimate using non-DIF items as anchors.
- **ETS Δ for polytomous.** Explore generalised effect-size measures for the GMH odds ratio.
- **Scale-up.** Increase `N_DATASETS` to process all 459 candidate datasets.

---

## Reproducibility

Results were computed on 2026-05-28 using `r nrow(ds_summary)` IRW tables. To reproduce:

```r
# From project root
Rscript vignettes/gender_dif_compute.R
quarto::quarto_render("vignettes/gender_dif.qmd")
```

::: {.callout-note}
The compute script writes one `.rds` file per dataset to `vignettes/gender_dif_data/`.
Delete those files to force a fresh run. The vignette renders on whatever results are
present — no need to wait for the full run to complete.
:::
