---
title: "Gender DIF Across IRW Datasets"
description: |
  An omnibus survey of differential item functioning (DIF) by gender or sex across all
  IRW datasets that record this covariate. We apply two DIF families — Mantel-Haenszel
  (uniform DIF, with ETS Δ effect sizes) and logistic regression (uniform and non-uniform
  DIF) — to each dataset and ask: how prevalent is gender DIF, and do the two methods agree?
bibliography: gender_dif_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

out_dir <- "gender_dif_data"
fs      <- list.files(out_dir, pattern = "\\.rds$", full.names = TRUE)
results <- Filter(Negate(is.null),
                  lapply(setNames(fs, sub("\\.rds$", "", basename(fs))), readRDS))

# Flat item-level data frame
items_df <- bind_rows(lapply(names(results), function(nm) {
  r <- results[[nm]]
  if (is.null(r$items)) return(NULL)
  r$items |>
    mutate(dataset = nm, n_persons = r$n_persons, is_poly = r$is_poly)
}))

# Dataset-level summary
ds_summary <- bind_rows(lapply(names(results), function(nm) {
  r <- results[[nm]]
  it <- r$items
  log_run <- if ("logistic_run" %in% names(it)) any(it$logistic_run, na.rm = TRUE) else NA
  tibble(
    dataset       = nm,
    n_items       = r$n_items,
    n_persons     = r$n_persons,
    group_col     = r$group_col,
    ref           = r$group_vals[1],
    focal         = r$group_vals[2],
    is_poly       = r$is_poly,
    logistic_run  = log_run,
    n_mh_sig      = sum(it$p_mh    < 0.05, na.rm = TRUE),
    n_unif_sig    = if (isTRUE(log_run)) sum(it$p_unif    < 0.05, na.rm = TRUE) else NA_integer_,
    n_nonunif_sig = if (isTRUE(log_run)) sum(it$p_nonunif < 0.05, na.rm = TRUE) else NA_integer_,
    median_r2     = if ("delta_r2_unif" %in% names(it))
                      median(it$delta_r2_unif, na.rm = TRUE) else NA_real_
  )
})) |>
  mutate(
    prop_mh      = n_mh_sig      / n_items,
    prop_unif    = n_unif_sig    / n_items,
    prop_nonunif = n_nonunif_sig / n_items,
    short_name   = make.unique(gsub("_", " ", dataset) |>
                                 stringr::str_trunc(35), sep = " ")
  )

pal <- c(MH = irw_blue, `Logistic (uniform)` = "#d6604d",
         `Logistic (non-uniform)` = "#4dac26")
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`gender_dif_compute.R`) directly if you're
relying on the numbers here.
:::

## Motivation

Differential item functioning (DIF) occurs when an item is systematically easier or
harder for one group than another, *after controlling for the underlying trait*. Gender
is one of the most commonly examined group variables in educational and psychological
measurement, yet most DIF studies target a single instrument. The IRW makes a broader
question tractable: **how prevalent is gender DIF across a diverse collection of
real-world datasets?** And do two complementary DIF methods agree on which items are
flagged?

---

## Data and methods

### Dataset selection

We identified all IRW datasets that record a `cov_gender` or `cov_sex` covariate using
`irw_filter()`, which also applies a default response-density filter (≥ 50%) to exclude
sparse datasets unlikely to support stable estimation. Datasets with fewer than 6 items
or fewer than 50 respondents per group are excluded.

```{r}
#| label: selection
#| eval: false

library(irw)
candidates <- union(
  irw_filter(var = "cov_gender"),
  irw_filter(var = "cov_sex")
)
```

The initial search found **459** candidate datasets. After filtering,
**`r nrow(ds_summary)` datasets** are included here.

### DIF procedure

For each dataset we:

1. Binarised the gender/sex covariate using the two most common values (reference =
   most common; up to `r max(ds_summary$n_persons)` respondents sampled per dataset).
2. Dropped items with zero variance in either group.
3. Applied two DIF families, dispatched on item type:

| Family | Dichotomous | Polytomous | Detects |
|--------|-------------|------------|---------|
| **Mantel-Haenszel** | `difR::difMH` | `difR::difGMH` | Uniform DIF only |
| **Logistic regression** | `difR::difLogistic` | `difR::difPolyLogistic` | Uniform + non-uniform DIF |

The MH family [@hollandthayer1988mh] uses total score as the matching criterion and
produces the **ETS Δ** effect size (= −2.35 × ln α̂~MH~) for dichotomous items
[@hollandthayer1985delta]: |Δ| < 1 negligible (A), 1–1.5 moderate (B), > 1.5 large (C).
The logistic family fits ordinal regression models per item and decomposes DIF into a
uniform component (group main effect) and a non-uniform component (group × score
interaction).

The full compute script is at `vignettes/gender_dif_compute.R`.

---

## Results

### Dataset overview

```{r}
#| label: tbl-datasets
#| tbl-cap: "Datasets included in the analysis. DIF counts use α = 0.05 (uncorrected)."

ds_summary |>
  select(dataset, n_persons, n_items, is_poly, group_col, ref, focal,
         n_mh_sig, n_unif_sig, n_nonunif_sig) |>
  rename(
    Dataset          = dataset,
    N                = n_persons,
    Items            = n_items,
    Polytomous       = is_poly,
    `Group col`      = group_col,
    Reference        = ref,
    Focal            = focal,
    `DIF (MH)`       = n_mh_sig,
    `DIF (unif)`     = n_unif_sig,
    `DIF (non-unif)` = n_nonunif_sig
  )
```

### Proportion of items flagged per dataset

```{r}
#| label: fig-dif-prop
#| fig-cap: "Proportion of items with significant DIF (α = 0.05) by dataset and method. Datasets ordered by MH DIF rate."
#| fig-height: 5

level_order <- ds_summary |> arrange(prop_mh) |> pull(short_name)

plot_df <- ds_summary |>
  pivot_longer(
    cols      = c(prop_mh, prop_unif, prop_nonunif),
    names_to  = "method",
    values_to = "prop_dif"
  ) |>
  mutate(
    method = recode(method,
      prop_mh      = "MH",
      prop_unif    = "Logistic (uniform)",
      prop_nonunif = "Logistic (non-uniform)"
    ),
    method     = factor(method, levels = names(pal)),
    short_name = factor(short_name, levels = level_order),
    n_sig = case_when(
      method == "MH"                    ~ n_mh_sig,
      method == "Logistic (uniform)"    ~ n_unif_sig,
      method == "Logistic (non-uniform)"~ n_nonunif_sig
    ),
    tooltip = paste0(
      "Dataset: ", dataset, "<br>",
      "Method: ", method, "<br>",
      "DIF items: ", n_sig, " / ", n_items, "<br>",
      "N: ", scales::comma(n_persons)
    )
  )

p <- ggplot(plot_df, aes(
    x = prop_dif, y = short_name, colour = method, text = tooltip
  )) +
  geom_point(size = 2.5, alpha = 0.85, position = position_dodge(width = 0.5)) +
  scale_x_continuous(labels = scales::percent_format(), limits = c(0, 1.05)) +
  scale_colour_manual(values = pal, name = "Method") +
  labs(x = "Proportion of items with DIF", y = NULL) +
  theme(
    panel.grid.major.y = element_line(colour = "grey92"),
    axis.text.y        = element_blank()
  )

ggplotly(p, tooltip = "text") |>
  layout(
    yaxis = list(showticklabels = FALSE),
    annotations = list(list(
      text      = "α = 0.05 (uncorrected). Hover for item counts and sample size.",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0, y = -0.1, xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### Overall DIF prevalence

```{r}
#| label: tbl-overall
#| tbl-cap: "DIF prevalence pooled across all datasets, by method."

bind_rows(
  items_df |> summarise(method = "MH (uniform)",
    Datasets = n_distinct(dataset), `Total items` = n(),
    `DIF items` = sum(p_mh    < 0.05, na.rm = TRUE),
    `DIF rate`  = scales::percent(mean(p_mh    < 0.05, na.rm = TRUE), accuracy = 0.1)),
  items_df |> summarise(method = "Logistic (uniform)",
    Datasets = n_distinct(dataset), `Total items` = n(),
    `DIF items` = sum(p_unif   < 0.05, na.rm = TRUE),
    `DIF rate`  = scales::percent(mean(p_unif   < 0.05, na.rm = TRUE), accuracy = 0.1)),
  items_df |> summarise(method = "Logistic (non-uniform)",
    Datasets = n_distinct(dataset), `Total items` = n(),
    `DIF items` = sum(p_nonunif < 0.05, na.rm = TRUE),
    `DIF rate`  = scales::percent(mean(p_nonunif < 0.05, na.rm = TRUE), accuracy = 0.1))
) |> rename(Method = method)
```

DIF is real and common by both methods, and the IRW's scale makes this a uniquely broad
cross-domain baseline — gender DIF is not confined to a few pathological instruments.
But the logistic non-uniform component (group × score interaction) flags noticeably
fewer items than either uniform measure, consistent with the broader measurement
literature: most gender DIF found here looks like a uniform difficulty shift rather
than a slope difference.

### Significance vs. effect size: the key divergence

A central finding of this analysis is that MH significance and ΔR² effect size can
come apart dramatically, particularly for polytomous items. Many datasets show MH
flagging rates near 100% while their median ΔR² stays in negligible territory (< 0.035).
This reflects a known limitation of Mantel-Haenszel for polytomous scales: when
groups differ substantially on the underlying trait mean, the total-score matching
criterion is contaminated and the GMH statistic accumulates to near-zero p-values for
every item — even when per-item effect sizes are trivial. Datasets that land in the
**lower-right** of the figure below exemplify this pattern. Datasets in the
**upper-right** (high flag rate *and* large ΔR²) represent genuine widespread DIF.

```{r}
#| label: fig-sig-vs-effect
#| fig-cap: "Dataset-level relationship between MH flagging rate and median logistic ΔR² effect size. Horizontal dashed lines mark the negligible/moderate (0.035) and moderate/large (0.07) ΔR² thresholds. Datasets without logistic results (polytomous, > 20 items) are excluded."
#| fig-height: 4.5

div_df <- ds_summary |>
  filter(!is.na(median_r2)) |>
  mutate(
    item_type = if_else(is_poly, "Polytomous", "Dichotomous"),
    tooltip   = paste0(
      "Dataset: ", dataset, "<br>",
      "Type: ", if_else(is_poly, "Polytomous", "Dichotomous"), "<br>",
      "MH flagged: ", scales::percent(prop_mh, accuracy = 1), "<br>",
      "Median ΔR²: ", round(median_r2, 4), "<br>",
      "N: ", scales::comma(n_persons)
    )
  )

p_div <- ggplot(div_df, aes(x = prop_mh, y = median_r2,
                             colour = item_type, size = n_persons, text = tooltip)) +
  geom_hline(yintercept = c(0.035, 0.07), linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.8) +
  scale_x_continuous(labels = scales::percent_format(), limits = c(0, 1.05)) +
  scale_colour_manual(values = c(Dichotomous = irw_blue, Polytomous = "#4dac26"),
                      name = "Item type") +
  scale_size_continuous(range = c(3, 9), name = "N", guide = "none") +
  labs(x = "Proportion of items flagged by MH (α = 0.05)",
       y = "Median ΔR² (logistic uniform)")

# Threshold labels are added as native plotly annotations (xref = "paper") rather
# than a ggplot annotate() layer, which ggplotly() clips at the left panel edge.
ggplotly(p_div, tooltip = "text") |>
  layout(annotations = list(
    list(
      text      = "Negligible / Moderate", showarrow = FALSE,
      xref = "paper", yref = "y", x = 0.01, y = 0.037,
      xanchor = "left", yanchor = "bottom",
      font = list(size = 11, color = "grey45")
    ),
    list(
      text      = "Moderate / Large", showarrow = FALSE,
      xref = "paper", yref = "y", x = 0.01, y = 0.072,
      xanchor = "left", yanchor = "bottom",
      font = list(size = 11, color = "grey45")
    ),
    list(
      text      = "Point size = sample size. Hover for dataset details.",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0, y = -0.12, xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    )
  ))
```

The same pattern holds at the item level: MH-significant polytomous items cluster near
zero on ΔR², while MH-significant dichotomous items span a wider effect-size range.

```{r}
#| label: fig-item-sig-vs-r2
#| fig-cap: "ΔR² distributions for MH-significant vs. non-significant items, by item type. Dashed line marks the negligible/moderate threshold (ΔR² = 0.035). Items without logistic results are excluded."
#| fig-height: 4

items_df |>
  filter(!is.na(delta_r2_unif), !is.na(p_mh)) |>
  mutate(
    mh_sig    = factor(if_else(p_mh < 0.05, "MH significant", "MH not significant"),
                       levels = c("MH not significant", "MH significant")),
    item_type = if_else(is_poly, "Polytomous", "Dichotomous")
  ) |>
  ggplot(aes(x = mh_sig, y = delta_r2_unif, fill = item_type)) +
  geom_hline(yintercept = 0.035, linetype = "dashed", colour = "#d6604d") +
  geom_boxplot(outlier.size = 0.8, alpha = 0.8) +
  facet_wrap(~item_type, scales = "free_y") +
  scale_fill_manual(values = c(Dichotomous = irw_blue, Polytomous = "#4dac26"),
                    guide = "none") +
  labs(x = NULL, y = expression(Delta * R^2 ~ "(logistic uniform)")) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

Even where MH significance is high, the majority of items have ΔR² below 0.035 — the
conventional "negligible" threshold [@jodoingierl2001effect]. Datasets with genuinely
large per-item effects (ΔR² > 0.07) are the exception. Read on its own, significance-only
DIF reporting would inflate the apparent prevalence of meaningful gender bias.

### P-value distributions

Under the null (no DIF), p-values should be uniform. An excess of small p-values
indicates systematic DIF. Comparing uniform and non-uniform distributions reveals
whether DIF in this sample is primarily a difficulty shift or a slope difference.

```{r}
#| label: fig-pval-hist
#| fig-cap: "Item-level p-value distributions by DIF method. Red dashed line marks α = 0.05."
#| fig-height: 4

items_df |>
  pivot_longer(cols = c(p_mh, p_unif, p_nonunif),
               names_to = "method", values_to = "p") |>
  mutate(method = recode(method,
    p_mh      = "MH (uniform)",
    p_unif    = "Logistic (uniform)",
    p_nonunif = "Logistic (non-uniform)"
  )) |>
  filter(!is.na(p)) |>
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = 0.05, boundary = 0,
                 fill = irw_blue, colour = "white", alpha = 0.75) +
  geom_vline(xintercept = 0.05, linetype = "dashed", colour = "#d6604d") +
  facet_wrap(~method) +
  labs(x = "p-value", y = "Count")
```

### ETS Δ effect sizes (dichotomous datasets)

For dichotomous items the MH odds ratio translates to the ETS Δ scale
[@hollandthayer1985delta], which classifies DIF severity: |Δ| < 1 = negligible (A),
1–1.5 = moderate (B), > 1.5 = large (C). Positive Δ means the item is harder for the
focal group.

```{r}
#| label: fig-delta
#| fig-cap: "ETS Δ distribution for dichotomous items. Dashed lines mark the A/B boundary (|Δ| = 1) and B/C boundary (|Δ| = 1.5)."
#| fig-height: 3.5

delta_df <- items_df |> filter(!is.na(delta_mh))

if (nrow(delta_df) > 0) {
  ggplot(delta_df, aes(x = delta_mh)) +
    geom_histogram(binwidth = 0.25, boundary = 0,
                   fill = irw_blue, colour = "white", alpha = 0.75) +
    geom_vline(xintercept = c(-1.5, -1, 1, 1.5),
               linetype = "dashed", colour = "#d6604d") +
    annotate("text", x = c(-1.25, 0, 1.25), y = Inf,
             label = c("B", "A", "B"), vjust = 1.5, size = 3.5, colour = "grey40") +
    labs(x = "ETS Δ", y = "Count")
} else {
  message("No dichotomous datasets in current results.")
}
```

### Logistic R² effect sizes (uniform DIF)

The logistic regression approach produces a pseudo-R² difference [@zumbo1999handbook]
for each item: the increase in model fit when a group term is added. This provides a
sample-size-independent effect size for both dichotomous and polytomous items.
Conventional cutoffs [@jodoingierl2001effect]: < 0.035 negligible, 0.035–0.07 moderate,
> 0.07 large.

```{r}
#| label: fig-delta-r2
#| fig-cap: "Distribution of pseudo-R² effect sizes for uniform DIF (logistic). Items with logistic skipped (polytomous datasets > 20 items) are excluded. Dashed lines mark negligible/moderate and moderate/large cutoffs."
#| fig-height: 3.5

r2_df <- items_df |>
  filter("delta_r2_unif" %in% names(items_df)) |>
  filter(!is.na(delta_r2_unif)) |>
  mutate(item_type = if_else(is_poly, "Polytomous", "Dichotomous"))

if (nrow(r2_df) > 0) {
  ggplot(r2_df, aes(x = delta_r2_unif, fill = item_type)) +
    geom_histogram(binwidth = 0.005, boundary = 0,
                   colour = "white", alpha = 0.8, position = "stack") +
    geom_vline(xintercept = c(0.035, 0.07),
               linetype = "dashed", colour = "#d6604d") +
    annotate("text", x = c(0.017, 0.0525, 0.11), y = Inf,
             label = c("Negligible", "Moderate", "Large"),
             vjust = 1.5, size = 3.5, colour = "grey40") +
    scale_fill_manual(values = c(Dichotomous = irw_blue, Polytomous = "#4dac26"),
                      name = "Item type") +
    labs(x = expression(Delta*R^2~"(uniform DIF)"), y = "Count")
} else {
  message("delta_r2_unif not available — rerun compute script.")
}
```

---

## Limitations

All analyses use uncorrected α = 0.05; Benjamini-Hochberg correction within each dataset
would reduce flagging rates. The GMH anticonservatism for polytomous items persists even
with purification when group trait-mean differences are large — this is a fundamental
limitation of total-score matching for polytomous data [@hollandthayer1988mh], not a
tuning issue. Logistic ΔR² is the more reliable metric in those settings. For polytomous
datasets with more than 20 items, `difPolyLogistic` was skipped due to computational
cost; those datasets contribute MH results only.

More generally: this vignette reports statistical *and* effect-size measures of DIF, but
doesn't establish which flagged items are substantively meaningful for score
interpretation in any given instrument, or attempt any correction for the fact that
gender is only one of many possible DIF grouping variables IRW tables could support.

---

## Reproducibility

Results were computed on 2026-05-28 using `r nrow(ds_summary)` IRW tables. To reproduce:

```r
# From project root
Rscript vignettes/gender_dif_compute.R
quarto::quarto_render("vignettes/gender_dif.qmd")
```

::: {.callout-note}
The compute script writes one `.rds` file per dataset to `vignettes/gender_dif_data/`.
Delete those files to force a fresh run. The vignette renders on whatever results are
present — no need to wait for the full run to complete.
:::

## References
