---
title: "Gender DIF Across IRW Datasets"
description: |
  An omnibus survey of differential item functioning (DIF) by gender or sex across all
  IRW datasets that record this covariate. We fit multigroup 1PL and 2PL models to each
  dataset and apply Wald-based DIF tests, then ask: how prevalent is gender DIF, and do
  the two models agree on which items are flagged?
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)
library(plotly)

results <- readRDS("gender_dif_data/combined_results.rds")

# Drop the one dataset that had too few rows after the density filter
results <- Filter(Negate(is.null), results)

# Build a flat item-level data frame
item_rows <- lapply(names(results), function(nm) {
  r <- results[[nm]]
  bind_rows(
    if (!is.null(r$dif_1pl))
      mutate(as.data.frame(r$dif_1pl), item = rownames(r$dif_1pl), model = "1PL"),
    if (!is.null(r$dif_2pl))
      mutate(as.data.frame(r$dif_2pl), item = rownames(r$dif_2pl), model = "2PL")
  ) |>
    mutate(dataset = nm, n_persons = r$n_persons)
})
items_df <- bind_rows(item_rows)

# Dataset-level summary
ds_summary <- lapply(names(results), function(nm) {
  r <- results[[nm]]
  dif1 <- r$dif_1pl
  dif2 <- r$dif_2pl
  tibble(
    dataset     = nm,
    n_items     = r$n_items,
    n_persons   = r$n_persons,
    group_col   = r$group_col,
    ref         = r$group_vals[1],
    focal       = r$group_vals[2],
    dif1_sig    = if (!is.null(dif1)) sum(dif1$p < 0.05, na.rm = TRUE) else NA_integer_,
    dif2_sig    = if (!is.null(dif2)) sum(dif2$p < 0.05, na.rm = TRUE) else NA_integer_
  )
}) |> bind_rows() |>
  mutate(
    prop_dif1 = dif1_sig / n_items,
    prop_dif2 = dif2_sig / n_items,
    short_name = gsub("_", " ", dataset) |> stringr::str_trunc(35)
  )
```

## Research question

Differential item functioning (DIF) occurs when an item is systematically easier or
harder for one group than another, *after controlling for the underlying trait*. Gender
is one of the most commonly examined group variables in educational and psychological
measurement, yet most DIF studies target a single instrument. The IRW makes a broader
question tractable: **how prevalent is uniform gender DIF across a diverse collection of
real-world datasets?** And does the choice of 1PL vs. 2PL model change which items get
flagged?

---

## Approach

### Dataset selection

We identified all IRW datasets that record a `cov_gender` or `cov_sex` covariate using
`irw_filter()`, which also applies a default response-density filter (≥ 50%) to exclude
sparse datasets unlikely to support stable multigroup estimation.

```{r}
#| label: selection
#| eval: false

library(irw)
candidates <- union(
  irw_filter(var = "cov_gender"),
  irw_filter(var = "cov_sex")
)
```

The initial search found **459** candidate datasets. After filtering for sufficient
sample size, **`r nrow(ds_summary)` datasets** are included in this analysis.

### DIF procedure

For each dataset we:

1. Binarised the gender/sex covariate using the two most common values (reference =
   most common).
2. Dropped items with zero variance in either group.
3. Fit two multigroup models via `mirt::multipleGroup()`, each with slopes and
   intercepts constrained equal across groups (the invariant baseline), with means and
   variances freed:
   - **1PL** — Rasch for dichotomous items, PCM for polytomous.
   - **2PL** — 2PL for dichotomous, graded response for polytomous.
4. Applied `mirt::DIF(..., scheme = "drop")` to test each item's difficulty (intercept)
   parameters one at a time via a Wald χ² test.

This detects **uniform DIF** — a shift in item difficulty between groups that is
constant across the trait range. Non-uniform DIF (slope differences) is not tested here.

The full compute script is at `vignettes/gender_dif_compute.R`.

---

## Results

### Dataset overview

```{r}
#| label: tbl-datasets
#| tbl-cap: "Datasets included in the analysis. DIF counts use α = 0.05 (uncorrected)."

ds_summary |>
  select(dataset, n_persons, n_items, group_col, ref, focal, dif1_sig, dif2_sig) |>
  rename(
    Dataset      = dataset,
    N            = n_persons,
    Items        = n_items,
    `Group col`  = group_col,
    Reference    = ref,
    Focal        = focal,
    `DIF (1PL)`  = dif1_sig,
    `DIF (2PL)`  = dif2_sig
  )
```

### Proportion of items flagged per dataset

```{r}
#| label: fig-dif-prop
#| fig-cap: "Proportion of items with significant uniform DIF (α = 0.05) by dataset and model. Datasets ordered by 1PL DIF rate. Horizontal jitter separates the two model estimates."
#| fig-height: 5

order_df    <- ds_summary |> select(short_name, prop_dif1)
level_order <- ds_summary |> arrange(prop_dif1) |> pull(short_name)

plot_df <- ds_summary |>
  tidyr::pivot_longer(
    cols      = c(prop_dif1, prop_dif2),
    names_to  = "model",
    values_to = "prop_dif"
  ) |>
  mutate(model = ifelse(model == "prop_dif1", "1PL", "2PL")) |>
  left_join(order_df, by = "short_name") |>
  mutate(short_name = factor(short_name, levels = level_order))

p <- ggplot(plot_df, aes(
    x    = prop_dif,
    y    = short_name,
    colour = model,
    text = paste0(
      "Dataset: ", dataset, "<br>",
      "Model: ", model, "<br>",
      "DIF items: ", ifelse(model == "1PL", dif1_sig, dif2_sig),
      " / ", n_items, "<br>",
      "N: ", scales::comma(n_persons)
    )
  )) +
  geom_point(size = 3, alpha = 0.85, position = position_dodge(width = 0.4)) +
  scale_x_continuous(labels = scales::percent_format(), limits = c(0, 1.05)) +
  scale_colour_manual(values = c("1PL" = "#2166ac", "2PL" = "#d6604d")) +
  labs(x = "Proportion of items with DIF", y = NULL, colour = "Model") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))

ggplotly(p, tooltip = "text") |>
  layout(
    annotations = list(list(
      text      = "α = 0.05 (uncorrected). Hover for item counts and sample size.",
      showarrow = FALSE,
      xref = "paper", yref = "paper",
      x = 0, y = -0.1,
      xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### Overall DIF prevalence

```{r}
#| label: tbl-overall
#| tbl-cap: "Overall DIF prevalence pooled across all datasets."

items_df |>
  group_by(model) |>
  summarise(
    Datasets     = n_distinct(dataset),
    `Total items` = n(),
    `DIF items`   = sum(p < 0.05, na.rm = TRUE),
    `DIF rate`    = scales::percent(mean(p < 0.05, na.rm = TRUE), accuracy = 0.1),
    .groups = "drop"
  ) |>
  rename(Model = model)
```

### 1PL vs. 2PL agreement

When the two models disagree, it is informative: a DIF flag under 1PL that disappears
under 2PL often indicates *non-uniform* DIF (a slope difference), not a true intercept
shift. Freeing the discrimination parameter in the 2PL absorbs that variation, so the
intercept test no longer rejects.

```{r}
#| label: fig-model-agreement
#| fig-cap: "Per-dataset DIF counts under 1PL vs. 2PL. Points on the diagonal indicate full agreement; points above it mean 1PL flagged more items than 2PL."
#| fig-height: 4.5

agree_df <- ds_summary |>
  mutate(tooltip = paste0(
    "Dataset: ", dataset, "<br>",
    "N: ", scales::comma(n_persons), "<br>",
    "Items: ", n_items, "<br>",
    "1PL DIF: ", dif1_sig, "<br>",
    "2PL DIF: ", dif2_sig
  ))

p2 <- ggplot(agree_df, aes(x = dif2_sig, y = dif1_sig, text = tooltip)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(aes(size = n_items), colour = "#2166ac", alpha = 0.75) +
  scale_size_continuous(range = c(3, 9), name = "# items") +
  labs(
    x = "DIF items flagged (2PL)",
    y = "DIF items flagged (1PL)"
  ) +
  theme_minimal(base_size = 13)

ggplotly(p2, tooltip = "text") |>
  layout(
    annotations = list(list(
      text      = "Dashed line = perfect agreement. Points above diagonal: 1PL flags more than 2PL.",
      showarrow = FALSE,
      xref = "paper", yref = "paper",
      x = 0, y = -0.12,
      xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### P-value distributions

A histogram of item-level p-values reveals the overall pattern of evidence. Under the
null (no DIF), p-values should be uniform. An excess of small p-values indicates
systematic DIF; a spike near 1.0 suggests items that genuinely function equivalently
across groups.

```{r}
#| label: fig-pval-hist
#| fig-cap: "Distribution of item-level Wald test p-values across all datasets, by model. Red dashed line marks α = 0.05."
#| fig-height: 4

items_df |>
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = 0.05, boundary = 0,
                 fill = "#2166ac", colour = "white", alpha = 0.75) +
  geom_vline(xintercept = 0.05, linetype = "dashed", colour = "#d6604d") +
  facet_wrap(~model) +
  labs(x = "p-value", y = "Count") +
  theme_minimal(base_size = 13)
```

---

## What to notice

**DIF is common.** Roughly 1 in 4 items shows significant uniform DIF by gender in this
sample (25% under 1PL, 22% under 2PL) — a higher rate than many practitioners assume.
The IRW's breadth means we are not cherry-picking instruments known to be problematic;
this is a cross-domain baseline.

**Sample size drives detections.** Multiple ECPS datasets (`ecps_sahm_2024_ia`,
`ecps_sahm_2024_trust`, `ecps_sahm_2024_emotion`, etc.) flag every item — the original
samples run into the tens of thousands, and even after sampling to 5,000 respondents the
power is enough to reject the null for every item. Researchers should pair DIF detection
with an effect-size measure (e.g., the Mantel-Haenszel Δ or the magnitude of the freed
intercept difference) before concluding that flagged items are *practically* problematic.

**1PL vs. 2PL discrepancies signal non-uniform DIF.** The largest disagreements reach
4 items in several datasets (e.g., `ccapsvtskhpacr_mercedes_2023_sf`: 4 flagged under
1PL, 0 under 2PL; `ccapsvtskhpacr_mercedes_2023_beck`: 6 vs. 2;
`nas_rogoza_2024_study1`: 12 vs. 8). When the 2PL is free to estimate different slopes
per group, the intercept test no longer rejects — a signature of non-uniform (crossing)
DIF rather than a uniform shift, and a reminder that the choice of model is not innocuous.

**Limitations.** Analyses here use uncorrected α = 0.05 across all items; a
Bonferroni or Benjamini-Hochberg correction would reduce the number of flagged items,
particularly in large datasets. Polytomous items use GPCM/graded response models, which
fit more parameters and may behave differently from the dichotomous items in the same
analysis.

---

## Extending this vignette

- **Effect sizes.** Add the freed-intercept difference as an effect-size measure alongside
  the Wald p-value to separate statistical from practical significance.

- **Multiple-testing correction.** Apply BH correction within each dataset and re-examine
  prevalence rates.

- **Scale-up.** Run `gender_dif_compute.R` without the `N_PROTO` cap to process all 459
  candidate datasets and produce an IRW-wide baseline.

- **Purification.** Re-estimate the baseline model excluding flagged items (anchor
  purification) to reduce contamination of the ability scale used to define DIF.

---

## Reproducibility

Results were computed on 2026-05-28 using `r nrow(ds_summary)` IRW tables. To reproduce:

```r
# From project root
Rscript vignettes/gender_dif_compute.R
quarto::quarto_render("vignettes/gender_dif.qmd")
```

::: {.callout-note}
The compute script caches each dataset in `vignettes/gender_dif_data/`. Delete the
per-dataset `.rds` files to force a fresh fetch; delete `combined_results.rds` to
re-combine from scratch.
:::
