---
title: "Guessing in Multiple-Choice Data: Which Correction Helps Out-of-Sample?"
description: |
  When guessing is present in multiple-choice data, which correction actually
  helps out-of-sample -- and when does the correction itself become a
  liability? Three published lenses on guessing (ability-based guessing,
  person-level purification, and a two-class mixture "safety valve") plus the
  standard IRT baselines, compared on real IRW tables using out-of-sample IMV.
date: today
bibliography: guessingdata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup

library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)

theme_set(theme_minimal(base_size = 13))

# Colors for the widget's 5 ICCs (Rasch/1PLg/3PL/1PL-AG/Mixture) and for the
# Results section's 6 non-Rasch models (adds 2PL and Purified Rasch, which
# don't appear as separate curves in the widget).
model_colors <- c(
  Rasch    = "#2a78d6",
  "2PL"    = "#6c757d",
  "1PLg"   = "#eb6834",
  "3PL"    = "#1baf7a",
  "1PL-AG" = "#e87ba4",
  Mixture  = "#4a3aa7",
  "Purified Rasch" = "#c9a227"
)
model_order <- c("2PL", "1PLg", "3PL", "1PL-AG", "Mixture", "Purified Rasch")

results_path      <- "guessingdata/guessing_results.rds"
provisional_path  <- "guessingdata/guessing_results_provisional.rds"
cache_ok <- file.exists(results_path) || file.exists(provisional_path)
is_provisional <- !file.exists(results_path) && file.exists(provisional_path)

if (cache_ok) {
  .res       <- readRDS(if (file.exists(results_path)) results_path else provisional_path)
  summary_df <- .res$summary
  tables_df  <- .res$tables
  date_run   <- .res$date_run
  holdout_frac <- .res$holdout_frac
  subsample_n  <- .res$subsample_n
  n_complete_provisional <- .res$n_complete %||% NA_integer_
  n_total_planned        <- .res$n_total_planned %||% NA_integer_

  # All IMVs reframed relative to Rasch ("1PL"), the common no-guessing
  # baseline, so every model's predictive gain (or cost) is on one shared
  # scale -- computed directly from each table's cached held-out
  # predictions, no refitting involved.
  model_keys <- c(pl2 = "2PL", plg = "1PLg", pl3 = "3PL", ag = "1PL-AG",
                   mix = "Mixture", purified = "Purified Rasch")
  if (!is.null(.res$raw)) {
    source("guessing_helpers.R")
    imv_vs_rasch_df <- map_dfr(.res$raw, function(r) {
      y <- r$true_vals
      p_rasch <- r$preds$rasch
      map_dfr(names(model_keys), function(key) {
        p <- r$preds[[key]]
        imv_val <- if (is.null(p) || all(is.na(p))) NA_real_ else compute_imv(p_rasch, p, y)
        tibble(table = r$table, model = model_keys[[key]], imv = imv_val)
      })
    }) |> mutate(model = factor(model, levels = model_order))

    # Full pairwise IMV(m1, m2) for every ordered pair of the 7 specs (IMV is
    # directional, so m1->m2 need not equal -m2->m1), per table -- computed
    # once here from the same cached held-out predictions, then embedded as
    # JSON for the user-selectable comparison widget below (no server-side
    # computation needed at view time, and the page stays a static file).
    all_model_keys <- c(rasch = "Rasch", pl2 = "2PL", plg = "1PLg", pl3 = "3PL",
                         ag = "1PL-AG", mix = "Mixture", purified = "Purified Rasch")
    pairwise_imv <- map(.res$raw, function(r) {
      y <- r$true_vals
      out <- list()
      for (k1 in names(all_model_keys)) {
        p1 <- r$preds[[k1]]
        if (is.null(p1) || all(is.na(p1))) next
        row <- list()
        for (k2 in names(all_model_keys)) {
          if (k1 == k2) next
          p2 <- r$preds[[k2]]
          row[[all_model_keys[[k2]]]] <- if (is.null(p2) || all(is.na(p2))) NA_real_ else compute_imv(p1, p2, y)
        }
        out[[all_model_keys[[k1]]]] <- row
      }
      out
    })
    names(pairwise_imv) <- map_chr(.res$raw, "table")
    pairwise_imv_json <- jsonlite::toJSON(pairwise_imv, auto_unbox = TRUE, na = "null", digits = 5)
  } else {
    imv_vs_rasch_df <- tibble(table = character(), model = factor(character(), levels = model_order), imv = numeric())
    pairwise_imv_json <- jsonlite::toJSON(list())
  }
} else {
  summary_df <- tibble(
    table = character(), m = integer(), m_verified = logical(),
    n_items = integer(), n_participants = integer(), n_heldout = integer(),
    converged_rasch = logical(), converged_2pl = logical(),
    converged_1plg = logical(), converged_3pl = logical(),
    converged_ag = logical(), converged_mix = logical(),
    pi_hat = numeric(), alpha_hat = numeric(), se_alpha = numeric(),
    lr_stat = numeric(), lr_p = numeric(), frac_flagged = numeric()
  )
  imv_vs_rasch_df <- tibble(table = character(), model = factor(character(), levels = model_order), imv = numeric())
  pairwise_imv_json <- jsonlite::toJSON(list())
  date_run <- Sys.Date(); holdout_frac <- 0.2; subsample_n <- 3000
  is_provisional <- FALSE
}
```

```{r}
#| label: provisional-banner
#| results: asis
if (cache_ok && isTRUE(is_provisional)) {
  cat(sprintf(
    "\n::: {.callout-warning}\n**Provisional preview.** %d of %d planned tables have finished fitting so far; this page reflects those %d only and will change once the remaining tables complete.\n:::\n",
    n_complete_provisional, n_total_planned, n_complete_provisional
  ))
}
```

## Motivation

Multiple-choice items let respondents get a question right without knowing
the answer. Three literatures locate that guessing behavior in three
different places, and each proposes a different fix:

- **@sanmartin2006** locate it in a **person &times; item interaction**: the
  1PL-AG model lets the probability of a correct guess itself depend on
  ability, via a weight $\alpha$ on top of a per-item floor. When $\alpha=0$
  the model reduces to a fixed per-item floor (1PL-G); $\alpha \ne 0$ means
  more able examinees guess more successfully -- partial knowledge leaking
  into the "guess" itself.
- **@torresirribarra2026** locate it at the **person level, in the
  calibration sample**: rather than adding a guessing parameter to the item
  response function, their Method A flags examinees whose response pattern
  looks more like a flat guessing process than a Rasch process, removes them
  from calibration, and refits. No new item parameter is introduced; the
  item parameters themselves are what improve.
- **@xiao2026safetyvalve** (the "Safety Valve" paper) locate it in a
  **discrete latent class**: a two-class mixture in which one class responds
  according to a Rasch model and the other responds at a fixed chance level,
  with the mixing proportion $\pi$ estimated from the data. As
  $\hat\pi \to 1$ the model collapses to a plain Rasch model, so (per their
  central finding) it costs almost nothing when guessing is absent, while a
  fixed-floor 1PLg can pay a large, asymmetric predictive penalty when its
  assumed floor doesn't match the population.

None of these three papers puts all three lenses, plus the standard
Rasch/2PL/1PLg/3PL baselines, on one predictive scale across a battery of
real datasets. That's the contribution here: fit all seven specifications on
real IRW tables and score every one of them on genuinely held-out responses,
using the InterModel Vigorish [IMV; @domingue2025imv].

## Explore the model shapes

Rasch and 1PLg apply the same floor to every examinee, regardless of whether
they're guessing or working the problem. 1PL-AG tilts that floor by ability.
The Mixture model only applies the floor to the estimated share of guessers.
Drag the sliders below to see how the five item characteristic curves (ICCs)
relate to one another.

```{=html}
<div class="guess-widget" style="border:1px solid #dee2e6;border-radius:6px;padding:1rem 1.25rem;margin:1rem 0;">
  <div style="display:grid;grid-template-columns:repeat(2,1fr);gap:12px 24px;margin-bottom:0.75rem;">
    <label style="font-size:0.9rem;">Item difficulty (b)
      <span style="font-size:0.72rem;font-weight:600;color:#495057;">-- all five curves</span>
      <input id="guess-b" type="range" min="-1" max="3" step="0.1" value="1.5" style="width:100%;">
      <span id="guess-b-val" style="font-family:monospace;font-size:0.85rem;"></span>
    </label>
    <label style="font-size:0.9rem;">Chance level (g, for an m-option item)
      <span style="font-size:0.72rem;font-weight:600;color:#495057;">-- all five curves</span>
      <input id="guess-g" type="range" min="0.1" max="0.5" step="0.05" value="0.25" style="width:100%;">
      <span id="guess-g-val" style="font-family:monospace;font-size:0.85rem;"></span>
    </label>
    <label style="font-size:0.9rem;">Ability weight in guessing (1PL-AG &alpha;)
      <span style="font-size:0.72rem;font-weight:600;color:#e87ba4;">-- 1PL-AG only (pink curve)</span>
      <input id="guess-alpha" type="range" min="0" max="1" step="0.05" value="0.2" style="width:100%;">
      <span id="guess-alpha-val" style="font-family:monospace;font-size:0.85rem;"></span>
    </label>
    <label style="font-size:0.9rem;">Engaged proportion (Mixture &pi;)
      <span style="font-size:0.72rem;font-weight:600;color:#4a3aa7;">-- Mixture only (purple curve)</span>
      <input id="guess-pi" type="range" min="0.5" max="1" step="0.01" value="0.85" style="width:100%;">
      <span id="guess-pi-val" style="font-family:monospace;font-size:0.85rem;"></span>
    </label>
  </div>
  <p style="font-size:0.78rem;color:#666;margin:0 0 0.5rem;">
    3PL (green) always uses discrimination fixed at 1.5 (rather than a slider)
    purely so its curve is visually distinguishable from 1PLg's at the same
    <em>b</em> and <em>g</em> -- it is not meant to represent a specific
    fitted discrimination value.
  </p>
  <div style="display:flex;flex-wrap:wrap;gap:14px;margin-bottom:6px;font-size:0.82rem;">
    <span><span style="display:inline-block;width:10px;height:10px;border-radius:2px;background:#2a78d6;"></span> Rasch</span>
    <span><span style="display:inline-block;width:10px;height:10px;border-radius:2px;background:#eb6834;"></span> 1PLg</span>
    <span><span style="display:inline-block;width:10px;height:10px;border-radius:2px;background:#1baf7a;"></span> 3PL</span>
    <span><span style="display:inline-block;width:10px;height:10px;border-radius:2px;background:#e87ba4;"></span> 1PL-AG</span>
    <span><span style="display:inline-block;width:10px;height:10px;border-radius:2px;background:#4a3aa7;"></span> Mixture</span>
  </div>
  <svg id="guess-svg" viewBox="0 0 640 380" style="width:100%;max-width:640px;height:auto;display:block;"></svg>
  <p style="font-size:0.85rem;color:#666;margin-top:0.5rem;">
    Rasch and 1PLg apply the same floor to every examinee. 1PL-AG tilts the
    floor by ability. The Mixture model only applies the floor to the
    estimated share of guessers.
  </p>
</div>

<script>
(function() {
  function plogis(x) { return 1 / (1 + Math.exp(-x)); }
  function qlogis(p) { return Math.log(p / (1 - p)); }

  var ids = ["b", "g", "alpha", "pi"];
  var els = {}, valEls = {};
  ids.forEach(function(id) {
    els[id] = document.getElementById("guess-" + id);
    valEls[id] = document.getElementById("guess-" + id + "-val");
  });
  var svg = document.getElementById("guess-svg");

  var W = 640, H = 380, mL = 50, mR = 68, mT = 10, mB = 40;
  var plotW = W - mL - mR, plotH = H - mT - mB;
  function px(theta) { return mL + (theta + 4) / 8 * plotW; }
  function py(p) { return mT + (1 - p) * plotH; }

  function pathFor(fn) {
    var d = "";
    for (var i = 0; i <= 160; i++) {
      var theta = -4 + i * (8 / 160);
      var x = px(theta), y = py(fn(theta));
      d += (i === 0 ? "M" : "L") + x.toFixed(2) + "," + y.toFixed(2) + " ";
    }
    return d;
  }

  function staticAxesSvg() {
    var s = "";
    s += '<line x1="' + mL + '" y1="' + mT + '" x2="' + mL + '" y2="' + (mT + plotH) + '" stroke="#888"/>';
    s += '<line x1="' + mL + '" y1="' + (mT + plotH) + '" x2="' + (mL + plotW) + '" y2="' + (mT + plotH) + '" stroke="#888"/>';
    [-4, -2, 0, 2, 4].forEach(function(t) {
      var x = px(t);
      s += '<line x1="' + x + '" y1="' + (mT + plotH) + '" x2="' + x + '" y2="' + (mT + plotH + 5) + '" stroke="#888"/>';
      s += '<text x="' + x + '" y="' + (mT + plotH + 18) + '" font-size="11" text-anchor="middle" fill="#333">' + t + '</text>';
    });
    [0, 0.25, 0.5, 0.75, 1].forEach(function(p) {
      var y = py(p);
      s += '<line x1="' + (mL - 5) + '" y1="' + y + '" x2="' + mL + '" y2="' + y + '" stroke="#888"/>';
      s += '<text x="' + (mL - 9) + '" y="' + (y + 4) + '" font-size="11" text-anchor="end" fill="#333">' + p + '</text>';
    });
    s += '<text x="' + (mL + plotW / 2) + '" y="' + (H - 4) + '" font-size="12" text-anchor="middle" fill="#333">&#952; (ability)</text>';
    s += '<text x="12" y="' + (mT + plotH / 2) + '" font-size="12" text-anchor="middle" fill="#333" transform="rotate(-90 12 ' + (mT + plotH / 2) + ')">P(correct)</text>';
    return s;
  }

  function draw() {
    var b = parseFloat(els.b.value), g = parseFloat(els.g.value),
        alpha = parseFloat(els.alpha.value), pi = parseFloat(els.pi.value);
    valEls.b.textContent = "b = " + b.toFixed(1);
    valEls.g.textContent = "g = " + g.toFixed(2);
    valEls.alpha.textContent = "α = " + alpha.toFixed(2);
    valEls.pi.textContent = "π = " + pi.toFixed(2);
    var gamma = qlogis(g);

    var rasch    = function(t) { return plogis(t - b); };
    var oneplg   = function(t) { return g + (1 - g) * plogis(t - b); };
    var threepl  = function(t) { return g + (1 - g) * plogis(1.5 * (t - b)); };
    var ag       = function(t) { var p = plogis(t - b); return p + (1 - p) * plogis(alpha * t + gamma); };
    var mixture  = function(t) { return pi * plogis(t - b) + (1 - pi) * g; };

    var curves = [
      { name: "Rasch",   fn: rasch,   color: "#2a78d6" },
      { name: "1PLg",    fn: oneplg,  color: "#eb6834" },
      { name: "3PL",     fn: threepl, color: "#1baf7a" },
      { name: "1PL-AG",  fn: ag,      color: "#e87ba4" },
      { name: "Mixture", fn: mixture, color: "#4a3aa7" }
    ];

    var html = staticAxesSvg();
    curves.forEach(function(c) {
      html += '<path d="' + pathFor(c.fn) + '" fill="none" stroke="' + c.color + '" stroke-width="2.5"/>';
    });

    // Direct end-of-line labels at theta=4 (right edge), so the text sits
    // level with the curve it names. Curves often converge near P=1 at high
    // theta, so labels are decluttered with a minimum vertical gap, sorted
    // by their true endpoint height (closest match to the curve retained).
    var labels = curves.map(function(c) {
      return { name: c.name, color: c.color, y: py(c.fn(4)) };
    });
    labels.sort(function(a, b) { return a.y - b.y; });
    var minGap = 13;
    for (var li = 1; li < labels.length; li++) {
      if (labels[li].y - labels[li - 1].y < minGap) {
        labels[li].y = labels[li - 1].y + minGap;
      }
    }
    labels.forEach(function(lab) {
      lab.y = Math.max(mT + 8, Math.min(mT + plotH, lab.y));
      html += '<text x="' + (mL + plotW + 6) + '" y="' + (lab.y + 4) +
              '" font-size="11" font-weight="600" fill="' + lab.color + '">' + lab.name + '</text>';
    });

    svg.innerHTML = html;
  }

  ids.forEach(function(id) { els[id].addEventListener("input", draw); });
  draw();
})();
</script>
```

## Data and methods

### Data selection

Guessing corrections only make sense for genuine single-answer,
multiple-choice items with a well-defined chance level -- so the candidate
pool needed to satisfy several requirements simultaneously: dichotomous
(correct/incorrect) responses, `construct_type == "Cognitive/educational"`
(guessing is a multiple-choice-test phenomenon, not a Likert/personality-item
one), sufficient sample size for the two custom estimators below to be
stable, and -- critically -- a verifiable, homogeneous number of response
options $m$ per table, since every model here needs a chance level $1/m$.

A first pass required IRW item text (to count response options directly)
plus `construct_type`/`n_categories`/`n_participants` metadata filters. That
produced 26 metadata-plausible candidates, but manual inspection of the
actual item text excluded most of them: some "options" columns turned out to
be scoring-rubric labels (Correct/Incorrect/Self-correct) rather than
response alternatives an examinee chooses from; some items were "choose
**two** words" formats with a non-standard chance structure; one was an
Agree/Disagree/Not-Sure belief inventory rather than a distractor-based
knowledge test. Only three tables survived as genuine, item-text-verified,
single-answer MC with a homogeneous option count: `gilbert_meta_1` (m=4),
`gilbert_meta_102` (m=5), and `gilbert_meta_103` (m=5), all from the same
underlying literacy-intervention study.

That pool alone was thin and not very representative of the high-stakes
settings where guessing corrections matter most. ENEM (Brazil's national
university-admissions exam) carries no item text in IRW, so its response
options can't be verified from IRW data directly -- but ENEM's format is
externally documented as a stable 5-option (A-E) multiple choice test across
subjects and years. We use that documented value ($m=5$) for 8 ENEM
tables spanning 2013, 2014, 2019, and 2024 and all four subject areas
(`mt`=math, `lc`=languages, `ch`=human sciences, `cn`=natural sciences),
restricted to the density=1 table/years (every respondent answered the
identical fixed item set, no rotated-booklet missingness). Each ENEM table
has roughly a million respondents; we work with a fixed-seed random
subsample of `r subsample_n` per table for compute tractability. **This is
an assumed, not IRW-verified, value of $m$** -- flagged explicitly rather
than silently treated as equivalent to the item-text-verified tables.

Per **@xiao2026safetyvalve** (Section 3.2), two-alternative forced-choice
(2AFC) items are excluded from this design on principle, not just by
happenstance: at $g_{\text{fit}}=0.5$ the Rasch model already predicts
roughly 0.5 on sufficiently difficult items, so the Mixture model's two
classes are not empirically separable -- the guessing class predicts a flat
0.5 that the engaged class already produces for low-ability respondents on
hard items. All tables used here have $m \ge 4$, safely inside the
class-separated region the Safety Valve paper establishes.

```{r}
#| label: table-list
if (cache_ok) {
  tables_df |>
    mutate(source = ifelse(m_verified, "item-text-verified", "documented (ENEM)")) |>
    select(table, m, source) |>
    knitr::kable(col.names = c("Table", "m", "Source"), align = "lrl",
                 caption = "Tables used")
}
```

### Models fit

Seven specifications per table:

1. **Rasch** -- no guessing correction.
2. **2PL** -- `mirt`, lognormal(0,1) prior on $a_1$ (this site's existing
   convention; see `imv.qmd`).
3. **1PLg** -- fixed floor $g=1/m$, discrimination fixed at 1, difficulty
   free (built via a constrained 3PL in `mirt`).
4. **3PL** -- `mirt`, lognormal(0,1) prior on $a_1$ (same as 2PL, chosen for
   internal consistency rather than replicating the Safety Valve paper's
   tighter lnorm(0,0.5)) and Beta(5,17) on $g$ (mean $\approx 0.23$), per
   the Safety Valve paper's Study 5.
5. **1PL-AG** [@sanmartin2006] -- standalone marginal-likelihood estimator
   (`vignettes/guessing_helpers.R::fit_1pl_ag()`), quadrature-integrated,
   with an analytic gradient verified against finite differences before use.
   Reports the likelihood-ratio test of $\alpha=0$.
6. **Mixture** ("safety valve") [@xiao2026safetyvalve] -- standalone
   estimator (`fit_mixture()`), multi-start over
   $\pi_0 \in \{0.6, 0.75, 0.9, 0.97\}$, gradient likewise verified.
7. **Purified Rasch** (Method A) [@torresirribarra2026] -- flags examinees
   whose response pattern favors a flat-guessing likelihood over the Rasch
   likelihood, removes them, refits Rasch on the retained sample
   (`purify_rasch()`). Its held-out predictions are evaluated on the full
   sample (including flagged persons), since the promise of purification is
   better item parameters, not that flagged persons are themselves predicted
   well.

Before use on real data, `fit_1pl_ag()` and `fit_mixture()` were validated
by simulation: both correctly return a non-significant likelihood-ratio test
/ $\hat\pi \approx 1$ under a true null (no guessing), and both recover
known parameters reasonably well under a true alternative (Mixture recovery
was excellent; 1PL-AG's $\beta$ recovered well but its $\gamma$ was noisy at
moderate sample sizes -- a limitation the source paper documents for its own
real-data application, not an implementation defect. See Limitations below.

### Cross-validation

A single 80/20 response-level (cell) holdout per table, masked once and
reused across all seven model fits, per this codebase's existing convention
([[feedback_cv_holdout]]; @stenhaug2022) -- masking is per-person and always
leaves at least one retained response. This is a **single split**, not the
20 repeated splits used in the Safety Valve paper's own empirical section;
that simplification was made deliberately for this first run.

## Results

All predictive comparisons below are framed relative to a single common
baseline -- the Rasch model, i.e. "1PL", the no-guessing reference every
other specification is trying to improve on. Rather than the scattered
pairwise comparisons a first pass might reach for (1PLg vs. Mixture, 3PL vs.
1PL-AG, and so on), every model's IMV is computed the same way,
`compute_imv(rasch_predictions, model_predictions, y)`, on the same
held-out cells. That puts all six non-Rasch specifications for a table on
one shared scale: positive means "beats plain Rasch out of sample," negative
means "costs something relative to just ignoring guessing entirely."

```{r}
#| label: results-table
if (cache_ok && nrow(summary_df) > 0) {
  summary_df |>
    mutate(
      alpha_lbl = sprintf("%.3f (p=%.3f)", alpha_hat, lr_p),
      pi_lbl    = sprintf("%.3f", pi_hat),
      frac_flagged = round(frac_flagged, 3)
    ) |>
    select(table, m, n_participants, n_items, n_heldout,
           pi_lbl, alpha_lbl, frac_flagged) |>
    knitr::kable(
      align = "lrrrrrl",
      col.names = c("Table", "m", "N", "J", "N held-out",
                    "π̂ (Mixture)", "α̂ (1PL-AG, p)", "Frac. flagged (Purified)"),
      caption = "Per-table model-internal diagnostics"
    )
} else {
  cat("Results not yet computed -- run vignettes/guessing_compute.R.")
}
```

```{r}
#| label: imv-vs-rasch-table
if (cache_ok && nrow(imv_vs_rasch_df) > 0) {
  imv_vs_rasch_df |>
    mutate(imv = round(imv, 4)) |>
    pivot_wider(names_from = model, values_from = imv) |>
    select(table, all_of(model_order)) |>
    knitr::kable(
      align = "lrrrrrr",
      caption = paste0("IMV(Rasch, model) for every model, on held-out cells ",
                        "(single ", holdout_frac * 100, "% split per table)")
    )
}
```

The table above fixes the baseline at Rasch. The widget below lets you pick
*any* two of the seven specifications and see `IMV(m1, m2)` -- m1 as
baseline, m2 as the model being evaluated -- across every table, without
re-fitting anything (every pairwise value was precomputed from the same
cached held-out predictions used everywhere else on this page). IMV is
directional, so `IMV(Rasch, 3PL)` and `IMV(3PL, Rasch)` are not
mirror images of each other -- pick both orders if you want to see that
asymmetry directly.

```{r}
#| label: pairwise-imv-widget
#| results: asis
if (cache_ok) {
  model_names_js <- jsonlite::toJSON(names(model_colors), auto_unbox = FALSE)
  widget_html <- sprintf('
<div class="pairwise-widget" style="border:1px solid #dee2e6;border-radius:6px;padding:1rem 1.25rem;margin:1rem 0;">
<div style="display:flex;flex-wrap:wrap;gap:1.25rem;align-items:center;margin-bottom:0.75rem;">
<label style="font-weight:600;font-size:0.9rem;">Baseline (m1):
<select id="pw-m1" style="margin-left:0.4rem;"></select>
</label>
<label style="font-weight:600;font-size:0.9rem;">Compared model (m2):
<select id="pw-m2" style="margin-left:0.4rem;"></select>
</label>
</div>
<svg id="pw-svg" viewBox="0 0 700 440" style="width:100%%;max-width:700px;height:auto;display:block;"></svg>
</div>

<script>
(function() {
var pairwiseData = %s;
var modelNames = %s;
var tableNames = Object.keys(pairwiseData);

var m1Sel = document.getElementById("pw-m1");
var m2Sel = document.getElementById("pw-m2");
var svg = document.getElementById("pw-svg");

modelNames.forEach(function(name) {
var o1 = document.createElement("option"); o1.value = name; o1.textContent = name;
m1Sel.appendChild(o1);
var o2 = document.createElement("option"); o2.value = name; o2.textContent = name;
m2Sel.appendChild(o2);
});
m1Sel.value = "Rasch";
m2Sel.value = "Mixture";

var W = 700, H = 440, mL = 170, mR = 20, mT = 10, mB = 50;
var plotW = W - mL - mR, plotH = H - mT - mB;

function niceStep(maxAbs) {
var raw = maxAbs / 3;
var mag = Math.pow(10, Math.floor(Math.log(raw) / Math.LN10));
var norm = raw / mag;
var mult = norm < 1.5 ? 1 : (norm < 3.5 ? 2 : (norm < 7.5 ? 5 : 10));
return mult * mag;
}

function fmtTick(v, step) {
var decimals = Math.max(0, -Math.floor(Math.log(step) / Math.LN10) + 1);
return v.toFixed(decimals);
}

function draw() {
var m1 = m1Sel.value, m2 = m2Sel.value;

var rows = tableNames.map(function(t) {
var v = (pairwiseData[t][m1] || {})[m2];
return { table: t, imv: (v === undefined || v === null) ? null : v };
});

if (m1 === m2) {
svg.innerHTML = "<text x=\\"20\\" y=\\"40\\" font-size=\\"13\\" fill=\\"#666\\">Pick two different models.</text>";
return;
}

// Fixed axis range (not rescaled per selected pair), so bar lengths are
// directly comparable across different m1/m2 choices. Values beyond the
// range are clipped to the panel edge but their true number is still shown
// as text rather than being cut off or silently rescaling the axis.
var AXIS_MAX = 0.015;
var rowH = plotH / rows.length;

function px(v) { return mL + plotW / 2 + (v / AXIS_MAX) * (plotW / 2); }
function clampX(x) { return Math.max(mL, Math.min(mL + plotW, x)); }

var html = "";
var axisY = mT + plotH;
html += "<line x1=\\"" + mL + "\\" y1=\\"" + axisY + "\\" x2=\\"" + (mL + plotW) + "\\" y2=\\"" + axisY + "\\" stroke=\\"#888\\"/>";

var step = niceStep(AXIS_MAX);
for (var tv = 0; tv <= AXIS_MAX + step / 2; tv += step) {
[tv, -tv].forEach(function(v) {
if (v !== 0 && Math.abs(v) < step / 2) return;
var x = px(v);
html += "<line x1=\\"" + x + "\\" y1=\\"" + mT + "\\" x2=\\"" + x + "\\" y2=\\"" + axisY +
"\\" stroke=\\"" + (v === 0 ? "#888" : "#e5e5e5") + "\\" stroke-width=\\"" + (v === 0 ? "1.5" : "1") + "\\"/>";
html += "<line x1=\\"" + x + "\\" y1=\\"" + axisY + "\\" x2=\\"" + x + "\\" y2=\\"" + (axisY + 5) + "\\" stroke=\\"#888\\"/>";
html += "<text x=\\"" + x + "\\" y=\\"" + (axisY + 18) + "\\" font-size=\\"10\\" text-anchor=\\"middle\\" fill=\\"#333\\">" + fmtTick(v, step) + "</text>";
});
}

rows.forEach(function(r, i) {
var yTop = mT + i * rowH + rowH * 0.15;
var barH = rowH * 0.7;
html += "<text x=\\"" + (mL - 8) + "\\" y=\\"" + (yTop + barH / 2 + 4) +
"\\" font-size=\\"11\\" text-anchor=\\"end\\" fill=\\"#333\\">" + r.table + "</text>";
if (r.imv === null) {
html += "<text x=\\"" + (mL + 4) + "\\" y=\\"" + (yTop + barH / 2 + 4) +
"\\" font-size=\\"11\\" fill=\\"#aaa\\">not available</text>";
return;
}
var x1 = px(0), x2Raw = px(r.imv), x2 = clampX(x2Raw);
var clipped = x2 !== x2Raw;
var barX = Math.min(x1, x2), barW = Math.abs(x2 - x1);
var color = r.imv >= 0 ? "#1baf7a" : "#d1495b";
html += "<rect x=\\"" + barX + "\\" y=\\"" + yTop + "\\" width=\\"" + Math.max(barW, 0.5) +
"\\" height=\\"" + barH + "\\" fill=\\"" + color + "\\"" + (clipped ? " opacity=\\"0.6\\"" : "") + "/>";
var lblX = r.imv >= 0 ? x2 + 4 : x2 - 4;
var lblAnchor = r.imv >= 0 ? "start" : "end";
if (clipped) { lblX = x2 - (r.imv >= 0 ? 4 : -4); lblAnchor = r.imv >= 0 ? "end" : "start"; }
html += "<text x=\\"" + lblX + "\\" y=\\"" + (yTop + barH / 2 + 4) + "\\" font-size=\\"10\\" fill=\\"#333\\" text-anchor=\\"" +
lblAnchor + "\\">" + r.imv.toFixed(4) + "</text>";
});

html += "<text x=\\"" + (mL + plotW / 2) + "\\" y=\\"" + (H - 6) +
"\\" font-size=\\"11\\" text-anchor=\\"middle\\" fill=\\"#333\\">IMV(" + m1 + ", " + m2 + ")</text>";
svg.innerHTML = html;
}

m1Sel.addEventListener("change", draw);
m2Sel.addEventListener("change", draw);
draw();
})();
</script>
', pairwise_imv_json, model_names_js)
  # results:asis inserts this text back into the Markdown stream, where
  # pandoc re-parses it as Markdown -- any line indented 4+ spaces reads as
  # an indented code block (swallowing the actual HTML as escaped text, so
  # the widget silently never renders). Strip leading whitespace from every
  # line so nothing in the generated HTML/JS can trigger that rule.
  cat(gsub("(?m)^[ \t]+", "", widget_html, perl = TRUE))
}
```

```{r}
#| label: highlight-picks
if (cache_ok && nrow(summary_df) > 0) {
  imv_wide <- imv_vs_rasch_df |> pivot_wider(names_from = model, values_from = imv)
  picks_df <- summary_df |> left_join(imv_wide, by = "table")

  no_guess_tbl <- picks_df |> arrange(desc(pi_hat)) |> slice(1)
  guess_tbl    <- picks_df |> arrange(pi_hat) |> slice(1)
  disagree_tbl <- picks_df |>
    filter(!is.na(Mixture), !is.na(`1PL-AG`)) |>
    mutate(disagree = abs(sign(Mixture) - sign(`1PL-AG`))) |>
    arrange(desc(disagree)) |> slice(1)
}
```

```{r}
#| label: what-to-notice
#| results: asis
if (cache_ok && nrow(summary_df) > 0) {
  cat("### What to notice\n\n")

  no_guess_lbl <- if (no_guess_tbl$pi_hat > 0.95) "Safety-valve-suppressed / least guessing detected" else "Least guessing detected (of tables run so far)"
  cat(sprintf(
    "**%s**: `%s` has \\u03c0\\u0302 = %.3f -- the Mixture model assigns the largest engaged share of the tables run so far, with IMV(Rasch,Mixture) of %.4f. %s\n\n",
    no_guess_lbl, no_guess_tbl$table, no_guess_tbl$pi_hat, no_guess_tbl$Mixture,
    if (no_guess_tbl$pi_hat > 0.95) "A near-zero IMV here is not a failure of the Mixture model; it is the safety-valve property working as intended -- there is nothing for it to correct." else ""
  ))

  cat(sprintf(
    "**Substantial guessing detected**: `%s` has \\u03c0\\u0302 = %.3f, meaning the Mixture model assigns roughly %.0f%% of response weight to the guessing class, with IMV(Rasch,Mixture) of %.4f.\n\n",
    guess_tbl$table, guess_tbl$pi_hat, 100 * (1 - guess_tbl$pi_hat), guess_tbl$Mixture
  ))

  if (nrow(disagree_tbl) > 0 && disagree_tbl$disagree > 0) {
    cat(sprintf(
      "**Where methods disagree**: `%s` shows the Mixture and 1PL-AG comparisons against Rasch pointing in different directions (IMV(Rasch,Mixture) = %.4f vs. IMV(Rasch,1PL-AG) = %.4f) -- a reminder that these two lenses model genuinely different mechanisms and won't always agree on which correction (if any) is worth its complexity.\n\n",
      disagree_tbl$table, disagree_tbl$Mixture, disagree_tbl$`1PL-AG`
    ))
  }
}
```

## Limitations

- **ENEM's $m=5$ is an assumed, not verified, value.** IRW carries no item
  text for these tables; the value comes from documented exam structure, not
  from counting options in the data itself. If that assumption is wrong for
  any particular year/subject, every ENEM result here inherits the error.
- **1PL-AG's $\gamma$ (and, to a lesser extent, $\alpha$) can be noisy.**
  Simulation validation recovered $\beta$ well but $\gamma$ poorly at
  moderate sample sizes -- consistent with @sanmartin2006's own real-data
  application, where some items had standard errors above 100. Point
  estimates of $\hat\alpha$ should be read alongside its LR test, not in
  isolation.
- **Single 80/20 split, not repeated cross-validation.** IMV estimates here
  reflect one holdout draw per table; they are not averaged over multiple
  splits the way the Safety Valve paper's own empirical section is.
- **Secondary (gilbert_meta_*) tables come from one study family.** All
  three item-text-verified tables share an underlying literacy-intervention
  study; they are not independent evidence of generalizability beyond that
  context.
- **Purified Rasch (Method A) assumes person-level, not item-level,
  guessing** -- an examinee is modeled as either guessing on everything or
  nothing, per @torresirribarra2026's own stated simplification relative to
  the (person &times; item) process that actually generates guessing.

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Fit all 7 specifications across all tables (from the project root)
source("vignettes/guessing_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/guessing.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(summary_df$table, output_file = "guessingdata/irw_references.bib")
```
:::

## References
