---
title: "The Reliability Paradox in Cognitive-Control Tasks"
description: |
  Hedge, Powell & Sumner (2018) showed that classic conflict tasks (Flanker,
  Stroop, stop-signal) produce large, dependable group effects but poor
  individual-differences reliability, because the within-subject difference
  score that measures the effect cancels out most of the between-person
  variance. We test whether the same collapse appears in IRW cognitive-control
  data: the Hearts & Flowers task, an alcohol Stroop, and a stop-signal task.
date: today
bibliography: hf_reliability/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "hf_reliability/hf_reliability_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache      <- readRDS(cache_path)
  step0       <- .cache$step0_summary
  results     <- .cache$results
  hedge_ref   <- .cache$hedge2018_reference
  n_splits    <- .cache$n_splits
  trial_floors <- .cache$trial_floors
  date_run    <- .cache$date_run
} else {
  step0    <- tibble(table = character())
  results  <- tibble(dataset = character(), task_type = character(), measure = character(),
                      reliability_type = character(), reliability_est = numeric(),
                      reliability_sd = numeric(), trial_floor = integer(),
                      n_respondents = integer(), mean_trials_congruent = numeric(),
                      mean_trials_incongruent = numeric(), var_congruent = numeric(),
                      var_incongruent = numeric(), cov_cong_incong = numeric(),
                      var_difference = numeric(), notes = character())
  hedge_ref <- tibble(task = character(), measure = character(), condition = character(),
                       icc_study1 = numeric(), icc_study2 = numeric(), icc_mean = numeric())
  n_splits <- NA_integer_
  trial_floors <- c(5L, 10L)
  date_run <- Sys.Date()
}

knitr::opts_chunk$set(eval = cache_ok)

dataset_labels <- c(
  imps2025_hf_blocked    = "Hearts & Flowers (blocked)",
  imps2025_hf_mixed      = "Hearts & Flowers (mixed-trial)",
  alcoholstroop_jones2024 = "Alcohol Stroop",
  matzke_stopsignal_2019 = "Stop-signal (OSARI)"
)

measure_labels <- c(rt = "Reaction time", acc = "Accuracy")
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/hf_reliability_compute.R\n```\n",
  ":::\n", sep = ""
)
```

## Motivation

Some cognitive tasks produce huge, dependable group-level effects yet turn
out to be nearly useless for studying individual differences. The clearest
examples are conflict-monitoring tasks: Flanker, Stroop, and their relatives.
Every study finds that people are slower and less accurate on incongruent
trials than congruent ones — the effect is close to universal. But
@hedge2018reliability showed that when you use these same tasks to *rank*
people (rather than just detect the group effect), reliability collapses:
individual conditions like congruent RT or incongruent RT are often reliable
enough on their own (test-retest ICCs around .6-.8), but the *difference
score* used to index the effect of interest — incongruent RT minus congruent
RT — is barely reliable at all.

The mechanism is a direct consequence of how difference scores propagate
variance. For two conditions $A$ and $B$ measured on the same people,

$$\mathrm{Var}(B - A) = \mathrm{Var}(A) + \mathrm{Var}(B) - 2\,\mathrm{Cov}(A, B)$$

Conflict tasks are robust precisely *because* almost everyone shows a similar
conflict cost — high scorers on congruent trials tend to also be high scorers
on incongruent trials (fast people are fast in both conditions, slow people
are slow in both). That shared individual differences signal is exactly what
$\mathrm{Cov}(A, B)$ captures, and subtracting it away leaves a difference
score with little between-person variance left, while the trial-level
measurement noise in each condition is *not* similarly reduced. The result:
low signal-to-noise for the individual-differences measure, even though the
group-level contrast ($\bar{B} - \bar{A}$) is large and rock solid.

This is a *structural* property of difference-score designs, not a defect
specific to Flanker or Stroop. Any task that measures an effect as a
within-subject contrast between two positively-correlated conditions is
vulnerable to it. The Hearts & Flowers (H&F) task [@imps2025_hf] — part of
the imps2025 datathon data in IRW — has exactly this structure: children
respond to a heart or flower stimulus, pressing the button on the *same*
side as the stimulus for hearts (congruent, low conflict) and the *opposite*
side for flowers (incongruent, high conflict). This is a spatial
stimulus-response compatibility manipulation in the same family as Simon and
Flanker effects, which makes it a natural candidate for replicating the
reliability paradox in IRW data. We also check two independent tasks with
the same congruent/incongruent structure: an alcohol-related
attentional-bias Stroop [@alcoholstroop_jones2024], and a stop-signal
(go/no-go family) task [@matzke-etal-2019].

## Methods

### Candidate dataset search

We screened IRW's table metadata for trial-level (not fixed-item-bank) tasks
with a response-time column and a plausible two-condition contrast, using a
keyword search over table names, dataset descriptions, and variable lists
(`heart`, `flower`, `flanker`, `stroop`, `simon`, `congruent`, `stop-signal`,
etc.), followed by direct inspection of each survivor's columns.

```{r}
#| label: tbl-step0
#| tbl-cap: "Candidate tables surviving the keyword screen, and why each was kept or excluded."

step0 %>%
  mutate(
    decision = case_when(
      table == "imps2025_hf" ~ "Kept (Hearts & Flowers, 2475 respondents, has wave variable)",
      table == "alcoholstroop_jones2024" ~ "Kept (156 respondents, single session)",
      grepl("safety_conflicts", table) ~ "Excluded: keyword false positive on \"conflicts\" (public-safety survey, no rt/condition structure)",
      TRUE ~ "Excluded"
    )
  ) %>%
  select(table, n_responses, n_participants, has_rt, decision) %>%
  knitr::kable(col.names = c("Table", "Responses", "Participants", "Has RT", "Decision"),
               format.args = list(big.mark = ","))
```

Two further candidates were located by widening the search to any table with
an `rt` column and inspecting item-name/column structure directly (not
capturable by a keyword match on task names): `wang2022_iat` (a fixed
20-item dichotomous bank despite its name — no trial-level RT, excluded) and
five `matzke-*` stop-signal tables (a `go_`/`stop_` item-prefix condition,
`cov_ssd` = stop-signal delay). Of those, `matzke-etal-2019` has 53
respondents and was kept; `matzke-experiment1-2022`,
`matzke-experiment3-standard`, and `matzke-experiment3-gamified-2022` have
7, 9, and 9 respondents respectively and were dropped for falling below our
50-respondent floor.

The final set has **four dataset variants** across three IRW tables:

- **Hearts & Flowers, blocked** (`imps2025_hf`): the `hearts test` block
  (congruent) vs. the `flowers test` block (incongruent) — a separate-block
  design.
- **Hearts & Flowers, mixed-trial** (`imps2025_hf`): within the `mixed test`
  block, trials are classified congruent/incongruent by stimulus shape
  (heart/flower) rather than by block — a trial-intermixed design, which is
  structurally closer to how Hedge et al.'s tasks are actually administered
  (their conditions are intermixed within blocks, not run in separate
  blocks; see their Fig. 1).
- **Alcohol Stroop** (`alcoholstroop_jones2024`): neutral word (congruent
  analogue) vs. alcohol-related word (incongruent analogue) — an
  attentional-bias variant of Stroop rather than a color-word conflict task,
  but with an identical per-trial rt/accuracy structure.
- **Stop-signal / OSARI** (`matzke-etal-2019`): go trials (congruent
  analogue) vs. stop-signal trials (incongruent analogue).

### Preprocessing

For each variant we dropped respondents with fewer than 5 trials in either
condition (sensitivity checked at a floor of 10; see
[Trial-count sensitivity]). Per person and condition we computed mean RT on
correct trials and mean accuracy, then the contrast

$$D_{rt} = \overline{RT}_{\text{incongruent}} - \overline{RT}_{\text{congruent}}, \qquad
  D_{acc} = \overline{acc}_{\text{congruent}} - \overline{acc}_{\text{incongruent}}$$

so that positive values of both indices reflect the "expected" direction of
a conflict cost.

**Stop-signal deviation.** By construction, a successfully inhibited
stop-signal trial has no response and hence no RT — there is no such thing
as "the RT of a correct stop trial." Hedge et al. instead estimate
Stop-Signal Reaction Time (SSRT) via a race-model calculation. We do not
implement the race model here; instead $D_{rt}$ for the stop-signal dataset
uses the RT of *failed* stop trials (a response did occur) as a naive
structural analogue, and $D_{acc}$ compares go-trial direction accuracy to
the stop-trial inhibition rate — two different kinds of accuracy. Treat the
stop-signal numbers below as a structural check only, not a validated
inhibition measure.

**Split-half reliability.** For each dataset/measure/condition we ran
`r ifelse(is.na(n_splits), 500, n_splits)` random split-halves: within each
person and condition, trials were randomly assigned to two halves, person
means were computed for each half, and the correlation between halves across
people was Spearman-Brown corrected ($r_{sb} = 2r/(1+r)$). The difference
score was split-half by using the same halves within each condition, so
$D_{\text{half 1}} = \overline{incong}_{\text{half 1}} - \overline{cong}_{\text{half 1}}$,
and correlating the two halves' difference scores. We report the mean and SD
of the Spearman-Brown-corrected reliability across splits, and (for
split-half only) restrict to each person's earliest available session, so
that pooling multiple sessions doesn't artificially inflate reliability for
respondents with repeat data.

**Test-retest.** Hearts & Flowers has a genuine wave variable, so for that
task alone we additionally computed real test-retest correlations across two
sessions (no splitting; raw condition means and $D$ from each full session,
correlated across sessions). Two wave pairs were available: `PLUS Fall` →
`PLUS Spring` and `2021-2022 Fall` → `2021-2022 Spring` (different cohorts).
**This is an important deviation from Hedge et al.**: their test-retest
interval is ~3 weeks in stable young adults, ours is roughly a school
semester (~6 months) in developing children aged ~5-11 — real developmental
change over that interval, not just measurement noise, will suppress our
test-retest correlations relative to theirs. The alcohol Stroop and
stop-signal datasets have no wave variable and support split-half only.

## Results

### Split-half reliability: conditions vs. difference scores

```{r}
#| label: fig-splithalf
#| fig-cap: "Split-half reliability (Spearman-Brown corrected; mean ± SD across splits) for each condition and the difference score, trial floor = 5. Difference-score reliability collapses relative to the conditions it's built from in three of the four dataset variants."
#| fig-height: 5

sh_plot_df <- results %>%
  filter(trial_floor == 5, !grepl("test-retest", reliability_type)) %>%
  mutate(
    dataset_lab = factor(dataset_labels[dataset], levels = rev(dataset_labels)),
    reliability_type = factor(reliability_type,
                               levels = c("congruent", "incongruent", "difference"),
                               labels = c("Congruent", "Incongruent", "Difference score")),
    measure_lab = measure_labels[measure]
  )

ggplot(sh_plot_df, aes(x = reliability_est, y = dataset_lab, colour = reliability_type)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_errorbarh(aes(xmin = reliability_est - reliability_sd, xmax = reliability_est + reliability_sd),
                  height = 0, position = position_dodge(width = 0.5)) +
  geom_point(size = 2.6, position = position_dodge(width = 0.5)) +
  facet_wrap(~measure_lab) +
  scale_colour_manual(values = c("Congruent" = irw_blue, "Incongruent" = irw_grey,
                                  "Difference score" = irw_red), name = NULL) +
  labs(x = "Split-half reliability (Spearman-Brown corrected)", y = NULL) +
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-splithalf
#| tbl-cap: "Split-half reliability by dataset, measure, and condition (trial floor = 5)."

results %>%
  filter(trial_floor == 5, !grepl("test-retest", reliability_type)) %>%
  transmute(
    Dataset = dataset_labels[dataset],
    Measure = measure_labels[measure],
    Condition = tools::toTitleCase(reliability_type),
    `N respondents` = n_respondents,
    `Reliability (mean ± SD)` = sprintf("%.2f ± %.2f", reliability_est, reliability_sd)
  ) %>%
  arrange(Dataset, Measure) %>%
  knitr::kable()
```

The pattern in @fig-splithalf reproduces the core finding of
@hedge2018reliability directly: congruent and incongruent RT reliabilities
sit at .78-.99 across all four dataset variants, but the RT difference-score
reliability drops to near zero for the mixed-trial Hearts & Flowers
(`r sprintf("%.2f", results$reliability_est[results$dataset=="imps2025_hf_mixed" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`),
the alcohol Stroop
(`r sprintf("%.2f", results$reliability_est[results$dataset=="alcoholstroop_jones2024" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`),
and drops substantially — though not to zero — for the blocked Hearts &
Flowers design
(`r sprintf("%.2f", results$reliability_est[results$dataset=="imps2025_hf_blocked" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`).

The blocked-vs-mixed-trial contrast within the same H&F task is informative:
Hedge et al.'s tasks intermix conditions within a block, and it's exactly
that design — the mixed-trial H&F variant — that shows the most severe
collapse. Separating congruent and incongruent trials into their own blocks
(as the blocked H&F variant does) leaves noticeably more difference-score
reliability intact, consistent with the variance-shrinkage mechanism: block
separation gives some room for per-block state (vigilance, strategy,
warm-up) to vary independently across conditions, which lowers
$\mathrm{Cov}(A,B)$ relative to $\mathrm{Var}(A)$ and $\mathrm{Var}(B)$ less
than a fully intermixed design does.

The stop-signal dataset is the exception: its difference-score reliability
stays high
(`r sprintf("%.2f", results$reliability_est[results$dataset=="matzke_stopsignal_2019" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`).
This is best read in light of the stop-signal deviation noted in Methods —
failed-stop RT and go RT are driven by different aspects of the race model
(a fast guess vs. the whole go-process distribution) and are less tightly
coupled across people than two congruency conditions of the same trial type,
so the variance-shrinkage mechanism has less to work with. This is a useful
negative case: the paradox is not automatic whenever you subtract two RTs,
it specifically depends on how correlated the two conditions are across
people.

### Where does the variance go?

```{r}
#| label: tbl-variance
#| tbl-cap: "Between-person variance components underlying the RT difference score (trial floor = 5). Var(D) = Var(incong) + Var(cong) − 2·Cov(cong, incong); a high covariance relative to the individual variances is what shrinks Var(D)."

results %>%
  filter(trial_floor == 5, measure == "rt", reliability_type == "difference") %>%
  transmute(
    Dataset = dataset_labels[dataset],
    `Var(congruent)` = round(var_congruent, 4),
    `Var(incongruent)` = round(var_incongruent, 4),
    `Cov(cong, incong)` = round(cov_cong_incong, 4),
    `Var(difference)` = round(var_difference, 4)
  ) %>%
  knitr::kable()
```

This table makes the mechanism from the Motivation section concrete: in
every dataset, $\mathrm{Cov}(\text{cong}, \text{incong})$ is a large fraction
of both individual variances, so subtracting $2\,\mathrm{Cov}$ removes most
of the variance that Var(congruent) and Var(incongruent) contribute,
leaving Var(difference) small relative to either component alone.

### True test-retest reliability (Hearts & Flowers only)

```{r}
#| label: fig-retest
#| fig-cap: "Split-half (single session) vs. true test-retest reliability for Hearts & Flowers RT. Test-retest spans roughly a school semester in developing children, not the ~3-week interval used by Hedge et al. in stable adults, so lower absolute values are expected."
#| fig-height: 4.5

retest_df <- results %>%
  filter(dataset %in% c("imps2025_hf_blocked", "imps2025_hf_mixed"), measure == "rt") %>%
  mutate(
    design = case_when(
      grepl("PLUS Fall -> PLUS Spring", reliability_type) ~ "Test-retest (PLUS cohort)",
      grepl("2021-2022 Fall -> 2021-2022 Spring", reliability_type) ~ "Test-retest (K cohort)",
      trial_floor == 5 ~ "Split-half (single session)",
      TRUE ~ NA_character_
    ),
    condition = case_when(
      grepl("^congruent", reliability_type) ~ "Congruent",
      grepl("^incongruent", reliability_type) ~ "Incongruent",
      grepl("^difference", reliability_type) ~ "Difference score",
      TRUE ~ NA_character_
    ),
    dataset_lab = dataset_labels[dataset]
  ) %>%
  filter(!is.na(design))

ggplot(retest_df, aes(x = reliability_est, y = condition, colour = design)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(size = 2.8, position = position_dodge(width = 0.5)) +
  facet_wrap(~dataset_lab) +
  scale_colour_manual(values = c("Split-half (single session)" = irw_blue,
                                  "Test-retest (PLUS cohort)" = irw_red,
                                  "Test-retest (K cohort)" = "#f4a582"), name = NULL) +
  labs(x = "Reliability", y = NULL,
       caption = "PLUS cohort: Fall -> Spring, one school year. K cohort: 2021-22 Fall -> Spring.") +
  theme(legend.position = "bottom")
```

Test-retest reliabilities are lower than split-half across the board, as
expected from the much longer, developmentally active retest interval. But
the *relative* pattern survives: the difference score's test-retest
reliability is consistently the lowest of the three, both in the blocked and
mixed-trial variants, indicating the paradox isn't an artifact of the
within-session split-half method — it also shows up when the same people are
tested again months later.

### Comparison to Hedge, Powell & Sumner (2018)

The table below places our estimates next to the point estimates reported in
Hedge et al.'s Table 1 (average of their Studies 1 and 2; two-way
random-effects ICC, absolute agreement, ~3-week test-retest in adults). Two
IRW rows are shown per comparable task: the split-half (single-session)
estimate, which is the most directly comparable design for the alcohol
Stroop and stop-signal datasets since they have no wave variable, and the
Hearts & Flowers test-retest estimate, which is the more conceptually
faithful comparison to Hedge et al.'s design despite the much longer
interval.

```{r}
#| label: tbl-hedge-comparison
#| tbl-cap: "IRW reliability estimates alongside Hedge et al. (2018) Table 1 point estimates (mean of their Studies 1 & 2). NA where a comparable estimate wasn't computed (e.g. no test-retest wave for a dataset)."

irw_split <- results %>%
  filter(trial_floor == 5, measure == "rt", !grepl("test-retest", reliability_type)) %>%
  transmute(pair_task = dataset_labels[dataset], source = "IRW, split-half",
            condition = reliability_type, reliability = reliability_est)

irw_retest <- results %>%
  filter(measure == "rt", grepl("test-retest: PLUS Fall", reliability_type)) %>%
  mutate(condition = sub(" \\(test-retest.*\\)", "", reliability_type)) %>%
  transmute(pair_task = dataset_labels[dataset], source = "IRW, test-retest (Fall->Spring)",
            condition, reliability = reliability_est)

hedge_tbl <- hedge_ref %>%
  filter(measure == "rt") %>%
  transmute(pair_task = task, source = "Hedge et al. (2018), test-retest",
            condition = condition, reliability = icc_mean)

pairing <- tribble(
  ~pair_task_irw,                    ~pair_task_hedge,
  "Hearts & Flowers (mixed-trial)",  "Flanker",
  "Hearts & Flowers (blocked)",      "Flanker",
  "Alcohol Stroop",                  "Stroop",
  "Stop-signal (OSARI)",             "Stop-signal"
)

comparison <- bind_rows(irw_split, irw_retest) %>%
  inner_join(pairing, by = c("pair_task" = "pair_task_irw")) %>%
  transmute(`Lab task analogue` = pair_task_hedge, `IRW dataset` = pair_task,
            Source = source, Condition = tools::toTitleCase(condition),
            Reliability = round(reliability, 2)) %>%
  bind_rows(
    hedge_tbl %>%
      transmute(`Lab task analogue` = pair_task, `IRW dataset` = "—",
                Source = source, Condition = tools::toTitleCase(condition),
                Reliability = round(reliability, 2))
  ) %>%
  arrange(`Lab task analogue`, Source, factor(Condition, levels = c("Congruent", "Incongruent", "Difference", "Go", "Ssrt")))

knitr::kable(comparison)
```

Two things stand out. First, the *direction* of the effect (condition
reliability >> difference-score reliability) matches Hedge et al. in every
task pair except the stop-signal analogue, for the reasons discussed above.
Second, our absolute IRW split-half numbers for individual conditions run
higher than Hedge et al.'s test-retest ICCs — expected, since split-half
reliability has no opportunity for between-session instability to erode it,
while ICC test-retest does.

### Trial-count sensitivity

```{r}
#| label: tbl-sensitivity
#| tbl-cap: "Split-half difference-score RT reliability at trial floors of 5 vs. 10 trials/person/condition."

results %>%
  filter(measure == "rt", reliability_type == "difference", !grepl("test-retest", reliability_type)) %>%
  transmute(Dataset = dataset_labels[dataset], `Trial floor` = trial_floor,
            `N respondents` = n_respondents,
            `Reliability (mean ± SD)` = sprintf("%.2f ± %.2f", reliability_est, reliability_sd)) %>%
  arrange(Dataset, `Trial floor`) %>%
  knitr::kable()
```

Raising the per-condition trial floor from 5 to 10 trims the sample (most
sharply for the mixed-trial H&F variant, which has fewer trials per
condition to begin with) but does not change the qualitative picture: the
difference score remains far less reliable than either condition alone.

## Discussion

The reliability paradox replicates in IRW cognitive-control data. Across
three independent tasks (Hearts & Flowers, alcohol Stroop, and — with an
important caveat — a stop-signal task) spanning both split-half and true
multi-month test-retest designs, individual congruent and incongruent
conditions are consistently far more reliable than the within-subject
difference score built from them. The mechanism is exactly the one
@hedge2018reliability describe: the two conditions are highly correlated
across people, so most of their between-person variance cancels in the
subtraction, while trial-level noise does not.

**This is a property of difference-score designs, not a general statement
about IRW data.** Most IRW tables are fixed-item-bank IRT designs, where
ability is estimated from many items measuring one construct rather than
subtracted between two conditions — the mechanism demonstrated here doesn't
apply to them, and this vignette should not be read as evidence that IRW
data are broadly unreliable for individual-differences work. It applies
specifically to trial-level tasks that measure an effect as a within-subject
contrast between two positively-correlated conditions, which in the current
IRW warehouse means Hearts & Flowers, the alcohol Stroop, and the
matzke-etal-2019 stop-signal task (plus, presumably, any future task with
the same congruent/incongruent or go/no-go structure).

**The blocked-vs-mixed-trial finding is worth follow-up.** That H&F's
blocked design retains more difference-score reliability than its
trial-intermixed mixed block is consistent with the variance-shrinkage
mechanism, but we only have one task to check it in. If it holds up in other
tasks, it would suggest a practical mitigation for researchers who need an
individual-differences difference score badly enough to trade away some
ecological validity: block, don't intermix.

**The stop-signal analogue needs a real SSRT, not this vignette's naive
substitute.** We used failed-stop RT in place of Hedge et al.'s race-model
SSRT purely to keep a go/no-go-family task in the comparison with the tools
already used elsewhere in this vignette. A proper `matzke-etal-2019`
replication attempt should use the integration method for SSRT (as Hedge et
al. recommend) rather than this shortcut.

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the compute script (from the project root)
source("vignettes/hf_reliability_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/hf_reliability_paradox.qmd")
```

---

## References
