---
title: "The Reliability Paradox in Cognitive-Control Tasks"
description: |
  Hedge, Powell & Sumner (2018) showed that classic conflict tasks (Flanker,
  Stroop, stop-signal) produce large, dependable group effects but poor
  individual-differences reliability, because the within-subject difference
  score that measures the effect cancels out most of the between-person
  variance. We test whether the same attenuation appears in IRW
  cognitive-control data: the Hearts & Flowers task, an alcohol Stroop, a
  stop-signal task, and
  three tasks (Stroop, Simon, ANT flanker) from the Enkavi et al. (2019)
  Self-Regulation Ontology battery.
date: today
bibliography: hf_reliability/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "hf_reliability/hf_reliability_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache      <- readRDS(cache_path)
  step0       <- .cache$step0_summary
  results     <- .cache$results
  hedge_ref   <- .cache$hedge2018_reference
  n_splits    <- .cache$n_splits
  trial_floors <- .cache$trial_floors
  date_run    <- .cache$date_run
} else {
  step0    <- tibble(table = character())
  results  <- tibble(dataset = character(), task_type = character(), measure = character(),
                      reliability_type = character(), reliability_est = numeric(),
                      reliability_sd = numeric(), trial_floor = integer(),
                      n_respondents = integer(), mean_trials_congruent = numeric(),
                      mean_trials_incongruent = numeric(), var_congruent = numeric(),
                      var_incongruent = numeric(), cov_cong_incong = numeric(),
                      var_difference = numeric(), notes = character())
  hedge_ref <- tibble(task = character(), measure = character(), condition = character(),
                       icc_study1 = numeric(), icc_study2 = numeric(), icc_mean = numeric())
  n_splits <- NA_integer_
  trial_floors <- c(5L, 10L)
  date_run <- Sys.Date()
}

knitr::opts_chunk$set(eval = cache_ok)

dataset_labels <- c(
  imps2025_hf_blocked    = "Hearts & Flowers (blocked)",
  imps2025_hf_mixed      = "Hearts & Flowers (mixed-trial)",
  alcoholstroop_jones2024 = "Alcohol Stroop",
  matzke_stopsignal_2019 = "Stop-signal (OSARI)",
  enkavi_stroop           = "Stroop (Enkavi et al. 2019)",
  enkavi_simon            = "Simon (Enkavi et al. 2019)",
  enkavi_ant_flanker      = "ANT Flanker (Enkavi et al. 2019)"
)

measure_labels <- c(rt = "Reaction time", acc = "Accuracy")
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/hf_reliability_compute.R\n```\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`hf_reliability_compute.R`) directly if you're
relying on the numbers here.
:::

## Motivation

Some cognitive tasks produce huge, dependable group-level effects yet turn
out to be nearly useless for studying individual differences. The clearest
examples are **conflict tasks**: paradigms where each trial pairs a
target feature that must be responded to with a second, task-irrelevant
feature that either matches (congruent) or clashes (incongruent) with the
correct response. Flanker (arrows flanked by same- or opposite-pointing
distractors), Stroop (color words printed in matching or mismatching ink),
and Simon-type spatial-compatibility tasks are the classic examples. In all
of them, incongruent trials require suppressing the irrelevant feature's
pull toward the wrong response, which costs time and accuracy relative to
congruent trials — the *conflict effect* (or congruency effect), usually
indexed as incongruent-minus-congruent RT or accuracy. Every study finds
that people are slower and less accurate on incongruent trials than
congruent ones — the effect is close to universal. But
@hedge2018reliability showed that when you use these same tasks to *rank*
people (rather than just detect the group effect), reliability collapses:
individual conditions like congruent RT or incongruent RT are often reliable
enough on their own (test-retest ICCs around .6-.8), but the *difference
score* used to index the effect of interest — incongruent RT minus congruent
RT — is barely reliable at all.

::: {.callout-note}
**What "reliability collapses" means, precisely.** Imagine testing the same
people twice (or splitting one session's trials into two halves) and
correlating each person's score across the two measurements. For congruent
RT alone, or incongruent RT alone, that correlation is reasonably strong
(ICC ~.6-.8) — a person who's fast in one measurement tends to still be
fast in the other. For the *difference score*, that same across-measurement
correlation is much weaker — in some datasets below, close to zero. This
doesn't mean the conflict effect itself is fake or unstable at the group
level; it means that *individual differences* in the size of the effect are
mostly swamped by trial-level noise, so a person's difference score in one
measurement predicts almost nothing about their difference score in
another. How close to zero depends on how strongly congruent and
incongruent RT covary across people — see [Where does the variance go?] for
why.
:::

```{r}
#| label: fig-task-examples
#| fig-cap: "Schematic examples of the conflict manipulation in Stroop and Hearts & Flowers. Stroop: name the ink colour, ignore the printed word; ink and word agree on congruent trials, disagree on incongruent trials. Hearts & Flowers: press the button on the same side as the stimulus for hearts (congruent), the opposite side for flowers (incongruent). These are illustrative reconstructions, not verbatim task stimuli."
#| fig-height: 3
#| echo: false
#| eval: true

stroop_df <- tibble(
  panel = factor(c("Congruent", "Incongruent"), levels = c("Congruent", "Incongruent")),
  ink   = c(irw_red, irw_blue)
)

stroop_plot <- ggplot(stroop_df, aes(x = 0, y = 0, label = "RED", colour = ink)) +
  geom_text(size = 13, fontface = "bold") +
  scale_colour_identity() +
  facet_wrap(~panel, nrow = 1) +
  coord_cartesian(xlim = c(-1, 1), ylim = c(-1, 1)) +
  labs(title = "Stroop") +
  theme_void(base_size = 13) +
  theme(strip.text = element_text(size = 11, face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        panel.spacing = unit(1.2, "lines"),
        plot.margin = margin(5, 15, 5, 5))

hf_df <- tibble(
  panel      = factor(c("Congruent (heart)", "Incongruent (flower)"),
                       levels = c("Congruent (heart)", "Incongruent (flower)")),
  symbol     = c("♥", "❀"),
  arrow_xend = c(-0.6, 0.6)
)

hf_plot <- ggplot(hf_df) +
  geom_text(aes(x = -0.6, y = 0.6, label = symbol), colour = "grey20", size = 11) +
  geom_rect(data = tibble(xmin = c(-0.9, 0.3), xmax = c(-0.3, 0.9)),
            aes(xmin = xmin, xmax = xmax), ymin = -0.9, ymax = -0.5,
            fill = NA, colour = irw_grey, inherit.aes = FALSE) +
  geom_segment(aes(x = -0.6, y = 0.3, xend = arrow_xend, yend = -0.5),
               arrow = arrow(length = unit(0.15, "cm")), colour = irw_grey) +
  facet_wrap(~panel, nrow = 1) +
  coord_cartesian(xlim = c(-1, 1), ylim = c(-1, 1)) +
  labs(title = "Hearts & Flowers") +
  theme_void(base_size = 13) +
  theme(strip.text = element_text(size = 11, face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        panel.spacing = unit(1.2, "lines"),
        plot.margin = margin(5, 5, 5, 15))

gridExtra::grid.arrange(stroop_plot, hf_plot, ncol = 2)
```

The mechanism is a direct consequence of how difference scores propagate
variance. For two conditions $A$ and $B$ measured on the same people,

$$\mathrm{Var}(B - A) = \mathrm{Var}(A) + \mathrm{Var}(B) - 2\,\mathrm{Cov}(A, B)$$ {#eq-vardiff}

Conflict tasks are robust precisely *because* almost everyone shows a similar
conflict cost — high scorers on congruent trials tend to also be high scorers
on incongruent trials (fast people are fast in both conditions, slow people
are slow in both). That shared individual differences signal is exactly what
$\mathrm{Cov}(A, B)$ captures, and subtracting it away leaves a difference
score with little between-person variance left, while the trial-level
measurement noise in each condition is *not* similarly reduced. The result:
low signal-to-noise for the individual-differences measure, even though the
group-level contrast ($\bar{B} - \bar{A}$) is large and rock solid.

This is a *structural* property of difference-score designs, not a defect
specific to Flanker or Stroop. Any task that measures an effect as a
within-subject contrast between two positively-correlated conditions is
vulnerable to it. The Hearts & Flowers (H&F) task [@imps2025_hf] — part of
the imps2025 datathon data in IRW — has exactly this structure: children
respond to a heart or flower stimulus, pressing the button on the *same*
side as the stimulus for hearts (congruent, low conflict) and the *opposite*
side for flowers (incongruent, high conflict). This is a spatial
stimulus-response compatibility manipulation in the same family as Simon and
Flanker effects, which makes it a natural candidate for replicating the
reliability paradox in IRW data. We also check independent tasks with the
same congruent/incongruent structure: an alcohol-related attentional-bias
Stroop [@alcoholstroop_jones2024], a stop-signal (go/no-go family) task
[@matzke-etal-2019], and three tasks (a classic color-word Stroop, a Simon
task, and the flanker component of the Attention Network Test) from
@enkavi2019's large-scale Self-Regulation Ontology test-retest battery.
Enkavi et al.'s own analysis is directly relevant here: across 372
behavioral-task dependent variables, they found contrast (difference-score)
DVs had a median reliability roughly 0.29 points lower than the raw,
noncontrast DVs they were built from (median ICC 0.17 vs. 0.50) — the same
qualitative pattern as @hedge2018reliability, in a much broader battery, and
the Stroop effect is one of the examples they use to illustrate it.

## Data and methods

### Candidate dataset search

We screened IRW's table metadata for trial-level (not fixed-item-bank) tasks
with a response-time column and a plausible two-condition contrast, using a
keyword search over table names, dataset descriptions, and variable lists
(`heart`, `flower`, `flanker`, `stroop`, `simon`, `congruent`, `stop-signal`,
etc.), followed by direct inspection of each survivor's columns.

```{r}
#| label: tbl-step0
#| tbl-cap: "Candidate tables surviving the keyword screen, and why each was kept or excluded."

step0 %>%
  mutate(
    decision = case_when(
      table == "imps2025_hf" ~ "Kept (Hearts & Flowers, 2475 respondents, has wave variable)",
      table == "alcoholstroop_jones2024" ~ "Kept (156 respondents, single session)",
      grepl("^enkavi_2019_", table) ~ "Kept (Enkavi et al. 2019 battery, 523 respondents, has wave variable)",
      grepl("safety_conflicts", table) ~ "Excluded: keyword false positive on \"conflicts\" (public-safety survey, no rt/condition structure)",
      TRUE ~ "Excluded"
    )
  ) %>%
  select(table, n_responses, n_participants, has_rt, decision) %>%
  knitr::kable(col.names = c("Table", "Responses", "Participants", "Has RT", "Decision"),
               format.args = list(big.mark = ","))
```

Two further candidates were located by widening the search to any table with
an `rt` column and inspecting item-name/column structure directly (not
capturable by a keyword match on task names): `wang2022_iat` (a fixed
20-item dichotomous bank despite its name — no trial-level RT, excluded) and
five `matzke-*` stop-signal tables (a `go_`/`stop_` item-prefix condition,
`cov_ssd` = stop-signal delay). Of those, `matzke-etal-2019` has 53
respondents and was kept; `matzke-experiment1-2022`,
`matzke-experiment3-standard`, and `matzke-experiment3-gamified-2022` have
7, 9, and 9 respondents respectively and were dropped for falling below our
50-respondent floor.

A third round located `enkavi_2019_stroop`, `enkavi_2019_simon`, and
`enkavi_2019_ant_flanker` — three tasks from @enkavi2019's Self-Regulation
Ontology battery — by direct inspection of IRW's table list rather than the
keyword screen above: at the time of writing these tables aren't yet indexed
in IRW's metadata catalog, so the keyword screen (which scans that catalog)
can't see them, even though they're fetchable like any other IRW table and
their table names alone would have matched the `stroop`/`simon`/`flanker`
keywords. Five further tasks in the same battery
(`enkavi_2019_navon`, `enkavi_2019_stopsignal`, `enkavi_2019_gonogo`,
`enkavi_2019_taskswitch`, `enkavi_2019_dpx_axcpt`) were located at the same
time but excluded from this round: they're related but not binary
congruent/incongruent contrasts (go/no-go, switch cost, AX-CPT probe types)
or need a bespoke 3-condition contrast (Navon), and are left for a future
extension of this analysis.

The final set has **seven dataset variants** across six IRW tables:

- **Hearts & Flowers, blocked** (`imps2025_hf`): the `hearts test` block
  (congruent) vs. the `flowers test` block (incongruent) — a separate-block
  design.
- **Hearts & Flowers, mixed-trial** (`imps2025_hf`): within the `mixed test`
  block, trials are classified congruent/incongruent by stimulus shape
  (heart/flower) rather than by block — a trial-intermixed design, which is
  structurally closer to how Hedge et al.'s tasks are actually administered
  (their conditions are intermixed within blocks, not run in separate
  blocks; see their Fig. 1).
- **Alcohol Stroop** (`alcoholstroop_jones2024`): neutral word (congruent
  analogue) vs. alcohol-related word (incongruent analogue) — an
  attentional-bias variant of Stroop rather than a color-word conflict task,
  but with an identical per-trial rt/accuracy structure.
- **Stop-signal / OSARI** (`matzke-etal-2019`): go trials (congruent
  analogue) vs. stop-signal trials (incongruent analogue).
- **Stroop** (`enkavi_2019_stroop`): classic color-word Stroop, congruent
  vs. incongruent, from @enkavi2019's Self-Regulation Ontology battery.
  N=523, an online (Amazon Mechanical Turk) adult sample (52% female, age
  21-60, mean 34.1).
- **Simon** (`enkavi_2019_simon`): spatial stimulus-response compatibility
  task, congruent vs. incongruent, same battery and sample as above.
- **ANT Flanker** (`enkavi_2019_ant_flanker`): the flanker component of the
  Attention Network Test, congruent vs. incongruent (a third "neutral"
  condition, no flankers, is excluded to keep the same two-condition
  contrast used throughout this vignette), same battery and sample as
  above.

### Preprocessing

For each variant we dropped respondents with fewer than 5 trials in either
condition (sensitivity checked at a floor of 10; see
[Trial-count sensitivity]). Per person and condition we computed mean RT on
correct trials and mean accuracy, then the contrast

$$D_{rt} = \overline{RT}_{\text{incongruent}} - \overline{RT}_{\text{congruent}}, \qquad
  D_{acc} = \overline{acc}_{\text{congruent}} - \overline{acc}_{\text{incongruent}}$$

so that positive values of both indices reflect the "expected" direction of
a conflict cost.

**Stop-signal deviation.** By construction, a successfully inhibited
stop-signal trial has no response and hence no RT — there is no such thing
as "the RT of a correct stop trial." Hedge et al. instead estimate
Stop-Signal Reaction Time (SSRT) via a race-model calculation. We do not
implement the race model here; instead $D_{rt}$ for the stop-signal dataset
uses the RT of *failed* stop trials (a response did occur) as a naive
structural analogue, and $D_{acc}$ compares go-trial direction accuracy to
the stop-trial inhibition rate — two different kinds of accuracy. Treat the
stop-signal numbers below as a structural check only, not a validated
inhibition measure.

**Split-half reliability.** For each dataset/measure/condition we ran
`r ifelse(is.na(n_splits), 500, n_splits)` random split-halves: within each
person and condition, trials were randomly assigned to two halves, person
means were computed for each half, and the correlation between halves across
people was Spearman-Brown corrected ($r_{sb} = 2r/(1+r)$). The difference
score was split-half by using the same halves within each condition, so
$D_{\text{half 1}} = \overline{incong}_{\text{half 1}} - \overline{cong}_{\text{half 1}}$,
and correlating the two halves' difference scores. We report the mean and SD
of the Spearman-Brown-corrected reliability across splits, and (for
split-half only) restrict to each person's earliest available session, so
that pooling multiple sessions doesn't artificially inflate reliability for
respondents with repeat data.

**Test-retest.** Two dataset groups have a genuine wave variable and
therefore support real test-retest correlations across sessions (no
splitting; raw condition means and $D$ from each full session, correlated
across sessions), rather than just split-half:

- **Hearts & Flowers**: two wave pairs were available, `PLUS Fall` →
  `PLUS Spring` and `2021-2022 Fall` → `2021-2022 Spring` (different
  cohorts). **This is an important deviation from Hedge et al.**: their
  test-retest interval is ~3 weeks in stable young adults, ours is roughly a
  school semester (~6 months) in developing children aged ~5-11 — real
  developmental change over that interval, not just measurement noise, will
  suppress our test-retest correlations relative to theirs.
- **The three Enkavi et al. (2019) tasks** (Stroop, Simon, ANT Flanker):
  `wave` is coded 1/2 for the two sessions. @enkavi2019 report a mean
  interval of 111 days between waves (median 115, range 60-228) in their
  Amazon Mechanical Turk adult sample — shorter than Hearts & Flowers'
  ~6-month gap and in stable adults rather than developing children, but
  still much longer than Hedge et al.'s ~3 weeks, and with substantial
  participant-to-participant variability in the exact interval (unlike the
  fixed-cohort H&F waves). Of the 523 respondents used for split-half,
  only respondents retained by Enkavi et al.'s own retest invitation and
  who cleared our trial floor at both sessions contribute to test-retest
  estimates (~150 respondents; see @fig-retest).

The alcohol Stroop and stop-signal datasets have no wave variable and
support split-half only.

## Results

### Split-half reliability: conditions vs. difference scores

```{r}
#| label: fig-splithalf
#| fig-cap: "Split-half reliability (Spearman-Brown corrected; mean ± SD across splits) for each condition and the difference score, trial floor = 5, for the six congruency-style dataset variants. Difference-score reliability drops relative to the conditions it's built from in every one of them, though the size of the drop varies widely. The stop-signal dataset is excluded here and shown separately in [Stop-signal: a structural comparison], since failed-stop RT isn't a congruency contrast in the same sense."
#| fig-height: 7

sh_plot_df <- results %>%
  filter(trial_floor == 5, !grepl("test-retest", reliability_type),
         dataset != "matzke_stopsignal_2019") %>%
  mutate(
    dataset_lab = factor(dataset_labels[dataset], levels = rev(dataset_labels)),
    reliability_type = factor(reliability_type,
                               levels = c("congruent", "incongruent", "difference"),
                               labels = c("Congruent", "Incongruent", "Difference score")),
    measure_lab = measure_labels[measure],
    tooltip = paste0(
      "Dataset: ", dataset_lab,
      "<br>Condition: ", reliability_type,
      "<br>Measure: ", measure_lab,
      "<br>Reliability: ", sprintf("%.2f", reliability_est), " ± ", sprintf("%.2f", reliability_sd),
      "<br>N: ", n_respondents
    )
  )

p_sh <- ggplot(sh_plot_df, aes(x = reliability_est, y = dataset_lab, colour = reliability_type, text = tooltip)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_errorbarh(aes(xmin = reliability_est - reliability_sd, xmax = reliability_est + reliability_sd),
                  height = 0, position = position_dodge(width = 0.5)) +
  geom_point(size = 2.6, position = position_dodge(width = 0.5)) +
  facet_wrap(~measure_lab) +
  scale_colour_manual(values = c("Congruent" = irw_grey, "Incongruent" = "black",
                                  "Difference score" = irw_red), name = NULL) +
  labs(x = "Split-half reliability (Spearman-Brown corrected)", y = NULL) +
  theme(legend.position = "bottom")

ggplotly(p_sh, tooltip = "text")
```

<details>
<summary>Show full split-half reliability table (all datasets, measures, and conditions)</summary>

```{r}
#| label: tbl-splithalf
#| tbl-cap: "Split-half reliability by dataset, measure, and condition (trial floor = 5)."

results %>%
  filter(trial_floor == 5, !grepl("test-retest", reliability_type)) %>%
  transmute(
    Dataset = dataset_labels[dataset],
    Measure = measure_labels[measure],
    Condition = tools::toTitleCase(reliability_type),
    `N respondents` = n_respondents,
    `Reliability (mean ± SD)` = sprintf("%.2f ± %.2f", reliability_est, reliability_sd)
  ) %>%
  arrange(Dataset, Measure) %>%
  knitr::kable()
```

</details>

The pattern in @fig-splithalf reproduces the core finding of
@hedge2018reliability directly: congruent and incongruent RT reliabilities
sit at .78-.99 across all six congruency-style dataset variants, but the RT
difference-score reliability drops in every one of them. The size of that
drop, however, varies considerably. At one extreme, it collapses
to near zero for the mixed-trial Hearts & Flowers
(`r sprintf("%.2f", results$reliability_est[results$dataset=="imps2025_hf_mixed" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`)
and the alcohol Stroop
(`r sprintf("%.2f", results$reliability_est[results$dataset=="alcoholstroop_jones2024" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`).
The blocked Hearts & Flowers design
(`r sprintf("%.2f", results$reliability_est[results$dataset=="imps2025_hf_blocked" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`)
and the Enkavi et al. (2019) Simon task
(`r sprintf("%.2f", results$reliability_est[results$dataset=="enkavi_simon" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`)
retain a moderate amount. And the Enkavi Stroop
(`r sprintf("%.2f", results$reliability_est[results$dataset=="enkavi_stroop" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`)
and ANT Flanker
(`r sprintf("%.2f", results$reliability_est[results$dataset=="enkavi_ant_flanker" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`)
retain a sizeable fraction of their component reliability, despite still
being well below their congruent/incongruent components (~.93-.96). This is
a useful complication of the "collapses to near zero" framing from the
Motivation section: the direction of the effect (component >> difference) is
essentially universal across these six variants, but its magnitude is not
— it depends on exactly how strongly the two conditions covary across
people in a given task, sample, and design, which is itself variable. The
Enkavi tasks — a much larger (N=523), adult, single-country online sample
than the other datasets here — retaining more difference-score reliability
than mixed-trial H&F or the alcohol Stroop is consistent with sample
composition (a wider age/ability range inflates between-person variance
relative to trial-level noise) contributing to that variability, though we
can't isolate that from task differences with only one dataset per task.

The blocked-vs-mixed-trial contrast within the same H&F task is informative:
Hedge et al.'s tasks intermix conditions within a block, and it's exactly
that design — the mixed-trial H&F variant — that shows the most severe
collapse. Separating congruent and incongruent trials into their own blocks
(as the blocked H&F variant does) leaves noticeably more difference-score
reliability intact, consistent with the variance-shrinkage mechanism: block
separation gives some room for per-block state (vigilance, strategy,
warm-up) to vary independently across conditions, which lowers
$\mathrm{Cov}(A,B)$ relative to $\mathrm{Var}(A)$ and $\mathrm{Var}(B)$ less
than a fully intermixed design does.

**This contrast deserves a caution, though.** Mean trials per condition are
similar between the two variants (~9.5-11.1 for blocked, ~9.7-9.9 for mixed;
see [Trial-count sensitivity]), but the *retained samples* are not: the
blocked split-half sample is
`r results$n_respondents[results$dataset=="imps2025_hf_blocked" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5]`
respondents (of the 2,475 in `imps2025_hf`), while the mixed-trial sample is
only
`r results$n_respondents[results$dataset=="imps2025_hf_mixed" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5]`
— about a quarter as many. Since per-condition trial counts are comparable
for those who clear the floor in each variant, this gap isn't the trial
floor thinning out the mixed sample; it looks more like most respondents
never received a "mixed test" block at all, so blocked and mixed aren't a
within-person manipulation on a shared sample but two considerably
different-sized (and potentially differently-composed, e.g. by cohort or
wave) subgroups of the full `imps2025_hf` respondent pool. We can't rule out
that this compositional difference, rather than the block/intermixed design
itself, contributes to the reliability gap between them.

### Stop-signal: a structural comparison

The stop-signal dataset was excluded from @fig-splithalf above because it
doesn't share the congruent/incongruent structure of the other six
variants, and its result is the opposite of the general pattern: its
difference-score reliability stays high
(`r sprintf("%.2f", results$reliability_est[results$dataset=="matzke_stopsignal_2019" & results$measure=="rt" & results$reliability_type=="difference" & results$trial_floor==5])`),
barely below its congruent (go,
`r sprintf("%.2f", results$reliability_est[results$dataset=="matzke_stopsignal_2019" & results$measure=="rt" & results$reliability_type=="congruent" & results$trial_floor==5])`)
and incongruent (stop,
`r sprintf("%.2f", results$reliability_est[results$dataset=="matzke_stopsignal_2019" & results$measure=="rt" & results$reliability_type=="incongruent" & results$trial_floor==5])`)
components. As detailed in Methods, this isn't really a like-for-like
difference score: $D_{rt}$ here is failed-stop RT minus go RT, not
Hedge et al.'s race-model SSRT, so this figure is a structural comparison,
not a validated inhibition measure. Failed-stop RT and go RT are driven by
different aspects of the race model (a fast guess vs. the whole
go-process distribution) and so are less tightly coupled across people
than two congruency conditions of the same trial type — the
variance-shrinkage mechanism has less to work with. That makes it a useful
negative case (the paradox is not automatic whenever you subtract two
RTs — it depends on how correlated the two quantities are across people),
but not a direct replication attempt, and it should not be read alongside
the congruency tasks above as if it were measuring the same kind of
contrast. A proper replication would need the race-model SSRT integration
method Hedge et al. use, which we do not implement here.

```{r}
#| label: tbl-stopsignal
#| tbl-cap: "Stop-signal split-half reliability alongside Hedge et al. (2018) Table 1 (mean of their Studies 1 & 2). The IRW 'Difference' row is failed-stop RT minus go RT, a structural stand-in for SSRT, not SSRT itself — the two are not the same quantity and shouldn't be read as a validation of one against the other."

ss_irw <- results %>%
  filter(trial_floor == 5, measure == "rt", !grepl("test-retest", reliability_type),
         dataset == "matzke_stopsignal_2019") %>%
  mutate(Condition = recode(reliability_type,
                             congruent = "Go", incongruent = "Stop",
                             difference = "Difference (failed-stop − go)")) %>%
  transmute(Source = "IRW, split-half", Condition,
            Reliability = round(reliability_est, 2))

ss_hedge <- hedge_ref %>%
  filter(measure == "rt", task == "Stop-signal") %>%
  transmute(Source = "Hedge et al. (2018), test-retest",
            Condition = tools::toTitleCase(condition),
            Reliability = round(icc_mean, 2))

bind_rows(ss_irw, ss_hedge) %>%
  knitr::kable()
```

### Where does the variance go?

```{r}
#| label: tbl-variance
#| tbl-cap: "Between-person variance components underlying the RT difference score (trial floor = 5). Var(D) = Var(incong) + Var(cong) − 2·Cov(cong, incong); a high covariance relative to the individual variances is what shrinks Var(D)."

results %>%
  filter(trial_floor == 5, measure == "rt", reliability_type == "difference") %>%
  transmute(
    Dataset = dataset_labels[dataset],
    `Var(congruent)` = round(var_congruent, 4),
    `Var(incongruent)` = round(var_incongruent, 4),
    `Cov(cong, incong)` = round(cov_cong_incong, 4),
    `Var(difference)` = round(var_difference, 4)
  ) %>%
  knitr::kable()
```

This table makes @eq-vardiff concrete: in every dataset,
$\mathrm{Cov}(\text{cong}, \text{incong})$ is a large fraction of both
individual variances, so subtracting $2\,\mathrm{Cov}$ (the last term in
@eq-vardiff) removes most of the variance that Var(congruent) and
Var(incongruent) contribute, leaving Var(difference) small relative to
either component alone.

@fig-variance-decomp shows @eq-vardiff concretely for a single task, the
alcohol Stroop. Var(congruent) and Var(incongruent) are both large and
similar in size, but so is $2\,\mathrm{Cov}(\text{cong}, \text{incong})$ —
almost as large as the two variances combined — so nearly all of that
variance is removed in the subtraction, leaving Var(difference) close to
zero. This is the arithmetic behind the "reliability collapses" claim
above: two individually well-behaved, reliable measurements can combine
into an almost-zero-variance difference score whenever they're highly
correlated across people.

```{r}
#| label: fig-variance-decomp
#| fig-cap: "Variance decomposition for the alcohol Stroop RT difference score (trial floor = 5). Var(difference) = Var(congruent) + Var(incongruent) − 2·Cov(congruent, incongruent); the covariance bar is almost as large as the sum of the two variance bars, leaving very little variance in the difference score."
#| fig-height: 4.2

vdf_row <- results %>%
  filter(trial_floor == 5, measure == "rt", reliability_type == "difference",
         dataset == "alcoholstroop_jones2024")

vdf <- tibble(
  component = factor(
    c("Var(congruent)", "Var(incongruent)", "2·Cov(cong, incong)", "Var(difference)"),
    levels = c("Var(congruent)", "Var(incongruent)", "2·Cov(cong, incong)", "Var(difference)")
  ),
  value = c(vdf_row$var_congruent, vdf_row$var_incongruent,
            2 * vdf_row$cov_cong_incong, vdf_row$var_difference),
  role  = c("congruent", "incongruent", "subtracted", "result")
)

ggplot(vdf, aes(x = component, y = value, fill = role)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.3f", value)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c(congruent = irw_grey, incongruent = "black",
                               subtracted = irw_blue, result = irw_red),
                     guide = "none") +
  labs(x = NULL, y = "Variance (RT, seconds²)",
       title = paste(dataset_labels["alcoholstroop_jones2024"], "RT: where the variance goes")) +
  expand_limits(y = max(vdf$value) * 1.15) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(size = 12, face = "bold"))
```

### True test-retest reliability

Five of the seven dataset variants have a genuine wave variable and
therefore support true test-retest correlations, not just split-half: the
two Hearts & Flowers variants (a ~6-month, within-school-year gap in
developing children) and the three Enkavi et al. (2019) tasks (a mean
111-day gap in adults; see Preprocessing).

```{r}
#| label: fig-retest
#| fig-cap: "Split-half (single session) vs. true test-retest reliability for RT, across all five dataset variants with a wave variable. Hearts & Flowers test-retest spans roughly a school semester in developing children; the Enkavi tasks span a mean 111-day gap in adults. Neither matches Hedge et al.'s ~3-week adult interval, so lower absolute values are expected in both."
#| fig-height: 7
#| fig-width: 9.5

retest_df <- results %>%
  filter(dataset %in% c("imps2025_hf_blocked", "imps2025_hf_mixed",
                         "enkavi_stroop", "enkavi_simon", "enkavi_ant_flanker"),
         measure == "rt") %>%
  mutate(
    design = case_when(
      grepl("PLUS Fall -> PLUS Spring", reliability_type) ~ "Test-retest (H&F, PLUS cohort)",
      grepl("2021-2022 Fall -> 2021-2022 Spring", reliability_type) ~ "Test-retest (H&F, K cohort)",
      grepl("test-retest: 1 -> 2", reliability_type) ~ "Test-retest (Enkavi et al. 2019, mean 111d)",
      trial_floor == 5 ~ "Split-half (single session)",
      TRUE ~ NA_character_
    ),
    condition = case_when(
      grepl("^congruent", reliability_type) ~ "Congruent",
      grepl("^incongruent", reliability_type) ~ "Incongruent",
      grepl("^difference", reliability_type) ~ "Difference score",
      TRUE ~ NA_character_
    ),
    dataset_lab = dataset_labels[dataset],
    tooltip = paste0(
      "Dataset: ", dataset_labels[dataset],
      "<br>Design: ", design,
      "<br>Condition: ", condition,
      "<br>Reliability: ", sprintf("%.2f", reliability_est)
    )
  ) %>%
  filter(!is.na(design))

p_retest <- ggplot(retest_df, aes(x = reliability_est, y = condition, colour = design, text = tooltip)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(size = 2.8, position = position_dodge(width = 0.5)) +
  facet_wrap(~dataset_lab, ncol = 2) +
  scale_colour_manual(values = c("Split-half (single session)" = irw_blue,
                                  "Test-retest (H&F, PLUS cohort)" = irw_red,
                                  "Test-retest (H&F, K cohort)" = "#f4a582",
                                  "Test-retest (Enkavi et al. 2019, mean 111d)" = "#1a9850"),
                       name = NULL, guide = guide_legend(nrow = 2)) +
  labs(x = "Reliability", y = NULL) +
  theme(legend.position = "bottom")

ggplotly(p_retest, tooltip = "text") |>
  layout(
    legend = list(orientation = "h", y = -0.15),
    annotations = list(list(
      text      = "H&F PLUS cohort: Fall -> Spring, one school year. H&F K cohort: 2021-22 Fall -> Spring. Enkavi et al. (2019): mean 111-day gap, range 60-228 days, Amazon Mechanical Turk adults.",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0, y = -0.25, xanchor = "left", yanchor = "top",
      font = list(size = 10, color = "grey50")
    ))
  )
```

Test-retest reliabilities are lower than split-half across the board, as
expected from the much longer retest interval in both the H&F and Enkavi
designs relative to Hedge et al.'s ~3 weeks. But the *relative* pattern
survives everywhere: the difference score's test-retest reliability is
consistently the lowest of the three conditions, in all five dataset
variants, indicating the paradox isn't an artifact of the within-session
split-half method — it also shows up when the same people are tested again
weeks or months later, in both a developmental (H&F) and a stable-adult
(Enkavi) sample. As with the split-half results, the *size* of the
test-retest collapse varies: the Enkavi Stroop difference score
(`r sprintf("%.2f", results$reliability_est[results$dataset=="enkavi_stroop" & results$measure=="rt" & results$reliability_type=="difference (test-retest: 1 -> 2)"])`)
retains noticeably more reliability across a true ~111-day retest than the
Enkavi Simon
(`r sprintf("%.2f", results$reliability_est[results$dataset=="enkavi_simon" & results$measure=="rt" & results$reliability_type=="difference (test-retest: 1 -> 2)"])`)
or ANT Flanker
(`r sprintf("%.2f", results$reliability_est[results$dataset=="enkavi_ant_flanker" & results$measure=="rt" & results$reliability_type=="difference (test-retest: 1 -> 2)"])`)
difference scores do, mirroring the split-half ordering for the same three
tasks.

### Comparison to Hedge, Powell & Sumner (2018)

The table below places our estimates next to the point estimates reported in
Hedge et al.'s Table 1 (average of their Studies 1 and 2; two-way
random-effects ICC, absolute agreement, ~3-week test-retest in adults), for
the six congruency-style tasks. (The stop-signal dataset is compared
separately in [Stop-signal: a structural comparison], since it isn't a
congruency contrast.) IRW rows are shown per comparable task: the
split-half (single-session) estimate, which is the most directly comparable
design for the alcohol Stroop since it has no wave variable, plus a
test-retest estimate for every dataset that has one (Hearts & Flowers and
the three Enkavi et al. 2019 tasks) — of the two, the Enkavi test-retest
rows are the closer match to Hedge et al.'s own true-test-retest,
stable-adult design.

```{r}
#| label: tbl-hedge-comparison
#| tbl-cap: "IRW reliability estimates alongside Hedge et al. (2018) Table 1 point estimates (mean of their Studies 1 & 2), for the six congruency-style tasks. NA where a comparable estimate wasn't computed (e.g. no test-retest wave for a dataset). Stop-signal is shown separately below."

irw_split <- results %>%
  filter(trial_floor == 5, measure == "rt", !grepl("test-retest", reliability_type),
         dataset != "matzke_stopsignal_2019") %>%
  transmute(pair_task = dataset_labels[dataset], source = "IRW, split-half",
            condition = reliability_type, reliability = reliability_est)

irw_retest_hf <- results %>%
  filter(measure == "rt", grepl("test-retest: PLUS Fall", reliability_type)) %>%
  mutate(condition = sub(" \\(test-retest.*\\)", "", reliability_type)) %>%
  transmute(pair_task = dataset_labels[dataset], source = "IRW, test-retest (H&F, Fall->Spring)",
            condition, reliability = reliability_est)

irw_retest_enkavi <- results %>%
  filter(measure == "rt", grepl("test-retest: 1 -> 2", reliability_type)) %>%
  mutate(condition = sub(" \\(test-retest.*\\)", "", reliability_type)) %>%
  transmute(pair_task = dataset_labels[dataset], source = "IRW, test-retest (Enkavi, mean 111d)",
            condition, reliability = reliability_est)

hedge_tbl <- hedge_ref %>%
  filter(measure == "rt", task != "Stop-signal", task != "Go/No-go") %>%
  transmute(pair_task = task, source = "Hedge et al. (2018), test-retest",
            condition = condition, reliability = icc_mean)

pairing <- tribble(
  ~pair_task_irw,                     ~pair_task_hedge,
  "Hearts & Flowers (mixed-trial)",   "Flanker",
  "Hearts & Flowers (blocked)",       "Flanker",
  "Alcohol Stroop",                   "Stroop",
  "Stroop (Enkavi et al. 2019)",      "Stroop",
  "ANT Flanker (Enkavi et al. 2019)", "Flanker"
)

comparison <- bind_rows(irw_split, irw_retest_hf, irw_retest_enkavi) %>%
  inner_join(pairing, by = c("pair_task" = "pair_task_irw")) %>%
  transmute(`Lab task analogue` = pair_task_hedge, `IRW dataset` = pair_task,
            Source = source, Condition = tools::toTitleCase(condition),
            Reliability = round(reliability, 2)) %>%
  bind_rows(
    hedge_tbl %>%
      transmute(`Lab task analogue` = pair_task, `IRW dataset` = "—",
                Source = source, Condition = tools::toTitleCase(condition),
                Reliability = round(reliability, 2))
  ) %>%
  arrange(`Lab task analogue`, Source, factor(Condition, levels = c("Congruent", "Incongruent", "Difference", "Go", "Ssrt")))

knitr::kable(comparison)
```

A few things stand out. First, the *direction* of the effect (condition
reliability >> difference-score reliability) matches Hedge et al. in every
task pair shown here (the stop-signal analogue, excluded from this table,
runs the opposite way — see [Stop-signal: a structural comparison]).
Second, our absolute IRW split-half numbers for individual conditions run
higher than Hedge et al.'s test-retest ICCs — expected, since split-half
reliability has no opportunity for between-session instability to erode it,
while ICC test-retest does. Third, the Enkavi test-retest rows are the
closest design match to Hedge et al.'s own true-test-retest methodology
among our datasets, but don't line up on magnitude in a consistent
direction: the Enkavi Flanker test-retest difference-score reliability
(.34) is closer to Hedge et al.'s Flanker ICC (.49) than Enkavi's own
split-half estimate (.69) is, while for Stroop it's the reverse — Enkavi's
split-half estimate (.65) is closer to Hedge et al.'s Stroop ICC (.63) than
Enkavi's own test-retest estimate (.44) is. We read this as a caution
against over-interpreting small magnitude differences across these
comparisons: different reliability estimators (Pearson split-half here vs.
Hedge et al.'s two-way random-effects ICC), samples, and retest intervals
all contribute noise to the exact numbers, even when the qualitative
pattern (component >> difference) is consistent throughout.

### Trial-count sensitivity

```{r}
#| label: tbl-sensitivity
#| tbl-cap: "Split-half difference-score RT reliability at trial floors of 5 vs. 10 trials/person/condition."

results %>%
  filter(measure == "rt", reliability_type == "difference", !grepl("test-retest", reliability_type)) %>%
  transmute(Dataset = dataset_labels[dataset], `Trial floor` = trial_floor,
            `N respondents` = n_respondents,
            `Reliability (mean ± SD)` = sprintf("%.2f ± %.2f", reliability_est, reliability_sd)) %>%
  arrange(Dataset, `Trial floor`) %>%
  knitr::kable()
```

Raising the per-condition trial floor from 5 to 10 trims the sample (most
sharply for the mixed-trial H&F variant, which has fewer trials per
condition to begin with) but does not change the qualitative picture: the
difference score remains far less reliable than either condition alone. The
three Enkavi tasks are unaffected by this trim (523 respondents at both
floors) — they simply have more trials per condition to begin with than
H&F does.

## Discussion

The reliability paradox replicates in IRW cognitive-control data — as an
**attenuation of variable size, not a uniform collapse to zero**. Across
six independent tasks (Hearts & Flowers, alcohol Stroop, a stop-signal task,
and the Enkavi et al. 2019 Stroop, Simon, and ANT Flanker tasks) spanning
split-half and true test-retest designs in both a developmental and a
stable-adult sample, individual congruent and incongruent conditions are
consistently far more reliable than the within-subject difference score
built from them, but several of those difference scores — the blocked H&F
design, the Enkavi Simon task, and (most strikingly) the Enkavi Stroop and
ANT Flanker tasks — retain moderate reliability rather than collapsing
outright. The mechanism is exactly the one @hedge2018reliability
describe: the two conditions are highly correlated across people, so most
of their between-person variance cancels in the subtraction (@eq-vardiff),
while trial-level noise does not. @enkavi2019's own analysis of a much
broader task battery reaches the same conclusion by a different route,
which is reassuring convergent evidence that this isn't an artifact of our
specific pipeline.

**The size of the collapse varies far more than its direction.** Every
congruency-style dataset here (i.e., excluding the stop-signal exception)
shows the qualitative pattern, but RT difference-score reliability spans
roughly .04 to .69 depending on the task, sample, and design — not a fixed
"paradox constant." The Enkavi tasks, run in a larger, adult, Amazon
Mechanical Turk sample, generally retained more difference-score
reliability than the mixed-trial H&F or alcohol Stroop datasets did, though
we can't cleanly separate sample effects from task effects with only one
dataset per task. Researchers relying on a congruency-style contrast score
for individual-differences work should treat this as a genuinely
task-and-sample-specific empirical question, not something to be predicted
from task family alone.

## Limitations

**This is a property of difference-score designs, not a general statement
about IRW data.** Most IRW tables are fixed-item-bank IRT designs, where
ability is estimated from many items measuring one construct rather than
subtracted between two conditions — the mechanism demonstrated here doesn't
apply to them, and this vignette should not be read as evidence that IRW
data are broadly unreliable for individual-differences work. It applies
specifically to trial-level tasks that measure an effect as a within-subject
contrast between two positively-correlated conditions, which in the current
IRW warehouse means Hearts & Flowers, the alcohol Stroop, the
matzke-etal-2019 stop-signal task, and the enkavi_2019_stroop,
enkavi_2019_simon, and enkavi_2019_ant_flanker tasks (plus, presumably, any
future task with the same congruent/incongruent or go/no-go structure —
including the five other Enkavi et al. 2019 tasks noted in Data and methods
as candidates for a future extension).

**The blocked-vs-mixed-trial finding is worth follow-up.** That H&F's
blocked design retains more difference-score reliability than its
trial-intermixed mixed block is consistent with the variance-shrinkage
mechanism, but we only have one task to check it in — none of the Enkavi
tasks offer a blocked/mixed contrast of the same kind. If it holds up in
other tasks, it would suggest a practical mitigation for researchers who
need an individual-differences difference score badly enough to trade away
some ecological validity: block, don't intermix.

**The stop-signal analogue needs a real SSRT, not this vignette's naive
substitute.** We used failed-stop RT in place of Hedge et al.'s race-model
SSRT purely to keep a go/no-go-family task in the comparison with the tools
already used elsewhere in this vignette. A proper `matzke-etal-2019`
replication attempt should use the integration method for SSRT (as Hedge et
al. recommend) rather than this shortcut.

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the compute script (from the project root)
source("vignettes/hf_reliability_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/hf_reliability_paradox.qmd")
```

---

## References
