---
title: "Home Advantage: A Century and a Half of It, and a Season Without Crowds"
description: |
  Teams win more often at home. Using the IRW's competition tables, we estimate
  home advantage net of team strength in every season of baseball, basketball,
  hockey, American football and English and Italian football back to the
  nineteenth century, ask how it has changed, and then use the 2020-21 seasons
  played in empty stadiums to ask how much of it the crowd supplies.
date: today
bibliography: home_advantage_data/references.bib
nocite: |
  @*
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "home_advantage_data/home_advantage_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  nocrowd_df       <- .cache$nocrowd
  nba_bubble       <- .cache$nba_bubble
  excluded_df      <- .cache$excluded
  nocrowd_window   <- .cache$nocrowd_window
  min_games        <- .cache$min_games %||% 50
  candidate_tables <- .cache$candidate_tables
  date_run         <- .cache$date_run
  is_pilot         <- isTRUE(.cache$pilot)
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_tables_run   <- length(candidate_tables)
} else {
  summary_df <- tibble(
    sport = character(), league = character(), table = character(),
    period = numeric(), first_date = as.Date(character()), h = numeric(),
    h_se = numeric(), method = character(), n_games = integer(),
    n_agents = integer(), n_components = integer(), draw_rate = numeric(),
    home_share = numeric()
  )
  nocrowd_df <- tibble(
    league = character(), sport = character(), definition = character(),
    h = numeric(), h_se = numeric(), delta = numeric(), delta_se = numeric(),
    n_nocrowd = integer(), n_games = integer()
  )
  nba_bubble       <- NULL
  excluded_df      <- tibble(table = character(), reason = character())
  nocrowd_window   <- as.Date(c("2020-03-13", "2021-05-31"))
  min_games        <- 50
  candidate_tables <- character()
  date_run         <- Sys.Date()
  is_pilot         <- FALSE
  r_all_candidates <- 0L
  r_n_tables_run   <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)

# the six series with one estimate per season
season_series <- c("Baseball (MLB)", "Basketball (NBA)", "Ice hockey (NHL)",
                   "American football (NFL)", "England (all divisions)", "Italy (Serie A)")
covid <- c(2020, 2021)
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
# inline values: evaluated only when the cache exists (arguments are lazy)
iv <- function(x) if (cache_ok) x else "\u2014"
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/home_advantage_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, partial run.** This render uses ",
  r_n_tables_run, " of the ", r_all_candidates, " competition tables chosen for ",
  "this page, to validate the pipeline and page layout. All numbers here are real ",
  "computed results, but sports missing from the pilot are missing from every ",
  "figure. Re-run `home_advantage_compute.R` with `PILOT <- FALSE` for the full set.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`home_advantage_compute.R`) directly if you're relying
on the numbers here.
:::

## Motivation

Home teams win more often than away teams. This is among the oldest and best
documented regularities in sport: @pollard1986home documented it for English
football, @pollard2005long traced it across professional team sports in North
America and England back to 1876, and @jamieson2010home's meta-analysis found it
across every sport and era studied. Two questions about it remain live.

1. **Is it shrinking?** It is widely believed to have declined, but a raw
   home-win percentage mixes home advantage with who happened to host whom. A
   strong team that plays more home games, or a season with lopsided schedules,
   moves the raw number without any change in what playing at home is worth.
2. **Where does it come from?** The usual suspects are the crowd, travel,
   familiarity with the venue, and referees responding to the crowd. In 2020 and
   2021 most professional leagues played without spectators, which separated the
   crowd from the rest for the first time on a large scale. Studies of German
   [@fischer2021crowd], European [@mccarrick2021home] and North American
   [@higgs2021bayesian] leagues have used that natural experiment.

The IRW's [competition tables](../comps_standard.qmd) are well suited to both. They
store one row per game between two agents, with a common coding of which side was
at home, across sports and back as far as 1871. Home advantage is then a single
parameter in a paired-comparison model, estimated net of the strength of the two
sides. In IRW terms it is something like an item parameter that every game shares,
estimated across hundreds of tables, and the question is whether it is stable.

## Data and methods

### Dataset selection

```{r}
#| label: filter-datasets
#| eval: false

library(irw)
# every comp table with a real home side and a long run of seasons
tables <- c(
  "retrosheet_mlb_1871_2025", "fivethirtyeight_nba_1946_2023",
  "fivethirtyeight_nhl_1917_2023", "fivethirtyeight_nfl_1920_2023",
  "nflverse_nfl_1999_2025", "footbayes_england", "footbayes_italy",
  "fivethirtyeight_soccer_2016_2023", "rugby_union_tier1_1871_2026",
  "intlfootball_men_1872-2026", "intlfootball_women_1956-2026"
)
games <- lapply(tables, irw_fetch, source = "comp")
```

The IRW's competition collection has over a hundred tables, from professional
leagues and esports to animal dominance encounters and pairwise judgments. Most
have no home side. We used the `r r_all_candidates` tables that have one and that
run over many seasons. Each table codes the home side the same way: `homefield` is
`"agent_a"` when `agent_a` is at home and missing at a neutral venue. Tables left
out, and why:

```{r}
#| label: tbl-excluded
#| echo: false

knitr::kable(excluded_df, col.names = c("Table(s)", "Why left out"))
```

### The model

The analysis unit is a league-season (for rugby and internationals, which have too
few games per year, a decade). Within each unit we fit a Bradley-Terry model
[@bradley1952rank] with a home term:

$$
\text{logit}\, P(a \text{ beats } b) = h \cdot \text{home}_{ab} + \theta_a - \theta_b,
$$

where $\text{home}_{ab}$ is 1 if $a$ is at home, $-1$ if $b$ is, and 0 at a neutral
venue. The $\theta$s are team strengths, so $h$ is the home advantage in logits,
*net of who played whom*. For a sense of scale, $h = 0.4$ means two evenly matched
teams would see the home side win about `r round(100 * plogis(0.4))`% of decided
games. Where a sport has draws, we fit the ordinal version of the model
[@agresti1992analysis], in which a draw sits between a win and a loss on the same
latent scale, so $h$ keeps the same meaning.

```{r}
#| label: fit-function
#| eval: false

# Every second game is written from the away side's point of view, so that the
# home column is not constant (and so not absorbed by the ordinal cutpoints).
fit_h <- function(a, b, y, home, nocrowd = NULL) {
  flip <- seq_along(a) %% 2 == 0
  a2 <- ifelse(flip, b, a); b2 <- ifelse(flip, a, b)
  y <- ifelse(flip, 1 - y, y); home <- ifelse(flip, -home, home)
  # team-strength design, one reference team per connected component
  cmp <- components(a2, b2)
  keep <- setdiff(sort(unique(c(a2, b2))), tapply(names(cmp), cmp, `[`, 1))
  X <- matrix(0, length(a2), length(keep))
  ia <- match(a2, keep); ib <- match(b2, keep)
  X[cbind(which(!is.na(ia)), ia[!is.na(ia)])] <- 1
  X[cbind(which(!is.na(ib)), ib[!is.na(ib)])] <- -1
  H <- cbind(home = home)
  if (!is.null(nocrowd)) H <- cbind(H, home_nocrowd = home * nocrowd)
  Z <- cbind(H, X)
  if (any(y == 0.5)) {   # draws: ordinal Bradley-Terry
    m <- MASS::polr(factor(y, levels = c(0, 0.5, 1)) ~ Z, Hess = TRUE)
  } else {
    m <- glm(y ~ Z - 1, family = binomial)
  }
  # ... returns h, its SE, and delta (below) when nocrowd is given
}
```

A few details matter for what follows:

- **Regular season only.** Playoff games are left out of the season fits: there
  are few of them, and home order in a playoff series is earned by the better
  record, not scheduled.
- **Divisions.** In English football a season has up to four divisions that never
  meet. They go in one fit, each with its own reference team, and share a single
  $h$.
- **Sparse or one-sided agents.** Teams with fewer than five games in a unit, and
  teams that won or lost every game, are dropped before fitting; their strength
  cannot be estimated (it is infinite for a perfect record). This matters mostly
  for the internationals.
- **NFL sources.** 538's table supplies seasons to 1998 and nflverse's from 1999,
  where the two overlap.
- Units with fewer than `r min_games` games are skipped.

**The no-crowd test.** For each league we pool the 2016-2023 seasons into one fit,
with a strength for every team-season and a second home term for games played in
the no-crowd window, `r format(nocrowd_window[1], "%d %B %Y")` to
`r format(nocrowd_window[2], "%d %B %Y")`:

$$
\text{logit}\, P(a \text{ beats } b) = (h + \delta \cdot \text{nocrowd}) \cdot \text{home}_{ab} + \theta_{a,s} - \theta_{b,s}.
$$

$\delta$ is the change in home advantage when the stands were empty. One window is
used for every league; for MLB, whose Retrosheet table records attendance, we also
define "no crowd" directly as a 2020 game with no recorded attendance.

Of the units built, `r iv(nrow(summary_df))` produced estimates; the season fits used
the ordinal model in `r iv(sum(summary_df$method == "ordinal"))` and the binary model
in `r iv(sum(summary_df$method == "binary"))`.

::: {.callout-note}
To reproduce with current IRW holdings, re-run `home_advantage_compute.R`
and commit the updated `home_advantage_data/home_advantage_results.rds`.
:::

## Results

### How big is home advantage?

A logit is not a natural unit, so it helps to ask what $h$ means for a single game
between two **equally strong teams**, $\theta_a = \theta_b$. Then the strengths
cancel and the model's prediction depends on $h$ alone.

- In a sport without draws, the home side wins with probability
  $P(\text{home win}) = \text{logit}^{-1}(h) = 1 / (1 + e^{-h})$. $h = 0$ is a
  coin flip; $h = 0.4$ is about `r round(100 * plogis(0.4))`%; and $e^{h}$ is the
  factor by which playing at home multiplies the odds of winning.
- In a sport with draws, the ordinal model has two cutpoints, $c_1 < c_2$, which set
  how often games are drawn. For equal teams, the away side wins with probability
  $\text{logit}^{-1}(c_1 - h)$, the home side with $1 - \text{logit}^{-1}(c_2 - h)$,
  and the rest are draws. Home advantage moves probability from away wins to draws
  and from draws to home wins.

To put every sport on one scale, we use the home side's **expected share of the
points** for equal teams, counting a win as 1 and a draw as ½: $P(\text{home win}) +
\tfrac{1}{2} P(\text{draw})$. It is 50% when there is no home advantage, and in a
sport without draws it is just the home win probability. The figure shows it by
era; hover over a point for $h$ and the win, draw and loss probabilities behind it.

```{r}
#| label: size-prep
#| include: false

# implied outcome probabilities for two equal teams, given h and the cutpoints
outcome_probs <- function(h, c1, c2) {
  binary <- is.na(c1)
  win  <- ifelse(binary, plogis(h), 1 - plogis(c2 - h))
  loss <- ifelse(binary, 1 - plogis(h), plogis(c1 - h))
  data.frame(p_win = win, p_draw = 1 - win - loss, p_loss = loss)
}
share_of <- function(h, c1, c2) with(outcome_probs(h, c1, c2), p_win + p_draw / 2)

unit_size <- summary_df |>
  filter(!period %in% covid) |>
  mutate(share = share_of(h, cut1, cut2),
         # SE of the share by the delta method, holding the cutpoints fixed
         share_se = abs(share_of(h + 1e-4, cut1, cut2) - share_of(h - 1e-4, cut1, cut2)) / 2e-4 * h_se)
unit_size <- bind_cols(unit_size, outcome_probs(unit_size$h, unit_size$cut1, unit_size$cut2))

# precision-weighted means within a group: h and share by their own SEs, the
# outcome probabilities with the weights used for h
pool <- function(d) {
  w <- 1 / d$h_se^2; ws <- 1 / d$share_se^2
  tibble(h = sum(w * d$h) / sum(w), h_se = sqrt(1 / sum(w)),
         share = sum(ws * d$share) / sum(ws), share_se = sqrt(1 / sum(ws)),
         p_win = sum(w * d$p_win) / sum(w), p_draw = sum(w * d$p_draw) / sum(w),
         p_loss = sum(w * d$p_loss) / sum(w), n_units = nrow(d), n_games = sum(d$n_games))
}
era_size <- unit_size |>
  filter(sport != "Football (538 leagues)") |>
  mutate(era = cut(period, c(-Inf, 1949, 1989, 2019), labels = c("before 1950", "1950-1989", "1990-2019"))) |>
  filter(!is.na(era)) |>
  group_by(league, era) |>
  group_modify(~ pool(.x)) |>
  ungroup()
recent <- filter(era_size, era == "1990-2019") |> arrange(desc(share))
r_of <- function(lg, v) recent[[v]][recent$league == lg]
pct <- function(x) paste0(round(100 * x), "%")

league_538 <- unit_size |>
  filter(sport == "Football (538 leagues)") |>
  group_by(league) |>
  group_modify(~ pool(.x)) |>
  ungroup() |>
  arrange(share)
```

```{r}
#| label: fig-size
#| fig-height: 5.5
#| fig-cap: "Home side's expected share of the points against an equally strong opponent (win = 1, draw = ½), by era, ±2 SE. 50% means no home advantage. Precision-weighted over the seasons (or decades) in each era; 2020 and 2021 are left out."

era_cols <- c("before 1950" = "grey70", "1950-1989" = "#92c5de", "1990-2019" = irw_blue)
pd <- era_size |>
  mutate(league = factor(league, levels = rev(recent$league)),
         text = paste0(league, ", ", era, "<br>h = ", fmt(h), " (SE ", fmt(h_se), ")",
                       "<br>equal teams: home win ", pct(p_win),
                       if_else(p_draw > 0.005, paste0(", draw ", pct(p_draw)), ""),
                       ", away win ", pct(p_loss),
                       "<br>home share of points ", pct(share),
                       "<br>", n_units, if_else(grepl("intern|Rugby", league), " decades, ", " seasons, "),
                       format(n_games, big.mark = ","), " games"))
p <- ggplot(pd, aes(share, league, colour = era, text = text)) +
  geom_vline(xintercept = 0.5, colour = irw_grey) +
  geom_linerange(aes(xmin = share - 2 * share_se, xmax = share + 2 * share_se),
                 position = position_dodge(width = 0.6)) +
  geom_point(size = 2, position = position_dodge(width = 0.6)) +
  scale_colour_manual(values = era_cols) +
  scale_x_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(x = "home side's expected share of the points, equal teams", y = NULL, colour = NULL)
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h", y = -0.15))
```

```{r}
#| label: tbl-size
#| echo: false

recent |>
  transmute(Series = league,
            `Odds ratio` = fmt(exp(h)),
            h = paste0(fmt(h), " (", fmt(h_se), ")"),
            `Home win` = pct(p_win),
            Draw = if_else(p_draw > 0.005, pct(p_draw), "\u2014"),
            `Away win` = pct(p_loss),
            `Home share of points` = pct(share)) |>
  relocate(h, .after = Series) |>
  knitr::kable(align = "lrrrrrr",
               caption = "1990-2019: home advantage h (SE), the odds ratio e^h, and what they imply for a game between two equally strong teams. A dash means the sport has (almost) no draws in this era.")
```

Every series has a home advantage, and they differ a great deal. In 1990-2019, two
equally strong NBA teams would see the home side win
`r iv(pct(r_of("Basketball (NBA)", "p_win")))` of the time; in the NFL,
`r iv(pct(r_of("American football (NFL)", "p_win")))`. Hockey, which still had ties
for much of this era, comes out at a home share of the points of
`r iv(pct(r_of("Ice hockey (NHL)", "share")))`. Baseball's is the smallest:
`r iv(pct(r_of("Baseball (MLB)", "p_win")))`, barely more than a coin flip. The
largest are in international matches, rugby tests and men's football, where the
home side takes about two-thirds of the points from an equal opponent.

The football numbers read differently because about a quarter to a third of games
are drawn. Between two equal English sides, the home team wins
`r iv(pct(r_of("England (all divisions)", "p_win")))`, the game is drawn
`r iv(pct(r_of("England (all divisions)", "p_draw")))` and the away team wins
`r iv(pct(r_of("England (all divisions)", "p_loss")))`. That is a home share of the
points of `r iv(pct(r_of("England (all divisions)", "share")))`, about the same
edge as in the NFL, even though the home side wins less than half its games. Serie A and men's
international football are larger still: an equal visiting side wins only about one
game in five.

Two cautions in reading the table. First, $h$ itself is comparable only within a
model family. In the ordinal model a given $h$ shifts a latent scale with a draw
band in the middle, so the same $h$ implies a smaller home *win* probability than
it would in a sport without draws; the share of points is the comparable number.
Second, "equally strong" is a model statement. In a real league most games are not
between equal teams, and the home side's actual win rate depends on the schedule.
The point of $h$ is that it does not.

```{r}
#| label: fig-size-538
#| fig-height: 8
#| fig-cap: "Home side's expected share of the points against an equal opponent, for the leagues in 538's club-football table (seasons 2016-2023 except 2020 and 2021), ±2 SE."

pd <- league_538 |>
  mutate(league = factor(league, levels = league),
         text = paste0(league, "<br>h = ", fmt(h), " (SE ", fmt(h_se), ")",
                       "<br>equal teams: home win ", pct(p_win), ", draw ", pct(p_draw),
                       ", away win ", pct(p_loss), "<br>", n_units, " seasons, ",
                       format(n_games, big.mark = ","), " games"))
p <- ggplot(pd, aes(share, league, text = text)) +
  geom_vline(xintercept = 0.5, colour = irw_grey) +
  geom_vline(xintercept = median(pd$share), colour = irw_blue, linetype = "dashed", linewidth = 0.3) +
  geom_linerange(aes(xmin = share - 2 * share_se, xmax = share + 2 * share_se), colour = irw_blue, alpha = 0.6) +
  geom_point(colour = irw_blue, size = 1.6) +
  scale_x_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(x = "home side's expected share of the points, equal teams", y = NULL)
ggplotly(p, tooltip = "text")
```

The same measure across `r iv(nrow(league_538))` club leagues in the late 2010s
and early 2020s ranges from `r iv(pct(min(league_538$share)))` in
`r iv(as.character(league_538$league[1]))` to
`r iv(pct(max(league_538$share)))` in
`r iv(as.character(league_538$league[nrow(league_538)]))` (median
`r iv(pct(median(league_538$share)))`). Some of that spread is noise, since a
league contributes only a few seasons here; the error bars show how much.

### Has it declined?

```{r}
#| label: fig-trend
#| fig-height: 8
#| fig-cap: "Home advantage h by season, with ±2 SE bars. 2020 and 2021, the seasons played largely without crowds, are in red. Rugby and internationals are estimated by decade."

trend_df <- summary_df |>
  filter(sport != "Football (538 leagues)") |>
  mutate(covid = period %in% covid & league %in% season_series,
         year = if_else(league %in% season_series, period, period + 5),
         text = paste0(league, " ", period, if_else(league %in% season_series, "", "s"),
                       "<br>h = ", fmt(h), " (SE ", fmt(h_se), ")<br>", n_games, " games, ",
                       n_agents, " teams<br>", table))
p <- ggplot(trend_df, aes(year, h, text = text)) +
  geom_hline(yintercept = 0, colour = irw_grey, linewidth = 0.3) +
  geom_linerange(aes(ymin = h - 2 * h_se, ymax = h + 2 * h_se, colour = covid), alpha = 0.4) +
  geom_point(aes(colour = covid), size = 0.9) +
  geom_smooth(data = filter(trend_df, !covid), mapping = aes(year, h, weight = 1 / h_se^2),
              method = "loess", span = 0.5, se = FALSE, colour = irw_blue, linewidth = 0.6,
              inherit.aes = FALSE) +
  scale_colour_manual(values = c(`FALSE` = "grey30", `TRUE` = irw_red), guide = "none") +
  facet_wrap(~ league, ncol = 3) +
  labs(x = NULL, y = "home advantage h (logits)")
ggplotly(p, tooltip = "text")
```

```{r}
#| label: tbl-slopes
#| echo: false

slope <- function(d) {
  m <- lm(h ~ I(period / 10), data = d, weights = 1 / h_se^2)
  co <- summary(m)$coefficients
  tibble(slope = co[2, 1], se = co[2, 2], n = nrow(d))
}
slopes <- summary_df |>
  filter(league %in% season_series, !period %in% covid) |>
  group_by(league) |>
  group_modify(~ bind_rows(
    slope(.x) |> mutate(span = "whole series"),
    slope(filter(.x, period >= 1990, period < 2020)) |> mutate(span = "1990-2019")
  )) |>
  ungroup() |>
  mutate(est = paste0(fmt(slope, 3), " (", fmt(se, 3), ")")) |>
  select(league, span, est) |>
  pivot_wider(names_from = span, values_from = est)
first_season <- summary_df |> filter(league %in% season_series) |> group_by(league) |> summarise(from = min(period))
slopes <- left_join(slopes, first_season, by = "league") |> select(league, from, everything())
knitr::kable(slopes, col.names = c("Series", "From", "Whole series", "1990-2019"),
             caption = "Change in h per decade (SE), from a precision-weighted regression of season estimates on year. 2020 and 2021 are left out of both.")
```

```{r}
#| label: slope-numbers
#| include: false
sl <- summary_df |>
  filter(league %in% season_series, !period %in% covid) |>
  group_by(league) |>
  group_modify(~ bind_cols(slope(.x), slope(filter(.x, period >= 1990, period < 2020)) |> rename_with(~ paste0(.x, "_90")))) |>
  ungroup()
n_down <- sum(sl$slope + 2 * sl$se < 0)
n_down90 <- sum(sl$slope_90 + 2 * sl$se_90 < 0)
```

Over the whole of each series, home advantage has fallen in `r iv(n_down)` of the
`r iv(nrow(sl))` leagues measured season by season, by more than twice its standard
error. The steepest declines are in European football: in English football $h$ has
roughly halved since before 1950, and in Serie A it has fallen by about 40%. Among
the North American sports the decline is clearest in hockey and in basketball,
where $h$ rose to a peak in the middle of the century before falling. Baseball's
small advantage has drifted down slowly, and has been essentially flat since 1990.
American football shows no long-run trend. Since 1990 the decline is significant
in `r iv(n_down90)` series (basketball, hockey and both football leagues); the NFL's
estimate points the same way but is too noisy to say.

This is the familiar result, and it is worth seeing that it survives conditioning
on team strength: the decline is not an artefact of who hosted whom. What the
model cannot say is *why*. Better travel, standardised stadiums, and changes in
officiating are all candidates, and all of them moved over the same decades.

Rugby and international football, estimated by decade, show no such decline. The
estimates are noisy, but they are also a different kind of home advantage: a
national side playing at home has often travelled less far, in a stadium it chose,
in front of a crowd that is almost entirely its own.

### Did empty stadiums shrink it?

```{r}
#| label: fig-nocrowd
#| fig-height: 9
#| fig-cap: "Change in home advantage (δ, logits) in games played without crowds, by league, ±2 SE. Negative means home advantage shrank. Each estimate is net of team-season strength, from the 2016-2023 seasons."

nc <- nocrowd_df |>
  mutate(label = if_else(definition == "attendance column", paste(league, "(attendance)"), league),
         group = if_else(sport == "Football (538 leagues)", "Club football", "North American leagues"),
         text = paste0(label, "<br>δ = ", fmt(delta), " (SE ", fmt(delta_se), ")<br>h with crowds = ",
                       fmt(h), "<br>", n_nocrowd, " home games without a crowd")) |>
  arrange(delta) |>
  mutate(label = factor(label, levels = label))
p <- ggplot(nc, aes(delta, label, text = text, colour = group)) +
  geom_vline(xintercept = 0, colour = irw_grey) +
  geom_linerange(aes(xmin = delta - 2 * delta_se, xmax = delta + 2 * delta_se), alpha = 0.6) +
  geom_point(size = 1.5) +
  scale_colour_manual(values = c("Club football" = irw_blue, "North American leagues" = irw_red)) +
  labs(x = "change in home advantage without crowds (δ, logits)", y = NULL, colour = NULL)
ggplotly(p, tooltip = "text") |> layout(legend = list(orientation = "h", y = -0.08))
```

```{r}
#| label: nocrowd-numbers
#| include: false
foot <- filter(nocrowd_df, sport == "Football (538 leagues)")
w <- 1 / foot$delta_se^2
pooled <- sum(w * foot$delta) / sum(w); pooled_se <- sqrt(1 / sum(w))
Q <- sum(w * (foot$delta - pooled)^2); Q_df <- nrow(foot) - 1
pooled_h <- sum(foot$h / foot$h_se^2) / sum(1 / foot$h_se^2)
na <- function(lg, def = "date window") filter(nocrowd_df, league == lg, definition == def)
```

In club football the answer is yes. Home advantage fell in
`r iv(sum(foot$delta < 0))` of the `r iv(nrow(foot))` leagues, and the precision-weighted
average change is `r iv(fmt(pooled))` logits (SE `r iv(fmt(pooled_se))`), against a typical
advantage with crowds of about `r iv(fmt(pooled_h))`: about
`r iv(round(-100 * pooled / pooled_h))`% of it went with the crowd. The leagues differ by more than chance would explain (heterogeneity
$Q$ = `r iv(fmt(Q, 1))` on `r iv(Q_df)` df). Several of the largest drops were in top divisions
(the Premier League, Ligue 1, the Bundesliga), while two second divisions, Ligue 2
and the 2. Bundesliga, show none (England's lower divisions did lose some). @fischer2021crowd found exactly this
pattern for Germany and traced it to occupancy: a stadium that is usually half
empty loses less when it is fully empty.

The North American leagues split. Basketball lost home advantage
($\delta$ = `r iv(fmt(na("Basketball (NBA)")$delta))`, SE `r iv(fmt(na("Basketball (NBA)")$delta_se))`),
as did American football, more noisily. Hockey barely moved. Baseball did not lose
any: the estimate is positive whether "no crowd" is read from the calendar
($\delta$ = `r iv(fmt(na("Baseball (MLB)")$delta))`) or from the attendance column
($\delta$ = `r iv(fmt(na("Baseball (MLB)", "attendance column")$delta))`, SE
`r iv(fmt(na("Baseball (MLB)", "attendance column")$delta_se))`). @higgs2021bayesian,
with a different model, found home advantage fell in the NBA's and NHL's 2020
playoffs, played in bubbles, and changed little in MLB and the NFL. We agree on
basketball and baseball. Our hockey and American football estimates come from the
regular season, not the bubble playoffs, which may be why they differ. Baseball's
advantage was small to begin with, and the sport has a structural home edge (the home side bats last)
that does not depend on anyone watching.

### Crowd or travel? The NBA bubble

```{r}
#| label: tbl-bubble
#| echo: false

if (!is.null(nba_bubble)) {
  tm <- nba_bubble$terms
  bub <- tibble(
    Setting = c("Normal seasons (crowd and travel)",
                "2020 bubble (no crowd, no travel)",
                "2020-21 season (travel, no crowd)"),
    `Home games` = c(NA, tm$n_games[tm$term == "bubble"], tm$n_games[tm$term == "season_2021"]),
    h = c(nba_bubble$h, nba_bubble$h + tm$delta),
    `Change from normal` = c(NA, tm$delta),
    SE = c(nba_bubble$h_se, tm$delta_se)
  )
  knitr::kable(bub, digits = 2, caption = "NBA home advantage, 2016-2023 seasons with playoffs; the 2021 playoffs, when crowds were partly back, are left out. The SE is for h in the first row and for the change in the others.")
}
```

The NBA gives one more contrast. Its 2020 restart was played in a bubble at Walt
Disney World: no crowd, and no travel either, since every team lived on site. The
season that followed was played in home arenas, with travel but with few or no
fans. 538 flags the bubble games as neutral, but its `agent_a` is still the
designated home team (in the 2020 Finals, the Lakers in games 1, 2 and 5 and the
Heat in 3, 4 and 6, as scheduled), so we can estimate a nominal home advantage for
them. It nearly vanished in the bubble and about halved in 2020-21. With only
`r iv(if (!is.null(nba_bubble)) nba_bubble$terms$n_games[1] else NA)` bubble games the
first estimate is imprecise, and the two do not differ significantly, but the
ordering is what you would expect if both the crowd and travel contribute.

## Limitations

- **One no-crowd window for every league.** Leagues closed and reopened at
  different times: some readmitted limited crowds in autumn 2020, MLB from April
  2021, while Brazil stayed closed into late 2021. Misclassified games pull
  $\delta$ toward zero, so the estimates understate any effect rather than invent
  one. The MLB attendance check, which needs no window, gives the same answer as
  the window does.
- **Neither design isolates the crowd.** The no-crowd seasons also changed
  schedules, travel restrictions, quarantines and, in some leagues, venues. The
  bubble comparison separates crowd from travel only for the NBA and only on 172
  games.
- **Strengths are fixed within a unit.** Teams improve and decline within a season,
  and much more within the decade used for rugby and the internationals. That
  matters little for $h$ in a balanced league schedule, more for the
  internationals, where who hosts is not balanced.
- **Dropped teams.** Teams with fewer than five games, or a perfect record, are
  dropped before fitting. For the internationals this removes the weakest
  sides, and in the 2000s the remaining women's schedule still separates, so that
  decade has no estimate.
- **The trend is descriptive.** A straight line per series summarises a decline;
  it does not say when or why it happened, and the per-season estimates are
  correlated within a series, so the slope SEs are optimistic.
- **Home advantage is averaged over teams.** Some venues (altitude, artificial
  pitches, long travel) carry more than others; one $h$ per league ignores that.

## Reproducibility

{{< include ../components/_data-version.qmd >}}

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the compute script (from the project root)
source("vignettes/home_advantage_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/home_advantage.qmd")
```

{{< include ../components/_source-links.qmd >}}

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(unique(summary_df$table), output_file = "home_advantage_data/references.bib",
                source = "comp")
```
:::

---

## References
