---
title: "Item-Level Heterogeneous Treatment Effects"
description: |
  When an RCT uses a psychometric instrument as its outcome, different items
  may respond differently to treatment. We fit item-level HTE models across
  IRW datasets and show how ignoring this variation can bias
  treatment-by-covariate interaction estimates.
date: today
bibliography: ilhtedata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "ilhtedata/il_hte_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache     <- readRDS(cache_path)
  summary_df <- .cache$summary
  item_df    <- .cache$item_df
  date_run   <- .cache$date_run
} else {
  summary_df <- tibble(tab = character(), n_items = integer(), n_persons = integer(),
                       ate_std = numeric(), sigma_zeta_std = numeric(), rho = numeric(),
                       sig_il_hte = logical(), sd_ratio = numeric(),
                       beta3_b = numeric(), beta3_c = numeric(), delta_beta3 = numeric())
  item_df    <- tibble(tab = character(), item = character(), total_std = numeric(),
                       ate_std = numeric())
  date_run   <- Sys.Date()
}
```

## Overview

Many RCTs in education, economics, and health research use psychometric
instruments — tests, surveys, symptom checklists — as outcome measures. A
standard analysis sums responses into a single score and estimates the average
treatment effect (ATE) on that score. But the individual items of the
instrument may not respond equally to treatment. An item that directly targets
the intervention's content may show a large positive effect while an unrelated
item shows none.

This item-level variation is called **item-level heterogeneous treatment
effects** (IL-HTE) [@gilbert2025estimating]. Formally, if the log-odds of a
correct response for item $i$ by person $j$ follows

$$\text{logit}\,P(Y_{ij} = 1) = \theta_j + b_i + \zeta_i\, T_j,$$

then $\zeta_i$ is the residual treatment effect on item $i$ above and beyond
the overall ATE $\beta_1$. When the variance $\sigma_\zeta^2 > 0$, items
respond heterogeneously to treatment. In `lme4` notation this is:

```r
glmer(resp ~ treat + (1|id) + (treat|item), family = binomial)
```

Beyond masking item-level variation, IL-HTE has a less obvious consequence
for causal identification. When $\zeta_i$ is correlated with item difficulty
$b_i$ (parameterized as $\rho \ne 0$), a conventional treatment-by-covariate
interaction model — the standard tool for asking "who benefits most?" — can
produce biased estimates even when no true person-level heterogeneity exists.
@gilbert2024disentangling prove this analytically and demonstrate it through
simulation. We illustrate both insights below using RCT datasets from the IRW.

## Item-level treatment effects

We fit the IL-HTE model to each dataset, extract empirical Bayes estimates of
the item-specific effects $\hat\beta_1 + \hat\zeta_i$, and standardize by
$\hat\sigma_\theta$ — the person-level SD from the baseline random-intercepts
model — so that effects are comparable across datasets. The ATE $\hat\beta_1$
is shown as a large point; individual item effects as small points. Datasets
are ordered by ATE. Color encodes whether the IL-HTE variance $\sigma_\zeta$
is statistically significant (likelihood ratio test, $p < 0.05$, with the
boundary correction of dividing the reported $p$-value by 2).

```{r}
#| label: fig-item-effects
#| fig-cap: "Empirical Bayes estimates of item-specific treatment effects, standardized by σ_θ. Small points: individual item effects (β₁ + ζ̂ᵢ)/σ_θ. Hover over the large dot to see dataset details."

library(plotly)

tab_order <- summary_df |> arrange(ate_std) |> pull(tab)

plot_df <- item_df |>
  left_join(summary_df |> select(tab, ate_std, sig_il_hte), by = "tab") |>
  mutate(
    y_pos     = as.integer(factor(tab, levels = tab_order)),
    sig_label = if_else(sig_il_hte, "Significant", "Not significant"),
    color_val = if_else(sig_il_hte, irw_blue, irw_red)
  )

ate_df <- summary_df |>
  mutate(
    y_pos     = as.integer(factor(tab, levels = tab_order)),
    sig_label = if_else(sig_il_hte, "Significant", "Not significant"),
    color_val = if_else(sig_il_hte, irw_blue, irw_red),
    hover_text = paste0(
      "<b>", tab, "</b><br>",
      "ATE: ", round(ate_std, 3), " SD<br>",
      "σζ: ", round(sigma_zeta_std, 3), " SD<br>",
      "ρ: ", round(rho, 2), "<br>",
      "Items: ", n_items, "<br>",
      "N: ", n_persons, "<br>",
      "Sig. IL-HTE: ", sig_il_hte
    )
  )

fig <- plot_ly() |>
  add_trace(
    data       = plot_df,
    x          = ~total_std,
    y          = ~y_pos,
    type       = "scatter",
    mode       = "markers",
    marker     = list(size = 4, opacity = 0.4),
    color      = ~sig_label,
    colors     = c("Significant" = irw_blue, "Not significant" = irw_red),
    hoverinfo  = "none",
    showlegend = TRUE
  ) |>
  add_trace(
    data       = ate_df,
    x          = ~ate_std,
    y          = ~y_pos,
    type       = "scatter",
    mode       = "markers",
    marker     = list(size = 10, line = list(width = 1, color = "white")),
    color      = ~sig_label,
    colors     = c("Significant" = irw_blue, "Not significant" = irw_red),
    text       = ~hover_text,
    hoverinfo  = "text",
    showlegend = FALSE
  ) |>
  layout(
    height = max(400, nrow(summary_df) * 14),
    xaxis = list(title = "Item-specific treatment effect (SDs)",
                 zeroline = TRUE, zerolinecolor = "#aaa", zerolinewidth = 1),
    yaxis = list(title = "", showticklabels = FALSE, showgrid = FALSE),
    legend = list(orientation = "h", x = 0, y = -0.05),
    margin = list(l = 20)
  )

fig
```

```{r}
#| label: il-hte-summary
#| echo: false

n_sig <- sum(summary_df$sig_il_hte, na.rm = TRUE)
n_tot <- nrow(summary_df)
```

Across `r n_tot` datasets, `r n_sig` show statistically significant IL-HTE
($p < 0.05$). In datasets with wide spreads of item effects, a single-number
ATE obscures meaningful variation in which aspects of the outcome respond to
the intervention — information that would be invisible in a sum-score analysis.

## The identification problem

A more subtle consequence of IL-HTE concerns the estimation of
treatment-by-covariate interactions. @gilbert2024disentangling show
analytically that when $\rho \ne 0$, the two distinct data-generating
processes — (1) true person-level HTE via a treatment × pretest interaction,
and (2) item-level HTE correlated with item difficulty — produce *identical*
patterns in sum-score outcomes. A model that omits IL-HTE therefore cannot
distinguish between these processes and will produce biased estimates of the
treatment-by-covariate interaction.

Specifically, when $\rho > 0$ (easier items respond more strongly to
treatment), treatment "stretches" the item difficulty distribution — the ratio
$\sigma_b^\text{treat}/\sigma_b^\text{ctrl}$ exceeds 1 — and the
constant-item model attributes this variation to a spurious negative
treatment-by-pretest interaction.

We illustrate this empirically by fitting two additional models per dataset:

- **M_B** (constant item effects + person HTE):
  `resp ~ treat + cov + treat:cov + (1|item) + (1|id)`
- **M_C** (flexible EIRM — IL-HTE + person HTE):
  `resp ~ treat + cov + treat:cov + (treat|item) + (1|id)`

where `cov` is the standardized pretest score. The $\hat\beta_3$ coefficient
on `treat:cov` represents the treatment-by-covariate interaction; M_B
estimates it without allowing for IL-HTE, M_C estimates it alongside IL-HTE.

```{r}
#| label: fig-identification
#| fig-cap: "Difference in treatment-by-covariate interaction estimates between M_B (constant item effects) and M_C (with IL-HTE), as a function of the ratio of item-difficulty SDs in treatment vs. control groups. Color encodes ρ; point size encodes number of items. Datasets with ratio > 1 and ρ > 0 show negative differences, consistent with the theoretical prediction."
#| fig-height: 4.5

label_df <- summary_df |>
  mutate(tab_short = sub("gilbert_meta_", "", tab))

ggplot(label_df, aes(x = sd_ratio, y = delta_beta3, colour = rho)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = irw_grey) +
  geom_point(aes(size = n_items)) +
  geom_text(aes(label = tab_short), vjust = -0.9, size = 3, show.legend = FALSE) +
  scale_colour_gradient2(
    low      = irw_blue,
    mid      = "grey80",
    high     = irw_red,
    midpoint = 0,
    name     = expression(rho)
  ) +
  scale_size_continuous(name = "Items", range = c(3, 8)) +
  labs(
    x = expression(sigma[b]^{treat} / sigma[b]^{ctrl}),
    y = expression(hat(beta)[3](M[B]) - hat(beta)[3](M[C]))
  )
```

Datasets to the right of the vertical reference line (ratio > 1) tend to show
negative $\hat\beta_3(M_B) - \hat\beta_3(M_C)$: the conventional model
produces a more negative interaction estimate than the flexible model,
consistent with the theoretical predictions of @gilbert2024disentangling.
This pattern arises because the correlation between item easiness and
item-specific treatment effects ($\rho$) creates the appearance of
heterogeneity by pretest score in a sum-score analysis, even if none exists
at the person level.

---

## Reproducibility

Results computed `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
Rscript vignettes/il_hte_compute.R
quarto::quarto_render("vignettes/il_hte.qmd")
```

---

## References
