---
title: "Item-Level Heterogeneous Treatment Effects"
description: |
  When an RCT uses a psychometric instrument as its outcome, different items
  may respond differently to treatment. We fit item-level HTE models across
  IRW datasets and show how ignoring this variation can bias
  treatment-by-covariate interaction estimates.
date: today
bibliography: ilhtedata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "ilhtedata/il_hte_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache     <- readRDS(cache_path)
  summary_df <- .cache$summary |> filter(is.finite(ate_std), is.finite(sigma_zeta_std))
  item_df    <- .cache$item_df |> filter(tab %in% summary_df$tab)
  date_run   <- .cache$date_run
} else {
  summary_df <- tibble(tab = character(), n_items = integer(), n_persons = integer(),
                       ate_std = numeric(), sigma_zeta_std = numeric(), rho = numeric(),
                       sig_il_hte = logical(), sd_ratio = numeric(),
                       beta3_b = numeric(), beta3_c = numeric(), delta_beta3 = numeric())
  item_df    <- tibble(tab = character(), item = character(), total_std = numeric(),
                       ate_std = numeric())
  date_run   <- Sys.Date()
}
```

## Motivation

Many RCTs in education, economics, and health research use psychometric
instruments — tests, surveys, symptom checklists — as outcome measures. A
standard analysis sums responses into a single score and estimates the average
treatment effect (ATE) on that score. But the individual items of the
instrument may not respond equally to treatment. An item that directly targets
the intervention's content may show a large positive effect while an unrelated
item shows none.

This item-level variation is called **item-level heterogeneous treatment
effects** (IL-HTE) [@gilbert2025estimating]. Beyond masking item-level
variation, IL-HTE has a less obvious consequence for causal identification.
When item-level effects are correlated with item difficulty, a conventional
treatment-by-covariate interaction model — the standard tool for asking "who
benefits most?" — can produce biased estimates even when no true person-level
heterogeneity exists. @gilbert2024disentangling prove this analytically and
demonstrate it through simulation. We illustrate both insights below using RCT
datasets from the IRW.

---

## Data and methods

### Dataset selection

We use IRW tables that record a binary `treat` indicator and have at least 100
participants (`irw_filter(var = "treat", n_participants = c(100, Inf))`). For
datasets with a `wave` column (multiple follow-up waves), only wave 1 is used.
Items are restricted to those with exactly two response categories; datasets
with fewer than three dichotomous items after that restriction are dropped.
Respondent counts are downsampled to at most 5,000 IDs for large datasets to
keep model fitting tractable.

### The IL-HTE model

Formally, if the log-odds of a correct response for item $i$ by person $j$
follows

$$
\begin{align}
    \text{logit}(\Pr(Y_{ij} = 1)) = \eta_{ij} &= \theta_j + b_{i} + \zeta_i T_j \\
    \theta_j &= \beta_0 + \beta_1 T_j + \varepsilon_j \\
    \begin{bmatrix}
        b_i \\
        \zeta_{i}
    \end{bmatrix}
     &\sim N\left(\begin{bmatrix}
         0 \\ 0
     \end{bmatrix},\begin{bmatrix}
         \sigma^2_b & \rho\sigma_b\sigma_\zeta \\
         \rho\sigma_b\sigma_\zeta & \sigma^2_\zeta
     \end{bmatrix}\right) \\
    \varepsilon_j &\sim N(0, \sigma^2_\theta)
\end{align}
$$

then $\zeta_i$ is the residual treatment effect on item $i$ above and beyond
the overall ATE $\beta_1$. When the variance $\sigma_\zeta^2 > 0$, items
respond heterogeneously to treatment. In `lme4` notation this is:

```r
glmer(resp ~ treat + (1|id) + (treat|item), family = binomial)
```

We fit this model to each dataset, extract empirical Bayes estimates of the
item-specific effects $\hat\beta_1 + \hat\zeta_i$, and standardize by
$\hat\sigma_\theta$ — the person-level SD from a baseline random-intercepts
model with no treatment slope — so that effects are comparable across
datasets. Statistical significance of $\sigma_\zeta$ uses a likelihood-ratio
test against that baseline model, with the standard boundary correction of
halving the reported p-value (the null variance of zero sits at the edge of
the parameter space).

### Model comparison for the identification problem

We illustrate the identification problem empirically by fitting two
additional models per dataset, wherever a pretest covariate is available:

- **M_B** (constant item effects + person HTE):
  `resp ~ treat + cov + treat:cov + (1|item) + (1|id)`
- **M_C** (flexible EIRM — IL-HTE + person HTE):
  `resp ~ treat + cov + treat:cov + (treat|item) + (1|id)`

where `cov` is the standardized pretest score: `std_baseline` where the
dataset provides it, or otherwise whichever `cov_`-prefixed column is most
correlated with the response in the control arm alone. Datasets with no
usable pretest covariate contribute to the item-level-effects analysis above
but are excluded from this comparison. The $\hat\beta_3$ coefficient on
`treat:cov` represents the treatment-by-covariate interaction; M_B estimates
it without allowing for IL-HTE, M_C estimates it alongside IL-HTE.

---

## Results

### Item-level treatment effects

The ATE $\hat\beta_1$ is shown as a large point; individual item effects as
small points. Datasets are ordered by ATE. Color encodes whether the IL-HTE
variance $\sigma_\zeta$ is statistically significant.

```{r}
#| label: fig-item-effects
#| fig-cap: "Empirical Bayes estimates of item-specific treatment effects, standardized by σ_θ. Small points: individual item effects (β₁ + ζ̂ᵢ)/σ_θ. Hover over the large dot to see dataset details."

tab_order <- summary_df |> arrange(ate_std) |> pull(tab)

plot_df <- item_df |>
  left_join(summary_df |> select(tab, ate_std, sig_il_hte), by = "tab") |>
  mutate(
    y_pos     = as.integer(factor(tab, levels = tab_order)),
    sig_label = if_else(sig_il_hte, "Significant", "Not significant"),
    color_val = if_else(sig_il_hte, irw_blue, irw_red)
  )

ate_df <- summary_df |>
  mutate(
    y_pos     = as.integer(factor(tab, levels = tab_order)),
    sig_label = if_else(sig_il_hte, "Significant", "Not significant"),
    color_val = if_else(sig_il_hte, irw_blue, irw_red),
    hover_text = paste0(
      "<b>", tab, "</b><br>",
      "ATE: ", round(ate_std, 3), " SD<br>",
      "σζ: ", round(sigma_zeta_std, 3), " SD<br>",
      "ρ: ", round(rho, 2), "<br>",
      "Items: ", n_items, "<br>",
      "N: ", n_persons, "<br>",
      "Sig. IL-HTE: ", sig_il_hte
    )
  )

fig <- plot_ly() |>
  add_trace(
    data       = plot_df,
    x          = ~total_std,
    y          = ~y_pos,
    type       = "scatter",
    mode       = "markers",
    marker     = list(size = 4, opacity = 0.4),
    color      = ~sig_label,
    colors     = c("Significant" = irw_blue, "Not significant" = irw_red),
    hoverinfo  = "none",
    showlegend = TRUE
  ) |>
  add_trace(
    data       = ate_df,
    x          = ~ate_std,
    y          = ~y_pos,
    type       = "scatter",
    mode       = "markers",
    marker     = list(size = 10, line = list(width = 1, color = "white")),
    color      = ~sig_label,
    colors     = c("Significant" = irw_blue, "Not significant" = irw_red),
    text       = ~hover_text,
    hoverinfo  = "text",
    showlegend = FALSE
  ) |>
  layout(
    height = max(400, nrow(summary_df) * 14),
    xaxis = list(title = "Item-specific treatment effect (SDs)",
                 zeroline = TRUE, zerolinecolor = "#aaa", zerolinewidth = 1),
    yaxis = list(title = "", showticklabels = FALSE, showgrid = FALSE),
    legend = list(orientation = "h", x = 0, y = -0.05),
    margin = list(l = 20)
  )

fig
```

```{r}
#| label: il-hte-summary
#| echo: false

n_sig <- sum(summary_df$sig_il_hte, na.rm = TRUE)
n_tot <- nrow(summary_df)
```

Across `r n_tot` datasets, `r n_sig` show statistically significant IL-HTE
($p < 0.05$). In datasets with wide spreads of item effects, a single-number
ATE obscures meaningful variation in which aspects of the outcome respond to
the intervention — information that would be invisible in a sum-score analysis.

### The identification problem

@gilbert2024disentangling show analytically that when item-level effects are
correlated with item difficulty ($\rho \ne 0$), two distinct data-generating
processes — (1) true person-level HTE via a treatment × pretest interaction,
and (2) item-level HTE correlated with item difficulty — produce *identical*
patterns in sum-score outcomes. A model that omits IL-HTE therefore cannot
distinguish between these processes and will produce biased estimates of the
treatment-by-covariate interaction.

Specifically, when $\rho > 0$ (easier items respond more strongly to
treatment), treatment "stretches" the item difficulty distribution — the ratio
$\sigma_b^\text{treat}/\sigma_b^\text{ctrl}$ exceeds 1 — and the
constant-item model attributes this variation to a spurious negative
treatment-by-pretest interaction.

```{r}
#| label: fig-identification
#| fig-cap: "Difference in treatment-by-covariate interaction estimates between M_B (constant item effects) and M_C (with IL-HTE), as a function of the ratio of item-difficulty SDs in treatment vs. control groups. Hover over a point for dataset details."

id_df <- summary_df |>
  filter(!is.na(delta_beta3)) |>
  mutate(
    hover_text = paste0(
      "<b>", tab, "</b><br>",
      "Δβ₃: ", round(delta_beta3, 3), "<br>",
      "SD ratio: ", round(sd_ratio, 3), "<br>",
      "ρ: ", round(rho, 2), "<br>",
      "Items: ", n_items, "<br>",
      "N: ", n_persons
    )
  )

plot_ly(id_df,
  x         = ~sd_ratio,
  y         = ~delta_beta3,
  type      = "scatter",
  mode      = "markers",
  marker    = list(
    size        = ~sqrt(n_items) * 2,
    color       = ~rho,
    colorscale  = list(c(0, irw_blue), c(0.5, "#d3d3d3"), c(1, irw_red)),
    cmin        = -1, cmax = 1,
    colorbar    = list(title = "ρ"),
    line        = list(width = 0.5, color = "white")
  ),
  text      = ~hover_text,
  hoverinfo = "text"
) |>
layout(
  xaxis = list(title = "σ_b(treat) / σ_b(ctrl)",
               zeroline = FALSE,
               zerolinecolor = "#aaa",
               showline = TRUE),
  yaxis = list(title = "β₃(M_B) − β₃(M_C)",
               zeroline = TRUE, zerolinecolor = "#aaa", zerolinewidth = 1),
  shapes = list(
    list(type = "line", x0 = 1, x1 = 1,
         y0 = ~min(delta_beta3, na.rm=TRUE), y1 = ~max(delta_beta3, na.rm=TRUE),
         line = list(dash = "dash", color = "#aaa"))
  )
)
```

Datasets to the right of the vertical reference line (ratio > 1) tend to show
negative $\hat\beta_3(M_B) - \hat\beta_3(M_C)$: the conventional model
produces a more negative interaction estimate than the flexible model,
consistent with the theoretical predictions of @gilbert2024disentangling.
This pattern arises because the correlation between item easiness and
item-specific treatment effects ($\rho$) creates the appearance of
heterogeneity by pretest score in a sum-score analysis, even if none exists
at the person level.

---

## Limitations

**Dichotomous items only.** Items are restricted to those with exactly two
response categories; polytomous RCT outcome instruments either lose those
items or, if fewer than three dichotomous items remain, are dropped from the
analysis entirely. IL-HTE in polytomous outcomes is not examined here.

**Only the first wave is used.** For datasets with multiple follow-up waves,
we use wave 1 only — this vignette doesn't examine whether item-level
treatment effects change over time or across follow-up assessments.

**The pretest covariate is chosen from the data.** Where a dataset has more
than one `cov_`-prefixed candidate, we pick whichever is most correlated with
the response in the control arm — a data-driven choice made before fitting
M_B and M_C, rather than a pre-registered or theory-driven selection. This
could inflate the apparent size of the M_B/M_C divergence relative to a
design that fixed the covariate in advance.

**Large datasets are downsampled** to at most 5,000 respondent IDs, so
results for the largest RCTs in this collection are based on a random subset
rather than the full sample.

---

## Reproducibility

Results computed `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
Rscript vignettes/il_hte_compute.R
quarto::quarto_render("vignettes/il_hte.qmd")
```

---

## References
