---
title: "Does Item Language Predict Difficulty? Text Features Across Cognitive Datasets"
description: |
  A Tier 2 IRW vignette linking item text to empirical difficulty. We extract 
  readability, length, negation, and other linguistic features from item text, 
  then ask how well they predict proportion correct across cognitive datasets.
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)
library(tidyr)

.cache          <- readRDS("itemtextdata/item_text_difficulty_results.rds")
all_data        <- .cache$all_data
eligible_tables <- .cache$eligible_tables
```

## Research question

Cognitive items vary enormously in difficulty. Some of that variation is substantive —
harder content is harder to answer. But some may be linguistic: items written in more
complex language, or containing negation, may be harder to process independent of their
content. Can simple text features extracted from item wording predict how often
respondents answer correctly?

The IRW's item text layer makes this question tractable. For tables where both response
data and item text are available, we can compute proportion correct per item, extract
linguistic features, and ask how much variance those features explain.

---

## Step 1: Select datasets with item text

We restrict to cognitive/educational dichotomous English tables that have item text
available via `irw_itemtext()`. Not all IRW tables have item text — this analysis
uses only those that do.

```{r}
#| label: filter
#| eval: false

# Tables with item text
text_tables <- irw_list_itemtext_tables()

# Cognitive/educational dichotomous English tables with enough respondents
meta <- irw_filter(
  construct_type   = "Cognitive/educational",
  n_categories     = 2,
  n_participants   = c(200, Inf),
  primary_language = "eng"
)

# Intersection: must satisfy both criteria
eligible_tables <- intersect(text_tables, meta)
```

This yielded `r length(eligible_tables)` tables
(as of `r format(.cache$date_run, "%B %Y")`), covering
`r n_distinct(all_data$table)` tables with usable item text after joining.

---

## Step 2: Compute proportion correct and extract text features

For each table we (1) fetch response data, (2) fetch item text, (3) coerce columns
to consistent types across tables, (4) compute proportion correct per item, and
(5) join the two on the `item` identifier.

```{r}
#| label: data-fetch
#| eval: false

process_table <- function(table_name) {
  df <- tryCatch(irw_fetch(table_name), error = function(e) NULL)
  if (is.null(df)) return(NULL)

  item_text <- tryCatch(irw_itemtext(table_name), error = function(e) NULL)
  if (is.null(item_text) || nrow(item_text) == 0) return(NULL)

  # Coerce to consistent types — resp and item vary across tables
  df <- df |> mutate(
    resp = suppressWarnings(as.numeric(resp)),
    item = as.character(item)
  )

  # Process item text — detect text column name and coerce item to character
  text_col  <- setdiff(names(item_text), "item")[1]
  item_text <- item_text |>
    rename(item_text_raw = all_of(text_col)) |>
    mutate(item = as.character(item)) |>
    filter(!is.na(item_text_raw), nchar(trimws(item_text_raw)) > 0)

  prop_correct <- df |>
    group_by(item) |>
    summarise(
      prop_correct = mean(resp == 1, na.rm = TRUE),
      n_resp       = sum(!is.na(resp)),
      .groups      = "drop"
    )

  inner_join(prop_correct, item_text, by = "item") |>
    mutate(table = table_name) |>
    select(table, item, prop_correct, n_resp, item_text_raw)
}
```

We then extract the following features from each item's text using `quanteda`:

| Feature | Description |
|---|---|
| `word_count` | Number of words |
| `sent_count` | Number of sentences |
| `avg_word_len` | Mean character length per word |
| `fk_grade` | Flesch-Kincaid grade level |
| `flesch_ease` | Flesch reading ease (higher = easier) |
| `ttr` | Type-token ratio (lexical diversity) |
| `has_negation` | Contains negation word (not, never, no, ...) |
| `has_question` | Question format (ends with ? or starts with question word) |
| `has_number` | Contains a numeral or number word |
| `cap_ratio` | Proportion of non-initial words that are capitalised |

---

## Step 3: Distributions of features and outcome

### Proportion correct

```{r}
#| label: fig-propcorrect
#| fig-cap: "Distribution of proportion correct across items. Each observation is one item."
#| fig-height: 3.5

ggplot(all_data, aes(x = prop_correct)) +
  geom_histogram(bins = 40, fill = "#2166ac", colour = "white", alpha = 0.8) +
  labs(x = "Proportion correct", y = "Count") +
  theme_minimal(base_size = 13)
```

### Continuous features vs. proportion correct

```{r}
#| label: fig-scatter
#| fig-cap: "Proportion correct vs. continuous text features. Each point is one item; lines are loess smooths."
#| fig-height: 8

continuous_features <- c("word_count", "avg_word_len", "fk_grade",
                          "flesch_ease", "ttr", "cap_ratio")

all_data |>
  select(prop_correct, all_of(continuous_features), table) |>
  pivot_longer(all_of(continuous_features),
               names_to = "feature", values_to = "value") |>
  mutate(feature = recode(feature,
    word_count   = "Word count",
    avg_word_len = "Avg word length",
    fk_grade     = "FK grade level",
    flesch_ease  = "Flesch ease",
    ttr          = "Type-token ratio",
    cap_ratio    = "Capitalisation ratio"
  )) |>
  ggplot(aes(x = value, y = prop_correct)) +
  geom_point(alpha = 0.15, size = 0.8, colour = "#2166ac") +
  geom_smooth(method = "loess", se = TRUE, colour = "#d6604d", linewidth = 0.8) +
  facet_wrap(~feature, scales = "free_x", ncol = 2) +
  labs(x = "Feature value", y = "Proportion correct") +
  theme_minimal(base_size = 12)
```

### Binary features vs. proportion correct

```{r}
#| label: fig-binary
#| fig-cap: "Proportion correct by binary text features. Points show means; bars show 95% CIs."
#| fig-height: 4

binary_features <- c("has_negation", "has_question", "has_number")

all_data |>
  select(prop_correct, all_of(binary_features)) |>
  pivot_longer(all_of(binary_features),
               names_to = "feature", values_to = "value") |>
  mutate(
    feature = recode(feature,
      has_negation = "Negation present",
      has_question = "Question format",
      has_number   = "Number present"
    ),
    value = factor(value, levels = c(0, 1), labels = c("No", "Yes"))
  ) |>
  group_by(feature, value) |>
  summarise(
    mean_pc = mean(prop_correct, na.rm = TRUE),
    se_pc   = sd(prop_correct, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) |>
  ggplot(aes(x = value, y = mean_pc, colour = value)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_pc - 1.96 * se_pc,
                    ymax = mean_pc + 1.96 * se_pc),
                width = 0.15) +
  scale_colour_manual(values = c("No" = "#888888", "Yes" = "#2166ac"),
                      guide = "none") +
  facet_wrap(~feature) +
  labs(x = NULL, y = "Mean proportion correct") +
  theme_minimal(base_size = 13)
```

---

## Step 4: How much variance do text features explain?

We fit a simple linear regression of proportion correct on all features, with
dataset fixed effects to account for the fact that overall difficulty varies
across instruments.

```{r}
#| label: regression

mod <- lm(
  prop_correct ~ word_count + avg_word_len + fk_grade + flesch_ease +
                 ttr + cap_ratio + has_negation + has_question + has_number +
                 table,
  data = all_data
)

# R-squared from model with and without table fixed effects
mod_no_fe <- lm(
  prop_correct ~ word_count + avg_word_len + fk_grade + flesch_ease +
                 ttr + cap_ratio + has_negation + has_question + has_number,
  data = all_data
)

r2_no_fe <- round(summary(mod_no_fe)$r.squared, 3)
r2_fe    <- round(summary(mod)$r.squared, 3)

# Coefficient table (excluding table fixed effects)
coef_tbl <- broom::tidy(mod, conf.int = TRUE) |>
  filter(!grepl("^table", term), term != "(Intercept)") |>
  mutate(
    term = recode(term,
      word_count   = "Word count",
      avg_word_len = "Avg word length",
      fk_grade     = "FK grade level",
      flesch_ease  = "Flesch ease",
      ttr          = "Type-token ratio",
      cap_ratio    = "Capitalisation ratio",
      has_negation = "Negation present",
      has_question = "Question format",
      has_number   = "Number present"
    ),
    across(c(estimate, conf.low, conf.high), ~round(.x, 3)),
    p.value = round(p.value, 3)
  ) |>
  select(Feature = term, Estimate = estimate,
         `95% CI low` = conf.low, `95% CI high` = conf.high,
         `p value` = p.value)

coef_tbl
```

Without dataset fixed effects, text features explain **`r r2_no_fe * 100`%** of
variance in proportion correct. Adding fixed effects brings this to
**`r r2_fe * 100`%**, suggesting most explainable variance is between instruments
rather than within them.

```{r}
#| label: fig-coefs
#| fig-cap: "Regression coefficients for text features predicting proportion correct (with dataset fixed effects). Error bars are 95% CIs."
#| fig-height: 4

coef_tbl |>
  mutate(Feature = reorder(Feature, Estimate)) |>
  ggplot(aes(x = Estimate, y = Feature)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = `95% CI low`, xmax = `95% CI high`),
                 height = 0.25, colour = "#888888") +
  geom_point(size = 3, colour = "#2166ac") +
  labs(x = "Coefficient (effect on proportion correct)", y = NULL) +
  theme_minimal(base_size = 13)
```

---

## What to notice

**Between- vs. within-instrument variance.** The jump in R² when adding fixed
effects is the key comparison. If most variance is between instruments, that
tells you overall difficulty is mostly a property of the test (content domain,
target population, test design) rather than item wording per se.

**Negation.** If items with negation show lower proportion correct, that's
consistent with processing load rather than content difficulty — respondents
may answer the item they wish had been asked. Worth checking whether negation
items also show lower discrimination (*a*) in a 2PL.

**Flesch-Kincaid vs. Flesch ease.** These two measures are highly correlated
but scaled in opposite directions. If both appear in the regression with similar
coefficients in opposite directions, that's expected collinearity — consider
dropping one.

**Limitations.** Proportion correct conflates item difficulty with sample ability.
A harder sample makes every item look more difficult. The dataset fixed effects
partially address this, but within-dataset ability variation is not controlled.
The natural next step is to use *b* parameters from the 2PL vignette instead.

---

## Extending this vignette

- **Use *b* instead of proportion correct.** Join `all_data` with `all_results`
  from the 2PL vignette for items that appear in both. IRT difficulty is
  ability-adjusted, making the text feature relationships cleaner.

- **LLM-predicted difficulty.** Use an embedding model to represent item text
  and ask whether semantic similarity predicts residual difficulty after
  accounting for surface features. A natural next step from this vignette.

- **Within-instrument analysis.** Fit the regression separately per table and
  ask whether the feature-difficulty relationship is consistent across
  instruments or highly variable.

---

## Reproducibility

Results computed on `r format(.cache$date_run, "%B %d, %Y")` using
`r length(eligible_tables)` IRW tables. To reproduce:

```r
source("vignettes/item_text_difficulty_compute.R")
quarto::quarto_render("vignettes/item_text_difficulty.qmd")
```

::: {.callout-tip}
To cite datasets used:

```r
for (tbl in eligible_tables) {
  irw_save_bibtex(tbl, output_file = "vignettes/itemtextdata/irw_references.bib",
                  append = TRUE)
}
```
:::
