---
title: "Does Item Language Predict Difficulty? Text Features Across Cognitive Datasets"
description: |
  A Tier 2 IRW vignette linking item text to empirical difficulty. We extract
  simple linguistic features from item text and ask how well they predict
  proportion correct across cognitive datasets.
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)
library(tidyr)
library(tibble)

.cache          <- readRDS("itemtextdata/item_text_difficulty_results.rds")
all_data        <- .cache$all_data
eligible_tables <- .cache$eligible_tables
```

## Research question

Cognitive items vary enormously in difficulty. Some of that variation is substantive —
harder content is harder to answer. But some may be linguistic: items written in more
complex language, or containing negation, may be harder to process independent of their
content. Can simple text features extracted from item wording predict how often
respondents answer correctly?

The IRW's item text layer makes this question tractable. For tables where both response
data and item text are available, we can compute proportion correct per item, extract
linguistic features, and ask how much variance those features explain.

---

## Step 1: Select datasets with item text

We restrict to cognitive/educational dichotomous English tables that have item text
available via `irw_itemtext()`. Not all IRW tables have item text — this analysis
uses only those that do.

```{r}
#| label: filter
#| eval: false

text_tables <- irw_list_itemtext_tables()

meta <- irw_filter(
  construct_type   = "Cognitive/educational",
  n_categories     = 2,
  n_participants   = c(200, Inf),
  primary_language = "eng"
)

eligible_tables <- intersect(text_tables, meta)
```

This yielded `r length(eligible_tables)` tables
(as of `r format(.cache$date_run, "%B %Y")`), covering
`r n_distinct(all_data$table)` tables with usable item text after joining.

---

## Step 2: Compute proportion correct and extract text features

For each table we (1) fetch response data, (2) fetch item text, (3) compute proportion
correct per item, and (4) join the two on the `item` identifier.

```{r}
#| label: data-fetch
#| eval: false

process_table <- function(table_name) {
  df        <- irw_fetch(table_name)
  item_text <- irw_itemtext(table_name)

  prop_correct <- df |>
    mutate(resp = as.numeric(resp), item = as.character(item)) |>
    group_by(item) |>
    summarise(
      prop_correct = mean(resp == 1, na.rm = TRUE),
      n_resp       = sum(!is.na(resp)),
      .groups      = "drop"
    )

  inner_join(prop_correct, item_text, by = "item") |>
    mutate(table = table_name)
}
```

We then extract two simple text features using base R:

| Feature | Description |
|---|---|
| `word_count` | Number of words |
| `avg_word_len` | Mean character length per word (excluding punctuation) |

---

## Step 3: Distributions of features and outcome

### Proportion correct

```{r}
#| label: fig-propcorrect
#| fig-cap: "Distribution of proportion correct across items. Each observation is one item."
#| fig-height: 3.5

ggplot(all_data, aes(x = prop_correct)) +
  geom_histogram(bins = 40, fill = "#2166ac", colour = "white", alpha = 0.8) +
  labs(x = "Proportion correct", y = "Count") +
  theme_minimal(base_size = 13)
```

---

## Step 4: Within-dataset relationships

Pooling all items together can be misleading — overall correlations between text
features and difficulty are heavily influenced by which datasets happen to have longer
or shorter items, not by within-instrument variation. Instead, we compute the
correlation between each feature and proportion correct *separately per dataset*, then
look at the distribution of those correlations across the `r n_distinct(all_data$table)`
datasets.

```{r}
#| label: within-cors

within_cors <- all_data |>
  group_by(table) |>
  summarise(
    n_items      = n(),
    r_word_count = round(cor(word_count,   prop_correct, use = "complete.obs"), 2),
    r_avg_wlen   = round(cor(avg_word_len, prop_correct, use = "complete.obs"), 2),
    .groups = "drop"
  )

within_cors
```

```{r}
#| label: fig-within-cors
#| fig-cap: "Within-dataset correlations between text features and proportion correct. Each point is one dataset; the dashed line marks zero."
#| fig-height: 4

within_cors |>
  pivot_longer(c(r_word_count, r_avg_wlen),
               names_to = "feature", values_to = "r") |>
  mutate(feature = recode(feature,
    r_word_count = "Word count",
    r_avg_wlen   = "Avg word length"
  )) |>
  ggplot(aes(x = r, y = feature)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(height = 0.15, size = 2.5, alpha = 0.7, colour = "#2166ac") +
  labs(x = "Correlation with proportion correct (within dataset)", y = NULL) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: fig-scatter-facet
#| fig-cap: "Word count vs. proportion correct, separately per dataset. Lines are OLS fits."
#| fig-height: 8

ggplot(all_data, aes(x = word_count, y = prop_correct)) +
  geom_point(alpha = 0.4, size = 0.9, colour = "#2166ac") +
  geom_smooth(method = "lm", se = FALSE, colour = "#d6604d", linewidth = 0.7) +
  facet_wrap(~table, scales = "free") +
  labs(x = "Word count", y = "Proportion correct") +
  theme_minimal(base_size = 10) +
  theme(strip.text = element_text(size = 7))
```

---

## What to notice

**Heterogeneity across datasets.** If the within-dataset correlations are scattered
around zero (some positive, some negative), that suggests any overall pooled
relationship is an artifact of dataset-level confounds rather than a genuine
text-difficulty signal.

**Direction of word count effect.** Longer items could be harder (more to parse) or
easier (more context provided). Looking within datasets avoids the confound that
some instruments simply use longer items throughout.

**Limitations.** Proportion correct conflates item difficulty with sample ability.
The natural next step is to use *b* parameters from the 2PL vignette instead, which
are ability-adjusted.

---

## Reproducibility

Results computed on `r format(.cache$date_run, "%B %d, %Y")` using
`r length(eligible_tables)` IRW tables. To reproduce:

```r
source("vignettes/item_text_difficulty_compute.R")
quarto::quarto_render("vignettes/item_text_difficulty.qmd")
```

::: {.callout-tip}
To cite datasets used:

```r
for (tbl in eligible_tables) {
  irw_save_bibtex(tbl, output_file = "itemtextdata/irw_references.bib",
                  append = TRUE)
}
```
:::
