---
title: "Does Item Language Predict Difficulty? Text Features Across Cognitive Datasets"
description: |
  A Tier 2 IRW vignette linking item text to empirical difficulty. We extract
  simple linguistic features from item text and ask how well they predict
  proportion correct across cognitive datasets.
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)
library(tidyr)

.cache          <- readRDS("itemtextdata/item_text_difficulty_results.rds")
all_data        <- .cache$all_data
eligible_tables <- .cache$eligible_tables
```

## Research question

Cognitive items vary enormously in difficulty. Some of that variation is substantive —
harder content is harder to answer. But some may be linguistic: items written in more
complex language, or containing negation, may be harder to process independent of their
content. Can simple text features extracted from item wording predict how often
respondents answer correctly?

The IRW's item text layer makes this question tractable. For tables where both response
data and item text are available, we can compute proportion correct per item, extract
linguistic features, and ask how much variance those features explain.

---

## Step 1: Select datasets with item text

We restrict to cognitive/educational dichotomous English tables that have item text
available via `irw_itemtext()`. Not all IRW tables have item text — this analysis
uses only those that do.

```{r}
#| label: filter
#| eval: false

text_tables <- irw_list_itemtext_tables()

meta <- irw_filter(
  construct_type   = "Cognitive/educational",
  n_categories     = 2,
  n_participants   = c(200, Inf),
  primary_language = "eng"
)

eligible_tables <- intersect(text_tables, meta)
```

This yielded `r length(eligible_tables)` tables
(as of `r format(.cache$date_run, "%B %Y")`), covering
`r n_distinct(all_data$table)` tables with usable item text after joining.

---

## Step 2: Compute proportion correct and extract text features

For each table we (1) fetch response data, (2) fetch item text, (3) compute proportion
correct per item, and (4) join the two on the `item` identifier.

```{r}
#| label: data-fetch
#| eval: false

process_table <- function(table_name) {
  df        <- irw_fetch(table_name)
  item_text <- irw_itemtext(table_name)

  prop_correct <- df |>
    mutate(resp = as.numeric(resp), item = as.character(item)) |>
    group_by(item) |>
    summarise(
      prop_correct = mean(resp == 1, na.rm = TRUE),
      n_resp       = sum(!is.na(resp)),
      .groups      = "drop"
    )

  inner_join(prop_correct, item_text, by = "item") |>
    mutate(table = table_name)
}
```

We then extract the following features from each item's text using base R:

| Feature | Description |
|---|---|
| `word_count` | Number of words |
| `avg_word_len` | Mean character length per word |
| `has_negation` | Contains a negation word (not, never, no, …) |
| `has_question` | Question format (ends with ? or starts with question word) |
| `has_number` | Contains a numeral or number word |

---

## Step 3: Distributions of features and outcome

### Proportion correct

```{r}
#| label: fig-propcorrect
#| fig-cap: "Distribution of proportion correct across items. Each observation is one item."
#| fig-height: 3.5

ggplot(all_data, aes(x = prop_correct)) +
  geom_histogram(bins = 40, fill = "#2166ac", colour = "white", alpha = 0.8) +
  labs(x = "Proportion correct", y = "Count") +
  theme_minimal(base_size = 13)
```

### Continuous features vs. proportion correct

```{r}
#| label: fig-scatter
#| fig-cap: "Proportion correct vs. continuous text features. Lines are loess smooths."
#| fig-height: 5

all_data |>
  select(prop_correct, word_count, avg_word_len, table) |>
  pivot_longer(c(word_count, avg_word_len),
               names_to = "feature", values_to = "value") |>
  mutate(feature = recode(feature,
    word_count   = "Word count",
    avg_word_len = "Avg word length"
  )) |>
  ggplot(aes(x = value, y = prop_correct)) +
  geom_point(alpha = 0.15, size = 0.8, colour = "#2166ac") +
  geom_smooth(method = "loess", se = TRUE, colour = "#d6604d", linewidth = 0.8) +
  facet_wrap(~feature, scales = "free_x", ncol = 2) +
  labs(x = "Feature value", y = "Proportion correct") +
  theme_minimal(base_size = 12)
```

### Binary features vs. proportion correct

```{r}
#| label: fig-binary
#| fig-cap: "Proportion correct by binary text features. Points show means; bars show 95% CIs."
#| fig-height: 4

all_data |>
  select(prop_correct, has_negation, has_question, has_number) |>
  pivot_longer(c(has_negation, has_question, has_number),
               names_to = "feature", values_to = "value") |>
  mutate(
    feature = recode(feature,
      has_negation = "Negation present",
      has_question = "Question format",
      has_number   = "Number present"
    ),
    value = factor(value, levels = c(0, 1), labels = c("No", "Yes"))
  ) |>
  group_by(feature, value) |>
  summarise(
    mean_pc = mean(prop_correct, na.rm = TRUE),
    se_pc   = sd(prop_correct, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) |>
  ggplot(aes(x = value, y = mean_pc, colour = value)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_pc - 1.96 * se_pc,
                    ymax = mean_pc + 1.96 * se_pc),
                width = 0.15) +
  scale_colour_manual(values = c("No" = "#888888", "Yes" = "#2166ac"),
                      guide = "none") +
  facet_wrap(~feature) +
  labs(x = NULL, y = "Mean proportion correct") +
  theme_minimal(base_size = 13)
```

---

## Step 4: How much variance do text features explain?

We fit a linear regression of proportion correct on all features, with dataset fixed
effects to account for overall difficulty varying across instruments.

```{r}
#| label: regression

mod <- lm(
  prop_correct ~ word_count + avg_word_len + has_negation + has_question +
                 has_number + table,
  data = all_data
)

mod_no_fe <- lm(
  prop_correct ~ word_count + avg_word_len + has_negation + has_question +
                 has_number,
  data = all_data
)

r2_no_fe <- round(summary(mod_no_fe)$r.squared, 3)
r2_fe    <- round(summary(mod)$r.squared, 3)

ci <- confint(mod)
coef_tbl <- as.data.frame(summary(mod)$coefficients) |>
  tibble::rownames_to_column("term") |>
  filter(!grepl("^table", term), term != "(Intercept)") |>
  mutate(
    `95% CI low`  = round(ci[term, 1], 3),
    `95% CI high` = round(ci[term, 2], 3),
    term = recode(term,
      word_count   = "Word count",
      avg_word_len = "Avg word length",
      has_negation = "Negation present",
      has_question = "Question format",
      has_number   = "Number present"
    ),
    Estimate = round(Estimate, 3),
    `p value` = round(`Pr(>|t|)`, 3)
  ) |>
  select(Feature = term, Estimate, `95% CI low`, `95% CI high`, `p value`)

coef_tbl
```

Without dataset fixed effects, text features explain **`r r2_no_fe * 100`%** of
variance in proportion correct. Adding fixed effects brings this to
**`r r2_fe * 100`%**, suggesting most explainable variance is between instruments
rather than within them.

```{r}
#| label: fig-coefs
#| fig-cap: "Regression coefficients for text features predicting proportion correct (with dataset fixed effects). Error bars are 95% CIs."
#| fig-height: 3.5

coef_tbl |>
  mutate(Feature = reorder(Feature, Estimate)) |>
  ggplot(aes(x = Estimate, y = Feature)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = `95% CI low`, xmax = `95% CI high`),
                 height = 0.25, colour = "#888888") +
  geom_point(size = 3, colour = "#2166ac") +
  labs(x = "Coefficient (effect on proportion correct)", y = NULL) +
  theme_minimal(base_size = 13)
```

---

## What to notice

**Between- vs. within-instrument variance.** The jump in R² when adding fixed effects
is the key comparison. If most variance is between instruments, overall difficulty is
mostly a property of the test (content domain, target population) rather than item
wording per se.

**Negation.** If items with negation show lower proportion correct, that's consistent
with processing load rather than content difficulty — respondents may answer the item
they wish had been asked.

**Limitations.** Proportion correct conflates item difficulty with sample ability.
A harder sample makes every item look more difficult. The dataset fixed effects
partially address this, but within-dataset ability variation is not controlled.
The natural next step is to use *b* parameters from the 2PL vignette instead.

---

## Reproducibility

Results computed on `r format(.cache$date_run, "%B %d, %Y")` using
`r length(eligible_tables)` IRW tables. To reproduce:

```r
source("vignettes/item_text_difficulty_compute.R")
quarto::quarto_render("vignettes/item_text_difficulty.qmd")
```

::: {.callout-tip}
To cite datasets used:

```r
for (tbl in eligible_tables) {
  irw_save_bibtex(tbl, output_file = "itemtextdata/irw_references.bib",
                  append = TRUE)
}
```
:::
