---
title: "Latent Space Item Response Models: Seeing Local Dependence as Geometry"
description: |
  The Rasch model assumes local independence, but nothing about a standard
  fit warns you when that assumption fails. The Latent Space Item Response
  Model (LSIRM) adds a single respondent-item distance term and turns the
  same question into a picture: an "interaction map" where items and
  respondents that share unexplained structure are pulled close together.
  We fit it to a small scouting set of IRW tables chosen to bracket the
  diagnostic -- known testlet structure, suspected speededness, and a
  presumed-Rasch-like negative control -- to see whether the geometry
  recovers what other diagnostics already tell us.
date: today
bibliography: lsirmdata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue   <- "#2166ac"
irw_red    <- "#d73027"
irw_grey   <- "grey50"
irw_purple <- "#762a83"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path_full  <- "lsirmdata/lsirm_interaction_maps_results_full.rds"
cache_path_scout <- "lsirmdata/lsirm_interaction_maps_results_scout.rds"
cache_path <- if (file.exists(cache_path_full)) cache_path_full else cache_path_scout
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  positions        <- .cache$positions
  item_order       <- .cache$item_order
  item_family_list <- .cache$item_family %||% list()
  scouting_tables  <- .cache$scouting_tables
  date_run         <- .cache$date_run
  is_pilot         <- isTRUE(.cache$scout)
} else {
  summary_df <- tibble(
    table = character(), is_dichotomous = logical(), n_categories = integer(),
    n_items = integer(), n_participants = integer(), gamma_mean = numeric(),
    gamma_lo = numeric(), gamma_hi = numeric(), pi_estimate = numeric(), bic = numeric()
  )
  positions <- list(); item_order <- list(); item_family_list <- list()
  scouting_tables <- character(); date_run <- Sys.Date(); is_pilot <- TRUE
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** -- figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/lsirm_interaction_maps_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, scouting run only.** This render uses the deliberately ",
  "fast MCMC settings (3,000 iterations, 1,000 burn-in) from the scouting ",
  "phase described below -- enough to check whether the pipeline works and ",
  "whether there's a signal worth writing up, not enough to trust as a final ",
  "estimate. Every number here is a real computed result, not a placeholder, ",
  "but treat point estimates (especially credible intervals) as provisional. ",
  "Re-run `lsirm_interaction_maps_compute.R --full` (15,000 iterations) to ",
  "replace this with the production run before treating this page as final.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette -- including the dataset search, the compute script, the
analysis design, and the writing below -- was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`lsirm_interaction_maps_compute.R`) directly if
you're relying on the numbers here.
:::

## Motivation

Every unidimensional model used elsewhere on this site -- the [2PL fits
across cognitive datasets](2pl_across_datasets.qmd), the IMV comparisons,
the [local dependence screen](local_dependence.qmd) itself -- assumes local
independence: once you condition on a respondent's position on the latent
trait, their responses to different items are unrelated. The [local
dependence vignette](local_dependence.qmd) checks this with Yen's Q3, a
*residual* diagnostic -- fit the model, subtract what it predicts, and see
whether anything is left to correlate between item pairs. That approach
answers "is independence violated" with a number per pair, but it doesn't
show *why* items and respondents interact, or which respondents are
involved.

The Latent Space Item Response Model [LSIRM; @jeon2021mapping] asks the
same question a different way: instead of testing residuals after the
fact, it adds a single extra term to the Rasch model up front -- a distance
penalty between each respondent and each item in a shared, unobserved 2D
space -- and lets the data decide whether that term is needed. When it
isn't, the model collapses back to an ordinary Rasch fit. When it is,
respondents and items that share unexplained structure are pulled close
together in the space, and the result is directly visualizable: an
**interaction map**. Local dependence stops being an abstract residual
correlation and becomes a picture -- clusters you can point at.

## Method in brief

For a respondent $j$ and item $i$ with an ordinary Rasch model, the
log-odds of a correct/endorsed response would be $\theta_j - \beta_i$
(respondent ability minus item difficulty). LSIRM adds one term:

$$\text{logit}\, P(Y_{ji} = 1) = \theta_j + \beta_i - \gamma \lVert z_j - w_i \rVert$$

where $z_j$ and $w_i$ are the respondent's and item's positions in a shared
2D Euclidean space, and $\gamma \geq 0$ scales how much that distance
matters. When $\gamma = 0$, the distance term vanishes and the model *is*
the Rasch model; the further $z_j$ sits from $w_i$, the more the distance
term drags the predicted log-odds down relative to what $\theta_j -
\beta_i$ alone would predict -- interaction the main effects didn't
capture. Fitting the positions and $\gamma$ jointly by MCMC produces the
interaction map: a scatter of item points and respondent points in the
same space, where proximity signals shared unexplained structure (a
testlet, a common stimulus, a shared response tendency) rather than an
ordinary ability/difficulty relationship.

Because $\gamma = 0$ is a single point on the boundary of $\gamma \geq 0$,
it can't be tested with an ordinary credible interval the way a normal
parameter could. The package used here (`lsirm12pl`) implements the
spike-and-slab model-selection extension from the original paper
[@ishwaran2005spike]: a mixture prior on $\log \gamma$ with a "spike"
component concentrated near $\gamma = 0$ and a "slab" component allowing
$\gamma > 0$, plus an indicator $\phi$ for which component is active at
each MCMC iteration. The posterior mean of $\phi$ -- reported below as the
**inclusion probability** -- is the model-selection answer: values near 0
mean the data are consistent with an ordinary Rasch fit; values near 1 mean
the interaction term is doing real work. This resolves an open question
going into this vignette (see Limitations) -- the current CRAN version of
`lsirm12pl` (2.0.3, not the 1.3.9 assumed when this vignette was scoped)
exposes spike-and-slab fitting directly via `lsirm1pl_ss()` (dichotomous
items) and `lsirmgrm_ss()` (graded/polytomous items), so no hand-rolled
WAIC comparison was needed.

## Data and methods

### Scouting-set selection

Rather than batch-fitting LSIRM across all of IRW -- MCMC here is far more
expensive per table than the mirt-based models used elsewhere on this site,
and the original paper's own examples ran 30-60 minutes per table at
N~200-700, I~7-24 -- this pass deliberately selects `r length(scouting_tables)`
tables chosen to bracket the diagnostic, each with a specific reason for
being included:

```{r}
#| label: scouting-set
#| eval: false

# 1. Tables with known item_family (testlet) structure -- ground truth on
#    which items *should* cluster if LSIRM is picking up real dependence.
#    Cross-referencing the local-dependence vignette's own item_family
#    census: only 2 of ~800 IRW tables carry a usable grouping at all.
#    g308_sirt is dichotomous; chakraborty2026_IWAH_IRW is graded (included
#    as a second confirmatory case via the GRM variant, even though it
#    doesn't meet the original dichotomous-only criterion for this bucket).

# 2. Tables suspected of speededness -- candidates for the "respondents
#    cluster near the last items" pattern described in the LSIRM paper's
#    discussion section. Cross-referenced against the (unmerged)
#    speededness-vignette branch's own classification:
#    credentialform_lnirt and nature_relatedness were both flagged
#    "likely speeded" there.

# 3. A presumed-Rasch-like negative control with no known family/testlet
#    structure and no speededness suspicion: blum_2018_imak_bin.

scouting_tables <- c("g308_sirt", "chakraborty2026_IWAH_IRW",
                     "credentialform_lnirt", "nature_relatedness",
                     "blum_2018_imak_bin")
```

```{r}
#| label: tbl-scouting-set
#| tbl-cap: "The scouting set: five tables chosen to bracket the diagnostic, not a representative sample of IRW."

summary_df |>
  mutate(role = case_when(
    table == "g308_sirt"                ~ "item_family (dichotomous)",
    table == "chakraborty2026_IWAH_IRW" ~ "item_family (graded)",
    table == "credentialform_lnirt"     ~ "speededness suspect (dichotomous)",
    table == "nature_relatedness"       ~ "speededness suspect (graded)",
    table == "blum_2018_imak_bin"       ~ "negative control"
  )) |>
  select(table, role, n_items, n_participants, n_categories) |>
  knitr::kable()
```

### Per-table computation

For each table, `fit_lsirm()` in the compute script:

1. Fetches the wide response matrix directly via `irw_fetch(table, resp =
   TRUE)`, and separately fetches the long format to recover `item_family`
   (if present) and a within-person item-presentation-order column (same
   fallback logic as the speededness vignette: a named order column if one
   exists, first-appearance order otherwise).
2. Downsamples respondents to at most 700 and items to at most 40,
   mirroring the downsampling pattern already used in [2PL across
   datasets](2pl_across_datasets.qmd). Item downsampling always keeps the
   true last 10 items by presentation order (so the "stragglers near the
   end" check below stays meaningful) and systematically samples the rest;
   this only triggered for `credentialform_lnirt` (200 -> 40 items).
3. Fits the 1PL LSIRM with spike-and-slab (`lsirm1pl_ss()`) for dichotomous
   tables, or the graded-response variant (`lsirmgrm_ss()`) for polytomous
   tables, both with a 2D latent space, matching the original paper's
   defaults.
4. Records posterior mean item and respondent positions, the $\gamma$
   posterior mean and 95% credible interval, the inclusion probability, and
   BIC.

```{r}
#| label: fit-function
#| eval: false

fit_lsirm <- function(table_name) {
  mat <- fetch_wide(table_name)  # irw_fetch(resp = TRUE), cleaned, downsampled
  fit <- if (is_dichotomous) lsirm1pl_ss(mat, ndim = 2) else lsirmgrm_ss(mat, ndim = 2)
  # gamma_estimate, pi_estimate (inclusion probability), z_estimate, w_estimate ...
}
```

::: {.callout-note}
To reproduce with current IRW holdings, re-run
`lsirm_interaction_maps_compute.R --full` and commit the updated
`lsirmdata/lsirm_interaction_maps_results_full.rds`.
:::

## Results

### Does the data need the interaction term at all?

```{r}
#| label: tbl-gamma
#| tbl-cap: "Posterior mean gamma (with 95% credible interval) and posterior inclusion probability for gamma > 0, one row per table. Inclusion probability above 0.5 favors the latent-space model over an ordinary Rasch fit; below 0.5 favors Rasch."

summary_df |>
  mutate(
    gamma_ci = sprintf("%.2f [%.2f, %.2f]", gamma_mean, gamma_lo, gamma_hi),
    favors   = ifelse(pi_estimate > 0.5, "Latent space (gamma > 0)", "Rasch (gamma = 0)")
  ) |>
  select(table, gamma_ci, pi_estimate, favors) |>
  knitr::kable(col.names = c("Table", "Gamma [95% CI]", "P(gamma > 0)", "Model favored"),
               digits = 3)
```

Three of the five tables show a clear, decisive signal for the interaction
term (inclusion probability > 0.98): `chakraborty2026_IWAH_IRW` (known
item_family structure), `credentialform_lnirt` and `nature_relatedness`
(both speededness suspects). Two show the opposite: `blum_2018_imak_bin`
(the negative control, as intended) and, more surprisingly,
`g308_sirt` -- the *dichotomous* item_family table. That contrast is
explored directly in the case study below rather than treated as a loose
end.

### Interaction maps

```{r}
#| label: fig-interaction-maps
#| fig-cap: "Interaction maps for all five scouting-set tables. Points are posterior mean positions in the shared 2D latent space: circles are respondents, triangles are items. Items sharing a known item_family grouping (where available) are colored; all other points are grey. Panels are titled with each table's posterior inclusion probability for gamma > 0."
#| fig-height: 8
#| fig-width: 10

build_position_df <- function(tbl) {
  pos <- positions[[tbl]]
  z <- as.data.frame(pos$z); names(z)[1:2] <- c("dim1", "dim2")
  z$type <- "respondent"; z$label <- rownames(pos$z); z$family <- NA_character_
  w <- as.data.frame(pos$w); names(w)[1:2] <- c("dim1", "dim2")
  w$type <- "item"; w$label <- rownames(pos$w)
  fam <- item_family_list[[tbl]]
  w$family <- if (!is.null(fam)) as.character(fam$item_family[match(w$label, fam$item)]) else NA_character_
  df <- bind_rows(z[, c("dim1","dim2","type","label","family")],
                  w[, c("dim1","dim2","type","label","family")])
  df$table <- tbl
  df
}

position_df <- bind_rows(lapply(summary_df$table, build_position_df))

panel_labels <- setNames(
  sprintf("%s (P[gamma>0] = %.2f)", summary_df$table, summary_df$pi_estimate),
  summary_df$table
)
position_df$panel <- panel_labels[position_df$table]

ggplot(position_df, aes(x = dim1, y = dim2, shape = type)) +
  geom_point(data = filter(position_df, is.na(family)), colour = irw_grey, alpha = 0.5, size = 1.6) +
  geom_point(data = filter(position_df, !is.na(family)), aes(colour = family), size = 2.4) +
  scale_shape_manual(values = c(respondent = 16, item = 17)) +
  facet_wrap(~panel, scales = "free", ncol = 2) +
  labs(x = NULL, y = NULL, shape = NULL, colour = "item_family") +
  theme(legend.position = "bottom", strip.text = element_text(size = 9))
```

The two high-inclusion-probability tables with distinguishable item
subgroups (`chakraborty2026_IWAH_IRW` by `item_family`) show visibly
tighter item clustering by group than the negative control's diffuse
scatter; `credentialform_lnirt` and `nature_relatedness` show broad,
non-null spread consistent with a real but less group-structured
interaction. `g308_sirt` and `blum_2018_imak_bin` both look close to
patternless noise -- consistent with their near-zero inclusion
probabilities, where the fitted positions carry little information (the
model has effectively decided it doesn't need them).

## Case study: LSIRM and Q3 on the same two tables

`g308_sirt` and `chakraborty2026_IWAH_IRW` aren't just this vignette's
item_family picks -- they're also the *only two* IRW tables the [local
dependence vignette](local_dependence.qmd) found with usable `item_family`
metadata, out of roughly 800 tables checked there. That gives a direct,
same-table comparison between a residual-based diagnostic (Q3) and a
geometric one (LSIRM):

```{r}
#| label: tbl-q3-vs-lsirm
#| tbl-cap: "Q3 (residual correlation) and LSIRM (interaction-map) diagnostics compared directly on the two IRW tables with known item_family structure. Q3 numbers are from local_dependence_results.rds (local_dependence.qmd)."

tibble(
  table              = c("g308_sirt", "chakraborty2026_IWAH_IRW"),
  q3_prop_flagged    = c(0.021, 0.260),
  q3_mean_abs        = c(0.065, 0.147),
  lsirm_gamma        = summary_df$gamma_mean[match(c("g308_sirt", "chakraborty2026_IWAH_IRW"), summary_df$table)],
  lsirm_p_gamma_gt_0 = summary_df$pi_estimate[match(c("g308_sirt", "chakraborty2026_IWAH_IRW"), summary_df$table)]
) |>
  knitr::kable(digits = 3, col.names = c("Table", "Q3: prop. flagged", "Q3: mean |Q3|",
                                          "LSIRM: gamma", "LSIRM: P(gamma>0)"))
```

The two diagnostics agree closely. `chakraborty2026_IWAH_IRW` -- a scale
where `item_family` groups the *same base item repeated across measurement
rounds* -- shows both a high proportion of Q3-flagged pairs (26%) and a
near-certain LSIRM inclusion probability (0.99); `g308_sirt` shows both a
low flagged proportion (2%) and a near-zero inclusion probability (0.01).
That agreement is reassuring on its own (two very different diagnostics
landing on the same conclusion from the same data), but `g308_sirt` is
also the more interesting result: its source paper -- "A Comparison of
Different Psychometric Approaches to Modeling Testlet Structures: An
Example with C-Tests" (Schroeders, Robitzsch & Schipolowski, 2014; see
`lsirmdata/irw_references.bib`) -- is explicitly *about* modeling testlet
structure in that instrument. Both diagnostics
here say the item_family grouping IRW carries for this particular table
doesn't translate into strong detectable local dependence in this data --
a useful caution that `item_family` metadata describes a testlet
*assignment*, not a guarantee that the assignment shows up as measurable
dependence in every dataset carrying it.

## Does the "stragglers near the end" pattern show up?

The original LSIRM paper's discussion section describes a specific
signature worth checking directly: in a speeded test, respondents who
don't finish should cluster near the *last* items in the interaction map,
because the shared "ran out of time" interaction is exactly the kind of
structure the distance term is built to absorb. Both `credentialform_lnirt`
and `nature_relatedness` were flagged `"likely speeded"` by the
(unmerged) speededness-vignette branch's independent omission-pattern
screen, making them direct candidates.

```{r}
#| label: tbl-position-clustering
#| tbl-cap: "Mean pairwise distance among the last 20% and first 20% of items (by presentation position) in the fitted item space, versus the mean pairwise distance across all items. A tighter cluster among last-position items than first-position items (or than the overall average) would support the stragglers-near-the-end pattern."

item_cluster_check <- function(tbl) {
  w   <- positions[[tbl]]$w
  ord <- item_order[[tbl]]
  ord <- ord[match(rownames(w), ord$item), ]
  n   <- nrow(ord)
  k   <- max(3, round(0.2 * n))
  last_items  <- ord$item[order(-ord$position)][1:k]
  first_items <- ord$item[order(ord$position)][1:k]
  d  <- as.matrix(dist(w))
  li <- rownames(w) %in% last_items
  fi <- rownames(w) %in% first_items
  tibble(
    table           = tbl,
    mean_dist_last  = mean(d[li, li][lower.tri(d[li, li])]),
    mean_dist_first = mean(d[fi, fi][lower.tri(d[fi, fi])]),
    mean_dist_all   = mean(d[lower.tri(d)])
  )
}

bind_rows(lapply(c("credentialform_lnirt", "nature_relatedness", "blum_2018_imak_bin"), item_cluster_check)) |>
  knitr::kable(digits = 2, col.names = c("Table", "Mean dist. (last 20%)", "Mean dist. (first 20%)", "Mean dist. (all pairs)"))
```

This check comes back **inconclusive at the scouting stage, not a clean
confirmation**. `credentialform_lnirt` shows *both* first- and
last-position items clustering tighter than the overall average -- not
specifically a last-item effect, and possibly confounded by the item
downsampling that table required (200 -> 40 items, which keeps the true
last 10 positions but samples the rest systematically rather than
contiguously). `nature_relatedness` shows the *opposite* of the predicted
direction: first-position items cluster tighter than last-position items.
Neither result rules out the pattern existing in the full, undownsampled
data with a properly converged chain -- this is exactly the kind of check
that deserves the full 15,000-iteration run rather than the 3,000-iteration
scouting pass before drawing a conclusion either way. Flagged here as
open, not resolved.

## What to notice

- **Clean separation is achievable and shows up where expected.**
  `chakraborty2026_IWAH_IRW`'s known repeated-item structure produces both
  a decisive inclusion probability (0.99) and visibly tighter clustering by
  `item_family` in the interaction map than the negative control shows.
- **A clean null is just as informative.** `blum_2018_imak_bin` (the
  intended negative control) behaves exactly as a Rasch-consistent table
  should: inclusion probability 0.007, a diffuse, patternless map.
- **The most interesting result wasn't the one we scoped for.**
  `g308_sirt` carries known testlet-structure metadata and was expected to
  show clustering; instead it came back Rasch-like on both LSIRM and Q3.
  That's convergent evidence across two independent methods, not a fitting
  artifact -- but it's a reminder that `item_family` in IRW describes an
  assignment made by the original data creators, not a guarantee that the
  assignment shows up as measurable dependence in every downstream
  analysis.
- **The speededness "stragglers" signature needs the full run to
  evaluate.** The quick position-clustering check above didn't confirm or
  rule out the pattern; it's confounded by item downsampling in one table
  and points the wrong direction in the other. Worth re-examining with the
  full item set and converged chains before writing a real conclusion here.
- **MCMC cost is the real constraint on scope.** Even with respondent and
  item downsampling, this analysis is far more expensive per table than
  the mirt-based models used elsewhere on the site, which is why this
  vignette covers five hand-picked tables rather than a systematic batch.

## Limitations

**This is a scouting pass, not a systematic survey.** Five tables were
hand-picked to bracket the diagnostic (known structure, suspected
speededness, presumed-null), not sampled to represent IRW as a whole.
Nothing here should be read as an estimate of how often LSIRM-detectable
interaction occurs across IRW generally.

**MCMC settings in this draft are intentionally fast, not final.** The
numbers above use 3,000 iterations with 1,000 burn-in (see the pilot
warning at the top of this page) -- enough to validate the pipeline and
get a first read on signal, not enough to trust point estimates or
credible intervals as final. A production run at 15,000 iterations is the
next step before this page is complete.

**Item downsampling is a compromise, and it may bias one result.**
`credentialform_lnirt`'s 200 -> 40 item reduction keeps the true last 10
items by position but samples the rest systematically; the inconclusive
stragglers-near-the-end check above may be partly an artifact of that
reduction rather than a fact about the full item set.

**The GRM extension is a package feature, not something the original paper
evaluated.** `chakraborty2026_IWAH_IRW` and `nature_relatedness` use
`lsirmgrm_ss()`, the graded-response variant `lsirm12pl` provides, rather
than the binary-response model the original paper's examples focus on.
The core distance-penalty logic is the same, but the graded extension's
own performance hasn't been validated here beyond "it ran and produced a
plausible-looking map."

**Only the 1PL (Rasch-equivalent) latent-space variant was fit.** The
package also exposes a 2PL LSIRM (`lsirm2pl_ss()`, allowing item
discrimination to vary). A flat interaction map under the 1PL variant --
as seen for the negative control here -- could in principle still mask
discrimination-driven structure that a 2PL fit would catch; this is
flagged as a natural extension rather than attempted in this pass.

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")` using
`r nrow(summary_df)` of `r length(scouting_tables)` scouting-set tables
(scouting-run MCMC settings; see the warning at the top of this page). To
reproduce:

```r
# 1. Re-run the compute script (from the project root)
#    -- fast scouting settings:
source("vignettes/lsirm_interaction_maps_compute.R")
#    -- or full production settings:
system("Rscript vignettes/lsirm_interaction_maps_compute.R --full")

# 2. Re-render this page
quarto::quarto_render("vignettes/lsirm_interaction_maps.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(scouting_tables, output_file = "lsirmdata/irw_references.bib")
```
:::

---

## References
