---
title: "Latent Space Item Response Models: Seeing Local Dependence as Geometry"
description: |
  The Rasch model assumes local independence, but nothing about a standard
  fit warns you when that assumption fails. The Latent Space Item Response
  Model (LSIRM) adds a single respondent-item distance term and turns the
  same question into a picture: an "interaction map" where items and
  respondents that share unexplained structure are pulled close together.
  We fit it to 16 IRW tables -- known testlet structure, suspected
  speededness, and blind exploratory draws -- and find a split by response
  format so clean it forced a validity check on the estimator itself.
date: today
bibliography: lsirmdata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue   <- "#2166ac"
irw_red    <- "#d73027"
irw_grey   <- "grey50"
irw_purple <- "#762a83"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path_full  <- "lsirmdata/lsirm_interaction_maps_results_full.rds"
cache_path_scout <- "lsirmdata/lsirm_interaction_maps_results_scout.rds"
cache_path <- if (file.exists(cache_path_full)) cache_path_full else cache_path_scout
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  positions        <- .cache$positions
  item_order       <- .cache$item_order
  item_family_list <- .cache$item_family %||% list()
  scouting_tables  <- .cache$scouting_tables
  date_run         <- .cache$date_run
  is_pilot         <- isTRUE(.cache$scout)
} else {
  summary_df <- tibble(
    table = character(), is_dichotomous = logical(), n_categories = integer(),
    n_items = integer(), n_participants = integer(), gamma_mean = numeric(),
    gamma_lo = numeric(), gamma_hi = numeric(), pi_estimate = numeric(), bic = numeric()
  )
  positions <- list(); item_order <- list(); item_family_list <- list()
  scouting_tables <- character(); date_run <- Sys.Date(); is_pilot <- TRUE
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** -- figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/lsirm_interaction_maps_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, scouting run only.** This render uses the deliberately ",
  "fast MCMC settings (3,000 iterations, 1,000 burn-in) from the scouting ",
  "phase described below -- enough to check whether the pipeline works and ",
  "whether there's a signal worth writing up, not enough to trust as a final ",
  "estimate. Every number here is a real computed result, not a placeholder, ",
  "but treat point estimates (especially credible intervals) as provisional. ",
  "Re-run `lsirm_interaction_maps_compute.R --full` (15,000 iterations) to ",
  "replace this with the production run before treating this page as final.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette -- including the dataset search, the compute script, the
analysis design, and the writing below -- was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`lsirm_interaction_maps_compute.R`) directly if
you're relying on the numbers here.
:::

## Motivation

Every unidimensional model used elsewhere on this site -- the [2PL fits
across cognitive datasets](2pl_across_datasets.qmd), the IMV comparisons,
the [local dependence screen](local_dependence.qmd) itself -- assumes local
independence: once you condition on a respondent's position on the latent
trait, their responses to different items are unrelated. The [local
dependence vignette](local_dependence.qmd) checks this with Yen's Q3, a
*residual* diagnostic -- fit the model, subtract what it predicts, and see
whether anything is left to correlate between item pairs. That approach
answers "is independence violated" with a number per pair, but it doesn't
show *why* items and respondents interact, or which respondents are
involved.

The Latent Space Item Response Model [LSIRM; @jeon2021mapping] asks the
same question a different way: instead of testing residuals after the
fact, it adds a single extra term to the Rasch model up front -- a distance
penalty between each respondent and each item in a shared, unobserved 2D
space -- and lets the data decide whether that term is needed. When it
isn't, the model collapses back to an ordinary Rasch fit. When it is,
respondents and items that share unexplained structure are pulled close
together in the space, and the result is directly visualizable: an
**interaction map**. Local dependence stops being an abstract residual
correlation and becomes a picture -- clusters you can point at.

## Method in brief

For a respondent $j$ and item $i$ with an ordinary Rasch model, the
log-odds of a correct/endorsed response would be $\theta_j - \beta_i$
(respondent ability minus item difficulty). LSIRM adds one term:

$$\text{logit}\, P(Y_{ji} = 1) = \theta_j + \beta_i - \gamma \lVert z_j - w_i \rVert$$

where $z_j$ and $w_i$ are the respondent's and item's positions in a shared
2D Euclidean space, and $\gamma \geq 0$ scales how much that distance
matters. When $\gamma = 0$, the distance term vanishes and the model *is*
the Rasch model; the further $z_j$ sits from $w_i$, the more the distance
term drags the predicted log-odds down relative to what $\theta_j -
\beta_i$ alone would predict -- interaction the main effects didn't
capture. Fitting the positions and $\gamma$ jointly by MCMC produces the
interaction map: a scatter of item points and respondent points in the
same space, where proximity signals shared unexplained structure (a
testlet, a common stimulus, a shared response tendency) rather than an
ordinary ability/difficulty relationship.

Because $\gamma = 0$ is a single point on the boundary of $\gamma \geq 0$,
it can't be tested with an ordinary credible interval the way a normal
parameter could. The package used here (`lsirm12pl`) implements the
spike-and-slab model-selection extension from the original paper
[@ishwaran2005spike]: a mixture prior on $\log \gamma$ with a "spike"
component concentrated near $\gamma = 0$ and a "slab" component allowing
$\gamma > 0$, plus an indicator $\phi$ for which component is active at
each MCMC iteration. The posterior mean of $\phi$ -- reported below as the
**inclusion probability** -- is the model-selection answer: values near 0
mean the data are consistent with an ordinary Rasch fit; values near 1 mean
the interaction term is doing real work. This resolves an open question
going into this vignette (see Limitations) -- the current CRAN version of
`lsirm12pl` (2.0.3, not the 1.3.9 assumed when this vignette was scoped)
exposes spike-and-slab fitting directly via `lsirm1pl_ss()` (dichotomous
items) and `lsirmgrm_ss()` (graded/polytomous items), so no hand-rolled
WAIC comparison was needed.

## Data and methods

### Study-set selection

MCMC here is far more expensive per table than the mirt-based models used
elsewhere on this site -- the original paper's own examples ran 30-60
minutes per table at N~200-700, I~7-24 -- so this vignette deliberately
selects `r length(scouting_tables)` tables rather than batch-fitting across
all of IRW, in two rounds:

**Round 1 (hypothesis-driven, 5 tables)**, chosen to bracket the diagnostic:

```{r}
#| label: scouting-set
#| eval: false

# 1. Tables with known item_family (testlet) structure -- ground truth on
#    which items *should* cluster if LSIRM is picking up real dependence.
#    Cross-referencing the local-dependence vignette's own item_family
#    census: only 2 of ~800 IRW tables carry a usable grouping at all.
#    g308_sirt is dichotomous; chakraborty2026_IWAH_IRW is graded.

# 2. Tables suspected of speededness -- candidates for the "respondents
#    cluster near the last items" pattern described in the LSIRM paper's
#    discussion section. Cross-referenced against the (unmerged)
#    speededness-vignette branch's own classification:
#    credentialform_lnirt and nature_relatedness were both flagged
#    "likely speeded" there.

# 3. A presumed-Rasch-like negative control with no known family/testlet
#    structure and no speededness suspicion: blum_2018_imak_bin.
```

**Round 2 (expansion, 11 more tables)**, added after Round 1 cleared the
go/no-go bar and raised a "what predicts a positive finding" question that
needed more data points:

```{r}
#| label: expansion-set
#| eval: false

# 4. The remaining 5 tables classified "likely speeded" by the same
#    speededness screen, completing that bucket in full: brain_hemisphere,
#    artistic_preferences, depression_anxiety_stress, fisher_temperment,
#    face_memory_test.

# 5. A *blind* random sample (seed 20260723) of 6 more tables from
#    irw_filter(n_items = c(5, 60), n_participants = c(100, 5000)),
#    excluding every table already in the study set -- deliberately not
#    hand-picked for any hypothesis: sd3ypl_klimczak_2019_ses,
#    BPAQ_Christopher_2024_PSS10, prpt_hellmann_2021_conscientiousness,
#    chen2026_sc, disgust_berger2014, autism_blotner_2025_s1_aq.
```

```{r}
#| label: tbl-scouting-set
#| tbl-cap: "The full 16-table study set. 'Domain' is a rough characterization from each table's construct/context (cognitive-ability task vs. personality/attitude/clinical self-report scale), not an IRW metadata field -- flagged because it turns out to matter (see below)."

summary_df |>
  mutate(
    role = case_when(
      table == "g308_sirt"                    ~ "item_family",
      table == "chakraborty2026_IWAH_IRW"     ~ "item_family",
      table %in% c("credentialform_lnirt", "nature_relatedness",
                   "brain_hemisphere", "artistic_preferences",
                   "depression_anxiety_stress", "fisher_temperment",
                   "face_memory_test")         ~ "speededness suspect",
      table == "blum_2018_imak_bin"           ~ "negative control",
      TRUE                                     ~ "blind exploratory"
    ),
    domain = case_when(
      table %in% c("g308_sirt", "credentialform_lnirt", "blum_2018_imak_bin") ~ "cognitive/ability",
      table == "face_memory_test" ~ "cognitive/ability (graded)",
      TRUE ~ "personality/attitude/clinical"
    )
  ) |>
  select(table, role, domain, n_items, n_participants, n_categories) |>
  arrange(role, domain) |>
  knitr::kable()
```

### Per-table computation

For each table, `fit_lsirm()` in the compute script:

1. Fetches the wide response matrix directly via `irw_fetch(table, resp =
   TRUE)`, and separately fetches the long format to recover `item_family`
   (if present) and a within-person item-presentation-order column (same
   fallback logic as the speededness vignette: a named order column if one
   exists, first-appearance order otherwise).
2. Downsamples respondents to at most 700 and items to at most 40,
   mirroring the downsampling pattern already used in [2PL across
   datasets](2pl_across_datasets.qmd). Item downsampling always keeps the
   true last 10 items by presentation order (so the "stragglers near the
   end" check below stays meaningful) and systematically samples the rest;
   this triggered for every table with more than 40 items
   (`credentialform_lnirt`, `depression_anxiety_stress`, `fisher_temperment`,
   `face_memory_test`, `artistic_preferences`).
3. Fits the 1PL LSIRM with spike-and-slab (`lsirm1pl_ss()`) for dichotomous
   tables, or the graded-response variant (`lsirmgrm_ss()`) for polytomous
   tables, both with a 2D latent space and the full 15,000-iteration MCMC
   settings (2,500 burn-in, thinned by 5), matching the original paper's
   defaults.
4. Records posterior mean item and respondent positions, the $\gamma$
   posterior mean and 95% credible interval, the inclusion probability, and
   BIC.

```{r}
#| label: fit-function
#| eval: false

fit_lsirm <- function(table_name) {
  mat <- fetch_wide(table_name)  # irw_fetch(resp = TRUE), cleaned, downsampled
  fit <- if (is_dichotomous) lsirm1pl_ss(mat, ndim = 2) else lsirmgrm_ss(mat, ndim = 2)
  # gamma_estimate, pi_estimate (inclusion probability), z_estimate, w_estimate ...
}
```

::: {.callout-note}
To reproduce with current IRW holdings, re-run
`lsirm_interaction_maps_compute.R --full` and commit the updated
`lsirmdata/lsirm_interaction_maps_results_full.rds`.
:::

## Results

### Does the data need the interaction term at all?

```{r}
#| label: tbl-gamma
#| tbl-cap: "Posterior mean gamma (with 95% credible interval) and posterior inclusion probability for gamma > 0, one row per table, sorted by response format then by P(gamma > 0). Inclusion probability above 0.5 favors the latent-space model over an ordinary Rasch fit; below 0.5 favors Rasch."

summary_df |>
  mutate(
    format   = ifelse(is_dichotomous, "dichotomous", "polytomous/graded"),
    gamma_ci = sprintf("%.2f [%.2f, %.2f]", gamma_mean, gamma_lo, gamma_hi),
    favors   = ifelse(pi_estimate > 0.5, "Latent space (gamma > 0)", "Rasch (gamma = 0)")
  ) |>
  arrange(is_dichotomous, desc(pi_estimate)) |>
  select(table, format, gamma_ci, pi_estimate, favors) |>
  knitr::kable(col.names = c("Table", "Format", "Gamma [95% CI]", "P(gamma > 0)", "Model favored"),
               digits = 3)
```

The pattern that jumps out is not subtle: **every one of the 13
polytomous/graded tables** -- including all 6 blind, hypothesis-free
exploratory tables -- shows a decisive inclusion probability above 0.96.
The 3 dichotomous tables split as originally scoped: two near-zero
(`g308_sirt`, `blum_2018_imak_bin`) and one decisive
(`credentialform_lnirt`, now the single highest inclusion probability in
the whole set at 0.9996). A split that tracks response format this cleanly,
appearing the moment the study set was widened with unrelated blind
tables, is exactly the kind of pattern that deserves a validity check
before being taken as a real substantive finding -- see the next section.

### Interaction maps

```{r}
#| label: fig-interaction-maps
#| fig-cap: "Interaction maps for all 16 study-set tables, grouped by response format (dichotomous first) and sorted by posterior inclusion probability within format. Points are posterior mean positions in the shared 2D latent space: circles are respondents, triangles are items. Items sharing a known item_family grouping (where available) are colored; all other points are grey. Panel titles show each table's posterior inclusion probability for gamma > 0."
#| fig-height: 16
#| fig-width: 12

build_position_df <- function(tbl) {
  pos <- positions[[tbl]]
  z <- as.data.frame(pos$z); names(z)[1:2] <- c("dim1", "dim2")
  z$type <- "respondent"; z$label <- rownames(pos$z); z$family <- NA_character_
  w <- as.data.frame(pos$w); names(w)[1:2] <- c("dim1", "dim2")
  w$type <- "item"; w$label <- rownames(pos$w)
  fam <- item_family_list[[tbl]]
  w$family <- if (!is.null(fam)) as.character(fam$item_family[match(w$label, fam$item)]) else NA_character_
  df <- bind_rows(z[, c("dim1","dim2","type","label","family")],
                  w[, c("dim1","dim2","type","label","family")])
  df$table <- tbl
  df
}

position_df <- bind_rows(lapply(summary_df$table, build_position_df))

panel_order <- summary_df |> arrange(is_dichotomous, desc(pi_estimate)) |> pull(table)
panel_labels <- setNames(
  sprintf("%s (P[gamma>0] = %.2f)", summary_df$table, summary_df$pi_estimate),
  summary_df$table
)
position_df$panel <- factor(panel_labels[position_df$table], levels = panel_labels[panel_order])

ggplot(position_df, aes(x = dim1, y = dim2, shape = type)) +
  geom_point(data = filter(position_df, is.na(family)), colour = irw_grey, alpha = 0.5, size = 1.4) +
  geom_point(data = filter(position_df, !is.na(family)), aes(colour = family), size = 2.2) +
  scale_shape_manual(values = c(respondent = 16, item = 17)) +
  facet_wrap(~panel, scales = "free", ncol = 4) +
  labs(x = NULL, y = NULL, shape = NULL, colour = "item_family") +
  theme(legend.position = "bottom", strip.text = element_text(size = 8))
```

`chakraborty2026_IWAH_IRW` (top group, colored by `item_family`) shows
visibly tighter item clustering by group than the two near-null dichotomous
panels' diffuse scatter. Beyond that one labeled case, though, the maps
mostly demonstrate the same thing the gamma table does: broad, structured
spread across nearly every polytomous panel regardless of role
(speededness suspect or blind exploratory), and near-patternless noise in
`g308_sirt` and `blum_2018_imak_bin`.

## Is this really about response format, or is the estimator biased?

A split this clean -- every polytomous table positive, most dichotomous
tables null -- is more consistent with an estimator quirk than with 13
independent substantive discoveries, so before interpreting it, the
question needs to be turned on the tool itself: does `lsirmgrm_ss()`
report high inclusion probability on *any* graded-response data, even data
built with zero true interaction?

```{r}
#| label: null-calibration
#| eval: false

# Simulate a clean single-factor GRM dataset: realistic item discrimination
# and category thresholds, N/I matched to the study set, gamma = 0 by
# construction (no interaction structure of any kind).
a <- matrix(rlnorm(20, 0, 0.3), ncol = 1)
d <- matrix(t(apply(matrix(rnorm(20 * 4), ncol = 4), 1, function(x) sort(x, decreasing = TRUE))), ncol = 4)
sim_dat <- mirt::simdata(a, d, N = 700, itemtype = "graded")
sim_fit <- lsirmgrm_ss(sim_dat, ndim = 2, niter = 15000, nburn = 2500, nthin = 5)
```

```{r}
#| label: tbl-null-calibration
#| tbl-cap: "Null-calibration check: lsirmgrm_ss() fit to simulated single-factor GRM data with no true interaction structure, versus the range of results actually observed across the 13 real polytomous study-set tables."

tibble(
  case             = c("Simulated null GRM data (this check)", "Real polytomous tables (range, n=13)"),
  pi_estimate      = c(.cache$null_calibration$pi_estimate,
                       paste0(sprintf("%.3f", min(summary_df$pi_estimate[!summary_df$is_dichotomous])),
                              " to ",
                              sprintf("%.3f", max(summary_df$pi_estimate[!summary_df$is_dichotomous])))),
  gamma            = c(sprintf("%.3f", .cache$null_calibration$gamma_mean),
                       paste0(sprintf("%.2f", min(summary_df$gamma_mean[!summary_df$is_dichotomous])),
                              " to ",
                              sprintf("%.2f", max(summary_df$gamma_mean[!summary_df$is_dichotomous]))))
) |>
  knitr::kable(col.names = c("Case", "P(gamma > 0)", "Gamma"))
```

The estimator passes this check: on data simulated with no true interaction
whatsoever, `lsirmgrm_ss()` correctly recovers a near-zero inclusion
probability (`r sprintf("%.3f", .cache$null_calibration$pi_estimate)`) and
a near-zero gamma. That rules out the crudest possible explanation (the
graded spike-and-slab implementation always says yes), and shifts the
weight of evidence toward this being a real property of the tables
themselves rather than a software bug -- but it does not fully resolve the
question, and the study-set design confounds two things that would need to
be teased apart to go further:

**Response format and instrument domain are nearly perfectly aliased in
this study set.** Every dichotomous table here is a cognitive-ability task
(testlet reading comprehension, a credentialing exam, figural analogies);
every polytomous table (with one partial exception, `face_memory_test`, a
graded confidence/memory task rather than a personality scale) is a
Likert-type personality, attitude, or clinical self-report scale
(conscientiousness, disgust sensitivity, autism screening, depression/
anxiety/stress, and so on). Likert personality batteries are well known to
carry genuine item redundancy -- reverse-worded twins, near-duplicate
phrasing, shared method variance, halo and acquiescence response styles --
in a way ability tests are usually designed to avoid. So "format predicts a
positive finding" in this particular batch may really be "domain predicts a
positive finding," with format riding along as a near-perfect proxy rather
than a causal driver in its own right. Disentangling the two would need
either a polytomous cognitive-ability battery or a dichotomous personality
scale (rare in IRW, per the format breakdown in the [local dependence
vignette](local_dependence.qmd)) in the study set -- neither was available
in this pass.

**One simulation is a spot check, not an exhaustive calibration.** The null
data above uses a fairly benign single-factor GRM with independently drawn,
moderate item discriminations. Real personality-scale data can have
features the simulation doesn't -- skewed trait distributions, floor/
ceiling effects, more categories with sparse endorsement -- that could in
principle inflate gamma through a channel other than genuine content
overlap. This check makes the "always positive on polytomous data" story
less likely, not impossible.

## Case study: LSIRM and Q3 on the same two tables

`g308_sirt` and `chakraborty2026_IWAH_IRW` aren't just this vignette's
item_family picks -- they're also the *only two* IRW tables the [local
dependence vignette](local_dependence.qmd) found with usable `item_family`
metadata, out of roughly 800 tables checked there. That gives a direct,
same-table comparison between a residual-based diagnostic (Q3) and a
geometric one (LSIRM):

```{r}
#| label: tbl-q3-vs-lsirm
#| tbl-cap: "Q3 (residual correlation) and LSIRM (interaction-map) diagnostics compared directly on the two IRW tables with known item_family structure. Q3 numbers are from local_dependence_results.rds (local_dependence.qmd)."

tibble(
  table              = c("g308_sirt", "chakraborty2026_IWAH_IRW"),
  q3_prop_flagged    = c(0.021, 0.260),
  q3_mean_abs        = c(0.065, 0.147),
  lsirm_gamma        = summary_df$gamma_mean[match(c("g308_sirt", "chakraborty2026_IWAH_IRW"), summary_df$table)],
  lsirm_p_gamma_gt_0 = summary_df$pi_estimate[match(c("g308_sirt", "chakraborty2026_IWAH_IRW"), summary_df$table)]
) |>
  knitr::kable(digits = 3, col.names = c("Table", "Q3: prop. flagged", "Q3: mean |Q3|",
                                          "LSIRM: gamma", "LSIRM: P(gamma>0)"))
```

The two diagnostics agree closely. `chakraborty2026_IWAH_IRW` -- a scale
where `item_family` groups the *same base item repeated across measurement
rounds* -- shows both a high proportion of Q3-flagged pairs (26%) and a
near-certain LSIRM inclusion probability (0.995); `g308_sirt` shows both a
low flagged proportion (2%) and a near-zero inclusion probability (0.003).
That agreement is reassuring on its own (two very different diagnostics
landing on the same conclusion from the same data), but `g308_sirt` is
also the more interesting result: its source paper -- "A Comparison of
Different Psychometric Approaches to Modeling Testlet Structures: An
Example with C-Tests" (Schroeders, Robitzsch & Schipolowski, 2014; see
`lsirmdata/irw_references.bib`) -- is explicitly *about* modeling testlet
structure in that instrument. Both diagnostics
here say the item_family grouping IRW carries for this particular table
doesn't translate into strong detectable local dependence in this data --
a useful caution that `item_family` metadata describes a testlet
*assignment*, not a guarantee that the assignment shows up as measurable
dependence in every dataset carrying it.

## Does the "stragglers near the end" pattern show up?

The original LSIRM paper's discussion section describes a specific
signature worth checking directly: in a speeded test, respondents who
don't finish should cluster near the *last* items in the interaction map,
because the shared "ran out of time" interaction is exactly the kind of
structure the distance term is built to absorb. All 7 tables classified
`"likely speeded"` by the (unmerged) speededness-vignette branch's
independent omission-pattern screen are now in the study set, giving a
much better-powered look at this than the original 2-table check.

```{r}
#| label: tbl-position-clustering
#| tbl-cap: "Mean pairwise distance among the last 20% and first 20% of items (by presentation position) in the fitted item space, versus the mean pairwise distance across all items, for all 7 speededness-suspect tables plus the two null-gamma tables as a reference. A tighter cluster among last-position items than first-position items (or than the overall average) would support the stragglers-near-the-end pattern; 'last < first' marks cases matching that prediction."

item_cluster_check <- function(tbl) {
  w   <- positions[[tbl]]$w
  ord <- item_order[[tbl]]
  ord <- ord[match(rownames(w), ord$item), ]
  n   <- nrow(ord)
  k   <- max(3, round(0.2 * n))
  last_items  <- ord$item[order(-ord$position)][1:k]
  first_items <- ord$item[order(ord$position)][1:k]
  d  <- as.matrix(dist(w))
  li <- rownames(w) %in% last_items
  fi <- rownames(w) %in% first_items
  tibble(
    table           = tbl,
    mean_dist_last  = mean(d[li, li][lower.tri(d[li, li])]),
    mean_dist_first = mean(d[fi, fi][lower.tri(d[fi, fi])]),
    mean_dist_all   = mean(d[lower.tri(d)])
  )
}

speeded_tables <- c("credentialform_lnirt", "nature_relatedness", "brain_hemisphere",
                    "artistic_preferences", "depression_anxiety_stress",
                    "fisher_temperment", "face_memory_test")

bind_rows(lapply(c(speeded_tables, "blum_2018_imak_bin", "g308_sirt"), item_cluster_check)) |>
  mutate(matches_prediction = ifelse(mean_dist_last < mean_dist_first, "yes (last < first)", "no")) |>
  knitr::kable(digits = 2, col.names = c("Table", "Mean dist. (last 20%)", "Mean dist. (first 20%)",
                                          "Mean dist. (all pairs)", "Matches prediction?"))
```

With the full speeded bucket in hand, this comes back **genuinely mixed,
not a confirmation or a rejection**. Of the 7 speeded-suspect tables, 3
show last-position items clustering tighter than first-position items as
predicted (`artistic_preferences`, `fisher_temperment` weakly,
`face_memory_test` strongly -- its last-item cluster is less than a
quarter the spread of its first-item cluster); the other 4 show the
opposite (`credentialform_lnirt`, `nature_relatedness`,
`brain_hemisphere`, `depression_anxiety_stress`). There's no obvious
feature separating the two groups (both directions include dichotomous
and polytomous tables, small and large item counts). `face_memory_test` is
the one clean example matching the paper's discussion-section pattern
qualitatively; the rest suggest that either the effect is real but
table-specific (present only for some speeded-test designs or item
types), or that this pairwise item-distance proxy is too indirect a test
without linking it to actual respondent-level omission behavior (available
in the speededness-vignette's own cached data, but not joined in here).
Reported as an open, mixed finding rather than forced into either
direction.

## What to notice

- **Response format (or the domain it's aliased with) is the dominant
  driver in this batch, not the specific mechanism we scoped for.** Every
  one of 13 polytomous tables -- known-structure, speeded-suspect, and 6
  hypothesis-free blind draws alike -- shows a decisive inclusion
  probability (>0.96). Only the 3 dichotomous tables show real variation.
  This dwarfs the original "does known structure produce a positive
  finding" question the study was scoped around.
- **The estimator itself checks out on synthetic data**, but the study
  design can't yet separate *format* (dichotomous vs. graded) from
  *domain* (cognitive-ability task vs. personality/attitude self-report) --
  they're nearly perfectly aliased here. That's the single most important
  open question this pass raises, not a settled conclusion.
- **Clean separation still shows up where there's independent ground
  truth.** `chakraborty2026_IWAH_IRW`'s known repeated-item structure
  produces visibly tighter clustering by `item_family` than the null
  dichotomous tables' diffuse scatter, and agrees closely with the
  existing Q3 diagnostic on the same table.
- **A clean null is just as informative.** `blum_2018_imak_bin` and
  `g308_sirt` (the latter despite carrying known testlet-structure
  metadata) both come back Rasch-like -- and agree with Q3 on `g308_sirt`
  specifically, which is convergent evidence across two independent
  methods, not a fitting artifact.
- **The speededness "stragglers" signature is a genuine mixed bag**, not a
  confirmation or a rejection, across the full 7-table speeded bucket: 3
  tables show the predicted last-item clustering, 4 don't. `face_memory_test`
  is the one strong, clean match to the paper's discussion-section pattern.
- **MCMC cost is still the real constraint on scope.** Even at 16 tables
  with respondent/item downsampling, this is a hand-assembled study set,
  not a systematic survey -- see Limitations.

## Limitations

**This is a hand-assembled study set, not a systematic survey -- even
after expansion.** 16 tables were chosen across two rounds (5
hypothesis-driven, 11 more completing a bucket or drawn blind), not
sampled to represent IRW as a whole. The 6 blind exploratory tables reduce
selection bias somewhat but were drawn from a size-filtered pool
(`n_items` 5-60, `n_participants` 100-5,000), not all of IRW. Nothing here
should be read as a calibrated estimate of how often LSIRM-detectable
interaction occurs across IRW generally.

**Response format and instrument domain are confounded in this study set,
and that's the central open question, not a footnote.** Every dichotomous
table is a cognitive-ability task; nearly every polytomous table is a
personality/attitude/clinical Likert scale. The null-calibration check
above rules out the crudest explanation (a biased estimator that always
says yes on graded data) but can't by itself tell us whether format or
domain is doing the causal work. Resolving this needs a polytomous
cognitive-ability battery and/or a dichotomous personality scale in the
study set -- both are rare to nonexistent in current IRW holdings, per the
format breakdown already documented in the [local dependence
vignette](local_dependence.qmd).

**Item downsampling is a compromise.** Every table with more than 40 items
(`credentialform_lnirt`, `depression_anxiety_stress`, `fisher_temperment`,
`face_memory_test`, `artistic_preferences`) had items reduced, keeping the
true last 10 by position and sampling the rest systematically. This could
plausibly distort the stragglers-near-the-end check above for those
specific tables, though the mixed result held across both downsampled and
non-downsampled speeded tables, which argues against downsampling being
the sole driver of the mixed picture.

**The GRM extension is a package feature, not something the original paper
evaluated.** 13 of 16 tables use `lsirmgrm_ss()`, the graded-response
variant `lsirm12pl` provides, rather than the binary-response model the
original paper's examples focus on. The null-calibration check validates
that this specific extension isn't trivially miscalibrated, but its
broader performance characteristics haven't been independently studied
here beyond that one simulated scenario.

**Only the 1PL (Rasch-equivalent) latent-space variant was fit.** The
package also exposes a 2PL LSIRM (`lsirm2pl_ss()`, allowing item
discrimination to vary). A flat interaction map under the 1PL variant --
as seen for the negative-control tables here -- could in principle still
mask discrimination-driven structure that a 2PL fit would catch; this is
flagged as a natural extension rather than attempted in this pass.

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")` using
`r nrow(summary_df)` of `r length(scouting_tables)` study-set tables at
full production MCMC settings (15,000 iterations, 2,500 burn-in). To
reproduce:

```r
# 1. Re-run the compute script (from the project root)
#    -- fast scouting settings (for quick iteration on new tables):
source("vignettes/lsirm_interaction_maps_compute.R")
#    -- full production settings (used for the numbers on this page):
system("Rscript vignettes/lsirm_interaction_maps_compute.R --full")

# 2. Re-render this page
quarto::quarto_render("vignettes/lsirm_interaction_maps.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(scouting_tables, output_file = "lsirmdata/irw_references.bib")
```
:::

---

## References
