---
title: "Text-Augmented IRT Calibration: A Mixed-Subjects Proof of Concept"
description: |
  Can surface text features (word count, average word length) generate synthetic calibration subjects for IRT?
  We use word-count predictors of item difficulty — from the item text vignette —
  to simulate synthetic respondents and apply the mixed-subjects IRT framework
  to a preschool social-emotional learning dataset.
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)
library(tidyr)

cache_path <- "mixed_subjects_data/results.rds"
if (!file.exists(cache_path)) {
  knitr::knit_exit("Cache not found — run `vignettes/mixed_subjects_irt_compute.R` first.")
}

.c             <- readRDS(cache_path)
gt_pars        <- .c$gt_pars
pilot_pars     <- .c$pilot_pars
lambda_final   <- .c$lambda_final
lambda_summary <- .c$lambda_summary
target_data    <- .c$target_data
r_text         <- .c$r_text
human_accuracy <- .c$human_accuracy
synth_accuracy <- .c$synth_accuracy
rmse_pilot     <- .c$rmse_pilot
rmse_mixed     <- .c$rmse_mixed
pilot_n        <- .c$pilot_n
n_synth        <- .c$n_synth
n_human_total  <- .c$n_human_total
sim_lambda     <- .c$sim_lambda_final
sim_summary    <- .c$sim_lambda_summary
b_sim_true     <- .c$sim_b_true
```

## The calibration cost problem

Fitting an IRT model requires pilot data — real respondents answering real items
before the test is deployed. Collecting that data is expensive, especially for
assessments targeting young children or clinical populations. A natural question:
can *synthetic* responses supplement a small human pilot and improve item
parameter recovery?

[Kanopka et al.](https://klintkanopka.com/mixedsubjectsirt/) address this with
a **mixed-subjects IRT** framework built on Prediction-Powered Inference (PPI).
The method combines human pilot responses with a much larger pool of synthetic
responses, using a tunable weight *λ* to control the synthetic contribution.
When synthetic responses are informative, *λ* is large; when they are
unreliable, *λ* shrinks toward zero — automatically guarding against poor
synthetic data degrading the estimates.

The IRW provides an ideal testbed. We have both item text (so we can predict
difficulty without actual test-takers) and large-scale human response data (so
we can measure actual parameter recovery). Here we demonstrate the approach on
a single social-emotional learning dataset, using **word count and average word length** as
the source of synthetic difficulty predictions rather than a live LLM API.

---

## Dataset: preschool_sel_box

`preschool_sel_box` contains responses from **`r n_human_total` children** to
**20 items** from a preschool social-emotional learning assessment. Items
involve brief scenarios about boxes (e.g., choosing, sharing, and describing
box contents); responses are binary (correct / incorrect).

```{r}
#| label: tbl-items
#| tbl-cap: "Item text and human proportion correct"

target_data |>
  select(item, item_text, word_count, avg_word_len, prop_correct, pred_p) |>
  mutate(word_count   = as.integer(word_count),
         avg_word_len = round(avg_word_len, 1),
         prop_correct = round(prop_correct, 3),
         pred_p       = round(pred_p, 3)) |>
  knitr::kable(col.names = c("Item", "Text", "Words", "Avg word len", "Human P(correct)", "Predicted P(correct)"))
```

---

## Text-feature difficulty predictions

The [item text difficulty vignette](item_text_difficulty.html) showed that
simple surface features — word count and average word length — predict item
difficulty across IRW datasets. We train a linear regression on **all IRW
datasets except `preschool_sel_box`**, then apply it to predict pass rates for
the 20 target items.

```{r}
#| label: fig-text-pred
#| fig-cap: !expr sprintf("Text-feature predictions vs. observed pass rates (Pearson r = %.3f). Predictions capture rank ordering but are compressed relative to the observed range.", r_text)

ggplot(target_data, aes(x = pred_p, y = prop_correct, label = item)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey70") +
  geom_point(size = 3, colour = "#2563EB", alpha = 0.8) +
  annotate("text", x = 0.45, y = 0.75,
           label = sprintf("r = %.3f", r_text),
           size = 4.5, colour = "#2563EB", fontface = "bold") +
  scale_x_continuous(limits = c(0.3, 0.7), labels = scales::percent) +
  scale_y_continuous(limits = c(0.4, 0.85), labels = scales::percent) +
  labs(
    x     = "Predicted P(correct) — word count + avg word length",
    y     = "Observed P(correct) — full human sample",
    title = "Text-feature predictions vs. observed pass rates"
  ) +
  theme_minimal(base_size = 13)
```

The correlation is `r round(r_text, 3)` — strong enough to identify easier and
harder items, but the predictions cluster tightly (all between 35% and 60%)
while observed pass rates span a wider range (45%–81%). This compression will
matter when we try to use the predictions for calibration.

---

## Generating synthetic subjects

We translate text-predicted pass rates into IRT difficulty parameters
(*b* = −logit of predicted *P*) and simulate **`r n_synth` synthetic subjects**
whose responses are drawn from a 1PL model. Ability is θ ~ N(0, 1), matching a
typical standardisation population.

```{r}
#| label: fig-synth-vs-human
#| fig-cap: "Synthetic vs. human accuracy per item. Synthetic responses match the observed ordering but are compressed toward 50%."

acc_df <- data.frame(
  item   = names(human_accuracy),
  human  = as.numeric(human_accuracy),
  synth  = as.numeric(synth_accuracy[names(human_accuracy)])
)

ggplot(acc_df, aes(x = human, y = synth)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey70") +
  geom_point(size = 3, colour = "#059669", alpha = 0.8) +
  scale_x_continuous(limits = c(0.4, 0.85), labels = scales::percent) +
  scale_y_continuous(limits = c(0.4, 0.85), labels = scales::percent) +
  labs(
    x     = "Human P(correct) — full sample",
    y     = "Synthetic P(correct) — text-predicted",
    title = "Synthetic vs. human accuracy per item"
  ) +
  theme_minimal(base_size = 13)
```

---

## Does the method work? A simulation first

Before applying to real data, we verify that *λ* tuning behaves correctly under
ideal conditions. We simulate 200 human pilot responses and 2,000 synthetic
responses from a 1PL model with items evenly spaced across b ∈ [−2, 2]
(sd = 1.22) — a wide difficulty range. The synthetic θ distribution is
N(0.5, 1), slightly offset from the human N(0, 1).

```{r}
#| label: fig-sim-risk
#| fig-cap: !expr sprintf("Lambda (λ) tuning in the simulation. λ controls how much weight the mixed-subjects estimator places on synthetic responses (λ=0 uses only the human pilot; larger λ blends in more synthetic data). The ability-risk criterion dips below its λ=0 value around λ = %.3f, confirming that synthetic data helps when predictions are high-quality.", sim_lambda)

sim_df <- sim_summary |>
  select(lambda, risk = selection_risk) |>
  mutate(min_risk = min(risk))

ggplot(sim_df, aes(x = lambda, y = risk)) +
  geom_hline(yintercept = sim_df$risk[1], linetype = "dashed",
             colour = "grey60", linewidth = 0.6) +
  geom_line(colour = "#2563EB", linewidth = 1) +
  geom_point(colour = "#2563EB", size = 2.5) +
  geom_point(data = sim_df |> filter(lambda == sim_lambda),
             colour = "#EF4444", size = 4) +
  annotate("text",
           x     = sim_lambda + 0.05,
           y     = min(sim_df$risk) - 0.00008,
           label = sprintf("λ* = %.3f", sim_lambda),
           colour = "#EF4444", size = 4) +
  labs(
    x     = "λ (weight on synthetic responses; 0 = pilot only)",
    y     = "Ability estimation risk",
    title = "Simulation: risk curve shows clear minimum"
  ) +
  theme_minimal(base_size = 13)
```

The risk curve has a clear minimum at λ = `r round(sim_lambda, 3)`. In these
ideal conditions — wide item difficulty spread, accurate synthetic data — a
small positive weight on synthetic responses reduces ability estimation risk
by about 0.7% relative to the pilot alone.

---

## Lambda tuning on real data

Now we apply the same framework to `preschool_sel_box`. The pilot is a random
sample of **`r pilot_n` children** from the full `r n_human_total`-subject
dataset. The synthetic pool is the `r n_synth` simulated subjects.

```{r}
#| label: fig-real-risk
#| fig-cap: "Lambda tuning on real data. The ability-risk criterion increases monotonically from λ=0 — no synthetic weight improves calibration."

real_df <- lambda_summary |>
  select(lambda, risk = selection_risk)

ggplot(real_df, aes(x = lambda, y = risk)) +
  geom_hline(yintercept = real_df$risk[1], linetype = "dashed",
             colour = "grey60", linewidth = 0.6) +
  geom_line(colour = "#059669", linewidth = 1) +
  geom_point(colour = "#059669", size = 2.5) +
  geom_point(data = real_df |> filter(lambda == lambda_final),
             colour = "#EF4444", size = 4) +
  annotate("text",
           x     = 0.15,
           y     = real_df$risk[1] + 0.0002,
           label = sprintf("λ* = %.0f (no benefit)", lambda_final),
           colour = "#EF4444", size = 4) +
  labs(
    x     = "λ (weight on synthetic responses; 0 = pilot only)",
    y     = "Ability estimation risk",
    title = "Real data: risk increases with any synthetic weight"
  ) +
  theme_minimal(base_size = 13)
```

The risk is monotonically increasing — adding any weight to the synthetic data
*hurts* ability estimation. The method sets λ = 0, falling back to the
pilot-only estimate.

---

## Parameter recovery

Since λ = 0, the mixed-subjects and pilot-only models are identical. We compare
their difficulty estimates against the ground truth from the full
`r n_human_total`-subject dataset.

```{r}
#| label: fig-recovery
#| fig-cap: !expr sprintf("Pilot-only difficulty recovery (n=%d). RMSE = %.3f.", pilot_n, rmse_pilot)

b_df <- inner_join(
  gt_pars    |> select(item, b_gt = b),
  pilot_pars |> select(item, b_pilot = b),
  by = "item"
)

ggplot(b_df, aes(x = b_gt, y = b_pilot)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey70") +
  geom_point(size = 3, colour = "#7C3AED", alpha = 0.85) +
  labs(
    x     = "Ground-truth b (full sample, n = 735)",
    y     = sprintf("Estimated b (pilot only, n = %d)", pilot_n),
    title = sprintf("Difficulty parameter recovery  |  RMSE = %.3f", rmse_pilot)
  ) +
  theme_minimal(base_size = 13)
```

---

## Why does λ = 0?

The simulation proved the method works when predictions are high-quality.
So why does it fail here? Two reasons:

**Compressed predictions.** The text regression predicts pass rates in the
range 35%–60%, while true pass rates span 45%–81%. In IRT terms, the predicted
*b* values all fall in [−0.4, +0.6], but true *b* values range from −1.8 to
+0.2. The synthetic subjects "live" in a narrow band of the difficulty scale,
adding little information about the harder items where calibration would benefit
most.

**Adequate pilot.** With `r pilot_n` subjects and 20 items — a 5:1
subject-to-item ratio — the pilot-only 1PL model already achieves reasonable
precision (RMSE = `r round(rmse_pilot, 3)`). The marginal benefit from
noisy synthetic data is small relative to the PPI correction cost the method
must pay for mixing populations.

Contrast the simulation: 200 subjects, 20 items, b spread across [−2, 2],
oracle synthetic data → λ = `r round(sim_lambda, 3)`. The gains are modest
(0.7% risk reduction) but real. The method correctly identifies when synthetic
data helps and when it does not.

---

## Discussion

This proof of concept demonstrates two things: the mixed-subjects IRT framework
works as designed, and simple surface features (word count, average word length) are not yet a strong enough signal
to improve IRT calibration.

For text-based synthetic data to help, we would likely need: (1) richer feature
engineering — item format, response options, cognitive demand taxonomies; (2)
domain-specific text models; or (3) actual LLM responses scored against answer
keys rather than statistical proxies. The IRW's combination of item text, answer
keys, and large human response datasets makes it well-positioned for all three
directions.

The `mixedsubjectsirt` package and the [mixed-subjects IRT
preprint](https://klintkanopka.com/mixedsubjectsirt/) provide the calibration
framework. The `r_text`-correlation finding from the [item text
difficulty vignette](item_text_difficulty.html) points to where feature
engineering effort is most needed.
