---
title: "Network Psychometrics vs. IRT: Does Centrality Track Discrimination?"
description: |
  IRT explains why items correlate with a single latent trait; network
  psychometrics explains it with a graph of direct partial-correlation
  edges between items, no latent variable required. Under a strong single
  common cause, the two should agree: a network's node centrality should
  track IRT discrimination, and the network should look densely and
  uniformly connected. We test this across real IRW instruments and check
  whether the relationship tracks tables already flagged as
  multidimensional or locally dependent elsewhere on this site.
date: today
bibliography: network_psych_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)
library(qgraph)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "network_psych_data/network_psych_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache               <- readRDS(cache_path)
  summary_df           <- .cache$summary
  graphs_list          <- .cache$graphs %||% list()
  strengths_list       <- .cache$strengths %||% list()
  discriminations_list <- .cache$discriminations %||% list()
  candidate_tables     <- .cache$candidate_tables
  date_run             <- .cache$date_run
  is_pilot             <- isTRUE(.cache$pilot)
  n_all_candidates     <- .cache$n_all_candidates %||% length(candidate_tables)
  n_binary_all         <- .cache$n_binary_all %||% NA_integer_
  n_poly_all           <- .cache$n_poly_all %||% NA_integer_
  r_n_tables_run       <- length(candidate_tables)
} else {
  summary_df <- tibble(
    table = character(), n_items = integer(), n_participants = integer(),
    n_categories = integer(), network_method = character(), network_density = numeric(),
    strength_a_cor = numeric(), construct_type = character(), item_format = character(),
    dim_ratio_12 = numeric(), dim_nfact_suggested = integer(), dim_unidimensional = logical(),
    ld_prop_flagged = numeric(), ld_mean_abs_q3 = numeric(),
    has_dim_match = logical(), has_ld_match = logical()
  )
  graphs_list <- list(); strengths_list <- list(); discriminations_list <- list()
  candidate_tables <- character()
  date_run <- Sys.Date()
  is_pilot <- FALSE
  n_all_candidates <- 0L; n_binary_all <- NA_integer_; n_poly_all <- NA_integer_
  r_n_tables_run <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/network_psych_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, small pilot run.** This render uses a deliberately small ",
  "sample of ", r_n_tables_run, " tables (a mix of hand-picked known-good ",
  "tables and a random top-up), run live against Redivis, to validate the ",
  "pipeline and page layout before committing to a full batch across all ",
  n_all_candidates, " eligible tables. All numbers here are real computed ",
  "results -- not placeholders -- but treat the *distributions* as a small, ",
  "noisy preview rather than a representative picture. Re-run ",
  "`network_psych_compute.R` with `PILOT <- FALSE` to scale up.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`network_psych_compute.R`) directly if you're
relying on the numbers here.
:::

## Motivation

Every other vignette on this site treats item covariation the same way
IRT does: responses correlate because they all depend, more or less, on
one latent trait. **Network psychometrics** proposes a different account.
Instead of a shared cause behind the correlations, it models items as
nodes in a graph and estimates direct, partial-correlation edges between
them — an Ising model for binary items, a Gaussian graphical model
(EBICglasso) for polytomous ones. No latent variable is fit at all; the
graph *is* the model.

These two accounts aren't just philosophically different — they make a
shared, testable prediction. Van der Maas et al. (2006)
[@vandermaas2006dynamical] and Epskamp, Maris, van Borkulo, & Borsboom
(2018) [@epskamp2018testing] show that if a single dominant common cause
really does explain a set of items, the two descriptions should coincide:
a node's **centrality** (how strongly and directly it connects to every
other node) should track that item's **IRT discrimination**, and the
estimated network should look densely and fairly uniformly connected —
not clustered into visibly separate sub-groups. This vignette tests that
prediction across real IRW instruments, and asks whether it holds up
better or worse for tables this site's other diagnostics already flag as
multidimensional ([dimensionality vignette](dimensionality.qmd)) or
locally dependent ([local-dependence vignette](local_dependence.qmd)).

## Data and methods

### Dataset selection

```{r}
#| label: filter-datasets
#| eval: false

binary_tables <- irw_filter(n_items = c(5, 30), n_participants = c(300, Inf),
                             n_categories = 2)
poly_tables   <- irw_filter(n_items = c(5, 30), n_participants = c(300, Inf),
                             n_categories = c(3, 7))
```

Items are capped at 30 (lower than the dimensionality vignette's 40) and
polytomous tables are restricted to 3-7 response categories, mainly for
compute reasons: EBIC-regularized network estimation scales worse than a
single IRT fit or eigen-decomposition. `r n_all_candidates` tables met
this criterion as of `r format(date_run, "%B %Y")` (`r n_binary_all`
binary, `r n_poly_all` polytomous).
`r if (is_pilot) "This page currently runs on a small pilot subset (see the warning above); a full pass analyzes all of them." else paste0("This pass analyzes all ", r_n_tables_run, " of them.")`

### Per-table computation

For each candidate table, `fit_network()`:

1. Fetches the table and builds a wide response matrix, downsampling to
   at most 10,000 respondents and dropping zero-variance items, mirroring
   the other IRT vignettes on this site.
2. Estimates a regularized network with `bootnet::estimateNetwork()`:
   `IsingFit` for binary items, `EBICglasso` for polytomous items
   (treating ordinal Likert data as continuous — standard network-
   literature practice, an approximation rather than an exact ordinal
   network model; see Limitations).
3. Computes node centrality with `qgraph::centrality_auto()`. In a
   genuinely weighted network this returns a `Strength` column (each
   item's summed edge weight); a fully empty regularized network — every
   edge shrunk to zero — gets auto-detected as unweighted and falls back
   to `Degree` instead. That's still a real, informative result (kept in
   the data below), just not one with a meaningful strength-discrimination
   correlation, so that comparison is left `NA` for those tables.
4. Fits the same one-factor model used elsewhere on this site — 2PL for
   binary, the graded response model for polytomous — and extracts item
   discrimination.
5. Records the correlation between node strength and discrimination
   across items — the vignette's core comparison — plus network density
   (proportion of possible edges present) and `construct_type`/
   `item_format` from IRW's tags.
6. Joins in, where the same table appears: the eigenvalue ratio /
   parallel-analysis outcome from the [dimensionality
   vignette](dimensionality.qmd)'s cache, and the proportion of Q3-flagged
   pairs from the [local-dependence vignette](local_dependence.qmd)'s
   cache — read directly from their saved results rather than
   recomputed.

```{r}
#| label: fit-function
#| eval: false

fit_network <- function(table_name) {
  resp <- fetch_wide(table_name)  # irw_fetch() + irw_long2resp(), cleaned
  net  <- estimateNetwork(resp, default = if (n_categories == 2) "IsingFit" else "EBICglasso")
  cent <- centrality_auto(net$graph)
  fit  <- mirt(resp, 1, itemtype = if (n_categories == 2) "2PL" else "graded")
  a    <- coef(fit, simplify = TRUE)$items[, "a1"]
  cor(cent$node.centrality$Strength, a)  # the core comparison
}
```

Of the `r r_n_tables_run` candidate tables in this pass, `r nrow(summary_df)`
produced usable results (the rest were dropped for too few usable items,
an unsupported column type on fetch, a network/IRT model that failed to
fit, or too few remaining response categories). This pass matched a
dimensionality-vignette entry for `r sum(summary_df$has_dim_match)` of
those tables and a local-dependence-vignette entry for
`r sum(summary_df$has_ld_match)` — both caches were built on an
overlapping but not identical candidate pool, so coverage is high but not
total.

::: {.callout-note}
To reproduce with current IRW holdings, re-run `network_psych_compute.R`
(after the dimensionality and local-dependence compute scripts, which it
depends on) and commit the updated
`network_psych_data/network_psych_results.rds`.
:::

## Results

### Two concrete examples, opposite ends of the relationship

```{r}
#| label: fig-example-network
#| fig-cap: "Estimated network (left) and IRT discrimination parameters (right) for the table with the strongest strength-discrimination correlation in this sample (top) and the weakest (bottom). Top: a dense, uniformly-connected graph with node positions that roughly track item discrimination -- exactly what Epskamp et al.'s theory predicts under a strong single factor. Bottom: edge thickness and node position barely track discrimination at all."
#| fig-height: 10
#| fig-width: 11

example_candidates <- summary_df %>%
  filter(!is.na(strength_a_cor), network_density > 0) %>%
  rowwise() %>%
  # Exclude numerically degenerate IRT fits -- e.g. a near-duplicate item
  # pair can drive discrimination to an implausible boundary value (a ~ 50),
  # which makes for a misleading rather than illustrative example.
  mutate(max_abs_a = max(abs(discriminations_list[[table]]), na.rm = TRUE)) %>%
  ungroup() %>%
  filter(max_abs_a < 5)
hi_table <- example_candidates %>% slice_max(strength_a_cor, n = 1, with_ties = FALSE) %>% pull(table)
lo_table <- example_candidates %>% slice_min(strength_a_cor, n = 1, with_ties = FALSE) %>% pull(table)

hi_cor <- summary_df$strength_a_cor[summary_df$table == hi_table]
lo_cor <- summary_df$strength_a_cor[summary_df$table == lo_table]

plot_example_pair <- function(table_name, label) {
  g <- graphs_list[[table_name]]
  s <- strengths_list[[table_name]]
  a <- discriminations_list[[table_name]]

  par(mar = c(4, 1, 4, 1))
  qgraph::qgraph(g, layout = "spring", labels = names(s),
                 label.cex = 0.9, vsize = 11, edge.color = irw_blue,
                 title = paste0(label, ": ", table_name, "\nestimated network"),
                 title.cex = 1)

  par(mar = c(4, 10, 4, 2))
  ord <- order(a)
  barplot(a[ord], names.arg = names(a)[ord], horiz = TRUE, las = 1,
          cex.names = 0.8, col = irw_blue, border = NA,
          xlab = "IRT discrimination (a)",
          main = paste0(label, ": ", table_name, "\ndiscrimination"), cex.main = 1)
}

par(mfrow = c(2, 2))
plot_example_pair(hi_table, "Highest correlation")
plot_example_pair(lo_table, "Lowest correlation")
```

```{r}
#| label: example-pair-inline
#| echo: false

hi_density <- summary_df$network_density[summary_df$table == hi_table]
lo_density <- summary_df$network_density[summary_df$table == lo_table]
lo_a       <- discriminations_list[[lo_table]]
n_lo_neg_a <- sum(lo_a < 0)
```

`r hi_table` has a strength-discrimination correlation of `r round(hi_cor, 2)`
— about as clean a confirmation of the theory as this sample produces.
`r lo_table` sits at the opposite end, at `r round(lo_cor, 2)`: the two
descriptions of the same items agree much less about which ones matter
most, to the point of actively disagreeing about direction.

What makes the contrast worth dwelling on is what *doesn't* differ between
them: both networks are fully connected (density
`r round(hi_density, 2)` and `r round(lo_density, 2)` respectively — every
possible edge present in both). Density and uniformity, the other half of
Epskamp et al.'s prediction, look identical here; only the
strength-discrimination correlation reveals that one table matches the
single-common-cause story and the other doesn't. A network that merely
*looks* dense and uniform is not, on its own, evidence that IRT and
network psychometrics agree about the items — this pair is a direct
counterexample.

The disagreement in `r lo_table` isn't subtle, either:
`r n_lo_neg_a` of its `r length(lo_a)` items get a *negative* IRT
discrimination, meaning the 2PL model estimates them as running opposite
the trait it fit from the rest — yet the network still connects every one
of those items to every other about as strongly as any pair in
`r hi_table`. The two frameworks aren't just disagreeing on a matter of
degree; they disagree on which items belong together at all. Practically,
that matters most for anyone using either statistic to pick or drop items
— shortening a scale by discrimination, or by network centrality, is not
a neutral choice, and a table like this one is exactly where the two
selections would part ways. Compare both examples to the aggregate
picture below: most tables fall somewhere between these two extremes, but
not always close to the middle.

### Does strength track discrimination in general?

```{r}
#| label: fig-cor-ranked
#| fig-cap: "Correlation between network node strength and IRT discrimination, one point per table, ranked from lowest to highest. Values near 1 match the theoretical prediction under a strong single factor; low or negative values indicate a breakdown. Hover over a point to see which table it is."
#| fig-height: 6

ranked_df <- summary_df %>%
  filter(!is.na(strength_a_cor)) %>%
  arrange(strength_a_cor) %>%
  mutate(
    rank = row_number(),
    tooltip = paste0(
      "Table: ", table,
      "<br>Strength-a correlation: ", round(strength_a_cor, 2),
      "<br>Network method: ", network_method,
      "<br>Network density: ", round(network_density, 2),
      "<br>n_items: ", n_items
    )
  )

p <- ggplot(ranked_df, aes(x = strength_a_cor, y = rank, text = tooltip)) +
  geom_point(colour = irw_blue, size = 1.6, alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  labs(x = "Correlation: node strength vs. IRT discrimination", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: cor-summary-inline
#| echo: false

med_cor <- median(summary_df$strength_a_cor, na.rm = TRUE)
prop_strong <- mean(summary_df$strength_a_cor > 0.7, na.rm = TRUE)
n_na_cor <- sum(is.na(summary_df$strength_a_cor))
```

The median strength-discrimination correlation across `r sum(!is.na(summary_df$strength_a_cor))`
tables with a usable comparison is `r round(med_cor, 2)`, and
`r scales::percent(prop_strong, accuracy = 1)` of tables exceed 0.7.
`r if (n_na_cor > 0) paste0(n_na_cor, " table(s) are excluded here because their regularized network had no edges at all -- an empty network, not a low correlation.") else ""`
The relationship is real and often strong, but — as the two comparisons
below make concrete — it is not universal.

### Cross-vignette payoff 1: does it track the dimensionality diagnostic?

```{r}
#| label: fig-cor-vs-dimensionality
#| fig-cap: "Strength-discrimination correlation against the dimensionality vignette's eigenvalue ratio (log scale, capped at 15 for display). If Epskamp et al.'s theory holds, the correlation should sit near 1 for tables with a high (clearly unidimensional) ratio and scatter lower as the ratio approaches the vignette's ratio = 4 heuristic cutoff or below. Hover for table details."
#| fig-height: 5

join_dim_df <- summary_df %>%
  filter(!is.na(strength_a_cor), !is.na(dim_ratio_12)) %>%
  mutate(tooltip = paste0(
    "Table: ", table,
    "<br>Strength-a cor: ", round(strength_a_cor, 2),
    "<br>Eigenvalue ratio: ", round(dim_ratio_12, 2),
    "<br>PA unidimensional: ", dim_unidimensional
  ))

p <- ggplot(join_dim_df, aes(x = pmin(dim_ratio_12, 15), y = strength_a_cor, text = tooltip)) +
  geom_point(colour = irw_blue, size = 2, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = irw_red, linewidth = 0.8) +
  scale_x_log10() +
  labs(x = "Eigenvalue ratio (dimensionality vignette, log scale, capped at 15)",
       y = "Strength-discrimination correlation")

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: dim-join-inline
#| echo: false

dim_join_full <- join_dim_df %>% filter(!is.na(dim_unidimensional))
n_dim_join    <- nrow(join_dim_df)
cor_dim_join  <- cor(join_dim_df$strength_a_cor, log(join_dim_df$dim_ratio_12))
n_flagged_uni   <- sum(dim_join_full$dim_unidimensional)
n_flagged_multi <- sum(!dim_join_full$dim_unidimensional)
med_cor_uni     <- median(dim_join_full$strength_a_cor[dim_join_full$dim_unidimensional])
med_cor_multi   <- median(dim_join_full$strength_a_cor[!dim_join_full$dim_unidimensional])
wilcox_p_dim    <- wilcox.test(
  dim_join_full$strength_a_cor[dim_join_full$dim_unidimensional],
  dim_join_full$strength_a_cor[!dim_join_full$dim_unidimensional]
)$p.value
```

Across the `r n_dim_join` tables with a match in both this vignette and the
[dimensionality vignette](dimensionality.qmd), the two diagnostics move
together weakly but reliably (r = `r round(cor_dim_join, 2)` between
strength-discrimination correlation and log eigenvalue ratio) — real
signal in the direction the theory predicts, not a tight relationship.
The theory's prediction shows up more clearly when split by the
dimensionality vignette's own parallel-analysis call: the
`r n_flagged_uni` tables it flags as unidimensional have a median
strength-discrimination correlation of `r round(med_cor_uni, 2)`, versus
`r round(med_cor_multi, 2)` for the `r n_flagged_multi` tables it flags as
multidimensional (Wilcoxon `r if (wilcox_p_dim < 0.001) "p < 0.001" else paste0("p = ", round(wilcox_p_dim, 3))`).
Parallel analysis is strict — only a small fraction of tables clear its
bar for "unidimensional" — but among the ones that do, network centrality
and IRT discrimination agree noticeably better.

### Cross-vignette payoff 2: do locally dependent tables look denser?

```{r}
#| label: fig-density-vs-localdep
#| fig-cap: "Network density against the local-dependence vignette's proportion of Q3-flagged item pairs. A testlet or shared-stimulus effect should show up as extra edges concentrated among the affected items -- more flagged pairs, denser estimated networks -- rather than the theory's predicted uniform density. Hover for table details."
#| fig-height: 5

join_ld_df <- summary_df %>%
  filter(!is.na(network_density), !is.na(ld_prop_flagged)) %>%
  mutate(tooltip = paste0(
    "Table: ", table,
    "<br>Network density: ", round(network_density, 2),
    "<br>Prop. Q3-flagged: ", round(ld_prop_flagged, 2)
  ))

p <- ggplot(join_ld_df, aes(x = ld_prop_flagged, y = network_density, text = tooltip)) +
  geom_point(colour = irw_blue, size = 2, alpha = 0.7) +
  labs(x = "Proportion of pairs Q3-flagged (local-dependence vignette)",
       y = "Network density")

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: ld-join-inline
#| echo: false

n_ld_join    <- nrow(join_ld_df)
cor_ld_join  <- cor(join_ld_df$network_density, join_ld_df$ld_prop_flagged)
cor_test_ld  <- cor.test(join_ld_df$network_density, join_ld_df$ld_prop_flagged)
```

Across the `r n_ld_join` tables with a match in both this vignette and the
[local-dependence vignette](local_dependence.qmd), network density and the
proportion of Q3-flagged pairs correlate at r = `r round(cor_ld_join, 2)`
(`r if (cor_test_ld$p.value < 0.001) "p < 0.001" else paste0("p = ", round(cor_test_ld$p.value, 3))`)
— the expected sign (more flagged pairs, denser network), and with
`r n_ld_join` tables behind it, a reliable one. Density alone doesn't distinguish a uniformly
dense network from one with a few tightly-connected clusters sitting
inside an otherwise sparse graph — the latter is the more specific
testlet signature, and it's exactly what the local-dependence vignette's
[Q3 heatmaps](local_dependence.qmd#a-worst-case-and-a-clean-case-side-by-side)
are built to catch directly. Detecting that cluster structure from the
network side properly needs community-detection tools this pass didn't
run at batch scale (see Limitations); density is the coarser proxy
available here.

## Practical implications

Where the strength-discrimination correlation is high and the network is
dense and uniform, network psychometrics and IRT are, practically
speaking, telling the same story about the same items — using either
framework to identify the "best" items should agree. Where it breaks
down, that's a signal worth taking seriously in its own right, independent
of which framework you prefer: it means the two ways of describing item
covariation actually disagree about which items matter most, and a table
flagged as multidimensional or locally dependent elsewhere on this site
is a reasonable place to expect exactly that disagreement.

## Limitations

- **EBICglasso on ordinal data** treats Likert-type responses as
  continuous, standard practice in the network-psychometrics literature
  but an approximation, not an exact ordinal network model.
- **Network estimation is more sample- and tuning-sensitive** than fitting
  a single IRT model. Small tables in this sample should be read
  cautiously; an empty regularized network (no edges survive
  regularization) is itself informative but excludes that table from the
  strength-discrimination comparison entirely.
- **No bootstrap stability checks.** `bootnet`'s case-dropping bootstrap
  (the standard way to check whether an estimated network's edges and
  centrality ordering are stable) was not run at full-batch scale for
  compute reasons. A natural follow-up: run it for a handful of the most
  interesting example tables from the full batch (the strongest
  confirmation, the sharpest breakdown) rather than all of them.
- **Density is a coarse proxy for "clustered."** As noted above, telling
  a uniformly dense network apart from one with tight sub-clusters needs
  community-detection tools this pass didn't run.
- **Cache overlap is partial.** The dimensionality and local-dependence
  caches were built on their own candidate pools (different item-count
  caps), so not every table here has a match in both — see the match-rate
  numbers in Data and methods.

## Reproducibility

```{r}
#| label: repro
#| echo: true
#| eval: false

# 1. Re-run the prerequisite compute scripts first (from the project root)
Rscript vignettes/dimensionality_compute.R
Rscript vignettes/local_dependence_compute.R

# 2. Re-run this vignette's compute script
Rscript vignettes/network_psych_compute.R

# 3. Re-render this page
quarto render vignettes/network_psych.qmd
```

Session info from the compute run: R `r if (cache_ok) .cache$session$R.version$version.string else "N/A"`,
run on `r format(date_run, "%B %d, %Y")`.

## References

::: {#refs}
:::
