---
title: "Network Psychometrics vs. IRT: Does Centrality Track Discrimination?"
description: |
  IRT explains why items correlate with a single latent trait; network
  psychometrics explains it with a graph of direct partial-correlation
  edges between items, no latent variable required. Under a strong single
  common cause, the two should agree: a network's node centrality should
  track IRT discrimination, and the network should look densely and
  uniformly connected. We test this across real IRW instruments and check
  whether the relationship tracks tables already flagged as
  multidimensional or locally dependent elsewhere on this site.
date: today
bibliography: network_psych_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(plotly)
library(qgraph)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "network_psych_data/network_psych_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache               <- readRDS(cache_path)
  summary_df           <- .cache$summary
  graphs_list          <- .cache$graphs %||% list()
  strengths_list       <- .cache$strengths %||% list()
  discriminations_list <- .cache$discriminations %||% list()
  edges_ggm_df         <- .cache$edges_ggm %||% tibble()
  edges_mrf_df         <- .cache$edges_mrf %||% tibble()
  option_b_tables      <- .cache$option_b_tables %||% character()
  candidate_tables     <- .cache$candidate_tables
  date_run             <- .cache$date_run
  is_pilot             <- isTRUE(.cache$pilot)
  n_all_candidates     <- .cache$n_all_candidates %||% length(candidate_tables)
  n_binary_all         <- .cache$n_binary_all %||% NA_integer_
  n_poly_all           <- .cache$n_poly_all %||% NA_integer_
  r_n_tables_run       <- length(candidate_tables)
} else {
  summary_df <- tibble(
    table = character(), n_items = integer(), n_participants = integer(),
    n_categories = integer(), network_method = character(), network_density = numeric(),
    strength_a_cor = numeric(), construct_type = character(), item_format = character(),
    dim_ratio_12 = numeric(), dim_nfact_suggested = integer(), dim_unidimensional = logical(),
    ld_prop_flagged = numeric(), ld_mean_abs_q3 = numeric(),
    has_dim_match = logical(), has_ld_match = logical(),
    has_bayes_match = logical(), has_mrf_match = logical(),
    n_used_ggm = integer(), relative_n_ggm = numeric(),
    prop_strong_presence_ggm = numeric(), prop_weak_presence_ggm = numeric(),
    prop_inconclusive_ggm = numeric(), prop_weak_absence_ggm = numeric(),
    prop_strong_absence_ggm = numeric(),
    n_used_mrf = integer(), relative_n_mrf = numeric(),
    prop_strong_presence_mrf = numeric(), prop_weak_presence_mrf = numeric(),
    prop_inconclusive_mrf = numeric(), prop_weak_absence_mrf = numeric(),
    prop_strong_absence_mrf = numeric()
  )
  graphs_list <- list(); strengths_list <- list(); discriminations_list <- list()
  edges_ggm_df <- tibble(); edges_mrf_df <- tibble(); option_b_tables <- character()
  candidate_tables <- character()
  date_run <- Sys.Date()
  is_pilot <- FALSE
  n_all_candidates <- 0L; n_binary_all <- NA_integer_; n_poly_all <- NA_integer_
  r_n_tables_run <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/network_psych_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, small pilot run.** This render uses a deliberately small ",
  "sample of ", r_n_tables_run, " tables (a mix of hand-picked known-good ",
  "tables and a random top-up), run live against Redivis, to validate the ",
  "pipeline and page layout before committing to a full batch across all ",
  n_all_candidates, " eligible tables. All numbers here are real computed ",
  "results -- not placeholders -- but treat the *distributions* as a small, ",
  "noisy preview rather than a representative picture. Re-run ",
  "`network_psych_compute.R` with `PILOT <- FALSE` to scale up.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`network_psych_compute.R`) directly if you're
relying on the numbers here.
:::

## Motivation

Every other vignette on this site treats item covariation the same way
IRT does: responses correlate because they all depend, more or less, on
one latent trait. **Network psychometrics** proposes a different account.
Instead of a shared cause behind the correlations, it models items as
nodes in a graph and estimates direct, partial-correlation edges between
them — an Ising model for binary items, a Gaussian graphical model
(EBICglasso) for polytomous ones. No latent variable is fit at all; the
graph *is* the model.

These two accounts aren't just philosophically different — they make a
shared, testable prediction. Van der Maas et al. (2006)
[@vandermaas2006dynamical] and Epskamp, Maris, van Borkulo, & Borsboom
(2018) [@epskamp2018testing] show that if a single dominant common cause
really does explain a set of items, the two descriptions should coincide.
This isn't just an informal observation: Marsman et al. (2018)
[@marsman2018introduction] prove a formal statistical equivalence between
the Ising model and the 2PL for binary items, and Marsman, van den Bergh,
& Haslbeck (2025) [@marsman2025bayesian] extend that equivalence to
ordinal items -- with a wrinkle noted in Data and methods below. Under
that equivalence, a node's **centrality** (how strongly and directly it connects to every
other node) should track that item's **IRT discrimination**, and the
estimated network should look densely and fairly uniformly connected —
not clustered into visibly separate sub-groups. This vignette tests that
prediction across real IRW instruments, and asks whether it holds up
better or worse for tables this site's other diagnostics already flag as
multidimensional ([dimensionality vignette](dimensionality.qmd)) or
locally dependent ([local-dependence vignette](local_dependence.qmd)).

## Data and methods

### Dataset selection

```{r}
#| label: filter-datasets
#| eval: false

binary_tables <- irw_filter(n_items = c(5, 30), n_participants = c(300, Inf),
                             n_categories = 2)
poly_tables   <- irw_filter(n_items = c(5, 30), n_participants = c(300, Inf),
                             n_categories = c(3, 7))
```

Items are capped at 30 (lower than the dimensionality vignette's 40) and
polytomous tables are restricted to 3-7 response categories, mainly for
compute reasons: EBIC-regularized network estimation scales worse than a
single IRT fit or eigen-decomposition. `r n_all_candidates` tables met
this criterion as of `r format(date_run, "%B %Y")` (`r n_binary_all`
binary, `r n_poly_all` polytomous).
`r if (is_pilot) "This page currently runs on a small pilot subset (see the warning above); a full pass analyzes all of them." else paste0("This pass analyzes all ", r_n_tables_run, " of them.")`

### Per-table computation

For each candidate table, `fit_network()`:

1. Fetches the table and builds a wide response matrix, downsampling to
   at most 10,000 respondents and dropping zero-variance items, mirroring
   the other IRT vignettes on this site.
2. Estimates a regularized network with `bootnet::estimateNetwork()`:
   `IsingFit` for binary items, `EBICglasso` for polytomous items. The
   polytomous correlations are computed via `qgraph::cor_auto()`
   (`corMethod = "cor_auto"`, `corArgs = list(forcePD = TRUE)`), which
   estimates polychoric correlations for ordinal data and projects to the
   nearest valid correlation matrix when the polychoric estimate isn't
   positive-definite — this still treats the *graph* estimation step as
   Gaussian, an approximation relative to an exact ordinal network model,
   but does not additionally treat Likert categories as literally
   continuous integers (see Limitations for why both of those still fall
   short of the ordinal Markov random field with a proven IRT
   equivalence).
3. Computes node centrality with `qgraph::centrality_auto()`. In a
   genuinely weighted network this returns a `Strength` column (each
   item's summed edge weight); a fully empty regularized network — every
   edge shrunk to zero — gets auto-detected as unweighted and falls back
   to `Degree` instead. That's still a real, informative result (kept in
   the data below), just not one with a meaningful strength-discrimination
   correlation, so that comparison is left `NA` for those tables.
4. Fits a one-factor IRT model and extracts item discrimination: 2PL for
   binary items; the **generalized partial credit model (GPCM)**, not the
   graded response model used elsewhere on this site, for polytomous items
   — GPCM is the model with a proven equivalence to the ordinal network
   model this vignette's polytomous branch estimates (see Limitations).
5. Records the correlation between node strength and discrimination
   across items — the vignette's core comparison — plus network density
   (proportion of possible edges present) and `construct_type`/
   `item_format` from IRW's tags.
6. Joins in, where the same table appears: the eigenvalue ratio /
   parallel-analysis outcome from the [dimensionality
   vignette](dimensionality.qmd)'s cache, and the proportion of Q3-flagged
   pairs from the [local-dependence vignette](local_dependence.qmd)'s
   cache — read directly from their saved results rather than
   recomputed.

```{r}
#| label: fit-function
#| eval: false

fit_network <- function(table_name) {
  resp <- fetch_wide(table_name)  # irw_fetch() + irw_long2resp(), cleaned
  net  <- if (n_categories == 2) {
    estimateNetwork(resp, default = "IsingFit")
  } else {
    estimateNetwork(resp, default = "EBICglasso", corMethod = "cor_auto")
  }
  cent <- centrality_auto(net$graph)
  fit  <- mirt(resp, 1, itemtype = if (n_categories == 2) "2PL" else "gpcm")
  a    <- coef(fit, simplify = TRUE)$items[, "a1"]
  cor(cent$node.centrality$Strength, a)  # the core comparison
}
```

Of the `r r_n_tables_run` candidate tables in this pass, `r nrow(summary_df)`
produced usable results (the rest were dropped for too few usable items,
an unsupported column type on fetch, a network/IRT model that failed to
fit, or too few remaining response categories). This pass matched a
dimensionality-vignette entry for `r sum(summary_df$has_dim_match)` of
those tables and a local-dependence-vignette entry for
`r sum(summary_df$has_ld_match)` — both caches were built on an
overlapping but not identical candidate pool, so coverage is high but not
total.

::: {.callout-note}
To reproduce with current IRW holdings, re-run `network_psych_compute.R`
(after the dimensionality and local-dependence compute scripts, which it
depends on) and commit the updated
`network_psych_data/network_psych_results.rds`.
:::

### Bayesian edge evidence: which model, and why

Everything above treats an edge as present or absent based on whether
regularized network estimation shrinks it to exactly zero — a single
point estimate, with no sense of how much statistical evidence actually
backs that call. Huth, Haslbeck, Keetelaar, van Holst, & Marsman (2026)
[@huth2026statistical] reanalyzed 293 published psychological networks
with a fully Bayesian alternative: fit each network as a Gaussian
graphical model (GGM) with a matrix-F prior on the partial correlations,
then compute an edge-inclusion Bayes factor (BF10) for every edge via the
Savage-Dickey density ratio. They found evidence was frequently much
weaker than a single regularized graph suggests — only 18.7% of edges in
the literature they reanalyzed had *strong* evidence either way.

Their software choice matters for how this section reproduces their
analysis. They fit **every** network in their reanalysis as a GGM via the
`BGGM` package [@williams2020bggm], accessed through the `easybgm` wrapper
[@huth2024easybgm] — uniformly, including networks built from binary or
ordinal items, explicitly as a simplifying choice for consistency across
a reanalysis spanning many published papers. They say directly that they
know this is an approximation for non-continuous data, and that the
ordinal Markov random field (fit via the `bgms` package) is the
conceptually better-suited model for that case — they just didn't use it.

IRW is overwhelmingly binary/ordinal item response data, which makes this
a real design decision here too, not just an aside:

- **Option A (primary, reported below):** fit every candidate table as a
  GGM via `easybgm(..., package = "BGGM")`, forced uniformly regardless of
  each table's actual `n_categories` — the same simplifying choice the
  paper made, for the same reason: it's the only way to compare IRW's
  breakdown directly against their headline percentages. `package = "BGGM"`
  has to be forced explicitly and paired with `type = "continuous"`: the
  installed `easybgm` (0.4.0) silently redirects back to `bgms` whenever
  `type = "binary"` is passed, even if `package = "BGGM"` was requested,
  which would otherwise quietly defeat this whole comparison.
- **Option B (secondary, small subset only):** fit the ordinal Markov
  random field via `bgms` instead — the model the paper itself calls more
  appropriate for this data type. Piloting this on real IRW tables found
  it costs roughly 20-50x Option A's runtime per table (138-192 sec vs.
  1-10 sec on the same two tables); a full batch across IRW's ~660
  candidate tables would take 7-27 hours. It runs here on a small
  stratified subset (`r length(option_b_tables)` tables, split across
  binary and polytomous items) as an illustrative robustness check, not a
  full-batch analysis.

Both use the paper's own matrix-F prior default (`prior_sd = 0.25`) and
classify each edge into the paper's five-way breakdown by its BF10:
strong presence (BF10 > 10), weak presence (3 < BF10 <= 10), inconclusive
(1/3 <= BF10 <= 3), weak absence (1/10 <= BF10 < 1/3), or strong absence
(BF10 < 1/10) — a small deviation from the paper's stated open-interval
boundaries (which don't specify which side of e.g. BF10 = 3 owns that
exact value), noted here since it's a real, if practically inconsequential,
difference from a literal reading of their method.

## Results

### Two concrete examples, opposite ends of the relationship

```{r}
#| label: fig-example-network
#| fig-cap: "Estimated network (left) and IRT discrimination parameters (right) for the table with the strongest strength-discrimination correlation in this sample (top) and the weakest (bottom). Edge color shows sign: blue = positive partial correlation, red = negative. Top: a dense, uniformly-connected, uniformly positive graph with node positions that roughly track item discrimination -- exactly what Epskamp et al.'s theory predicts under a strong single factor. Bottom: edge thickness and node position barely track discrimination at all, and every edge is negative."
#| fig-height: 10
#| fig-width: 11

example_candidates <- summary_df %>%
  filter(!is.na(strength_a_cor), network_density > 0) %>%
  rowwise() %>%
  # Exclude numerically degenerate IRT fits -- e.g. a near-duplicate item
  # pair can drive discrimination to an implausible boundary value (a ~ 50),
  # which makes for a misleading rather than illustrative example.
  mutate(max_abs_a = max(abs(discriminations_list[[table]]), na.rm = TRUE)) %>%
  ungroup() %>%
  filter(max_abs_a < 5)
hi_table <- example_candidates %>% slice_max(strength_a_cor, n = 1, with_ties = FALSE) %>% pull(table)
lo_table <- example_candidates %>% slice_min(strength_a_cor, n = 1, with_ties = FALSE) %>% pull(table)

hi_cor <- summary_df$strength_a_cor[summary_df$table == hi_table]
lo_cor <- summary_df$strength_a_cor[summary_df$table == lo_table]

plot_example_pair <- function(table_name, label) {
  g <- graphs_list[[table_name]]
  s <- strengths_list[[table_name]]
  a <- discriminations_list[[table_name]]

  par(mar = c(4, 1, 4, 1))
  qgraph::qgraph(g, layout = "spring", labels = names(s),
                 label.cex = 0.9, vsize = 11, posCol = irw_blue, negCol = irw_red,
                 title = paste0(label, ": ", table_name, "\nestimated network"),
                 title.cex = 1)

  par(mar = c(4, 10, 4, 2))
  ord <- order(a)
  barplot(a[ord], names.arg = names(a)[ord], horiz = TRUE, las = 1,
          cex.names = 0.8, col = irw_blue, border = NA,
          xlab = "IRT discrimination (a)",
          main = paste0(label, ": ", table_name, "\ndiscrimination"), cex.main = 1)
}

par(mfrow = c(2, 2))
plot_example_pair(hi_table, "Highest correlation")
plot_example_pair(lo_table, "Lowest correlation")
```

```{r}
#| label: example-pair-inline
#| echo: false

hi_density <- summary_df$network_density[summary_df$table == hi_table]
lo_density <- summary_df$network_density[summary_df$table == lo_table]
lo_a       <- discriminations_list[[lo_table]]
n_lo_neg_a <- sum(lo_a < 0)
lo_g       <- graphs_list[[lo_table]]
lo_all_neg <- all(lo_g[upper.tri(lo_g)] <= 0)
lo_irt_model <- if (summary_df$n_categories[summary_df$table == lo_table] == 2) {
  "2PL model"
} else {
  "graded response model"
}
```

`r hi_table` has a strength-discrimination correlation of `r round(hi_cor, 2)`
— about as clean a confirmation of the theory as this sample produces.
`r lo_table` sits at the opposite end, at `r round(lo_cor, 2)`: the two
descriptions of the same items agree much less about which ones matter
most, to the point of actively disagreeing about direction.

What makes the contrast worth dwelling on is what *doesn't* differ much
between them: both networks are comparably dense (density
`r round(hi_density, 2)` and `r round(lo_density, 2)` respectively).
Density and uniformity, the other half of
Epskamp et al.'s prediction, don't distinguish these two cases well; only the
strength-discrimination correlation reveals that one table matches the
single-common-cause story and the other doesn't. A network that merely
*looks* dense and uniform is not, on its own, evidence that IRT and
network psychometrics agree about the items — this pair is a direct
counterexample.

The disagreement in `r lo_table` isn't subtle, either:
`r n_lo_neg_a` of its `r length(lo_a)` items get a *negative* IRT
discrimination, meaning the `r lo_irt_model` estimates them as running
opposite the trait it fit from the rest.
`r if (lo_all_neg) "Every edge in the estimated network is negative (visible from the red edges above), not merely non-uniform in strength -- the two frameworks aren't just disagreeing on a matter of degree; the network sees a single tightly (negatively) coupled cluster where IRT sees a trait some items run backward on." else "The network still connects those items to every other about as strongly as any pair in hi_table, despite the sign disagreement in the IRT fit."`
Practically,
that matters most for anyone using either statistic to pick or drop items
— shortening a scale by discrimination, or by network centrality, is not
a neutral choice, and a table like this one is exactly where the two
selections would part ways. Compare both examples to the aggregate
picture below: most tables fall somewhere between these two extremes, but
not always close to the middle.

```{r}
#| label: clifford-aside-inline
#| echo: false

clifford_table <- "clifford_2018_police_blame"
clifford_present <- clifford_table %in% summary_df$table
if (clifford_present) {
  clifford_cor <- summary_df$strength_a_cor[summary_df$table == clifford_table]
}
```

::: {.callout-note collapse="true"}
#### A worked case for *why* a single common cause can fail, not just that it does

`r if (clifford_present) paste0("One table worth a specific look regardless of where it happens to fall in the ranking above -- its strength-discrimination correlation this pass is ", round(clifford_cor, 2), " -- is ", clifford_table, ".") else paste0(clifford_table, " didn't appear among this pass's usable tables, but the underlying item design below is worth keeping in mind for similar multi-target blame/attribution items elsewhere in IRW.")` Each item asks the identically-worded "how much blame does X
deserve" about a different target in the 2018 Sacramento police shooting of
Stephon Clark -- state officials and the responding officers on one side,
and the victim himself (`item_blame_clark`) on the other. Blaming the
victim is a substantively different judgment from blaming officials, and
the raw item correlations bear that out: `item_blame_clark` correlates
negatively with every other item (fetched directly from IRW, not shown
here), while the official-blame items mostly correlate positively with
each other. That's a real bipolar attitude structure, not a data-entry or
reverse-coding artifact -- there's no oppositely-worded item to un-reverse;
every item shares the same "how much blame" wording, only the target
changes. It's a reminder that a low (or negative) strength-discrimination
correlation doesn't always mean the same thing: sometimes it's a
genuinely multidimensional or bipolar construct like this one, and
sometimes -- as with the current lowest-correlation example above -- the
network and the IRT model simply weight the same roughly-unidimensional
items differently.
:::

### Does strength track discrimination in general?

```{r}
#| label: fig-cor-ranked
#| fig-cap: "Correlation between network node strength and IRT discrimination, one point per table. The y-axis carries no information of its own -- it is each table's rank by that correlation (low to high), used only to spread overlapping points out vertically so individual tables are separable; read the plot along the x-axis only. Values near 1 match the theoretical prediction under a strong single factor; low or negative values indicate a breakdown. Hover over a point to see which table it is."
#| fig-height: 6

ranked_df <- summary_df %>%
  filter(!is.na(strength_a_cor)) %>%
  arrange(strength_a_cor) %>%
  mutate(
    rank = row_number(),
    tooltip = paste0(
      "Table: ", table,
      "<br>Strength-a correlation: ", round(strength_a_cor, 2),
      "<br>Network method: ", network_method,
      "<br>Network density: ", round(network_density, 2),
      "<br>n_items: ", n_items
    )
  )

p <- ggplot(ranked_df, aes(x = strength_a_cor, y = rank, text = tooltip)) +
  geom_point(colour = irw_blue, size = 1.6, alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  labs(x = "Correlation: node strength vs. IRT discrimination",
       y = "Rank (spreads points out only -- carries no meaning of its own)") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: cor-summary-inline
#| echo: false

med_cor <- median(summary_df$strength_a_cor, na.rm = TRUE)
prop_strong <- mean(summary_df$strength_a_cor > 0.7, na.rm = TRUE)
n_na_cor <- sum(is.na(summary_df$strength_a_cor))
```

The median strength-discrimination correlation across `r sum(!is.na(summary_df$strength_a_cor))`
tables with a usable comparison is `r round(med_cor, 2)`, and
`r scales::percent(prop_strong, accuracy = 1)` of tables exceed 0.7.
`r if (n_na_cor > 0) paste0(n_na_cor, " table(s) are excluded here because their regularized network had no edges at all -- an empty network, not a low correlation.") else ""`
The relationship is real and often strong, but — as the two comparisons
below make concrete — it is not universal.

### Cross-vignette payoff 1: does it track the dimensionality diagnostic?

```{r}
#| label: fig-cor-vs-dimensionality
#| fig-cap: "Strength-discrimination correlation against the dimensionality vignette's eigenvalue ratio (log scale, capped at 15 for display). If Epskamp et al.'s theory holds, the correlation should sit near 1 for tables with a high (clearly unidimensional) ratio and scatter lower as the ratio approaches the vignette's ratio = 4 heuristic cutoff or below. Hover for table details."
#| fig-height: 5

join_dim_df <- summary_df %>%
  filter(!is.na(strength_a_cor), !is.na(dim_ratio_12)) %>%
  mutate(tooltip = paste0(
    "Table: ", table,
    "<br>Strength-a cor: ", round(strength_a_cor, 2),
    "<br>Eigenvalue ratio: ", round(dim_ratio_12, 2),
    "<br>PA unidimensional: ", dim_unidimensional
  ))

p <- ggplot(join_dim_df, aes(x = pmin(dim_ratio_12, 15), y = strength_a_cor, text = tooltip)) +
  geom_point(colour = irw_blue, size = 2, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = irw_red, linewidth = 0.8) +
  scale_x_log10() +
  labs(x = "Eigenvalue ratio (dimensionality vignette, log scale, capped at 15)",
       y = "Strength-discrimination correlation")

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: dim-join-inline
#| echo: false

dim_join_full <- join_dim_df %>% filter(!is.na(dim_unidimensional))
n_dim_join    <- nrow(join_dim_df)
cor_dim_join  <- cor(join_dim_df$strength_a_cor, log(join_dim_df$dim_ratio_12))
n_flagged_uni   <- sum(dim_join_full$dim_unidimensional)
n_flagged_multi <- sum(!dim_join_full$dim_unidimensional)
med_cor_uni     <- median(dim_join_full$strength_a_cor[dim_join_full$dim_unidimensional])
med_cor_multi   <- median(dim_join_full$strength_a_cor[!dim_join_full$dim_unidimensional])
wilcox_p_dim    <- wilcox.test(
  dim_join_full$strength_a_cor[dim_join_full$dim_unidimensional],
  dim_join_full$strength_a_cor[!dim_join_full$dim_unidimensional]
)$p.value
```

Across the `r n_dim_join` tables with a match in both this vignette and the
[dimensionality vignette](dimensionality.qmd), the two diagnostics move
together weakly but reliably (r = `r round(cor_dim_join, 2)` between
strength-discrimination correlation and log eigenvalue ratio) — real
signal in the direction the theory predicts, not a tight relationship.
The theory's prediction shows up more clearly when split by the
dimensionality vignette's own parallel-analysis call: the
`r n_flagged_uni` tables it flags as unidimensional have a median
strength-discrimination correlation of `r round(med_cor_uni, 2)`, versus
`r round(med_cor_multi, 2)` for the `r n_flagged_multi` tables it flags as
multidimensional (Wilcoxon `r if (wilcox_p_dim < 0.001) "p < 0.001" else paste0("p = ", round(wilcox_p_dim, 3))`).
Parallel analysis is strict — only a small fraction of tables clear its
bar for "unidimensional" — but among the ones that do, network centrality
and IRT discrimination agree noticeably better.

### Cross-vignette payoff 2: do locally dependent tables look denser?

```{r}
#| label: fig-density-vs-localdep
#| fig-cap: "Network density against the local-dependence vignette's proportion of Q3-flagged item pairs. A testlet or shared-stimulus effect should show up as extra edges concentrated among the affected items -- more flagged pairs, denser estimated networks -- rather than the theory's predicted uniform density. Hover for table details."
#| fig-height: 5

join_ld_df <- summary_df %>%
  filter(!is.na(network_density), !is.na(ld_prop_flagged)) %>%
  mutate(tooltip = paste0(
    "Table: ", table,
    "<br>Network density: ", round(network_density, 2),
    "<br>Prop. Q3-flagged: ", round(ld_prop_flagged, 2)
  ))

p <- ggplot(join_ld_df, aes(x = ld_prop_flagged, y = network_density, text = tooltip)) +
  geom_point(colour = irw_blue, size = 2, alpha = 0.7) +
  labs(x = "Proportion of pairs Q3-flagged (local-dependence vignette)",
       y = "Network density")

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: ld-join-inline
#| echo: false

n_ld_join    <- nrow(join_ld_df)
cor_ld_join  <- cor(join_ld_df$network_density, join_ld_df$ld_prop_flagged)
cor_test_ld  <- cor.test(join_ld_df$network_density, join_ld_df$ld_prop_flagged)
```

Across the `r n_ld_join` tables with a match in both this vignette and the
[local-dependence vignette](local_dependence.qmd), network density and the
proportion of Q3-flagged pairs correlate at r = `r round(cor_ld_join, 2)`
(`r if (cor_test_ld$p.value < 0.001) "p < 0.001" else paste0("p = ", round(cor_test_ld$p.value, 3))`)
— the expected sign (more flagged pairs, denser network), and with
`r n_ld_join` tables behind it, a reliable one. Density alone doesn't distinguish a uniformly
dense network from one with a few tightly-connected clusters sitting
inside an otherwise sparse graph — the latter is the more specific
testlet signature, and it's exactly what the local-dependence vignette's
[Q3 heatmaps](local_dependence.qmd#a-worst-case-and-a-clean-case-side-by-side)
are built to catch directly. Detecting that cluster structure from the
network side properly needs community-detection tools this pass didn't
run at batch scale (see Limitations); density is the coarser proxy
available here.

## How much statistical evidence actually supports these edges?

```{r}
#| label: bayes-setup-inline
#| echo: false

n_bayes_matched <- sum(summary_df$has_bayes_match, na.rm = TRUE)
n_mrf_matched   <- sum(summary_df$has_mrf_match, na.rm = TRUE)

evidence_levels <- c("strong_presence", "weak_presence", "inconclusive",
                     "weak_absence", "strong_absence")
evidence_labels <- c("Strong presence", "Weak presence", "Inconclusive",
                     "Weak absence", "Strong absence")

irw_overall_ggm <- if (nrow(edges_ggm_df) > 0) {
  table(factor(edges_ggm_df$evidence_category_ggm, levels = evidence_levels)) /
    sum(!is.na(edges_ggm_df$evidence_category_ggm))
} else rep(NA_real_, 5)
names(irw_overall_ggm) <- evidence_labels

irw_inconclusive <- unname(irw_overall_ggm["Inconclusive"])
irw_weak         <- unname(irw_overall_ggm["Weak presence"] + irw_overall_ggm["Weak absence"])
irw_strong       <- unname(irw_overall_ggm["Strong presence"] + irw_overall_ggm["Strong absence"])
```

The paper's headline numbers, across the 293 published networks they
reanalyzed: 31.7% of edges had *inconclusive* evidence, 49.7% had only
*weak* evidence either way (4.4% weak presence + 45.2% weak absence), and
just 18.7% had *strong* evidence (13.0% strong presence + 5.7% strong
absence). Most published edges, in other words, are asserted with far
less statistical support than a single regularized network graph implies.

Using Option A (GGM via BGGM, forced uniformly, matching their method
exactly), this pass produced usable Bayesian edge evidence for
`r n_bayes_matched` of `r nrow(summary_df)` tables. Pooling every edge
across every table (not averaging per-table percentages, matching how the
paper itself pools across networks for its headline numbers):

```{r}
#| label: fig-evidence-breakdown
#| fig-cap: "IRW's overall Bayesian edge-evidence breakdown (this pass, pooled across all tables) against Huth et al.'s (2026) published benchmark. Each bar is one evidence category from the same five-way classification."
#| fig-height: 4

breakdown_df <- bind_rows(
  tibble(source = "IRW (this pass)", category = evidence_labels,
         proportion = as.numeric(irw_overall_ggm)),
  tibble(source = "Huth et al. (2026)", category = evidence_labels,
         proportion = c(0.130, 0.044, 0.317, 0.452, 0.057))
) %>%
  mutate(category = factor(category, levels = evidence_labels))

p <- ggplot(breakdown_df, aes(x = category, y = proportion, fill = source)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("IRW (this pass)" = irw_blue, "Huth et al. (2026)" = irw_grey)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion of edges", fill = NULL) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

ggplotly(p) |> layout(margin = list(t = 60), legend = list(orientation = "h", y = -0.2))
```

IRW's item-network data lands `r if (!is.na(irw_strong) && irw_strong > 0.187) "better-supported than" else if (!is.na(irw_strong) && irw_strong < 0.187) "less well-supported than" else "comparably supported to"`
the published literature the paper reanalyzed: `r scales::percent(irw_inconclusive, accuracy = 0.1)`
inconclusive (vs. their 31.7%), `r scales::percent(irw_weak, accuracy = 0.1)` weak either way
(vs. their 49.7%), and `r scales::percent(irw_strong, accuracy = 0.1)` strong either way (vs.
their 18.7%). Read this comparison with real caution: the paper's 293
networks span whatever sample sizes and item counts happened to appear in
the published literature they sampled, while this pass's 5-30-item,
300+-participant filter (see Data and methods above) is an IRW-specific
compute-budget choice, not a matched comparison design -- the gap above
is suggestive, not a controlled replication.

### Does evidence strength track relative sample size?

Huth et al.'s central moderator isn't raw sample size — it's **relative**
sample size, observations per possible edge: N / (p(p-1)/2). Their Fig. 3
bins networks into relative-sample-size groups and shows the evidence
breakdown shift toward stronger evidence as relative sample size rises.

```{r}
#| label: fig-relative-n-bins
#| fig-cap: "Evidence-category breakdown by relative sample size bin (quartiles of n_participants / (n_items * (n_items - 1) / 2) across tables with usable Bayesian edge evidence), pooling edges within each bin. If IRW shows the same pattern Huth et al. found, strong presence/absence should grow and inconclusive should shrink from left to right."
#| fig-height: 5

rel_n_bins_df <- summary_df %>%
  filter(has_bayes_match, !is.na(relative_n_ggm)) %>%
  mutate(rel_n_bin = ntile(relative_n_ggm, 4)) %>%
  select(table, rel_n_bin, relative_n_ggm)

bin_ranges <- rel_n_bins_df %>%
  group_by(rel_n_bin) %>%
  summarise(lo = min(relative_n_ggm), hi = max(relative_n_ggm), .groups = "drop") %>%
  mutate(label = paste0("Q", rel_n_bin, "\n[", round(lo, 1), "-", round(hi, 1), "]"))

bin_breakdown_df <- edges_ggm_df %>%
  inner_join(rel_n_bins_df, by = "table") %>%
  filter(!is.na(evidence_category_ggm)) %>%
  count(rel_n_bin, evidence_category_ggm) %>%
  group_by(rel_n_bin) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup() %>%
  left_join(bin_ranges, by = "rel_n_bin") %>%
  mutate(
    evidence_category_ggm = factor(evidence_category_ggm, levels = evidence_levels, labels = evidence_labels),
    label = factor(label, levels = bin_ranges$label)
  )

p <- ggplot(bin_breakdown_df, aes(x = label, y = proportion, fill = evidence_category_ggm)) +
  geom_col(position = "stack") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = setNames(
    c(irw_blue, "#67a9cf", irw_grey, "#f4a582", irw_red), evidence_labels
  )) +
  labs(x = "Relative sample size quartile (N / possible edges)", y = "Proportion of edges", fill = NULL)

ggplotly(p) |> layout(margin = list(t = 40), legend = list(orientation = "h", y = -0.3))
```

```{r}
#| label: relative-n-inline
#| echo: false

strong_by_bin <- bin_breakdown_df %>%
  filter(evidence_category_ggm %in% c("Strong presence", "Strong absence")) %>%
  group_by(rel_n_bin) %>%
  summarise(prop_strong = sum(proportion), .groups = "drop") %>%
  arrange(rel_n_bin)
monotonic_strong <- all(diff(strong_by_bin$prop_strong) >= -0.02)  # small tolerance for noise
```

`r if (monotonic_strong) "IRW shows essentially the same monotonic pattern Huth et al. found: strong evidence (presence or absence combined) rises from the lowest to the highest relative-sample-size quartile." else "IRW does not cleanly reproduce the paper's monotonic pattern here -- strong evidence does not rise monotonically across relative-sample-size quartiles in this pass, even though the same relative-sample-size definition and binning approach were used."`
Strong evidence (presence + absence) by quartile:
`r paste(scales::percent(strong_by_bin$prop_strong, accuracy = 1), collapse = " -> ")`
(lowest to highest relative sample size).

### Does edge weight predict evidence strength?

The paper's Fig. 4 found no sharp relationship between an edge's partial
correlation magnitude and its evidence strength (BF10) — similar-sized
partial correlations landed in every evidence category — though the
correspondence was cleaner in their higher-relative-sample-size networks.

```{r}
#| label: fig-pcor-vs-bf
#| fig-cap: "Partial correlation magnitude against log10(BF10) for every edge with usable evidence (Option A/GGM), colored by evidence category, split by whether the edge's table falls above or below the median relative sample size. A clean relationship would show points aligning along a smooth curve rather than every category scattered across the same pcor range."
#| fig-height: 6

median_rel_n <- median(summary_df$relative_n_ggm[summary_df$has_bayes_match], na.rm = TRUE)

pcor_bf_df <- edges_ggm_df %>%
  inner_join(summary_df %>% select(table, relative_n_ggm), by = "table") %>%
  filter(!is.na(evidence_category_ggm), !is.na(relative_n_ggm)) %>%
  mutate(
    rel_n_group = ifelse(relative_n_ggm >= median_rel_n,
                          "Above-median relative sample size", "Below-median relative sample size"),
    evidence_category_ggm = factor(evidence_category_ggm, levels = evidence_levels, labels = evidence_labels),
    log10_bf = log10(pmin(pmax(BF10_ggm, 1e-10), 1e10))  # cap for display; classification itself is unaffected
  )

p <- ggplot(pcor_bf_df, aes(x = pcor_ggm, y = log10_bf, colour = evidence_category_ggm)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_hline(yintercept = c(-1, log10(1/3), log10(3), 1), linetype = "dashed", colour = "grey70") +
  scale_colour_manual(values = setNames(
    c(irw_blue, "#67a9cf", irw_grey, "#f4a582", irw_red), evidence_labels
  )) +
  facet_wrap(~ rel_n_group, ncol = 1) +
  labs(x = "Partial correlation (GGM)", y = expression(log[10](BF[10])), colour = NULL)

ggplotly(p, tooltip = c("x", "y", "colour")) |> layout(margin = list(t = 40))
```

```{r}
#| label: pcor-bf-inline
#| echo: false

cor_pcor_bf_all <- cor(abs(pcor_bf_df$pcor_ggm), pcor_bf_df$log10_bf, use = "complete.obs")
cor_pcor_bf_hi  <- with(filter(pcor_bf_df, rel_n_group == "Above-median relative sample size"),
                         cor(abs(pcor_ggm), log10_bf, use = "complete.obs"))
cor_pcor_bf_lo  <- with(filter(pcor_bf_df, rel_n_group == "Below-median relative sample size"),
                         cor(abs(pcor_ggm), log10_bf, use = "complete.obs"))
```

Overall, `|`partial correlation`|` and log(BF10) correlate at r = `r round(cor_pcor_bf_all, 2)`
across all edges — real but far from a tight relationship, matching the
paper's "no sharp relationship" finding. Splitting by relative sample size:
r = `r round(cor_pcor_bf_hi, 2)` in above-median-relative-sample-size
tables versus r = `r round(cor_pcor_bf_lo, 2)` below the median.
`r if (!is.na(cor_pcor_bf_hi) && !is.na(cor_pcor_bf_lo) && cor_pcor_bf_hi > cor_pcor_bf_lo) "The correspondence is cleaner in higher-relative-sample-size tables here too, matching what Huth et al. found in their own data." else "This pass does not show the same higher-relative-sample-size cleanup Huth et al. found -- the relationship isn't obviously tighter in IRW's better-powered tables."`

```{r}
#| label: mrf-agreement-inline
#| echo: false
#| output: asis

if (nrow(edges_mrf_df) > 0) {
  agree_df <- edges_ggm_df %>%
    filter(table %in% option_b_tables) %>%
    inner_join(edges_mrf_df, by = c("table", "item_i", "item_j")) %>%
    filter(!is.na(evidence_category_ggm), !is.na(evidence_category_mrf))
  n_agree_edges <- nrow(agree_df)
  pct_exact_agree <- mean(agree_df$evidence_category_ggm == agree_df$evidence_category_mrf)
  pct_direction_agree <- mean(
    grepl("presence", agree_df$evidence_category_ggm) == grepl("presence", agree_df$evidence_category_mrf) |
    agree_df$evidence_category_ggm == "inconclusive" | agree_df$evidence_category_mrf == "inconclusive"
  )

  cat("### Does the ordinal MRF agree with the GGM?\n\n")
  cat("Huth et al. acknowledge the ordinal Markov random field (fit via `bgms`) as ",
      "conceptually better-suited to binary/ordinal data than the GGM they actually used. ",
      "Testing that acknowledgment directly, on the ", length(option_b_tables),
      "-table Option B subset (", n_agree_edges, " edges with usable evidence under both models): ",
      "the two models put the same edge in the *exact same* one of five evidence categories ",
      scales::percent(pct_exact_agree, accuracy = 1), " of the time, and agree on ",
      "presence-vs-absence direction (treating inconclusive calls under either model as a pass) ",
      scales::percent(pct_direction_agree, accuracy = 1), " of the time.\n\n",
      if (pct_exact_agree < 0.5) {
        "That's a real disagreement rate, not just estimation noise -- for a meaningful share of edges, which model IRW's own data says is theoretically more appropriate changes the conclusion about whether that edge is real. Given the paper's own acknowledgment that GGM is a simplifying approximation for this data type, the exact-category numbers above should be read as a lower bound on how much Option A's uniform-GGM choice matters for binary/ordinal data specifically, not just a robustness footnote."
      } else {
        "The two models mostly agree, which is some reassurance that Option A's uniform-GGM simplification isn't badly distorting the picture for IRW's binary/ordinal tables -- though this check only covers a small illustrative subset, not the full candidate pool."
      }, "\n", sep = "")
} else {
  cat("### Does the ordinal MRF agree with the GGM?\n\n",
      "Option B (ordinal MRF) results were not available in this cache -- this comparison ",
      "requires re-running `network_psych_compute.R` with the current amendment.\n", sep = "")
}
```

### Do the cross-vignette checks look cleaner using only strong-evidence edges?

Both cross-vignette comparisons above used the frequentist regularized
network's edges, kept or dropped by LASSO shrinkage. Here, instead,
**node strength is rebuilt using only edges the Bayesian model calls
strong evidence for** (BF10 > 10) — a stricter, statistically-justified
filter rather than a regularization penalty — to see whether the two
cross-vignette relationships get cleaner once weakly-supported edges are
dropped entirely rather than merely down-weighted.

```{r}
#| label: fig-strong-only-dim
#| fig-cap: "Strength-discrimination correlation computed from strong-evidence-only edges, against the dimensionality vignette's eigenvalue ratio (compare to the all-edges version earlier on this page). Node strength here sums |partial correlation| over only that node's strong-presence edges; a node with none gets strength 0."
#| fig-height: 5

strong_edges_long <- edges_ggm_df %>%
  filter(evidence_category_ggm == "strong_presence") %>%
  { bind_rows(
      transmute(., table, item = item_i, w = abs(pcor_ggm)),
      transmute(., table, item = item_j, w = abs(pcor_ggm))
    ) } %>%
  group_by(table, item) %>%
  summarise(strength = sum(w), .groups = "drop")

strong_strength_cor_df <- map_dfr(intersect(unique(edges_ggm_df$table), names(discriminations_list)), function(tbl) {
  a <- discriminations_list[[tbl]]
  if (is.null(a) || length(a) < 2) return(NULL)
  items <- names(a)
  strength <- setNames(rep(0, length(items)), items)
  tbl_strong <- strong_edges_long[strong_edges_long$table == tbl, ]
  if (nrow(tbl_strong) > 0) strength[tbl_strong$item] <- tbl_strong$strength
  if (length(unique(strength)) < 2) return(NULL)
  tibble(table = tbl, strong_strength_a_cor = suppressWarnings(cor(strength, a[items])))
})

join_dim_strong_df <- strong_strength_cor_df %>%
  inner_join(summary_df %>% select(table, dim_ratio_12, dim_unidimensional), by = "table") %>%
  filter(!is.na(strong_strength_a_cor), !is.na(dim_ratio_12))

p <- ggplot(join_dim_strong_df, aes(x = pmin(dim_ratio_12, 15), y = strong_strength_a_cor)) +
  geom_point(colour = irw_blue, size = 2, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = irw_red, linewidth = 0.8) +
  scale_x_log10() +
  labs(x = "Eigenvalue ratio (dimensionality vignette, log scale, capped at 15)",
       y = "Strong-evidence-only strength-discrimination correlation")

ggplotly(p) |> layout(margin = list(t = 40))
```

```{r}
#| label: strong-dim-inline
#| echo: false

n_strong_dim_join <- nrow(join_dim_strong_df)
med_all_cor    <- median(summary_df$strength_a_cor, na.rm = TRUE)
med_strong_cor <- median(join_dim_strong_df$strong_strength_a_cor, na.rm = TRUE)

strong_ld_df <- summary_df %>%
  filter(has_bayes_match, !is.na(prop_strong_presence_ggm), !is.na(ld_prop_flagged))
cor_strong_ld <- cor(strong_ld_df$prop_strong_presence_ggm, strong_ld_df$ld_prop_flagged)
cor_all_ld    <- cor(join_ld_df$network_density, join_ld_df$ld_prop_flagged)
```

Across `r n_strong_dim_join` tables, the strong-evidence-only
strength-discrimination correlation has a median of `r round(med_strong_cor, 2)`,
versus `r round(med_all_cor, 2)` for the all-edges version earlier on this
page.
`r if (!is.na(med_strong_cor) && !is.na(med_all_cor) && med_strong_cor > med_all_cor) "Restricting to strong-evidence edges sharpens the correspondence with IRT discrimination -- consistent with the idea that the earlier relationship was partly diluted by weakly-supported edges the regularized network kept anyway." else "Restricting to strong-evidence edges doesn't obviously sharpen the correspondence with IRT discrimination here -- the earlier relationship doesn't appear to be primarily a weak-edge dilution effect."`

For the local-dependence check, the proportion of an table's edges
classified strong presence correlates with the proportion of Q3-flagged
pairs at r = `r round(cor_strong_ld, 2)`, versus r = `r round(cor_all_ld, 2)`
for network density against the same Q3-flagged proportion earlier.
`r if (!is.na(cor_strong_ld) && !is.na(cor_all_ld) && abs(cor_strong_ld) > abs(cor_all_ld)) "The testlet signal is somewhat cleaner using strong-evidence-only edges than using raw regularized-network density." else "The testlet signal isn't obviously cleaner using strong-evidence-only edges than using raw regularized-network density -- density from LASSO-regularized edges and density from Bayesian strong-evidence edges appear to carry similar information here."`

## Practical implications

Where the strength-discrimination correlation is high and the network is
dense and uniform, network psychometrics and IRT are, practically
speaking, telling the same story about the same items — using either
framework to identify the "best" items should agree. Where it breaks
down, that's a signal worth taking seriously in its own right, independent
of which framework you prefer: it means the two ways of describing item
covariation actually disagree about which items matter most, and a table
flagged as multidimensional or locally dependent elsewhere on this site
is a reasonable place to expect exactly that disagreement.

## Limitations

- **EBICglasso on ordinal data, two gaps found and one fixed by
  refitting.** `bootnet::estimateNetwork()`'s `EBICglasso` preset does not
  call `qgraph::cor_auto()` by default — as of `bootnet` 1.4, the
  `EBICglasso` default set switched from `corMethod = "cor_auto"` to
  plain Pearson correlations (`cor`). Neither `bootnet` nor `qgraph` was
  pinned in `renv.lock` at the time (now fixed — see below), so an earlier
  pass of this vignette likely ran on whatever version was on the machine
  at the time and may have silently treated 7-point Likert integers as
  literally continuous. This pass instead passes `corMethod = "cor_auto"`
  explicitly (polychoric/polyserial correlations for ordinal items) with
  `corArgs = list(forcePD = TRUE)`, since a polychoric correlation matrix
  over several ordinal items is not always positive-definite and
  `bootnet` errors out by default rather than correcting it; `forcePD`
  projects to the nearest valid correlation matrix instead. The impact was
  real and sometimes large: `clifford_2018_police_blame`, the lowest
  strength-discrimination correlation table in an earlier pass of this
  vignette (r = −0.69), moved to r = 0.09 once the network side used
  polychoric correlations and the IRT side used GPCM instead of GRM (next
  point) — a different table is now the lowest-correlation example (see
  the worked-case callout above). Separately, and regardless of that fix:
  even an exact polychoric-correlation Gaussian graphical model is still
  not the network model with a proven equivalence to an ordinal IRT
  model. Marsman, van den Bergh, & Haslbeck (2025) [@marsman2025bayesian]
  derive that equivalence for the **ordinal Markov random field**, and
  show it corresponds to the **generalized partial credit model (GPCM)**,
  not the graded response model most applied network-IRT comparisons use
  for polytomous items — this vignette now fits GPCM for that reason (see
  Data and methods), but an exact ordinal-MRF network fit (rather than a
  Gaussian approximation on polychoric correlations) is still not
  implemented here.
- **Network estimation is more sample- and tuning-sensitive** than fitting
  a single IRT model. Small tables in this sample should be read
  cautiously; an empty regularized network (no edges survive
  regularization) is itself informative but excludes that table from the
  strength-discrimination comparison entirely.
- **No bootstrap stability checks.** `bootnet`'s case-dropping bootstrap
  (the standard way to check whether an estimated network's edges and
  centrality ordering are stable) was not run at full-batch scale for
  compute reasons. A natural follow-up: run it for a handful of the most
  interesting example tables from the full batch (the strongest
  confirmation, the sharpest breakdown) rather than all of them.
- **Density is a coarse proxy for "clustered."** As noted above, telling
  a uniformly dense network apart from one with tight sub-clusters needs
  community-detection tools this pass didn't run.
- **Cache overlap is partial.** The dimensionality and local-dependence
  caches were built on their own candidate pools (different item-count
  caps), so not every table here has a match in both — see the match-rate
  numbers in Data and methods.
- **`bootnet` and `qgraph` were not pinned in `renv.lock` when this pass
  was run.** This pass ran against whatever versions of these two packages
  happened to be installed, which is how the `cor_auto` default-behavior
  change above went unnoticed. Both are now pinned in `renv.lock`, which
  prevents a silent version drift like this going forward but doesn't
  retroactively confirm which version produced the results above.
- **Matrix-F prior SD was not sensitivity-tested at IRW scale.** Both
  Option A and Option B above use the package default (`prior_sd = 0.25`),
  matching Huth et al.'s own default. They themselves ran a
  prior-sensitivity check across three SD values on their own data; a
  natural follow-up here would be the same check on a handful of IRW
  tables, not run in this pass for compute reasons.
- **Option A's uniform-GGM treatment of binary/ordinal data is an
  approximation this vignette inherited deliberately, not a claim it's the
  ideal model for IRW's data.** It's the same simplifying choice Huth et
  al. made in their own reanalysis, adopted here specifically so the
  headline-number comparison above is apples-to-apples with their
  benchmark. The Option A/B agreement check above is the direct test of
  how much that approximation costs for IRW's mostly-binary/ordinal data;
  where the two disagree, Option B (the ordinal MRF) is the theoretically
  better-suited model, and Option A should be read as an approximation to
  it, not the ground truth.
- **Option B ran on a small illustrative subset, not the full candidate
  pool.** Piloting found `bgms`'s ordinal MRF sampler costs roughly
  20-50x Option A's BGGM runtime per table; a full batch would take
  7-27 hours for compute that a full Option A batch does in well under
  two. The `r length(option_b_tables)`-table subset is enough to test
  whether the two models systematically disagree, but any specific
  IRW table not in that subset has only the Option A (GGM) result
  available.

## Reproducibility

```{r}
#| label: repro
#| echo: true
#| eval: false

# 1. Re-run the prerequisite compute scripts first (from the project root)
Rscript vignettes/dimensionality_compute.R
Rscript vignettes/local_dependence_compute.R

# 2. Re-run this vignette's compute script
Rscript vignettes/network_psych_compute.R

# 3. Re-render this page
quarto render vignettes/network_psych.qmd
```

Session info from the compute run: R `r if (cache_ok) .cache$session$R.version$version.string else "N/A"`,
run on `r format(date_run, "%B %d, %Y")`.

## References

::: {#refs}
:::
