---
title: "Network Psychometrics vs. IRT: Does Centrality Track Discrimination?"
description: |
  IRT explains why items correlate with a single latent trait; network
  psychometrics explains it with a graph of direct partial-correlation
  edges between items, no latent variable required. Under a strong single
  common cause, the two should agree: a network's node centrality should
  track IRT discrimination, and the network should look densely and
  uniformly connected. We test this across real IRW instruments and check
  whether the relationship tracks tables already flagged as
  multidimensional or locally dependent elsewhere on this site.
date: today
bibliography: network_psych_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(plotly)
library(qgraph)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "network_psych_data/network_psych_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache               <- readRDS(cache_path)
  summary_df           <- .cache$summary
  graphs_list          <- .cache$graphs %||% list()
  strengths_list       <- .cache$strengths %||% list()
  discriminations_list <- .cache$discriminations %||% list()
  edges_ggm_df         <- .cache$edges_ggm %||% tibble()
  edges_mrf_df         <- .cache$edges_mrf %||% tibble()
  option_b_tables      <- .cache$option_b_tables %||% character()
  candidate_tables     <- .cache$candidate_tables
  date_run             <- .cache$date_run
  is_pilot             <- isTRUE(.cache$pilot)
  n_all_candidates     <- .cache$n_all_candidates %||% length(candidate_tables)
  n_binary_all         <- .cache$n_binary_all %||% NA_integer_
  n_poly_all           <- .cache$n_poly_all %||% NA_integer_
  r_n_tables_run       <- length(candidate_tables)
} else {
  summary_df <- tibble(
    table = character(), n_items = integer(), n_participants = integer(),
    n_categories = integer(), network_method = character(), network_density = numeric(),
    strength_a_cor = numeric(), construct_type = character(), item_format = character(),
    dim_ratio_12 = numeric(), dim_nfact_suggested = integer(), dim_unidimensional = logical(),
    ld_prop_flagged = numeric(), ld_mean_abs_q3 = numeric(),
    has_dim_match = logical(), has_ld_match = logical(),
    has_bayes_match = logical(), has_mrf_match = logical(),
    n_used_ggm = integer(), relative_n_ggm = numeric(),
    prop_strong_presence_ggm = numeric(), prop_weak_presence_ggm = numeric(),
    prop_inconclusive_ggm = numeric(), prop_weak_absence_ggm = numeric(),
    prop_strong_absence_ggm = numeric(),
    n_used_mrf = integer(), relative_n_mrf = numeric(),
    prop_strong_presence_mrf = numeric(), prop_weak_presence_mrf = numeric(),
    prop_inconclusive_mrf = numeric(), prop_weak_absence_mrf = numeric(),
    prop_strong_absence_mrf = numeric()
  )
  graphs_list <- list(); strengths_list <- list(); discriminations_list <- list()
  edges_ggm_df <- tibble(); edges_mrf_df <- tibble(); option_b_tables <- character()
  candidate_tables <- character()
  date_run <- Sys.Date()
  is_pilot <- FALSE
  n_all_candidates <- 0L; n_binary_all <- NA_integer_; n_poly_all <- NA_integer_
  r_n_tables_run <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/network_psych_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, small pilot run.** This render uses a deliberately small ",
  "sample of ", r_n_tables_run, " tables (a mix of hand-picked known-good ",
  "tables and a random top-up), run live against Redivis, to validate the ",
  "pipeline and page layout before committing to a full batch across all ",
  n_all_candidates, " eligible tables. All numbers here are real computed ",
  "results -- not placeholders -- but treat the *distributions* as a small, ",
  "noisy preview rather than a representative picture. Re-run ",
  "`network_psych_compute.R` with `PILOT <- FALSE` to scale up.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`network_psych_compute.R`) directly if you're
relying on the numbers here.
:::

## Motivation

Every other vignette on this site treats item covariation the same way
IRT does: responses correlate because they all depend, more or less, on
one latent trait. **Network psychometrics** proposes a different account.
Instead of a shared cause behind the correlations, it models items as
nodes in a graph and estimates direct, partial-correlation edges between
them — an Ising model for binary items, a Gaussian graphical model
(EBICglasso) for polytomous ones. No latent variable is fit at all; the
graph *is* the model.

These two accounts aren't just philosophically different — they make a
shared, testable prediction. Van der Maas et al. (2006)
[@vandermaas2006dynamical] and Epskamp, Maris, van Borkulo, & Borsboom
(2018) [@epskamp2018testing] show that if a single dominant common cause
really does explain a set of items, the two descriptions should coincide.
This isn't just an informal observation: Marsman et al. (2018)
[@marsman2018introduction] prove a formal statistical equivalence between
the Ising model and the 2PL for binary items, and Marsman, van den Bergh,
& Haslbeck (2025) [@marsman2025bayesian] extend that equivalence to
ordinal items -- with a wrinkle noted in Data and methods below. Under
that equivalence, a node's **centrality** (how strongly and directly it connects to every
other node) should track that item's **IRT discrimination**, and the
estimated network should look densely and fairly uniformly connected —
not clustered into visibly separate sub-groups. This vignette tests that
prediction across real IRW instruments, and asks whether it holds up
better or worse for tables this site's other diagnostics already flag as
multidimensional ([dimensionality vignette](dimensionality.qmd)) or
locally dependent ([local-dependence vignette](local_dependence.qmd)).

## Data and methods

### Dataset selection

```{r}
#| label: filter-datasets
#| eval: false

binary_tables <- irw_filter(n_items = c(5, 30), n_participants = c(300, Inf),
                             n_categories = 2)
poly_tables   <- irw_filter(n_items = c(5, 30), n_participants = c(300, Inf),
                             n_categories = c(3, 7))
```

Items are capped at 30 (lower than the dimensionality vignette's 40) and
polytomous tables are restricted to 3-7 response categories, mainly for
compute reasons: EBIC-regularized network estimation scales worse than a
single IRT fit or eigen-decomposition. `r n_all_candidates` tables met
this criterion as of `r format(date_run, "%B %Y")` (`r n_binary_all`
binary, `r n_poly_all` polytomous).
`r if (is_pilot) "This page currently runs on a small pilot subset (see the warning above); a full pass analyzes all of them." else paste0("This pass analyzes all ", r_n_tables_run, " of them.")`

### Per-table computation

For each candidate table, `fit_network()`:

1. Fetches the table and builds a wide response matrix, downsampling to
   at most 10,000 respondents and dropping zero-variance items, mirroring
   the other IRT vignettes on this site.
2. Estimates a regularized network with `bootnet::estimateNetwork()`:
   `IsingFit` for binary items, `EBICglasso` for polytomous items. The
   polytomous correlations are computed via `qgraph::cor_auto()`
   (`corMethod = "cor_auto"`, `corArgs = list(forcePD = TRUE)`), which
   estimates polychoric correlations for ordinal data and projects to the
   nearest valid correlation matrix when the polychoric estimate isn't
   positive-definite — this still treats the *graph* estimation step as
   Gaussian, an approximation relative to an exact ordinal network model,
   but does not additionally treat Likert categories as literally
   continuous integers (see Limitations for why both of those still fall
   short of the ordinal Markov random field with a proven IRT
   equivalence).
3. Computes node centrality with `qgraph::centrality_auto()`. In a
   genuinely weighted network this returns a `Strength` column (each
   item's summed edge weight); a fully empty regularized network — every
   edge shrunk to zero — gets auto-detected as unweighted and falls back
   to `Degree` instead. That's still a real, informative result (kept in
   the data below), just not one with a meaningful strength-discrimination
   correlation, so that comparison is left `NA` for those tables. This is
   rare — 7 of the `r nrow(summary_df)` tables in this pass (about 1%):
   `erf_breuer_2017_frmmc`, `gilbert_meta_38`, `lsat`,
   `project_kids_wj_ak_grade` (all `IsingFit`/binary), and `Resistance`,
   `sun_2025_morality_study2_peoplerespect`,
   `sun_2025_morality_study2_peopletrust` (all `EBICglasso`/polytomous).
   Small enough in number, and split across both estimators, that it reads
   as ordinary regularization behavior on these specific items rather than
   a systematic estimation problem — but a table-by-table check of
   whether any of the seven point to a data or fitting issue specifically
   (rather than a genuinely null network) hasn't been done and would be
   worth doing.
4. Fits a one-factor IRT model and extracts item discrimination: 2PL for
   binary items; the **generalized partial credit model (GPCM)**, not the
   graded response model used elsewhere on this site, for polytomous items
   — GPCM is the model with a proven equivalence to the ordinal network
   model this vignette's polytomous branch estimates (see Limitations).
5. Records the correlation between node strength and discrimination
   across items — the vignette's core comparison — plus network density
   (proportion of possible edges present) and `construct_type`/
   `item_format` from IRW's tags.
6. Joins in, where the same table appears: the eigenvalue ratio /
   parallel-analysis outcome from the [dimensionality
   vignette](dimensionality.qmd)'s cache, and the proportion of Q3-flagged
   pairs from the [local-dependence vignette](local_dependence.qmd)'s
   cache — read directly from their saved results rather than
   recomputed.

```{r}
#| label: fit-function
#| eval: false

fit_network <- function(table_name) {
  resp <- fetch_wide(table_name)  # irw_fetch() + irw_long2resp(), cleaned
  net  <- if (n_categories == 2) {
    estimateNetwork(resp, default = "IsingFit")
  } else {
    estimateNetwork(resp, default = "EBICglasso", corMethod = "cor_auto")
  }
  cent <- centrality_auto(net$graph)
  fit  <- mirt(resp, 1, itemtype = if (n_categories == 2) "2PL" else "gpcm")
  a    <- coef(fit, simplify = TRUE)$items[, "a1"]
  cor(cent$node.centrality$Strength, a)  # the core comparison
}
```

Of the `r r_n_tables_run` candidate tables in this pass, `r nrow(summary_df)`
produced usable results (the rest were dropped for too few usable items,
an unsupported column type on fetch, a network/IRT model that failed to
fit, or too few remaining response categories). This pass matched a
dimensionality-vignette entry for `r sum(summary_df$has_dim_match)` of
those tables and a local-dependence-vignette entry for
`r sum(summary_df$has_ld_match)` — both caches were built on an
overlapping but not identical candidate pool, so coverage is high but not
total.

::: {.callout-note}
To reproduce with current IRW holdings, re-run `network_psych_compute.R`
(after the dimensionality and local-dependence compute scripts, which it
depends on) and commit the updated
`network_psych_data/network_psych_results.rds`.
:::

### Bayesian edge evidence: which model, and why

Everything above treats an edge as present or absent based on whether
regularized network estimation shrinks it to exactly zero — a single
point estimate, with no sense of how much statistical evidence actually
backs that call. Huth, Haslbeck, Keetelaar, van Holst, & Marsman (2026)
[@huth2026statistical] reanalyzed 293 published psychological networks
with a fully Bayesian alternative: fit each network as a Gaussian
graphical model (GGM) with a matrix-F prior on the partial correlations,
then compute an edge-inclusion Bayes factor (BF10) for every edge via the
Savage-Dickey density ratio. They found evidence was frequently much
weaker than a single regularized graph suggests — only 18.7% of edges in
the literature they reanalyzed had *strong* evidence either way.

Their software choice matters for how this section reproduces their
analysis. They fit **every** network in their reanalysis as a GGM via the
`BGGM` package [@williams2020bggm], accessed through the `easybgm` wrapper
[@huth2024easybgm] — uniformly, including networks built from binary or
ordinal items, explicitly as a simplifying choice for consistency across
a reanalysis spanning many published papers. They say directly that they
know this is an approximation for non-continuous data, and that the
ordinal Markov random field (fit via the `bgms` package) is the
conceptually better-suited model for that case — they didn't use it
because it was newly available and considerably slower to fit at the
time of their reanalysis (`bgms` has since gotten substantially faster),
and because it still isn't how ordinal data is typically fit in empirical
network-psychometrics practice.

IRW is overwhelmingly binary/ordinal item response data, which makes this
a real design decision here too, not just an aside:

- **Option A (primary, reported below):** fit every candidate table as a
  GGM via `easybgm(..., package = "BGGM")`, forced uniformly regardless of
  each table's actual `n_categories` — the same simplifying choice the
  paper made, for the same reason: it's the only way to compare IRW's
  breakdown directly against their headline percentages, and it also
  matches how ordinal network data is typically handled in current
  empirical practice, not just this paper's method. `package = "BGGM"` is
  forced explicitly and paired with `type = "continuous"`, so every table
  is fit by the same estimator regardless of its response format.
- **Option B (secondary, small subset only):** fit the ordinal Markov
  random field via `bgms` instead — the model the paper itself calls more
  appropriate for this data type. Piloting this on real IRW tables found
  it costs roughly 20-50x Option A's runtime per table (138-192 sec vs.
  1-10 sec on the same two tables); a full batch across IRW's ~660
  candidate tables would take 7-27 hours. It runs here on a small
  stratified subset (`r length(option_b_tables)` tables, split across
  binary and polytomous items) as an illustrative robustness check, not a
  full-batch analysis.

Option A uses the paper's own matrix-F prior default (`prior_sd = 0.25`)
for its GGM partial correlations; Option B's ordinal MRF (`bgms`) does not
use this prior at all — it has its own default prior structure for the
ordinal model, left at `easybgm`'s defaults rather than matched to
Option A's `prior_sd`. Both are classified into the paper's five-way
breakdown by BF10:
strong presence (BF10 > 10), weak presence (3 < BF10 <= 10), inconclusive
(1/3 <= BF10 <= 3), weak absence (1/10 <= BF10 < 1/3), or strong absence
(BF10 < 1/10) — a small deviation from the paper's stated open-interval
boundaries (which don't specify which side of e.g. BF10 = 3 owns that
exact value), noted here since it's a real, if practically inconsequential,
difference from a literal reading of their method.

## Results

### Two concrete examples, opposite ends of the relationship

```{r}
#| label: fig-example-network
#| fig-cap: "Estimated network (left) and IRT discrimination parameters (right) for the table with the strongest strength-discrimination correlation in this sample (top) and the weakest (bottom). Edge color shows sign: blue = positive partial correlation, red = negative. Top: a dense, uniformly-connected, uniformly positive graph with node positions that roughly track item discrimination -- exactly what Epskamp et al.'s theory predicts under a strong single factor. Bottom: edge thickness and node position barely track discrimination at all, and every edge is negative."
#| fig-height: 10
#| fig-width: 11

example_candidates <- summary_df %>%
  filter(!is.na(strength_a_cor), network_density > 0) %>%
  rowwise() %>%
  # Exclude numerically degenerate IRT fits -- e.g. a near-duplicate item
  # pair can drive discrimination to an implausible boundary value (a ~ 50),
  # which makes for a misleading rather than illustrative example.
  mutate(max_abs_a = max(abs(discriminations_list[[table]]), na.rm = TRUE)) %>%
  ungroup() %>%
  filter(max_abs_a < 5)
hi_table <- example_candidates %>% slice_max(strength_a_cor, n = 1, with_ties = FALSE) %>% pull(table)
lo_table <- example_candidates %>% slice_min(strength_a_cor, n = 1, with_ties = FALSE) %>% pull(table)

hi_cor <- summary_df$strength_a_cor[summary_df$table == hi_table]
lo_cor <- summary_df$strength_a_cor[summary_df$table == lo_table]

plot_example_pair <- function(table_name, label) {
  g <- graphs_list[[table_name]]
  s <- strengths_list[[table_name]]
  a <- discriminations_list[[table_name]]

  par(mar = c(4, 1, 4, 1))
  qgraph::qgraph(g, layout = "spring", labels = names(s),
                 label.cex = 0.9, vsize = 11, posCol = irw_blue, negCol = irw_red,
                 title = paste0(label, ": ", table_name, "\nestimated network"),
                 title.cex = 1)

  par(mar = c(4, 10, 4, 2))
  ord <- order(a)
  barplot(a[ord], names.arg = names(a)[ord], horiz = TRUE, las = 1,
          cex.names = 0.8, col = irw_blue, border = NA,
          xlab = "IRT discrimination (a)",
          main = paste0(label, ": ", table_name, "\ndiscrimination"), cex.main = 1)
}

par(mfrow = c(2, 2))
plot_example_pair(hi_table, "Highest correlation")
plot_example_pair(lo_table, "Lowest correlation")
```

```{r}
#| label: example-pair-inline
#| echo: false

hi_density <- summary_df$network_density[summary_df$table == hi_table]
lo_density <- summary_df$network_density[summary_df$table == lo_table]
lo_a       <- discriminations_list[[lo_table]]
n_lo_neg_a <- sum(lo_a < 0)
lo_g       <- graphs_list[[lo_table]]
hi_g       <- graphs_list[[hi_table]]
lo_all_neg <- all(lo_g[upper.tri(lo_g)] <= 0)

# network_density above is presence density (proportion of possible edges
# nonzero); these additionally weight by edge strength and track sign, to
# separate "how many edges survived" from "how strong/uniform they are"
edge_stats <- function(g) {
  ut <- g[upper.tri(g)]
  nz <- ut[ut != 0]
  list(weighted_density = sum(abs(ut)) / length(ut),
       mean_abs_weight  = mean(abs(nz)),
       prop_negative    = mean(nz < 0))
}
hi_edge_stats <- edge_stats(hi_g)
lo_edge_stats <- edge_stats(lo_g)
lo_irt_model <- if (summary_df$n_categories[summary_df$table == lo_table] == 2) {
  "2PL model"
} else {
  "generalized partial credit model"
}
```

`r hi_table` has a strength-discrimination correlation of `r round(hi_cor, 2)`
— about as clean a confirmation of the theory as this sample produces.
`r lo_table` sits at the opposite end, at `r round(lo_cor, 2)`: the two
descriptions of the same items agree much less about which ones matter
most, to the point of actively disagreeing about direction.

What makes the contrast worth dwelling on is what *doesn't* differ much
between them, at least by one measure: both networks have a comparable
*presence* density (proportion of possible edges nonzero: `r round(hi_density, 2)`
and `r round(lo_density, 2)` respectively). But presence density treats
every nonzero edge the same regardless of strength, and by that count
alone the two graphs look more alike than they actually are. Weighting by
edge strength instead (mean \|partial correlation\| over present edges,
and a weighted-density measure that also counts absent edges as zero)
tells a different story: `r hi_table` averages
`r round(hi_edge_stats$mean_abs_weight, 2)` per edge (weighted density
`r round(hi_edge_stats$weighted_density, 2)`), while `r lo_table` averages
`r round(lo_edge_stats$mean_abs_weight, 2)` (weighted density
`r round(lo_edge_stats$weighted_density, 2)`) — visibly the sparser,
weaker-signal network once strength is taken into account, even though
its raw edge count is not far off. The more distinctive difference,
though, is sign: every edge in `r hi_table` is positive, while
`r round(lo_edge_stats$prop_negative * 100, 0)`% of `r lo_table`'s nonzero
edges are negative. Density and uniformity, the other half of Epskamp et
al.'s prediction, don't cleanly distinguish these two cases on a
presence-only count; only the strength-discrimination correlation, and a
closer look at edge weight and sign, reveal that one table matches the
single-common-cause story and the other doesn't. A network that merely
*looks* dense on a presence-only count is not, on its own, evidence that
IRT and network psychometrics agree about the items — this pair is a
direct counterexample, and the discrepancy is smaller once weighted
density is used instead of presence density.

The disagreement in `r lo_table` isn't subtle, either:
`r n_lo_neg_a` of its `r length(lo_a)` items get a *negative* IRT
discrimination, meaning the `r lo_irt_model` estimates them as running
opposite the trait it fit from the rest.
`r if (lo_all_neg) "Every edge in the estimated network is negative (visible from the red edges above), not merely non-uniform in strength -- the two frameworks aren't just disagreeing on a matter of degree; the network sees a single tightly (negatively) coupled cluster where IRT sees a trait some items run backward on." else "The network still connects those items to every other about as strongly as any pair in hi_table, despite the sign disagreement in the IRT fit."`
That has a plausible practical upshot for anyone using either statistic
to pick or drop items — shortening a scale by discrimination or by
network centrality would not be a neutral choice for a table like this
one, since the two selections would part ways — though this vignette
only shows that the two statistics disagree here, not how much that
disagreement would actually move item-selection decisions in practice;
that's a natural follow-up, not something demonstrated above. Compare
both examples to the aggregate picture below: most tables fall somewhere
between these two extremes, but not always close to the middle.

```{r}
#| label: clifford-aside-inline
#| echo: false

clifford_table <- "clifford_2018_police_blame"
clifford_present <- clifford_table %in% summary_df$table
if (clifford_present) {
  clifford_cor <- summary_df$strength_a_cor[summary_df$table == clifford_table]
}
```

::: {.callout-note collapse="true"}
#### A worked case for *why* a single common cause can fail, not just that it does

`r if (clifford_present) paste0("One table worth a specific look regardless of where it happens to fall in the ranking above -- its strength-discrimination correlation this pass is ", round(clifford_cor, 2), " -- is ", clifford_table, ".") else paste0(clifford_table, " didn't appear among this pass's usable tables, but the underlying item design below is worth keeping in mind for similar multi-target blame/attribution items elsewhere in IRW.")` Each item asks the identically-worded "how much blame does X
deserve" about a different target in the 2018 Sacramento police shooting of
Stephon Clark -- state officials and the responding officers on one side,
and the victim himself (`item_blame_clark`) on the other. Blaming the
victim is a substantively different judgment from blaming officials, and
the raw item correlations bear that out: `item_blame_clark` correlates
negatively with every other item (fetched directly from IRW, not shown
here), while the official-blame items mostly correlate positively with
each other. That's a real bipolar attitude structure, not a data-entry or
reverse-coding artifact -- there's no oppositely-worded item to un-reverse;
every item shares the same "how much blame" wording, only the target
changes. It's a reminder that a low (or negative) strength-discrimination
correlation doesn't always mean the same thing: sometimes it's a
genuinely multidimensional or bipolar construct like this one, and
sometimes -- as with the current lowest-correlation example above -- the
network and the IRT model simply weight the same roughly-unidimensional
items differently. IRW tables do get an independent unidimensionality
check elsewhere on this site -- see [Cross-vignette payoff
1](#cross-vignette-payoff-1-does-it-track-the-dimensionality-diagnostic)
below, which splits this vignette's strength-discrimination correlation
by the [dimensionality vignette](dimensionality.qmd)'s own
parallel-analysis unidimensional/multidimensional call, rather than
relying on a single worked example.
:::

### Do negative edges predict a worse match, generally?

The `r lo_table` example above pairs a low strength-discrimination
correlation with a network that's substantially negative-edged. Is that
a coincidence of this one example, or does it hold across the sample?

```{r}
#| label: fig-negedge-vs-cor
#| fig-cap: "Strength-discrimination correlation against the proportion of each table's nonzero edges that are negative. One point per table with a usable comparison. If negative edges are a marker of a breakdown between the two frameworks (as in the low-correlation example above), points should trend downward left to right."
#| fig-height: 5

negedge_df <- summary_df %>%
  filter(!is.na(strength_a_cor)) %>%
  rowwise() %>%
  mutate(prop_neg_edges = {
    g <- graphs_list[[table]]
    ut <- g[upper.tri(g)]
    nz <- ut[ut != 0]
    if (length(nz) == 0) NA_real_ else mean(nz < 0)
  }) %>%
  ungroup() %>%
  filter(!is.na(prop_neg_edges)) %>%
  mutate(tooltip = paste0(
    "Table: ", table,
    "<br>Strength-a cor: ", round(strength_a_cor, 2),
    "<br>Prop. negative edges: ", round(prop_neg_edges, 2)
  ))

p <- ggplot(negedge_df, aes(x = prop_neg_edges, y = strength_a_cor, text = tooltip)) +
  geom_point(colour = irw_blue, size = 1.8, alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, colour = irw_red, linewidth = 0.8) +
  labs(x = "Proportion of nonzero edges that are negative",
       y = "Strength-discrimination correlation")

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: negedge-inline
#| echo: false

n_negedge <- nrow(negedge_df)
cor_negedge <- cor(negedge_df$prop_neg_edges, negedge_df$strength_a_cor)
cor_negedge_test <- cor.test(negedge_df$prop_neg_edges, negedge_df$strength_a_cor)
```

It isn't a coincidence. Across all `r n_negedge` tables with a usable
comparison, the proportion of negative edges correlates with the
strength-discrimination correlation at r = `r round(cor_negedge, 2)`
(`r if (cor_negedge_test$p.value < 0.001) "p < 0.001" else paste0("p = ", round(cor_negedge_test$p.value, 3))`)
— a strong, negative relationship. Networks with mostly-positive edges
tend to match IRT discrimination well; networks with a substantial share
of negative edges tend not to. That's consistent with the theory itself:
Epskamp et al.'s prediction is specifically for a single dominant common
cause producing a densely, uniformly, *positively* connected network, so
a network with many negative edges is close to a direct signal that the
single-common-cause premise doesn't hold for that table (a bipolar or
multidimensional structure like the `clifford_2018_police_blame` case
above), not an independent nuisance factor layered on top of the
strength-discrimination comparison.

### Is the negative-edge problem partly a sign-mismatch problem, not just a theory breakdown?

These are validated scales, so a negative IRT discrimination on a handful
of items is usually a sign that the item is reverse-keyed and wasn't
recoded before fitting, not evidence of a genuinely negative relationship
with the trait. **We don't have a reliable way to confirm that directly for
IRW as a whole: `irw_tags()` metadata is table-level (age range, sample,
construct type, measurement tool, item format, construct name), not
item-level, so there's no reverse-coding flag to check against.** What
follows is an empirical check on item-level sign agreement instead, not a
validated audit of which items are actually reverse-keyed.

```{r}
#| label: sign-mismatch-inline
#| echo: false

item_level <- purrr::map_dfr(names(discriminations_list), function(tbl) {
  a <- discriminations_list[[tbl]]
  g <- graphs_list[[tbl]]
  s <- strengths_list[[tbl]]
  if (is.null(a) || is.null(g) || length(a) < 2) return(NULL)
  items <- names(a)
  edge_sum <- rowSums(g[items, items, drop = FALSE])
  tibble(table = tbl, item = items, a = a, edge_sum = edge_sum, strength = s[items])
})

any_neg_strength <- any(item_level$strength < 0, na.rm = TRUE)
tab_sign <- table(a_neg = item_level$a < 0, edge_neg = item_level$edge_sum < 0)
prop_edgeneg_given_aneg  <- mean(item_level$edge_sum[item_level$a < 0]  < 0)
prop_edgeneg_given_apos  <- mean(item_level$edge_sum[item_level$a >= 0] < 0)
chisq_sign <- chisq.test(tab_sign)
n_a_neg <- sum(item_level$a < 0)
n_items_total <- nrow(item_level)

alt_cor_df <- summary_df %>%
  filter(!is.na(strength_a_cor)) %>%
  rowwise() %>%
  mutate(
    abs_a_cor = {
      a <- discriminations_list[[table]]
      s <- strengths_list[[table]]
      items <- names(a)
      s <- s[items]
      if (length(unique(s)) < 2) NA_real_ else suppressWarnings(cor(s, abs(a)))
    },
    n_neg_a = sum(discriminations_list[[table]] < 0)
  ) %>%
  ungroup()

has_neg_a <- alt_cor_df %>% filter(n_neg_a > 0, !is.na(abs_a_cor))
med_orig_all <- median(alt_cor_df$strength_a_cor, na.rm = TRUE)
med_abs_all  <- median(alt_cor_df$abs_a_cor, na.rm = TRUE)
med_orig_neg <- median(has_neg_a$strength_a_cor, na.rm = TRUE)
med_abs_neg  <- median(has_neg_a$abs_a_cor, na.rm = TRUE)
wilcox_sign  <- wilcox.test(has_neg_a$strength_a_cor, has_neg_a$abs_a_cor, paired = TRUE)
```

Node strength, as used throughout this vignette (and as `qgraph`/`bootnet`
compute it), is a sum of *absolute* edge weights — it can never be negative
(confirmed directly against all `r n_items_total` items here:
`r if (any_neg_strength) "some are negative, contrary to that claim" else "none are"`).
IRT discrimination is signed. That mismatch alone
means any item mirt fits with a negative discrimination is close to
guaranteed to drag the correlation down, independent of whether the
network "gets it right" for that item. Two things follow from that.

First, negative discrimination and negative edges aren't independent —
they concentrate on the same items far more than chance. Among the
`r n_a_neg` of `r n_items_total` items with a negative discrimination,
`r scales::percent(prop_edgeneg_given_aneg, accuracy = 1)` also have a net-negative
summed edge weight, versus only `r scales::percent(prop_edgeneg_given_apos, accuracy = 1)`
of items with non-negative discrimination
(`r if (chisq_sign$p.value < 0.001) "chi-squared p < 0.001" else paste0("chi-squared p = ", round(chisq_sign$p.value, 3))`).
That's consistent with both frameworks periodically picking up on the same
under-the-hood problem — most plausibly a reverse-keyed item — even though
neither framework is set up to flag it directly.

Second, a sign-invariant version of the same metric — correlating node
strength against `abs(a)` instead of signed `a` — does noticeably better
whenever a table has at least one negative-discrimination item. Across all
`r nrow(alt_cor_df)` tables the median moves modestly, from
`r round(med_orig_all, 2)` (signed) to `r round(med_abs_all, 2)` (absolute).
Restricted to the `r nrow(has_neg_a)` tables with at least one
negative-discrimination item, the gap is much larger — median
`r round(med_orig_neg, 2)` to `r round(med_abs_neg, 2)`
(`r if (wilcox_sign$p.value < 0.001) "paired Wilcoxon p < 0.001" else paste0("paired Wilcoxon p = ", round(wilcox_sign$p.value, 3))`).
For a meaningful share of tables, part of what looks like a
strength-discrimination *disagreement* is really a metric that penalizes
sign conventions the network side was never going to reproduce, on top of
whatever genuine disagreement remains. The rest of this vignette keeps the
signed correlation throughout for consistency and because it's the more
conservative (harder-to-pass) choice, not because it's the better-justified
metric for this specific question.

### Does strength track discrimination in general?

```{r}
#| label: fig-cor-distribution
#| fig-cap: "Distribution of the strength-discrimination correlation across all tables with a usable comparison. The boxplot shows the median and interquartile range; each point is one table, jittered vertically only to avoid overplotting -- read the plot along the x-axis. Values near 1 match the theoretical prediction under a strong single factor; low or negative values indicate a breakdown. Hover over a point to see which table it is."
#| fig-height: 3

dist_df <- summary_df %>%
  filter(!is.na(strength_a_cor)) %>%
  mutate(tooltip = paste0(
    "Table: ", table,
    "<br>Strength-a correlation: ", round(strength_a_cor, 2),
    "<br>Network method: ", network_method,
    "<br>Network density: ", round(network_density, 2),
    "<br>n_items: ", n_items
  ))

p <- ggplot(dist_df, aes(x = strength_a_cor, y = "", text = tooltip)) +
  geom_boxplot(colour = irw_grey, fill = NA, outlier.shape = NA, width = 0.5) +
  geom_jitter(colour = irw_blue, height = 0.15, size = 1.4, alpha = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  labs(x = "Strength-discrimination correlation", y = NULL)

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 40))
```

```{r}
#| label: fig-cor-vs-weighted-density
#| fig-cap: "Strength-discrimination correlation against each table's weighted network density -- mean absolute edge weight over all possible edges, counting absent edges as zero, so higher values mean a denser and/or more strongly-connected network. Unlike a plotting-only axis, this one carries real meaning: a theory that predicts a densely, uniformly connected network under a strong single factor also predicts higher weighted density should go with better strength-discrimination agreement. The two estimators are shown separately and on their own x scales because their edge weights are not in the same units -- EBICglasso weights are partial correlations, bounded by 1, while IsingFit weights are log-odds and unbounded, so pooling them onto one axis would compare quantities that are not comparable. One IsingFit table (eurpar2_mudfold, weighted density 14.1, more than ten times the next-largest value) is off-scale and omitted from the panel; the correlation reported for IsingFit excludes it. One point per table with a usable comparison. Hover for table details."
#| fig-height: 5

density_df <- summary_df %>%
  filter(!is.na(strength_a_cor)) %>%
  rowwise() %>%
  mutate(weighted_density = edge_stats(graphs_list[[table]])$weighted_density) %>%
  ungroup() %>%
  filter(!is.na(weighted_density)) %>%
  mutate(tooltip = paste0(
    "Table: ", table,
    "<br>Strength-a correlation: ", round(strength_a_cor, 2),
    "<br>Weighted density: ", round(weighted_density, 2),
    "<br>Network method: ", network_method,
    "<br>n_items: ", n_items
  ))

# Edge weights are not on a common scale across the two estimators (partial
# correlations vs. log-odds), so each is plotted on its own x axis. One
# IsingFit table sits an order of magnitude beyond the rest and is dropped
# from the panel so the bulk of the distribution is legible.
DENSITY_PLOT_CUTOFF <- 10
density_plot_df <- filter(density_df, weighted_density <= DENSITY_PLOT_CUTOFF)

p <- ggplot(density_plot_df, aes(x = weighted_density, y = strength_a_cor, text = tooltip)) +
  geom_point(colour = irw_blue, size = 1.6, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = irw_red, linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = irw_grey) +
  facet_wrap(~ network_method, scales = "free_x") +
  labs(x = "Weighted network density (mean |edge weight| over all possible edges)",
       y = "Strength-discrimination correlation")

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: cor-summary-inline
#| echo: false

med_cor <- median(summary_df$strength_a_cor, na.rm = TRUE)
prop_strong <- mean(summary_df$strength_a_cor > 0.7, na.rm = TRUE)
n_na_cor <- sum(is.na(summary_df$strength_a_cor))
EMPTY_TABLES <- summary_df$table[is.na(summary_df$strength_a_cor) & summary_df$network_density == 0]
# Correlations are computed within estimator, for the same reason the figure
# facets: a pooled correlation mixes partial correlations with log-odds.
# The IsingFit correlation is computed on the same points the panel shows,
# i.e. excluding the one off-scale table.
density_by_method <- density_plot_df %>%
  group_by(network_method) %>%
  summarise(
    n = n(),
    r = cor(weighted_density, strength_a_cor),
    p = cor.test(weighted_density, strength_a_cor)$p.value,
    .groups = "drop"
  )
fmt_p <- function(p) if (p < 0.001) "p < 0.001" else paste0("p = ", round(p, 3))
dens_ggm  <- filter(density_by_method, network_method == "EBICglasso")
dens_ising <- filter(density_by_method, network_method == "IsingFit")
```

The median strength-discrimination correlation across `r sum(!is.na(summary_df$strength_a_cor))`
tables with a usable comparison is `r round(med_cor, 2)`, and
`r scales::percent(prop_strong, accuracy = 1)` of tables exceed 0.7.
`r if (n_na_cor > 0) paste0(n_na_cor, " table(s) are excluded here because their regularized network had no edges at all -- an empty network, not a low correlation.") else ""`
The relationship is real and often strong, but — as the two comparisons
below make concrete — it is not universal. The density half of Epskamp
et al.'s prediction — that a strong single factor should produce a densely,
uniformly connected network *and* good agreement with discrimination, so the
two should move together — gets only weak support. Among the
`r dens_ggm$n` tables fit with EBICglasso, where edge weights are partial
correlations, weighted density correlates with the strength-discrimination
correlation at r = `r sprintf("%.2f", dens_ggm$r)` (`r fmt_p(dens_ggm$p)`): in the
predicted direction, but small enough that density explains barely more than
one percent of the variation in agreement. Among the `r dens_ising$n`
IsingFit tables it is r = `r sprintf("%.2f", dens_ising$r)`
(`r fmt_p(dens_ising$p)`) — no reliable relationship at all. The two
estimators are kept apart here because their edge weights aren't in the same
units: pooling them yields a single number that describes neither.

### Why are 7 networks empty? A closer look

The `r n_na_cor` tables excluded above for having no edges at all
(`r paste(EMPTY_TABLES, collapse = ", ")`) were audited directly: each was
re-fetched and re-fit outside the main batch to check whether the empty
result reproduces and, if so, what's driving it — a data or estimation
artifact (too few usable respondents, degenerate items, a numerically
unstable correlation matrix) would be a different finding than a real
absence of detectable partial-correlation structure.

```{r}
#| label: empty-audit-table
#| echo: false

empty_audit_path <- "network_psych_data/network_psych_empty_audit.rds"
empty_audit_ok <- file.exists(empty_audit_path)
if (empty_audit_ok) {
  empty_audit_df <- readRDS(empty_audit_path)
  knitr::kable(
    empty_audit_df %>%
      transmute(
        Table = table,
        `N (participants)` = n_participants_raw,
        `N items (kept)` = n_items_kept,
        `Prop. missing` = round(prop_missing, 2),
        `Prop. extreme marginals` = ifelse(is.na(extreme_marginal), "--", round(extreme_marginal, 2)),
        `Max |r|` = round(max_abs_cor, 2),
        `Mean |r|` = round(mean_abs_cor, 2),
        `Density (re-fit)` = density_reproduced
      ),
    caption = "Diagnostics for the 7 tables with an empty regularized network, re-fit independently of the main batch."
  )
}

gilbert_extreme_marginal <- empty_audit_df$extreme_marginal[empty_audit_df$table == "gilbert_meta_38"]
missing_tables <- c("sun_2025_morality_study2_peoplerespect",
                     "sun_2025_morality_study2_peopletrust", "project_kids_wj_ak_grade")
max_missing_prop <- max(empty_audit_df$prop_missing[empty_audit_df$table %in% missing_tables])
```

All 7 reproduced an empty network on an independent re-fit (`Density
(re-fit)` = 0 throughout) — this isn't a one-off fluke from the original
batch run. What they share is uniformly weak pairwise item correlations:
mean |r| sits between `r round(min(empty_audit_df$mean_abs_cor), 2)` and
`r round(max(empty_audit_df$mean_abs_cor), 2)` across the 7 tables, and even
the single strongest pairwise correlation in each table (`Max |r|`) tops out
around `r round(max(empty_audit_df$max_abs_cor), 2)`. EBIC/LASSO
regularization is designed to drop exactly this kind of weak, likely-noise
edge to control false positives — so an empty network here is the
regularization doing its job on genuinely weak signal, not an artifact of
the fitting pipeline. A couple of tables point to a plausible reason the
signal is weak to begin with: `gilbert_meta_38` has
`r scales::percent(gilbert_extreme_marginal, accuracy = 1)` of its binary
items at an extreme (under 5% or over 95%) endorsement rate, which caps how
much correlation those items can show with anything; the two
`sun_2025_morality_study2_*` tables and `project_kids_wj_ak_grade` have
substantial missingness (`r scales::percent(max_missing_prop, accuracy = 1)`
at the high end), which thins out the pairwise-complete data each
correlation is estimated on. Sample size (`N (participants)`,
348-1000) is modest but in line with plenty of tables elsewhere in this
sample that *do* get a usable network, so it reads as a contributing factor
alongside weak marginal correlations rather than the primary driver on its
own. None of this points to a data-pipeline bug — the empty result looks
like a genuine (if under-powered) null.

*Caveat: this audit re-fits each table once outside the main batch's
`furrr`-parallelized run; it confirms reproducibility under the same
`estimateNetwork()` call, not stability under resampling (no bootstrap was
run here — see Limitations).*

### Cross-vignette payoff 1: does it track the dimensionality diagnostic?

```{r}
#| label: fig-cor-vs-dimensionality
#| fig-cap: "Strength-discrimination correlation against the dimensionality vignette's eigenvalue ratio (log scale, capped at 15 for display). If Epskamp et al.'s theory holds, the correlation should sit near 1 for tables with a high (clearly unidimensional) ratio and scatter lower as the ratio approaches the vignette's ratio = 4 heuristic cutoff or below. Hover for table details."
#| fig-height: 5

join_dim_df <- summary_df %>%
  filter(!is.na(strength_a_cor), !is.na(dim_ratio_12)) %>%
  mutate(tooltip = paste0(
    "Table: ", table,
    "<br>Strength-a cor: ", round(strength_a_cor, 2),
    "<br>Eigenvalue ratio: ", round(dim_ratio_12, 2),
    "<br>PA unidimensional: ", dim_unidimensional
  ))

p <- ggplot(join_dim_df, aes(x = log10(pmin(dim_ratio_12, 15)), y = strength_a_cor, text = tooltip)) +
  geom_point(colour = irw_blue, size = 2, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = irw_red, linewidth = 0.8) +
  # xintercept in the same log10 units as the x aesthetic above -- ggplotly does not
  # reliably re-apply scale_x_log10()'s transform to geom_vline, so the axis is built
  # by hand here rather than trusting that transform to carry through to the shape.
  geom_vline(xintercept = log10(4), linetype = "dashed", colour = irw_grey) +
  scale_x_continuous(labels = function(x) round(10^x, 1)) +
  labs(x = "Eigenvalue ratio (dimensionality vignette, log scale, capped at 15; dashed line = ratio 4 unidimensionality heuristic)",
       y = "Strength-discrimination correlation")

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: dim-join-inline
#| echo: false

dim_join_full <- join_dim_df %>% filter(!is.na(dim_unidimensional))
n_dim_join    <- nrow(join_dim_df)
cor_dim_join  <- cor(join_dim_df$strength_a_cor, log(join_dim_df$dim_ratio_12))
n_flagged_uni   <- sum(dim_join_full$dim_unidimensional)
n_flagged_multi <- sum(!dim_join_full$dim_unidimensional)
med_cor_uni     <- median(dim_join_full$strength_a_cor[dim_join_full$dim_unidimensional])
med_cor_multi   <- median(dim_join_full$strength_a_cor[!dim_join_full$dim_unidimensional])
wilcox_p_dim    <- wilcox.test(
  dim_join_full$strength_a_cor[dim_join_full$dim_unidimensional],
  dim_join_full$strength_a_cor[!dim_join_full$dim_unidimensional]
)$p.value
```

Across the `r n_dim_join` tables with a match in both this vignette and the
[dimensionality vignette](dimensionality.qmd), the two diagnostics move
together weakly but reliably (r = `r round(cor_dim_join, 2)` between
strength-discrimination correlation and log eigenvalue ratio) — real
signal in the direction the theory predicts, not a tight relationship.
The theory's prediction shows up more clearly when split by the
dimensionality vignette's own parallel-analysis call: the
`r n_flagged_uni` tables it flags as unidimensional have a median
strength-discrimination correlation of `r round(med_cor_uni, 2)`, versus
`r round(med_cor_multi, 2)` for the `r n_flagged_multi` tables it flags as
multidimensional (Wilcoxon `r if (wilcox_p_dim < 0.001) "p < 0.001" else paste0("p = ", round(wilcox_p_dim, 3))`).
Parallel analysis is strict — only a small fraction of tables clear its
bar for "unidimensional" — but among the ones that do, network centrality
and IRT discrimination agree noticeably better.

### A network-native multidimensionality check: the stochastic block model

The comparison above uses the dimensionality vignette's eigenvalue-ratio
screen — a call made entirely from the IRT side, applied here as a proxy
for the network side. `bgms` includes a stochastic block model (SBM) test
— documented in the package as implementing the model from Sekulovski et
al. (2025), not independently verified against that paper here — that asks
the same question natively from inside
the network model itself: does the estimated edge structure look like one
tightly-connected cluster, or more than one? Following a reviewer
suggestion, this fits the SBM (`bgms::bgm(..., edge_prior =
"Stochastic-Block")`) on the same 20-table Option B stratified subset used
for the ordinal-MRF robustness check elsewhere on this page, and derives a
literal Bayes factor for one cluster vs. more than one (via Savage-Dickey
against `bgm()`'s own zero-truncated-Poisson prior on cluster count), to
compare against the strength-discrimination correlation the same way the
eigenvalue-ratio check above does.

```{r}
#| label: sbm-check-inline
#| echo: false

sbm_path <- "network_psych_data/network_psych_sbm_reliable.rds"
sbm_ok <- file.exists(sbm_path)
if (sbm_ok) {
  sbm_reliable <- readRDS(sbm_path)
  n_sbm_reliable <- nrow(sbm_reliable)
  max_bf_k1 <- max(sbm_reliable$bf_k1)
  cor_bf_strength <- cor(sbm_reliable$bf_k1, sbm_reliable$strength_a_cor)
  cor_p_strength  <- cor(sbm_reliable$post_p_k1, sbm_reliable$strength_a_cor)
}
```

`r if (sbm_ok) paste0("Of the 20 candidate tables, one failed to fetch (an unsupported data type), one had no strength-discrimination match, and four had to be dropped for a severe MCMC convergence failure -- three tables' chains diverged on 100% of post-warmup samples, one on 25% -- leaving ", n_sbm_reliable, " tables with a trustworthy SBM fit and a usable comparison.") else "SBM results not available."`
Two things came out of it. First, **no table clears the BF > 10 bar for
one-cluster structure** — the highest Bayes factor across all
`r if (sbm_ok) n_sbm_reliable else "--"` reliable tables is
`r if (sbm_ok) round(max_bf_k1, 2) else "--"`, well short of "strong
evidence" by the same BF10 convention used throughout this vignette. The
two-group comparison a reviewer originally suggested (compare
strength-discrimination correlation for BF>10-favoring-one-cluster tables
against the rest) isn't possible in this sample — there is no
one-cluster-favoring group. Second, the *continuous* version of the same
signal still points the right way: both the Bayes factor and the raw
posterior probability of one cluster correlate positively with the
strength-discrimination correlation
(r = `r if (sbm_ok) round(cor_bf_strength, 2) else "--"` and
r = `r if (sbm_ok) round(cor_p_strength, 2) else "--"` respectively) —
weak, on a sample this small, but in the direction the theory predicts, and
independent evidence from a completely different diagnostic (the network's
own cluster structure rather than an IRT-side eigenvalue ratio) pointing
the same way as the dimensionality cross-check above. Given the small,
divergence-thinned sample here, this reads as a plausible corroboration of
the eigenvalue-ratio result, not a stronger replacement for it.

### Cross-vignette payoff 2: do locally dependent tables look denser?

```{r}
#| label: fig-density-vs-localdep
#| fig-cap: "Network density against the local-dependence vignette's proportion of Q3-flagged item pairs. A testlet or shared-stimulus effect should show up as extra edges concentrated among the affected items -- more flagged pairs, denser estimated networks -- rather than the theory's predicted uniform density. Hover for table details."
#| fig-height: 5

join_ld_df <- summary_df %>%
  filter(!is.na(network_density), !is.na(ld_prop_flagged)) %>%
  mutate(tooltip = paste0(
    "Table: ", table,
    "<br>Network density: ", round(network_density, 2),
    "<br>Prop. Q3-flagged: ", round(ld_prop_flagged, 2)
  ))

p <- ggplot(join_ld_df, aes(x = ld_prop_flagged, y = network_density, text = tooltip)) +
  geom_point(colour = irw_blue, size = 2, alpha = 0.7) +
  labs(x = "Proportion of pairs Q3-flagged (local-dependence vignette)",
       y = "Network density")

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: ld-join-inline
#| echo: false

n_ld_join    <- nrow(join_ld_df)
cor_ld_join  <- cor(join_ld_df$network_density, join_ld_df$ld_prop_flagged)
cor_test_ld  <- cor.test(join_ld_df$network_density, join_ld_df$ld_prop_flagged)
```

Across the `r n_ld_join` tables with a match in both this vignette and the
[local-dependence vignette](local_dependence.qmd), network density and the
proportion of Q3-flagged pairs correlate at r = `r round(cor_ld_join, 2)`
(`r if (cor_test_ld$p.value < 0.001) "p < 0.001" else paste0("p = ", round(cor_test_ld$p.value, 3))`)
— the expected sign (more flagged pairs, denser network), and with
`r n_ld_join` tables behind it, a reliable one. Density alone doesn't distinguish a uniformly
dense network from one with a few tightly-connected clusters sitting
inside an otherwise sparse graph — the latter is the more specific
testlet signature, and it's exactly what the local-dependence vignette's
[Q3 heatmaps](local_dependence.qmd#a-worst-case-and-a-clean-case-side-by-side)
are built to catch directly. Detecting that cluster structure from the
network side properly needs community-detection tools this pass didn't
run at batch scale (see Limitations); density is the coarser proxy
available here.

### Do dimensionality and local dependence predict the correspondence jointly?

Most tables in this sample clear the dimensionality vignette's
unidimensionality bar, yet the median strength-discrimination correlation
sits around 0.7, not close to 1 — and the local-dependence cross-check
above is the weaker of the two on its own (r = `r round(cor_ld_join, 2)`
against network density, not strength-discrimination correlation directly).
Neither diagnostic looking decisive by itself doesn't rule out the two
being jointly informative: a table could clear the eigenvalue-ratio bar
while still carrying enough Q3-flagged pairs to suppress the
strength-discrimination correlation in a way an eigenvalue ratio alone
wouldn't show.

```{r}
#| label: joint-dim-ld-inline
#| echo: false

joint_dim_ld_df <- summary_df %>%
  filter(!is.na(strength_a_cor), !is.na(dim_ratio_12), !is.na(ld_prop_flagged)) %>%
  mutate(log_dim_ratio = log10(pmin(dim_ratio_12, 15)))

m_dim_only <- lm(strength_a_cor ~ log_dim_ratio, data = joint_dim_ld_df)
m_ld_only  <- lm(strength_a_cor ~ ld_prop_flagged, data = joint_dim_ld_df)
m_additive <- lm(strength_a_cor ~ log_dim_ratio + ld_prop_flagged, data = joint_dim_ld_df)
m_interact <- lm(strength_a_cor ~ log_dim_ratio * ld_prop_flagged, data = joint_dim_ld_df)

r2_dim <- summary(m_dim_only)$r.squared
r2_ld  <- summary(m_ld_only)$r.squared
r2_add <- summary(m_additive)$r.squared
r2_int <- summary(m_interact)$r.squared
int_coef <- coef(m_interact)["log_dim_ratio:ld_prop_flagged"]
int_p    <- summary(m_interact)$coefficients["log_dim_ratio:ld_prop_flagged", "Pr(>|t|)"]
anova_int <- anova(m_additive, m_interact)
anova_p   <- anova_int[["Pr(>F)"]][2]
```

A joint model bears that out, modestly. Fit separately, log eigenvalue
ratio explains `r scales::percent(r2_dim, accuracy = 1)` of the variance in
strength-discrimination correlation across the `r nrow(joint_dim_ld_df)`
tables with both diagnostics available, and Q3-flagged proportion alone
explains `r scales::percent(r2_ld, accuracy = 1)`. Together, additively,
that only rises to `r scales::percent(r2_add, accuracy = 1)` — mostly what
you'd expect from two weak, largely independent signals. Adding an
interaction term brings it to `r scales::percent(r2_int, accuracy = 1)`,
a small but statistically real improvement
(`r if (anova_p < 0.001) "F-test p < 0.001" else paste0("F-test p = ", round(anova_p, 3))`
comparing the additive and interaction models). The interaction itself
(coefficient `r round(int_coef, 2)`,
`r if (int_p < 0.001) "p < 0.001" else paste0("p = ", round(int_p, 3))`) is
positive: local dependence's negative pull on the correlation is sharper
for tables that *don't* clear the unidimensionality bar, and fades for
tables that do — a plausible story (a strong single factor can absorb a
little local dependence without breaking the network-IRT correspondence;
a weaker or absent one can't), but a small effect against a large amount of
unexplained variance. Combined, dimensionality and local dependence still
account for well under a tenth of the spread in strength-discrimination
correlation — whatever explains most of the ~0.7-not-1 gap for
"cleanly unidimensional" tables, it mostly isn't either diagnostic used
here, alone or together.

## How much statistical evidence actually supports these edges?

```{r}
#| label: bayes-setup-inline
#| echo: false

n_bayes_matched <- sum(summary_df$has_bayes_match, na.rm = TRUE)
n_mrf_matched   <- sum(summary_df$has_mrf_match, na.rm = TRUE)

evidence_levels <- c("strong_presence", "weak_presence", "inconclusive",
                     "weak_absence", "strong_absence")
evidence_labels <- c("Strong presence", "Weak presence", "Inconclusive",
                     "Weak absence", "Strong absence")

irw_overall_ggm <- if (nrow(edges_ggm_df) > 0) {
  table(factor(edges_ggm_df$evidence_category_ggm, levels = evidence_levels)) /
    sum(!is.na(edges_ggm_df$evidence_category_ggm))
} else rep(NA_real_, 5)
names(irw_overall_ggm) <- evidence_labels

irw_inconclusive <- unname(irw_overall_ggm["Inconclusive"])
irw_weak         <- unname(irw_overall_ggm["Weak presence"] + irw_overall_ggm["Weak absence"])
irw_strong       <- unname(irw_overall_ggm["Strong presence"] + irw_overall_ggm["Strong absence"])
```

The paper's headline numbers, across the 293 published networks they
reanalyzed: 31.7% of edges had *inconclusive* evidence, 49.7% had only
*weak* evidence either way (4.4% weak presence + 45.2% weak absence), and
just 18.7% had *strong* evidence (13.0% strong presence + 5.7% strong
absence). Most published edges, in other words, are asserted with far
less statistical support than a single regularized network graph implies.

Using Option A (GGM via BGGM, forced uniformly, matching their method
exactly), this pass produced usable Bayesian edge evidence for
`r n_bayes_matched` of `r nrow(summary_df)` tables. Pooling every edge
across every table (not averaging per-table percentages, matching how the
paper itself pools across networks for its headline numbers):

```{r}
#| label: fig-evidence-breakdown
#| fig-cap: "IRW's overall Bayesian edge-evidence breakdown (this pass, pooled across all tables) against Huth et al.'s (2026) published benchmark. Each bar is one evidence category from the same five-way classification."
#| fig-height: 4

breakdown_df <- bind_rows(
  tibble(source = "IRW (this pass)", category = evidence_labels,
         proportion = as.numeric(irw_overall_ggm)),
  tibble(source = "Huth et al. (2026)", category = evidence_labels,
         proportion = c(0.130, 0.044, 0.317, 0.452, 0.057))
) %>%
  mutate(category = factor(category, levels = evidence_labels))

p <- ggplot(breakdown_df, aes(x = category, y = proportion, fill = source)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("IRW (this pass)" = irw_blue, "Huth et al. (2026)" = irw_grey)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion of edges", fill = NULL) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

ggplotly(p) |> layout(margin = list(t = 60), legend = list(orientation = "h", y = -0.2))
```

IRW's item-network data lands `r if (!is.na(irw_strong) && irw_strong > 0.187) "better-supported than" else if (!is.na(irw_strong) && irw_strong < 0.187) "less well-supported than" else "comparably supported to"`
the published literature the paper reanalyzed: `r scales::percent(irw_inconclusive, accuracy = 0.1)`
inconclusive (vs. their 31.7%), `r scales::percent(irw_weak, accuracy = 0.1)` weak either way
(vs. their 49.7%), and `r scales::percent(irw_strong, accuracy = 0.1)` strong either way (vs.
their 18.7%). Read this comparison with real caution: the paper's 293
networks span whatever sample sizes and item counts happened to appear in
the published literature they sampled, while this pass's 5-30-item,
300+-participant filter (see Data and methods above) is an IRW-specific
compute-budget choice, not a matched comparison design -- the gap above
is suggestive, not a controlled replication.

### Does evidence strength track relative sample size?

Huth et al.'s central moderator isn't raw sample size — it's **relative**
sample size, observations per possible edge: N / (p(p-1)/2). Their Fig. 3
bins networks into relative-sample-size groups and shows the evidence
breakdown shift toward stronger evidence as relative sample size rises.

```{r}
#| label: fig-relative-n-bins
#| fig-cap: "Evidence-category breakdown by relative sample size bin (quartiles of n_participants / (n_items * (n_items - 1) / 2) across tables with usable Bayesian edge evidence), pooling edges within each bin. If IRW shows the same pattern Huth et al. found, strong presence/absence should grow and inconclusive should shrink from left to right."
#| fig-height: 5

rel_n_bins_df <- summary_df %>%
  filter(has_bayes_match, !is.na(relative_n_ggm)) %>%
  mutate(rel_n_bin = ntile(relative_n_ggm, 4)) %>%
  select(table, rel_n_bin, relative_n_ggm)

bin_ranges <- rel_n_bins_df %>%
  group_by(rel_n_bin) %>%
  summarise(lo = min(relative_n_ggm), hi = max(relative_n_ggm), .groups = "drop") %>%
  mutate(label = paste0("Q", rel_n_bin, "\n[", round(lo, 1), "-", round(hi, 1), "]"))

bin_breakdown_df <- edges_ggm_df %>%
  inner_join(rel_n_bins_df, by = "table") %>%
  filter(!is.na(evidence_category_ggm)) %>%
  count(rel_n_bin, evidence_category_ggm) %>%
  group_by(rel_n_bin) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup() %>%
  left_join(bin_ranges, by = "rel_n_bin") %>%
  mutate(
    evidence_category_ggm = factor(evidence_category_ggm, levels = evidence_levels, labels = evidence_labels),
    label = factor(label, levels = bin_ranges$label)
  )

p <- ggplot(bin_breakdown_df, aes(x = label, y = proportion, fill = evidence_category_ggm)) +
  geom_col(position = "stack") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = setNames(
    c(irw_blue, "#67a9cf", irw_grey, "#f4a582", irw_red), evidence_labels
  )) +
  labs(x = "Relative sample size quartile (N / possible edges)", y = "Proportion of edges", fill = NULL)

ggplotly(p) |> layout(margin = list(t = 40), legend = list(orientation = "h", y = -0.3))
```

```{r}
#| label: relative-n-inline
#| echo: false

strong_by_bin <- bin_breakdown_df %>%
  filter(evidence_category_ggm %in% c("Strong presence", "Strong absence")) %>%
  group_by(rel_n_bin) %>%
  summarise(prop_strong = sum(proportion), .groups = "drop") %>%
  arrange(rel_n_bin)
monotonic_strong <- all(diff(strong_by_bin$prop_strong) >= -0.02)  # small tolerance for noise
```

`r if (monotonic_strong) "IRW shows essentially the same monotonic pattern Huth et al. found: strong evidence (presence or absence combined) rises from the lowest to the highest relative-sample-size quartile." else "IRW does not cleanly reproduce the paper's monotonic pattern here -- strong evidence does not rise monotonically across relative-sample-size quartiles in this pass, even though the same relative-sample-size definition and binning approach were used."`
Strong evidence (presence + absence) by quartile:
`r paste(scales::percent(strong_by_bin$prop_strong, accuracy = 1), collapse = " -> ")`
(lowest to highest relative sample size).

### Does edge weight predict evidence strength?

The paper's Fig. 4 found no sharp relationship between an edge's partial
correlation magnitude and its evidence strength (BF10) — similar-sized
partial correlations landed in every evidence category — though the
correspondence was cleaner in their higher-relative-sample-size networks.

```{r}
#| label: fig-pcor-vs-bf
#| fig-cap: "Partial correlation magnitude against log10(BF10) for every edge with usable evidence (Option A/GGM), colored by evidence category, split by whether the edge's table falls above or below the median relative sample size. A clean relationship would show points aligning along a smooth curve rather than every category scattered across the same pcor range."
#| fig-height: 6

median_rel_n <- median(summary_df$relative_n_ggm[summary_df$has_bayes_match], na.rm = TRUE)

pcor_bf_df <- edges_ggm_df %>%
  inner_join(summary_df %>% select(table, relative_n_ggm), by = "table") %>%
  filter(!is.na(evidence_category_ggm), !is.na(relative_n_ggm)) %>%
  mutate(
    rel_n_group = ifelse(relative_n_ggm >= median_rel_n,
                          "Above-median relative sample size", "Below-median relative sample size"),
    evidence_category_ggm = factor(evidence_category_ggm, levels = evidence_levels, labels = evidence_labels),
    log10_bf = log10(pmin(pmax(BF10_ggm, 1e-10), 1e10))  # cap for display; classification itself is unaffected
  )

p <- ggplot(pcor_bf_df, aes(x = pcor_ggm, y = log10_bf, colour = evidence_category_ggm)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_hline(yintercept = c(-1, log10(1/3), log10(3), 1), linetype = "dashed", colour = "grey70") +
  scale_colour_manual(values = setNames(
    c(irw_blue, "#67a9cf", irw_grey, "#f4a582", irw_red), evidence_labels
  )) +
  facet_wrap(~ rel_n_group, ncol = 1) +
  labs(x = "Partial correlation (GGM)", y = expression(log[10](BF[10])), colour = NULL)

ggplotly(p, tooltip = c("x", "y", "colour")) |> layout(margin = list(t = 40))
```

```{r}
#| label: pcor-bf-inline
#| echo: false

cor_pcor_bf_all <- cor(abs(pcor_bf_df$pcor_ggm), pcor_bf_df$log10_bf, use = "complete.obs")
cor_pcor_bf_hi  <- with(filter(pcor_bf_df, rel_n_group == "Above-median relative sample size"),
                         cor(abs(pcor_ggm), log10_bf, use = "complete.obs"))
cor_pcor_bf_lo  <- with(filter(pcor_bf_df, rel_n_group == "Below-median relative sample size"),
                         cor(abs(pcor_ggm), log10_bf, use = "complete.obs"))
```

Overall, `|`partial correlation`|` and log(BF10) correlate at r = `r round(cor_pcor_bf_all, 2)`
across all edges — real but far from a tight relationship, matching the
paper's "no sharp relationship" finding. Splitting by relative sample size:
r = `r round(cor_pcor_bf_hi, 2)` in above-median-relative-sample-size
tables versus r = `r round(cor_pcor_bf_lo, 2)` below the median.
`r if (!is.na(cor_pcor_bf_hi) && !is.na(cor_pcor_bf_lo) && cor_pcor_bf_hi > cor_pcor_bf_lo) "The correspondence is cleaner in higher-relative-sample-size tables here too, matching what Huth et al. found in their own data." else "This pass does not show the same higher-relative-sample-size cleanup Huth et al. found -- the relationship isn't obviously tighter in IRW's better-powered tables."`

```{r}
#| label: mrf-agreement-inline
#| echo: false
#| output: asis

if (nrow(edges_mrf_df) > 0) {
  agree_df <- edges_ggm_df %>%
    filter(table %in% option_b_tables) %>%
    inner_join(edges_mrf_df, by = c("table", "item_i", "item_j")) %>%
    filter(!is.na(evidence_category_ggm), !is.na(evidence_category_mrf))
  n_agree_edges <- nrow(agree_df)
  pct_exact_agree <- mean(agree_df$evidence_category_ggm == agree_df$evidence_category_mrf)
  pct_direction_agree <- mean(
    grepl("presence", agree_df$evidence_category_ggm) == grepl("presence", agree_df$evidence_category_mrf) |
    agree_df$evidence_category_ggm == "inconclusive" | agree_df$evidence_category_mrf == "inconclusive"
  )

  cat("### Does the ordinal MRF agree with the GGM?\n\n")
  cat("Huth et al. acknowledge the ordinal Markov random field (fit via `bgms`) as ",
      "conceptually better-suited to binary/ordinal data than the GGM they actually used. ",
      "Testing that acknowledgment directly, on the ", length(option_b_tables),
      "-table Option B subset (", n_agree_edges, " edges with usable evidence under both models): ",
      "the two models put the same edge in the *exact same* one of five evidence categories ",
      scales::percent(pct_exact_agree, accuracy = 1), " of the time, and agree on ",
      "presence-vs-absence direction (treating inconclusive calls under either model as a pass) ",
      scales::percent(pct_direction_agree, accuracy = 1), " of the time.\n\n",
      if (pct_exact_agree < 0.5) {
        "That's a real disagreement rate, not just estimation noise -- for a meaningful share of edges, which model IRW's own data says is theoretically more appropriate changes the conclusion about whether that edge is real. Given the paper's own acknowledgment that GGM is a simplifying approximation for this data type, the exact-category numbers above should be read as a lower bound on how much Option A's uniform-GGM choice matters for binary/ordinal data specifically, not just a robustness footnote."
      } else {
        "The two models mostly agree, which is some reassurance that Option A's uniform-GGM simplification isn't badly distorting the picture for IRW's binary/ordinal tables -- though this check only covers a small illustrative subset, not the full candidate pool."
      }, "\n", sep = "")
} else {
  cat("### Does the ordinal MRF agree with the GGM?\n\n",
      "Option B (ordinal MRF) results were not available in this cache -- this comparison ",
      "requires re-running `network_psych_compute.R` with the current amendment.\n", sep = "")
}
```

### Do the cross-vignette checks look cleaner using only strong-evidence edges?

Both cross-vignette comparisons above used the frequentist regularized
network's edges, kept or dropped by LASSO shrinkage. Here, instead,
**node strength is rebuilt using only edges the Bayesian model calls
strong evidence for** (BF10 > 10) — a stricter, statistically-justified
filter rather than a regularization penalty — to see whether the two
cross-vignette relationships get cleaner once weakly-supported edges are
dropped entirely rather than merely down-weighted.

```{r}
#| label: fig-strong-only-dim
#| fig-cap: "Strength-discrimination correlation computed from strong-evidence-only edges, against the dimensionality vignette's eigenvalue ratio (compare to the all-edges version earlier on this page). Node strength here sums |partial correlation| over only that node's strong-presence edges; a node with none gets strength 0."
#| fig-height: 5

strong_edges_long <- edges_ggm_df %>%
  filter(evidence_category_ggm == "strong_presence") %>%
  { bind_rows(
      transmute(., table, item = item_i, w = abs(pcor_ggm)),
      transmute(., table, item = item_j, w = abs(pcor_ggm))
    ) } %>%
  group_by(table, item) %>%
  summarise(strength = sum(w), .groups = "drop")

strong_strength_cor_df <- map_dfr(intersect(unique(edges_ggm_df$table), names(discriminations_list)), function(tbl) {
  a <- discriminations_list[[tbl]]
  if (is.null(a) || length(a) < 2) return(NULL)
  items <- names(a)
  strength <- setNames(rep(0, length(items)), items)
  tbl_strong <- strong_edges_long[strong_edges_long$table == tbl, ]
  if (nrow(tbl_strong) > 0) strength[tbl_strong$item] <- tbl_strong$strength
  if (length(unique(strength)) < 2) return(NULL)
  tibble(table = tbl, strong_strength_a_cor = suppressWarnings(cor(strength, a[items])))
})

join_dim_strong_df <- strong_strength_cor_df %>%
  inner_join(summary_df %>% select(table, dim_ratio_12, dim_unidimensional), by = "table") %>%
  filter(!is.na(strong_strength_a_cor), !is.na(dim_ratio_12))

p <- ggplot(join_dim_strong_df, aes(x = pmin(dim_ratio_12, 15), y = strong_strength_a_cor)) +
  geom_point(colour = irw_blue, size = 2, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = irw_red, linewidth = 0.8) +
  scale_x_log10() +
  labs(x = "Eigenvalue ratio (dimensionality vignette, log scale, capped at 15)",
       y = "Strong-evidence-only strength-discrimination correlation")

ggplotly(p) |> layout(margin = list(t = 40))
```

```{r}
#| label: strong-dim-inline
#| echo: false

n_strong_dim_join <- nrow(join_dim_strong_df)
med_all_cor    <- median(summary_df$strength_a_cor, na.rm = TRUE)
med_strong_cor <- median(join_dim_strong_df$strong_strength_a_cor, na.rm = TRUE)

strong_ld_df <- summary_df %>%
  filter(has_bayes_match, !is.na(prop_strong_presence_ggm), !is.na(ld_prop_flagged))
cor_strong_ld <- cor(strong_ld_df$prop_strong_presence_ggm, strong_ld_df$ld_prop_flagged)
cor_all_ld    <- cor(join_ld_df$network_density, join_ld_df$ld_prop_flagged)
```

Across `r n_strong_dim_join` tables, the strong-evidence-only
strength-discrimination correlation has a median of `r round(med_strong_cor, 2)`,
versus `r round(med_all_cor, 2)` for the all-edges version earlier on this
page.
`r if (!is.na(med_strong_cor) && !is.na(med_all_cor) && med_strong_cor > med_all_cor) "Restricting to strong-evidence edges sharpens the correspondence with IRT discrimination -- consistent with the idea that the earlier relationship was partly diluted by weakly-supported edges the regularized network kept anyway." else "Restricting to strong-evidence edges doesn't obviously sharpen the correspondence with IRT discrimination here -- the earlier relationship doesn't appear to be primarily a weak-edge dilution effect."`

For the local-dependence check, the proportion of a table's edges
classified strong presence correlates with the proportion of Q3-flagged
pairs at r = `r round(cor_strong_ld, 2)`, versus r = `r round(cor_all_ld, 2)`
for network density against the same Q3-flagged proportion earlier.
`r if (!is.na(cor_strong_ld) && !is.na(cor_all_ld) && abs(cor_strong_ld) > abs(cor_all_ld)) "The testlet signal is somewhat cleaner using strong-evidence-only edges than using raw regularized-network density." else "The testlet signal isn't obviously cleaner using strong-evidence-only edges than using raw regularized-network density -- density from LASSO-regularized edges and density from Bayesian strong-evidence edges appear to carry similar information here."`

### Does the correspondence hold up better in overall more-stable networks?

Dropping weak edges (above) changes what node strength is computed
*from* — it leaves out edges that still contribute to the frequentist
network's centrality ordering, which conflates two different questions:
whether more stable estimation gives a better correspondence, and
whether specifically discarding weakly-supported edges does. Here's a
cleaner version of the first question: instead of rebuilding node
strength from a trimmed edge set, keep each table's **full** regularized
network and its already-computed strength-discrimination correlation
unchanged, and instead split tables into groups by how much overall
Bayesian support their edge set has — the proportion of a table's edges
classified strong evidence (presence or absence combined) under Option A.

```{r}
#| label: fig-stability-vs-cor
#| fig-cap: "Strength-discrimination correlation (full network, not edge-trimmed) against each table's overall proportion of strong-evidence edges (presence + absence combined). If more stable overall evidence goes with a better correspondence, points should trend upward left to right."
#| fig-height: 5

stability_df <- summary_df %>%
  filter(has_bayes_match, !is.na(strength_a_cor),
         !is.na(prop_strong_presence_ggm), !is.na(prop_strong_absence_ggm)) %>%
  mutate(prop_strong_overall = prop_strong_presence_ggm + prop_strong_absence_ggm,
         tooltip = paste0(
           "Table: ", table,
           "<br>Strength-a cor: ", round(strength_a_cor, 2),
           "<br>Prop. strong evidence: ", round(prop_strong_overall, 2)
         ))

p <- ggplot(stability_df, aes(x = prop_strong_overall, y = strength_a_cor, text = tooltip)) +
  geom_point(colour = irw_blue, size = 1.8, alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, colour = irw_red, linewidth = 0.8) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Proportion of table's edges with strong Bayesian evidence (presence + absence)",
       y = "Strength-discrimination correlation (full network)")

ggplotly(p, tooltip = "text") |> layout(margin = list(t = 60))
```

```{r}
#| label: stability-inline
#| echo: false

STABILITY_CUT <- 0.7
n_stability     <- nrow(stability_df)
cor_stability   <- cor(stability_df$prop_strong_overall, stability_df$strength_a_cor)
stable_hi       <- stability_df$strength_a_cor[stability_df$prop_strong_overall >= STABILITY_CUT]
stable_lo       <- stability_df$strength_a_cor[stability_df$prop_strong_overall <  STABILITY_CUT]
med_stable_hi   <- median(stable_hi)
med_stable_lo   <- median(stable_lo)
wilcox_stability <- wilcox.test(stable_hi, stable_lo)$p.value
```

Across `r n_stability` tables, overall stability and the
strength-discrimination correlation move together (r =
`r round(cor_stability, 2)`). Splitting at
`r scales::percent(STABILITY_CUT, accuracy = 1)` strong-evidence edges —
`r length(stable_hi)` tables at or above that bar, `r length(stable_lo)`
below it — the more-stable group has a median correlation of
`r round(med_stable_hi, 2)` versus `r round(med_stable_lo, 2)` for the
less-stable group (Wilcoxon
`r if (wilcox_stability < 0.001) "p < 0.001" else paste0("p = ", round(wilcox_stability, 3))`).
Unlike the strong-evidence-only edge trimming above, this comparison
never removes an edge from the network used to compute strength — it
only asks whether tables whose *overall* edge set is better-supported
tend to show a cleaner correspondence with IRT discrimination, and by
that test the answer is a reasonably clear yes: more stable estimation
does go with better agreement between the two frameworks, independent of
whether any specific edges get trimmed.

### How sensitive is the evidence classification to the prior?

Every result in this section rests on one prior choice: the matrix-F
`prior_sd = 0.25` for Option A's GGM partial correlations, Huth et al.'s own
default rather than anything tuned for IRW's data. Huth et al. ran their own
prior-sensitivity check across 3 SD values on their data; here's the same
idea on a 6-table handful drawn from the Option B stratified subset
(`ecuador_2011_safety_avoidance`, `geiser_tam`, `gilbert_meta_12`,
`2024_online_addiction_sabas`, `amatus_cipora_2024_pisa_me`,
`che_2026_regulatory_self_efficacy`), refitting at `prior_sd` = 0.1, 0.25,
and 0.5 and comparing every edge's evidence-category classification back to
the 0.25 default.

```{r}
#| label: prior-sensitivity-inline
#| echo: false

prior_sens_path <- "network_psych_data/network_psych_prior_sensitivity_results.rds"
prior_sens_ok <- file.exists(prior_sens_path)
if (prior_sens_ok) {
  prior_sens <- readRDS(prior_sens_path)
  ps_comp <- prior_sens$comparison
  n_ps_edges <- nrow(ps_comp)
  flip_01 <- mean(ps_comp$evidence_category_sd0.25 != ps_comp$evidence_category_sd0.1, na.rm = TRUE)
  flip_05 <- mean(ps_comp$evidence_category_sd0.25 != ps_comp$evidence_category_sd0.5, na.rm = TRUE)
  sev <- c(strong_absence = 1, weak_absence = 2, inconclusive = 3,
           weak_presence = 4, strong_presence = 5)
  sev25 <- sev[ps_comp$evidence_category_sd0.25]
  sev01 <- sev[ps_comp$evidence_category_sd0.1]
  sev05 <- sev[ps_comp$evidence_category_sd0.5]
  mean_step_01 <- mean(abs(sev25 - sev01), na.rm = TRUE)
  mean_step_05 <- mean(abs(sev25 - sev05), na.rm = TRUE)
  prop_sign_flip <- mean(
    (sign(sev25 - 3) != sign(sev01 - 3) & sev25 != 3 & sev01 != 3) |
    (sign(sev25 - 3) != sign(sev05 - 3) & sev25 != 3 & sev05 != 3),
    na.rm = TRUE
  )
  cor_bf_01 <- cor(log10(pmax(ps_comp$bf10_sd0.25, 1e-10)),
                    log10(pmax(ps_comp$bf10_sd0.1, 1e-10)), use = "complete.obs")
  cor_bf_05 <- cor(log10(pmax(ps_comp$bf10_sd0.25, 1e-10)),
                    log10(pmax(ps_comp$bf10_sd0.5, 1e-10)), use = "complete.obs")
}
```

`r if (prior_sens_ok) paste0("Across ", n_ps_edges, " edges, ", scales::percent(flip_01, accuracy = 1), " change evidence category when the prior tightens to SD = 0.1, and ", scales::percent(flip_05, accuracy = 1), " when it loosens to SD = 0.5.") else "Prior-sensitivity results not available."`
Read on its own, that looks like the five-way classification is quite
unstable. It's less alarming once decomposed: the average edge only moves
`r if (prior_sens_ok) round(mean_step_01, 2) else "--"` categories (out of
4 possible steps) under the tighter prior and
`r if (prior_sens_ok) round(mean_step_05, 2) else "--"` under the looser
one — mostly a single step, e.g. `weak_absence` to `inconclusive`, not a
jump from `strong_presence` to `strong_absence`. Genuine
presence-vs-absence reversals (crossing `inconclusive` entirely) happen for
only `r if (prior_sens_ok) scales::percent(prop_sign_flip, accuracy = 1) else "--"`
of edges. And the underlying continuous evidence, before any threshold is
applied, is reasonably stable: `log10(BF10)` correlates at
`r if (prior_sens_ok) round(cor_bf_01, 2) else "--"` (SD 0.1) and
`r if (prior_sens_ok) round(cor_bf_05, 2) else "--"` (SD 0.5) against the
default prior's values. The five-way classification's apparent instability
is mostly an artifact of hard thresholds sitting close to where edges
naturally cluster (near BF10 = 1/3 and BF10 = 10) — a known property of
discretizing a continuous quantity, not evidence that the underlying
Bayesian estimates themselves move around unpredictably with the prior.
This was checked on 6 tables only, not the full candidate pool — enough to
see the pattern, not enough to rule out a table where prior choice matters
more than it did here.

### Decomposing the gap: dimensionality, local dependence, stability, and negative edges together

Van der Maas et al. (2006) and Epskamp et al. (2018) derive a tight
analytical correspondence between network centrality and IRT
discrimination under a single dominant common cause; empirically, across
this sample, the correspondence is real but partial. Every check above
tested one candidate explanation for that gap at a time — multidimensionality,
local dependence, negative edges, estimation stability — each on its own
axis. None of them ruled the others out, and some plausibly overlap (the
sign-mismatch section above found negative edges and negative
discriminations concentrate on the same items; local dependence's effect
above only showed up in interaction with dimensionality, not as a main
effect). A single joint model, fit once with all four together, is a more
honest way to ask how much each one contributes once the others are held
fixed.

```{r}
#| label: joint-decomposition-inline
#| echo: false

negedge_join_df <- summary_df %>%
  filter(!is.na(strength_a_cor)) %>%
  rowwise() %>%
  mutate(prop_neg_edges = {
    g <- graphs_list[[table]]
    ut <- g[upper.tri(g)]
    nz <- ut[ut != 0]
    if (length(nz) == 0) NA_real_ else mean(nz < 0)
  }) %>%
  ungroup()

decomp_df <- summary_df %>%
  select(table, strength_a_cor, dim_ratio_12, ld_prop_flagged,
         prop_strong_presence_ggm, prop_strong_absence_ggm, has_bayes_match) %>%
  inner_join(negedge_join_df %>% select(table, prop_neg_edges), by = "table") %>%
  filter(!is.na(strength_a_cor), !is.na(dim_ratio_12), !is.na(ld_prop_flagged),
         !is.na(prop_neg_edges), has_bayes_match,
         !is.na(prop_strong_presence_ggm), !is.na(prop_strong_absence_ggm)) %>%
  mutate(log_dim_ratio = log10(pmin(dim_ratio_12, 15)),
         prop_strong_overall = prop_strong_presence_ggm + prop_strong_absence_ggm)

m_no_negedge <- lm(strength_a_cor ~ log_dim_ratio + ld_prop_flagged + prop_strong_overall,
                    data = decomp_df)
m_decomp <- lm(strength_a_cor ~ log_dim_ratio + ld_prop_flagged + prop_neg_edges + prop_strong_overall,
               data = decomp_df)
decomp_coef <- summary(m_decomp)$coefficients
r2_no_negedge <- summary(m_no_negedge)$r.squared
r2_decomp <- summary(m_decomp)$r.squared
pred_cor_max <- max(abs(cor(decomp_df[, c("log_dim_ratio","ld_prop_flagged",
                                           "prop_neg_edges","prop_strong_overall")])[upper.tri(diag(4))]))
```

Across the `r nrow(decomp_df)` tables with all four measures available,
dimensionality, local dependence, and stability together explain
`r scales::percent(r2_no_negedge, accuracy = 1)` of the variance in
strength-discrimination correlation — already more than any pair of them
managed above, but still modest. Adding negative-edge share to the same
model more than doubles that, to
`r scales::percent(r2_decomp, accuracy = 1)`, and it dominates the other
three: its coefficient (`r round(decomp_coef["prop_neg_edges", "Estimate"], 2)`,
`r if (decomp_coef["prop_neg_edges", "Pr(>|t|)"] < 0.001) "p < 0.001" else paste0("p = ", round(decomp_coef["prop_neg_edges", "Pr(>|t|)"], 3))`)
dwarfs the others in both size and significance. Local dependence's main
effect, significant on its own above only via the interaction with
dimensionality, drops to essentially nothing once negative-edge share is
in the model (`r round(decomp_coef["ld_prop_flagged", "Estimate"], 3)`,
`r if (decomp_coef["ld_prop_flagged", "Pr(>|t|)"] < 0.001) "p < 0.001" else paste0("p = ", round(decomp_coef["ld_prop_flagged", "Pr(>|t|)"], 3))`)
— consistent with local dependence's apparent effect being partly a proxy
for negative-edge share rather than an independent driver, though the
predictors aren't strongly collinear here (largest pairwise correlation
among the four: `r round(pred_cor_max, 2)`), so this looks like a real
dominance relationship rather than a multicollinearity artifact.
Dimensionality and stability both keep small, roughly stable coefficients
with or without negative-edge share in the model — real but minor
independent contributions.

Two things follow. First, negative edges aren't just one nuisance factor
among several — they're the dominant one this pass can measure, and given
the sign-mismatch finding above, a meaningful share of that is plausibly a
metric artifact (unsigned strength vs. signed discrimination) rather than
a genuine breakdown in the network-IRT correspondence. Second, even the
full four-predictor model leaves most of the variance unexplained
(`r scales::percent(1 - r2_decomp, accuracy = 1)`). Whatever else separates
Epskamp et al.'s analytical prediction from what this sample shows — sampling
noise in a single network estimate, item-level idiosyncrasies no
table-level diagnostic captures, network-estimation choices this vignette
didn't vary (EBIC tuning, alternative regularization defaults) — isn't
captured by any of the four candidate explanations tested across this
vignette, alone or together.

## Practical implications

Where the strength-discrimination correlation is high and the network is
dense, uniform, and positively connected, network psychometrics and IRT
are telling a similar story about the same items — using either
framework to identify the "best" items would likely agree, though this
vignette measures agreement between the two statistics rather than
agreement between the item-selection decisions each would produce. Where
it breaks down, that's worth taking seriously in its own right,
independent of which framework you prefer: it means the two ways of
describing item covariation disagree about which items matter most, and
a table flagged as multidimensional or locally dependent elsewhere on
this site, or with a substantial share of negative network edges (see
above), is a reasonable place to expect exactly that disagreement. These
recommendations follow from a relatively small, first-pass empirical
scan rather than a full mechanistic account of *why* the two frameworks
diverge when they do — read them as a starting point for where to look
more carefully, not a settled prescription.

## Limitations

- **EBICglasso on ordinal data, two gaps found and one fixed by
  refitting.** `bootnet::estimateNetwork()`'s `EBICglasso` preset does not
  call `qgraph::cor_auto()` by default — as of `bootnet` 1.4, the
  `EBICglasso` default set switched from `corMethod = "cor_auto"` to
  plain Pearson correlations (`cor`). Neither `bootnet` nor `qgraph` was
  pinned in `renv.lock` at the time (now fixed — see below), so an earlier
  pass of this vignette likely ran on whatever version was on the machine
  at the time and may have silently treated 7-point Likert integers as
  literally continuous. This pass instead passes `corMethod = "cor_auto"`
  explicitly (polychoric/polyserial correlations for ordinal items) with
  `corArgs = list(forcePD = TRUE)`, since a polychoric correlation matrix
  over several ordinal items is not always positive-definite and
  `bootnet` errors out by default rather than correcting it; `forcePD`
  projects to the nearest valid correlation matrix instead. The impact was
  real and sometimes large: `clifford_2018_police_blame`, the lowest
  strength-discrimination correlation table in an earlier pass of this
  vignette (r = −0.69), moved to r = 0.09 once the network side used
  polychoric correlations and the IRT side used GPCM instead of GRM (next
  point) — a different table is now the lowest-correlation example (see
  the worked-case callout above). Separately, and regardless of that fix:
  even an exact polychoric-correlation Gaussian graphical model is still
  not the network model with a proven equivalence to an ordinal IRT
  model. Marsman, van den Bergh, & Haslbeck (2025) [@marsman2025bayesian]
  derive that equivalence for the **ordinal Markov random field**, and
  show it corresponds to the **generalized partial credit model (GPCM)**,
  not the graded response model most applied network-IRT comparisons use
  for polytomous items — this vignette now fits GPCM for that reason (see
  Data and methods), but an exact ordinal-MRF network fit (rather than a
  Gaussian approximation on polychoric correlations) is still not
  implemented here.
- **Weighted density is not comparable across the two estimators.**
  EBICglasso edge weights are partial correlations (bounded by 1);
  IsingFit weights are log-odds and unbounded, so the same nominal
  "weighted density" means different things in the two halves of the
  sample. The density figure and the correlations reported with it are
  therefore split by estimator rather than pooled, and one IsingFit table
  (`eurpar2_mudfold`, weighted density 14.1, over ten times the
  next-largest value) is excluded from the IsingFit panel and its
  correlation as an outlier on that unbounded scale.
- **Network estimation is more sample- and tuning-sensitive** than fitting
  a single IRT model. Small tables in this sample should be read
  cautiously; an empty regularized network (no edges survive
  regularization) is itself informative but excludes that table from the
  strength-discrimination comparison entirely.
- **No bootstrap stability checks.** `bootnet`'s case-dropping bootstrap
  (the standard way to check whether an estimated network's edges and
  centrality ordering are stable) was not run at full-batch scale for
  compute reasons. A natural follow-up: run it for a handful of the most
  interesting example tables from the full batch (the strongest
  confirmation, the sharpest breakdown) rather than all of them.
- **Density is a coarse proxy for "clustered."** As noted above, telling
  a uniformly dense network apart from one with tight sub-clusters needs
  community-detection tools this pass didn't run.
- **Cache overlap is partial.** The dimensionality and local-dependence
  caches were built on their own candidate pools (different item-count
  caps), so not every table here has a match in both — see the match-rate
  numbers in Data and methods.
- **`bootnet` and `qgraph` were not pinned in `renv.lock` when this pass
  was run.** This pass ran against whatever versions of these two packages
  happened to be installed, which is how the `cor_auto` default-behavior
  change above went unnoticed. Both are now pinned in `renv.lock`, which
  prevents a silent version drift like this going forward but doesn't
  retroactively confirm which version produced the results above.
- **Matrix-F prior sensitivity was checked on a 6-table handful, not the
  full candidate pool** (see "How sensitive is the evidence classification
  to the prior?" above). The five-way evidence-category classification
  looks unstable at first glance (roughly half of edges change category
  under a tighter or looser prior), but the underlying continuous BF10 is
  much more stable (log-scale correlation ~0.6 against the default prior)
  and genuine presence-vs-absence reversals are rare (~2%) — most flips are
  single-step moves across a nearby threshold, not evidence the Bayesian
  estimates themselves are prior-dominated. Not checked at full-batch
  scale, and the three SD values used (0.1/0.25/0.5) were chosen as a
  reasonable spread rather than independently confirmed against Huth et
  al.'s own three values.
- **The stochastic-block-model check (see "A network-native
  multidimensionality check" above) ran on the 20-table Option B subset,
  and lost 6 of those 20 to data/convergence issues** (1 unsupported data
  type, 1 no strength-discrimination match, 4 severe MCMC divergence — 3 at
  100% of post-warmup samples). No table in the remaining 14 cleared the
  BF > 10 bar for one-cluster structure, so the originally-planned
  two-group comparison (one-cluster-favoring tables vs. the rest) couldn't
  run; the continuous version (Bayes factor / posterior probability of one
  cluster vs. strength-discrimination correlation) showed a weak positive
  relationship consistent with the eigenvalue-ratio cross-check, but on a
  small, divergence-thinned sample. Not run at full-batch scale, for the
  same runtime-cost reasons noted throughout this section for `bgms`.
- **Option A's uniform-GGM treatment of binary/ordinal data is an
  approximation this vignette inherited deliberately, not a claim it's the
  ideal model for IRW's data.** It's the same simplifying choice Huth et
  al. made in their own reanalysis, adopted here specifically so the
  headline-number comparison above is apples-to-apples with their
  benchmark. The Option A/B agreement check above is the direct test of
  how much that approximation costs for IRW's mostly-binary/ordinal data;
  where the two disagree, Option B (the ordinal MRF) is the theoretically
  better-suited model, and Option A should be read as an approximation to
  it, not the ground truth.
- **Option B ran on a small illustrative subset, not the full candidate
  pool.** Piloting found `bgms`'s ordinal MRF sampler costs roughly
  20-50x Option A's BGGM runtime per table; a full batch would take
  7-27 hours for compute that a full Option A batch does in well under
  two. The `r length(option_b_tables)`-table subset is enough to test
  whether the two models systematically disagree, but any specific
  IRW table not in that subset has only the Option A (GGM) result
  available.
- **The negative-edge / reverse-coding link is checked empirically, not
  validated against ground truth.** The "Is the negative-edge problem
  partly a sign-mismatch problem" section above shows negative
  discrimination and negative edge-sum concentrate on the same items far
  more than chance, and that a sign-invariant metric (`abs(a)` instead of
  signed `a`) recovers a noticeably better correspondence wherever a table
  has at least one negative-discrimination item. But **`irw_tags()`
  metadata is table-level, not item-level — there is no reverse-coding flag
  to confirm any specific item is actually reverse-keyed**, so this remains
  an inference from sign co-occurrence, not a validated audit. A firmer
  version would need item-level keying information IRW doesn't currently
  carry.
- **The joint dimensionality x local-dependence model (see "Do
  dimensionality and local dependence predict the correspondence jointly?"
  above) only used an interaction of two predictors and a linear
  specification.** The interaction is statistically real but small (R²
  0.065 to 0.074); a more flexible functional form (e.g. a GAM, or
  including the local-dependence vignette's own testlet-cluster count
  rather than just proportion-flagged) might capture more of the joint
  relationship than a two-way linear interaction does.
- **The full four-predictor decomposition (dimensionality, local
  dependence, stability, negative-edge share — see "Decomposing the gap"
  above) explains more variance than any pairwise comparison but still
  leaves most of it unaccounted for** (`r scales::percent(1 - r2_decomp, accuracy = 1)`
  of variance in the fitted sample). It's also purely observational
  bivariate/multivariate regression across tables, not a causal or
  simulation-based account of the mechanism — e.g. it can't distinguish
  "negative edges cause a worse correspondence" from "both are downstream
  of the same underlying multidimensionality/reverse-coding problem." A
  natural follow-up would be a small simulation (generate data under a
  known single-factor model, deliberately reverse-code a subset of items
  without recoding, refit both frameworks) to establish the mechanism
  directly rather than inferring it from cross-table correlations. Not run
  here.

## Reproducibility

{{< include ../components/_source-links.qmd >}}

```{r}
#| label: repro
#| echo: true
#| eval: false

# 1. Re-run the prerequisite compute scripts first (from the project root)
Rscript vignettes/dimensionality_compute.R
Rscript vignettes/local_dependence_compute.R

# 2. Re-run this vignette's compute script
Rscript vignettes/network_psych_compute.R

# 3. Ancillary analyses referenced above -- each is a standalone script, not
#    part of the main compute pipeline, and each writes a small .rds this
#    page reads directly (see file.exists() guards throughout):
Rscript vignettes/network_psych_empty_audit.R          # "Why are 7 networks empty?"
Rscript vignettes/network_psych_prior_sensitivity.R     # "How sensitive is the evidence classification to the prior?"
Rscript vignettes/network_psych_sbm_check.R             # "A network-native multidimensionality check"

# 4. Re-render this page
quarto render vignettes/network_psych.qmd
```

Session info from the compute run: R `r if (cache_ok) .cache$session$R.version$version.string else "N/A"`,
run on `r format(date_run, "%B %d, %Y")`.

## Acknowledgements

We are grateful to **Karoline Huth** (University of Amsterdam) for reviewing
an earlier version of this page in detail and sending line-numbered notes.
Several of the analyses above exist because of that review: the audit of the
empty-network tables, the stochastic-block-model multidimensionality check,
the matrix-F prior sensitivity analysis, and the comparison that splits tables
by overall network stability rather than trimming weak-evidence edges — the
last of these was her suggested alternative to what this page did before, and
is a cleaner test of the same question. She also corrected our framing of why
the ordinal Markov random field was not used, and caught a claim that both
model options shared the matrix-F prior when only one does.

This revised version incorporates her notes but was not itself reviewed by
her; the analytic choices, interpretations and any remaining errors are ours
alone.

## References

::: {#refs}
:::
