---
title: "What Is a Wrong Answer Worth?"
description: |
  Multiple-choice data usually reach a model as right or wrong. The IRW's
  nominal tables keep the option chosen. Does knowing which wrong answer a
  respondent picked help predict how they answer items we have not seen? We
  compare a 2PL on the 0/1 scores with the nominal and nested logit models on
  the options, out of sample, using the InterModel Vigorish (IMV), first for
  right/wrong and then for the option chosen.
date: today
bibliography: nominal_imv_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13) +
            theme(panel.grid.minor = element_blank(), legend.position = "top"))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"
col_yard <- irw_grey   # Rasch -> 2PL, the yardstick
col_nom  <- irw_blue   # -> nominal
col_nest <- irw_red    # -> nested logit

`%||%` <- function(a, b) if (is.null(a)) b else a

cache_path <- "nominal_imv_data/nominal_imv_results.rds"
cache_ok   <- file.exists(cache_path)
if (cache_ok) {
  .cache           <- readRDS(cache_path)
  res              <- .cache$results
  candidate_tables <- .cache$candidate_tables
  date_run         <- as.Date(.cache$date_run)
  is_pilot         <- isTRUE(.cache$pilot)
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_tables_run   <- length(candidate_tables)
} else {
  res <- list(); candidate_tables <- character()
  date_run <- Sys.Date(); is_pilot <- FALSE
  r_all_candidates <- 0L; r_n_tables_run <- 0L
}
fitted  <- Filter(function(r) is.na(r$skipped), res)
skipped <- Filter(function(r) !is.na(r$skipped), res)

LABELS <- c(
  `wilmer-rmet-normative-data-set-2022`  = "Reading the Mind in the Eyes (RMET)",
  `wilmer-mrmet-normative-data-set-2022` = "Multiracial RMET",
  borges_brazil_residency_2024_pbt       = "Residency exam (paper)",
  borges_brazil_residency_2024_cbt       = "Residency exam (computer)",
  choi_2020_ednet_listening              = "EdNet TOEIC listening",
  choi_2020_ednet_reading                = "EdNet TOEIC reading",
  vocabulary_iq                          = "Vocabulary IQ test (VIQT)",
  zorowitz_2023_marsib                   = "Matrix reasoning (MaRs-IB)",
  fitz_2024_numeracy                     = "Numeracy (Fitz)",
  psychtools_ability                     = "ICAR ability items",
  experimental_iq                        = "Experimental IQ items",
  myszkowski_2018_spmls                  = "Raven's SPM (last series)",
  suarez_2026_statistics                 = "Statistics exam (Suárez)",
  voropaeva_2026_health_media            = "Health media literacy",
  nomt_hooper_2024_study2                = "Novel object memory (NOMT)",
  papousek_2017_anatomy                  = "Anatom practice",
  geography_nom                          = "Geography practice",
  cifar10h                               = "CIFAR-10H image labels",
  blum_2018_imak                         = "IMAK"
)
label_of <- function(tn) {
  tn <- sub("_nom$", "", tn)
  if (grepl("^enem_", tn)) {
    p <- strsplit(tn, "_")[[1]]
    area <- c(ch = "humanities", cn = "natural sciences", lc = "languages", mt = "math")[p[4]]
    return(paste("ENEM", p[2], area))
  }
  if (tn %in% names(LABELS)) LABELS[[tn]] else tn
}
family_of <- function(tn) ifelse(grepl("^enem_", tn), "ENEM", "Other tests")

empty_summ <- tibble(table = character(), label = character(), family = character(),
                     n = integer(), n_full = integer(), items = integer(), p = numeric(),
                     dense = logical(), yard = numeric(), nom = numeric(), nest = numeric(),
                     cat_yard = numeric(), cat_nom = numeric(), cat_nest = numeric(),
                     wr_nom = numeric(), wr_nest = numeric(), nest_folds_pos = integer(),
                     cat_nest_folds_pos = integer(), wr_nest_folds_pos = integer())
summ <- bind_rows(empty_summ, lapply(fitted, function(r) {
  f <- r$imv_by_fold
  tibble(
    table = r$table, label = label_of(r$table), family = family_of(r$table),
    n = r$n_persons, n_full = r$n_full, items = r$n_items, p = r$p_correct,
    dense = !is.null(r$dense_block),
    yard = r$imv_all[["rasch_to_2pl"]], nom = r$imv_all[["twopl_to_nom"]],
    nest = r$imv_all[["twopl_to_nest"]],
    cat_yard = r$imv_all[["cat_rasch_to_2pl"]], cat_nom = r$imv_all[["cat_twopl_to_nom"]],
    cat_nest = r$imv_all[["cat_twopl_to_nest"]],
    wr_nom = r$imv_all[["wrong_shares_to_nom"]], wr_nest = r$imv_all[["wrong_shares_to_nest"]],
    nest_folds_pos = sum(f[, "twopl_to_nest"] > 0),
    cat_nest_folds_pos = sum(f[, "cat_twopl_to_nest"] > 0),
    wr_nest_folds_pos = sum(f[, "wrong_shares_to_nest"] > 0)
  )
})) |> mutate(share = nest / yard)

# Numbers quoted in the text.
f4  <- function(x) sprintf("%.4f", x)
pct <- function(x) scales::percent(x, 1)
and_list <- function(x) if (length(x) < 2) x else paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
n_tab     <- nrow(summ)
n_folds   <- 5 * n_tab
nom_wins  <- summ$label[summ$nom > summ$nest]
top_share <- slice_max(summ, share, n = 1)
low_share <- slice_min(summ, share, n = 1)
cat_over_yard <- summ$label[summ$cat_nest > summ$cat_yard]

folds <- terc <- tibble()
if (length(fitted)) {
  folds <- bind_rows(lapply(fitted, function(r)
    as_tibble(r$imv_by_fold) |>
      mutate(fold = row_number(), label = label_of(r$table), family = family_of(r$table))))

  terc <- bind_rows(lapply(fitted, function(r) {
    as.data.frame(r$imv_by_tercile) |>
      tibble::rownames_to_column("tercile") |>
      mutate(table = r$table, label = label_of(r$table), family = family_of(r$table))
  })) |> mutate(tercile = factor(tercile, c("low", "mid", "high"), c("Bottom", "Middle", "Top")))
}

# A dot per table with its fold-to-fold range, for any set of IMV columns.
dot_plot <- function(cols, names, colours, xlab) {
  pts <- summ |> select(label, family, all_of(names(cols))) |>
    pivot_longer(-c(label, family), names_to = "cmp", values_to = "imv")
  rng <- folds |> select(label, all_of(unname(cols))) |>
    pivot_longer(-label, names_to = "col", values_to = "v") |>
    group_by(label, col) |> summarise(lo = min(v), hi = max(v), .groups = "drop") |>
    mutate(cmp = names(cols)[match(col, cols)])
  d <- left_join(pts, rng, by = c("label", "cmp")) |>
    mutate(cmp = factor(cmp, names(cols), names))
  ord <- summ |> arrange(.data[[names(cols)[length(cols)]]]) |> pull(label)
  d$label <- factor(d$label, ord)
  ggplot(d, aes(imv, label, colour = cmp)) +
    geom_vline(xintercept = 0, colour = "grey80") +
    geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0, linewidth = 0.6,
                   position = position_dodge(width = 0.6)) +
    geom_point(size = 2.2, position = position_dodge(width = 0.6)) +
    facet_grid(family ~ ., scales = "free_y", space = "free_y") +
    scale_colour_manual(values = colours, name = NULL) +
    labs(x = xlab, y = NULL)
}
knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/nominal_imv_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: pilot-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && is_pilot) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, partial run.** This render uses ", r_n_tables_run,
  " pilot tables out of ", r_all_candidates, " candidates, to check the ",
  "pipeline and page layout. The numbers are real computed results, but treat ",
  "them as a small preview. Re-run `nominal_imv_compute.R` with ",
  "`PILOT <- FALSE` to analyse every candidate.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`nominal_imv_compute.R`) directly if you're relying
on the numbers here.
:::

## Motivation

Score a multiple-choice item 0/1 and two respondents who both got it wrong look the same, though one may have chosen the answer a nearly-right reader would choose and the other an answer nobody who understood the item would pick. The [course lesson on nominal models](https://ben-domingue.github.io/irw-course/lessons/nominal.html) shows that the wrong answers carry *information* about the respondents who miss items. It leaves open whether that information improves *prediction*. This page asks that across the multiple-choice tables in the IRW's nominal source, in two forms:

1. **Right or wrong.** Does a model that has seen which wrong answers a respondent chose predict better whether they will get an unseen item right?
2. **Which option.** Does it predict better *which option* they will choose, and in particular *which wrong answer*, given that they miss?

The IMV has been used on this site to ask the same kind of question about other information beyond the 0/1 scores. [Response time and accuracy](rt_imv.qmd) asks how much response times add to predictions of correctness over ability alone, and [Guessing in multiple-choice data](guessing.qmd) asks, on several of the same ENEM tests, whether modelling guessing helps out of sample. [Calculating IMV values](imv.qmd) walks through the basic computation.

## Data and methods

### Dataset selection

```{r}
#| label: candidate-tables
#| eval: false

cands <- c(
  paste0("enem_2013_1mil_", c("ch", "cn", "lc", "mt"), "_nom"),  # one ENEM test per area
  "wilmer-rmet-normative-data-set-2022_nom", "borges_brazil_residency_2024_pbt_nom",
  "choi_2020_ednet_listening_nom", "vocabulary_iq_nom", "psychtools_ability_nom"
  # ... and every other irw_nominal table with at least 10 items
)
```

The candidates are four ENEM tests, one per subject area from 2013, and every other table in `irw_nominal` with at least 10 items: `r r_all_candidates` tables in all, listed by hand in the compute script because the nominal metadata lags behind releases. A table is analysed if it is single-answer multiple choice: one keyed option per item, and no item with more than 12 answers each chosen by at least 1% of respondents or with its rarer answers together making up more than 10% of responses. Each table is cut to 3,000 respondents at most.

**Rare options.** A wrong option chosen by fewer than 1% of respondents is pooled with the item's other rare options. If the pooled group is still under 1% (ENEM's blank and double-mark codes, about 0.5% of responses), those responses are dropped. Respondents left with fewer than half the items are dropped, and so is any table with fewer than 1,000 respondents remaining.

**Sparse tables.** Some tables come from adaptive or self-paced systems, where each respondent sees a small part of a large item bank: EdNet's TOEIC practice app, for example. There we keep a dense block: the 40 most-answered items, and the respondents who answered at least half of them. These items are not a random draw from the bank, so the result describes the block. Such tables are marked † below.

### Models

Three item response models are fitted to every table, with the Rasch model alongside as a reference point.

- **2PL.** The usual model for right/wrong. It sees only the 0/1 scores.
- **Nominal response model** [@bock1972]. Every option, the key included, gets its own slope and intercept on θ. The probability of option $k$ is a softmax over the options' linear predictors.
- **Nested logit** [@suh2010]. A 2PL for the key, and a nominal model *among the distractors* for the respondents who miss. The key's curve keeps the 2PL's shape; the wrong answers only add information about who misses the item and how.

The course lesson walks through both option models on one test; this page does not re-derive them.

### Per-table computation

For each candidate table, `fit_one()` in the compute script:

1. Fetches the table with `irw_fetch(source = "nom")`, keeps one response per person and item, keeps a dense block if the table is sparse, and draws up to 3,000 respondents.
2. Builds two person × item matrices, one holding the option chosen (with the key coded as the last category) and one holding the 0/1 score. Rare options are pooled, and thin items and respondents are dropped, as described above.
3. Splits the observed person-item cells at random into five folds. For each fold, it fits the Rasch model and the 2PL to the 0/1 scores, and the nominal model and the nested logit to the options, using `mirt` with all the fold's cells hidden.
4. Takes each model's EAP θ, estimated from the visible cells, and predicts every hidden cell. It records two probabilities: P(key), and the probability of the option actually chosen.
5. Computes the three IMVs described below, pooled over all held-out cells, fold by fold, and within thirds of the 2PL θ.

```{r}
#| label: fit-function
#| eval: false

fit_one <- function(tn) {
  P <- prep(readRDS(cache_file(tn)))            # options (key last) and 0/1 matrices
  fold <- sample(rep_len(1:5, length(obs)))     # folds over observed person-item cells
  per_fold <- lapply(1:5, function(k) {
    opt_tr <- opt; opt_tr[obs[fold == k]] <- NA # hide this fold
    x_tr   <- x01; x_tr[obs[fold == k]]   <- NA
    mods <- list(
      rasch   = mirt(x_tr, 1, "Rasch"),
      twopl   = mirt(x_tr, 1, "2PL"),
      nominal = mirt(opt_tr, 1, "nominal", technical = list(NCYCLES = 2000)),
      nested  = mirt(opt_tr, 1, "2PLNRM", key = ncat, technical = list(NCYCLES = 2000))
    )
    # Each model's EAP theta -> P(key) and P(option chosen) on the hidden cells.
    # Rasch and the 2PL split 1 - P(key) by the training-cell distractor shares.
  })
  # imv.binary() on right/wrong; imv_cat() on the option chosen and on which wrong answer
}
```

Of the `r r_all_candidates` candidate tables, `r n_tab` produced usable results. The rest are listed, with the reason, under @tbl-results. `r if (all(sapply(fitted, function(r) all(r$converged)))) "Every fold's fit converged for all four models on every analysed table." else paste0("Some fold fits did not converge: ", paste(sapply(Filter(function(r) !all(r$converged), fitted), function(r) label_of(r$table)), collapse = ", "), ".")`

::: {.callout-note}
To reproduce with current IRW holdings, re-run `nominal_imv_compute.R`
and commit the updated `nominal_imv_data/nominal_imv_results.rds`.
:::

### Three versions of the IMV

The [InterModel Vigorish](https://www.cambridge.org/core/journals/psychometrika/article/intermodel-vigorish-as-a-lens-for-understanding-and-quantifying-the-value-of-item-response-models-for-dichotomously-coded-items/F61C75F6F945A5B13F73C6128EB83998) compares two models' out-of-sample predictions. For binary outcomes it converts each model's mean log-likelihood on the held-out cells into a *weighted coin*: the coin whose own expected log-likelihood equals the model's. If a model's predictions are as good as betting on a coin that lands heads with probability $w$, the IMV of model 1 over model 0 is the relative gain in that weight, $(w_1 - w_0)/w_0$. The Rasch → 2PL gain on the same cells is our yardstick: the value of letting items differ in discrimination, which most analysts already consider worth having.

**1. Right/wrong (binary IMV).** Every model predicts P(key) on the hidden cells, and `imv::imv.binary()` compares them. All four models are judged on the same outcome.

**2. The option chosen (multinomial IMV).** Here the outcome is which of the item's $K$ options the respondent chose. The coin becomes a $K$-sided die that lands on the chosen option with probability $w$ and on each of the others with probability $(1 - w)/(K - 1)$. We solve for the $w$ whose expected log-likelihood matches the model's,

$$
\frac{1}{n}\sum_{i=1}^{n} \Big[\, w \log w + (1 - w) \log\frac{1 - w}{K_i - 1} \Big] \;=\; \frac{1}{n}\sum_{i=1}^{n} \log \hat p_i,
$$

where $\hat p_i$ is the probability the model gave the option actually chosen in held-out cell $i$ and $K_i$ is the number of options on that cell's item. The IMV is again $(w_1 - w_0)/w_0$. With two options everywhere this is exactly the binary IMV. Because $K$ differs between tests (four options on RMET, five on ENEM, fewer where rare options were pooled), the multinomial IMV is best compared across models within a table, not across tables.

This is not the only way to extend the IMV beyond two outcomes. @kanopka2023 [ch. 3] builds two polytomous IMVs out of binary bets. One, $\omega_t$, bets on whether a response falls at or below each threshold category. The other, $\omega_c$, bets on which of each pair of categories was chosen, given that it was one of the two, and averages the pairwise IMVs within items, weighting by how often each pair occurs. Because every piece is an ordinary binary IMV, both stay on the dichotomous IMV's scale, so values can be compared across tests whatever the number of categories. We depart from them for three reasons. First, $\omega_t$ needs ordered categories, and distractors have no order. Second, the question here is a single bet, which of $K$ options is chosen, and the die scores the whole predicted distribution with one held-out log-likelihood. $\omega_c$ would instead estimate a separate IMV for every pair of options on every item, and with 3,000 respondents and distractors chosen by as few as 1% of them, many of those pairs rest on a handful of held-out responses. Third, the die splits cleanly into the option chosen and, below, which wrong answer was chosen; the pairs that set the key against a distractor would bring the right/wrong signal back into $\omega_c$. The cost is the one their construction avoids: the die's scale depends on $K$, which is why we compare models within a table rather than across tables.

A 2PL predicts only P(key), so to enter this comparison it needs a rule for the wrong options. We use the simplest: **2PL + shares**. The 2PL gives P(key), and 1 − P(key) is split among the wrong options in the proportions in which they were chosen on that item's training cells. That baseline already knows which distractors are popular. What it lacks, and what the option models add, is any link between θ and the choice of distractor.

**3. Which wrong answer.** The third version keeps only the held-out cells answered *wrongly* and asks which wrong option was chosen, among the $K - 1$. Each model's prediction is its probability of the chosen option divided by its P(wrong). For the 2PL + shares baseline that is just the share, so this IMV measures one thing: how much knowing θ tells you about *which* wrong answer someone picks.

## Results

### Does knowing the wrong answers predict right/wrong better?

All three comparisons in @fig-gains are the same bet: on each held-out cell, which model gave the better probability that the respondent chose the key? Each model makes that prediction from its own θ, estimated from the respondent's visible cells. The 2PL and Rasch read P(key) straight off their item curves. The nominal model and the nested logit predict a probability for every option, and we keep only the key's; all the wrong options together make up "incorrect". So the option models are scored on right/wrong exactly as the 2PL is.

Moving from the 2PL to an option model changes two things at once:

- **θ.** The option models estimate θ from which wrong answers a respondent chose, not just whether they were wrong, so two people with the same score can get different θs.
- **The key's curve.** In the nominal model the key is one category of a softmax over all the options, so its curve depends on every distractor's slope and intercept and need not have the 2PL's shape. The nested logit keeps the 2PL's curve for the key and uses the distractors only below it.

So the orange dot (2PL → nested logit) is close to the value of the better θ alone. The blue dot (2PL → nominal) is the better θ plus whatever re-drawing the key's curve adds or costs. The gap between them is the price, or payoff, of that new curve.

```{r}
#| label: fig-gains
#| fig-cap: "Out-of-sample binary IMV for three model comparisons. Blue: the nominal model changes both θ and the key's curve. Orange: the nested logit changes θ but keeps the 2PL's curve for the key. Points pool the five folds; bars span the fold-to-fold range."
#| fig-height: 8
#| fig-width: 9

dot_plot(c(yard = "rasch_to_2pl", nom = "twopl_to_nom", nest = "twopl_to_nest"),
         c("Rasch → 2PL (yardstick)", "2PL → nominal", "2PL → nested logit"),
         c(col_yard, col_nom, col_nest), "IMV for held-out right/wrong")
```

Knowing which wrong answers a respondent chose elsewhere always helps, but only a little. The nested logit beats the 2PL on all `r n_tab` tables and in `r sum(summ$nest_folds_pos)` of `r n_folds` folds, with IMVs from `r f4(min(summ$nest))` to `r f4(max(summ$nest))`. Every one of those is smaller than the Rasch → 2PL gain on the same cells.

The bars show the spread across folds, which understates the uncertainty. Refitting with a different random split into folds moved some of these IMVs by up to about 0.001 (ENEM humanities, 2PL → nested, went from .0041 to .0035). Read differences smaller than that as ties.

The nested logit also matches or beats the nominal model on `r sum(summ$nest >= summ$nom)` of the `r n_tab` tables`r if (length(nom_wins)) paste0(" (the exception", if (length(nom_wins) > 1) "s are " else " is ", and_list(nom_wins), ", by a hair)") else ""`. Apart from that exception, blue sits at or left of orange, so re-drawing the key's curve mostly costs accuracy rather than adding it. The cost is clearest on EdNet listening and the computer-based residency exam, where the nominal model spends most of what the distractors add to θ on a noisier curve for the key. For predicting right and wrong, the better way to use the distractors is to let them inform θ and leave the key's curve alone.

How far orange gets toward grey varies a lot. The nested logit's gain is `r pct(top_share$share)` of the Rasch → 2PL gain on `r top_share$label` but only `r pct(low_share$share)` on the `r low_share$label`. The "Nested ÷ yardstick" column of @tbl-results gives it for every table; it has no clear relation to how hard the test is. It is lowest on the two residency exams. These are the longest tests here, at around 100 items, so a likely reason is that θ is already well estimated from the right/wrong scores alone, leaving the distractors little to add.

### Does it predict which option is chosen?

Now the bet changes. Instead of right or wrong, the outcome on each held-out cell is the option chosen, scored with the multinomial IMV. The 2PL and Rasch enter with the "+ shares" rule: their P(key), with the rest split among the distractors by how popular each was on the training cells. Grey is again the yardstick, here the Rasch → 2PL gain on the same multinomial scale. The left panel of @fig-option scores every held-out cell; the right panel, described below, scores only the wrong answers.

```{r}
#| label: fig-option
#| fig-cap: "Out-of-sample multinomial IMV. Left: the option chosen on every held-out cell, with 2PL + shares as the baseline for the option models. Right: which wrong answer was chosen, over held-out wrong answers only, with the distractors' training shares as the baseline. Points pool the five folds; bars span the fold-to-fold range. The two panels have different scales."
#| fig-height: 8
#| fig-width: 11

panels <- list(
  `Option chosen (vs. 2PL + shares)`         = c(yard = "cat_rasch_to_2pl", nom = "cat_twopl_to_nom", nest = "cat_twopl_to_nest"),
  `Which wrong answer (vs. distractor shares)` = c(nom = "wrong_shares_to_nom", nest = "wrong_shares_to_nest")
)
d <- bind_rows(lapply(names(panels), function(pn) {
  cols <- panels[[pn]]
  rng <- folds |> select(label, family, all_of(unname(cols))) |>
    pivot_longer(-c(label, family), names_to = "col", values_to = "v") |>
    group_by(label, family, col) |>
    summarise(lo = min(v), hi = max(v), .groups = "drop")
  pts <- bind_rows(lapply(fitted, function(r) tibble(
    label = label_of(r$table), col = unname(cols), imv = r$imv_all[unname(cols)])))
  left_join(rng, pts, by = c("label", "col")) |>
    mutate(panel = pn, cmp = names(cols)[match(col, cols)])
})) |>
  mutate(panel = factor(panel, names(panels)),
         cmp = factor(cmp, c("yard", "nom", "nest"),
                      c("Rasch → 2PL (yardstick)", "→ nominal", "→ nested logit")),
         label = factor(label, summ |> arrange(wr_nest) |> pull(label)))

ggplot(d, aes(imv, label, colour = cmp)) +
  geom_vline(xintercept = 0, colour = "grey80") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0, linewidth = 0.6,
                 position = position_dodge(width = 0.6)) +
  geom_point(size = 2.2, position = position_dodge(width = 0.6)) +
  facet_grid(family ~ panel, scales = "free", space = "free_y") +
  scale_colour_manual(values = c(col_yard, col_nom, col_nest), name = NULL) +
  labs(x = "IMV", y = NULL)
```

Judged on the option chosen, the option models beat 2PL + shares on every table. That is not guaranteed, because the baseline already knows how popular each distractor is. On `r length(cat_over_yard)` of the `r n_tab` tables (`r and_list(cat_over_yard)`), the nested logit's gain over 2PL + shares is larger than the gain from adding discrimination on the same scale.

That gain mixes two things: predicting the key better, which the first section already measured, and predicting which distractor is chosen. The right panel of @fig-option isolates the second. It keeps only the held-out cells answered wrongly and asks which wrong answer was chosen. The baseline is the distractors' shares alone, which know nothing about θ, so this IMV measures what knowing θ tells you about the choice of distractor.

Among held-out wrong answers, θ predicts which one was chosen on every table, with IMVs from `r f4(min(summ$wr_nest))` to `r f4(max(summ$wr_nest))`. That is a median `r round(median(summ$wr_nest / summ$nest))` times the right/wrong gain on the same table. The two IMVs are for different outcomes, so the ratio is only a rough guide, but the gap is too large to be an artefact of scale. The nominal model and the nested logit are indistinguishable here, as expected, since both model the choice among distractors the same way. The wrong answers carry far more information about *which wrong answer* comes next than about *whether* the next answer is wrong. They enter a right/wrong prediction only through θ, and θ is already well estimated from the scores.

### Who gains?

The next two figures split the held-out cells into thirds of the 2PL θ and compute each IMV within each third. Each line is one table. Click a line, or choose a table from the menu above the figure, to pick it out from the rest; double-click to clear. Hover over a point for its table and value.

```{r}
#| label: terc-plot-fn
#| echo: false

# One line per table across the thirds of theta, with a menu (and click) to
# pick out one table; the others fade.
terc_plot <- function(ycol, ylab, group) {
  d <- terc |> mutate(y = .data[[ycol]],
                      text = sprintf("%s<br>%s third: %.4f", label, tercile, y))
  sd <- highlight_key(d, ~label, group = group)
  p <- ggplot(sd, aes(tercile, y, group = label, colour = family, text = text)) +
    geom_hline(yintercept = 0, colour = "grey80") +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    scale_colour_manual(values = c(ENEM = irw_red, `Other tests` = irw_blue), name = NULL) +
    labs(x = "Third of the 2PL θ", y = ylab)
  ggplotly(p, tooltip = "text") |>
    highlight(on = "plotly_click", off = "plotly_doubleclick", selectize = TRUE,
              persistent = FALSE, opacityDim = 0.12) |>
    layout(legend = list(orientation = "h", y = 1.08))
}
```

```{r}
#| label: fig-terciles
#| fig-cap: "Binary IMV, 2PL → nested logit, within thirds of the 2PL θ. One line per table."

terc_plot("twopl_to_nest", "IMV, 2PL → nested logit", "Pick out a table")
```

For right/wrong, where the gain lands depends on the test. On most tables it goes to the bottom third, as the course lesson's information curves predict: low scorers are the respondents with wrong answers to read. ENEM never shows that pattern. On humanities and languages the gain peaks in the middle third. On the two hardest tests (natural sciences and math, about a quarter of responses correct), it rises toward the top, and the bottom third gains least. Low scorers who guess close to at random would produce this pattern, because their wrong answers carry no signal. [Issue #284](https://github.com/datapages/irw/issues/284) tests this directly.

```{r}
#| label: fig-terciles-wrong
#| fig-cap: "Which wrong answer: multinomial IMV, shares → nested logit, over held-out wrong answers within thirds of the 2PL θ. One line per table."

terc_plot("wrong_shares_to_nest", "IMV, shares → nested logit", "Pick out a table ")
```

For which wrong answer, the middle third gains least on every table. On most tables both ends gain a lot. The exceptions are again the two hardest ENEM tests, natural sciences and math, where the top third gains most and the bottom third much less, as near-random guessing at the bottom would predict. The dip in the middle is partly built in. The shares are the distractors' popularity averaged over everyone who missed, so they describe a typical wrong answer, from the middle of the distribution, best. The rest is the nominal model's crossing option curves at work. If low scorers who miss are drawn to one distractor and high scorers who miss to another, then knowing which end of the scale a respondent sits at tells you which to expect. Which distractors those are, and why they pull, is the question of [issue #285](https://github.com/datapages/irw/issues/285).

### Every table

```{r}
#| label: tbl-results
summ |>
  arrange(family, desc(nest)) |>
  transmute(Table = paste0(label, ifelse(dense, " †", "")), N = format(n, big.mark = ","),
            Items = items, `% correct` = pct(p),
            `Rasch → 2PL` = f4(yard), `2PL → nested` = f4(nest), `Nested ÷ yardstick` = pct(share),
            `Option: 2PL + shares → nested` = f4(cat_nest),
            `Wrong answer: shares → nested` = f4(wr_nest)) |>
  knitr::kable(align = "lrrrrrrrr")
```

† A dense block of a sparse table: the 40 most-answered items and the respondents who answered at least half of them.

```{r}
#| label: not-analysed
#| output: asis
reason <- function(x) dplyr::case_when(
  grepl("^N = ", x)                  ~ "Fewer than 1,000 respondents",
  grepl("^sparse design", x)         ~ "Sparse design with no dense block (each respondent saw a small part of a large item bank)",
  grepl("option set changes", x)     ~ "Option set changes from one presentation of an item to the next (adaptive practice)",
  grepl("free text|not one keyed|resp is not 0/1|no resp/text", x) ~
    "Not single-answer multiple choice (free text, typed numbers, multi-select, or unscored)",
  grepl("MIN_ITEMS", x)              ~ "Fewer than 10 usable items (an item needs at least two wrong answers each chosen by 1% or more)",
  TRUE ~ x)
sk <- tibble(table = sapply(skipped, `[[`, "table"), why = reason(sapply(skipped, `[[`, "skipped")))
cat("**Not analysed.** Four tables have fewer than 10 items and were not attempted (`cos101_2026`, ",
    "`goldberg_2018_spa_computer_use`, `himmelstein-berlin_numeracy-2025`, `mthimkhulu_2023_pirls_reading`). ",
    "The rest of the nominal source was tried and set aside:\n\n", sep = "")
for (w in unique(sk$why)) cat("- ", w, ": ", paste0("`", sub("_nom$", "", sk$table[sk$why == w]), "`", collapse = ", "), "\n", sep = "")
```

## Limitations

**One sample of respondents, one split into folds.** Each table is cut to 3,000 respondents, and the folds are drawn once. Refitting with a different split moved some IMVs by up to about 0.001, more than the fold-to-fold bars suggest. Differences smaller than that are ties.

**Four ENEM tests stand in for 52.** We used one 2013 test per subject area. The differences between areas are large here; whether they hold in other years is untested.

**Dense blocks are not random samples of items.** For EdNet, the 40 most-answered questions and the learners who answered at least half of them are a particular slice of the app's use. The EdNet results describe that block, not the whole item bank.

**Rare answers are pooled or dropped.** Options chosen by fewer than 1% of respondents are pooled, and ENEM's blanks and double marks are dropped. An unusual distractor may be exactly where a wrong answer says the most, and these IMVs never score it.

**The which-wrong-answer baseline knows nothing about θ.** The large IMVs in the right panel of @fig-option show that θ is informative about which distractor is chosen. They do not show that the nominal model beats other ways of using ability, such as distractor shares computed separately by score band.

**The multinomial IMV depends on the number of options.** Its scale changes with $K$, so compare models within a table rather than tables with each other. It is also not the polytomous IMV of @kanopka2023; see above for how the two differ.

**"Who gains" groups respondents by the 2PL θ.** Grouping by an option model's θ would move some respondents between thirds and could change the shape of the curves at the margins.

**Only unidimensional models, and no guessing.** We did not fit a 3PL or any other guessing model; the [guessing vignette](guessing.qmd) covers those on ENEM.

**Cross-validation is over cells, not people.** The held-out cells belong to respondents whose other answers the models have seen, so these IMVs measure predicting a missing response from a known respondent. Scoring new respondents from a few items could show a different gain.

## Reproducibility

{{< include ../components/_source-links.qmd >}}

{{< include ../components/_data-version.qmd >}}

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the compute script (from the project root). It caches fetched
#    tables in ~/.cache/irw_nominal_imv and per-table results in
#    vignettes/nominal_imv_data/fits/, so an interrupted run resumes.
source("vignettes/nominal_imv_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/nominal_imv.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(summ$table, output_file = "nominal_imv_data/references.bib", source = "nom")
```
:::

---

## References

::: {#refs}
:::
