---
title: "Polytomous IRT: Binomial vs. Native Models"
description: |
  How do parsimonious binomial IRT models compare to native polytomous IRT models
  (GPCM, GRM, PCM, sequential) when evaluated by cross-validated prediction error?
  We fit seven models to each of a diverse set of IRW polytomous datasets and compare
  out-of-sample RMSE.
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(plotly)
})

results  <- readRDS("poly_compare_data/multi_results.rds")
all_rmse <- do.call(rbind, lapply(results, `[[`, "rmse"))

meta <- do.call(rbind, lapply(names(results), function(nm) {
  r <- results[[nm]]
  data.frame(dataset = nm, K_range = r$K_range, ni = r$ni, N = r$N,
             stringsAsFactors = FALSE)
}))

# Model ordering for plots
model_levels <- c("PCM", "GPCM", "GRM", "Tutz", "1PL+logit", "2PL+logit", "1PL+cloglog")
model_type   <- c(PCM = "Native", GPCM = "Native", GRM = "Native", Tutz = "Native",
                  `1PL+logit` = "Binomial", `2PL+logit` = "Binomial",
                  `1PL+cloglog` = "Binomial")

all_rmse <- all_rmse |>
  mutate(
    model     = factor(model, levels = model_levels),
    mod_type  = model_type[as.character(model)],
    short_name = gsub("_", " ", dataset) |> stringr::str_trunc(35)
  )

# Dataset-level: median RMSE across models for ordering
ds_med <- all_rmse |>
  group_by(dataset, short_name) |>
  summarise(med_rmse = median(rmse), .groups = "drop") |>
  arrange(med_rmse)

all_rmse <- all_rmse |>
  mutate(short_name = factor(short_name, levels = ds_med$short_name))

# Winner per dataset
winners <- all_rmse |>
  group_by(dataset) |>
  slice_min(rmse, n = 1, with_ties = FALSE) |>
  ungroup()

# Mean rank
ranks <- all_rmse |>
  group_by(dataset) |>
  mutate(rank = rank(rmse, ties.method = "min")) |>
  ungroup()

mean_ranks <- ranks |>
  group_by(model, mod_type) |>
  summarise(mean_rank = mean(rank), .groups = "drop") |>
  arrange(mean_rank)
```

## Research question

Polytomous IRT models (GPCM, GRM, PCM, sequential) are the standard tools for ordered
responses with three or more categories. But an alternative strategy is to treat the
response as a **binomial count** — fitting a binomial distribution B(*K*−1, *p*) where
*p* is a function of the latent trait. Binomial IRT models are parsimonious (two
parameters per item regardless of *K*) and can use different link functions (logit,
log-log).

The IRW makes a large-scale empirical test straightforward: **do binomial models match
or beat native polytomous IRT when evaluated by out-of-sample prediction error?** And
does the choice of link function matter?

---

## Approach

### Models

We compare seven models across two families:

**Native polytomous** (one set of parameters per response category):

- **PCM** — Partial Credit Model (Rasch-constrained GPCM; discrimination fixed at 1)
- **GPCM** — Generalized Partial Credit Model (free discrimination)
- **GRM** — Graded Response Model (cumulative logistic)
- **Tutz** — Sequential (adjacent-category) model

**Binomial IRT** (two parameters per item; response treated as B(*K*−1, *p*)):

- **1PL+logit** — logistic link, discrimination fixed
- **2PL+logit** — logistic link, free discrimination
- **1PL+cloglog** — complementary log-log link, discrimination fixed

All models are fit with `mirt`. The custom binomial items use `mirt::createItem()`.

### Datasets

We used `irw_filter(n_categories = c(3, 6))` to identify IRW datasets with 3–6
ordered response categories. Additional filters: N ≥ 200, ≥ 3 items, all items share
the same *K*. Datasets larger than 5,000 respondents were downsampled.

```{r}
#| label: selection
#| eval: false

library(irw)
candidates <- irw_filter(n_categories = c(3, 6))
```

### Cross-validation

Each dataset was split 80/20 by person. Models were fit on the training set; predictions
were generated on the held-out test set using expected scores E[*Y*|θ̂] / (*K*−1),
putting all models on a common [0, 1] scale. RMSE was computed over person–item
pairs with non-missing test responses.

The full compute script is at `vignettes/poly_compare_multi_compute.R`.

---

## Results

### Dataset overview

```{r}
#| label: tbl-datasets
#| tbl-cap: "Polytomous datasets included in the analysis."

meta |>
  left_join(
    winners |> select(dataset, best_model = model),
    by = "dataset"
  ) |>
  rename(Dataset = dataset, `K range` = K_range, Items = ni,
         `N (train+test)` = N, `Best model` = best_model) |>
  arrange(`K range`, Items)
```

### RMSE by dataset and model

```{r}
#| label: fig-rmse-dot
#| fig-cap: "Cross-validated RMSE for each model on each dataset. Datasets ordered by median RMSE across models. Native models in blue, binomial models in orange."
#| fig-height: 6.5

pal <- c(Native = "#2166ac", Binomial = "#d6604d")

p <- ggplot(all_rmse, aes(
    x    = rmse,
    y    = short_name,
    colour = mod_type,
    shape  = model,
    text = paste0(
      "Dataset: ", dataset, "<br>",
      "Model: ", model, "<br>",
      "RMSE: ", round(rmse, 4), "<br>",
      "K: ", meta$K_range[match(dataset, meta$dataset)],
      "  items: ", meta$ni[match(dataset, meta$dataset)],
      "  N: ", meta$N[match(dataset, meta$dataset)]
    )
  )) +
  geom_point(size = 2.5, alpha = 0.85,
             position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = pal, name = "Family") +
  scale_shape_manual(values = c(16, 17, 15, 3, 1, 2, 4), name = "Model") +
  labs(x = "CV RMSE (rescaled response)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))

ggplotly(p, tooltip = "text") |>
  layout(
    annotations = list(list(
      text      = "80/20 train/test split. Lower RMSE = better prediction.",
      showarrow = FALSE,
      xref = "paper", yref = "paper",
      x = 0, y = -0.07,
      xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### Model rankings

```{r}
#| label: fig-ranks
#| fig-cap: "Mean rank across all datasets (rank 1 = best). Error bars show ±1 SE."
#| fig-height: 3.5

rank_se <- ranks |>
  group_by(model, mod_type) |>
  summarise(mean_rank = mean(rank), se = sd(rank) / sqrt(n()), .groups = "drop")

ggplot(rank_se, aes(
    x = reorder(model, mean_rank),
    y = mean_rank,
    fill = mod_type,
    ymin = mean_rank - se,
    ymax = mean_rank + se
  )) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(width = 0.25, colour = "grey40") +
  scale_fill_manual(values = pal, name = "Family") +
  coord_flip() +
  labs(x = NULL, y = "Mean rank (lower = better)") +
  theme_minimal(base_size = 13)
```

### GPCM vs. 2PL+logit

The two best-performing models — GPCM (best mean rank) and 2PL+logit (most wins) — are
the flexible members of each family. How closely do they track each other?

```{r}
#| label: fig-gpcm-vs-2pl
#| fig-cap: "Per-dataset RMSE for GPCM vs. 2PL+logit. Points above the diagonal: GPCM fits better; below: 2PL+logit fits better. Point size reflects number of items."
#| fig-height: 4.5

comp <- all_rmse |>
  filter(model %in% c("GPCM", "2PL+logit")) |>
  select(dataset, model, rmse) |>
  pivot_wider(names_from = model, values_from = rmse) |>
  left_join(meta, by = "dataset") |>
  mutate(tooltip = paste0(
    "Dataset: ", dataset, "<br>",
    "GPCM: ", round(GPCM, 4), "<br>",
    "2PL+logit: ", round(`2PL+logit`, 4), "<br>",
    "K: ", K_range, "  items: ", ni, "  N: ", N
  ))

p3 <- ggplot(comp, aes(x = `2PL+logit`, y = GPCM, size = ni, text = tooltip)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(colour = "#2166ac", alpha = 0.75) +
  scale_size_continuous(range = c(3, 10), name = "# items") +
  labs(x = "CV RMSE — 2PL+logit", y = "CV RMSE — GPCM") +
  theme_minimal(base_size = 13)

ggplotly(p3, tooltip = "text") |>
  layout(
    annotations = list(list(
      text      = "Dashed line = equal performance. Above diagonal: GPCM better.",
      showarrow = FALSE,
      xref = "paper", yref = "paper",
      x = 0, y = -0.12,
      xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

---

## What to notice

**Discrimination matters more than the model family.** The clearest pattern in the
rankings is the within-family split: GPCM (rank 2.5) substantially beats PCM (rank 4.1),
and 2PL+logit (rank 2.8) substantially beats 1PL+logit (rank 4.9). Freeing the
discrimination parameter consistently improves predictions — whether you are inside the
native polytomous family or the binomial family.

**Binomial models are competitive.** 2PL+logit wins on `r sum(winners$model == "2PL+logit")`
of `r nrow(winners)` datasets, compared with `r sum(winners$model == "GPCM")` for
GPCM. The two models track each other closely across datasets (see the scatter above),
with neither dominating. Given that the binomial model uses just 2 parameters per item
regardless of *K*, this is a favorable result for parsimony.

**The cloglog link sometimes wins.** 1PL+cloglog takes top spot on
`r sum(winners$model == "1PL+cloglog")` datasets despite being a constrained 1-parameter
model. This is unexpected and may reflect an asymmetric response distribution in those
instruments — the cloglog link has a longer right tail than the logistic, so it fits
well when low-category responses are rare.

**GRM lags GPCM.** Among the native polytomous models, GRM (mean rank 3.9) trails
GPCM (2.5) despite also estimating discrimination. The cumulative logit parameterisation
appears to be less competitive in practice on these datasets than the adjacent-category
approach.

---

## Extending this vignette

- **Effect sizes.** The RMSE differences are often small. Pair them with a permutation
  test or bootstrap CI to assess whether GPCM's edge over 2PL+logit is reliable.

- **More link functions.** Add probit and cauchit binomial variants to map out the full
  link-function landscape.

- **IMV.** Replace (or supplement) RMSE with Inter-Model Vigorish to directly quantify
  how much each model gains over a reference baseline.

- **Scale-up.** Bump `N_DATASETS` in `poly_compare_multi_compute.R` and re-run to
  cover more of the IRW polytomous catalogue.

---

## Reproducibility

Results were computed on 2026-05-28 using `r nrow(meta)` IRW tables. To reproduce:

```r
# From project root
Rscript vignettes/poly_compare_multi_compute.R
quarto::quarto_render("vignettes/poly_compare.qmd")
```

::: {.callout-note}
The compute script caches each dataset in `vignettes/poly_compare_data/`. Delete the
per-dataset `.rds` files to force a fresh run; delete `multi_results.rds` to
re-combine from scratch.
:::
