---
title: "Polytomous IRT: Binomial vs. Standard Polytomous Models"
description: |
  How do parsimonious binomial IRT models compare to standard polytomous IRT models
  (GPCM, GRM, PCM, sequential) when evaluated by cross-validated prediction error and
  Inter-Model Vigorish? We fit seven models to each of a diverse set of IRW polytomous
  datasets and compare out-of-sample RMSE, IMV vs. a prevalence baseline, and direct
  pairwise IMV between the best models of each family.
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(plotly)
  library(mirt)
})

fs       <- list.files("poly_compare_data", pattern = "^multi_.+\\.rds$", full.names = TRUE)
fs       <- fs[!grepl("multi_(results|summary)\\.rds$", fs)]
results  <- Filter(Negate(is.null), lapply(fs, readRDS))
names(results) <- sub("^multi_", "", sub("\\.rds$", "", basename(fs)))
# backward compat: old files store K_range, new files store K
results  <- lapply(results, function(r) {
  if (is.null(r[["K"]])) r$K <- r$K_range + 1L
  r
})
results  <- Filter(function(r) r$K >= 3, results)  # drop binary (K < 3)
all_rmse     <- do.call(rbind, lapply(results, `[[`, "rmse"))
all_imv      <- do.call(rbind, Filter(Negate(is.null), lapply(results, `[[`, "imv")))
all_pair     <- do.call(rbind, Filter(Negate(is.null), lapply(results, `[[`, "pair_imv")))
all_imv_1pl  <- do.call(rbind, Filter(Negate(is.null), lapply(results, `[[`, "imv_1pl")))
all_asym     <- do.call(rbind, Filter(Negate(is.null), lapply(results, `[[`, "asym")))
all_cat_decomp <- do.call(rbind, Filter(Negate(is.null), lapply(results, `[[`, "cat_decomp")))

meta <- do.call(rbind, lapply(names(results), function(nm) {
  r <- results[[nm]]
  data.frame(dataset = nm, K = r$K, ni = r$ni, N = r$N,
             stringsAsFactors = FALSE)
}))

# Model ordering for plots
model_levels <- c("PCM", "GPCM", "GRM", "Tutz",
                  "Binom 1PL (logit)", "Binom 2PL (logit)", "Binom 1PL (cloglog)",
                  "NP spline")
model_type   <- c(PCM = "Standard", GPCM = "Standard", GRM = "Standard", Tutz = "Standard",
                  `Binom 1PL (logit)` = "Binomial", `Binom 2PL (logit)` = "Binomial",
                  `Binom 1PL (cloglog)` = "Binomial", `NP spline` = "Binomial")

all_rmse <- all_rmse |>
  mutate(
    model     = factor(model, levels = model_levels),
    mod_type  = model_type[as.character(model)],
    short_name = gsub("_", " ", dataset) |> stringr::str_trunc(35)
  )

# ensure short_name is unique per dataset (truncation can collapse distinct names)
sn_map <- all_rmse |>
  distinct(dataset, short_name) |>
  mutate(short_name = make.unique(short_name, sep = " "))
all_rmse <- all_rmse |>
  select(-short_name) |>
  left_join(sn_map, by = "dataset")

# Dataset-level: median RMSE across models for ordering
ds_med <- all_rmse |>
  group_by(dataset, short_name) |>
  summarise(med_rmse = median(rmse), .groups = "drop") |>
  arrange(med_rmse)

all_rmse <- all_rmse |>
  mutate(short_name = factor(short_name, levels = ds_med$short_name))

# Winner per dataset
winners <- all_rmse |>
  group_by(dataset) |>
  slice_min(rmse, n = 1, with_ties = FALSE) |>
  ungroup()

# Mean rank
ranks <- all_rmse |>
  group_by(dataset) |>
  mutate(rank = rank(rmse, ties.method = "min")) |>
  ungroup()

mean_ranks <- ranks |>
  group_by(model, mod_type) |>
  summarise(mean_rank = mean(rank), .groups = "drop") |>
  arrange(mean_rank)

# Head-to-head: GPCM vs Binom 2PL (logit)
h2h <- all_rmse |>
  filter(model %in% c("GPCM", "Binom 2PL (logit)")) |>
  select(dataset, model, rmse) |>
  tidyr::pivot_wider(names_from = model, values_from = rmse)
n_gpcm_beats_2pl <- sum(h2h$GPCM < h2h$`Binom 2PL (logit)`, na.rm = TRUE)

# IMV summaries
all_imv <- all_imv |>
  mutate(model    = factor(model, levels = model_levels),
         mod_type = model_type[as.character(model)])

mean_imv <- all_imv |>
  group_by(model, mod_type) |>
  summarise(mean_imv_c = mean(imv_c, na.rm = TRUE),
            se_imv_c   = sd(imv_c,   na.rm = TRUE) / sqrt(sum(!is.na(imv_c))),
            mean_imv_t = mean(imv_t, na.rm = TRUE),
            se_imv_t   = sd(imv_t,   na.rm = TRUE) / sqrt(sum(!is.na(imv_t))),
            .groups = "drop")

all_imv_1pl <- all_imv_1pl |>
  mutate(model    = factor(model, levels = model_levels),
         mod_type = model_type[as.character(model)])

mean_imv_1pl <- all_imv_1pl |>
  group_by(model, mod_type) |>
  summarise(mean_imv_c = mean(imv_c, na.rm = TRUE),
            se_imv_c   = sd(imv_c,   na.rm = TRUE) / sqrt(sum(!is.na(imv_c))),
            mean_imv_t = mean(imv_t, na.rm = TRUE),
            se_imv_t   = sd(imv_t,   na.rm = TRUE) / sqrt(sum(!is.na(imv_t))),
            .groups = "drop")

# Per-dataset GPCM vs Binom 2PL (logit): RMSE + pairwise IMV
h2h_imv <- h2h |>
  left_join(all_pair |> filter(comparison == "Binom 2PL (logit)→GPCM") |>
              select(dataset, imv_c, imv_t),
            by = "dataset") |>
  left_join(meta |> select(dataset, K, ni, N), by = "dataset") |>
  arrange(GPCM) |>
  mutate(tooltip = paste0(
    "Dataset: ", dataset, "<br>",
    "RMSE GPCM: ", round(GPCM, 4), "<br>",
    "RMSE Binom 2PL (logit): ", round(`Binom 2PL (logit)`, 4), "<br>",
    "IMV_c: ", round(imv_c, 4), "<br>",
    "IMV_t: ", round(imv_t, 4), "<br>",
    "K: ", K, "  items: ", ni, "  N: ", N
  ))
```

## Research question

Polytomous IRT models (GPCM, GRM, PCM, sequential) are the standard tools for ordered
responses with three or more categories. But an alternative strategy is to treat the
response as a **binomial count** — fitting a binomial distribution B(*K*−1, *p*) where
*p* is a function of the latent trait. Binomial IRT models are parsimonious (two
parameters per item regardless of *K*) and can use different link functions (logit,
log-log).

The IRW makes a large-scale empirical test straightforward: **do binomial models match
or beat standard polytomous IRT when evaluated by out-of-sample prediction error?** And
does the choice of link function matter?

---

## Approach

### Models

We compare seven models across two families:

**Standard polytomous** (one set of parameters per response category):

- **PCM** — Partial Credit Model (Rasch-constrained GPCM; discrimination fixed at 1)
- **GPCM** — Generalized Partial Credit Model (free discrimination)
- **GRM** — Graded Response Model (cumulative logistic)
- **Tutz** — Sequential (adjacent-category) model

**Binomial IRT** (two parameters per item; response treated as B(*K*−1, *p*)):

- **Binom 1PL (logit)** — logistic link, discrimination fixed
- **Binom 2PL (logit)** — logistic link, free discrimination
- **Binom 1PL (cloglog)** — complementary log-log link, discrimination fixed
- **NP spline** — monotone spline link, binomial category distribution; fitted per item via GPCM EAP thetas

All models are fit with `mirt`. The custom binomial items use `mirt::createItem()`.

### Datasets

We used `irw_filter(n_categories = c(3, 7))` to identify IRW datasets with 3–7
ordered response categories. Additional filters: N ≥ 200, ≥ 3 items, all items share
the same *K*. Datasets larger than 5,000 respondents were downsampled.

```{r}
#| label: selection
#| eval: false

library(irw)
candidates <- irw_filter(n_categories = c(3, 7))
```

### Parameter counts

The two families differ substantially in how many parameters they use per item:

| Model | Params per item | Grows with K? |
|---|---|---|
| PCM | K−1 | yes |
| GPCM | K | yes |
| GRM | K | yes |
| Tutz | K | yes |
| Binom 1PL (logit) | 1 | no |
| **Binom 2PL (logit)** | **2** | **no** |
| Binom 1PL (cloglog) | 1 | no |
| NP spline | spline knots | no |

PCM fixes discrimination at 1, leaving only K−1 threshold parameters free. GPCM, GRM,
and Tutz each add a free discrimination parameter for K total. At K = 4 (the most common
case in our data), GPCM uses 4 parameters per item vs. 2 for Binom 2PL (logit). The binomial
models' competitiveness despite this parsimony advantage is the central question.

### Cross-validation

20% of observed responses were randomly held out at the **cell level** — individual
person–item pairs, not whole persons. Models were fit on the remaining 80% of responses
(held-out cells treated as missing). Each person was then scored via EAP using only
their observed training responses, and held-out cells were predicted as expected scores
E[*Y*|θ̂] / (*K*−1), putting all models on a common [0, 1] scale. RMSE was computed
over the held-out cells.

The full compute script is at `vignettes/poly_compare_multi_compute.R`.

### Inter-Model Vigorish (IMV)

Alongside RMSE we compute two IMV metrics using the `imv` package
([Domingue et al.](https://www.cambridge.org/core/journals/psychometrika/article/intermodel-vigorish-as-a-lens-for-understanding-and-quantifying-the-value-of-item-response-models-for-dichotomously-coded-items/F61C75F6F945A5B13F73C6128EB83998)).
IMV extends the binary vigorish to polytomous responses in two ways:

- **ω_c (category)** — all pairwise category comparisons, weighted by category frequency. The "c" indexes categories.
- **ω_t (thresholded)** — cumulative dichotomizations at each threshold, weighted by prevalence below each cut.

Both are computed against a **prevalence baseline** (training-set category proportions), so positive values mean the model improves on simply knowing the overall category distribution. We also report **pairwise IMV** — ω_c and ω_t with Binom 2PL (logit) as baseline and GPCM as the enhanced model — to directly quantify how much the standard polytomous parameterisation adds over its binomial counterpart.

---

```{r}
#| label: sim-ref
#| echo: false

sim_cache <- "poly_compare_data/sim_gpcm_k5.rds"
if (!file.exists(sim_cache)) {
  set.seed(20260528)
  ni <- 25L; N_sim <- 1000L; ncat <- 5L
  a <- matrix(rlnorm(ni, 0, 0.5), ni, 1)
  d <- matrix(rnorm(ni * (ncat - 1L)), ni, ncat - 1L)
  sim_dat <- simdata(a = a, d = d, N = N_sim, itemtype = "gpcm")

  non_missing <- which(!is.na(sim_dat), arr.ind = TRUE)
  holdout_idx <- sample(nrow(non_missing), floor(0.2 * nrow(non_missing)))
  train_mat   <- sim_dat
  train_mat[non_missing[holdout_idx, , drop = FALSE]] <- NA
  test_cells  <- non_missing[holdout_idx, , drop = FALSE]
  obs_vals    <- sim_dat[test_cells]

  mod   <- mirt(train_mat, 1, itemtype = rep("gpcm", ni),
                verbose = FALSE, technical = list(NCYCLES = 2000))
  theta <- fscores(mod, response.pattern = train_mat, method = "EAP")[, 1]
  K <- ncat
  pred <- numeric(nrow(test_cells))
  for (j in seq_len(ni)) {
    rows_j <- which(test_cells[, 2] == j)
    if (length(rows_j) == 0L) next
    Theta_j <- matrix(theta[test_cells[rows_j, 1]], ncol = 1)
    pr      <- probtrace(extract.item(mod, j), Theta_j)
    pred[rows_j] <- as.numeric(pr %*% 0:(ncol(pr) - 1L)) / (K - 1L)
  }
  saveRDS(sqrt(mean((pred - obs_vals / (K - 1L))^2)), sim_cache)
}
sim_rmse_gpcm <- readRDS(sim_cache)
```

## Results

### Dataset overview

```{r}
#| label: tbl-datasets
#| tbl-cap: "Polytomous datasets included in the analysis."

meta |>
  left_join(
    winners |> select(dataset, best_model = model),
    by = "dataset"
  ) |>
  rename(Dataset = dataset, `K` = K, Items = ni,
         `N (train+test)` = N, `Best model` = best_model) |>
  arrange(K, Items)
```

### RMSE by dataset and model

```{r}
#| label: fig-rmse-dot
#| fig-cap: "Cross-validated RMSE for each model on each dataset. Datasets ordered by median RMSE across models. Standard polytomous models in blue, binomial models in orange. Vertical dashed line: GPCM RMSE on simulated data (N=1000, 25 items, K=5; discriminations ~ LogNormal(0, 0.5), thresholds ~ Normal(0, 1), generated via `mirt::simdata()`)."
#| fig-height: 6.5

pal <- c(Standard = "#2166ac", Binomial = "#d6604d")

p <- ggplot(all_rmse, aes(
    x    = rmse,
    y    = short_name,
    colour = mod_type,
    shape  = model,
    text = paste0(
      "Dataset: ", dataset, "<br>",
      "Model: ", model, "<br>",
      "RMSE: ", round(rmse, 4), "<br>",
      "K: ", meta$K[match(dataset, meta$dataset)],
      "  items: ", meta$ni[match(dataset, meta$dataset)],
      "  N: ", meta$N[match(dataset, meta$dataset)]
    )
  )) +
  geom_vline(xintercept = sim_rmse_gpcm, linetype = "dashed",
             colour = "grey30", linewidth = 0.7) +
  geom_point(size = 2.5, alpha = 0.85,
             position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = pal, name = "Family") +
  scale_shape_manual(values = c(16, 17, 15, 3, 1, 2, 4, 8), name = "Model") +
  labs(x = "CV RMSE (rescaled response)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_line(colour = "grey92"),
    axis.text.y        = element_blank()
  )

ggplotly(p, tooltip = "text") |>
  layout(
    yaxis = list(showticklabels = FALSE),
    annotations = list(list(
      text      = "80/20 train/test split. Lower RMSE = better prediction.",
      showarrow = FALSE,
      xref = "paper", yref = "paper",
      x = 0, y = -0.07,
      xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### Model rankings

```{r}
#| label: fig-ranks
#| fig-cap: "Mean rank across all datasets (rank 1 = best). Error bars show ±1 SE."
#| fig-height: 3.5

rank_se <- ranks |>
  group_by(model, mod_type) |>
  summarise(mean_rank = mean(rank), se = sd(rank) / sqrt(n()), .groups = "drop")

ggplot(rank_se, aes(
    x = reorder(model, mean_rank),
    y = mean_rank,
    fill = mod_type,
    ymin = mean_rank - se,
    ymax = mean_rank + se
  )) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(width = 0.25, colour = "grey40") +
  scale_fill_manual(values = pal, name = "Family") +
  coord_flip() +
  labs(x = NULL, y = "Mean rank (lower = better)") +
  theme_minimal(base_size = 13)
```

### IMV vs. prevalence baseline

```{r}
#| label: fig-imv-bars
#| fig-cap: "Mean ω_c (category) and ω_t (thresholded) IMV vs. the training-set prevalence baseline. Higher = more improvement over simply predicting category proportions. Error bars show ±1 SE."
#| fig-height: 4

imv_long <- mean_imv |>
  tidyr::pivot_longer(c(mean_imv_c, mean_imv_t),
                      names_to = "metric", values_to = "imv") |>
  mutate(se     = ifelse(metric == "mean_imv_c", se_imv_c, se_imv_t),
         metric = recode(metric,
                         mean_imv_c = "ω_c (category)",
                         mean_imv_t = "ω_t (thresholded)"))

ggplot(imv_long, aes(x = reorder(model, -imv), y = imv,
                     fill = mod_type, ymin = imv - se, ymax = imv + se)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(width = 0.25, colour = "grey40") +
  scale_fill_manual(values = pal, name = "Family") +
  facet_wrap(~metric) +
  coord_flip() +
  labs(x = NULL, y = "Mean IMV (higher = better)") +
  theme_minimal(base_size = 13)
```

### IMV vs. Binom 1PL (logit) floor

```{r}
#| label: fig-imv-1pl-bars
#| fig-cap: "Mean ω_c (category) and ω_t (thresholded) IMV vs. the Binom 1PL (logit) floor. Positive = model beats the simplest binomial baseline. Error bars show ±1 SE. The Binom 1PL (logit) row is zero by definition."
#| fig-height: 4

imv_1pl_long <- mean_imv_1pl |>
  tidyr::pivot_longer(c(mean_imv_c, mean_imv_t),
                      names_to = "metric", values_to = "imv") |>
  mutate(se     = ifelse(metric == "mean_imv_c", se_imv_c, se_imv_t),
         metric = recode(metric,
                         mean_imv_c = "ω_c (category)",
                         mean_imv_t = "ω_t (thresholded)"))

ggplot(imv_1pl_long, aes(x = reorder(model, -imv), y = imv,
                          fill = mod_type, ymin = imv - se, ymax = imv + se)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(width = 0.25, colour = "grey40") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  scale_fill_manual(values = pal, name = "Family") +
  facet_wrap(~metric) +
  coord_flip() +
  labs(x = NULL, y = "Mean IMV vs. Binom 1PL (logit) (higher = better)") +
  theme_minimal(base_size = 13)
```

### Pairwise IMV distributions

For each dataset, pairwise IMV is computed with the binomial model as baseline and
the standard polytomous model as the enhanced model. Positive values indicate the
standard polytomous model improves over its binomial counterpart. The four panels
cover both IMV variants (ω_c, ω_t) and both discrimination-constrained (Binom 1PL (logit) vs.
PCM) and free-discrimination (Binom 2PL (logit) vs. GPCM) comparisons.

```{r}
#| label: fig-pair-imv-dist
#| fig-cap: "Distribution of pairwise IMV across datasets. Each panel shows ω_c (category) or ω_t (thresholded) for the named comparison. Positive values indicate the standard polytomous model improves over its binomial counterpart. Vertical dashed line at zero."
#| fig-height: 5

pair_long_raw <- all_pair |>
  filter(comparison %in% c("Binom 1PL (logit)→PCM", "Binom 2PL (logit)→GPCM")) |>
  tidyr::pivot_longer(c(imv_c, imv_t), names_to = "metric", values_to = "imv") |>
  mutate(
    metric     = recode(metric, imv_c = "ω_c (category)", imv_t = "ω_t (thresholded)"),
    comparison = factor(comparison, levels = c("Binom 1PL (logit)→PCM", "Binom 2PL (logit)→GPCM"))
  )

panel_stats <- pair_long_raw |>
  group_by(metric, comparison) |>
  summarise(
    mean_imv = mean(imv, na.rm = TRUE),
    n_clip   = sum(imv < -0.1 | imv > 0.1, na.rm = TRUE),
    .groups  = "drop"
  ) |>
  mutate(
    clip_label = ifelse(n_clip > 0, paste0(n_clip, " clipped"), ""),
    mean_label = paste0("mean = ", round(mean_imv, 3)),
    mean_x     = pmin(pmax(mean_imv, -0.1), 0.1)
  )

pair_long <- pair_long_raw |>
  group_by(metric, comparison) |>
  mutate(imv = pmin(pmax(imv, -0.1), 0.1)) |>
  ungroup()

ggplot(pair_long, aes(x = imv)) +
  geom_histogram(bins = 20, fill = "#2166ac", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_vline(data = panel_stats, aes(xintercept = mean_x),
             colour = "firebrick", linewidth = 0.8) +
  geom_text(data = panel_stats, aes(label = mean_label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
            size = 3.3, colour = "firebrick", inherit.aes = FALSE) +
  geom_text(data = panel_stats, aes(label = clip_label),
            x = Inf, y = -Inf, hjust = 1.1, vjust = -0.5,
            size = 3.3, colour = "grey50", inherit.aes = FALSE) +
  facet_grid(metric ~ comparison) +
  labs(x = "Pairwise IMV (positive = standard polytomous model better)", y = "Count",
       caption = "Values winsorized to [−0.1, 0.1]. Red line = mean (computed on full data).") +
  theme_minimal(base_size = 13)
```

```{r}
#| label: tbl-h2h
#| tbl-cap: "Per-dataset comparison: RMSE and pairwise IMV for GPCM vs. Binom 2PL (logit). Sorted by GPCM RMSE."

h2h_imv |>
  select(Dataset = dataset, K, Items = ni, N,
         `RMSE GPCM` = GPCM, `RMSE Binom 2PL` = `Binom 2PL (logit)`,
         `ω_c` = imv_c, `ω_t` = imv_t) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

### Best standard polytomous vs. best binomial (and their constrained counterparts)

GPCM is the best-performing standard polytomous model and Binom 2PL (logit) the best binomial
model; PCM and Binom 1PL (logit) are their discrimination-constrained counterparts. The two
panels show how each flexible model compares to its constrained analogue from the other
family.

```{r}
#| label: fig-model-pairs
#| fig-cap: "Left: GPCM vs. Binom 2PL (logit) (flexible models). Right: PCM vs. Binom 1PL (logit) (constrained models). Points above the diagonal: standard polytomous model fits better. Coloured dotted lines show the mean RMSE for the two featured models (orange = binomial, blue = standard); grey dotted lines show the means for the corresponding constrained/flexible counterpart."
#| fig-height: 4.5

make_scatter <- function(mod_x, mod_y, col, other_x = NULL, other_y = NULL) {
  d <- all_rmse |>
    filter(model %in% c(mod_x, mod_y)) |>
    select(dataset, model, rmse) |>
    pivot_wider(names_from = model, values_from = rmse) |>
    left_join(meta, by = "dataset") |>
    mutate(tooltip = paste0(
      "Dataset: ", dataset, "<br>",
      mod_y, ": ", round(.data[[mod_y]], 4), "<br>",
      mod_x, ": ", round(.data[[mod_x]], 4), "<br>",
      "K: ", K, "  items: ", ni, "  N: ", N
    ))
  mean_x <- mean(d[[mod_x]], na.rm = TRUE)
  mean_y <- mean(d[[mod_y]], na.rm = TRUE)
  other_mean_x <- if (!is.null(other_x))
    mean(all_rmse$rmse[all_rmse$model == other_x], na.rm = TRUE) else NULL
  other_mean_y <- if (!is.null(other_y))
    mean(all_rmse$rmse[all_rmse$model == other_y], na.rm = TRUE) else NULL
  p <- ggplot(d, aes(x = .data[[mod_x]], y = .data[[mod_y]],
                size = ni, text = tooltip)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
    geom_vline(xintercept = mean_x, linetype = "dotted", colour = "#d6604d", linewidth = 0.8) +
    geom_hline(yintercept = mean_y, linetype = "dotted", colour = "#2166ac", linewidth = 0.8)
  if (!is.null(other_mean_x))
    p <- p + geom_vline(xintercept = other_mean_x, linetype = "dotted", colour = "grey50", linewidth = 0.8)
  if (!is.null(other_mean_y))
    p <- p + geom_hline(yintercept = other_mean_y, linetype = "dotted", colour = "grey50", linewidth = 0.8)
  p +
    geom_point(colour = col, alpha = 0.75) +
    scale_size_continuous(range = c(3, 9), name = "# items") +
    labs(x = paste("CV RMSE —", mod_x), y = paste("CV RMSE —", mod_y)) +
    theme_minimal(base_size = 12)
}

p_flex  <- make_scatter("Binom 2PL (logit)", "GPCM", "#2166ac", other_x = "Binom 1PL (logit)", other_y = "PCM")
p_const <- make_scatter("Binom 1PL (logit)", "PCM",  "#2166ac", other_x = "Binom 2PL (logit)", other_y = "GPCM")

subplot(
  ggplotly(p_flex,  tooltip = "text"),
  ggplotly(p_const, tooltip = "text"),
  nrows = 1, shareY = FALSE, titleX = TRUE, titleY = TRUE
) |>
  layout(
    annotations = list(list(
      text      = "Above diagonal: standard polytomous model better.",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0, y = -0.1, xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### Performance gap vs. number of response categories

With more categories, standard polytomous models gain parameters while binomial models
stay at 2 — does the prediction gap widen with *K*?

```{r}
#| label: fig-by-k
#| fig-cap: "RMSE difference (Binom 2PL (logit) minus GPCM) by number of response categories K. Positive values mean GPCM predicts better. Each point is one dataset; black diamonds are per-K means."
#| fig-height: 4

gap_k <- all_rmse |>
  filter(model %in% c("GPCM", "Binom 2PL (logit)")) |>
  select(dataset, model, rmse) |>
  pivot_wider(names_from = model, values_from = rmse) |>
  left_join(meta |> select(dataset, K, ni, N), by = "dataset") |>
  mutate(
    gap     = `Binom 2PL (logit)` - GPCM,
    tooltip = paste0(
      "Dataset: ", dataset, "<br>",
      "K: ", K, "<br>",
      "Gap: ", round(gap, 4), "<br>",
      "items: ", ni, "  N: ", N
    )
  )

p_gap <- ggplot(gap_k, aes(x = K, y = gap, text = tooltip)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(width = 0.08, size = 3, colour = "#2166ac", alpha = 0.75) +
  stat_summary(aes(x = K, y = gap), fun = mean,
               geom = "point", shape = 18, size = 5, colour = "black",
               inherit.aes = FALSE) +
  scale_x_continuous(breaks = 3:8) +
  labs(x = "K (number of response categories)", y = "RMSE(Binom 2PL (logit)) − RMSE(GPCM)") +
  theme_minimal(base_size = 13)

ggplotly(p_gap, tooltip = "text") |>
  layout(
    annotations = list(list(
      text      = "Above zero: GPCM better. Black diamonds: per-K means.",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0, y = -0.1, xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

---

## What to notice

**Discrimination matters — and so does model family.** The clearest within-family
pattern is that freeing discrimination helps substantially: GPCM (mean rank
`r round(mean_ranks$mean_rank[mean_ranks$model=="GPCM"],2)`) beats PCM
(`r round(mean_ranks$mean_rank[mean_ranks$model=="PCM"],2)`), and Binom 2PL (logit)
(`r round(mean_ranks$mean_rank[mean_ranks$model=="Binom 2PL (logit)"],2)`) beats Binom 1PL (logit)
(`r round(mean_ranks$mean_rank[mean_ranks$model=="Binom 1PL (logit)"],2)`). But across families,
the standard polytomous models take the top three spots — GPCM, GRM, Tutz all rank
ahead of Binom 2PL (logit). The per-category parameterisation carries predictive value beyond
what a two-parameter binomial model can capture.

**GPCM dominates.** GPCM wins on `r sum(winners$model == "GPCM")` of
`r nrow(winners)` datasets and beats Binom 2PL (logit) head-to-head on `r n_gpcm_beats_2pl` of
`r nrow(h2h)`. Binom 2PL (logit) wins on only `r sum(winners$model == "Binom 2PL (logit)")` datasets
despite using fewer parameters. The parsimony advantage of the binomial model does not
translate to better out-of-sample prediction at this scale.

**Constrained binomial models perform poorly.** Binom 1PL (logit) and Binom 1PL (cloglog) rank last,
well below even PCM. Fixing discrimination hurts far more in the binomial family than
in the standard polytomous family — PCM at least captures per-category thresholds
precisely, while Binom 1PL (logit) reduces the entire response distribution to a single location
parameter.

**IMV amplifies the gap between families.** By RMSE, Binom 2PL (logit) is nearly tied with
GPCM/GRM/Tutz. But by ω_c vs. the prevalence baseline, standard polytomous models score
substantially higher (~0.22–0.24 vs. ~0.17 for Binom 2PL (logit)), indicating they do a better
job discriminating *which category* a person falls in — signal that expected-score RMSE
compresses away. ω_t tells a similar story but more conservatively.

**Pairwise IMV confirms the advantage is real but variable.** The distributions of
ω_c and ω_t for both the constrained (Binom 1PL (logit) → PCM) and flexible (Binom 2PL (logit) → GPCM)
comparisons are right-skewed with a mode near zero, but most mass lies above zero — the
standard polytomous model typically wins. The constrained comparison (Binom 1PL (logit) → PCM)
tends to show larger gains, consistent with PCM's additional threshold parameters
compensating for the fixed-discrimination constraint.

---

## Extending this vignette

- **Effect sizes.** The RMSE differences are often small. Pair them with a permutation
  test or bootstrap CI to assess whether GPCM's edge over Binom 2PL (logit) is reliable.

- **More link functions.** Add probit and cauchit binomial variants to map out the full
  link-function landscape.

- **Scale-up.** Bump `N_DATASETS` in `poly_compare_multi_compute.R` and re-run to
  cover more of the IRW polytomous catalogue.

---

## Reproducibility

Results were computed on 2026-05-28 using `r nrow(meta)` IRW tables. To reproduce:

```r
# From project root
Rscript vignettes/poly_compare_multi_compute.R
quarto::quarto_render("vignettes/poly_compare.qmd")
```

::: {.callout-note}
The compute script caches each dataset in `vignettes/poly_compare_data/`. Delete the
per-dataset `.rds` files to force a fresh run; delete `multi_results.rds` to
re-combine from scratch.
:::
