---
title: "Polytomous IRT: Binomial vs. Standard Polytomous Models"
description: |
  How do parsimonious binomial IRT models compare to standard polytomous IRT models
  (GPCM, GRM, PCM, sequential) when evaluated by cross-validated prediction error?
  We fit seven models to each of a diverse set of IRW polytomous datasets and compare
  out-of-sample RMSE.
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(plotly)
  library(mirt)
})

fs       <- list.files("poly_compare_data", pattern = "^multi_.+\\.rds$", full.names = TRUE)
fs       <- fs[!grepl("multi_(results|summary)\\.rds$", fs)]
results  <- Filter(Negate(is.null), lapply(fs, readRDS))
names(results) <- sub("^multi_", "", sub("\\.rds$", "", basename(fs)))
# backward compat: old files store K_range, new files store K
results  <- lapply(results, function(r) {
  if (is.null(r[["K"]])) r$K <- r$K_range + 1L
  r
})
results  <- Filter(function(r) r$K >= 3, results)  # drop binary (K < 3)
all_rmse <- do.call(rbind, lapply(results, `[[`, "rmse"))

meta <- do.call(rbind, lapply(names(results), function(nm) {
  r <- results[[nm]]
  data.frame(dataset = nm, K = r$K, ni = r$ni, N = r$N,
             stringsAsFactors = FALSE)
}))

# Model ordering for plots
model_levels <- c("PCM", "GPCM", "GRM", "Tutz", "1PL+logit", "2PL+logit", "1PL+cloglog")
model_type   <- c(PCM = "Standard", GPCM = "Standard", GRM = "Standard", Tutz = "Standard",
                  `1PL+logit` = "Binomial", `2PL+logit` = "Binomial",
                  `1PL+cloglog` = "Binomial")

all_rmse <- all_rmse |>
  mutate(
    model     = factor(model, levels = model_levels),
    mod_type  = model_type[as.character(model)],
    short_name = gsub("_", " ", dataset) |> stringr::str_trunc(35)
  )

# ensure short_name is unique per dataset (truncation can collapse distinct names)
sn_map <- all_rmse |>
  distinct(dataset, short_name) |>
  mutate(short_name = make.unique(short_name, sep = " "))
all_rmse <- all_rmse |>
  select(-short_name) |>
  left_join(sn_map, by = "dataset")

# Dataset-level: median RMSE across models for ordering
ds_med <- all_rmse |>
  group_by(dataset, short_name) |>
  summarise(med_rmse = median(rmse), .groups = "drop") |>
  arrange(med_rmse)

all_rmse <- all_rmse |>
  mutate(short_name = factor(short_name, levels = ds_med$short_name))

# Winner per dataset
winners <- all_rmse |>
  group_by(dataset) |>
  slice_min(rmse, n = 1, with_ties = FALSE) |>
  ungroup()

# Mean rank
ranks <- all_rmse |>
  group_by(dataset) |>
  mutate(rank = rank(rmse, ties.method = "min")) |>
  ungroup()

mean_ranks <- ranks |>
  group_by(model, mod_type) |>
  summarise(mean_rank = mean(rank), .groups = "drop") |>
  arrange(mean_rank)

# Head-to-head: GPCM vs 2PL+logit
h2h <- all_rmse |>
  filter(model %in% c("GPCM", "2PL+logit")) |>
  select(dataset, model, rmse) |>
  tidyr::pivot_wider(names_from = model, values_from = rmse)
n_gpcm_beats_2pl <- sum(h2h$GPCM < h2h$`2PL+logit`, na.rm = TRUE)
```

## Research question

Polytomous IRT models (GPCM, GRM, PCM, sequential) are the standard tools for ordered
responses with three or more categories. But an alternative strategy is to treat the
response as a **binomial count** — fitting a binomial distribution B(*K*−1, *p*) where
*p* is a function of the latent trait. Binomial IRT models are parsimonious (two
parameters per item regardless of *K*) and can use different link functions (logit,
log-log).

The IRW makes a large-scale empirical test straightforward: **do binomial models match
or beat standard polytomous IRT when evaluated by out-of-sample prediction error?** And
does the choice of link function matter?

---

## Approach

### Models

We compare seven models across two families:

**Standard polytomous** (one set of parameters per response category):

- **PCM** — Partial Credit Model (Rasch-constrained GPCM; discrimination fixed at 1)
- **GPCM** — Generalized Partial Credit Model (free discrimination)
- **GRM** — Graded Response Model (cumulative logistic)
- **Tutz** — Sequential (adjacent-category) model

**Binomial IRT** (two parameters per item; response treated as B(*K*−1, *p*)):

- **1PL+logit** — logistic link, discrimination fixed
- **2PL+logit** — logistic link, free discrimination
- **1PL+cloglog** — complementary log-log link, discrimination fixed

All models are fit with `mirt`. The custom binomial items use `mirt::createItem()`.

### Datasets

We used `irw_filter(n_categories = c(3, 7))` to identify IRW datasets with 3–7
ordered response categories. Additional filters: N ≥ 200, ≥ 3 items, all items share
the same *K*. Datasets larger than 5,000 respondents were downsampled.

```{r}
#| label: selection
#| eval: false

library(irw)
candidates <- irw_filter(n_categories = c(3, 7))
```

### Parameter counts

The two families differ substantially in how many parameters they use per item:

| Model | Params per item | Grows with K? |
|---|---|---|
| PCM | K−1 | yes |
| GPCM | K | yes |
| GRM | K | yes |
| Tutz | K | yes |
| 1PL+logit | 1 | no |
| **2PL+logit** | **2** | **no** |
| 1PL+cloglog | 1 | no |

PCM fixes discrimination at 1, leaving only K−1 threshold parameters free. GPCM, GRM,
and Tutz each add a free discrimination parameter for K total. At K = 4 (the most common
case in our data), GPCM uses 4 parameters per item vs. 2 for 2PL+logit. The binomial
models' competitiveness despite this parsimony advantage is the central question.

### Cross-validation

20% of observed responses were randomly held out at the **cell level** — individual
person–item pairs, not whole persons. Models were fit on the remaining 80% of responses
(held-out cells treated as missing). Each person was then scored via EAP using only
their observed training responses, and held-out cells were predicted as expected scores
E[*Y*|θ̂] / (*K*−1), putting all models on a common [0, 1] scale. RMSE was computed
over the held-out cells.

The full compute script is at `vignettes/poly_compare_multi_compute.R`.

---

```{r}
#| label: sim-ref
#| echo: false

sim_cache <- "poly_compare_data/sim_gpcm_k5.rds"
if (!file.exists(sim_cache)) {
  set.seed(20260528)
  ni <- 25L; N_sim <- 1000L; ncat <- 5L
  a <- matrix(rlnorm(ni, 0, 0.5), ni, 1)
  d <- matrix(rnorm(ni * (ncat - 1L)), ni, ncat - 1L)
  sim_dat <- simdata(a = a, d = d, N = N_sim, itemtype = "gpcm")

  non_missing <- which(!is.na(sim_dat), arr.ind = TRUE)
  holdout_idx <- sample(nrow(non_missing), floor(0.2 * nrow(non_missing)))
  train_mat   <- sim_dat
  train_mat[non_missing[holdout_idx, , drop = FALSE]] <- NA
  test_cells  <- non_missing[holdout_idx, , drop = FALSE]
  obs_vals    <- sim_dat[test_cells]

  mod   <- mirt(train_mat, 1, itemtype = rep("gpcm", ni),
                verbose = FALSE, technical = list(NCYCLES = 2000))
  theta <- fscores(mod, response.pattern = train_mat, method = "EAP")[, 1]
  K <- ncat
  pred <- numeric(nrow(test_cells))
  for (j in seq_len(ni)) {
    rows_j <- which(test_cells[, 2] == j)
    if (length(rows_j) == 0L) next
    Theta_j <- matrix(theta[test_cells[rows_j, 1]], ncol = 1)
    pr      <- probtrace(extract.item(mod, j), Theta_j)
    pred[rows_j] <- as.numeric(pr %*% 0:(ncol(pr) - 1L)) / (K - 1L)
  }
  saveRDS(sqrt(mean((pred - obs_vals / (K - 1L))^2)), sim_cache)
}
sim_rmse_gpcm <- readRDS(sim_cache)
```

## Results

### Dataset overview

```{r}
#| label: tbl-datasets
#| tbl-cap: "Polytomous datasets included in the analysis."

meta |>
  left_join(
    winners |> select(dataset, best_model = model),
    by = "dataset"
  ) |>
  rename(Dataset = dataset, `K` = K, Items = ni,
         `N (train+test)` = N, `Best model` = best_model) |>
  arrange(K, Items)
```

### RMSE by dataset and model

```{r}
#| label: fig-rmse-dot
#| fig-cap: "Cross-validated RMSE for each model on each dataset. Datasets ordered by median RMSE across models. Standard polytomous models in blue, binomial models in orange. Vertical dashed line: GPCM RMSE on simulated data (N=1000, 25 items, K=5; discriminations ~ LogNormal(0, 0.5), thresholds ~ Normal(0, 1), generated via `mirt::simdata()`)."
#| fig-height: 6.5

pal <- c(Standard = "#2166ac", Binomial = "#d6604d")

p <- ggplot(all_rmse, aes(
    x    = rmse,
    y    = short_name,
    colour = mod_type,
    shape  = model,
    text = paste0(
      "Dataset: ", dataset, "<br>",
      "Model: ", model, "<br>",
      "RMSE: ", round(rmse, 4), "<br>",
      "K: ", meta$K[match(dataset, meta$dataset)],
      "  items: ", meta$ni[match(dataset, meta$dataset)],
      "  N: ", meta$N[match(dataset, meta$dataset)]
    )
  )) +
  geom_vline(xintercept = sim_rmse_gpcm, linetype = "dashed",
             colour = "grey30", linewidth = 0.7) +
  geom_point(size = 2.5, alpha = 0.85,
             position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = pal, name = "Family") +
  scale_shape_manual(values = c(16, 17, 15, 3, 1, 2, 4), name = "Model") +
  labs(x = "CV RMSE (rescaled response)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_line(colour = "grey92"),
    axis.text.y        = element_blank()
  )

ggplotly(p, tooltip = "text") |>
  layout(
    yaxis = list(showticklabels = FALSE),
    annotations = list(list(
      text      = "80/20 train/test split. Lower RMSE = better prediction.",
      showarrow = FALSE,
      xref = "paper", yref = "paper",
      x = 0, y = -0.07,
      xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### Model rankings

```{r}
#| label: fig-ranks
#| fig-cap: "Mean rank across all datasets (rank 1 = best). Error bars show ±1 SE."
#| fig-height: 3.5

rank_se <- ranks |>
  group_by(model, mod_type) |>
  summarise(mean_rank = mean(rank), se = sd(rank) / sqrt(n()), .groups = "drop")

ggplot(rank_se, aes(
    x = reorder(model, mean_rank),
    y = mean_rank,
    fill = mod_type,
    ymin = mean_rank - se,
    ymax = mean_rank + se
  )) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(width = 0.25, colour = "grey40") +
  scale_fill_manual(values = pal, name = "Family") +
  coord_flip() +
  labs(x = NULL, y = "Mean rank (lower = better)") +
  theme_minimal(base_size = 13)
```

### Best standard polytomous vs. best binomial (and their constrained counterparts)

GPCM is the best-performing standard polytomous model and 2PL+logit the best binomial
model; PCM and 1PL+logit are their discrimination-constrained counterparts. The two
panels show how each flexible model compares to its constrained analogue from the other
family.

```{r}
#| label: fig-model-pairs
#| fig-cap: "Left: GPCM vs. 2PL+logit (flexible models). Right: PCM vs. 1PL+logit (constrained models). Points above the diagonal: standard polytomous model fits better. Coloured dotted lines show the mean RMSE for the two featured models (orange = binomial, blue = standard); grey dotted lines show the means for the corresponding constrained/flexible counterpart."
#| fig-height: 4.5

make_scatter <- function(mod_x, mod_y, col, other_x = NULL, other_y = NULL) {
  d <- all_rmse |>
    filter(model %in% c(mod_x, mod_y)) |>
    select(dataset, model, rmse) |>
    pivot_wider(names_from = model, values_from = rmse) |>
    left_join(meta, by = "dataset") |>
    mutate(tooltip = paste0(
      "Dataset: ", dataset, "<br>",
      mod_y, ": ", round(.data[[mod_y]], 4), "<br>",
      mod_x, ": ", round(.data[[mod_x]], 4), "<br>",
      "K: ", K, "  items: ", ni, "  N: ", N
    ))
  mean_x <- mean(d[[mod_x]], na.rm = TRUE)
  mean_y <- mean(d[[mod_y]], na.rm = TRUE)
  other_mean_x <- if (!is.null(other_x))
    mean(all_rmse$rmse[all_rmse$model == other_x], na.rm = TRUE) else NULL
  other_mean_y <- if (!is.null(other_y))
    mean(all_rmse$rmse[all_rmse$model == other_y], na.rm = TRUE) else NULL
  p <- ggplot(d, aes(x = .data[[mod_x]], y = .data[[mod_y]],
                size = ni, text = tooltip)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
    geom_vline(xintercept = mean_x, linetype = "dotted", colour = "#d6604d", linewidth = 0.8) +
    geom_hline(yintercept = mean_y, linetype = "dotted", colour = "#2166ac", linewidth = 0.8)
  if (!is.null(other_mean_x))
    p <- p + geom_vline(xintercept = other_mean_x, linetype = "dotted", colour = "grey50", linewidth = 0.8)
  if (!is.null(other_mean_y))
    p <- p + geom_hline(yintercept = other_mean_y, linetype = "dotted", colour = "grey50", linewidth = 0.8)
  p +
    geom_point(colour = col, alpha = 0.75) +
    scale_size_continuous(range = c(3, 9), name = "# items") +
    labs(x = paste("CV RMSE —", mod_x), y = paste("CV RMSE —", mod_y)) +
    theme_minimal(base_size = 12)
}

p_flex  <- make_scatter("2PL+logit", "GPCM", "#2166ac", other_x = "1PL+logit", other_y = "PCM")
p_const <- make_scatter("1PL+logit", "PCM",  "#2166ac", other_x = "2PL+logit", other_y = "GPCM")

subplot(
  ggplotly(p_flex,  tooltip = "text"),
  ggplotly(p_const, tooltip = "text"),
  nrows = 1, shareY = FALSE, titleX = TRUE, titleY = TRUE
) |>
  layout(
    annotations = list(list(
      text      = "Above diagonal: standard polytomous model better.",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0, y = -0.1, xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

### Performance gap vs. number of response categories

With more categories, standard polytomous models gain parameters while binomial models
stay at 2 — does the prediction gap widen with *K*?

```{r}
#| label: fig-by-k
#| fig-cap: "RMSE difference (2PL+logit minus GPCM) by number of response categories K. Positive values mean GPCM predicts better. Each point is one dataset; black diamonds are per-K means."
#| fig-height: 4

gap_k <- all_rmse |>
  filter(model %in% c("GPCM", "2PL+logit")) |>
  select(dataset, model, rmse) |>
  pivot_wider(names_from = model, values_from = rmse) |>
  left_join(meta |> select(dataset, K, ni, N), by = "dataset") |>
  mutate(
    gap     = `2PL+logit` - GPCM,
    tooltip = paste0(
      "Dataset: ", dataset, "<br>",
      "K: ", K, "<br>",
      "Gap: ", round(gap, 4), "<br>",
      "items: ", ni, "  N: ", N
    )
  )

p_gap <- ggplot(gap_k, aes(x = K, y = gap, text = tooltip)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(width = 0.08, size = 3, colour = "#2166ac", alpha = 0.75) +
  stat_summary(aes(x = K, y = gap), fun = mean,
               geom = "point", shape = 18, size = 5, colour = "black",
               inherit.aes = FALSE) +
  scale_x_continuous(breaks = 3:8) +
  labs(x = "K (number of response categories)", y = "RMSE(2PL+logit) − RMSE(GPCM)") +
  theme_minimal(base_size = 13)

ggplotly(p_gap, tooltip = "text") |>
  layout(
    annotations = list(list(
      text      = "Above zero: GPCM better. Black diamonds: per-K means.",
      showarrow = FALSE, xref = "paper", yref = "paper",
      x = 0, y = -0.1, xanchor = "left", yanchor = "top",
      font = list(size = 11, color = "grey50")
    ))
  )
```

---

## What to notice

**Discrimination matters — and so does model family.** The clearest within-family
pattern is that freeing discrimination helps substantially: GPCM (mean rank
`r round(mean_ranks$mean_rank[mean_ranks$model=="GPCM"],2)`) beats PCM
(`r round(mean_ranks$mean_rank[mean_ranks$model=="PCM"],2)`), and 2PL+logit
(`r round(mean_ranks$mean_rank[mean_ranks$model=="2PL+logit"],2)`) beats 1PL+logit
(`r round(mean_ranks$mean_rank[mean_ranks$model=="1PL+logit"],2)`). But across families,
the standard polytomous models take the top three spots — GPCM, GRM, Tutz all rank
ahead of 2PL+logit. The per-category parameterisation carries predictive value beyond
what a two-parameter binomial model can capture.

**GPCM dominates.** GPCM wins on `r sum(winners$model == "GPCM")` of
`r nrow(winners)` datasets and beats 2PL+logit head-to-head on `r n_gpcm_beats_2pl` of
`r nrow(h2h)`. 2PL+logit wins on only `r sum(winners$model == "2PL+logit")` datasets
despite using fewer parameters. The parsimony advantage of the binomial model does not
translate to better out-of-sample prediction at this scale.

**Constrained binomial models perform poorly.** 1PL+logit and 1PL+cloglog rank last,
well below even PCM. Fixing discrimination hurts far more in the binomial family than
in the standard polytomous family — PCM at least captures per-category thresholds
precisely, while 1PL+logit reduces the entire response distribution to a single location
parameter.

---

## Extending this vignette

- **Effect sizes.** The RMSE differences are often small. Pair them with a permutation
  test or bootstrap CI to assess whether GPCM's edge over 2PL+logit is reliable.

- **More link functions.** Add probit and cauchit binomial variants to map out the full
  link-function landscape.

- **IMV.** Replace (or supplement) RMSE with Inter-Model Vigorish to directly quantify
  how much each model gains over a reference baseline.

- **Scale-up.** Bump `N_DATASETS` in `poly_compare_multi_compute.R` and re-run to
  cover more of the IRW polytomous catalogue.

---

## Reproducibility

Results were computed on 2026-05-28 using `r nrow(meta)` IRW tables. To reproduce:

```r
# From project root
Rscript vignettes/poly_compare_multi_compute.R
quarto::quarto_render("vignettes/poly_compare.qmd")
```

::: {.callout-note}
The compute script caches each dataset in `vignettes/poly_compare_data/`. Delete the
per-dataset `.rds` files to force a fresh run; delete `multi_results.rds` to
re-combine from scratch.
:::
