---
title: "Extreme and Midpoint Responding in Likert-Format IRW Data"
description: |
  How prevalent are extreme (ERS) and midpoint (MRS) response styles in
  Likert-format IRW data, and can you tell from the responses alone whether
  style is confounded with content? We screen IRW's Likert-scale tables
  (4-7 response categories, at least 5 items), compute person-level ERS/MRS
  indices, and show that the dependence between a sum-score style index and
  the substantive score is strong but non-monotone -- so a Pearson
  correlation, the obvious screen, sees almost none of it.
date: today
bibliography: response_style_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "response_style_data/response_style_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  binned_df        <- .cache$binned
  persons_df       <- .cache$persons
  candidate_tables <- .cache$candidate_tables
  date_run         <- .cache$date_run
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_tables_run   <- length(candidate_tables)
} else {
  summary_df  <- tibble()
  binned_df   <- tibble()
  persons_df  <- tibble()
  candidate_tables <- character()
  date_run <- Sys.Date()
  r_all_candidates <- 0L
  r_n_tables_run   <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — the IRW figures will not appear (the simulation ",
  "sections below still run). Generate results by running the compute ",
  "script, then re-render:\n\n",
  "```r\nRscript vignettes/response_style_compute.R\n```\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
## About this vignette

The dataset search, compute script, analysis design, and writing here were
produced largely by Claude (Anthropic), working from a task specification
and with human review. Check the compute script
(`response_style_compute.R`) directly if you're relying on the numbers.

We thank **Carl F. Falk** (McGill University) for reviewing this vignette
and for directing us to the literature it is built on
[@falk2020mnrm; @savalei2014recovering; @falk2016flexible].
:::

## Motivation

Likert-type items [@likert1932technique] ask respondents to place themselves
on an ordered scale — "strongly disagree" to "strongly agree," typically with
4 to 7 response categories. The scale is meant to be a transparent window
onto the underlying construct: someone who agrees more strongly with an item
should pick a higher category. In practice, people also differ systematically
in *how* they use the response scale itself, independent of the construct
being measured. Three response styles are especially well documented:

- **Extreme response style (ERS)** — a tendency to favor the endpoints of the
  scale (the lowest and highest categories) regardless of item content.
- **Midpoint response style (MRS)** — a tendency to favor the middle
  category, when one exists (only scales with an odd number of categories
  have a true midpoint).
- **Acquiescence (net agreement bias)** — a tendency toward the "agree" end
  of the scale across items regardless of content, distinct from ERS/MRS but
  related in that it's also a property of *how* someone uses the scale
  rather than *what* they think.

The cheap way to measure the first two is a **sum-score index**: count the
proportion of a person's responses that land on an endpoint (ERS) or on the
midpoint (MRS). It requires no model, runs in one line, and is what most
applied papers do. This vignette is largely about why that index cannot, on
its own, tell you whether style is contaminating your measurement — and what
the honest version of that question looks like.

Three concrete questions of the Likert-format tables in the IRW:

1. **How common is extreme responding, and does it vary a lot across
   datasets?**
2. **Can we tell, from a sum-score style index alone, whether style is
   entangled with the substantive score?**
3. **Does a simple, widely used "style control" fix anything?**

The short answers turn out to be: it varies enormously; no, and the usual
screen for it is actively misleading; and no.

### See extreme responding in action

Drag the slider to change how much a simulated respondent's answers are
driven by response style rather than by item content. At $p = 0$, every
answer reflects a fixed, moderate underlying attitude (with a little
item-level noise); as $p$ rises, a growing share of answers instead land on
a scale endpoint by chance, regardless of what the item actually asks. The
resulting **ERS** — the proportion of this one simulated respondent's own
answers sitting at an endpoint — is exactly the person-level statistic this
vignette computes for every respondent in every IRW Likert-format table
below.

```{ojs}
viewof style_p = Inputs.range(
  [0, 0.8], {value: 0.2, step: 0.05, label: "Extreme-response tendency (p)"}
)
```

```{ojs}
n_items_demo = 12
rng = d3.randomLcg(42)

demo_responses = {
  const resp = [];
  for (let i = 0; i < n_items_demo; i++) {
    if (rng() < style_p) {
      resp.push(rng() < 0.5 ? 1 : 5);
    } else {
      const honest = Math.round(3 + (rng() - 0.5) * 1.5);
      resp.push(Math.min(5, Math.max(1, honest)));
    }
  }
  return resp;
}

demo_ers = demo_responses.filter((r) => r === 1 || r === 5).length / n_items_demo
```

```{ojs}
Plot.plot({
  width: 700,
  height: 110,
  marginLeft: 40,
  x: {domain: [0.5, n_items_demo + 0.5], label: "Item"},
  y: {domain: [0.5, 5.5], label: "Response (1-5)", ticks: [1, 2, 3, 4, 5]},
  marks: [
    Plot.ruleY([1, 5], {stroke: "#d73027", strokeDasharray: "3,3"}),
    Plot.dot(
      demo_responses.map((r, i) => ({item: i + 1, resp: r})),
      {
        x: "item", y: "resp", r: 6,
        fill: (d) => (d.resp === 1 || d.resp === 5) ? "#d73027" : "#2166ac"
      }
    )
  ]
})
```

```{ojs}
html`<div style="font-size:1.05rem;margin-top:0.5rem;">
  This respondent's <strong>ERS = ${demo_ers.toFixed(2)}</strong> —
  ${Math.round(demo_ers * n_items_demo)} of ${n_items_demo} responses
  (<span style="color:#d73027;">red</span>) landed on an endpoint.
</div>`
```

## Data and methods

### Dataset selection

```{r}
#| label: filter-datasets
#| eval: false

all_candidates <- irw_filter(
  item_format    = "Likert Scale/selected response",
  n_categories   = c(4, 7),
  n_items        = c(5, Inf),
  n_participants = c(100, Inf)
)

# A manageable random sample for this pass, not the full candidate list
tables <- sample(all_candidates, min(150, length(all_candidates)))
```

We restrict to tables tagged as Likert-format selected-response items, with
4 to 7 response categories (covering both odd category counts with a true
midpoint — 5, 7 — and even counts without one — 4, 6), **at least 5 items
per person**, and **at least 100 respondents**. The item-count floor matters
more here than in most IRW vignettes: ERS and MRS are computed as a
*proportion of a person's own responses*, so with very few items per person
that proportion is built from very few trials and is correspondingly noisy
(a person who happens to answer 2 of 3 items at the scale endpoint gets an
ERS of 0.67 by chance alone). Five items is still a fairly low floor — treat
person-level style estimates from tables near it as noisier than those from
longer batteries. The respondent floor rules out tables too small for
table-level ERS/MRS means to be anything but sampling noise.

This selection is deliberately not exhaustive: `r r_all_candidates` tables
met these criteria as of `r format(date_run, "%B %Y")`, and this pass
analyzes a **random sample of `r r_n_tables_run` of them** rather than the
full list — some candidate tables run past 100 items and tens of thousands
of respondents, and running all of them in one pass isn't practical.
`fit_to_disk()`'s on-disk cache means a later pass with a larger or
different sample adds to what's already here rather than redoing it.

### Per-table computation

For each candidate table, `fit_response_style()`:

1. Fetches the table and builds a wide response matrix (`irw_fetch(table,
   resp = TRUE)`, falling back to a direct `irw_long2resp()` call if that
   internal conversion fails for a given table's shape).
2. Determines the response category range from the data directly (not just
   the table-level metadata) and **drops the table if per-item category
   ranges are inconsistent** — e.g. some items coded 1-4 and others 1-7 —
   since that indicates either a coding problem or a mixed response format
   that a single ERS/MRS definition can't cleanly apply to.
3. Computes, per person: ERS (proportion of non-missing responses at the
   minimum or maximum category), MRS (proportion at the midpoint category —
   left `NA` for tables with an even number of categories), and the person's
   mean response across items, which is the substantive raw score for a
   single-factor Likert battery.
4. Characterises the **shape** of the relationship between each style index
   and that raw score, using three statistics of increasing generality:
   the Pearson correlation (linear association only), a binned $\eta^2$
   (proportion of style variance explained by 20 quantile bins of the raw
   score — assumption-free, detects any shape), and the $R^2$ from
   regressing the style index on a natural spline of the raw score. The gap
   between $r^2$ and the other two is the object of interest.

::: {.callout-note}
## What this vignette does not do

A natural next step would be to fit a single-factor CFA to raw responses and
again to **ipsatized** responses (each person's items minus their own mean),
and read the change in standardized loadings as a measure of how much
response style distorts the substantive solution. This vignette does not do
that, because the comparison is not interpretable — wrong target and wrong
estimator. Since the method is common, the reasons are worked through in
[their own section below](#why-person-mean-centering-isnt-the-fix) rather
than left implicit.
:::

## Results

### How common is extreme responding?

```{r}
#| label: fig-ers-ranked
#| fig-cap: "Every table's mean ERS (proportion of responses at a scale endpoint), ranked from lowest to highest. Hover over a point to see which table it is. Error bars show +/- 1 SD across people within each table."
#| fig-height: 6

ers_plot_df <- summary_df %>%
  filter(!is.na(mean_ers)) %>%
  arrange(mean_ers) %>%
  mutate(
    rank    = row_number(),
    tooltip = paste0(
      "Table: ", table,
      "<br>Mean ERS: ", round(mean_ers, 2),
      "<br>SD ERS: ", round(sd_ers, 2),
      "<br>n_items: ", n_items,
      "<br>N: ", n_participants
    )
  )

p <- ggplot(ers_plot_df, aes(x = mean_ers, y = rank, text = tooltip)) +
  geom_errorbarh(aes(xmin = pmax(mean_ers - sd_ers, 0), xmax = pmin(mean_ers + sd_ers, 1)),
                  height = 0, colour = irw_grey) +
  geom_point(colour = irw_blue, size = 2) +
  labs(x = "Mean ERS (proportion of endpoint responses)", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggplotly(p, tooltip = "text") |>
  layout(margin = list(t = 60))
```

If ERS were a small, uniform nuisance, this ranking would be flat and low.
Instead — see the spread in @fig-ers-ranked — a handful of tables sit far to
the right, with a substantial fraction of *all* responses landing on a scale
endpoint, while others sit much lower. Table-level ERS is not a fixed
property of "using a Likert scale" in general; it varies with the
instrument, the sample, and the substantive content being asked about.

That last clause is doing more work than it looks. If mean ERS varies with
content, then ERS and content are not separate things you can measure
independently — which is exactly what the next section is about.

### A correlation is the wrong test for confounding {#a-correlation-is-the-wrong-test}

The natural screen is: correlate each person's ERS with their raw score. Near
zero means style is construct-irrelevant noise; far from zero means style and
construct are entangled. It is an appealing test, and it is not a valid one.

The reason is that the relationship between a sum-score style index and the
raw score is typically **strong but non-monotone** — U-shaped for ERS,
inverted-U for MRS — and a Pearson correlation is close to blind to a
symmetric U. @falk2020mnrm make this point directly (their Figure 4 shows "a
distinct U-shaped non-linear relationship" between the substantive trait and
sum-score ERS, and conclude that "it would be difficult for such methods to
disentangle ERS from the construct of interest").

Worse, part of the U is **arithmetically forced**. On a scale running from
$c_{\min}$ to $c_{\max}$, a person can avoid the endpoints entirely only
while their mean stays within the range the interior categories span,
$[c_{\min}+1,\ c_{\max}-1]$. Push the mean outside that band and endpoint
responses become unavoidable: reaching a mean $m < c_{\min}+1$ requires at
least a fraction $(c_{\min}+1-m)$ of responses at $c_{\min}$. So ERS must
rise toward 1 at both ends of the score range, in any dataset, whether or
not a single respondent has anything resembling a response style.

Here is what that does to the screen. We simulate a strictly unidimensional
5-category battery — a content trait, item noise, and **no response-style
process at all**:

```{r}
#| label: fig-sim-ushape
#| eval: true
#| fig-cap: "Simulated data with NO response-style process: 12 items, 5 categories, responses driven only by a single content trait plus noise. Each point is one simulated respondent. The dashed line is the arithmetic lower bound on ERS given a person's mean score. The Pearson correlation is essentially zero; the dependence is nearly total."
#| fig-height: 4.5

set.seed(7)
K <- 5; n_items_sim <- 12; n_sim <- 3000

theta_sim <- rnorm(n_sim)
sim_mat <- sapply(seq_len(n_items_sim), function(j) {
  z <- theta_sim * 0.9 + rnorm(n_sim, 0, sqrt(1 - 0.9^2))
  as.integer(cut(z, breaks = c(-Inf, -1.2, -0.4, 0.4, 1.2, Inf), labels = 1:K))
})

sim_ers  <- rowMeans(sim_mat == 1 | sim_mat == K)
sim_mean <- rowMeans(sim_mat)

# Arithmetic lower bound on ERS given the person's mean (see text)
ers_lower_bound <- function(m, min_cat, max_cat) {
  pmin(1, pmax(pmax(0, (min_cat + 1) - m), pmax(0, m - (max_cat - 1))))
}

sim_r    <- cor(sim_ers, sim_mean)
sim_eta2 <- {
  bin   <- cut(sim_mean, breaks = unique(quantile(sim_mean, seq(0, 1, length.out = 21))),
               include.lowest = TRUE)
  grand <- mean(sim_ers)
  sum(tapply(sim_ers, bin, function(v) length(v) * (mean(v) - grand)^2), na.rm = TRUE) /
    sum((sim_ers - grand)^2)
}

bound_df <- tibble(m = seq(1, K, length.out = 400)) %>%
  mutate(bound = ers_lower_bound(m, 1, K))

ggplot(tibble(sim_mean, sim_ers), aes(sim_mean, sim_ers)) +
  geom_point(alpha = 0.12, colour = irw_blue, size = 1) +
  geom_line(data = bound_df, aes(m, bound),
            linetype = "dashed", colour = irw_red, linewidth = 0.8) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              colour = "black", linewidth = 0.9) +
  annotate("label", x = 3, y = 0.92, hjust = 0.5, size = 4.2, label.size = 0,
           label = sprintf("Pearson r = %.3f  (r² = %.3f)\nbinned η² = %.2f",
                           sim_r, sim_r^2, sim_eta2)) +
  labs(x = "Person mean response (the substantive score)",
       y = "Person ERS") +
  coord_cartesian(ylim = c(0, 1))
```

In @fig-sim-ushape the correlation-based screen reports
`r sprintf("r = %.3f", sim_r)` — an $r^2$ of
`r sprintf("%.3f", sim_r^2)`, i.e. the style index and the score share
essentially none of their variance, apparently the cleanest possible bill of
health. The binned $\eta^2$ for the same data is
`r sprintf("%.2f", sim_eta2)`: the raw score determines almost all of the
variance in ERS. Both numbers describe the same 3,000 points. Only one of
them is right.

The practical upshot is that a near-zero ERS-score correlation is not
evidence of anything. It is the expected result when style and content are
maximally confounded in the ordinary, symmetric way.

### The same pattern in IRW data

```{r}
#| label: fig-r2-vs-eta2
#| fig-cap: "For each IRW table: variance in person ERS explained by the raw score, measured linearly (r-squared, x-axis) and without a shape assumption (binned eta-squared, y-axis). The dashed line is equality. Points sit far above it -- a correlation-based screen sees a small fraction of the dependence that is actually there."
#| fig-height: 5

shape_df <- summary_df %>%
  filter(!is.na(ers_score_cor), !is.na(ers_score_eta2)) %>%
  mutate(
    r2      = ers_score_cor^2,
    tooltip = paste0(
      "Table: ", table,
      "<br>Pearson r: ", round(ers_score_cor, 3),
      "<br>r²: ", round(r2, 3),
      "<br>binned η²: ", round(ers_score_eta2, 3),
      "<br>spline R²: ", round(ers_score_r2_spline, 3),
      "<br>n_items: ", n_items, " | N: ", n_participants
    )
  )

p <- ggplot(shape_df, aes(x = r2, y = ers_score_eta2, text = tooltip)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(colour = irw_red, alpha = 0.75, size = 2) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = expression("Linear only:"~r^2), y = expression("Any shape:"~eta^2))

ggplotly(p, tooltip = "text") |>
  layout(margin = list(t = 60))
```

```{r}
#| label: shape-numbers
#| echo: false

med_r2   <- median(shape_df$r2, na.rm = TRUE)
med_eta2 <- median(shape_df$ers_score_eta2, na.rm = TRUE)
n_hidden <- sum(shape_df$r2 < 0.05 & shape_df$ers_score_eta2 > 0.25, na.rm = TRUE)
n_shape  <- nrow(shape_df)
```

Across the `r n_shape` tables in @fig-r2-vs-eta2, the median $r^2$ is
`r sprintf("%.3f", med_r2)` while the median $\eta^2$ is
`r sprintf("%.3f", med_eta2)`. **`r n_hidden`** tables would pass a
correlation screen outright ($r^2 < 0.05$) while the raw score in fact
accounts for more than a quarter of the variance in their ERS index. Those
are precisely the datasets where an analyst using the standard screen would
conclude that response style is a harmless nuisance.

```{r}
#| label: fig-binned-curves
#| fig-cap: "Binned ERS as a function of the raw score, one line per table, with each table's score rescaled onto [0, 1] within its own category range. Faceted by number of response categories, because the arithmetic lower bound (dashed red) is steeper and narrower on scales with more categories. Panel headers give the number of tables. The U is close to universal."
#| fig-height: 6

cat_lab <- summary_df %>%
  count(n_categories, name = "n_tab") %>%
  mutate(facet = paste0(n_categories, " categories (", n_tab, " tables)"))

curves_df <- binned_df %>%
  left_join(select(summary_df, table, n_categories), by = "table") %>%
  left_join(select(cat_lab, n_categories, facet), by = "n_categories")

# Exact bound in rescaled units: on a K-category scale, x = (m - c_min)/(K - 1),
# so the arithmetically forced ERS is max(0, 1 - x(K-1), 1 - (1-x)(K-1)).
bound_df_irw <- cat_lab %>%
  rowwise() %>%
  reframe(
    facet = facet,
    x     = seq(0, 1, length.out = 300),
    bound = pmin(1, pmax(0, 1 - seq(0, 1, length.out = 300) * (n_categories - 1),
                         1 - (1 - seq(0, 1, length.out = 300)) * (n_categories - 1)))
  )

ggplot(curves_df, aes(x = score_rel, y = ers_mean, group = table)) +
  geom_line(alpha = 0.25, colour = irw_blue, linewidth = 0.4) +
  geom_line(data = bound_df_irw, aes(x = x, y = bound, group = 1),
            colour = irw_red, linewidth = 1, linetype = "dashed") +
  facet_wrap(~facet) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Raw score, rescaled within the table's own category range",
       y = "Mean ERS in bin")
```

@fig-binned-curves shows the shape directly. Tables differ in how deep the
trough is and how steep the walls are — that variation is real and is where
any genuine, content-independent style variance would live — but essentially
every table traces the same U, and the floor of that U is the dashed red
line: what arithmetic alone requires. Note that the bound bites hardest on
4-category scales, where a person cannot get their mean below 2 or above 3
without using endpoints at all, and relaxes as categories are added.

<details>
<summary>Show three individual tables at person level</summary>

```{r}
#| label: fig-exemplars
#| fig-cap: "Person-level ERS against raw score for the three tables with the largest gap between eta-squared and r-squared. Dashed red line: the arithmetic lower bound for that table's category range. Black line: loess fit."
#| fig-height: 7

top_gap <- shape_df %>%
  mutate(gap = ers_score_eta2 - r2) %>%
  slice_max(gap, n = 3) %>%
  select(table, gap)

ex_df <- persons_df %>%
  filter(table %in% top_gap$table) %>%
  left_join(select(summary_df, table, min_cat, max_cat), by = "table")

ex_bounds <- ex_df %>%
  distinct(table, min_cat, max_cat) %>%
  purrr::pmap_dfr(function(table, min_cat, max_cat) {
    m <- seq(min_cat, max_cat, length.out = 300)
    tibble(table = table, m = m,
           bound = ers_lower_bound(m, min_cat, max_cat))
  })

ggplot(ex_df, aes(person_mean, ers)) +
  geom_point(alpha = 0.15, colour = irw_blue, size = 0.9) +
  geom_line(data = ex_bounds, aes(m, bound),
            linetype = "dashed", colour = irw_red, linewidth = 0.8) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              colour = "black", linewidth = 0.8) +
  facet_wrap(~table, scales = "free_x", ncol = 1) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Person mean response", y = "Person ERS") +
  theme(strip.text = element_text(size = 9))
```

</details>

### Midpoint responding

MRS shows the mirror-image problem: an inverted U, peaking for people whose
raw score sits near the scale midpoint, and a Pearson correlation that is
near zero for the same symmetry reason.

```{r}
#| label: fig-mrs-curves
#| fig-cap: "Binned MRS against the rescaled raw score, one line per table with a true midpoint (odd number of categories). The inverted U is the MRS analogue of the ERS trough."
#| fig-height: 4.5

mid_tables <- summary_df$table[summary_df$has_midpoint]

binned_df %>%
  filter(table %in% mid_tables, !is.na(mrs_mean)) %>%
  ggplot(aes(x = score_rel, y = mrs_mean, group = table)) +
  geom_line(alpha = 0.2, colour = irw_blue, linewidth = 0.4) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Raw score, rescaled within the table's own category range",
       y = "Mean MRS in bin")
```

<details>
<summary>Show full MRS table (all tables with a true midpoint)</summary>

```{r}
#| label: tbl-mrs
#| tbl-cap: "Tables with an odd number of categories (a true midpoint) support an MRS index; even-category tables are recorded as NA and excluded rather than assigned an arbitrary category. Compare the r-squared and eta-squared columns."

summary_df %>%
  filter(has_midpoint) %>%
  transmute(
    Table = table, `N categories` = n_categories,
    `Mean MRS` = round(mean_mrs, 3), `SD MRS` = round(sd_mrs, 3),
    `MRS-score r` = round(mrs_score_cor, 3),
    `r²` = round(mrs_score_cor^2, 3),
    `η²` = round(mrs_score_eta2, 3)
  ) %>%
  arrange(desc(`Mean MRS`)) %>%
  knitr::kable()
```

</details>

## Why person-mean-centering isn't the fix {#why-person-mean-centering-isnt-the-fix}

A common response to all of this is **ipsatization**: subtract each person's
own mean across items from each of their responses (applied researchers often
also divide by the person's SD), then analyse the centered scores as if the
style component had been removed — typically by fitting a single-factor CFA
before and after and reading the change in standardized loadings as the
footprint of response style. Two things are wrong with that.

**First, it targets the wrong style.** Person-mean-centering removes
differences in *overall response level* — which is acquiescence, or net
agreement bias. It is not a control for ERS or MRS, which are about how much
a person concentrates responses at the endpoints or the middle, not where
their responses sit on average. @savalei2014recovering compare three
approaches for acquiescence specifically — ipsatization following
@chan1993ipsative, a CFA with an acquiescence factor carrying equal loadings,
and an EFA with target rotation — and none of them is put forward as an
ERS/MRS control. A person with high ERS and a high person-mean and a person
with low ERS and the same person-mean are treated identically by
ipsatization.

**Second, the CFA that follows is not identified.** Ipsatized scores sum to
zero within each person by construction, so the $p$ item columns are linearly
dependent and their covariance matrix has rank $p-1$, with an exact zero
eigenvalue. Fitting an ordinary $p$-item factor model to that matrix asks the
estimator to fit a singular covariance structure. @chan1993ipsative give the
proper treatment — drop one item and impose the corresponding constraints on
the retained ones — and @savalei2014recovering implement it. Doing it the
naive way produces numbers, but they are artifacts. Here is the failure, on
clean simulated single-factor data:

```{r}
#| label: ipsatization-failure
#| eval: true
#| echo: true

library(lavaan)

set.seed(1)
n_ips <- 1500; p_ips <- 8
theta_ips <- rnorm(n_ips)
x_ips <- outer(theta_ips, rep(0.7, p_ips)) +
  matrix(rnorm(n_ips * p_ips, 0, sqrt(1 - 0.49)), n_ips, p_ips)
x_ips <- round(scale(x_ips) * 1.1 + 3)
x_ips[x_ips < 1] <- 1; x_ips[x_ips > 5] <- 5

ips <- x_ips - rowMeans(x_ips)          # the ipsatization step

eig <- eigen(cov(ips), only.values = TRUE)$values
mod_ips <- paste0("F =~ ", paste0("i", seq_len(p_ips), collapse = " + "))

d_raw <- as.data.frame(x_ips); names(d_raw) <- paste0("i", seq_len(p_ips))
d_ips <- as.data.frame(ips);   names(d_ips) <- paste0("i", seq_len(p_ips))

fit_raw_ips <- cfa(mod_ips, data = d_raw, std.lv = TRUE)
fit_ips_ips <- suppressWarnings(
  try(cfa(mod_ips, data = d_ips, std.lv = TRUE), silent = TRUE)
)
```

```{r}
#| label: tbl-ipsatization-failure
#| eval: true
#| echo: false
#| tbl-cap: "Standardized loadings from a single-factor CFA on clean simulated data, raw versus naively ipsatized. The true loading is 0.7 for every item and there is no response-style process in the simulation at all."

l_raw_ips <- standardizedSolution(fit_raw_ips) %>%
  filter(op == "=~") %>% pull(est.std)
l_ips_ips <- if (inherits(fit_ips_ips, "try-error")) rep(NA_real_, p_ips) else {
  standardizedSolution(fit_ips_ips) %>% filter(op == "=~") %>% pull(est.std)
}

tibble(
  Item        = paste0("i", seq_len(p_ips)),
  Raw         = round(l_raw_ips, 3),
  Ipsatized   = round(l_ips_ips, 3)
) %>%
  knitr::kable()
```

```{r}
#| label: ipsatization-diagnostics
#| eval: true
#| echo: false
#| output: asis

converged_ips <- !inherits(fit_ips_ips, "try-error") &&
  isTRUE(lavInspect(fit_ips_ips, "converged"))

cat(sprintf(
  paste0("Smallest eigenvalue of the ipsatized covariance matrix: **%.2e** ",
         "(the matrix has rank %d of %d). The ipsatized CFA reports ",
         "`converged = %s`, and its standardized loadings %s.\n"),
  min(eig), qr(cov(ips))$rank, p_ips, converged_ips,
  if (any(abs(l_ips_ips) > 1, na.rm = TRUE))
    "include out-of-bounds values above 1 in absolute terms"
  else "are unstable across items"
))
```

The true loading is 0.7 on every item, and the simulation contains no
response style whatsoever. The raw-data CFA recovers that. The ipsatized fit
returns sign flips and out-of-bounds standardized loadings. Read without the
diagnostics, that pattern is easy to mistake for a dramatic substantive
result — items whose apparent relationship to the construct reverses once
style is "controlled." It is a rank-deficient covariance matrix.

Estimating on data with missing values under FIML is what allows fits like
this to report convergence at all: the missingness perturbs the exact linear
dependency enough that the estimator can limp to a solution. That makes the
artifact harder to notice, not less of an artifact — so convergence is not a
safeguard here.

## What would actually work

Separating style from content requires a model in which both are latent
variables estimated jointly from the full response pattern, rather than a
transformation of the raw scores. @bolt2025review review this literature and
organise it into three families:

- **Multidimensional IRT**, including the multidimensional nominal response
  model (MNRM), where each response style is a latent dimension whose
  *scoring function* encodes which categories it favours — for ERS on a
  7-category item, $\mathbf{s}_{\text{ERS}} = [1\ 0\ 0\ 0\ 0\ 0\ 1]$,
  alongside the substantive dimension's $[0\ 1\ 2\ 3\ 4\ 5\ 6]$
  [@falk2016flexible; @falk2020mnrm].
- **Mixture IRT**, which represents style heterogeneity as discrete latent
  classes of respondents with different item parameters.
- **Multiple-process / IRTree models**, which decompose each response into a
  sequence of sub-decisions — first whether to respond extremely, then in
  which direction — and model each node separately [@bockenholt2012modeling].

@falk2020mnrm is a tutorial for the MNRM route specifically, with code for
`mirt` (scoring functions go in via the `gpcm_mats` argument), flexMIRT and
Mplus; it is also the source of the nonlinearity argument above, and shows
that response-pattern-based EAP estimates from the MNRM break the U-shape
that sum scores cannot escape.

Two cautions worth carrying into that work. @bolt2025review recommend
matching the model to the goal — IRTrees when the interest is in the response
*process*, MIRT or mixtures when style is a nuisance to be controlled — and
argue for dataset-specific sensitivity analyses rather than universal
conclusions about "how much response style matters." And @schoenmakers2024correcting
show that the MNRM and IRTree parametrizations conceptualise ERS differently
enough to reach different conclusions about group differences on the same
data, so the choice of style model is itself a substantive assumption rather
than a technicality.

Fitting one of these across IRW's Likert holdings is a substantially heavier
computation than anything on this page, and is planned as a follow-up pass
rather than something this vignette delivers.

## Limitations

**This vignette diagnoses, it does not correct.** Everything above
establishes that sum-score style indices cannot be separated from content
by any transformation of those same sum scores, and that the standard
correlation screen conceals the problem. It does not estimate how much
genuine, content-independent style variance exists in any IRW table — that
requires the models in the previous section.

**The raw score is an internal criterion.** ERS is compared against the
person's own mean across the same items it is computed from, which is the
right substantive score for a single-factor battery but guarantees some
mechanical dependence. An external validity criterion, independent of the
item set, would be a cleaner comparison and is not available for most IRW
tables.

**The arithmetic bound is a floor, not an explanation.** It shows that *some*
U-shape is unavoidable, and @fig-binned-curves shows observed curves sitting
above it. The gap between the observed curve and the bound is where real
style variance would show up, but attributing that gap to style rather than
to the distribution of the trait requires, again, a model.

**Item-count floor.** Five items per person is a low bar for a proportion.
Tables near that floor have noisy person-level ERS/MRS, which attenuates
$\eta^2$ and $R^2$ alike — so the dependence reported here is, if anything,
understated for short instruments.

**Other approaches not covered.** Factor-analytic treatments of response
styles outside the three IRT families above — including work in the
Lorenzo-Seva and Ferrando line — are not reviewed here and may be relevant
to the same question.

## Reproducibility

{{< include ../components/_source-links.qmd >}}

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the compute script (from the project root)
source("vignettes/response_style_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/response_style.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(summary_df$table, output_file = "response_style_data/irw_references.bib")
```
:::

## Acknowledgements

We are grateful to **Carl F. Falk** (McGill University) for a detailed review
of this vignette — in particular for the points that a correlation cannot
test style-content confounding when the relationship is non-monotone, and
that ipsatization addresses acquiescence rather than ERS or MRS — and for
directing us to the relevant literature. Any remaining errors are ours.

---

## References
