---
title: "Extreme and Midpoint Responding in Likert-Format IRW Data"
description: |
  How prevalent are extreme (ERS) and midpoint (MRS) response styles in
  Likert-format IRW data, and do they distort substantive conclusions --
  like factor loadings -- if left unmodeled? We screen IRW's Likert-scale
  tables (4-7 response categories, at least 5 items), compute person-level
  ERS/MRS indices, and compare single-factor CFA loadings before and after a
  simple ipsatization-based style control.
date: today
bibliography: response_style_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "response_style_data/response_style_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache           <- readRDS(cache_path)
  summary_df       <- .cache$summary
  loadings_df      <- .cache$loadings
  candidate_tables <- .cache$candidate_tables
  date_run         <- .cache$date_run
  draft_from_local <- isTRUE(.cache$DRAFT_FROM_LOCAL_FIXTURES)
  r_all_candidates <- .cache$n_all_candidates %||% length(candidate_tables)
  r_n_tables_run   <- length(candidate_tables)
} else {
  summary_df <- tibble(
    table = character(), construct_type = character(), n_items = integer(),
    n_participants = integer(), min_cat = numeric(), max_cat = numeric(),
    n_categories = numeric(), has_midpoint = logical(), mean_ers = numeric(),
    sd_ers = numeric(), mean_mrs = numeric(), sd_mrs = numeric(),
    ers_score_cor = numeric(), mrs_score_cor = numeric(),
    style_sensitivity = numeric(), cfa_raw_converged = logical(),
    cfa_ips_converged = logical()
  )
  loadings_df <- tibble(table = character(), item = character(),
                         loading_raw = numeric(), loading_ipsatized = numeric())
  candidate_tables <- character()
  date_run <- Sys.Date()
  draft_from_local <- FALSE
  r_all_candidates <- 0L
  r_n_tables_run   <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures and tables will not appear. Generate results ",
  "by running the compute script, then re-render:\n\n",
  "```r\nRscript vignettes/response_style_compute.R\n```\n",
  ":::\n", sep = ""
)
```

```{r}
#| label: draft-warning
#| echo: false
#| eval: true
#| output: asis

if (cache_ok && draft_from_local) cat(
  "::: {.callout-warning}\n",
  "**Draft preview, not final results.** This render uses a local snapshot ",
  "of IRW `.Rdata` fixtures (not a live `irw_filter()`/`irw_fetch()` run ",
  "against Redivis), selected with the same 4-7 category / >=5 item logic ",
  "described below but without the `item_format` tag filter (unavailable ",
  "offline). Numbers here are real computed results from real IRW tables, ",
  "run through the actual `analyze_response_style()` function -- but the ",
  "table set and exact figures will change once `response_style_compute.R` ",
  "is run for real against Redivis. Treat this page as a layout/content ",
  "draft only.\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the
analysis design, and the writing below — was produced largely by Claude
(Anthropic), working from a task specification and with human review.
Treat the methodological choices and interpretations accordingly, and
check the compute script (`response_style_compute.R`) directly if you're
relying on the numbers here.
:::

## Motivation

Likert-type items [@likert1932technique] ask respondents to place themselves
on an ordered scale — "strongly disagree" to "strongly agree," typically with
4 to 7 response categories. The scale is meant to be a transparent window
onto the underlying construct: someone who agrees more strongly with an item
should pick a higher category. In practice, people also differ systematically
in *how* they use the response scale itself, independent of the construct
being measured. Three response styles are especially well documented:

- **Extreme response style (ERS)** — a tendency to favor the endpoints of the
  scale (the lowest and highest categories) regardless of item content.
- **Midpoint response style (MRS)** — a tendency to favor the middle
  category, when one exists (only scales with an odd number of categories
  have a true midpoint).
- **Acquiescence (net agreement bias)** — a tendency toward the "agree" end
  of the scale across items regardless of content, distinct from ERS/MRS but
  related in that it's also a property of *how* someone uses the scale
  rather than *what* they think.

These aren't just a nuisance for individual items — they're a **known
confound**: if response style correlates with the substantive score a scale
is trying to measure (rather than being pure, construct-irrelevant noise),
then style and construct are statistically entangled, and any downstream
model (a factor loading, a group comparison, a validity correlation) that
treats the raw score as a clean measure of the construct is contaminated by
however people use the scale.

This vignette asks three concrete questions of the Likert-format tables in
the IRW:

1. **How common is extreme responding, and does it vary a lot across
   datasets?**
2. **Is response style entangled with the substantive score, or does it
   behave like independent noise?**
3. **Does controlling for style — even with a crude method — change a
   basic substantive conclusion (a factor loading)?**

### See extreme responding in action

Drag the slider to change how much a simulated respondent's answers are
driven by response style rather than by item content. At $p = 0$, every
answer reflects a fixed, moderate underlying attitude (with a little
item-level noise); as $p$ rises, a growing share of answers instead land on
a scale endpoint by chance, regardless of what the item actually asks. The
resulting **ERS** — the proportion of this one simulated respondent's own
answers sitting at an endpoint — is exactly the person-level statistic this
vignette computes for every respondent in every IRW Likert-format table
below.

```{ojs}
viewof style_p = Inputs.range(
  [0, 0.8], {value: 0.2, step: 0.05, label: "Extreme-response tendency (p)"}
)
```

```{ojs}
n_items_demo = 12
rng = d3.randomLcg(42)

demo_responses = {
  const resp = [];
  for (let i = 0; i < n_items_demo; i++) {
    if (rng() < style_p) {
      resp.push(rng() < 0.5 ? 1 : 5);
    } else {
      const honest = Math.round(3 + (rng() - 0.5) * 1.5);
      resp.push(Math.min(5, Math.max(1, honest)));
    }
  }
  return resp;
}

demo_ers = demo_responses.filter((r) => r === 1 || r === 5).length / n_items_demo
```

```{ojs}
Plot.plot({
  width: 700,
  height: 110,
  marginLeft: 40,
  x: {domain: [0.5, n_items_demo + 0.5], label: "Item"},
  y: {domain: [0.5, 5.5], label: "Response (1-5)", ticks: [1, 2, 3, 4, 5]},
  marks: [
    Plot.ruleY([1, 5], {stroke: "#d73027", strokeDasharray: "3,3"}),
    Plot.dot(
      demo_responses.map((r, i) => ({item: i + 1, resp: r})),
      {
        x: "item", y: "resp", r: 6,
        fill: (d) => (d.resp === 1 || d.resp === 5) ? "#d73027" : "#2166ac"
      }
    )
  ]
})
```

```{ojs}
html`<div style="font-size:1.05rem;margin-top:0.5rem;">
  This respondent's <strong>ERS = ${demo_ers.toFixed(2)}</strong> —
  ${Math.round(demo_ers * n_items_demo)} of ${n_items_demo} responses
  (<span style="color:#d73027;">red</span>) landed on an endpoint.
</div>`
```

## Data and methods

### Dataset selection

```{r}
#| label: filter-datasets
#| eval: false

all_candidates <- irw_filter(
  item_format    = "Likert Scale/selected response",
  n_categories   = c(4, 7),
  n_items        = c(5, Inf),
  n_participants = c(100, Inf)
)

# A manageable random sample for this pass, not the full candidate list
tables <- sample(all_candidates, min(150, length(all_candidates)))
```

We restrict to tables tagged as Likert-format selected-response items, with
4 to 7 response categories (covering both odd category counts with a true
midpoint — 5, 7 — and even counts without one — 4, 6), **at least 5 items
per person**, and **at least 100 respondents**. The item-count floor matters
more here than in most IRW vignettes: ERS and MRS are computed as a
*proportion of a person's own responses*, so with very few items per person
that proportion is built from very few trials and is correspondingly noisy
(a person who happens to answer 2 of 3 items at the scale endpoint gets an
ERS of 0.67 by chance alone). Five items is still a fairly low floor — treat
person-level style estimates from tables near it as noisier than those from
longer batteries. The respondent floor rules out tables too small for
table-level ERS/MRS means and CFA fits to be anything but sampling noise.

This selection is deliberately not exhaustive: `r r_all_candidates` tables
met these criteria as of `r format(date_run, "%B %Y")`, and this pass
analyzes a **random sample of `r r_n_tables_run` of them** rather than the
full list — some candidate tables run past 100 items and tens of thousands
of respondents, and running all of them in one pass isn't practical.
`fit_to_disk()`'s on-disk cache (see below) means a later pass with a larger
or different sample adds to what's already here rather than redoing it.

### Per-table computation

For each candidate table, `fit_response_style()`:

1. Fetches the table and builds a wide response matrix (`irw_fetch(table,
   resp = TRUE)`, falling back to a direct `irw_long2resp()` call if that
   internal conversion fails for a given table's shape).
2. Determines the response category range from the data directly (not just
   the table-level metadata) and **drops the table if per-item category
   ranges are inconsistent** — e.g. some items coded 1-4 and others 1-7 —
   since that indicates either a coding problem or a mixed response format
   that a single ERS/MRS definition can't cleanly apply to. Tables dropped
   for this reason are logged (with their observed min/max ranges) in the
   compute script's console output.
3. Computes, per person: ERS (proportion of non-missing responses at the
   minimum or maximum category), MRS (proportion at the midpoint category —
   left `NA` for tables with an even number of categories, since there is no
   midpoint to speak of), and the person's mean response level across items
   (used below both as a net-acquiescence proxy and as the substantive
   "raw score").
4. Correlates ERS and MRS with that raw score across people, within each
   table.
5. Fits a single-factor CFA (`lavaan::cfa()`, as in the
   [CFA tutorial](cfa.qmd)) two ways: once on the raw item responses, and
   once on **ipsatized** responses (each person's own item responses minus
   their own mean across items — a simple, common style control). We
   compare standardized loadings between the two fits and summarize the
   mean absolute change as a table-level **style-sensitivity** metric. If a
   table's CFA doesn't converge cleanly on both the raw and ipsatized data,
   we record `NA` for style sensitivity rather than force a comparison on an
   unstable fit.

```{r}
#| label: fit-function
#| eval: false

analyze_response_style <- function(table_name, wide, tags_meta) {
  # ... category-range consistency check, per-person ERS/MRS/mean level,
  #     ERS/MRS-score correlations, raw vs. ipsatized single-factor CFA ...
}

fit_response_style <- function(table_name) {
  wide <- fetch_wide(table_name)  # irw_fetch(resp = TRUE) with a long2resp fallback
  analyze_response_style(table_name, wide, tags_meta)
}
```

Of the `r r_n_tables_run` sampled tables, `r nrow(summary_df)` produced
usable response-style results after the checks above (the rest were dropped
for an inconsistent category range, too few respondents after cleaning, or a
CFA that didn't converge).

::: {.callout-note}
To reproduce with current IRW holdings, re-run `response_style_compute.R`
and commit the updated `response_style_data/response_style_results.rds`.
:::

## Results

### How common is extreme responding?

```{r}
#| label: fig-ers-ranked
#| fig-cap: "Every table's mean ERS (proportion of responses at a scale endpoint), ranked from lowest to highest. Hover over a point to see which table it is. Error bars show +/- 1 SD across people within each table."
#| fig-height: 6

ers_plot_df <- summary_df %>%
  filter(!is.na(mean_ers)) %>%
  arrange(mean_ers) %>%
  mutate(
    rank    = row_number(),
    tooltip = paste0(
      "Table: ", table,
      "<br>Mean ERS: ", round(mean_ers, 2),
      "<br>SD ERS: ", round(sd_ers, 2),
      "<br>n_items: ", n_items,
      "<br>N: ", n_participants
    )
  )

p <- ggplot(ers_plot_df, aes(x = mean_ers, y = rank, text = tooltip)) +
  geom_errorbarh(aes(xmin = pmax(mean_ers - sd_ers, 0), xmax = pmin(mean_ers + sd_ers, 1)),
                  height = 0, colour = irw_grey) +
  geom_point(colour = irw_blue, size = 2) +
  labs(x = "Mean ERS (proportion of endpoint responses)", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggplotly(p, tooltip = "text") |>
  layout(margin = list(t = 60))
```

If ERS were a small, uniform nuisance, this ranking would be flat and low.
Instead — see the spread in @fig-ers-ranked — a handful of tables sit far to
the right, with a substantial fraction of *all* responses landing on a scale
endpoint, while others sit much lower. Table-level ERS is not a fixed
property of "using a Likert scale" in general; it varies with the
instrument, the sample, and (very plausibly, though we can't isolate it
here) the substantive content being asked about.

### Is response style entangled with the substantive score?

```{r}
#| label: fig-ers-score-cor
#| fig-cap: "ERS-score correlation for each table, plotted against that table's mean ERS. Points far from zero indicate a table where extreme responding is not independent of the substantive score it's measured alongside."
#| fig-height: 5

ggplot(summary_df %>% filter(!is.na(ers_score_cor)),
       aes(x = mean_ers, y = ers_score_cor)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(colour = irw_red, alpha = 0.75, size = 2) +
  labs(x = "Mean ERS", y = "Correlation: person ERS vs. person mean score") +
  coord_cartesian(ylim = c(-1, 1))
```

A correlation near zero in @fig-ers-score-cor is the reassuring case: ERS
behaves like independent, construct-irrelevant noise for that table, and a
substantive score built from the raw responses isn't obviously contaminated
by it. A correlation far from zero is the concerning case — it means people
who respond more extremely also score systematically higher or lower on the
scale itself, so ERS and the construct are statistically entangled rather
than separable. Both patterns appear across IRW's Likert tables; this
variability is itself the point — you cannot assume from a scale's category
count or subject area alone whether style will be a nuisance or a confound in
a given dataset.

<details>
<summary>Show full MRS table (all tables with a true midpoint)</summary>

```{r}
#| label: tbl-mrs
#| tbl-cap: "Tables with an odd number of categories (a true midpoint) support an MRS index; even-category tables are recorded as NA and excluded from this MRS summary rather than assigned an arbitrary category."

summary_df %>%
  filter(has_midpoint) %>%
  transmute(
    Table = table, `N categories` = n_categories,
    `Mean MRS` = round(mean_mrs, 3), `SD MRS` = round(sd_mrs, 3),
    `MRS-score r` = round(mrs_score_cor, 3)
  ) %>%
  arrange(desc(`Mean MRS`)) %>%
  knitr::kable()
```

</details>

### Does controlling for style change the substantive conclusions?

```{r}
#| label: fig-style-sensitivity-ranked
#| fig-cap: "Style-sensitivity metric (mean absolute change in standardized loading, raw vs. ipsatized CFA) across tables where both fits converged cleanly. Hover over a bar to see which table it is."
#| fig-height: 6

sens_df <- summary_df %>%
  filter(!is.na(style_sensitivity)) %>%
  arrange(style_sensitivity) %>%
  mutate(
    rank    = row_number(),
    tooltip = paste0(
      "Table: ", table,
      "<br>Style sensitivity: ", round(style_sensitivity, 3),
      "<br>n_items: ", n_items,
      "<br>N: ", n_participants
    )
  )

p <- ggplot(sens_df, aes(x = style_sensitivity, y = rank, text = tooltip)) +
  geom_col(fill = irw_blue, width = 0.6, orientation = "y") +
  labs(x = "Style sensitivity (mean |Δ standardized loading|)", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggplotly(p, tooltip = "text") |>
  layout(margin = list(t = 60))
```

We highlight the tables with the largest style-sensitivity metric — the
biggest average shift in standardized loadings between the raw and
ipsatized single-factor CFA — to make concrete what "controlling for style
changes the answer" actually looks like at the item level.

```{r}
#| label: fig-loading-examples
#| fig-cap: "Standardized loadings before (raw) and after (person-mean-centered / ipsatized) a simple style control, for the tables with the largest style-sensitivity metric (restricted to tables with 15 or fewer items, so all three panels stay legible at a shared panel height). Ipsatization can flip loading signs, not just shrink them -- see the Limitations section."
#| fig-height: 7

# Capped at <=15 items so three stacked facets (one panel height each) stay
# readable; a table with many more items than the others would otherwise be
# squeezed into the same vertical space and its item labels would overlap.
top_tables <- summary_df %>%
  filter(!is.na(style_sensitivity), n_items <= 15) %>%
  slice_max(style_sensitivity, n = 3) %>%
  pull(table)

example_loadings <- loadings_df %>%
  filter(table %in% top_tables) %>%
  mutate(table = factor(table, levels = top_tables)) %>%
  pivot_longer(cols = c(loading_raw, loading_ipsatized),
               names_to = "model", values_to = "loading") %>%
  mutate(model = recode(model, loading_raw = "Raw", loading_ipsatized = "Ipsatized"))

ggplot(example_loadings, aes(x = loading, y = item, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_line(aes(group = item), colour = irw_grey) +
  geom_point(size = 2.4) +
  facet_wrap(~table, scales = "free_y", ncol = 1) +
  scale_colour_manual(values = c(Raw = irw_grey, Ipsatized = irw_red), name = NULL) +
  labs(x = "Standardized loading", y = NULL) +
  theme(legend.position = "bottom", strip.text = element_text(size = 9))
```

@fig-loading-examples is where "style-sensitivity" stops being an abstract
number and becomes a concrete analysis decision. Read each panel item by
item: the grey dot is an item's loading from the raw responses, the red dot
is the same item's loading after ipsatization, and the line between them is
the size of the shift. In all three tables here the shift isn't confined to
a couple of unstable items — nearly every item flips from a positive raw
loading to a negative ipsatized one, or vice versa (`item_AD4` goes from
-0.53 to +0.78; `item_LOC_1` from -0.74 to +0.59; `item_Empathy1` from +0.71
to -0.50, and almost every other item in each of these three tables does
the same). That is the practical stakes of this vignette's third question:
if you used the raw-score loadings to decide which items are the strongest
measures of the construct — or even just to check that a scale's items all
point the same direction — a style control would hand you the *opposite*
answer for the same data, not because the construct changed, but because
what looked like signal was substantially response style. Which set of
loadings is "correct" isn't something this comparison settles (see the
ipsatization caveat below); the point is that the choice is consequential,
not cosmetic.

## Limitations

**Ipsatization is a crude style control, not a validated one.**
Person-mean-centering removes *all* between-person variance in mean response
level — including genuine, construct-relevant differences in overall
standing, not just style-driven ones. For a single-factor scale, that can
push item covariances negative in a way raw scores never show, which is why
@fig-loading-examples can show loadings that don't just shrink but flip
sign entirely between the raw and ipsatized fits. That isn't a bug in the
comparison; it's the classic, well-documented behavior of ipsative data
(forcing a fixed row sum/mean induces artificial negative dependence among
the remaining items). Treat the *direction and size of the shift* as
evidence that style materially affects the loading structure for that
table, not as evidence that the ipsatized loadings are the "corrected,"
trustworthy ones.

**A more principled alternative: IRTree models.** Tree-based IRT models
(e.g. Böckenholt-style IRTree models) explicitly separate a response-style
process (does this person tend to pick an extreme/midpoint category at all?)
from a content process (given that, do they agree or disagree?), estimated
jointly rather than removed by a blunt transformation of the raw scores.
That's a heavier modeling commitment than this vignette makes — it requires
specifying a tree structure per item type and fitting a multidimensional
IRT model — and is flagged here as a natural extension rather than something
implemented in this pass.

**Other scope limitations.** The category-range consistency check drops any
table where items don't share a single response format, which likely
removes some legitimately mixed-format instruments along with genuinely
miscoded ones — we can't distinguish the two from the data alone. The
ERS/MRS-score correlation uses the person's own mean item response as the
"score," which is appropriate for a single-factor instrument but conflates
score and style more directly than an external criterion would; a validity
correlate independent of the item set itself would be a stronger test of
confounding, and is left for future work. Style-sensitivity is computed only
where both CFA fits converge cleanly — tables where a single-factor model is
a poor fit to begin with are silently excluded from that specific
comparison (though they still contribute to the ERS/MRS distributions
above).

## Reproducibility

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
# 1. Re-run the compute script (from the project root)
source("vignettes/response_style_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/response_style.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(summary_df$table, output_file = "response_style_data/irw_references.bib")
```
:::

---

## References
