---
title: "Does Robust Polychoric Correlation Disagree with Maximum Likelihood on IRW Data?"
description: |
  Polychoric correlation assumes every respondent's answers come from one
  underlying bivariate normal. A robust estimator (Welz, Mair & Alfons, 2026)
  relaxes that for a minority of observations and is asymptotically equivalent
  to maximum likelihood when the assumption holds -- so a gap between the two
  is a diagnostic, not a disagreement about which is right. We scan IRW's
  ordinal tables for that gap.
date: today
bibliography: robust_polychoricdata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "robust_polychoricdata/robust_polychoric_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache      <- readRDS(cache_path)
  summary_df  <- .cache$summary
  pairs_df    <- .cache$pairs
  failed_df   <- .cache$failed
  date_run    <- .cache$date_run
  tuning_c    <- .cache$tuning_c
  max_cats    <- .cache$max_categories
  items_cap   <- .cache$items_cap
  n_cand      <- .cache$n_all_candidates
  n_attempted <- length(.cache$candidate_tables)
} else {
  summary_df <- tibble(); pairs_df <- tibble(); failed_df <- tibble()
  date_run <- Sys.Date(); tuning_c <- 0.6; max_cats <- 5; items_cap <- 15
  n_cand <- 0L; n_attempted <- 0L
}

knitr::opts_chunk$set(eval = cache_ok)
```

```{r}
#| label: cache-warning
#| echo: false
#| eval: true
#| output: asis

if (!cache_ok) cat(
  "::: {.callout-warning}\n",
  "**Cache not found** — figures will not appear. Generate results with:\n\n",
  "```bash\nPILOT=FALSE Rscript vignettes/robust_polychoric_compute.R\n```\n",
  ":::\n", sep = ""
)
```

::: {.callout-note}
This vignette — including the dataset search, the compute script, the analysis
design, and the writing below — was produced largely by Claude (Anthropic),
working from a task specification and with human review. Check the compute
script (`robust_polychoric_compute.R`) directly if you're relying on the
numbers here.
:::

## Motivation

Polychoric correlation [@olsson1979maximum] is the workhorse for ordinal data
in psychometrics: it estimates what the correlation between two items *would*
be if the discrete responses were coarsenings of two continuous, jointly
normal latent variables. Nearly every ordinal CFA, ordinal SEM, and
factor-analytic scale evaluation in the field rests on it, usually without
comment.

The assumption is strong, and it is assumed to hold for **every respondent**.
One bivariate normal is supposed to have generated the whole table. That is
exactly the kind of assumption real survey data strains: careless responders,
bots, people misreading a reverse-worded item, or an attentive but genuinely
different subgroup all produce response patterns the single-normal model
cannot fit.

@welz2026robust propose a robust alternative — a "C-estimator" — designed for
partial misspecification: the model is taken to hold for most observations but
not necessarily all. The estimator automatically downweights cells of the
contingency table whose observed frequencies the polychoric model cannot fit,
governed by a tuning constant $c$ (default `r tuning_c`; $c = \infty$ recovers
ordinary maximum likelihood).

The property that makes this useful as a *diagnostic* rather than a
replacement is this: **when the model is correctly specified for all
observations, the robust estimator is asymptotically equivalent to maximum
likelihood.** Agreement is therefore the null hypothesis. A gap between the
two estimates is not a disagreement about which number is right — it is
evidence that some observations are not described by the single-bivariate-normal
model that ML assumes throughout.

That reframes the question this vignette can ask of IRW. Not "which estimator
should you use," but:

1. **How often, and how much, do the two estimators diverge across real
   ordinal IRW tables?**
2. **Is the divergence one-directional?** If contamination attenuates ML,
   the robust estimate should sit systematically *above* it, not scatter
   symmetrically around it.
3. **What does the divergence look like at the cell level** for the pair where
   it is largest — which response combinations drive it?

## Data and methods

### Which tables, and why the category ceiling matters

```{r}
#| label: selection
#| eval: false

all_candidates <- irw_filter(
  n_categories   = c(3, 5),
  n_items        = c(6, Inf),
  n_participants = c(1000, Inf)
)
```

We take IRW tables with **3 to 5 response categories**, at least 6 items, and
at least 1,000 respondents; `r n_cand` tables qualified as of
`r format(date_run, "%B %Y")`, and this pass analyzes a random sample of
`r n_attempted`. Two of those bounds are not the obvious ones and deserve
explanation.

**The upper category bound is a computational necessity, not a modelling
choice.** The robust estimator's cost is driven overwhelmingly by the number
of response categories, and the growth is a cliff rather than a slope. Timing
a single item pair at 1,000 respondents:

| Categories | Robust | ML (`psych::polychoric`) | Ratio |
|---|---|---|---|
| 3 | 0.18 s | 0.012 s | 15× |
| 4 | 0.59 s | 0.014 s | 42× |
| 5 | 1.11 s | 0.020 s | 56× |
| 7 | 50.37 s | 0.040 s | 1259× |

At 5 categories a 15-item table is a few minutes; at 7 it is most of a day.
An earlier attempt at this vignette was abandoned as infeasible after
benchmarking a 7-category table — the worst case in the range — and
concluding the estimator was uniformly ~1000× too slow. It isn't; that
conclusion was an artifact of the table chosen. **This does mean 6- and
7-category instruments are out of scope here**, which is a real limitation:
if divergence behaves differently on longer scales, this vignette cannot see
it.

**The respondent floor is high for the opposite of the usual reason.** Small
samples are *more* expensive here, not less: at 5 categories the same pair took
8.4 s at N=500 against ~1.3 s at N=1000, 2000 and 5000, because sparse cells
make the estimator struggle to converge. So respondents are never subsampled —
every table is analyzed at full N — and the 1,000 floor keeps the estimator
out of the regime where it is both slowest and least stable.

### Per-table computation

Following the analysis pattern in the paper's own
[replication material](https://github.com/mwelz/robust-polycor-replication),
for each table we build a complete-case response matrix, sample at most
`r items_cap` items (so at most `r choose(items_cap, 2)` pairs), recode
categories to consecutive integers, and fit both estimators to the *same*
matrix:

```{r}
#| label: fit-pattern
#| eval: false

fit_mle <- polycormat_mle(mat, variance = FALSE)              # maximum likelihood
fit_rob <- polycormat(mat, c = 0.6, variance = FALSE,         # robust C-estimator
                      parallel = FALSE)
diff <- fit_rob$cormat - fit_mle$cormat                       # signed, per pair
```

Both estimators see identical data, so any difference is attributable to the
estimator rather than to differing exclusions. The run is deliberately serial.

## Results

```{r}
#| label: headline
#| echo: false

n_tab   <- nrow(summary_df)
n_pairs <- nrow(pairs_df)
med_abs <- if (n_pairs) median(pairs_df$abs_diff, na.rm = TRUE) else NA
max_abs <- if (n_pairs) max(pairs_df$abs_diff, na.rm = TRUE) else NA
pct_pos <- if (n_pairs) 100 * mean(pairs_df$diff > 0, na.rm = TRUE) else NA
```

`r n_tab` tables produced usable results, giving `r n_pairs` item pairs with
both estimates.

### How large is the gap?

```{r}
#| label: fig-diff-hist
#| fig-cap: "Signed difference between the robust and ML polychoric estimate, one observation per item pair, pooled across tables. Zero (dashed) is what the null of a correctly specified model predicts."
#| fig-height: 4

ggplot(pairs_df, aes(x = diff)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_histogram(bins = 60, fill = irw_blue, colour = NA) +
  labs(x = expression(rho[robust] - rho[ML]), y = "Item pairs")
```

The median absolute gap is `r sprintf("%.3f", med_abs)` and the largest is
`r sprintf("%.3f", max_abs)`. For context, @welz2026robust report differences
"as much as 0.3" in their own Big Five application.

### Is the divergence one-directional?

```{r}
#| label: fig-scatter
#| fig-cap: "Robust against ML estimate for every item pair. The diagonal is exact agreement. Systematic displacement to one side is the signature of contamination attenuating ML, rather than noise."
#| fig-height: 5.2

p <- ggplot(pairs_df, aes(x = rho_mle, y = rho_rob,
                          text = paste0(table, "<br>", item_x, " / ", item_y,
                                        "<br>ML: ", round(rho_mle, 3),
                                        "<br>robust: ", round(rho_rob, 3)))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(alpha = 0.35, colour = irw_blue, size = 1.4) +
  labs(x = expression(rho[ML]), y = expression(rho[robust]))

ggplotly(p, tooltip = "text")
```

`r sprintf("%.1f%%", pct_pos)` of pairs have a robust estimate *above* the ML
estimate. Under a correctly specified model the split would be near 50/50, so
a strong asymmetry is the informative result: it is what contamination looks
like when it attenuates the ML estimate and downweighting restores it.

### Which tables diverge most?

```{r}
#| label: tbl-ranking
#| tbl-cap: "Tables ranked by the largest absolute robust-vs-ML gap among their sampled item pairs."

summary_df %>%
  arrange(desc(max_abs_diff)) %>%
  transmute(
    Table = table, Items = n_items, N = n_respondents, Pairs = n_pairs,
    `Max |diff|`    = round(max_abs_diff, 3),
    `Median |diff|` = round(median_abs_diff, 3),
    `Mean signed`   = round(mean_diff, 3)
  ) %>%
  head(20) %>%
  knitr::kable()
```

## Limitations

**Scope is capped at 5 response categories** for the computational reasons
given above. Six- and seven-point scales are extremely common in practice, and
nothing here speaks to them.

**Complete-case matrices.** Both estimators are fit to listwise-complete data,
so tables with scattered missingness lose respondents and some are dropped
entirely for falling under the floor. This selects toward completely-answered
instruments, which may itself be related to careless responding — the very
thing the robust estimator is meant to detect.

**Divergence localizes assumption failure; it does not identify careless
respondents.** The cell-level Pearson residuals are a property of an item
pair's contingency table, not a per-person score. The paper's own replication
code flags outlying *cells*; attributing those to particular respondents is an
inference beyond what this analysis supports.

**Both estimators can fail together.** @welz2026robust are explicit (their
Section 8) that when the latent distribution is nonnormal for *all*
observations rather than a contaminated minority, neither ML nor the robust
estimator recovers the truth; the robust one helps mainly when nonnormality
concentrates in the tails. Agreement between the two is therefore not proof
that the model is right.

**One tuning constant.** Everything here uses $c = `r tuning_c`$, the package
default and the paper's recommendation. Divergence is a function of $c$, and
no sensitivity analysis over it was run.

## Reproducibility

{{< include ../components/_source-links.qmd >}}

Results computed `r format(date_run, "%B %d, %Y")`. To regenerate:

```bash
PILOT=FALSE Rscript vignettes/robust_polychoric_compute.R
quarto render vignettes/robust_polychoric.qmd
```

## Acknowledgements

We thank **Max Welz** (Erasmus University Rotterdam) for correspondence
clarifying what the robust estimator is and is not designed to do — in
particular that it functions as a sanity check on modelling assumptions rather
than as a competitor to maximum likelihood, and that both estimators can fail
together under global distributional misspecification. The analysis here
follows the pattern in the paper's own replication material. He has not
reviewed this page.

## References
