---
title: "Response Time and Accuracy: An IMV Analysis"
description: |
  How much does knowing how fast a person responded improve predictions of
  whether they answered correctly? We fit Rasch IRT to binary accuracy data
  from IRW datasets with response times, then quantify the gain from adding
  within-item-centered log(RT) using the InterModel Vigorish (IMV).
  Motivated by Domingue et al. (2022).
date: today
bibliography: rtimvdata/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(plotly)

theme_set(theme_minimal(base_size = 13))
irw_blue <- "#2166ac"
irw_red  <- "#d73027"
irw_grey <- "grey50"

cache_path <- "rtimvdata/rt_imv_results.rds"
cache_ok   <- file.exists(cache_path)

if (cache_ok) {
  .cache   <- readRDS(cache_path)
  results  <- .cache$results |> filter(N >= 150)
  date_run <- .cache$date_run
} else {
  results  <- tibble(table_name = character(), N = integer(), J = integer(),
                     n_obs = integer(), imv_linear = numeric(),
                     imv_spline = numeric(), imv_flex = numeric())
  date_run <- Sys.Date()
}

n_datasets <- nrow(results)
```

## Motivation

IRT models predict binary accuracy from a latent ability estimate — but they
ignore *when* a person responded. Response time (RT) carries its own signal
about item-person fit: very fast incorrect responses may reflect guessing, while
slow correct responses may reflect effortful retrieval. The question is whether
RT adds *predictive* value over and above the IRT probability.

---

## Data and methods

We answer this using the **InterModel Vigorish** [IMV; @domingue2022speed], a
log-score measure of how much one model's predictions improve on another's. For
each IRW dataset with RT and binary accuracy, we compare three nested models:

- **M0**: `resp ~ (1|item) + (1|id)` — a random-item Rasch model [@deboeck2011lmer]
  with crossed random effects for item difficulty and person ability, no RT
- **M1**: `resp ~ rt_cwi + (1|item) + (1|id)` — add within-item-centered log(RT), linear
- **M2**: `resp ~ spl1 + spl2 + spl3 + spl4 + (1|item) + (1|id)` — add flexible spline RT

`rt_cwi = log(rt) − mean_item(log(rt))` centers RT within items, isolating
the *micro-speed–accuracy tradeoff*: is this person faster or slower than
average on this item, and does that predict correctness? The random effects
for item and person replace a separately-fitted Rasch model, estimating item
difficulty and person ability jointly with the RT effect.

`spl1`–`spl4` are the four columns of a B-spline basis (`bs(rt_cwi, df=4)`)
computed from `rt_cwi` on the full dataset and stored as pre-computed columns.
Precomputing the basis (rather than evaluating `bs()` inline) ensures that
`imv()`'s internal model updates operate on the same feature matrix.

### Datasets

Datasets are restricted to those with at least 150 respondents (`N >= 150`)
for stable random-effects estimation.

```{r}
#| label: tbl-datasets
#| tbl-cap: "IRW datasets used in this analysis. All have binary accuracy responses and response times."

results |>
  select(table_name, N, J, n_obs) |>
  arrange(desc(n_obs)) |>
  knitr::kable(
    col.names = c("Dataset", "N persons", "J items", "Obs."),
    format.args = list(big.mark = ",")
  )
```

## Results

### Does RT improve predictions over IRT?

```{r}
#| label: fig-imv-main
#| fig-cap: "IMV gain from adding RT to IRT predictions. Blue = linear RT (M0→M1); red = spline RT (M0→M2). Positive values indicate RT improves predictions. Datasets ordered by spline IMV."
#| fig-height: 4.5

plot_df <- results |>
  mutate(table_name = reorder(table_name, imv_spline)) |>
  pivot_longer(c(imv_linear, imv_spline),
               names_to = "comparison", values_to = "imv") |>
  mutate(
    comparison = recode(comparison,
      imv_linear = "Linear RT  (M0 → M1)",
      imv_spline = "Spline RT  (M0 → M2)"
    ),
    tooltip = paste0(
      "Dataset: ", table_name,
      "<br>Comparison: ", comparison,
      "<br>IMV: ", round(imv, 4),
      "<br>N: ", N, "  J: ", J
    )
  )

p <- ggplot(plot_df, aes(x = imv, y = table_name, colour = comparison, text = tooltip)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = irw_grey) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  scale_colour_manual(
    values = c("Linear RT  (M0 → M1)" = irw_blue,
               "Spline RT  (M0 → M2)" = irw_red),
    name = NULL
  ) +
  labs(x = "IMV gain over IRT-only baseline", y = NULL) +
  theme(legend.position = "bottom")

ggplotly(p, tooltip = "text") |>
  layout(
    legend = list(orientation = "h", y = -0.25),
    margin = list(b = 80)
  )
```

```{r}
#| label: summary-stats
#| echo: false

med_linear <- median(results$imv_linear, na.rm = TRUE)
med_spline <- median(results$imv_spline, na.rm = TRUE)
pct_pos_linear <- mean(results$imv_linear > 0, na.rm = TRUE) * 100
pct_pos_spline <- mean(results$imv_spline > 0, na.rm = TRUE) * 100
```

Across `r n_datasets` datasets, the median IMV gain from adding linear RT is
`r round(med_linear, 4)` (`r round(pct_pos_linear, 0)`% of datasets positive);
the spline gain is `r round(med_spline, 4)`
(`r round(pct_pos_spline, 0)`% positive).

---

## Discussion

**RT adds modest but consistent predictive value.** Positive IMV for both
linear and spline RT indicates that response time generally adds value in prediction of accuracy.

**In some cases, the gains come from adding RT nonlinearly.** RT can be a highly nonlinear predictor of response time. Indeed, the relationship between response time and accuracy may even be nonmonotonic.

---

## Limitations

**Only a narrow slice of the RT signal is tested here.** `rt_cwi` isolates the
*within-item* micro speed-accuracy tradeoff (is this person faster or slower
than average *on this item*); it doesn't examine between-item RT variation or
a person's overall response speed across the whole test, either of which
could carry additional predictive signal not captured by this comparison.

**The IRT baseline is Rasch, not 2PL.** M0 fixes equal item discrimination
(a random-item Rasch model). Whether RT's predictive gain changes once
discrimination is allowed to vary by item — as in the [2PL
vignette](2pl_across_datasets.qmd) — is not tested here.

**Small datasets are excluded.** The `N >= 150` floor drops any dataset too
small for stable random-effects estimation, so the median IMV gains above
describe the datasets that clear that bar, not IRW's full RT-bearing holdings.

---

## Reproducibility

Results computed `r format(date_run, "%B %d, %Y")`. To regenerate:

```r
Rscript vignettes/rt_imv_compute.R
quarto::quarto_render("vignettes/rt_imv.qmd")
```

---

## References
