---
title: "Speeded vs. Power Tests in the IRW"
description: |
  Which IRW instruments show evidence of a fixed time limit (speeded tests)
  versus untimed administration (power tests)? We screen response-time-carrying
  IRW tables for the classic speededness signature — a rising rate of
  unanswered items toward the end of the test — and ask what it implies for
  how each table's item responses should be modeled.
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| echo: false

library(dplyr)
library(ggplot2)

# Load precomputed results (produced by speededness_compute.R)
.cache              <- readRDS("speededness_data/speededness_results.rds")
table_summary       <- .cache$table_summary
item_level_results  <- .cache$item_level_results
candidate_tables    <- .cache$candidate_tables
```

## Research question

Classical test theory draws a sharp distinction between **power tests**, where
every respondent has enough time to attempt every item, and **speeded tests**,
where a fixed time limit means slower respondents run out of time before
reaching the end. The distinction matters beyond test administration: speeded
data violate the local independence assumption that most IRT models rely on.
When items are left blank because time ran out — not because the respondent
didn't know the answer — item difficulty and discrimination estimates near
the end of the test become entangled with time pressure rather than reflecting
the construct alone.

Most IRW tables don't come labeled "speeded" or "power." But the two designs
leave different fingerprints in the data: speeded tests show a rising rate of
missing (not-reached) responses as item position increases, while power tests
show a flat, position-independent missingness rate. We use this signature to
screen the response-time-carrying tables in the IRW and classify each one.

---

## Step 1: Select candidate tables

We use `irw_filter()` to find tables that carry a response-time (`rt`)
variable at all, with at least 10 items (fewer than that, a position trend
isn't very informative). We also disable `irw_filter()`'s default density
filter (`density = NULL`): its default keeps only tables with 50-100%
response density, but speededness *is* elevated missingness — screening on
density up front would exclude exactly the tables most likely to show the
signature we're looking for.

```{r}
#| label: filter-datasets
#| eval: false

candidate_tables <- irw_filter(var = "rt", n_items = c(10, Inf), density = NULL)
```

`irw_filter()` only knows metadata-level facts — it can't tell us whether
`rt` is recorded per-response or is really a single total-test-time value
copied onto every row for a person, and it can't tell us whether `rt` is in
seconds or milliseconds. Both are checked after fetching each table, and
tables that fail either check are dropped (see Step 2).

This selected `r length(candidate_tables)` candidate tables from the IRW
(as of `r format(.cache$date_run, "%B %Y")`), of which
`r nrow(table_summary)` had usable response and RT data and were classified.

::: {.callout-note}
To reproduce with current IRW holdings, re-run `speededness_compute.R` and
commit the updated `speededness_data/speededness_results.rds`.
:::

---

## Step 2: Per-table speededness diagnostics

For each candidate table, `fit_speededness()`:

1. Fetches the table and sanity-checks `rt` — drops non-positive values,
   converts millisecond-scale `rt` to seconds, and drops tables where `rt`
   looks like a single total-test-time value (more than 90% of respondents
   have only one distinct `rt` value) rather than a genuine per-response
   measurement.
2. Derives an **item position** variable. When a `date` timestamp varies
   meaningfully within respondents, position is each item's median rank in
   that within-person timestamp order. Otherwise, position falls back to
   order of first appearance in the fetched data — a real limitation, since
   the IRW standard doesn't guarantee that row order reflects presentation
   order. Tables using the fallback are flagged via `order_source` in the
   results and should be read with that caveat in mind.
3. Computes the **omission trend**: proportion of missing responses at each
   item position, regressed on position. A positive, permutation-significant
   slope (position labels shuffled 500 times) is the classic speededness
   signature — unfinished items cluster at the end rather than scattering
   randomly — and tags the table `"likely speeded"`; otherwise `"likely power"`.
4. Computes secondary diagnostics: the **RT-by-position trend** (declining RT
   near the end suggests rushing), the **accuracy-by-position trend**
   (reported as a raw diagnostic — it's confounded with any difficulty
   ordering of items and isn't used for classification), and the
   **person-level speed-accuracy correlation** (total RT vs. total score;
   near zero suggests power, positive suggests a speed-accuracy tradeoff
   consistent with time pressure).

```{r}
#| label: fit-function
#| eval: false

fit_speededness <- function(table_name) {
  df <- irw_fetch(table_name)
  # ... rt sanity checks, item-position derivation ...

  omit_by_pos <- df |>
    group_by(position) |>
    summarise(p_missing = mean(is.na(resp)), .groups = "drop")
  omit_slope <- coef(lm(p_missing ~ position, data = omit_by_pos))[["position"]]

  # permutation test: shuffle position labels, refit, compare slope magnitude
  perm_slopes <- replicate(500, {
    shuffled <- omit_by_pos
    shuffled$position <- sample(shuffled$position)
    coef(lm(p_missing ~ position, data = shuffled))[["position"]]
  })
  omit_perm_p <- mean(abs(perm_slopes) >= abs(omit_slope))

  classification <- if (omit_slope > 0 && omit_perm_p < 0.05) "likely speeded" else "likely power"
  # ... RT trend, accuracy trend, person-level speed-accuracy correlation ...
}

# Run across all candidate tables; each result written to disk as it completes
# (see speededness_compute.R for full details, including the rt plausibility
# checks and construct_type join)
```

---

## Step 3: Results

### Distribution of omission slopes

```{r}
#| label: fig-omit-slopes
#| fig-cap: "Omission-by-position slope for each classified table, colored by classification. Positive slopes indicate a rising rate of unanswered items toward the end of the test."
#| fig-height: 4

ggplot(table_summary, aes(x = omit_slope, fill = classification)) +
  geom_histogram(bins = 30, alpha = 0.85, position = "identity") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("likely speeded" = "#b2182b", "likely power" = "#2166ac")) +
  labs(
    x    = "Omission-by-position slope",
    y    = "Number of tables",
    fill = NULL
  ) +
  theme_minimal(base_size = 13)
```

### Summary statistics

```{r}
#| label: tbl-summary
#| tbl-cap: "Classification counts and summary statistics across screened IRW tables"

table_summary |>
  group_by(classification) |>
  summarise(
    Tables            = n(),
    `Median items`    = round(median(n_items), 1),
    `Median N`        = round(median(n_participants), 0),
    `Median RT slope` = round(median(rt_slope, na.rm = TRUE), 4),
    `Median speed-accuracy r` = round(median(speed_acc_cor, na.rm = TRUE), 3),
    .groups = "drop"
  )
```

### Example tables

We highlight the most clearly speeded table (steepest positive, significant
omission slope) and the flattest power table (omission slope closest to zero)
among those classified.

```{r}
#| label: fig-examples
#| fig-cap: "Proportion of missing responses by item position for a clearly speeded table (top) and a clearly power table (bottom)."
#| fig-height: 7

speeded_example <- table_summary |>
  filter(classification == "likely speeded") |>
  slice_max(omit_slope, n = 1, with_ties = FALSE) |>
  pull(table)

power_example <- table_summary |>
  filter(classification == "likely power") |>
  slice_min(abs(omit_slope), n = 1, with_ties = FALSE) |>
  pull(table)

example_data <- item_level_results |>
  filter(table %in% c(speeded_example, power_example)) |>
  mutate(
    label = case_when(
      table == speeded_example ~ paste0(table, " (likely speeded)"),
      table == power_example   ~ paste0(table, " (likely power)")
    )
  )

ggplot(example_data, aes(x = position, y = p_missing)) +
  geom_point(colour = "#2166ac") +
  geom_smooth(method = "lm", se = FALSE, colour = "#b2182b", linewidth = 0.8) +
  facet_wrap(~label, ncol = 1, scales = "free_y") +
  labs(x = "Item position", y = "Proportion missing") +
  theme_minimal(base_size = 13)
```

### Classification by construct type

Does speededness cluster in cognitive/ability instruments rather than
affective or survey-type constructs? This is a plausible pattern — cognitive
tests are more often administered under time pressure — but not an assumed
one; we report what the classified tables show.

```{r}
#| label: tbl-crosstab
#| tbl-cap: "Classification by construct type"

table_summary |>
  filter(!is.na(construct_type)) |>
  count(construct_type, classification) |>
  tidyr::pivot_wider(names_from = classification, values_from = n, values_fill = 0)
```

---

## Why this matters

Speeded data violate local independence: an item's missingness (or, before
that, the accuracy of a rushed final answer) reflects the respondent's
position in the test and remaining time, not just their standing on the
measured construct. Fitting a standard IRT model to speeded data without
accounting for this can bias item parameters — particularly for items near
the end of the test, where difficulty and discrimination estimates become
entangled with time-pressure effects rather than reflecting the construct
alone. Tables flagged `"likely speeded"` here are candidates for models that
explicitly separate a speed and an accuracy process, or for pre-processing
that treats not-reached items differently from items the respondent actually
attempted and got wrong.

**Limitations.** Where no natural presentation-order variable (`date`) is
available, item position falls back to order of first appearance in the
fetched data, which need not reflect the order items were actually
administered — the `order_source` column in the results flags which tables
rely on this fallback. The accuracy-by-position trend is reported as a raw
diagnostic only; it is confounded with any difficulty ordering of items and
does not factor into classification. Classification itself is a screen, not
a definitive test: the permutation-based significance threshold (p < .05)
will misclassify some tables, especially ones near the boundary between
speeded and power administration.

---

## Reproducibility

Results on this page were computed on `r format(.cache$date_run, "%B %d, %Y")`
using `r length(candidate_tables)` candidate IRW tables. To reproduce:

```r
# 1. Re-run the compute script (from project root)
source("vignettes/speededness_compute.R")

# 2. Re-render this page
quarto::quarto_render("vignettes/speededness.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
for (tbl in table_summary$table) {
  irw_save_bibtex(tbl, output_file = "speededness_data/irw_references.bib", append = TRUE)
}
```
:::
