---
title: "Binomial Approximations to the GPCM: A Simulation"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 8
execute:
  echo: false
  message: false
  warning: false
---

```{r setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(ggrepel)
  library(dplyr)
  library(mgcv)
})

res <- readRDS("gpcm_binomial_results.rds")

# ── Model probability helpers (used for ERF illustration) ─────────────────────
gpcm_pr_fn <- function(theta, a, b) {
  K <- length(b) + 1L; ln <- matrix(0, nrow = length(theta), ncol = K)
  for (k in seq_len(K - 1L)) ln[, k + 1L] <- ln[, k] + a * (theta - b[k])
  ln <- ln - apply(ln, 1, max); ex <- exp(ln); ex / rowSums(ex)
}
grm_pr_fn <- function(theta, a, b) {
  K <- length(b) + 1L
  cum <- cbind(1, plogis(outer(theta, b, function(th, bk) a * (th - bk))), 0)
  pmax(cum[, 1:K] - cum[, 2:(K + 1L)], 1e-15)
}
binom_pr_fn <- function(theta, a, b, K, link) {
  eta <- a * theta - b
  p <- switch(link, logit = plogis(eta), probit = pnorm(eta),
              cloglog = 1 - exp(-exp(eta)), cauchit = 0.5 + atan(eta) / pi)
  p <- pmin(pmax(p, 1e-10), 1 - 1e-10)
  outer(p, 0:(K - 1L), function(pp, k) dbinom(k, K - 1L, pp))
}
es_fn <- function(pr, K) as.numeric(pr %*% 0:(K - 1L))

model_labels <- c(
  gpcm              = "GPCM",
  pcm               = "PCM",
  grm               = "GRM",
  binom_2pl_logit   = "Binom 2PL (logit)",
  binom_1pl_logit   = "Binom 1PL (logit)",
  binom_1pl_cloglog = "Binom 1PL (cloglog)",
  binom_1pl_cauchit = "Binom 1PL (cauchit)",
  binom_1pl_probit  = "Binom 1PL (probit)"
)
model_colors <- c(
  "GPCM"               = "#1b7837",
  "PCM"                = "#762a83",
  "GRM"                = "#2166ac",
  "Binom 2PL (logit)"  = "#d6604d",
  "Binom 1PL (logit)"  = "#f4a582",
  "Binom 1PL (cloglog)"= "#c2a5cf",
  "Binom 1PL (cauchit)"= "#a6dba0",
  "Binom 1PL (probit)" = "#bababa"
)

res <- res |>
  mutate(
    model_label = factor(model_labels[as.character(model)], levels = model_labels),
    K_label     = paste0("K = ", K),
    a_label     = paste0("a = ", a_true)
  )

lpos_fn <- function(data, y_var) {
  data |>
    group_by(model_label, K_label, a_label) |>
    filter(asym >= quantile(asym, 0.9)) |>
    summarise(asym = mean(asym),
              y_val      = mean(.data[[y_var]], na.rm = TRUE),
              .groups    = "drop")
}

theme_set(
  theme_bw(base_size = 11) +
    theme(strip.background = element_rect(fill = "grey92"),
          legend.position  = "none",
          plot.margin      = margin(5, 5, 5, 5))
)

make_plot <- function(data, y_var, y_lab) {
  lpos <- lpos_fn(data, y_var)
  ggplot(data, aes(x = asym, y = .data[[y_var]],
                   colour = model_label, group = model_label)) +
    annotate("segment", x = -log(8), xend = log(8), y = 0, yend = 0,
             linetype = "dashed", colour = "grey60") +
    geom_smooth(data = function(d) filter(d, model_label != "GPCM"),
                method = "gam", formula = y ~ s(x, bs = "cs"),
                se = TRUE, linewidth = 0.8, alpha = 0.12) +
    annotate("rect", xmin = log(8), xmax = log(8) + 3, ymin = -Inf, ymax = Inf,
             fill = "white", colour = NA) +
    geom_label_repel(
      data          = lpos,
      aes(x = asym, y = y_val, label = model_label),
      size          = 2.6, hjust = 0,
      nudge_x       = log(8) + 0.3 - lpos$asym,
      direction     = "y", segment.size = 0.3, segment.color = "grey50",
      box.padding   = 0.2, label.padding = 0.15,
      show.legend   = FALSE, xlim = c(log(8), log(8) + 3)
    ) +
    facet_grid(K_label ~ a_label) +
    scale_colour_manual(values = model_colors) +
    scale_x_continuous(
      name   = "Gap asymmetry  (0 = equal gaps;  negative = wider lower;  positive = wider upper)",
      limits = c(-log(8), log(8) + 3),
      breaks = round(c(-log(8), -1, 0, 1, log(8)), 2),
      labels = c("-ln8\n(wider lower)", "-1", "0\n(symmetric)", "1",
                 "ln8\n(wider upper)"),
      expand = expansion(mult = c(0.02, 0))
    ) +
    coord_cartesian(clip = "off") +
    ylab(y_lab)
}
```

## Overview

The Generalized Partial Credit Model (GPCM) is a natural choice for polytomous item response data, but simpler *binomial approximations* are sometimes used instead: the polytomous response $X \in \{0,\ldots,K{-}1\}$ is modeled as if it were the outcome of $K{-}1$ independent Bernoulli trials, i.e., $X \sim \text{Binomial}(K{-}1,\, p)$, where $p$ is a function of the latent trait through a chosen link function.

This simulation asks: **when do binomial approximations pay a meaningful cost relative to the true GPCM?** We quantify that cost using IMV(model, GPCM) — the InterModel Vigorish of GPCM over each alternative — where GPCM's fitted probabilities serve as the reference. Positive IMV means GPCM predicts better; negative means the alternative wins.

---

## Simulation Setup

### Data generating model

Responses are generated from the GPCM. For a $K$-category item with discrimination $a$ and ordered thresholds $b_1 < \cdots < b_{K-1}$, the probability of response $k$ given latent trait $\theta$ is

$$
P(X = k \mid \theta) = \frac{\exp\!\left(\sum_{j=1}^{k} a(\theta - b_j)\right)}{\sum_{h=0}^{K-1} \exp\!\left(\sum_{j=1}^{h} a(\theta - b_j)\right)},
$$

with the empty sum defined as zero for $k=0$. When $a=1$ this reduces to the PCM.

Latent abilities $\theta_i \sim \mathcal{N}(0,1)$ are drawn once per replication and treated as **known** throughout, isolating item-level misspecification from ability estimation noise.

### Conditions

| Factor | Levels |
|---|---|
| Discrimination $a$ | 1.0 (PCM), 1.5 |
| Categories $K$ | 5, 7 |
| Gap asymmetry | Uniform($-\ln 8$, $\ln 8$) per replication |
| Items per condition | 10 (locations shifted uniformly in $[-1,1]$) |
| Persons per replication | 1000 |
| Replications | 100 |

**Threshold shape** is controlled by a gap asymmetry parameter $s$, drawn independently for each replication from Uniform($-\ln 8$, $\ln 8$). All threshold vectors span the same total range $[-1.5, 1.5]$. When $s=0$ the $K-1$ thresholds are evenly spaced. When $s>0$ gaps widen toward the upper end (higher categories are more spread out); when $s<0$ gaps widen toward the lower end. The max/min gap ratio equals $e^{|s|}$ and is comparable across $K$ values. Results are summarized with a GAM smooth over the continuous $s$ values.

```{r fig-thresholds, fig.height=4, fig.cap="Threshold locations for K=5 and K=7 across the full asymmetry range. Coloured segments connect consecutive thresholds; colour encodes gap width (yellow = wide, purple = narrow). Dots mark threshold positions. All vectors span the same total range [-1.5, 1.5]."}
make_thresholds <- function(K, asym = 0) {
  n           <- K - 1L
  total_range <- 3.0
  sym         <- seq(-total_range / 2, total_range / 2, length.out = n)
  if (asym == 0 || n <= 2L) return(sym)
  k_idx       <- 0:(n - 2L)
  gap_weights <- exp(asym * k_idx / (n - 2L))
  gaps        <- gap_weights / sum(gap_weights) * total_range
  b           <- cumsum(c(-total_range / 2, gaps[-length(gaps)]))
  c(b, b[n - 1L] + gaps[n - 1L])
}

asym_ex <- c(-log(8), -1, 0, 1, log(8))
asym_labels <- c("-ln8\n(wider lower)", "-1", "0\n(symmetric)", "1",
                 "ln8\n(wider upper)")
K_vals  <- c(5, 7)

thresh_df <- do.call(rbind, lapply(K_vals, function(K) {
  do.call(rbind, lapply(seq_along(asym_ex), function(i) {
    b <- make_thresholds(K, asym_ex[i])
    data.frame(K = K, asym = asym_ex[i], asym_lab = asym_labels[i],
               threshold = seq_along(b), value = b)
  }))
}))
thresh_df$K_label   <- paste0("K = ", thresh_df$K)
thresh_df$asym_lab  <- factor(thresh_df$asym_lab, levels = asym_labels)

gaps_df <- thresh_df |>
  arrange(K, asym, threshold) |>
  group_by(K_label, asym_lab) |>
  mutate(val_next = lead(value), gap = val_next - value) |>
  filter(!is.na(gap))

ggplot() +
  annotate("rect", xmin = -1.5, xmax = 1.5, ymin = -Inf, ymax = Inf,
           fill = "grey95") +
  geom_segment(data = gaps_df,
               aes(x = value, xend = val_next, y = asym_lab, yend = asym_lab,
                   colour = gap),
               linewidth = 5) +
  geom_point(data = thresh_df, aes(x = value, y = asym_lab),
             shape = 21, fill = "white", colour = "grey30", size = 2.2) +
  scale_colour_viridis_c(name = "gap", option = "plasma", direction = -1) +
  facet_wrap(~ K_label, ncol = 2) +
  scale_x_continuous(name = "Threshold location", limits = c(-2, 2),
                     breaks = seq(-1.5, 1.5, by = 0.5)) +
  scale_y_discrete(name = "Gap asymmetry (s)") +
  theme_bw(base_size = 11) +
  theme(strip.background = element_rect(fill = "grey92"),
        panel.grid.minor = element_blank(),
        legend.position  = "right")
```

### Models compared

The figure shows expected response functions (ERF: $E[X|\theta]$ vs $\theta$) for each model at K=5, $a=1.5$, $s=\ln 8$ (maximum upper asymmetry). Each model uses its best-fitting single parameters; the GPCM ERF (black) is the target.

:::: {.columns}

::: {.column width="52%"}

| Model | Parameters | Description |
|---|---|---|
| **GPCM** | $a$, $b_1,\ldots,b_{K-1}$ | True generating model |
| **PCM** | $b_1,\ldots,b_{K-1}$ | GPCM with $a=1$ fixed |
| **GRM** | $a$, $b_1,\ldots,b_{K-1}$ | Cumulative logit; $P(X \ge k) = \text{logis}(a(\theta-b_k))$ |
| **Binom 2PL (logit)** | $a$, $b$ | $X \sim \text{Binom}(K{-}1, \text{logis}(a\theta - b))$ |
| **Binom 1PL** | $b$ | Same with $a=1$, four link functions |

:::

::: {.column width="48%"}

```{r fig-erf, fig.height=3.8, fig.width=4.5}
K_ex <- 5L; a_ex <- 1.5; s_ex <- log(8)   # high asymmetry, wider upper gaps
b_ex <- make_thresholds(K_ex, s_ex)
theta_g <- seq(-3, 3, by = 0.05)

gpcm_es <- es_fn(gpcm_pr_fn(theta_g, a_ex, b_ex), K_ex)
pcm_es  <- es_fn(gpcm_pr_fn(theta_g, 1.0,   b_ex), K_ex)
grm_es  <- es_fn(grm_pr_fn( theta_g, a_ex,  b_ex), K_ex)

# Best-fit single parameters for each binomial model
opt1 <- function(link) {
  optim(0, function(b) mean((gpcm_es - es_fn(binom_pr_fn(theta_g,1,b,K_ex,link),K_ex))^2),
        method = "Brent", lower = -3, upper = 3)$par
}
opt2 <- function() {
  optim(c(a_ex, 0),
        function(p) mean((gpcm_es - es_fn(binom_pr_fn(theta_g,p[1],p[2],K_ex,"logit"),K_ex))^2),
        method = "BFGS")$par
}
p2   <- opt2()

erf_df <- rbind(
  data.frame(model="GPCM",               theta=theta_g, es=gpcm_es),
  data.frame(model="PCM",                theta=theta_g, es=pcm_es),
  data.frame(model="GRM",                theta=theta_g, es=grm_es),
  data.frame(model="Binom 2PL\n(logit)", theta=theta_g,
             es=es_fn(binom_pr_fn(theta_g,p2[1],p2[2],K_ex,"logit"),K_ex)),
  data.frame(model="Binom 1PL\n(logit)", theta=theta_g,
             es=es_fn(binom_pr_fn(theta_g,1,opt1("logit"),  K_ex,"logit"),  K_ex)),
  data.frame(model="Binom 1PL\n(probit)",theta=theta_g,
             es=es_fn(binom_pr_fn(theta_g,1,opt1("probit"), K_ex,"probit"), K_ex)),
  data.frame(model="Binom 1PL\n(cloglog)",theta=theta_g,
             es=es_fn(binom_pr_fn(theta_g,1,opt1("cloglog"),K_ex,"cloglog"),K_ex)),
  data.frame(model="Binom 1PL\n(cauchit)",theta=theta_g,
             es=es_fn(binom_pr_fn(theta_g,1,opt1("cauchit"),K_ex,"cauchit"),K_ex))
)

erf_df$is_gpcm <- erf_df$model == "GPCM"

ggplot(erf_df, aes(x = theta, y = es, group = model,
                   colour = model, linewidth = is_gpcm)) +
  geom_line() +
  scale_linewidth_manual(values = c("TRUE" = 1.2, "FALSE" = 0.55),
                         guide = "none") +
  scale_colour_manual(values = c(
    "GPCM"              = "black",
    "PCM"               = "#762a83",
    "GRM"               = "#2166ac",
    "Binom 2PL\n(logit)"  = "#d6604d",
    "Binom 1PL\n(logit)"  = "#f4a582",
    "Binom 1PL\n(probit)" = "#bababa",
    "Binom 1PL\n(cloglog)"= "#c2a5cf",
    "Binom 1PL\n(cauchit)"= "#a6dba0"
  ), name = NULL) +
  scale_x_continuous(breaks = -3:3) +
  labs(x = expression(theta), y = "E[X | θ]") +
  theme_bw(base_size = 10) +
  theme(legend.key.size  = unit(0.5, "lines"),
        legend.text      = element_text(size = 7),
        legend.position  = "right",
        panel.grid.minor = element_blank())
```

:::

::::

Each model is fit to each item with $\theta$ held fixed; item parameters are recovered by MLE, and predicted probabilities are formed at the true $\theta$ values.

---

## Results

### IMV_c

```{r fig-imv-c, fig.cap="IMV_c(model, GPCM): GAM smooth over 100 replications with gap asymmetry s drawn from Uniform(-ln8, ln8). Positive = GPCM predicts better. The dashed line marks zero at s=0 (symmetric)."}
make_plot(res, "imv_c",
          "IMV_c(model, GPCM)  [positive = GPCM wins]")
```

At $s=0$ (symmetric thresholds) the alternatives are close to or below zero — the constrained models match or slightly beat the GPCM, as the extra GPCM parameters gain little when the gap structure is simple. As $|s|$ increases in either direction the binomial approximations diverge upward, with GPCM's advantage growing more sharply for K=7 than K=5.

The two directions of asymmetry are not symmetric in their effect. Wider-upper gaps ($s > 0$) push more IMV disadvantage onto the binomial models than wider-lower gaps ($s < 0$) of the same magnitude. This makes sense: upper-category scarcity is harder for a single $b$ parameter to track than lower-category scarcity, because the binomial mean is sensitive to the center of the distribution, not its tails.

A notable finding is that the asymmetric link functions (cloglog, probit) do *not* benefit under asymmetric thresholds — they perform worse than logit regardless of direction. For wider-upper gaps ($s > 0$): the cloglog link rises slowly at low $\theta$ and saturates quickly at high $\theta$, opposite to the GPCM pattern where the expected score plateaus as high categories remain rare even at high ability. For wider-lower gaps ($s < 0$): the cloglog's fast initial rise is also a poor match. The ERF figure illustrates this directly. The logit link, despite its symmetry, provides a more neutral approximation across the full ability range.

The GRM tracks GPCM closely near $s=0$ (same number of threshold parameters, different parameterization) and diverges at both extremes, more so for K=7.

---

### IMV_t

```{r fig-imv-t, fig.cap="IMV_t(model, GPCM): threshold-based IMV variant. The pattern mirrors IMV_c but at smaller scale; IMV_t averages over K-1 cumulative binary thresholds rather than all pairwise category comparisons."}
make_plot(res, "imv_t",
          "IMV_t(model, GPCM)  [positive = GPCM wins]")
```

IMV_t tells the same directional story. Values near zero at $s=0$ grow steadily in both directions as asymmetry increases, with the wider-upper side showing a steeper rise.

---

### RMSE

```{r fig-rmse, fig.cap="Expected-score RMSE (rescaled 0-1). Lower is better. RMSE differences are compressed relative to IMV but rank models consistently."}
make_plot(res, "rmse",
          "RMSE (expected score, 0-1 scale)")
```

RMSE rankings are consistent with IMV. GPCM and GRM converge near $s=0$; binomial models add meaningful RMSE as $|s|$ grows, with the effect larger on the wider-upper side.

---

## Summary

| Condition | Key finding |
|---|---|
| $s=0$ (symmetric) | Alternatives match or beat GPCM; constrained models act as implicit regularization |
| $|s| \uparrow$, either direction | GPCM advantage grows; binomial single-$b$ can't track the asymmetric gap structure |
| $s > 0$ vs $s < 0$ | Wider-upper gaps hurt binomial models more than wider-lower gaps of equal magnitude |
| Best binomial link | Logit most competitive across all conditions; cloglog and probit degrade fastest |
| GRM | Tracks GPCM closely near $s=0$; diverges at both extremes, more so for K=7 |
| Discrimination $a$ | Scales IMV magnitude; asymmetry effect qualitatively the same at $a=1$ and $a=1.5$ |
