---
title: 'Estimation Infrastructure for Asymmetric IRT Models'
output:
  html_document: default
---

This supplement provides the estimation infrastructure used in the manuscript, *On the asymmetry, complexity, and predicted data pattern of nontraditional item response theory models*.

It includes: (1) analysis settings, (2) Gauss-Hermite quadrature setup, (3) custom item definitions for AO, LPE, and RH with analytic gradients and Hessians, (4) prior specification and starting-value transfer routines, and (5) an LSAT7 illustration that demonstrates the full estimation pipeline from a baseline 2PLM fit through all three asymmetric models.

# Analysis settings

```{r}
library(mirt)
library(statmod)
library(dplyr)
library(tidyr)

em_cycles     <- 5000
quadpts_fit   <- 50
optimizer_fit <- "nlminb"
start_beta_RH <- 0.2
dlogit        <- 1.702

eps_prob_clip_lower <- 1e-32
eps_prob_clip_upper <- .Machine$double.eps

make_shape_prior_profile <- function(){
  list(
    AO  = list(mean = 0, sd = 1),
    LPE = list(mean = 0, sd = 1),
    RH  = list(mean = 0, sd = 1)
  )
}

EMP_PRIORS <- make_shape_prior_profile()
```

# Gauss-Hermite quadrature N(0,1)

```{r}
gh <- statmod::gauss.quad.prob(quadpts_fit, dist = "normal")
ord <- order(gh$nodes)

theta <- gh$nodes[ord]
w_theta <- gh$weights[ord]
Theta_mat <- matrix(theta, ncol = 1)

prior_GH50 <- local({
  theta_ref <- as.numeric(theta)
  w_ref <- as.numeric(w_theta)
  tol <- 1e-12

  stopifnot(isTRUE(all.equal(sum(w_ref), 1, tolerance = tol)))

  function(Theta, Etable){
    Theta <- as.matrix(Theta)
    if(ncol(Theta) != 1L)
      stop("prior_GH50(): only 1D supported")
    th <- as.numeric(Theta[,1])
    if(length(th) != length(theta_ref) || max(abs(th - theta_ref)) > tol){
      stop("prior_GH50(): Theta mismatch. Use customTheta = Theta_mat")
    }
    w_ref
  }
})

assert_fixed_GH50_contract <- local({
  theta_ref <- as.numeric(theta)
  Theta_ref <- as.matrix(Theta_mat)

  function(theta_in, Theta_mat_in, quadpts_fit = NULL){
    theta_in <- as.numeric(theta_in)
    Theta_mat_in <- as.matrix(Theta_mat_in)
    if(ncol(Theta_mat_in) != 1L)
      stop("This script assumes a 1D Theta_mat.")
    if(length(theta_in) != length(theta_ref) ||
       !isTRUE(all.equal(theta_in, theta_ref, tolerance = 0, check.attributes = FALSE))){
      stop("This script is locked to the global GH50 'theta'.")
    }
    if(nrow(Theta_mat_in) != nrow(Theta_ref) ||
       !isTRUE(all.equal(as.numeric(Theta_mat_in[,1]), theta_ref,
                         tolerance = 0, check.attributes = FALSE))){
      stop("This script is locked to the global GH50 'Theta_mat'.")
    }
    if(!is.null(quadpts_fit) &&
       (!is.numeric(quadpts_fit) || length(quadpts_fit) != 1L ||
        quadpts_fit != length(theta_ref))){
      stop(sprintf(
        "quadpts_fit is retained only for interface compatibility. Set quadpts_fit = %d.",
        length(theta_ref)
      ))
    }
    TRUE
  }
})
```

# Helpers

```{r}
bound_prob_mirt <- local({
  eps_lo <- eps_prob_clip_lower
  eps_hi <- eps_prob_clip_upper
  function(P){
    P[P <= eps_lo]     <- eps_lo
    P[P >= 1 - eps_hi] <- 1 - eps_hi
    P
  }
})

clip_mask_mirt <- local({
  eps_lo <- eps_prob_clip_lower
  eps_hi <- eps_prob_clip_upper
  function(P){
    (P <= eps_lo) | (P >= 1 - eps_hi)
  }
})

validate_prior_norm <- function(mean, sd, nm = "shape prior"){
  if(!is.numeric(mean) || length(mean) != 1L || !is.finite(mean)){
    stop(nm, ": mean must be a single finite number")
  }
  if(!is.numeric(sd) || length(sd) != 1L || !is.finite(sd) || sd <= 0){
    stop(nm, ": sd must be a single positive finite number")
  }
  invisible(TRUE)
}

shape_par_name <- function(model_type){
  switch(model_type,
         "AO"  = "eta",
         "LPE" = "S",
         "RH"  = "delta",
         stop("Unknown model_type: ", model_type))
}

set_norm_prior_in_values <- function(sv, par_names, mean = 0, sd = 1){
  validate_prior_norm(mean, sd)

  nmcol <- if("name" %in% names(sv)) "name" else if("parname" %in% names(sv)) "parname" else
    stop("set_norm_prior_in_values(): no name/parname column found")

  if(!("prior.type" %in% names(sv))) sv$prior.type <- NA_character_
  if(!("prior_1" %in% names(sv))) sv$prior_1 <- NA_real_
  if(!("prior_2" %in% names(sv))) sv$prior_2 <- NA_real_

  sv$prior.type <- as.character(sv$prior.type)

  idx <- sv[[nmcol]] %in% par_names
  if(any(idx)){
    sv$prior.type[idx] <- "norm"
    sv$prior_1[idx] <- mean
    sv$prior_2[idx] <- sd
  }

  sv
}

get_col <- function(df, cand){
  hit <- intersect(cand, names(df))
  if(length(hit) == 0) stop("Missing column: ", paste(cand, collapse = ","))
  hit[1]
}

convert_ad_logit_to_probit <- function(ad_df, D = 1.702){
  stopifnot(is.data.frame(ad_df))
  req <- c("item","a1","d")
  if(!all(req %in% names(ad_df))){
    stop("convert_ad_logit_to_probit(): need columns: ", paste(req, collapse = ", "))
  }
  out <- ad_df
  out$a1 <- out$a1 / D
  out$d  <- out$d  / D
  out
}
```

# Custom link IRT models with analytic gradients/Hessians

## AO (lambda = exp(eta))

```{r}
AO_def <- local({

  log1pexp <- function(x){
    pmax(x, 0) + log1p(exp(-abs(x)))
  }

  P_AO <- function(par, Theta, ncat){
    a   <- par[1]
    d   <- par[2]
    eta <- par[3]

    Theta <- as.matrix(Theta); th <- Theta[,1]
    z <- a * th + d
    lambda <- exp(eta)

    if(lambda == 0){
      logP0 <- -exp(z)
      P0 <- exp(logP0)
      P1 <- -expm1(logP0)
      return(bound_prob_mirt(cbind(P0, P1)))
    }
    if(is.infinite(lambda)){
      P0 <- rep(1, length(z))
      P1 <- rep(0, length(z))
      return(bound_prob_mirt(cbind(P0, P1)))
    }

    logA  <- log1pexp(eta + z)
    logP0 <- -logA / lambda
    P0 <- exp(logP0)
    P1 <- -expm1(logP0)
    bound_prob_mirt(cbind(P0, P1))
  }

  dP_AO <- function(par, Theta, ncat){
    a   <- par[1]
    d   <- par[2]
    eta <- par[3]

    Theta <- as.matrix(Theta); th <- Theta[,1]
    z <- a * th + d
    lambda <- exp(eta)

    n <- length(z)
    npar <- length(par)
    dp1 <- array(0, dim = c(n, ncat, npar))

    if(!(lambda == 0 || is.infinite(lambda))){
      logA <- log1pexp(eta + z)
      g    <- exp(z - logA)
      P0   <- exp(-logA / lambda)

      dP_dz   <- P0 * g
      dP_da   <- th * dP_dz
      dP_dd   <- dP_dz
      dP_deta <- P0 * (g - logA / lambda)

      dP_da[!is.finite(dP_da)] <- 0
      dP_dd[!is.finite(dP_dd)] <- 0
      dP_deta[!is.finite(dP_deta)] <- 0

      dp1[,2,1] <- dP_da
      dp1[,2,2] <- dP_dd
      dp1[,2,3] <- dP_deta
      dp1[,1,]  <- -dp1[,2,]
    }

    Pbound <- P_AO(par, Theta, ncat)
    clip <- clip_mask_mirt(Pbound)
    for(cat in 1:ncat){
      if(any(clip[,cat])) dp1[clip[,cat], cat, ] <- 0
    }

    as.vector(dp1)
  }

  d2P_AO <- function(par, Theta, ncat){
    a   <- par[1]
    d   <- par[2]
    eta <- par[3]

    Theta <- as.matrix(Theta); th <- Theta[,1]
    z <- a * th + d
    lambda <- exp(eta)

    n <- length(z)
    npar <- length(par)
    dp2 <- array(0, dim = c(n, ncat, npar, npar))

    if(!(lambda == 0 || is.infinite(lambda))){
      logA <- log1pexp(eta + z)
      g    <- exp(z - logA)
      P0   <- exp(-logA / lambda)
      h    <- g - logA / lambda

      d2_dz2     <- P0 * (g - (1 + lambda) * g^2)
      d2_dz_deta <- -P0 * g * (h + lambda * g)
      dh_deta    <- (logA / lambda) - g - lambda * g^2
      d2_deta2   <- P0 * (dh_deta - h^2)

      d2_dz2[!is.finite(d2_dz2)] <- 0
      d2_dz_deta[!is.finite(d2_dz_deta)] <- 0
      d2_deta2[!is.finite(d2_deta2)] <- 0

      dp2[,2,1,1] <- th^2 * d2_dz2
      dp2[,2,2,2] <- d2_dz2
      dp2[,2,1,2] <- th * d2_dz2
      dp2[,2,2,1] <- dp2[,2,1,2]

      dp2[,2,1,3] <- th * d2_dz_deta
      dp2[,2,3,1] <- dp2[,2,1,3]
      dp2[,2,2,3] <- d2_dz_deta
      dp2[,2,3,2] <- dp2[,2,2,3]

      dp2[,2,3,3] <- d2_deta2
      dp2[,1,,]   <- -dp2[,2,,]
    }

    Pbound <- P_AO(par, Theta, ncat)
    clip <- clip_mask_mirt(Pbound)
    for(cat in 1:ncat){
      if(any(clip[,cat])) dp2[clip[,cat], cat, , ] <- 0
    }

    as.vector(dp2)
  }

  gr_AO <- function(x, Theta){
    dp1 <- array(dP_AO(x@par, Theta, x@ncat),
                 dim = c(nrow(Theta), x@ncat, length(x@par)))
    g <- mirt:::symbolicGrad_par(x, Theta, dp1 = dp1)
    g[x@est]
  }

  hss_AO <- function(x, Theta){
    dp1 <- array(dP_AO(x@par, Theta, x@ncat),
                 dim = c(nrow(Theta), x@ncat, length(x@par)))
    dp2 <- array(d2P_AO(x@par, Theta, x@ncat),
                 dim = c(nrow(Theta), x@ncat, length(x@par), length(x@par)))
    H <- mirt:::symbolicHessian_par(x, Theta, dp1 = dp1, dp2 = dp2)
    H[x@est, x@est, drop = FALSE]
  }

  list(P = P_AO, dP = dP_AO, d2P = d2P_AO, gr = gr_AO, hss = hss_AO)
})
```

## LPE (lambda = exp(S))

```{r}
LPE_def <- local({

  P_LPE <- function(par, Theta, ncat){
    a  <- par[1]
    d  <- par[2]
    S  <- par[3]

    Theta <- as.matrix(Theta); th <- Theta[,1]
    z <- a * th + d

    lambda <- exp(S)
    logPsi <- plogis(z, log.p = TRUE)

    if(lambda == 0){
      P1 <- rep(1, length(z))
      P0 <- rep(0, length(z))
      return(bound_prob_mirt(cbind(P0, P1)))
    }
    if(is.infinite(lambda)){
      P1 <- ifelse(logPsi == 0, 1, 0)
      P0 <- 1 - P1
      return(bound_prob_mirt(cbind(P0, P1)))
    }

    t  <- lambda * logPsi
    P1 <- exp(t)
    P0 <- -expm1(t)
    bound_prob_mirt(cbind(P0, P1))
  }

  dP_LPE <- function(par, Theta, ncat){
    a  <- par[1]
    d  <- par[2]
    S  <- par[3]

    Theta <- as.matrix(Theta); th <- Theta[,1]
    z <- a * th + d
    lambda <- exp(S)

    n <- length(z)
    npar <- length(par)
    dp1 <- array(0, dim = c(n, ncat, npar))

    if(!(lambda == 0 || is.infinite(lambda))){
      Psi    <- plogis(z)
      logPsi <- plogis(z, log.p = TRUE)
      Q      <- 1 - Psi

      Praw <- P_LPE(par, Theta, ncat)
      P1 <- Praw[,2]

      dP_dz <- lambda * P1 * Q
      dP_da <- th * dP_dz
      dP_dd <- dP_dz
      dP_dS <- lambda * P1 * logPsi

      dP_da[!is.finite(dP_da)] <- 0
      dP_dd[!is.finite(dP_dd)] <- 0
      dP_dS[!is.finite(dP_dS)] <- 0

      dp1[,2,1] <- dP_da
      dp1[,2,2] <- dP_dd
      dp1[,2,3] <- dP_dS
      dp1[,1,]  <- -dp1[,2,]
    }

    Pbound <- P_LPE(par, Theta, ncat)
    clip <- clip_mask_mirt(Pbound)
    for(cat in 1:ncat){
      if(any(clip[,cat])) dp1[clip[,cat], cat, ] <- 0
    }

    as.vector(dp1)
  }

  d2P_LPE <- function(par, Theta, ncat){
    a  <- par[1]
    d  <- par[2]
    S  <- par[3]

    Theta <- as.matrix(Theta); th <- Theta[,1]
    z <- a * th + d
    lambda <- exp(S)

    n <- length(z)
    npar <- length(par)
    dp2 <- array(0, dim = c(n, ncat, npar, npar))

    if(!(lambda == 0 || is.infinite(lambda))){
      Psi    <- plogis(z)
      logPsi <- plogis(z, log.p = TRUE)
      Q      <- 1 - Psi

      Praw <- P_LPE(par, Theta, ncat)
      P1 <- Praw[,2]

      d2_dz2     <- lambda * P1 * Q * (lambda * Q - Psi)
      d2_dz_dS   <- lambda * P1 * Q * (lambda * logPsi + 1)
      d2_dS2     <- lambda * P1 * logPsi * (1 + lambda * logPsi)

      d2_dz2[!is.finite(d2_dz2)] <- 0
      d2_dz_dS[!is.finite(d2_dz_dS)] <- 0
      d2_dS2[!is.finite(d2_dS2)] <- 0

      dp2[,2,1,1] <- th^2 * d2_dz2
      dp2[,2,2,2] <- d2_dz2
      dp2[,2,1,2] <- th * d2_dz2
      dp2[,2,2,1] <- dp2[,2,1,2]

      dp2[,2,1,3] <- th * d2_dz_dS
      dp2[,2,3,1] <- dp2[,2,1,3]
      dp2[,2,2,3] <- d2_dz_dS
      dp2[,2,3,2] <- dp2[,2,2,3]

      dp2[,2,3,3] <- d2_dS2
      dp2[,1,,]   <- -dp2[,2,,]
    }

    Pbound <- P_LPE(par, Theta, ncat)
    clip <- clip_mask_mirt(Pbound)
    for(cat in 1:ncat){
      if(any(clip[,cat])) dp2[clip[,cat], cat, , ] <- 0
    }

    as.vector(dp2)
  }

  gr_LPE <- function(x, Theta){
    dp1 <- array(dP_LPE(x@par, Theta, x@ncat),
                 dim = c(nrow(Theta), x@ncat, length(x@par)))
    g <- mirt:::symbolicGrad_par(x, Theta, dp1 = dp1)
    g[x@est]
  }

  hss_LPE <- function(x, Theta){
    dp1 <- array(dP_LPE(x@par, Theta, x@ncat),
                 dim = c(nrow(Theta), x@ncat, length(x@par)))
    dp2 <- array(d2P_LPE(x@par, Theta, x@ncat),
                 dim = c(nrow(Theta), x@ncat, length(x@par), length(x@par)))
    H <- mirt:::symbolicHessian_par(x, Theta, dp1 = dp1, dp2 = dp2)
    H[x@est, x@est, drop = FALSE]
  }

  list(P = P_LPE, dP = dP_LPE, d2P = d2P_LPE, gr = gr_LPE, hss = hss_LPE)
})
```

## RH (delta)

```{r}
RH_def <- local({

  P_RH <- function(par, Theta, ncat){
    a <- par[1]
    d <- par[2]
    delta <- par[3]

    Theta <- as.matrix(Theta); th <- Theta[,1]
    z <- a * th + d

    s <- plogis(delta * th)
    sigma <- sqrt(2) * sqrt(s)
    u <- z / sigma
    u[(is.nan(u)) & (z == 0) & (sigma == 0)] <- 0

    P0 <- numeric(length(u))
    P1 <- numeric(length(u))
    pos <- (u >= 0)

    P0[pos]  <- pnorm(u[pos], lower.tail = FALSE)
    P1[pos]  <- 1 - P0[pos]
    P1[!pos] <- pnorm(u[!pos])
    P0[!pos] <- 1 - P1[!pos]

    bound_prob_mirt(cbind(P0, P1))
  }

  dP_RH <- function(par, Theta, ncat){
    a <- par[1]
    d <- par[2]
    delta <- par[3]

    Theta <- as.matrix(Theta); th <- Theta[,1]
    z <- a * th + d

    s <- plogis(delta * th)
    sigma <- sqrt(2) * sqrt(s)
    u <- z / sigma
    u[(is.nan(u)) & (z == 0) & (sigma == 0)] <- 0

    phi <- dnorm(u)

    n <- length(u)
    npar <- length(par)
    dp1 <- array(0, dim = c(n, ncat, npar))

    u_a <- th / sigma
    u_d <- 1 / sigma
    c   <- th * (1 - s) / 2
    u_delta <- -u * c

    dP_da <- phi * u_a
    dP_dd <- phi * u_d
    dP_ddelta <- phi * u_delta

    dP_da[!is.finite(dP_da)] <- 0
    dP_dd[!is.finite(dP_dd)] <- 0
    dP_ddelta[!is.finite(dP_ddelta)] <- 0

    dp1[,2,1] <- dP_da
    dp1[,2,2] <- dP_dd
    dp1[,2,3] <- dP_ddelta
    dp1[,1,]  <- -dp1[,2,]

    Pbound <- P_RH(par, Theta, ncat)
    clip <- clip_mask_mirt(Pbound)
    for(cat in 1:ncat){
      if(any(clip[,cat])) dp1[clip[,cat], cat, ] <- 0
    }

    as.vector(dp1)
  }

  d2P_RH <- function(par, Theta, ncat){
    a <- par[1]
    d <- par[2]
    delta <- par[3]

    Theta <- as.matrix(Theta); th <- Theta[,1]
    z <- a * th + d

    s <- plogis(delta * th)
    sigma <- sqrt(2) * sqrt(s)
    u <- z / sigma
    u[(is.nan(u)) & (z == 0) & (sigma == 0)] <- 0

    phi <- dnorm(u)

    n <- length(u)
    npar <- length(par)
    dp2 <- array(0, dim = c(n, ncat, npar, npar))

    u_a <- th / sigma
    u_d <- 1 / sigma
    c   <- th * (1 - s) / 2
    u_delta <- -u * c

    P_aa <- -phi * u * (u_a^2)
    P_dd <- -phi * u * (u_d^2)
    P_ad <- -phi * u * (u_a * u_d)

    P_ad2 <- phi * u_a * c * (u^2 - 1)
    P_dd2 <- phi * u_d * c * (u^2 - 1)

    u_dd2 <- u * th^2 * (1 - s^2) / 4
    P_d2d2 <- phi * (u_dd2 - u * (u_delta^2))

    P_aa[!is.finite(P_aa)] <- 0
    P_dd[!is.finite(P_dd)] <- 0
    P_ad[!is.finite(P_ad)] <- 0
    P_ad2[!is.finite(P_ad2)] <- 0
    P_dd2[!is.finite(P_dd2)] <- 0
    P_d2d2[!is.finite(P_d2d2)] <- 0

    dp2[,2,1,1] <- P_aa
    dp2[,2,2,2] <- P_dd
    dp2[,2,1,2] <- P_ad; dp2[,2,2,1] <- P_ad
    dp2[,2,1,3] <- P_ad2; dp2[,2,3,1] <- P_ad2
    dp2[,2,2,3] <- P_dd2; dp2[,2,3,2] <- P_dd2
    dp2[,2,3,3] <- P_d2d2

    dp2[,1,,]   <- -dp2[,2,,]

    Pbound <- P_RH(par, Theta, ncat)
    clip <- clip_mask_mirt(Pbound)
    for(cat in 1:ncat){
      if(any(clip[,cat])) dp2[clip[,cat], cat, , ] <- 0
    }

    as.vector(dp2)
  }

  gr_RH <- function(x, Theta){
    dp1 <- array(dP_RH(x@par, Theta, x@ncat),
                 dim = c(nrow(Theta), x@ncat, length(x@par)))
    g <- mirt:::symbolicGrad_par(x, Theta, dp1 = dp1)
    g[x@est]
  }

  hss_RH <- function(x, Theta){
    dp1 <- array(dP_RH(x@par, Theta, x@ncat),
                 dim = c(nrow(Theta), x@ncat, length(x@par)))
    dp2 <- array(d2P_RH(x@par, Theta, x@ncat),
                 dim = c(nrow(Theta), x@ncat, length(x@par), length(x@par)))
    H <- mirt:::symbolicHessian_par(x, Theta, dp1 = dp1, dp2 = dp2)
    H[x@est, x@est, drop = FALSE]
  }

  list(P = P_RH, dP = dP_RH, d2P = d2P_RH, gr = gr_RH, hss = hss_RH)
})

inject_helpers_into_def <- function(def){
  stopifnot(is.list(def))
  for(nm in c("P","dP","d2P","gr","hss")){
    if(!is.function(def[[nm]])) stop("def$", nm, " is not a function")
    environment(def[[nm]])$bound_prob_mirt <- bound_prob_mirt
    environment(def[[nm]])$clip_mask_mirt  <- clip_mask_mirt
  }
  def
}

AO_def  <- inject_helpers_into_def(AO_def)
LPE_def <- inject_helpers_into_def(LPE_def)
RH_def  <- inject_helpers_into_def(RH_def)
```

# Create custom items

```{r}
make_x_AO <- function(deriv_mode = c("analytic","richardson")){
  deriv_mode <- match.arg(deriv_mode)
  args <- list(
    name = "AO",
    par = c(a1 = 1, d = 0, eta = 0),
    est = c(TRUE, TRUE, TRUE),
    P = AO_def$P
  )
  if(deriv_mode == "analytic"){
    args$gr  <- AO_def$gr
    args$hss <- AO_def$hss
  } else {
    args$derivType <- "Richardson"
    args$derivType.hss <- "Richardson"
  }
  do.call(createItem, args)
}

make_x_LPE <- function(deriv_mode = c("analytic","richardson")){
  deriv_mode <- match.arg(deriv_mode)
  args <- list(
    name = "LPE",
    par = c(a1 = 1, d = 0, S = 0),
    est = c(TRUE, TRUE, TRUE),
    P = LPE_def$P
  )
  if(deriv_mode == "analytic"){
    args$gr  <- LPE_def$gr
    args$hss <- LPE_def$hss
  } else {
    args$derivType <- "Richardson"
    args$derivType.hss <- "Richardson"
  }
  do.call(createItem, args)
}

make_x_RH <- function(deriv_mode = c("analytic","richardson")){
  deriv_mode <- match.arg(deriv_mode)
  args <- list(
    name = "RH",
    par = c(a1 = 1, d = 0, delta = 0),
    est = c(TRUE, TRUE, TRUE),
    P = RH_def$P
  )
  if(deriv_mode == "analytic"){
    args$gr  <- RH_def$gr
    args$hss <- RH_def$hss
  } else {
    args$derivType <- "Richardson"
    args$derivType.hss <- "Richardson"
  }
  do.call(createItem, args)
}

make_custom_bank <- function(deriv_mode = c("analytic","richardson")){
  deriv_mode <- match.arg(deriv_mode)
  list(
    AO  = make_x_AO(deriv_mode = deriv_mode),
    LPE = make_x_LPE(deriv_mode = deriv_mode),
    RH  = make_x_RH(deriv_mode = deriv_mode)
  )
}
```

# Starting values and custom fits

```{r}
extract_ad <- function(mod){
  tab <- mirt::mod2values(mod)
  nmcol  <- if("name" %in% names(tab)) "name" else "parname"
  valcol <- "value"
  if(!("item" %in% names(tab))) stop("mod2values() has no 'item' column")

  item_num <- suppressWarnings(as.integer(gsub("[^0-9]+","", as.character(tab$item))))
  tab$item_num <- item_num
  tab <- tab[!is.na(tab$item_num) & tab$item_num > 0, , drop = FALSE]

  out_long <- tab[, c("item_num", nmcol, valcol), drop = FALSE]
  names(out_long) <- c("item","par","value")
  out_long$par <- as.character(out_long$par)
  out_long$value <- suppressWarnings(as.numeric(out_long$value))

  out_wide <- tidyr::pivot_wider(out_long, names_from = par, values_from = value)

  if(!("a1" %in% names(out_wide)) && "a"  %in% names(out_wide)) out_wide$a1 <- out_wide$a
  if(!("d"  %in% names(out_wide)) && "d1" %in% names(out_wide)) out_wide$d  <- out_wide$d1

  out_wide <- out_wide[, c("item","a1","d"), drop = FALSE]
  out_wide
}

apply_startvals_to_sv <- function(sv, ad_df = NULL, model_type, shape_init = NULL, shape_df = NULL){

  nmcol  <- get_col(sv, c("name","parname"))
  valcol <- get_col(sv, c("value"))
  estcol <- get_col(sv, c("est"))

  if(!("item" %in% names(sv))) stop("pars table has no 'item' column")

  item_num <- suppressWarnings(as.integer(gsub("[^0-9]+","", as.character(sv$item))))

  if(!is.null(ad_df) && nrow(ad_df) > 0){
    ad_df <- as.data.frame(ad_df)
    if(!all(c("item","a1","d") %in% names(ad_df))){
      stop("apply_startvals_to_sv(): ad_df must contain item, a1, d")
    }
    for(i in ad_df$item){
      a_i <- ad_df$a1[ad_df$item == i][1]
      d_i <- ad_df$d [ad_df$item == i][1]
      if(is.finite(a_i)){
        idx <- (item_num == i) & (sv[[nmcol]] %in% c("a1","a"))
        sv[[valcol]][idx] <- a_i
      }
      if(is.finite(d_i)){
        idx <- (item_num == i) & (sv[[nmcol]] %in% c("d","d1"))
        sv[[valcol]][idx] <- d_i
      }
    }
  }

  shape_name <- shape_par_name(model_type)

  if(!is.null(shape_df) && nrow(shape_df) > 0){
    shape_df <- as.data.frame(shape_df)
    if(!all(c("item", shape_name) %in% names(shape_df))){
      stop("apply_startvals_to_sv(): shape_df must contain item and ", shape_name)
    }
    for(i in shape_df$item){
      s_i <- shape_df[[shape_name]][shape_df$item == i][1]
      if(is.finite(s_i)){
        idx <- (item_num == i) & (sv[[nmcol]] == shape_name)
        sv[[valcol]][idx] <- s_i
        sv[[estcol]][idx] <- TRUE
      }
    }
  } else if(!is.null(shape_init) && is.finite(shape_init)){
    idxs <- (item_num > 0) & (sv[[nmcol]] == shape_name)
    sv[[valcol]][idxs] <- shape_init
    sv[[estcol]][idxs] <- TRUE
  }

  sv
}

fit_custom <- function(dat_items, wts, model_type, custom_item_obj,
                       ad_start = NULL, shape_init = NULL, shape_df = NULL,
                       shape_prior_mean = 0, shape_prior_sd = 1,
                       em_cycles = 5000, quadpts_fit = 50, optimizer_fit = "nlminb"){

  warns <- character(0)

  assert_fixed_GH50_contract(theta, Theta_mat, quadpts_fit)

  fit_technical <- list(
    NCYCLES = em_cycles,
    customTheta = Theta_mat,
    customPriorFun = prior_GH50
  )

  out <- tryCatch({

    sv_raw <- withCallingHandlers(
      mirt(dat_items, model = 1, itemtype = model_type,
           customItems = setNames(list(custom_item_obj), model_type),
           pars = "values", method = "EM",
           survey.weights = wts,
           technical = fit_technical,
           optimizer = optimizer_fit,
           verbose = FALSE),
      warning = function(w){
        warns <<- c(warns, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    )

    sv2 <- apply_startvals_to_sv(
      sv_raw,
      ad_df = ad_start,
      model_type = model_type,
      shape_init = shape_init,
      shape_df = shape_df
    )

    sv2 <- set_norm_prior_in_values(
      sv2,
      par_names = shape_par_name(model_type),
      mean = shape_prior_mean,
      sd = shape_prior_sd
    )

    mod <- withCallingHandlers(
      mirt(dat_items, model = 1, itemtype = model_type,
           customItems = setNames(list(custom_item_obj), model_type),
           pars = sv2, method = "EM",
           survey.weights = wts,
           technical = fit_technical,
           optimizer = optimizer_fit,
           verbose = FALSE),
      warning = function(w){
        warns <<- c(warns, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    )

    list(mod = mod, error = FALSE, message = NA_character_, warnings = unique(warns))

  }, error = function(e){
    list(mod = NULL, error = TRUE, message = conditionMessage(e), warnings = unique(warns))
  })

  out
}
```

# Safe extractors

```{r}
has_valid_mod <- function(fit_obj, require_converged = FALSE){
  ok <- !is.null(fit_obj) && !isTRUE(fit_obj$error) && !is.null(fit_obj$mod)
  if(require_converged && ok){
    ok <- tryCatch(isTRUE(fit_obj$mod@OptimInfo$converged), error = function(e) FALSE)
  }
  ok
}

extract_param_table <- function(mod){
  tab <- mod2values(mod)
  item_num <- suppressWarnings(as.integer(gsub("[^0-9]+", "", as.character(tab$item))))
  tab$item_num <- item_num
  tab <- tab[!is.na(tab$item_num) & tab$item_num > 0, , drop = FALSE]

  nmcol <- if("name" %in% names(tab)) "name" else "parname"

  out_long <- tab[, c("item_num", nmcol, "value"), drop = FALSE]
  names(out_long) <- c("item", "par", "value")
  out_long$par <- as.character(out_long$par)
  out_long$value <- suppressWarnings(as.numeric(out_long$value))

  out_wide <- tidyr::pivot_wider(out_long, names_from = par, values_from = value)

  if(!("a1" %in% names(out_wide)) && "a"  %in% names(out_wide)) out_wide$a1 <- out_wide$a
  if(!("d"  %in% names(out_wide)) && "d1" %in% names(out_wide)) out_wide$d  <- out_wide$d1

  out_wide[order(out_wide$item), , drop = FALSE]
}
```

# Fixed defaults

```{r}
.em_cycles_global <- em_cycles
.quadpts_fit_global <- quadpts_fit
.optimizer_fit_global <- optimizer_fit
.start_beta_RH_global <- start_beta_RH
.dlogit_global <- dlogit
.EMP_PRIORS_global <- EMP_PRIORS
```

# LSAT7 illustration

```{r}
fit_lsat7 <- function(
  em_cycles = .em_cycles_global,
  quadpts_fit = .quadpts_fit_global,
  optimizer_fit = .optimizer_fit_global,
  start_beta_RH = .start_beta_RH_global,
  dlogit = .dlogit_global,
  priors = .EMP_PRIORS_global
){

  assert_fixed_GH50_contract(theta, Theta_mat, quadpts_fit)

  data(LSAT7)
  lsat7_full <- as.data.frame(expand.table(LSAT7))
  ni_lsat7 <- ncol(lsat7_full)

  lsat7_mirt_df <- lsat7_full
  lsat7_mirt_df[] <- lapply(lsat7_mirt_df, function(x) as.integer(as.character(x)))
  names(lsat7_mirt_df) <- paste0("I", seq_len(ni_lsat7))
  lsat7_mirt_df$r <- 1
  rownames(lsat7_mirt_df) <- NULL

  dat_items <- lsat7_mirt_df[, 1:ni_lsat7, drop = FALSE]
  wts <- lsat7_mirt_df$r

  fit_technical <- list(
    NCYCLES = em_cycles,
    customTheta = Theta_mat,
    customPriorFun = prior_GH50
  )

  custom_bank <- make_custom_bank(deriv_mode = "analytic")

  # 1) Baseline 2PLM
  fit_2PL <- mirt(
    dat_items, model = 1, itemtype = "2PL",
    method = "EM",
    survey.weights = wts,
    technical = fit_technical,
    optimizer = optimizer_fit,
    verbose = FALSE
  )

  ad_2PL <- extract_ad(fit_2PL)

  ad_RH_start <- if(!is.null(ad_2PL) && nrow(ad_2PL) > 0){
    convert_ad_logit_to_probit(ad_2PL, D = dlogit)
  } else {
    NULL
  }

  # 2) Custom AO
  fit_AO <- fit_custom(
    dat_items, wts,
    model_type = "AO",
    custom_item_obj = custom_bank$AO,
    ad_start = ad_2PL,
    shape_init = 0,
    shape_prior_mean = priors$AO$mean,
    shape_prior_sd = priors$AO$sd,
    em_cycles = em_cycles,
    quadpts_fit = quadpts_fit,
    optimizer_fit = optimizer_fit
  )

  # 3) Custom LPE
  fit_LPE <- fit_custom(
    dat_items, wts,
    model_type = "LPE",
    custom_item_obj = custom_bank$LPE,
    ad_start = ad_2PL,
    shape_init = 0,
    shape_prior_mean = priors$LPE$mean,
    shape_prior_sd = priors$LPE$sd,
    em_cycles = em_cycles,
    quadpts_fit = quadpts_fit,
    optimizer_fit = optimizer_fit
  )

  # 4) Custom RH
  fit_RH <- fit_custom(
    dat_items, wts,
    model_type = "RH",
    custom_item_obj = custom_bank$RH,
    ad_start = ad_RH_start,
    shape_init = start_beta_RH,
    shape_prior_mean = priors$RH$mean,
    shape_prior_sd = priors$RH$sd,
    em_cycles = em_cycles,
    quadpts_fit = quadpts_fit,
    optimizer_fit = optimizer_fit
  )

  # 5) Built-in LPE via restricted 5PLM (g = 0, u = 1; for comparison with custom LPE)
  fit_LPE_builtin <- tryCatch({
    warns_b <- character(0)

    sv_b <- withCallingHandlers(
      mirt(dat_items, model = 1, itemtype = "5PL",
           pars = "values", method = "EM",
           survey.weights = wts,
           technical = fit_technical,
           optimizer = optimizer_fit,
           verbose = FALSE),
      warning = function(w){
        warns_b <<- c(warns_b, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    )

    nmcol_b <- if("name" %in% names(sv_b)) "name" else "parname"

    # Fix g = 0, u = 1
    for(pn in c("g","c","c1")){
      idx <- sv_b[[nmcol_b]] == pn
      if(any(idx)){
        sv_b$value[idx] <- 0; sv_b$est[idx] <- FALSE
        if("lbound" %in% names(sv_b)) sv_b$lbound[idx] <- 0
        if("ubound" %in% names(sv_b)) sv_b$ubound[idx] <- 0
      }
    }
    for(pn in c("u","u1")){
      idx <- sv_b[[nmcol_b]] == pn
      if(any(idx)){
        sv_b$value[idx] <- 1; sv_b$est[idx] <- FALSE
        if("lbound" %in% names(sv_b)) sv_b$lbound[idx] <- 1
        if("ubound" %in% names(sv_b)) sv_b$ubound[idx] <- 1
      }
    }

    # Set shape parameter start to 0 and apply prior
    is_shape <- sv_b[[nmcol_b]] %in% c("S","logS")
    if(any(is_shape)) sv_b$value[is_shape] <- 0
    sv_b <- set_norm_prior_in_values(sv_b, par_names = c("S","logS"),
                                      mean = priors$LPE$mean, sd = priors$LPE$sd)

    mod_b <- withCallingHandlers(
      mirt(dat_items, model = 1, itemtype = "5PL",
           pars = sv_b, method = "EM",
           survey.weights = wts,
           technical = fit_technical,
           optimizer = optimizer_fit,
           verbose = FALSE),
      warning = function(w){
        warns_b <<- c(warns_b, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    )

    list(mod = mod_b, error = FALSE, message = NA_character_, warnings = unique(warns_b))
  }, error = function(e){
    list(mod = NULL, error = TRUE, message = conditionMessage(e), warnings = character(0))
  })

  list(
    meta = list(
      em_cycles = em_cycles,
      quadpts_fit = quadpts_fit,
      optimizer_fit = optimizer_fit,
      start_beta_RH = start_beta_RH,
      dlogit = dlogit,
      shape_priors = priors
    ),
    fit = list(
      `2PLM` = list(mod = fit_2PL, error = FALSE, message = NA_character_, warnings = character(0)),
      AO = fit_AO,
      LPE = fit_LPE,
      RH = fit_RH,
      LPE_builtin = fit_LPE_builtin
    )
  )
}
```

# Run LSAT7 illustration

```{r}
res_lsat7 <- fit_lsat7()

cat("\n--- 2PLM ---\n")
print(extract_param_table(res_lsat7$fit$`2PLM`$mod))

cat("\n--- Custom AO ---\n")
if(has_valid_mod(res_lsat7$fit$AO)){
  print(extract_param_table(res_lsat7$fit$AO$mod))
} else {
  cat("AO fit failed:", res_lsat7$fit$AO$message, "\n")
}

cat("\n--- Custom LPE ---\n")
if(has_valid_mod(res_lsat7$fit$LPE)){
  print(extract_param_table(res_lsat7$fit$LPE$mod))
} else {
  cat("LPE fit failed:", res_lsat7$fit$LPE$message, "\n")
}

cat("\n--- Custom RH ---\n")
if(has_valid_mod(res_lsat7$fit$RH)){
  print(extract_param_table(res_lsat7$fit$RH$mod))
} else {
  cat("RH fit failed:", res_lsat7$fit$RH$message, "\n")
}

cat("\n--- Built-in LPE (5PLM with g = 0, u = 1; for comparison) ---\n")
if(has_valid_mod(res_lsat7$fit$LPE_builtin)){
  print(extract_param_table(res_lsat7$fit$LPE_builtin$mod))
} else {
  cat("Built-in LPE fit failed:", res_lsat7$fit$LPE_builtin$message, "\n")
}
```
