---
title: "Replication at Scale: Does Algorithm Selection for Class Detection Transport to Item Response Data?"
description: |
  Sotoudeh and DiMaggio (2021) predicted which of ten distance measures would
  best recover latent classes in nine survey modules, using regressions trained
  on 198,000 simulated matrices, and validated those predictions against five
  internal cluster-quality criteria. We run the same procedure on a stratified
  sample of IRW tables and report which of their findings survive.
date: today
bibliography: plenitude_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(knitr)
})

res    <- readRDS("plenitude_data/plenitude_results.rds")
corpus <- read.csv("plenitude_data/corpus.csv", stringsAsFactors = FALSE)
table9 <- read.csv("vendor/table9.csv", stringsAsFactors = FALSE)
date_run <- file.info("plenitude_data/plenitude_results.rds")$mtime

mf <- res$metafeatures; pr <- res$predicted; so <- res$solutions
va <- res$validity;     rk <- res$reverse_key

n_tables <- dplyr::n_distinct(mf$table)
n_fits   <- res$n_fits

METHODS <- c("eJaccard","Podani","Euclidean","Cosine","eDice","Correlation",
             "ACE","Original Relationality","Recursive Relationality","Ensemble")
CRITERIA <- c(pearsongamma = "Pearson's Gamma", within_ss = "Within-class SS",
              wb_ratio = "Within-between ratio", ch = "Calinski-Harabasz",
              aic_improvement = "SEM AIC improvement")
## within_ss and wb_ratio are cost-like: smaller is better. The rest are gains.
LOWER_BETTER <- c("within_ss", "wb_ratio")

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
```

## Motivation

Sotoudeh and DiMaggio [-@sotoudeh2021plenitude] set out to make the choice of a
class-detection algorithm an empirical question rather than a matter of taste.
They simulated 198,000 response matrices under a known class structure, measured
how accurately each of ten distance measures recovered that structure, and
regressed accuracy on fourteen *meta-features* of the matrix — its skewness, its
principal-component structure, the distribution of respondent-by-respondent
correlations, and so on. The fitted regressions let an analyst compute those
fourteen numbers on a real dataset and read off which measure is predicted to do
best. They then applied the procedure to nine General Social Survey and American
National Election Studies modules and checked the predictions against five
internal cluster-quality criteria.

Their headline results were that predicted accuracy correlates roughly .54 to
.68 with the internal criteria (their Table 12); that no single measure wins
everywhere; and that new (recursive) relationality and correlation are
generalists while Euclidean distance, cosine and the original Goldberg
relationality underperform.

Nine datasets is a thin evidentiary base for a claim about which algorithm to
use on *your* data. This page runs their published procedure — their code, their
fitted models, their criteria — on `r n_tables` item-response tables drawn from
the Item Response Warehouse. This is a replication. We are not proposing an
alternative framework or new meta-features; where their procedure is
underspecified we follow what their code does and record the choice.

## Scope, and what could not be replicated

Their artifact chain runs through a personal Google Drive folder with no DOI
(the linked GitHub repository contains only a README). The files are vendored in
`analysis/vendor/` and pinned by SHA-256.

Two limits on scope follow from what that folder contains:

- **Nine of the ten methods have fitted models.** `metafeature_models_cleaned.RDS`
  holds nine `lm` objects, one per distance measure. The pairwise-voting
  ensemble has none — it is a combination rule rather than a measure, so there is
  nothing to predict its accuracy from. We therefore *run* all ten methods and
  compute all five validity criteria for all ten, but predicted accuracy exists
  for nine.
- **Table 11 is out of scope.** `models_w_noise.RDS` — the regressions fitted
  under varying assumed error variance, which the paper's discussion of when the
  ensemble wins depends on — downloads as a valid gzip stream but fails
  `readRDS`. The noise-assumption models are unrecoverable from the released
  artifacts, so we do not attempt the Table 11 comparison rather than
  approximating it with something else.

### Five defects in the released code

The scouting pass that preceded this run audited the released code. Four defects
were carried into this replication as explicit decisions; a fifth was found while
wiring up the prediction step.

```{r}
#| label: tbl-defects
data.frame(
  Defect = c("`num_vars` defaults to 10",
             "`PercentOut` scale",
             "`OverallRightKurt` on a logical vector",
             "Louvain hardcoded",
             "`make_prediction()` matches positionally"),
  What = c(
    "`evaluate.metafeatures()` takes `num_vars = 10` and `select.method()` never passes it. It is the denominator of `PercentOut` and `intrinsicDimProp`.",
    "The released formula is a count of outlier-bearing *variables* over item count, times 100 — a 0–100 lattice. Table 9's column runs 0.000–0.048.",
    "`Kurt(df > mean(df))` is the kurtosis of a Bernoulli indicator. The evidently-intended `right_half_kurt` is defined at their line 217 and never called.",
    "`metaRCA()` fixes community detection at `cluster_louvain()` with no override, though the paper concedes the choice was consequential.",
    "Coefficients are multiplied by meta-features by position. The column order of `evaluate.metafeatures()` does not match the coefficient order of the fitted models — 7 of 14 positions are swapped."),
  Handling = c(
    "`num_vars = ncol(df)` passed explicitly everywhere; `select.method()` never used.",
    "Both readings computed and carried through the whole downstream analysis.",
    "Released behaviour is primary — their coefficients were fitted on that quantity. Corrected version reported as a sensitivity check.",
    "Kept, matching the paper. Recorded as fixed rather than treated as a free parameter.",
    "Matched by name for the primary analysis; released positional behaviour reported as a sensitivity check.")
) |> kable(caption = "Defects in the released code and how each is handled here.")
```

The smoke test for the first of these: on an 18-item table, the released default
returns `PercentOut = 130` — a percentage above 100, impossible by construction —
and `intrinsicDimProp = 0.400` against a correct 0.222. Passing `num_vars`
explicitly puts both back in range, which is the check that we are on the right
code path.

## Data and methods

### Corpus

```{r}
#| label: tbl-strata
corpus |>
  count(affect_cog, catband = cut(n_categories, c(3,5,7,1e6),
                                  labels = c("4-5","6-7","8+"))) |>
  pivot_wider(names_from = catband, values_from = n, values_fill = 0) |>
  rename(`Construct tag` = affect_cog) |>
  kable(caption = "Realized stratification of the drawn corpus, on response-category count and the affective/cognitive tag.")
```

The sampling frame is the `r nrow(read.csv("scout_eligible_tables.csv"))`-table
eligible set established by the scouting pass: complete (`density == 1`), 6–40
items, at least 300 respondents, at least 4 response categories. Tables with more
than 50,000 respondents are excluded on download cost alone
(`r nrow(read.csv("scout_eligible_tables.csv")) - nrow(read.csv("scout_eligible_tables.csv")[read.csv("scout_eligible_tables.csv")$n_participants <= 50000, ])`
of them), following the same decision the scouting run made. From the remainder
we drew `r nrow(corpus)` tables stratified on response-category band and on the
construct tag. Note that `r sum(corpus$affect_cog == "untagged")` of the drawn
tables carry no construct tag at all, so any split by construct type runs on
fewer tables than the total.

**Single tables only.** The merge path — stacking sibling tables that share a
respondent sample — is not used. The scouting run found it produced the worst
reverse-keying contamination observed and depends on an id-comparability
assumption verified on only 24 of 114 candidate groups.

### Subsampling

Respondents are subsampled to a fixed cap of N = 2,000, with three replicates per
table at recorded seeds (101, 202, 303). Several meta-features are N-sensitive and
their nine datasets ran roughly 1,500–2,500 respondents, so a fixed cap keeps the
comparison to their Table 9 interpretable and keeps our tables comparable to each
other.

```{r}
#| label: n-shortfall
n_short <- sum(!corpus$n_participants >= 2000)
```

The cap binds rarely: `r n_short` of `r nrow(corpus)` tables
(`r fmt(100*n_short/nrow(corpus), 0)`%) have fewer than 2,000 respondents and use
all of them. For those tables the three replicates are the same respondents, and
differ only through the stochastic community-detection step — they measure
algorithmic stability, not sampling stability. The list is in
`plenitude_data/metafeatures.csv` (`N_capped`).

### Procedure

For each table × replicate we compute the fourteen meta-features, apply the nine
fitted `lm` objects to obtain predicted accuracy per method, run all nine distance
measures plus the pairwise-voting ensemble to recover a class solution, and
compute the five internal validity criteria. Following their construction, the
criteria are computed on a distance matrix built from **absolute response
correlations**.

::: {.callout-warning}
## Their caveat, restated
Building the validity distance matrix from absolute response correlations biases
these criteria toward the measures that are themselves correlation-based —
correlation, cosine and ACE. The paper says so, and it applies with equal force
here. Agreement between predicted accuracy and the criteria is not independent
evidence for a correlation-based measure winning.
:::

The fifth criterion, the SEM comparison, fits a saturated covariance model
grouped by recovered class and takes the AIC improvement over the pooled model.
That model is not identified in any class with fewer members than items, and
these methods routinely return such classes. We report their construction
unchanged — `try()`, `NA` on failure — and alongside it a *trimmed* variant that
pools all classes smaller than p + 1 into one. The trimmed version is reported in
addition to, never in place of, theirs.

## Results

```{r}
#| label: run-scale
tot_hours <- sum(so$secs, na.rm = TRUE) / 3600
```

`r n_fits` table × replicate fits completed, covering `r n_tables` tables and
`r fmt(tot_hours, 1)` core-hours of clustering.

### Predicted accuracy, and a scale problem that decides itself

```{r}
#| label: tbl-readings
pr |>
  group_by(reading) |>
  summarise(min = min(predicted), median = median(predicted), max = max(predicted),
            `out of [0,1]` = sprintf("%.0f%%", 100*mean(predicted < 0 | predicted > 1)),
            .groups = "drop") |>
  mutate(across(c(min, median, max), ~ fmt(.x, 3))) |>
  kable(caption = "Predicted accuracy under each reading of the ambiguous inputs. Accuracy is a proportion; values outside [0,1] are impossible.")
```

The brief for this replication treated the `PercentOut` scale as unresolved and
asked for both readings to be carried through. The data resolve it. Feeding the
released formula's 0–100 values into the fitted models puts predicted accuracy
outside [0,1] on essentially every table, because the `PercentOut` coefficient
(−0.022) was fitted against inputs two orders of magnitude smaller. The
cell-outlier proportion — the reading whose magnitudes match Table 9 — produces
predictions in range.

The conclusion is narrow but firm: **the released `evaluate.metafeatures()` does
not produce the input that the released models expect.** A user following the
documented workflow gets predicted accuracies that are not accuracies. We use the
cell-proportion reading (`pct_cells`) as the working primary below, and report
the winner-map comparison across readings that the brief asked for.

```{r}
#| label: winner-stability
winner_by <- function(reading) {
  pr |> filter(reading == !!reading) |>
    group_by(table, replicate) |>
    slice_max(predicted, n = 1, with_ties = FALSE) |>
    ungroup() |> select(table, replicate, winner = method)
}
w <- Reduce(function(a, b) full_join(a, b, by = c("table","replicate")),
            lapply(c("primary","pct_cells","rk_corrected","positional"), function(r)
              winner_by(r) |> rename(!!r := winner)))
agree <- sapply(c("primary","rk_corrected","positional"),
                function(r) mean(w[[r]] == w$pct_cells, na.rm = TRUE))
```

Winner-per-table agreement with the `pct_cells` reading:
released `PercentOut` `r fmt(100*agree[["primary"]], 0)`%,
corrected `OverallRightKurt` `r fmt(100*agree[["rk_corrected"]], 0)`%,
released positional coefficient matching `r fmt(100*agree[["positional"]], 0)`%.

```{r}
#| label: tbl-oor
mf_long <- mf |> select(table, replicate, all_of(c("OverallSD","OverallKurt","OverallRightKurt",
  "VarColPC1","SkewColPC1","SkewColPC2","Skewness","Mean_Row_Corr","Row_Corr_Kurtosis",
  "CorrRightKurt","Mean_Col_Kurtosis","Mean_SD_Ratio","intrinsicDemnsionalityProp"))) |>
  pivot_longer(-c(table, replicate), names_to = "feature", values_to = "value")
t9 <- table9 |> select(-dataset) |>
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  group_by(feature) |> summarise(lo = min(value), hi = max(value), .groups = "drop")
oor <- mf_long |> inner_join(t9, by = "feature") |>
  mutate(outside = value < lo | value > hi) |>
  group_by(feature) |>
  summarise(`% outside their Table 9 range` = sprintf("%.0f%%", 100*mean(outside)), .groups = "drop") |>
  arrange(desc(`% outside their Table 9 range`))
kable(oor, caption = "How often IRW meta-feature values fall outside the range their nine datasets spanned. Their simulation training data is not distributed, so Table 9 is the closest available reference envelope.")
```

Their simulation training data is not distributed and the fitted `lm` objects
were stripped of their model frames, so the training envelope itself cannot be
recovered. Table 9 — the nine empirical datasets they applied the models to — is
the closest available reference, and it is a weaker one: falling inside it means
only that we look like their nine datasets, not that we look like the
simulations.

### Q1 — Does prediction track validation?

Their Table 12 reported correlations of roughly .54 to .68 between predicted
accuracy and the five internal validity criteria across their nine datasets.

Two of the criteria — within-class sum of squares and Calinski–Harabasz — scale
with the number of respondents and items, and our tables span
`r min(corpus$n_items)`–`r max(corpus$n_items)` items and
`r min(corpus$eff_N)`–`r max(corpus$eff_N)` respondents where their nine were
far more alike. Pooling raw criterion
values across such tables lets between-table scale differences dominate the
within-table ranking that the question is actually about. So we report the
correlation **within each table** and summarise across tables, which is the
like-for-like quantity, and give the pooled correlation alongside it for direct
comparison with their published number.

```{r}
#| label: q1
d <- pr |> filter(reading == "pct_cells") |>
  inner_join(va, by = c("table","replicate","method"))

## sign convention: flip the cost-like criteria so a positive correlation always
## means "predicted accuracy agrees with the criterion"
orient <- function(cr) if (cr %in% LOWER_BETTER) -1 else 1

q1 <- lapply(names(CRITERIA), function(cr) {
  dd <- d |> mutate(val = orient(cr) * .data[[cr]]) |> filter(is.finite(val), is.finite(predicted))
  if (nrow(dd) < 10) return(NULL)

  ## pooled, as their Table 12 reads
  ct <- cor.test(dd$predicted, dd$val)

  ## within-table, then summarised across tables
  wt <- dd |> group_by(table) |>
    filter(n() >= 5, sd(val) > 0, sd(predicted) > 0) |>
    summarise(r = cor(predicted, val), .groups = "drop")
  ci <- if (nrow(wt) > 2) t.test(wt$r)$conf.int else c(NA, NA)

  data.frame(Criterion = CRITERIA[[cr]],
             `Tables` = nrow(wt),
             `Mean within-table r` = mean(wt$r),
             `lo` = ci[1], `hi` = ci[2],
             `Pooled r` = unname(ct$estimate),
             `plo` = ct$conf.int[1], `phi` = ct$conf.int[2],
             check.names = FALSE)
}) |> bind_rows()

q1 |>
  transmute(Criterion, Tables,
            `Mean within-table r [95% CI]` = sprintf("%s [%s, %s]",
              fmt(`Mean within-table r`), fmt(lo), fmt(hi)),
            `Pooled r [95% CI]` = sprintf("%s [%s, %s]",
              fmt(`Pooled r`), fmt(plo), fmt(phi))) |>
  kable(caption = "Q1. Agreement between predicted accuracy and each internal validity criterion. Signs are oriented so positive always means agreement. Their nine-dataset values were .54-.68.")
```

```{r}
#| label: fig-q1-dist
#| fig-cap: "Distribution of the within-table correlation between predicted accuracy and each criterion. One value per table."
lapply(names(CRITERIA), function(cr) {
  d |> mutate(val = orient(cr) * .data[[cr]]) |>
    filter(is.finite(val), is.finite(predicted)) |>
    group_by(table) |> filter(n() >= 5, sd(val) > 0, sd(predicted) > 0) |>
    summarise(r = cor(predicted, val), .groups = "drop") |>
    mutate(criterion = CRITERIA[[cr]])
}) |> bind_rows() |>
  ggplot(aes(r)) +
  geom_histogram(bins = 25, fill = "#4292c6") +
  geom_vline(xintercept = 0, colour = "grey30") +
  facet_wrap(~ criterion) +
  labs(x = "Within-table correlation of predicted accuracy with the criterion",
       y = "Tables") +
  theme_minimal(base_size = 11)
```

```{r}
#| label: fig-q1
#| fig-cap: "Predicted accuracy against each validity criterion, pooled. One point per table x replicate x method. The two scale-dependent criteria show the between-table spread that motivates the within-table summary above."
d |> select(predicted, all_of(names(CRITERIA))) |>
  pivot_longer(-predicted, names_to = "criterion", values_to = "value") |>
  filter(is.finite(value)) |>
  mutate(criterion = CRITERIA[criterion]) |>
  ggplot(aes(value, predicted)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE, colour = "#2c7fb8") +
  facet_wrap(~ criterion, scales = "free_x") +
  labs(x = "Validity criterion", y = "Predicted accuracy") +
  theme_minimal(base_size = 11)
```

### Q2 — Does the winner vary?

```{r}
#| label: q2
top_pred <- pr |> filter(reading == "pct_cells") |>
  group_by(table, replicate) |> slice_max(predicted, n = 1, with_ties = FALSE) |>
  ungroup() |> transmute(table, replicate, basis = "Predicted accuracy", winner = method)

top_val <- lapply(names(CRITERIA), function(cr) {
  v <- va |> select(table, replicate, method, val = all_of(cr)) |> filter(is.finite(val))
  if (nrow(v) == 0) return(NULL)
  if (cr %in% LOWER_BETTER) v <- v |> group_by(table, replicate) |> slice_min(val, n = 1, with_ties = FALSE)
  else                      v <- v |> group_by(table, replicate) |> slice_max(val, n = 1, with_ties = FALSE)
  v |> ungroup() |> transmute(table, replicate, basis = CRITERIA[[cr]], winner = method)
}) |> bind_rows()

wins <- bind_rows(top_pred, top_val)
wins |> count(basis, winner) |>
  group_by(basis) |> mutate(pct = sprintf("%.0f%%", 100*n/sum(n))) |>
  select(-n) |> pivot_wider(names_from = basis, values_from = pct, values_fill = "-") |>
  arrange(match(winner, METHODS)) |> rename(Method = winner) |>
  kable(caption = "Q2. Share of table x replicate runs each method wins, by basis. Predicted accuracy covers nine methods; the criteria cover all ten.")
```

```{r}
#| label: q2-concentration
conc <- wins |> count(basis, winner) |> group_by(basis) |>
  summarise(top_share = max(n)/sum(n), top_method = winner[which.max(n)],
            n_distinct_winners = n(), .groups = "drop")
```

```{r}
#| label: tbl-q2-conc
conc |> mutate(top_share = sprintf("%.0f%%", 100*top_share)) |>
  rename(Basis = basis, `Top method` = top_method,
         `Its share of wins` = top_share, `Distinct winners` = n_distinct_winners) |>
  kable(caption = "Q2. Concentration of wins. A high share for one method on one basis is evidence against 'no free lunch' on that basis.")
```

### Q3 — Do their method-level characterizations hold?

```{r}
#| label: q3
rank_within <- function(df, col, lower_better) {
  df |> group_by(table, replicate) |>
    mutate(rk = if (lower_better) rank(.data[[col]], na.last = "keep")
                else rank(-.data[[col]], na.last = "keep")) |> ungroup()
}
ranks <- lapply(names(CRITERIA), function(cr) {
  va |> select(table, replicate, method, val = all_of(cr)) |> filter(is.finite(val)) |>
    rank_within("val", cr %in% LOWER_BETTER) |>
    transmute(method, criterion = CRITERIA[[cr]], rk)
}) |> bind_rows()

q3 <- ranks |> group_by(method) |>
  summarise(`Mean rank across criteria` = mean(rk, na.rm = TRUE),
            `Top-3 rate` = mean(rk <= 3, na.rm = TRUE), .groups = "drop") |>
  left_join(pr |> filter(reading == "pct_cells") |> group_by(method) |>
              summarise(`Mean predicted accuracy` = mean(predicted), .groups = "drop"),
            by = "method") |>
  arrange(`Mean rank across criteria`)

q3 |> mutate(across(where(is.numeric), ~ fmt(.x, 3))) |>
  rename(Method = method) |>
  kable(caption = "Q3. Mean rank across the five validity criteria (lower is better; ten methods so 5.5 is chance), the rate of finishing top-3, and mean predicted accuracy. The ensemble has no fitted model.")
```

### Class solutions and degeneracy

```{r}
#| label: tbl-degen
so |> filter(status == "ok") |>
  group_by(Method = method) |>
  summarise(`median k` = median(k), `median substantive k` = median(n_substantive),
            `median largest class` = sprintf("%.0f%%", 100*median(largest_prop)),
            `degenerate` = sprintf("%.0f%%", 100*mean(degenerate)),
            `median secs` = fmt(median(secs), 2), .groups = "drop") |>
  arrange(match(Method, METHODS)) |>
  kable(caption = "Recovered class solutions. Degenerate means k = 1 or a largest class above 90%. Degenerate runs are flagged, not dropped.")
```

```{r}
#| label: degen-rate
degen_rate <- mean(so$degenerate[so$status == "ok"], na.rm = TRUE)
err_rate   <- mean(so$status != "ok")
```

Overall, `r fmt(100*degen_rate, 1)`% of method runs are degenerate and
`r fmt(100*err_rate, 1)`% errored. Degenerate solutions are retained in every
table above.

## The reverse-keying diagnostic

Recovered "classes" in item response data can be nothing more than which way a
respondent leans on the first principal component — agreement with the sign of
the PC1 loadings. We report that directly: for each table × method, the share of
variance in a respondent's PC1-polarity score that is explained by their class
assignment.

```{r}
#| label: tbl-rk
rk |> filter(is.finite(eta2)) |> group_by(Method = method) |>
  summarise(Q1 = quantile(eta2, .25), Median = median(eta2), Q3 = quantile(eta2, .75),
            Max = max(eta2), `share > 0.5` = sprintf("%.0f%%", 100*mean(eta2 > 0.5)),
            .groups = "drop") |>
  mutate(across(c(Q1, Median, Q3, Max), ~ fmt(.x, 3))) |>
  arrange(match(Method, METHODS)) |>
  kable(caption = "Reverse-keying eta-squared by method. High values mean the recovered partition is largely response polarity rather than heterogeneous covariance structure.")
```

```{r}
#| label: fig-rk
#| fig-cap: "Distribution of reverse-keying eta-squared across tables, by method."
rk |> filter(is.finite(eta2)) |>
  mutate(method = factor(method, levels = rev(METHODS))) |>
  ggplot(aes(eta2, method)) +
  geom_boxplot(outlier.alpha = 0.3, fill = "#deebf7") +
  geom_vline(xintercept = 0.5, linetype = 2, colour = "grey40") +
  labs(x = expression(eta^2 ~ "of PC1 polarity explained by class"), y = NULL) +
  theme_minimal(base_size = 11)
```

This is a caveat on what the recovered classes *are*, not a reason to discard
them. But it bears directly on Q3: a measure that scores well on the internal
criteria while also carrying a high polarity η² is winning at something narrower
than "finding heterogeneous covariance patterns."

## Transportability: how far outside their training design are we?

Their regressions were trained entirely on simulated matrices with **10 items and
100 respondents**. Item count was not among the parameters varied in their grid
search (their Table 3), so every one of the 198,000 training matrices was exactly
10 items wide.

```{r}
#| label: tbl-transport
mf |> distinct(table, .keep_all = TRUE) |>
  summarise(
    `Items: min` = min(n_items), `Items: median` = median(n_items),
    `Items: max` = max(n_items), `Items: % at 10` = sprintf("%.0f%%", 100*mean(n_items == 10)),
    `N: min` = min(n_resp), `N: median` = median(n_resp), `N: max` = max(n_resp)) |>
  mutate(across(everything(), as.character)) |>
  pivot_longer(everything(), names_to = "Quantity", values_to = "Value") |>
  kable(caption = "Our corpus against their training design of 10 items and 100 respondents.")
```

```{r}
#| label: fig-transport
#| fig-cap: "Item count and respondent count of the corpus. The dashed lines are their training design."
mf |> distinct(table, .keep_all = TRUE) |>
  select(n_items, n_resp) |>
  pivot_longer(everything()) |>
  mutate(name = recode(name, n_items = "Items", n_resp = "Respondents")) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 30, fill = "#4292c6") +
  geom_vline(data = data.frame(name = c("Items","Respondents"), v = c(10, 100)),
             aes(xintercept = v), linetype = 2, colour = "grey30") +
  facet_wrap(~ name, scales = "free") +
  labs(x = NULL, y = "Tables") + theme_minimal(base_size = 11)
```

This is a stated limitation, not something we correct for. Two of the fourteen
meta-features (`PercentOut`, `intrinsicDimProp`) are explicitly normalised by
item count, and several others — `VarColPC1` and `OverallSD` in particular — are
scale-dependent in ways that interact with the response-category mix. Applying
coefficients fitted on 10 × 100 matrices to a 40 × 2,000 matrix is an
extrapolation in both dimensions at once, and nothing in the released artifacts
bounds the error that introduces.

## Limitations

- The paper's simulation training data is not distributed and the fitted models
  were stripped of their model frames, so we cannot check whether a table's
  meta-features fall inside the *training* envelope — only inside the envelope
  their nine empirical datasets happened to span.
- The noise-assumption models fail to load, so the paper's claim that the
  ensemble increasingly wins as assumed error variance rises is untested here.
- Community detection is Louvain throughout, hardcoded in their code. The paper
  itself reports that this choice accounted for a meaningful share of measured
  improvement, so it is a fixed and consequential parameter, not a neutral one.
- The five validity criteria are computed on correlation-based distances and
  therefore favour correlation-based measures. This is their construction and
  their stated caveat.
- `r sum(corpus$affect_cog == "untagged")` of `r nrow(corpus)` drawn tables lack a
  construct tag, so construct-type splits rest on a smaller base than the corpus.
- Tables above 50,000 respondents were excluded from the sampling frame on
  download cost. This is a pragmatic restriction, not a statistical one.
- Two 40-item tables (`hexaco_ashton_2014_c`, `hexaco_ashton_2014_h`) were
  dropped after the draw, again on compute cost — at the full N = 2,000 cap they
  ran to roughly 19 core-hours each, about a quarter of the whole run between
  them. They are not replaced, because backfilling from the frame would swap
  wide tables for narrow ones and bias the corpus on exactly the dimension the
  transportability section is about. The practical effect is that the corpus
  tops out at `r max(corpus$n_items)` items rather than 40, which makes the
  extrapolation from their 10-item training design slightly *less* severe than
  the eligible frame would have allowed.

## Reproducibility

```{r}
#| output: asis
## components/source_links.R hardcodes the vignettes/ directory; this page lives
## in analysis/, so the links are emitted here against the same URL pattern.
files <- c("plenitude_replication.qmd", "plenitude_replication_compute.R",
           "plenitude_lib.R", "plenitude_data_prep.R", "plenitude_sample.R",
           "plenitude_collate.R")
files <- files[file.exists(files)]
cat("\n**Source code for this page:** ",
    paste(sprintf("[`%s`](https://github.com/datapages/irw/blob/main/analysis/%s)", files, files),
          collapse = " \u00b7 "), "\n\n", sep = "")
```

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate, from
the project root:

```r
# 1. Draw the corpus (deterministic given the seed)
source("analysis/plenitude_sample.R")

# 2. Fetch, fit and collate. Skip-if-exists at the table x replicate level,
#    so the run is resumable. Needs REDIVIS_API_TOKEN for the fetch stage.
#    ~140 core-hours; PLENITUDE_WORKERS controls parallelism.
source("analysis/plenitude_replication_compute.R")

# 3. Re-render this page
quarto::quarto_render("analysis/plenitude_replication.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(unique(read.csv("analysis/plenitude_data/corpus.csv")$table),
                output_file = "analysis/plenitude_data/references.bib")
```
:::

---

## References

::: {#refs}
:::
