---
title: "Extending Algorithm Selection for Class Detection to Item Response Data"
description: |
  Sociologists use several methods to identify groups of survey respondents who
  share a cultural schema. These methods often produce different results,
  leaving researchers with little basis for choosing among them. Sotoudeh and
  DiMaggio (2021) address this problem by using 198,000 simulated datasets to
  predict which method will work best on a given dataset. We apply their
  released code and models to 117 item-response tables from the Item Response
  Warehouse. Their predictions agree with the internal validity criteria about
  as well as they did on the original survey modules, and their assessments of
  particular measures largely hold. The prediction models are less useful for
  choosing among measures: they recommend the same measure for most tables. One
  possible explanation is that all the datasets used to train the models had
  exactly ten items.
date: today
bibliography: plenitude_data/references.bib
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(knitr)
})

res    <- readRDS("plenitude_data/plenitude_results.rds")
corpus <- read.csv("plenitude_data/corpus.csv", stringsAsFactors = FALSE)
table9 <- read.csv("vendor/table9.csv", stringsAsFactors = FALSE)
date_run <- file.info("plenitude_data/plenitude_results.rds")$mtime

mf <- res$metafeatures; pr <- res$predicted; so <- res$solutions
va <- res$validity;     rk <- res$reverse_key

n_tables <- dplyr::n_distinct(mf$table)
n_fits   <- res$n_fits

METHODS <- c("eJaccard","Podani","Euclidean","Cosine","eDice","Correlation",
             "ACE","Original Relationality","Recursive Relationality","Ensemble")
CRITERIA <- c(pearsongamma = "Pearson's Gamma", within_ss = "Within-class SS",
              wb_ratio = "Within-between ratio", ch = "Calinski-Harabasz",
              aic_improvement = "SEM AIC improvement")
## All five criteria are higher-is-better as computed here. The four partition
## criteria come from fpc::cluster.stats() run on |r|, a SIMILARITY matrix, so a
## tighter partition has a larger within-class SS and within-between ratio, not
## a smaller one. check_criteria_orientation() in the compute script asserts
## this on planted classes. An earlier draft treated within_ss and wb_ratio as
## lower-is-better, which reversed their sign in Q1 and their winners in Q2/Q3.
LOWER_BETTER <- character(0)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
```

```{r}
#| label: run-status
#| output: asis
n_drawn   <- nrow(corpus)
cov       <- mf |> count(table, name = "reps")
full_reps <- sum(cov$reps == 3)
## A run is complete when every drawn table x replicate has been attempted --
## whether it produced a usable matrix or was excluded for cause.
n_excl    <- if (!is.null(res$excluded)) nrow(res$excluded) else 0
attempted <- if (!is.null(res$n_attempted)) res$n_attempted else n_fits
complete  <- attempted >= 3 * n_drawn && all(cov$reps == 3)
if (!complete) {
  cat(sprintf(
"::: {.callout-important}
## Draft: partial run

The compute run is still in progress. Every number on this page is computed from
**%d of the %d drawn tables** (%d fits; %d tables have all three replicates,
%d have fewer). Tables were fitted cheapest-first, so the coverage is *not* a
random subset of the corpus — it is biased toward tables with few items and few
respondents, and the widest and largest tables are disproportionately still
outstanding. Treat every quantity below as provisional, and read the method
comparisons in particular with care: the measures whose cost scales steeply in
item count are being judged here mostly on narrow tables.
:::

",
    nrow(cov), n_drawn, n_fits, full_reps, nrow(cov) - full_reps))
}
```

## Background

Since Goldberg [-@goldberg2011mapping] introduced relational class analysis,
sociologists have developed related methods for identifying groups of
respondents who share a cultural schema. Members of these groups organize
attitudes in similar ways, even when they differ in the attitudes they hold
[@boutyline2017correlational]. The methods differ mainly in how they measure similarity
between respondents' answers. These differences can lead to quite different
groupings of the same respondents. Researchers often choose a method by
convention, leaving a potentially consequential choice largely unexamined.

Sotoudeh and DiMaggio [-@sotoudeh2021plenitude] propose a way to choose among
these methods. They simulated 198,000 response matrices with known group
structures and measured how accurately each of ten distance measures recovered
those groups. They then regressed accuracy on fourteen summary characteristics
of each matrix, which they call *meta-features*. These include skewness,
principal-component structure, and correlations between respondents. An analyst
can calculate the same features for a new dataset and use the fitted models to
predict which measure will work best. The authors applied the procedure to nine
modules from the General Social Survey and the American National Election
Studies. They assessed the predictions using five internal measures of cluster
quality.

Predicted accuracy correlated with these criteria at about .54 to .68 (their
Table 12). No measure was best on every dataset. Recursive relationality and
correlation performed well across settings, while Euclidean distance, cosine,
and Goldberg's original relationality measure generally performed less well.

The nine survey modules provide an initial test of the procedure. Its
usefulness also depends on how well it works with other kinds of data. The
authors' released code and fitted models allow us to examine that question.

We apply the procedure to `r n_tables` item-response tables from the Item
Response Warehouse (IRW). These tables vary more widely in the number of items,
sample size, and subject matter than the original survey modules. We ask three
questions: Do the predictions agree with the validity criteria? Does the best
measure vary across tables? Do the authors' assessments of particular measures
hold up? We retain the original approach and document the implementation
choices described below.

### The released materials

The authors' files are stored in a personal Google Drive folder without a DOI.
The linked GitHub repository contains only a README. We keep local copies in
`analysis/vendor/` and record their SHA-256 hashes so the analysis can be
reproduced if the original files change.

The files limit what we can test in two ways.

- **No prediction model for the ensemble.** `metafeature_models_cleaned.RDS`
  contains nine `lm` objects, one for each distance measure. It includes no
  prediction model for the pairwise-voting ensemble, which combines the other
  measures. We run all ten methods and calculate all five criteria for each, but
  can compare predicted accuracy for only nine.
- **No usable noise models.** `models_w_noise.RDS` contains regressions fitted
  under different assumptions about error variance. It downloads as a valid gzip
  file but cannot be opened with `readRDS`. The comparison in the paper's Table
  11 depends on these models, so we cannot reproduce it.

### Issues in the released code

We identified six issues in the released code (@tbl-defects). Four emerged
during the preliminary review, one when we applied the prediction models, and
one after a reader asked about the signs of two criteria in Q1. The table
records how we handled each issue.

```{r}
#| label: tbl-defects
data.frame(
  Issue = c("`num_vars` defaults to 10",
             "`PercentOut` scale",
             "`OverallRightKurt` on a logical vector",
             "Louvain hardcoded",
             "`make_prediction()` matches positionally",
             "Validity criteria computed on a similarity matrix"),
  What = c(
    "`evaluate.metafeatures()` defaults to `num_vars = 10`, and `select.method()` never overrides it. This value is the denominator of `PercentOut` and `intrinsicDimProp`.",
    "The released formula counts the items that contain outliers, divides by the number of items, and multiplies by 100, giving values from 0 to 100. The same column in their Table 9 runs from 0.000 to 0.048.",
    "`Kurt(df > mean(df))` is the kurtosis of a Bernoulli indicator. The function that seems to have been intended, `right_half_kurt`, is defined at line 217 but never called.",
    "`metaRCA()` fixes community detection at `cluster_louvain()` with no option to change it, although the paper notes that this choice matters.",
    "Coefficients are multiplied by meta-features by position. The column order of `evaluate.metafeatures()` does not match the order of the coefficients in the fitted models; 7 of 14 positions are swapped.",
    "`fpc::cluster.stats()` expects distances but is given `abs(cor(t(x)))`, which is a similarity. On it, a better partition has a *larger* within-class SS and within-between ratio, and negative Pearson's Γ and Calinski–Harabasz values, which the released code's `abs()` turns positive."),
  Handling = c(
    "We pass `num_vars = ncol(df)` explicitly and do not use `select.method()`.",
    "We compute both versions and compare them (@sec-appendix).",
    "We use the released version, since the coefficients were fitted on it, and report the corrected one in @sec-appendix.",
    "Kept as in the paper and treated as fixed.",
    "We match by name and report the released positional matching in @sec-appendix.",
    "Kept as in the paper, with all four criteria read as higher-is-better. The compute script checks this on planted classes before fitting.")
) |> kable(caption = "Places where the released code needs care, and how each is handled here.")
```

An 18-item table illustrates the first issue. The default calculation gives
`PercentOut = 130`, which is outside the possible range of a percentage. It also
gives `intrinsicDimProp = 0.400`, rather than 0.222. Passing the actual item
count through `num_vars` corrects both calculations.

The sixth issue does not change any criterion value; it changes how the values
should be read. With two planted classes, the true partition scores higher than
a shuffled one on all four partition criteria as the released code computes
them. The compute script now runs this check before fitting and stops if any
criterion fails it.

## Data and methods

### Corpus

```{r}
#| label: tbl-strata
corpus |>
  count(affect_cog, catband = cut(n_categories, c(3,5,7,1e6),
                                  labels = c("4-5","6-7","8+"))) |>
  pivot_wider(names_from = catband, values_from = n, values_fill = 0) |>
  rename(`Construct tag` = affect_cog) |>
  kable(caption = "Realized stratification of the drawn corpus, on response-category count and the affective/cognitive tag.")
```

```{r}
#| label: frame
elig <- read.csv("scout_eligible_tables.csv")
n_big <- sum(elig$n_participants > 50000)
```

The sampling frame contains `r nrow(elig)` IRW tables that meet four criteria:
`density == 1`, 6–40 items, at least 300 respondents, and at least four response
categories. We excluded `r n_big` tables with more than 50,000 respondents
because of download costs. We then drew `r nrow(corpus)` tables, stratified by
response-category count and construct type (@tbl-strata).
`r sum(corpus$affect_cog == "untagged")` sampled tables have no construct tag,
so comparisons by construct type cover a smaller sample.

One table, `sd3ypl_klimczak_2019_sd3`, has no complete cases, leaving
`r n_tables` tables for analysis. IRW lists it as complete, but the density flag
records whether a response row exists, not whether it contains a value. Of its
18,980 responses, 1,460 are `NA`. The missing values are concentrated in a few
items, but every one of the 730 respondents is missing at least one answer. The
authors' `grab_measurements()` function uses complete-case filtering, which
removes all respondents. An IRW density of one therefore does not guarantee
complete responses.

We analyse each table separately. Some IRW tables share respondents and could be
combined, but doing so produced substantial reverse-keying problems in
preliminary runs (see below). Combining tables also requires consistent
respondent IDs. We could confirm this for only 24 of 114 candidate groups.

### Subsampling

```{r}
#| label: n-shortfall
n_short <- sum(!corpus$n_participants >= 2000)
```

We use at most 2,000 respondents per table and run three replicates with seeds
101, 202, and 303. Several meta-features depend on sample size. The cap limits
differences between our tables and the authors' nine empirical datasets, which
each had roughly 1,500–2,500 respondents.

`r n_short` of the `r nrow(corpus)` sampled tables
(`r fmt(100*n_short/nrow(corpus), 0)`%) have fewer than 2,000 respondents, so we
use all their respondents. For these tables, the three replicates contain the
same people. Any differences in their results arise from stochastic community
detection and reflect algorithmic stability. For larger tables, the replicates
also reflect sampling variability. The `N_capped` field in
`plenitude_data/metafeatures.csv` identifies tables affected by the cap.

### Procedure

For each table and replicate, we calculate the fourteen meta-features and use
the nine fitted models to predict each measure's accuracy. We then apply the
nine measures and the ensemble to identify classes. Finally, we calculate the
five internal validity criteria. Following the original paper, these criteria
are computed from absolute response correlations between respondents.

::: {.callout-warning}
## A caveat from the original paper
The criteria favor correlation-based measures, including correlation, cosine,
and ACE, because the criteria are themselves computed from correlations. The
authors acknowledge this limitation. Agreement between the predictions and the
criteria may therefore partly reflect their shared basis in correlation.
:::

The fifth criterion compares a saturated covariance model fitted separately
within each class with a single pooled model. It records the improvement in AIC.
Classes with too few respondents relative to the number of items can prevent
estimation, and these methods often produce small classes. We retain the
authors' code, which uses `try()` and returns `NA` if estimation fails. As a
supplement, we report a trimmed version that combines all classes with fewer
than p + 1 respondents into one group, where p is the number of items.

## Results

```{r}
#| label: run-scale
tot_hours <- sum(so$secs, na.rm = TRUE) / 3600
```

All `r n_fits` runs (`r n_tables` tables × 3 replicates) finished, using about
`r fmt(tot_hours, 0)` core-hours of clustering.

### Predicted accuracy and the scale of PercentOut

```{r}
#| label: readings-summary
below0 <- function(r) mean(pr$predicted[pr$reading == r] < 0, na.rm = TRUE)
above1 <- function(r) mean(pr$predicted[pr$reading == r] > 1, na.rm = TRUE)
```

The intended scale of `PercentOut` is unclear. The released formula returns
values on a 0–100 scale, and using these values with the fitted models produces
predicted accuracies below zero in `r fmt(100*below0("primary"), 0)`% of cases.
The alternative, the proportion of outlying cells, is closer in magnitude to the
values in the authors' Table 9. Under it, no predictions fall below zero,
although `r fmt(100*above1("pct_cells"), 0)`% exceed one. These results suggest a
mismatch between the released `evaluate.metafeatures()` calculation and the
inputs expected by the prediction models. We use the cell-proportion
specification (`pct_cells`) for the remaining analysis, though we cannot confirm
that it matches the definition used in the training data. The choice matters:
@sec-appendix shows that it, the `OverallRightKurt` correction, and positional
coefficient matching each change which measure is predicted to be best in most
tables.

```{r}
#| label: tbl-oor
mf_long <- mf |> select(table, replicate, all_of(c("OverallSD","OverallKurt","OverallRightKurt",
  "VarColPC1","SkewColPC1","SkewColPC2","Skewness","Mean_Row_Corr","Row_Corr_Kurtosis",
  "CorrRightKurt","Mean_Col_Kurtosis","Mean_SD_Ratio","intrinsicDemnsionalityProp"))) |>
  pivot_longer(-c(table, replicate), names_to = "feature", values_to = "value")
t9 <- table9 |> select(-dataset) |>
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  group_by(feature) |> summarise(lo = min(value), hi = max(value), .groups = "drop")
oor <- mf_long |> inner_join(t9, by = "feature") |>
  mutate(outside = value < lo | value > hi) |>
  group_by(feature) |>
  summarise(`% outside their Table 9 range` = sprintf("%.0f%%", 100*mean(outside)), .groups = "drop") |>
  arrange(desc(`% outside their Table 9 range`))
kable(oor, caption = "How often IRW meta-feature values fall outside the range their nine datasets spanned. Their simulation training data is not distributed, so Table 9 is the closest available reference envelope.")
```

The simulated data were not released, and the fitted models were saved without
their model frames. We therefore cannot compare our tables directly with the
training data. The authors' nine empirical datasets (their Table 9) provide the
closest available reference. Skewness falls outside the reported range for more
than half of our tables. The mean correlation between respondents falls outside
it for more than a third. Values within these ranges indicate similarity to the
GSS and ANES modules, but not necessarily to the simulations.

### Q1. Do the predictions agree with the validity criteria?

Across their nine datasets, Sotoudeh and DiMaggio found correlations of about
.54 to .68 between predicted accuracy and the five internal criteria (their
Table 12).

Pooling results across our tables requires care. Within-class sum of squares and
the Calinski–Harabasz index depend on the numbers of respondents and items. Our
tables range from `r min(corpus$n_items)` to `r max(corpus$n_items)` items and
include up to 2,000 respondents per run. A pooled correlation may therefore
reflect differences between tables as well as differences between methods. To
focus on comparisons among methods for a given table, we calculate correlations
within each table and average them across tables. We also report pooled
correlations for comparison with the published figures.

```{r}
#| label: q1
d <- pr |> filter(reading == "pct_cells") |>
  inner_join(va, by = c("table","replicate","method"))

## sign convention: a positive correlation always means "predicted accuracy
## agrees with the criterion" (LOWER_BETTER is empty; see setup)
orient <- function(cr) if (cr %in% LOWER_BETTER) -1 else 1

q1 <- lapply(names(CRITERIA), function(cr) {
  dd <- d |> mutate(val = orient(cr) * .data[[cr]]) |> filter(is.finite(val), is.finite(predicted))
  if (nrow(dd) < 10) return(NULL)

  ## pooled, as their Table 12 reads
  ct <- cor.test(dd$predicted, dd$val)

  ## within-table, then summarised across tables
  wt <- dd |> group_by(table) |>
    filter(n() >= 5, sd(val) > 0, sd(predicted) > 0) |>
    summarise(r = cor(predicted, val), .groups = "drop")
  ci <- if (nrow(wt) > 2) t.test(wt$r)$conf.int else c(NA, NA)

  data.frame(Criterion = CRITERIA[[cr]],
             `Tables` = nrow(wt),
             `Mean within-table r` = mean(wt$r),
             `lo` = ci[1], `hi` = ci[2],
             `Pooled r` = unname(ct$estimate),
             `plo` = ct$conf.int[1], `phi` = ct$conf.int[2],
             check.names = FALSE)
}) |> bind_rows()

q1 |>
  transmute(Criterion, Tables,
            `Mean within-table r [95% CI]` = sprintf("%s [%s, %s]",
              fmt(`Mean within-table r`), fmt(lo), fmt(hi)),
            `Pooled r [95% CI]` = sprintf("%s [%s, %s]",
              fmt(`Pooled r`), fmt(plo), fmt(phi))) |>
  kable(caption = "Q1. Agreement between predicted accuracy and each internal validity criterion. All five criteria are higher-is-better, so positive means agreement. Their nine-dataset values were .54-.68.")
```

```{r}
#| label: fig-q1-dist
#| fig-cap: "Distribution of the within-table correlation between predicted accuracy and each criterion. One value per table."
lapply(names(CRITERIA), function(cr) {
  d |> mutate(val = orient(cr) * .data[[cr]]) |>
    filter(is.finite(val), is.finite(predicted)) |>
    group_by(table) |> filter(n() >= 5, sd(val) > 0, sd(predicted) > 0) |>
    summarise(r = cor(predicted, val), .groups = "drop") |>
    mutate(criterion = CRITERIA[[cr]])
}) |> bind_rows() |>
  ggplot(aes(r)) +
  geom_histogram(bins = 25, fill = "#4292c6") +
  geom_vline(xintercept = 0, colour = "grey30") +
  facet_wrap(~ criterion) +
  labs(x = "Within-table correlation of predicted accuracy with the criterion",
       y = "Tables") +
  theme_minimal(base_size = 11)
```

```{r}
#| label: fig-q1
#| fig-cap: "Predicted accuracy against each validity criterion, pooled. One point per table x replicate x method. The spread between tables on the two scale-dependent criteria is why we summarise within tables."
d |> select(predicted, all_of(names(CRITERIA))) |>
  pivot_longer(-predicted, names_to = "criterion", values_to = "value") |>
  filter(is.finite(value)) |>
  mutate(criterion = CRITERIA[criterion]) |>
  ggplot(aes(value, predicted)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE, colour = "#2c7fb8") +
  facet_wrap(~ criterion, scales = "free_x") +
  labs(x = "Validity criterion", y = "Predicted accuracy") +
  theme_minimal(base_size = 11)
```

```{r}
#| label: q1-readout
q1r <- setNames(q1$`Mean within-table r`, q1$Criterion)
```

The predictions agree with the criteria about as well as they did in the
original study. The mean within-table correlations for within-class sum of
squares (`r fmt(q1r[["Within-class SS"]])`), the within-between ratio
(`r fmt(q1r[["Within-between ratio"]])`), Pearson's Γ
(`r fmt(q1r[["Pearson's Gamma"]])`), and Calinski–Harabasz
(`r fmt(q1r[["Calinski-Harabasz"]])`) all fall within the published range.
The SEM correlation is also positive, but weaker
(`r fmt(q1r[["SEM AIC improvement"]])`). As the warning above notes, some of this
agreement may reflect the shared basis in correlation.

The number of classes each method identifies may also contribute. Partition-based
criteria depend partly on how many classes a partition contains. @tbl-kconfound
reports the associations between class count and the five criteria.

```{r}
#| label: tbl-kconfound
kj <- so |> select(table, replicate, method, k) |>
  inner_join(va, by = c("table","replicate","method"))

krow <- function(cr) {
  w <- kj |> mutate(v = .data[[cr]]) |> filter(is.finite(v)) |>
    group_by(table, replicate) |> filter(n() >= 5, sd(k) > 0, sd(v) > 0) |>
    summarise(rho = cor(k, v, method = "spearman"), .groups = "drop")
  data.frame(Criterion = CRITERIA[[cr]], `rho with k` = mean(w$rho), check.names = FALSE)
}
kt <- bind_rows(lapply(names(CRITERIA), krow))

kp <- pr |> filter(reading == "pct_cells") |>
  inner_join(so |> select(table, replicate, method, k), by = c("table","replicate","method")) |>
  group_by(table, replicate) |> filter(sd(k) > 0, sd(predicted) > 0) |>
  summarise(rho = cor(predicted, k, method = "spearman"), .groups = "drop")
rho_pred_k <- mean(kp$rho)

kt |> mutate(`rho with k` = fmt(`rho with k`)) |>
  kable(caption = "Mean within-run Spearman correlation between the number of recovered classes and each criterion. All five criteria are higher-is-better, so a positive value means the criterion rewards finer partitions.")
```

Methods with higher predicted accuracy tend to identify more classes (mean
within-run ρ = `r fmt(rho_pred_k)`). Four of the five criteria also favor
partitions with more classes; Calinski–Harabasz is the exception and favors
fewer. Agreement on those four may therefore partly reflect a shared preference
for finer partitions. Agreement on Calinski–Harabasz runs against that tendency,
so it is not explained by class count.

We retain the published comparisons without adjusting for class count. These
associations do not tell us how much of the agreement class count explains.

For every criterion, the pooled correlation is smaller than the mean
within-table correlation. The two summaries therefore give different impressions
of agreement. The within-table correlations address our question more directly:
how do the methods compare for a given dataset?

### Q2. Does the best measure vary across tables?

```{r}
#| label: q2
top_pred <- pr |> filter(reading == "pct_cells") |>
  group_by(table, replicate) |> slice_max(predicted, n = 1, with_ties = FALSE) |>
  ungroup() |> transmute(table, replicate, basis = "Predicted accuracy", winner = method)

top_val <- lapply(names(CRITERIA), function(cr) {
  v <- va |> select(table, replicate, method, val = all_of(cr)) |> filter(is.finite(val))
  if (nrow(v) == 0) return(NULL)
  if (cr %in% LOWER_BETTER) v <- v |> group_by(table, replicate) |> slice_min(val, n = 1, with_ties = FALSE)
  else                      v <- v |> group_by(table, replicate) |> slice_max(val, n = 1, with_ties = FALSE)
  v |> ungroup() |> transmute(table, replicate, basis = CRITERIA[[cr]], winner = method)
}) |> bind_rows()

wins <- bind_rows(top_pred, top_val)
wins |> count(basis, winner) |>
  group_by(basis) |> mutate(pct = sprintf("%.0f%%", 100*n/sum(n))) |>
  select(-n) |> pivot_wider(names_from = basis, values_from = pct, values_fill = "-") |>
  arrange(match(winner, METHODS)) |> rename(Method = winner) |>
  kable(caption = "Q2. Share of table x replicate runs each method wins, by basis. Predicted accuracy covers nine methods; the criteria cover all ten.")
```

```{r}
#| label: q2-concentration
conc <- wins |> count(basis, winner) |> group_by(basis) |>
  summarise(top_share = max(n)/sum(n), top_method = winner[which.max(n)],
            n_distinct_winners = n(), .groups = "drop")
```

```{r}
#| label: tbl-q2-conc
conc |> mutate(top_share = sprintf("%.0f%%", 100*top_share)) |>
  rename(Basis = basis, `Top method` = top_method,
         `Its share of wins` = top_share, `Distinct winners` = n_distinct_winners) |>
  kable(caption = "Q2. Concentration of wins. A high share means that basis consistently favours a single measure.")
```

```{r}
#| label: q2-readout
cs <- function(b) conc |> filter(basis == b)
pct <- function(x) sprintf("%.0f%%", 100*x)
pred_top <- cs("Predicted accuracy")
corr_bases <- conc |> filter(basis != "Predicted accuracy", top_method == "Correlation")
PROSE <- c(`Pearson's Gamma` = "Pearson's Γ", `Within-class SS` = "within-class SS",
           `Within-between ratio` = "the within-between ratio")
```

Recursive relationality has the highest predicted accuracy in
`r pct(pred_top$top_share)` of runs, and only `r pred_top$n_distinct_winners`
measures ever rank first. The models thus recommend the same measure for most of
the `r n_tables` tables. The differences between these tables and the 10-item,
100-respondent training matrices may help explain this result, as discussed
below.

The validity criteria also concentrate, but on a different measure. Correlation
ranks first under `r c("one","two","three","four","five")[nrow(corr_bases)]` of the five criteria:
`r paste(sprintf("%s of runs under %s", pct(corr_bases$top_share), PROSE[corr_bases$basis]), collapse = ", ")`.
The ensemble ranks first in `r pct(cs("Calinski-Harabasz")$top_share)` of runs
under Calinski–Harabasz. Only the SEM criterion is widely spread: its leading
method, `r cs("SEM AIC improvement")$top_method`, accounts for
`r pct(cs("SEM AIC improvement")$top_share)` of runs, and
`r cs("SEM AIC improvement")$n_distinct_winners` methods rank first at least
once. Correlation's strong showing needs to be read in light of the
correlation-based criteria.

The best measure therefore varies across tables, but less than the original
study suggested. Even under the most concentrated criterion, a different measure
wins in about a third of runs, so no measure is best everywhere. But on four of
the five criteria a single measure wins most runs, and three of those favor the
measure that the criteria are built to favor. The predictions and the criteria
also disagree about which measure that is: the models favor recursive
relationality, while the criteria mostly favor correlation.

### Q3. Do the authors' assessments of particular measures hold?

```{r}
#| label: q3
rank_within <- function(df, col, lower_better) {
  df |> group_by(table, replicate) |>
    mutate(rk = if (lower_better) rank(.data[[col]], na.last = "keep")
                else rank(-.data[[col]], na.last = "keep")) |> ungroup()
}
ranks <- lapply(names(CRITERIA), function(cr) {
  va |> select(table, replicate, method, val = all_of(cr)) |> filter(is.finite(val)) |>
    rank_within("val", cr %in% LOWER_BETTER) |>
    transmute(method, criterion = CRITERIA[[cr]], rk)
}) |> bind_rows()

q3 <- ranks |> group_by(method) |>
  summarise(`Mean rank across criteria` = mean(rk, na.rm = TRUE),
            `Top-3 rate` = mean(rk <= 3, na.rm = TRUE), .groups = "drop") |>
  left_join(pr |> filter(reading == "pct_cells") |> group_by(method) |>
              summarise(`Mean predicted accuracy` = mean(predicted), .groups = "drop"),
            by = "method") |>
  arrange(`Mean rank across criteria`)

q3 |> mutate(across(where(is.numeric), ~ fmt(.x, 3))) |>
  rename(Method = method) |>
  kable(caption = "Q3. Mean rank across the five validity criteria (lower is better; ten methods so 5.5 is chance), the rate of finishing top-3, and mean predicted accuracy. The ensemble has no fitted model.")
```

```{r}
#| label: q3-readout
best <- q3$method[1]; worst <- q3$method[nrow(q3)]
rng  <- range(q3$`Mean rank across criteria`)
mr   <- setNames(q3$`Mean rank across criteria`, q3$method)
t3   <- setNames(q3$`Top-3 rate`, q3$method)
```

Mean ranks range from `r fmt(rng[1])` to `r fmt(rng[2])`, compared with an
expected rank of 5.5 under random ordering, so the criteria separate the methods
clearly. `r best` ranks best on average and `r worst` worst.

The authors' assessments largely hold. Correlation (mean rank
`r fmt(mr[["Correlation"]])`, top three in `r pct(t3[["Correlation"]])` of
runs) and recursive relationality (`r fmt(mr[["Recursive Relationality"]])`)
both perform well, and recursive relationality also has the highest mean
predicted accuracy. Euclidean distance (`r fmt(mr[["Euclidean"]])`) and original
relationality (`r fmt(mr[["Original Relationality"]])`) are among the weakest.
Cosine sits near the middle (`r fmt(mr[["Cosine"]])`), close to chance
rather than clearly poor. The ensemble also performs well
(`r fmt(mr[["Ensemble"]])`), though the released models provide no prediction of
its accuracy. Correlation's first place should be read with the same caveat as
in Q2.

### Class solutions and degeneracy

```{r}
#| label: tbl-degen
so |> filter(status == "ok") |>
  group_by(Method = method) |>
  summarise(`median k` = median(k), `median substantive k` = median(n_substantive),
            `median largest class` = sprintf("%.0f%%", 100*median(largest_prop)),
            .groups = "drop") |>
  arrange(match(Method, METHODS)) |>
  kable(caption = "Recovered class solutions.")
```

```{r}
#| label: degen-rate
degen_rate <- mean(so$degenerate[so$status == "ok"], na.rm = TRUE)
err_rate   <- mean(so$status != "ok")
```

`r if (degen_rate == 0 && err_rate == 0) "No run produced a degenerate solution (a single class, or one class holding more than 90% of respondents), and none failed." else sprintf("%s%% of method runs produced a degenerate solution (a single class, or one class holding more than 90%% of respondents) and %s%% failed. Degenerate solutions are retained in every table above.", fmt(100*degen_rate, 1), fmt(100*err_rate, 1))`

## The reverse-keying diagnostic

Recovered classes may separate respondents by response polarity rather than by
distinct covariance patterns. For each table and method, we assess how closely
class membership corresponds to respondents' alignment with the signs of the
first principal-component loadings. We summarize this relationship as the share
of variance in PC1 polarity scores explained by class membership (η²).

```{r}
#| label: tbl-rk
rk |> filter(is.finite(eta2)) |> group_by(Method = method) |>
  summarise(Q1 = quantile(eta2, .25), Median = median(eta2), Q3 = quantile(eta2, .75),
            Max = max(eta2), `share > 0.5` = sprintf("%.0f%%", 100*mean(eta2 > 0.5)),
            .groups = "drop") |>
  mutate(across(c(Q1, Median, Q3, Max), ~ fmt(.x, 3))) |>
  arrange(match(Method, METHODS)) |>
  kable(caption = "Reverse-keying eta-squared by method. High values mean the classes mostly separate respondents by the direction of their answers.")
```

```{r}
#| label: fig-rk
#| fig-cap: "Distribution of reverse-keying eta-squared across tables, by method."
rk |> filter(is.finite(eta2)) |>
  mutate(method = factor(method, levels = rev(METHODS))) |>
  ggplot(aes(eta2, method)) +
  geom_boxplot(outlier.alpha = 0.3, fill = "#deebf7") +
  geom_vline(xintercept = 0.5, linetype = 2, colour = "grey40") +
  labs(x = expression(eta^2 ~ "of PC1 polarity explained by class"), y = NULL) +
  theme_minimal(base_size = 11)
```

eJaccard and cosine show the strongest associations with response polarity.
Their η² exceeds .5 in about a third and a quarter of tables, respectively. This
does not establish that the classes lack substantive meaning, but it affects
their interpretation. When class membership closely follows response polarity, a
high validity score provides limited evidence that the method has identified
distinct covariance patterns.

## How far the tables are from the training data

Every one of the 198,000 simulated matrices used to train the models had 10
items and 100 respondents. Item count was not among the parameters the authors
varied (their Table 3).

```{r}
#| label: tbl-transport
mf |> distinct(table, .keep_all = TRUE) |>
  summarise(
    `Items: min` = min(n_items), `Items: median` = median(n_items),
    `Items: max` = max(n_items), `Items: % at 10` = sprintf("%.0f%%", 100*mean(n_items == 10)),
    `N: min` = min(n_resp), `N: median` = median(n_resp), `N: max` = max(n_resp)) |>
  mutate(across(everything(), as.character)) |>
  pivot_longer(everything(), names_to = "Quantity", values_to = "Value") |>
  kable(caption = "Our corpus against their training design of 10 items and 100 respondents.")
```

```{r}
#| label: fig-transport
#| fig-cap: "Item count and respondent count of the corpus. The dashed lines are their training design."
mf |> distinct(table, .keep_all = TRUE) |>
  select(n_items, n_resp) |>
  pivot_longer(everything()) |>
  mutate(name = recode(name, n_items = "Items", n_resp = "Respondents")) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 30, fill = "#4292c6") +
  geom_vline(data = data.frame(name = c("Items","Respondents"), v = c(10, 100)),
             aes(xintercept = v), linetype = 2, colour = "grey30") +
  facet_wrap(~ name, scales = "free") +
  labs(x = NULL, y = "Tables") + theme_minimal(base_size = 11)
```

The IRW tables differ from the training matrices in both item count and sample
size. These differences may help explain the results, although we cannot
establish how much they matter. Two meta-features, `PercentOut` and
`intrinsicDimProp`, are divided by item count. Others, including `VarColPC1` and
`OverallSD`, depend on scale and the number of response categories.

Applying models trained on 10-item, 100-respondent matrices to tables with up to
`r max(corpus$n_items)` items and 2,000 respondents therefore requires
extrapolation. The released materials do not allow us to quantify the resulting
prediction error. Simulations covering a wider range of item counts and sample
sizes would allow a more direct test of the procedure for these data.

## Conclusion

The results largely support Sotoudeh and DiMaggio's assessments. Predicted
accuracy agrees with four of the five validity criteria about as well as it did
on the original survey modules. The measures they recommended, recursive
relationality and correlation, rank near the top, and Euclidean distance and
original relationality rank near the bottom. Support for their broader finding
that no measure is best across datasets is weaker: the best measure still varies,
but four of the five criteria favor a single measure in most runs, three of them
correlation. The prediction models provide less useful guidance: they recommend
the same measure for most tables. The narrow simulation design may contribute to
this result, but its role remains uncertain. A wider simulation design and
clearer documentation of the code issues in @tbl-defects would help establish
when the procedure can guide method selection for item-response data.

## Limitations

- The simulated training data were not released, and the fitted models do not
  retain their model frames. We can compare our tables with the authors' nine
  empirical datasets, but cannot compare them directly with the data used to
  train the models.
- The noise-assumption models cannot be loaded, so we do not test the claim that
  the ensemble's advantage grows with error variance.
- The released code uses Louvain community detection throughout. The authors
  report that this choice affected performance, so our results are conditional
  on it.
- The validity criteria are computed from correlations and so favour
  correlation-based measures, as the authors acknowledge.
- `r sum(corpus$affect_cog == "untagged")` of the `r nrow(corpus)` sampled tables
  have no construct tag, so comparisons by construct type rest on fewer tables.
- One sampled table, `sd3ypl_klimczak_2019_sd3`, could not be analysed (see
  Corpus), so the results are based on `r n_tables` tables, not
  `r nrow(corpus)`.
- We excluded tables with more than 50,000 respondents because of download
  costs. This practical restriction limits the sampling frame.
- Two 40-item tables (`hexaco_ashton_2014_c` and `hexaco_ashton_2014_h`) were
  excluded after sampling because of computational costs. At 2,000 respondents,
  each required about 19 core-hours. We did not replace them because substitutes
  would have had fewer items, underrepresenting the widest tables. The analytic
  sample therefore contains at most `r max(corpus$n_items)` items per table,
  making extrapolation from the 10-item training matrices less extensive than the
  full sampling frame would permit.

## Appendix: implementation sensitivity {#sec-appendix}

```{r}
#| label: tbl-readings
## OverallRightKurt_corrected is undefined when every value above the mean is
## identical (Kurt of a constant), so the rk_corrected reading is summarised over
## its finite cases and the undefined count is reported rather than hidden.
pr |>
  group_by(reading) |>
  summarise(min = min(predicted, na.rm = TRUE),
            median = median(predicted, na.rm = TRUE),
            max = max(predicted, na.rm = TRUE),
            `below 0` = sprintf("%.0f%%", 100*mean(predicted < 0, na.rm = TRUE)),
            `above 1` = sprintf("%.0f%%", 100*mean(predicted > 1, na.rm = TRUE)),
            undefined = sum(!is.finite(predicted)),
            .groups = "drop") |>
  mutate(across(c(min, median, max), ~ fmt(.x, 3))) |>
  kable(caption = "Predicted accuracy under each reading of the ambiguous inputs. Accuracy is a proportion, so values outside [0,1] are impossible by construction.")
```

```{r}
#| label: winner-stability
winner_by <- function(reading) {
  pr |> filter(reading == !!reading) |>
    group_by(table, replicate) |>
    slice_max(predicted, n = 1, with_ties = FALSE) |>
    ungroup() |> select(table, replicate, winner = method)
}
w <- Reduce(function(a, b) full_join(a, b, by = c("table","replicate")),
            lapply(c("primary","pct_cells","rk_corrected","positional"), function(r)
              winner_by(r) |> rename(!!r := winner)))
agree <- sapply(c("primary","rk_corrected","positional"),
                function(r) mean(w[[r]] == w$pct_cells, na.rm = TRUE))
```

@tbl-readings compares predicted accuracy under four specifications: the
released `PercentOut` formula (`primary`), the cell-proportion `PercentOut` used
in the main analysis (`pct_cells`), the corrected `OverallRightKurt`
(`rk_corrected`), and the released positional matching of coefficients to
meta-features (`positional`). Under the released `PercentOut` formula,
`r fmt(100*below0("primary"), 0)`% of predictions fall below zero; under
positional matching, `r fmt(100*above1("positional"), 0)`% exceed one.

These choices substantially affect which measure is selected. Relative to
`pct_cells`, the same measure ranks first in only
`r fmt(100*agree[["primary"]], 0)`% of tables when we use the released
`PercentOut` calculation. Agreement falls to
`r fmt(100*agree[["rk_corrected"]], 0)`% when `OverallRightKurt` is corrected
and to `r fmt(100*agree[["positional"]], 0)`% when coefficients are matched by
position, as in the released code.

## Reproducibility

```{r}
#| output: asis
## components/source_links.R hardcodes the vignettes/ directory; this page lives
## in analysis/, so the links are emitted here against the same URL pattern.
files <- c("plenitude_replication.qmd", "plenitude_replication_compute.R",
           "plenitude_lib.R", "plenitude_data_prep.R", "plenitude_sample.R",
           "plenitude_collate.R")
files <- files[file.exists(files)]
cat("\n**Source code for this page:** ",
    paste(sprintf("[`%s`](https://github.com/datapages/irw/blob/main/analysis/%s)", files, files),
          collapse = " · "), "\n\n", sep = "")
```

Results were computed on `r format(date_run, "%B %d, %Y")`. To regenerate the
analysis, run the following commands from the project root:

```r
# 1. Draw the corpus (deterministic given the seed)
source("analysis/plenitude_sample.R")

# 2. Fetch, fit and collate. Skip-if-exists at the table x replicate level,
#    so the run is resumable. Needs REDIVIS_API_TOKEN for the fetch stage.
#    ~140 core-hours; PLENITUDE_WORKERS controls parallelism. Stops before
#    fitting if check_criteria_orientation() fails.
source("analysis/plenitude_replication_compute.R")

# 3. Re-render this page
quarto::quarto_render("analysis/plenitude_replication.qmd")
```

::: {.callout-tip}
To cite the datasets used, run:

```r
irw_save_bibtex(unique(read.csv("analysis/plenitude_data/corpus.csv")$table),
                output_file = "analysis/plenitude_data/references.bib")
```
:::

---

## References

::: {#refs}
:::
